### 2020-05-03 casecontrol.Rmd
scripts for case-control analysis and Rmarkdown report

---
title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'

author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol.pdf")

---

```{r functions, echo=FALSE}

pvalue.latex <- function(x, n=1, nexp=1) {
  sapply(x, function(z) {
    if (is.na(z)) {
        return(NA)
    } else if(z > 0.001) {
        return(signif(z, 1))
    } else {
    z <- sprintf("%.*E", 0, signif(z, n)) # default is 1 sig fig
#    z <- sprintf("%.*E", n, signif(z, n + 1))
    z <- as.numeric(unlist(strsplit(z, "E"))) # split z at E
    sprintf("\\ensuremath{%.*f\\times 10^{%0*d}}", 0, z[1], nexp, z[2])
                                        #    sprintf("\\ensuremath{%.*f\\times 10^{%0*d}}", n, z[1], nexp + 1, z[2])
    }
  })
}

## create and format a 95% confidence interval around an odds/hazard ratio
or.ci <- function(coeff, se, ndigits=2) {
  ci.lo <- exp(coeff - 1.96 * se)
  ci.up <- exp(coeff + 1.96 * se)
  sprintf("%.*f (%.*f, %.*f)", ndigits, exp(coeff), ndigits, ci.lo, ndigits, ci.up)
}

```
\newpage

# Background
Case series data from hospitals has suggested a preponderance of certain conditions including diabetes and cardiovascular disease, and elevated risks in those of non-European ethnic origin. Quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator. Furthermore it has not been possible to disentangle whether drug effects may underlie the apparent association with certain co-morbid groups.

In Scotland the existence of a laboratory database with nationwide coverage and a unique health care identifier on all health records has enabled us to ascertain all those with tests positive for SARS-CoV-2 and link to other health records.  Here we describe a population wide case-control study of test confirmed COVID-19 disease in Scotland. From this we report the demographic factors, co-morbidities and concurrent medications associated with being hospitalised for COVID-19 and for becoming a severe COVID-19 case i.e. either warranting critical care support or dying within 28 days of a positive test. This study is intended to inform policy on defining vulnerable groups, identify any drug classes that influence risk of severe disease, and lay a basis for evaluating the feasibility of risk-stratified control measures.

# Methods
## Data sources 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.  .This allowed all those with a positive viral nucleic acid test for SARS-CoV-2 to be ascertained, hereafter denoted as test positive cases.  At the time writing, serology testing results for auto-antibodies indicating prior infection with SARS-CoV-2 are not yet available.

The National Community Health Index (CHI) database assigns a unique health care identifier to everyone in the population, and is used in all care systems in Scotland.  Using the CHI identifier, records from ECOSS were linked to hospitalization and discharge data from the weekly updated Scottish Morbidity Record 01 and the daily updated RAPID-24 databases. Admission to critical care was ascertained from the national audit database of critical care units (SICSAG). Death registrations and cause of death were obtained from linkage to the National Register of Scotland.  All these databases have nationwide coverage. At the time of writing the national policy on testing has evolved since the start of the epidemic with testing initially being restricted to possible cases that required hospital admission but widening to a "four-pillar" approach that has increased testing in health care workers residential care home settings. Accordingly the probability of being tested has changed over time.  For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case. Note that this is an incidence density sampling design i.e. a control who becomes a case at a later point is retained in the control set. With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.

For all cases and controls co-morbid conditions, the Charlson Comorbidity Index and drug exposure data were available from i) a five year look back query of the Scottish Morbidity Record 01 for sets of International Classification of Disease (ICD)-10 codes for various conditions ii) querying the national prescribing records for exposure to various drug classes iii) linkage to the national diabetes register SCI-diabetes.

We initially examined history of a specific list of conditions and drug classes that have been advised as high risk conditions for COVID-19 by Public Health England, World Health organisation or the US Centres for Disease Control. ICD-10 codes for the conditions are listed in Supplementary Materials.

Ethnicity was ascertained from ethnic group coding on prior SMR records and from applying the ONOMAP algorithm to names available in the CHI database. All cases and controls were assigned to a quintile of the Scottish Index of Multiple Deprivation- this is a residential postcode based indicator of socioeconomic status.

## Case definition
We classified test-positive cases into three groups:

1. Severe or fatal COVID-19, defined by entering critical care in the Scottish Intensive Care Society and Audit Group database (SICSAG) OR death within 28 days of a positive nucleic acid test (regardless of the cause of death given on the death certificate).

2. Hospitalised but not severe as defined above

3. Not hospitalized

The focus on severe disease is relevant to policy makers because decisions about control measures have been based on projections of the requirement for critical care. By restricting the case definition to this group, which combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of test-positive cases, whatever triage procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

The proportion of fatal cases that were not tested can be examined in national mortality data ...

# Statistical methods
We used  the R statistical package. We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)

We tabulated the prevalence of co-morbidities and current drug use by case status.  We used conditional logistic regression to estimate the univariate odds associated with each covariate of interest conditioned on age sex and GP practice. 
We also report a conditional logistic regression with all covariates entered simultaneously.

We tested each effect for interaction with age and sex.

Models were run overall and by age band.

To select variables to be retained in the model, we used backward selection with a termination criterion that the reduction in deviance is less than the value corresponding to a threshold p-value of 0.01 
Data were split into a 70:30 training:test sets. 
We evaluated the performance of the resulting backward selection model for predicting severe COVID-19 disease in the test set. For this we used the expected information for discrimination as described and used previously

### The initial demographic vars of interest are ( give that age and sex is matched)

 ethnic group
 SIMD

### The initial comorbidities of interest are listed below

- Circulatory Disorders
  - Ischaemic heart disease
  - Other heart disease
  - Cerebrovascular Disease
  - Other circulatory system diseases
- Neurological diseases except inflammatory
  - Epilepsy
  - Mono and  polyneuropathies
  - Other neurological conditions
- Respiratory diseases
  -  Acute respiratory infections
  -  Asthma
  -  Other Chronic lower respiratory disease
- Tuberculosis 
- Connective tissue diseases
- Decompensated liver disease
- Kidney disease
  - Advanced Chronic kidney disease or RRT
  - Other chronic kidney disease
- Diabetes
  - type 1
  - type 2
- Malignant neoplasms
  - Lung cancers
  - Blood cancers
  - Other cancers
- Immunological diseases
  - HIV
  - Certain disorders involving the immune system
- Sickle cell disease
- Cystic fibrosis
- Organ transplantation other than kidney
- Treated hypertension (on any antihypertensive drugs in past year)
- Charlson comorbidity index

​
### Other recent health care exposure variables of interest are

- Any inpatient admission in past year and excluding the three weeks prior to testing 
- Any emergency admission in past year and excluding the three weeks prior to testing
- Any ICU/HDU/Admission in past year and excluding the three weeks prior to testing

### Initial drug exposures of interest are

- Any lipid  drug 
- Any anticoag or platelet  
- Any Antihypertensive drugs disaggregated to: 
  - Vasodilator antihypertensive drugs
  - Centrally-acting antihypertensive drugs
  - Adrenergic neurone blocking drugs
  - Alpha-adrenoceptor blocking drugs
  - Angiotensin-converting enzyme inhibitors
  - Angiotensin-II receptor antagonists
  - Renin inhibitors
  - Thiazides and related diuretics
  - Calcium-channel blocker
- Any prescribed NSAID
- Any cytotoxic or immunosuppressant drug 
- Any chloroquine drug or derivatives of chloroquine

Derivations ICD10 and BNF codes for these comorbidities / drugs are in the appendix

Further comorbidities will be included in future protocol versions

Note that at this point we are not including high risk of exposure as covariates ie HCW, residential care home, prisoner etc this will form part of future work.

# Information governance
This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data. Datasets were de-identified before analysis.

# Results
## Comparison of severe cases with others who tested positive 
We emphasize that because testing was used selectively over this period, those who tested positive and were never hospitalised are unlikely to be representative of never-hospitalized cases in the population.

## Association of demographic factors with severe disease
Table \ref{tab:demog} shows univariate associations of demographic factors with disease.  Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home (rate ratio 8.6).  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.

```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
pvalue <- pvalue.latex(univariate.demog[, 5])
ci <- or.ci(univariate.demog[, 1], univariate.demog[, 3]) 
table.demog.aug <- data.frame(table.demog,
                                   ci, 
                                   pvalue)
colnames(table.demog.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
rownames(table.demog.aug) <- replace.names(rownames(table.demog.aug))

pander(table.demog.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       split.cells=c(15, 15, 25, 8),
       caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")

```

## Associations of comorbid conditions with severe disease
In univariate logistic regression models, both Type 1 and Type 2 diabetes were associated with severe disease (Table \ref{tab:conditions}. There were strong associations with circulatory disease, respiratory disease, kidney disease and neurological disease.  No association of severe disease with cystic fibrosis was detected.  A possible explanation for this may be that this group was effectively shielded from infection in accordance with official guidance.

```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}

pvalue <- pvalue.latex(univariate.conditions[, 5])
ci <- or.ci(univariate.conditions[, 1], univariate.conditions[, 3]) 
table.conditions.aug <- data.frame(table.conditions,
                                   ci, 
                                   pvalue)
colnames(table.conditions.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
rownames(table.conditions.aug) <- replace.names(rownames(table.conditions.aug))

pander(table.conditions.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       split.cells=c(20, 20, 20, 7),
       caption="\\label{tab:conditions}Univariate association of severe disease with comorbid conditions")

```

Further analyses to adjust for co-morbidity ...

### Counts of co-morbid conditions

Table xx shows the distribution of numbers of the co-morbid conditions listed by WHO or CDC or PHE by age band in severe cases versus their controls. 
Across each age band there were people with severe COVID-19 who did not appear to have any of these co-morbid conditions.

## Associations of prescribed drugs with severe disease 
Table \ref{tab:drugs} shows that severe disease was associated with use of antihypertensive agents, anticoagulants, and lipid-lowering agents.  Use of ACE inhibitors or angiotensin receptor blockers was not associated with higher risk than use of other classes of antihypertensive agent.  Use of non-steroidal anti-inflammatory agents was weakly associated with severe disease.  Use of hydroxychloroquine was associated with severe disease.

Further analyses to adjust for co-morbidity ...

```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}

pvalue <- pvalue.latex(univariate.drugs[, 5])
ci <- or.ci(univariate.drugs[, 1], univariate.drugs[, 3]) 
table.drugs.aug <- data.frame(table.drugs,
                                   ci, 
                                   pvalue)
colnames(table.drugs.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
rownames(table.drugs.aug) <- replace.names(rownames(table.drugs.aug))

pander(table.drugs.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       split.cells=c(20, 20, 20, 7), 
       caption="\\label{tab:drugs}Univariate association of severe disease with drugs prescribed in last year")

```

## Examining those with no apparent co-morbid conditions

A more in depth focus on severe cases at younger ages

As shown in table  among severe case below age 50  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found. Among severe case below age 50-70  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.  However in those with none of these co morbid conditions xx% were on   drugs for xxx xxx and xxx.

## Multivariate analysis of risk factors for severe disease

A stepwise regression procedure was carried out with all demographic, morbidity and drug exposure variables, using alternating forward-backward selection.  The `r nrow(stepwise.all)` variables retained are shown in Table \ref{tab:stepwise}

```{r model, echo=FALSE, warning=FALSE, message=FALSE}

table.stepwise <- as.data.frame(stepwise.all[, c(1, 3, 5)])
colnames(table.stepwise) <- c("Coefficient (log rate ratio)", "Standard error", "p.value")
rownames(table.stepwise) <- replace.names(rownames(table.stepwise))

table.stepwise$p.value <- pvalue.latex(table.stepwise$p.value)

pander(table.stepwise, table.style="multiline",
       # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
	   digits=c(3, 3, 1), 
	   split.table="Inf",
       caption="\\label{tab:stepwise}Stepwise regression: variables retained in model for severe disease")

```

## Predictive performance of a risk classifier

The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.  Table \ref{tab:wevid} shows that the C-statistic was `r summary(stepwise.densities)[3]` and the expected information for discrimination was  `r summary(stepwise.densities)[5]` bits.

show incremental comparison of different models ...

```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}

pander(summary(stepwise.densities), table.style="multiline",
       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )

```

```{r plotwdists, echo=FALSE, fig.cap="Cross-validation of model chosen by stepwise regression: distributions of weight of evidence", out.width="0.8\\textwidth"}

plotWdists(stepwise.densities) + theme_grey(base_size = 18) + 
	theme(legend.title=element_blank())

```

### 2020-05-03 casecontrol.R

@@ -1,0 +1,490 @@
+## analysis script for case-control study
+
+replace.names <- function(varnames) {
+    names <- varnames
+    found <- names %in% lookup.names$varname
+    names[found] <- as.character(lookup.names$longname[match(names[found],
+                                                             lookup.names$varname)])
+    return(names)
+}
+
+traintest.fold <- function(train.data, test.data, start.model,
+                           lower.formula, upper.formula) { # stepwise clogit on training fold, predicted probs on test fold
+    start.model <- clogit(formula=upper.formula, data=train.data)
+    stepwise.model <- step(start.model,
+                           scope=list(lower=lower.formula, upper=upper.formula),
+                           direction="both", trace=-1, method="approximate")
+    ## use reference=sample, then normalize separately within each stratum
+    unnorm.p <- predict(object=stepwise.model, newdata=test.data,
+                        na.action="na.pass", 
+                        type="risk", reference="sample")
+    norm.predicted <- normalize.predictions(unnorm.p=unnorm.p,
+                                            stratum=test.data$stratum,
+                                            y=test.data$CASE)
+}
+
+nonmissing.obs <- function(x, varnames) { ## subset rows nonmissing for varnames
+    keep <- rep(TRUE, nrow(x))
+    for(j in 1:length(varnames)) {
+        keep[is.na(x[, match(varnames[j], colnames(x))])] <- FALSE
+    }
+    return(x[keep, ])
+}
+
+normalize.predictions <- function(unnorm.p, stratum, y) { # format a pvalue in latex
+    ## normalize probs so that they sum to 1 within each stratum
+    ## also compute prior probs
+    ## this is appropriate if there is only 1 case per stratum
+    unnorm.p <- as.numeric(unnorm.p)
+    stratum <- as.factor(as.integer(as.character(stratum)))
+    nonmissing <- !is.na(unnorm.p)
+
+    unnorm.p <- unnorm.p[nonmissing]
+    stratum <- stratum[nonmissing]
+    y <- y[nonmissing]
+
+    ## compute normalizing constant and prior for each stratum, then merge
+    norm.constant <- as.numeric(tapply(unnorm.p, stratum, sum))
[…]

### 2020-05-06 casecontrol.Rmd
a few edits

@@
 title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'
-
 author: 
@@
 
-pvalue.latex <- function(x, n=1, nexp=1) {
-  sapply(x, function(z) {
-    if (is.na(z)) {
-        return(NA)
-    } else if(z > 0.001) {
-        return(signif(z, 1))
-    } else {
-    z <- sprintf("%.*E", 0, signif(z, n)) # default is 1 sig fig
-#    z <- sprintf("%.*E", n, signif(z, n + 1))
-    z <- as.numeric(unlist(strsplit(z, "E"))) # split z at E
-    sprintf("\\ensuremath{%.*f\\times 10^{%0*d}}", 0, z[1], nexp, z[2])
-                                        #    sprintf("\\ensuremath{%.*f\\times 10^{%0*d}}", n, z[1], nexp + 1, z[2])
-    }
-  })
-}
-
-## create and format a 95% confidence interval around an odds/hazard ratio
-or.ci <- function(coeff, se, ndigits=2) {
-  ci.lo <- exp(coeff - 1.96 * se)
-  ci.up <- exp(coeff + 1.96 * se)
-  sprintf("%.*f (%.*f, %.*f)", ndigits, exp(coeff), ndigits, ci.lo, ndigits, ci.up)
-}
+source("helperfunctions.R")
 
 ```
+
 \newpage
@@
 # Background
-Case series data from hospitals has suggested a preponderance of certain conditions including diabetes and cardiovascular disease, and elevated risks in those of non-European ethnic origin.
+Reports based on cases seen in hospitals hve suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected, and that a higher than expected proportion of severe cases are of non-European ethnic origin.
 Quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator.
@@
 
-In Scotland the existence of a laboratory database with nationwide coverage and a unique health care identifier on all health records has enabled us to ascertain all those with tests positive for SARS-CoV-2 and link to other health records.
-Here we describe a population wide case-control study of test confirmed COVID-19 disease in Scotland.
-From this we report the demographic factors, co-morbidities and concurrent medications associated with being hospitalised for COVID-19 and for becoming a severe COVID-19 case i.e.
-either warranting critical care support or dying within 28 days of a positive test.
-This study is intended to inform policy on defining vulnerable groups, identify any drug classes that influence risk of severe disease, and lay a basis for evaluating the feasibility of risk-stratified control measures.
+The aim of this study was to investigate risk factors for severe or fatal cases of COVID-19 confirmed by testing.
+Specific objectives were to test hypotheses about drugs classes associated with altered risk, and to construct a risk prediction algorithm that can be used to identify high-risk individuals who could be offered shielding during the epidemic.
+To achieve this aim we conducted a case-control study of all severe and fatal cases in Scotland, with controls matched for age, sex and primary care practice.
 
 # Methods
-## Data sources 
-The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.
-.This allowed all those with a positive viral nucleic acid test for SARS-CoV-2 to be ascertained, hereafter denoted as test positive cases.
-At the time writing, serology testing results for auto-antibodies indicating prior infection with SARS-CoV-2 are not yet available.
+## Case definition 
+The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
+The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.
+All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
+Admissions to critical care were obtained from the national audit database of the Scottish Intensive Care Society and Audit Group (SICSAG)ical care units (SICSAG).
+Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day.
+Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
+and cause of death were obtained from linkage to the National Register of Scotland.
 
-The National Community Health Index (CHI) database assigns a unique health care identifier to everyone in the population, and is used in all care systems in Scotland.
-Using the CHI identifier, records from ECOSS were linked to hospitalization and discharge data from the weekly updated Scottish Morbidity Record 01 and the daily updated RAPID-24 databases.
-Admission to critical care was ascertained from the national audit database of critical care units (SICSAG).
-Death registrations and cause of death were obtained from linkage to the National Register of Scotland.
-All these databases have nationwide coverage.
-At the time of writing the national policy on testing has evolved since the start of the epidemic with testing initially being restricted to possible cases that required hospital admission but widening to a "four-pillar" approach that has increased testing in health care workers residential care home settings.
-Accordingly the probability of being tested has changed over time.
+We classified test-positive cases into three groups:
+
+1.
+Severe or fatal COVID-19, defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test regardless of the cause of death given on the death certificate.
+
+2. Hospitalised but not severe as defined above
+
+3. Not hospitalised
+
+By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
+This focus on severe or fatal disease is relevant to policy because decisions about control measures have been based on projections of the requirement for critical care.
+
+In national mortality data, XX % of those certified with COVID-19 as underlying case of death had a positive test record.
+
+## Matched controls
 For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case.
@@
 
-For all cases and controls co-morbid conditions, the Charlson Comorbidity Index and drug exposure data were available from i) a five year look back query of the Scottish Morbidity Record 01 for sets of International Classification of Disease (ICD)-10 codes for various conditions ii) querying the national prescribing records for exposure to various drug classes iii) linkage to the national diabetes register SCI-diabetes.
+## Demographic data
+Ethnicity of cases and controls was classified using a name classification algorithm [@lakha_name_2011] with the names in the CHI database.
+This is able to distinguish South Asian and Chinese groups from the White group, but misclassifies Black-Caribbean people (a relatively small group in Scotland) as White.
+For those with a hospital admission record, a separate coding of ethnicity was available using self-described categories based on those used in the national census.
+Socioeconomic status was scored as the Scottish Index of Multiple Deprivation, based on the postcode in the CHI database.
+
+## Morbidity and drug prescribing
+
+All cases and Using the CHI identifier, records from ECOSS were linked to hospital discharge records in the Morbidity Record 01 (SMR01).  and
+
+For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01, excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.
+BNF drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 15 days before testing positive for SARS-CoV-2.
+Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
 
 We initially examined history of a specific list of conditions and drug classes that have been advised as high risk conditions for COVID-19 by Public Health England, World Health organisation or the US Centres for Disease Control.
-ICD-10 codes for the conditions are listed in Supplementary Materials.
 
-Ethnicity was ascertained from ethnic group coding on prior SMR records and from applying the ONOMAP algorithm to names available in the CHI database.
-All cases and controls were assigned to a quintile of the Scottish Index of Multiple Deprivation- this is a residential postcode based indicator of socioeconomic status.
+Co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
 
-## Case definition
-We classified test-positive cases into three groups:
+The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.
 
-1.
-Severe or fatal COVID-19, defined by entering critical care in the Scottish Intensive Care Society and Audit Group database (SICSAG) OR death within 28 days of a positive nucleic acid test (regardless of the cause of death given on the death certificate).
+## Statistical methods
 
-2. Hospitalised but not severe as defined above
+We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)
 
-3. Not hospitalized
+The univariate association of each variable with severe COVID-19 was modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`
 
-The focus on severe disease is relevant to policy makers because decisions about control measures have been based on projections of the requirement for critical care.
-By restricting the case definition to this group, which combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of test-positive cases, whatever triage procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
+We tested each effect for interaction with age and sex.  Effect of interest
 
-The proportion of fatal cases that were not tested can be examined in national mortality data ...
+The analyses were repeated after excluding those aged over 75 years.
 
-# Statistical methods
-We used  the R statistical package.
-We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020.
-(CHECK)
+To construct a risk prediction model, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.
 
-We tabulated the prevalence of co-morbidities and current drug use by case status.
-We used conditional logistic regression to estimate the univariate odds associated with each covariate of interest conditioned on age sex and GP practice.
-We also report a conditional logistic regression with all covariates entered simultaneously.
+The performance of the risk prediction model as a classifier was examined by n-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination and (more conventionally) as the C-statistic.
 
-We tested each effect for interaction with age and sex.
-
-Models were run overall and by age band.
-
-To select variables to be retained in the model, we used backward selection with a termination criterion that the reduction in deviance is less than the value corresponding to a threshold p-value of 0.01
-Data were split into a 70:30 training:test sets. 
-We evaluated the performance of the resulting backward selection model for predicting severe COVID-19 disease in the test set.
-For this we used the expected information for discrimination as described and used previously
-
-### The initial demographic vars of interest are ( give that age and sex is matched)
-
- ethnic group
- SIMD
-
-### The initial comorbidities of interest are listed below
+### Comorbidities of initial interest
 
@@
 
-# Information governance
+## Information governance
 This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data.
@@
 ## Association of demographic factors with severe disease
+
+For the case-control comparison, this version of the manuscript restricts the dataset to those aged less than `r max(cc.severe$AGE) + 1` years.
+
 Table \ref{tab:demog} shows univariate associations of demographic factors with disease.
-Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home (rate ratio 8.6).
+Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.
 In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.
@@
 	   split.tables="Inf",
-       split.cells=c(15, 15, 25, 8),
+       align=rep("centre", 4), 
+       split.cells=c(16, 16, 25, 8),
        caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")
@@
 
-table.stepwise <- as.data.frame(stepwise.all[, c(1, 3, 5)])
-colnames(table.stepwise) <- c("Coefficient (log rate ratio)", "Standard error", "p.value")
+table.stepwise <- as.data.frame(stepwise.all[, c(1, 5)])
+table.stepwise[, 1] <- exp(table.stepwise[, 1])
+colnames(table.stepwise) <- c("Rate ratio", "p.value")
 rownames(table.stepwise) <- replace.names(rownames(table.stepwise))
@@
        # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
-	   digits=c(3, 3, 1), 
+	   digits=c(3, 1), 
 	   split.table="Inf",
@@
 ## Predictive performance of a risk classifier
-
 The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.
-Table \ref{tab:wevid} shows that the C-statistic was `r summary(stepwise.densities)[3]` and the expected information for discrimination was  `r summary(stepwise.densities)[5]` bits.
+Table \ref{tab:wevid} shows that the C-statistic was `r summary(all.densities)[3]` and the expected information for discrimination was  `r summary(all.densities)[5]` bits.
 
@@
 
-pander(summary(stepwise.densities), table.style="multiline",
+pander(summary(all.densities), table.style="multiline",
        split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
@@
 
-plotWdists(stepwise.densities) + theme_grey(base_size = 18) + 
+plotWdists(all.densities) + theme_grey(base_size = 18) + 
 	theme(legend.title=element_blank())
@@
 ```
+
+# Discussion
+
+What is the explanation of the 14-fold higher incidence of severe disease in those who have encashed at least one prescription in the last year?
+
+# References {-}
+
+<div id="refs"></div>

### 2020-05-07 casecontrol.Rmd
Fixed to read ICD subchapters and BNF section codes

@@
 ---
-title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'
+title: 'Risk factors for severe COVID-19: a case-control study based on the Scottish population'
 author: 
@@
   \usepackage{array}
+  \usepackage{booktabs}
+  \usepackage{longtable}
+  \usepackage{lscape}
+
 output: 
@@
 
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
+
+library(knitr)
+library(kableExtra)
+
 ```
 
-\newpage
+# Abstract
+
+_Objectives_: to investigate risk factors for severe COVID-19 in the population.
+Specifically, we wish to establish how far the association of severe disease with pre-existing morbidity is accounted for by a few circumscribed conditions that confer high vulnerability, rather than by associations with poor health that span multiple disease categories.
+Other objectives were to identify any drug exposures that are independently associated with high or low risk, and to assess the performance of a risk classifier constructed from medical records to discriminate between those at high and low risk for severe disease
+
+_Methods_: age-sex-matched case-control study of severe disease, with each case matched to 10 controls registered with the same primary care practice.
+
+_Results_:
+
+* No new conditions associated with high risk independently of age and general health.
+
+* Drug exposures unexpectedly associated with high risk: anticoagulants
+
+* Performance of a risk classifier: at least 1.0 bits independently of age and sex
+
+* A rule-based classifier based on ... can extract XX% of the information obtained from a model-based classifier.
+
+_Conclusions_- These results show that risk stratification based on medical records can discriminate better than relying on age plus a list of conditions that confer highly vulnerable status.
+They lay the basis for policies based on stratify and shield.
 
@@
 
-### Comorbidities of initial interest
+te that at this point we are not including high risk of exposure as covariates ie HCW, residential care home, prisoner etc this will form part of future work.
+
+## Information governance
+This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data.
+Datasets were de-identified before analysis.
+
+# Results
+## Comparison of severe cases with others who tested positive 
+We emphasize that because testing was used selectively over this period, those who tested positive and were never hospitalised are unlikely to be representative of never-hospitalized cases in the population.
+
+```{r agegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.agegr) <- replace.names(rownames(table.agegr))
+knitr::kable(table.agegr,
+             escape=TRUE, 
+             booktabs=TRUE,
+             label="agegr",
+             align=rep("c", 8), 
+             caption="Associations of severe disease with health status by age group") %>%
+    column_spec(column=1, width="3.5cm") %>%
+    group_rows("Hospital discharge diagnoses", 3, 6) %>%
+    add_header_above(c(" "=1, "0-34 years"=2, "35-54 years"=2,
+                       "55-74 years"=2, "75+ years"=2)) %>%
+    landscape()
+
+```
+
+## Association of demographic factors with severe disease
+
+For the case-control comparison, this version of the manuscript restricts the dataset to those aged less than `r max(cc.severe$AGE) + 1` years.
+
+Table \ref{tab:demog} shows univariate associations of demographic factors with disease.
+Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.
+In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.
+
+\newpage
+
+```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
+
+table.demog.aug <- combine.tables3(table.demog, univariate.demog, multivariate.demog)
+rownames(table.demog.aug) <- replace.names(rownames(table.demog.aug))
+knitr::kable(table.demog.aug,
+             escape=TRUE, 
+             booktabs=TRUE,
+             label="demog",
+             digits=c(0, 0, 0, 20, 0, 100), 
+             col.names=c(colnames(table.demog.aug)[1:2],
+                         "Rate ratio (95% CI)", "p-value",
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with demographic factors") %>%
+    column_spec(column=1, width="4cm") %>%
+    group_rows("Ethnicity", 1, 2) %>%
+    group_rows("SIMD quintile", 3, 6) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+
+\newpage
+
+## Associations of comorbid conditions with severe disease
+In univariate logistic regression models, both Type 1 and Type 2 diabetes were associated with severe disease (Table \ref{tab:conditions}.
+There were strong associations with circulatory disease, respiratory disease, kidney disease and neurological disease.
+No association of severe disease with cystic fibrosis was detected.
+A possible explanation for this may be that this group was effectively shielded from infection in accordance with official guidance.
+
+```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
+
+table.conditions.aug <- combine.tables3(table.conditions, univariate.conditions,
+                                        multivariate.conditions)
+rownames(table.conditions.aug) <- replace.names(rownames(table.conditions.aug))
+knitr::kable(table.conditions.aug,
+             escape=TRUE, 
+             booktabs=TRUE,
+             label="conditions",
+             digits=c(0, 0, 0, 100, 0, 100), 
+             col.names=c(colnames(table.conditions.aug)[1:2],
+                         "Rate ratio (95% CI)", "p-value",
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with pre-existing conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    group_rows("Diabetes", 2, 4) %>%
+    group_rows("ICD-10 chapter", 5, nrow(table.conditions.aug)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+
+Further analyses to adjust for co-morbidity ...
+
+### Counts of co-morbid conditions
+
+Table xx shows the distribution of numbers of the co-morbid conditions listed by WHO or CDC or PHE by age band in severe cases versus their controls. 
+Across each age band there were people with severe COVID-19 who did not appear to have any of these co-morbid conditions.
+
+## Associations of prescribed drugs with severe disease 
+Table \ref{tab:drugs} shows that severe disease was associated with use of antihypertensive agents, anticoagulants, and lipid-lowering agents.
+Use of ACE inhibitors or angiotensin receptor blockers was not associated with higher risk than use of other classes of antihypertensive agent.
+Use of non-steroidal anti-inflammatory agents was weakly associated with severe disease.
+Use of hydroxychloroquine was associated with severe disease.
+
+Further analyses to adjust for co-morbidity ...
+
+```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
+
+table.drugs.aug <- combine.tables3(table.drugs, univariate.drugs, multivariate.drugs)
+rownames(table.drugs.aug) <- replace.names(rownames(table.drugs.aug))
+knitr::kable(table.drugs.aug,
+             escape=TRUE, 
+             booktabs=TRUE,
+             label="drugs",
+             digits=c(0, 0, 0, 100, 0, 100), 
+             col.names=c(colnames(table.drugs.aug)[1:2],
+                         "Rate ratio (95% CI)", "p-value",
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs") %>%
+    column_spec(column=1, width="4.5cm") %>%
+    group_rows("BNF chapter", 2, nrow(table.drugs.aug)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+
+## Examining those with no apparent co-morbid conditions
+
+A more in depth focus on severe cases at younger ages
+
+As shown in table  among severe case below age 50  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.
+Among severe case below age 50-70  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.
+However in those with none of these co morbid conditions xx% were on   drugs for xxx xxx and xxx.
+
+## Multivariate analysis of risk factors for severe disease
+
+A stepwise regression procedure was carried out with all demographic, morbidity and drug exposure variables, using alternating forward-backward selection.
+The `r nrow(stepwise.all)` variables retained are shown in Table \ref{tab:stepwise}
+
+```{r model, echo=FALSE, warning=FALSE, message=FALSE}
+
+table.stepwise <- as.data.frame(stepwise.all[, c(1, 5)])
+table.stepwise[, 1] <- exp(table.stepwise[, 1])
+table.stepwise[, 1] <- round(table.stepwise[, 1], 2)
+table.stepwise[, 2] <- signif(table.stepwise[, 2], 1)
+rownames(table.stepwise) <- replace.names(rownames(table.stepwise))
+
+# table.stepwise$p.value <- pvalue.latex(table.stepwise$p.value)
+
+knitr::kable(table.stepwise,
+             booktabs=TRUE,
+             escape=TRUE,
+             col.names=c("Rate ratio", "p-value"), 
+             label="stepwise",
+             align=rep("c", 2),
+             caption="Stepwise regression: variables retained in model for severe disease") %>%
+    group_rows("Diabetes", 3, 5)
+
+```
+
+## Predictive performance of a risk classifier
+The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.
+Table \ref{tab:wevid} shows that the C-statistic was `r summary(all.densities)[3]` and the expected information for discrimination was  `r summary(all.densities)[5]` bits.
+
+show incremental comparison of different models ...
+
+```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}
+
+pander(summary(all.densities), table.style="multiline",
+       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
+       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )
+
+```
+
+```{r plotwdists, echo=FALSE, fig.cap="Cross-validation of model chosen by stepwise regression: distributions of weight of evidence", out.width="0.8\\textwidth"}
+
+plotWdists(all.densities) + theme_grey(base_size = 18) + 
+	theme(legend.title=element_blank())
+
+```
+
+# Discussion
+
+What is the explanation of the 14-fold higher incidence of severe disease in those who have encashed at least one prescription in the last year?
+
+# References {-}
+
+<div id="refs"></div>
+
+\newpage
+
+# Appendix
+
+# Supplementary tables
+
+```{r icdchapter1, echo=FALSE}
+i <- 1
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r icdchapter2, echo=FALSE}
+i <- 2
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r icdchapter3, echo=FALSE}
+i <- 3
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r icdchapter4, echo=FALSE}
+i <- 4
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter5, echo=FALSE}
+i <- 5
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter6, echo=FALSE}
+i <- 6
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter7, echo=FALSE}
+i <- 7
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter8, echo=FALSE}
+i <- 8
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter9, echo=FALSE}
+i <- 9
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter10, echo=FALSE}
+i <- 10
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter11, echo=FALSE}
+i <- 11
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter12, echo=FALSE}
+i <- 12
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter13, echo=FALSE}
+i <- 13
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+```{r icdchapter14, echo=FALSE}
+i <- 14
+knitr::kable(tables.icd.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("icdchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with conditions in ICD-10 chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+#################### BNF chapters ############################
+
+```{r bnfchapter1, echo=FALSE}
+i <- 1
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter2, echo=FALSE}
+i <- 2
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter3, echo=FALSE}
+i <- 3
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter4, echo=FALSE}
+i <- 4
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter5, echo=FALSE}
+i <- 5
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter6, echo=FALSE}
+i <- 6
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter7, echo=FALSE}
+i <- 7
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter8, echo=FALSE}
+i <- 8
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter9, echo=FALSE}
+i <- 9
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter10, echo=FALSE}
+i <- 10
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter11, echo=FALSE}
+i <- 11
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter12, echo=FALSE}
+i <- 12
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter13, echo=FALSE}
+i <- 13
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+```{r bnfchapter14, echo=FALSE}
+i <- 14
+knitr::kable(tables.bnf.list[[i]],
+             escape=TRUE, 
+             booktabs=TRUE,
+             label=paste0("bnfchapter", i),
+             col.names=c(colnames(tables.icd.list[[i]])[1:2],
+                         "Rate ratio (95% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption=paste("Univariate associations with drug sections in BNF chapter", i)) %>%
+    column_spec(column=1, width="8cm") %>%
+    landscape()
+```
+
+## Comorbidities of initial interest
 
@@
 ​
-### Other recent health care exposure variables of interest are
+## Other recent health care exposure variables of interest are
 
@@
 Further comorbidities will be included in future protocol versions
-
-Note that at this point we are not including high risk of exposure as covariates ie HCW, residential care home, prisoner etc this will form part of future work.
-
-## Information governance
-This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data.
-Datasets were de-identified before analysis.
-
-# Results
-## Comparison of severe cases with others who tested positive 
-We emphasize that because testing was used selectively over this period, those who tested positive and were never hospitalised are unlikely to be representative of never-hospitalized cases in the population.
-
-## Association of demographic factors with severe disease
-
-For the case-control comparison, this version of the manuscript restricts the dataset to those aged less than `r max(cc.severe$AGE) + 1` years.
-
-Table \ref{tab:demog} shows univariate associations of demographic factors with disease.
-Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.
-In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.
-
-```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
-pvalue <- pvalue.latex(univariate.demog[, 5])
-ci <- or.ci(univariate.demog[, 1], univariate.demog[, 3]) 
-table.demog.aug <- data.frame(table.demog,
-                                   ci, 
-                                   pvalue)
-colnames(table.demog.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
-rownames(table.demog.aug) <- replace.names(rownames(table.demog.aug))
-
-pander(table.demog.aug, 
-       digits=c(0, 0, 3, 3, 3, 1),
-       round=c(0, 0, 3, 3, 3, 10),
-	   split.tables="Inf",
-       align=rep("centre", 4), 
-       split.cells=c(16, 16, 25, 8),
-       caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")
-
-```
-
-## Associations of comorbid conditions with severe disease
-In univariate logistic regression models, both Type 1 and Type 2 diabetes were associated with severe disease (Table \ref{tab:conditions}.
-There were strong associations with circulatory disease, respiratory disease, kidney disease and neurological disease.
-No association of severe disease with cystic fibrosis was detected.
-A possible explanation for this may be that this group was effectively shielded from infection in accordance with official guidance.
-
-```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
-
-pvalue <- pvalue.latex(univariate.conditions[, 5])
-ci <- or.ci(univariate.conditions[, 1], univariate.conditions[, 3]) 
-table.conditions.aug <- data.frame(table.conditions,
-                                   ci, 
-                                   pvalue)
-colnames(table.conditions.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
-rownames(table.conditions.aug) <- replace.names(rownames(table.conditions.aug))
-
-pander(table.conditions.aug, 
-       digits=c(0, 0, 3, 3, 3, 1),
-       round=c(0, 0, 3, 3, 3, 10),
-	   split.tables="Inf",
-       split.cells=c(20, 20, 20, 7),
-       caption="\\label{tab:conditions}Univariate association of severe disease with comorbid conditions")
-
-```
-
-Further analyses to adjust for co-morbidity ...
-
-### Counts of co-morbid conditions
-
-Table xx shows the distribution of numbers of the co-morbid conditions listed by WHO or CDC or PHE by age band in severe cases versus their controls. 
-Across each age band there were people with severe COVID-19 who did not appear to have any of these co-morbid conditions.
-
-## Associations of prescribed drugs with severe disease 
-Table \ref{tab:drugs} shows that severe disease was associated with use of antihypertensive agents, anticoagulants, and lipid-lowering agents.
-Use of ACE inhibitors or angiotensin receptor blockers was not associated with higher risk than use of other classes of antihypertensive agent.
-Use of non-steroidal anti-inflammatory agents was weakly associated with severe disease.
-Use of hydroxychloroquine was associated with severe disease.
-
-Further analyses to adjust for co-morbidity ...
-
-```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
-
-pvalue <- pvalue.latex(univariate.drugs[, 5])
-ci <- or.ci(univariate.drugs[, 1], univariate.drugs[, 3]) 
-table.drugs.aug <- data.frame(table.drugs,
-                                   ci, 
-                                   pvalue)
-colnames(table.drugs.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
-rownames(table.drugs.aug) <- replace.names(rownames(table.drugs.aug))
-
-pander(table.drugs.aug, 
-       digits=c(0, 0, 3, 3, 3, 1),
-       round=c(0, 0, 3, 3, 3, 10),
-	   split.tables="Inf",
-       split.cells=c(20, 20, 20, 7), 
-       caption="\\label{tab:drugs}Univariate association of severe disease with drugs prescribed in last year")
-
-```
-
-## Examining those with no apparent co-morbid conditions
-
-A more in depth focus on severe cases at younger ages
-
-As shown in table  among severe case below age 50  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.
-Among severe case below age 50-70  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.
-However in those with none of these co morbid conditions xx% were on   drugs for xxx xxx and xxx.
-
-## Multivariate analysis of risk factors for severe disease
-
-A stepwise regression procedure was carried out with all demographic, morbidity and drug exposure variables, using alternating forward-backward selection.
-The `r nrow(stepwise.all)` variables retained are shown in Table \ref{tab:stepwise}
-
-```{r model, echo=FALSE, warning=FALSE, message=FALSE}
-
-table.stepwise <- as.data.frame(stepwise.all[, c(1, 5)])
-table.stepwise[, 1] <- exp(table.stepwise[, 1])
-colnames(table.stepwise) <- c("Rate ratio", "p.value")
-rownames(table.stepwise) <- replace.names(rownames(table.stepwise))
-
-table.stepwise$p.value <- pvalue.latex(table.stepwise$p.value)
-
-pander(table.stepwise, table.style="multiline",
-       # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
-	   digits=c(3, 1), 
-	   split.table="Inf",
-       caption="\\label{tab:stepwise}Stepwise regression: variables retained in model for severe disease")
-
-```
-
-## Predictive performance of a risk classifier
-The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.
-Table \ref{tab:wevid} shows that the C-statistic was `r summary(all.densities)[3]` and the expected information for discrimination was  `r summary(all.densities)[5]` bits.
-
-show incremental comparison of different models ...
-
-```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}
-
-pander(summary(all.densities), table.style="multiline",
-       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
-       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )
-
-```
-
-```{r plotwdists, echo=FALSE, fig.cap="Cross-validation of model chosen by stepwise regression: distributions of weight of evidence", out.width="0.8\\textwidth"}
-
-plotWdists(all.densities) + theme_grey(base_size = 18) + 
-	theme(legend.title=element_blank())
-
-```
-
-# Discussion
-
-What is the explanation of the 14-fold higher incidence of severe disease in those who have encashed at least one prescription in the last year?
-
-# References {-}
-
-<div id="refs"></div>

### 2020-05-07 casecontrol.R

@@ -22,13 +22,14 @@ diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 record
 procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
 scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
 bnfcodes <- read_excel("./data/BNF_Code_Information.xlsx", sheet=4)
-
+colnames(bnfcodes) <- c("chaptername", "chapternum", "sectionname", "sectioncode")
+    
 bnfchapters <- bnfcodes[, 1:2]
 bnfchapters <- base::unique(bnfchapters)
-colnames(bnfchapters) <- c("fullname", "chapternum")
-truncate.at <- StrPos(bnfchapters$fullname, " |,|-") -1
-truncate.at[is.na(truncate.at)] <- nchar(bnfchapters$fullname[is.na(truncate.at)])
-bnfchapters$shortname <- substr(bnfchapters$fullname, 1, truncate.at) 
+#colnames(bnfchapters) <- c("fullname", "chapternum")
+truncate.at <- StrPos(bnfchapters$chaptername, " |,|-") -1
+truncate.at[is.na(truncate.at)] <- nchar(bnfchapters$chaptername[is.na(truncate.at)])
+bnfchapters$shortname <- substr(bnfchapters$chaptername, 1, truncate.at) 
 
 names(cc.all) <- gsub("CASE_NO", "stratum", names(cc.all))
 names(cc.all) <- gsub("^SEX$", "sex", names(cc.all))
@@ -37,14 +38,20 @@ names(cc.all) <- gsub("^ethnic$", "ethnic.old", names(cc.all))
 names(cc.all) <- gsub("^CAREHOME$", "care.home", names(cc.all))
 names(cc.all) <- gsub("^simd$", "SIMD.quintile", names(cc.all))
 
-cc.all$stratum <- as.integer(cc.all$stratum)
+## exclude controls already dead on date of testing case they were matched to
+controls.deceased <- with(cc.all, CASE==0 &
+                                  !is.na(Date.Death) &
+                                  Date.Death <= SPECDATE)
+cc.all <- cc.all[!controls.deceased, ]                         
 
+cc.all$stratum <- as.integer(cc.all$stratum)
 ## check that each stratum contains a single case
+cat("checking that each stratum contains a single case ...")
[…]

### 2020-05-12 casecontrol.Rmd
Fixed to use refactored version

--- 2020-05-06 casecontrol.Rmd
@@
 ---
-title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'
+title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population'
 author: 
@@
     affiliation: HPS
-  - name: Sharon Hutchinson 
-    affiliation: GCU
   - name: David McAllister 
@@
     affiliation: Usher
+  - name: Janet Murray
+    affiliation: HPS 
   - name: Tom Caparrotta 
     affiliation: IGMM
+  - name: Sharon Hutchinson 
+    affiliation: GCU
   - name: Helen M Colhoun
@@
      address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
-
 geometry: margin=1in
@@
 date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
-
 header-includes: |
@@
   \usepackage{array}
+  \usepackage{booktabs}
+  \usepackage{longtable}
+  \usepackage{lscape}
 output: 
@@
 
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
+
+library(knitr)
+library(kableExtra)
+
+table.densities <- rbind(summary(listed.densities),
+                         summary(full.densities))[, 1:6]
+
 ```
+
+# Abstract
+
+_Background_-- Why some people infected with Sars-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.
+Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.
+
+_Methods_ -- We linked the records of those tested positive for SARS-C0V-2 viral RNA in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data.
+Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test.
+We selected 10 sex, age and general practice based controls per test-positive case.
+For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year.
+We evaluated the risk of severe COVID-19 associated with  a pre-defined  set of  conditions using conditional logistic regression.
+We examined how much predictive performance  for severe COVID-19  a risk classifier built  and tested with four-fold cross validation based on these conditions could achieve and whether this could be improved using information from the rest of the hospitalisation and prescribing data.
+
+_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
+In the case-control analysis, the odds ratios were
+
+`r table.agegr.all[1, 3]` for residence in a care home status, 
+`r table.agegr.all[4, 3]` for diabetes, 
+`r table.agegr.all[8, 3]` for chronic renal failure, 
+`r table.agegr.all[5, 3]` for ischemic heart disease
+`r table.agegr.all[9, 3]` for neurological disease, 
+`r table.agegr.all[10, 3]` for chronic liver disease and 
+`r table.agegr.all[11, 3]` for immune deficiency or suppresssion
+
+Overall but INSERT xx% of severe cases versus INSERT xx% of controls had at least one of these INSERT ( xx versus xx% in those aged under 40 years).
+The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
+These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.
+
+In comparison with a baseline model based on care home residence only, adding listed conditions increased the information for discrimination to  INSERT 0.9 bits.
+Adding indicator variables for ICD-10 and BNF chapters increased the information for discrimination further to  INSERT 1.5 bits.
+From a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
+
+*Conclusions* --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and the vast majority of younger persons developing severe disease have an underlying medical condition.
+The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.
+A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.
+The strong association with prescribing history we have observed may reflect co-morbidity or adverse drug effects and we are investigating this further.
 
@@
 # Background
-Reports based on cases seen in hospitals hve suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected, and that a higher than expected proportion of severe cases are of non-European ethnic origin.
-Quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator.
-Furthermore it has not been possible to disentangle whether drug effects may underlie the apparent association with certain co-morbid groups.
+It is clear that  there is wide susceptibility to risk of severe COVID-19 disease  with older age being a predominant factor.
+Case series from many countries have  suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected.
+However there are also many anecdotal reports of apparently healthy young persons succumbing to disease.
+Apart from co-morbidities the potential  role of ethnicity in disease severity has been highlighted.
+However quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator.
+In particular a  systematic evaluation of the past medical history in those with and without severe disease has not been reported.
 
-The aim of this study was to investigate risk factors for severe or fatal cases of COVID-19 confirmed by testing.
-Specific objectives were to test hypotheses about drugs classes associated with altered risk, and to construct a risk prediction algorithm that can be used to identify high-risk individuals who could be offered shielding during the epidemic.
-To achieve this aim we conducted a case-control study of all severe and fatal cases in Scotland, with controls matched for age, sex and primary care practice.
+Here we  used extensive health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID -19 disease.
+We defined  severe disease as viral RNA-confirmed diagnosis of COVID-19 as  that required critical care support or was associated with death within 28 days of testing.
+We focused on severe disease it is of most  concern but also  because it can be completely ascertained  whereas  testing policies means that   complete ascertainment of less severe disease is not possible in most countries.
 
@@
 All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
-Admissions to critical care were obtained from the national audit database of the Scottish Intensive Care Society and Audit Group (SICSAG)ical care units (SICSAG).
-Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day.
+Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01.
+Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG)  database that captures admission to all critical care units across Scotland.
 Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
@@
 By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
-This focus on severe or fatal disease is relevant to policy because decisions about control measures have been based on projections of the requirement for critical care.
-
-In national mortality data, XX % of those certified with COVID-19 as underlying case of death had a positive test record.
 
 ## Matched controls
-For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case.
+For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
 Note that this is an incidence density sampling design i.e.
-a control who becomes a case at a later point is retained in the control set.
+a control who becomes a case at a later point is correctly retained in the control set.
 With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.
@@
 ## Demographic data
-Ethnicity of cases and controls was classified using a name classification algorithm [@lakha_name_2011] with the names in the CHI database.
-This is able to distinguish South Asian and Chinese groups from the White group, but misclassifies Black-Caribbean people (a relatively small group in Scotland) as White.
-For those with a hospital admission record, a separate coding of ethnicity was available using self-described categories based on those used in the national census.
-Socioeconomic status was scored as the Scottish Index of Multiple Deprivation, based on the postcode in the CHI database.
+Socioeconomic status was scored using the postcode based indicator  the Scottish Index of Multiple Deprivation using the postcode in the CHI database.
+Ethnic group was captured in two ways: self designated ethnicity  was derived from past records in the hospital admissions and outpatient attendances  (Scottish Morbidity Records) using  categories based on those used in the national census.
+However since this is incomplete  we also used the a name classification algorithm ONOMAP [@lakha_name_2011] with the names in the CHI database.
+This is able to distinguish South Asian and Chinese groups but not Black-Caribbean people from the White group.
+Residential care home status was available via the CHI database.
 
 ## Morbidity and drug prescribing
-
-All cases and Using the CHI identifier, records from ECOSS were linked to hospital discharge records in the Morbidity Record 01 (SMR01).  and
-
-For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01, excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.
-BNF drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 15 days before testing positive for SARS-CoV-2.
+For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
+British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
 
-We initially examined history of a specific list of conditions and drug classes that have been advised as high risk conditions for COVID-19 by Public Health England, World Health organisation or the US Centres for Disease Control.
-
-Co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
-
+We first extracted history of a specific list of conditions and drug classes that have been  included as  risk conditions for COVID-19 by  public health agencies https://www.nhs.uk/conditions/coronavirus-covid-19/people-at-higher-risk-from-coronavirus/.
+These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
 The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.
+In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes  and BNF  codes to chapter level.
+We did not  specifically  tabulate  the conditions that have been listed as making people " clinically extremely vulnerable"  such as cystic fibrosis or leukaemia  since there is no  clinical equipoise  regarding risk in such persons and since they will be underrepresented in the case data as they have been in complete shielding since early in the epidemic.
 
 ## Statistical methods
-
-We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)
-
-The univariate association of each variable with severe COVID-19 was modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`
-
-We tested each effect for interaction with age and sex.  Effect of interest
-
-The analyses were repeated after excluding those aged over 75 years.
-
-To construct a risk prediction model, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.
-
-The performance of the risk prediction model as a classifier was examined by n-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination and (more conventionally) as the C-statistic.
-
-### Comorbidities of initial interest
-
-- Circulatory Disorders
-  - Ischaemic heart disease
-  - Other heart disease
-  - Cerebrovascular Disease
-  - Other circulatory system diseases
-- Neurological diseases except inflammatory
-  - Epilepsy
-  - Mono and  polyneuropathies
-  - Other neurological conditions
-- Respiratory diseases
-  -  Acute respiratory infections
-  -  Asthma
-  -  Other Chronic lower respiratory disease
-- Tuberculosis 
-- Connective tissue diseases
-- Decompensated liver disease
-- Kidney disease
-  - Advanced Chronic kidney disease or RRT
-  - Other chronic kidney disease
-- Diabetes
-  - type 1
-  - type 2
-- Malignant neoplasms
-  - Lung cancers
-  - Blood cancers
-  - Other cancers
-- Immunological diseases
-  - HIV
-  - Certain disorders involving the immune system
-- Sickle cell disease
-- Cystic fibrosis
-- Organ transplantation other than kidney
-- Treated hypertension (on any antihypertensive drugs in past year)
-- Charlson comorbidity index
-
-​
-### Other recent health care exposure variables of interest are
-
-- Any inpatient admission in past year and excluding the three weeks prior to testing 
-- Any emergency admission in past year and excluding the three weeks prior to testing
-- Any ICU/HDU/Admission in past year and excluding the three weeks prior to testing
-
-### Initial drug exposures of interest are
-
-- Any lipid  drug 
-- Any anticoag or platelet  
-- Any Antihypertensive drugs disaggregated to: 
-  - Vasodilator antihypertensive drugs
-  - Centrally-acting antihypertensive drugs
-  - Adrenergic neurone blocking drugs
-  - Alpha-adrenoceptor blocking drugs
-  - Angiotensin-converting enzyme inhibitors
-  - Angiotensin-II receptor antagonists
-  - Renin inhibitors
-  - Thiazides and related diuretics
-  - Calcium-channel blocker
-- Any prescribed NSAID
-- Any cytotoxic or immunosuppressant drug 
-- Any chloroquine drug or derivatives of chloroquine
-
-Derivations ICD10 and BNF codes for these comorbidities / drugs are in the appendix
-
-Further comorbidities will be included in future protocol versions
-
-Note that at this point we are not including high risk of exposure as covariates ie HCW, residential care home, prisoner etc this will form part of future work.
+We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020.
+(CHECK)  The univariate association of  sociodemographic variables, pre-defined conditions and discharge and prescribing history  with severe COVID-19 were modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit` We examined the associations by age band and overall.
+Among those  cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF  chapter with severe COVID-19.
+To construct a risk prediction models, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.
+The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.
 
@@
 # Results
-## Comparison of severe cases with others who tested positive 
-We emphasize that because testing was used selectively over this period, those who tested positive and were never hospitalised are unlikely to be representative of never-hospitalized cases in the population.
+<!---
+Still needed
 
-## Association of demographic factors with severe disease
+- A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code
+- Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
+-  Run table 2 analyses for all subjects combined 
+- what is currently tables 3,4,5  and 7  could be put into appendix and the table reference in text changed as appropriate ?
+- tables currently 9,11,12 should be removed as will go in pharmacoepi  paper  --->
+## Incidence and mortality from severe COVID-19 in the Scottish population
+Logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.
+Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.
+The relationship of mortality to age was almost exactly linear on a logit scale.
+For fatal and non-fatal cases combined, the mortality differentials between males and females widened in middle age before narrowing again after about age 75 years.
+In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.
+The odds ratio associated  with male sex was is INSERT  for all severe disease and  INSERT  for fatal disease.
+As shown in Table INSERT  almost all of the severe cases  death over aged 75years died  but  the majority of severe cases under aged 60 survived.
 
-For the case-control comparison, this version of the manuscript restricts the dataset to those aged less than `r max(cc.severe$AGE) + 1` years.
-
-Table \ref{tab:demog} shows univariate associations of demographic factors with disease.
+## Association with demographic factors
+Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.
 Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.
 In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.
+Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.
+Although the numbers are small, there is no evidence that the Black minority is at higher risk of severe disease.
 
-```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
-pvalue <- pvalue.latex(univariate.demog[, 5])
-ci <- or.ci(univariate.demog[, 1], univariate.demog[, 3]) 
-table.demog.aug <- data.frame(table.demog,
-                                   ci, 
-                                   pvalue)
-colnames(table.demog.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
-rownames(table.demog.aug) <- replace.names(rownames(table.demog.aug))
+## Associations with listed conditions
+Overall  INSERT xx of cases and xx of controls had at least one of the listed conditions.
+The  prevalence of the listed conditions in cases and controls by age band is shown in Table  INSERT .
+Almost half of cases aged under 40 years had at least one listed condition, compared with only 9% (43 / 470) of controls.
+In those 75+ years INSERT xxx 80% of cases had at least one listed condition but  INSERT xx% of controls also did.
 
-pander(table.demog.aug, 
-       digits=c(0, 0, 3, 3, 3, 1),
-       round=c(0, 0, 3, 3, 3, 10),
-	   split.tables="Inf",
-       align=rep("centre", 4), 
-       split.cells=c(16, 16, 25, 8),
-       caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")
+As shown  in the age band specific univariate logistic regression models in tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}.
+Here we collapsed the two younger age bands to <60 years.
+All of the listed conditions were significantly associated with severe disease  in each age band.
+Diabetes was associated with a 6.6.
+fold rate of severe COVID-19 under age 60 years.
+Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.
 
+Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}  also show  multivariate rate ratios by age group.
+The key point of this  analysis is to demonstrate that at all ages there was a very strong  and highly significant  association of having had any prescription in the last year  and of having had any admission in the past five years even adjusted for all the listed conditions and care home residence.
+
+## Systematic analysis of diagnoses associated with severe disease
+The approximately two-fold  risk of  with severe  COVID-19  associated with prior hospital admission independently  of other conditions and factors was further explored.
+The association of hospital diagnoses of each ICD-10 chapter with severe COVID-19, restricted  to those without any of the listed conditions.
+These results are shown in Table \ref{tab:conditions}.
+In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
+In a multivariate analysis of all chapters, the most strongly associated chapter-level diagnosis was Chapter 2 (neoplasms).
+This was further disaggregated with subchapters in as shown in Supplement  Table \ref{tab:icdchapter2}.
+Unexpectedly, the strongest association was with the subchapter "Malignant Neoplasms of Digestive Organs".
+Several other subchapters corresponding to cancer sites were also associated with severe disease.
+
+## Associations of prescribed drugs with severe disease 
+Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.
+The rate ratio associated with this variable varies from INSERT 9.2 in those aged under 60 years to INSERT 42.2 in those aged 75 years and over.
+In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to  INSERT3.8 and INSERT11.9 respectively.
+It is notable that about one third of those over 75 had not encashed a prescription in the previous year.
+
+To investigate this further, we partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.
+Table \ref{tab:drugs} shows these associations.
+In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
+In a multivariate analysis of all chapters, these associations were much weaker.
+The BNF chapter with the strongest association with severe disease was Chapter 1 (gastrointestinal).
+Other chapters associated with severe disease were Chapter 2 (cardiovascular) and Chapter 4 (central nervous system).
+
+## Construction of a multivariate risk prediction model 
+Predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.
+
+For each of variable set, a stepwise regression procedure was carried out using alternating forward-backward selection.
+The variables retained with each variable set are shown in Table \ref{tab:stepwise} .
+The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.
+Observed and predicted case status were compared over all test folds.
+Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[1, 3]` to `r table.densities[2, 3]` and the expected information for discrimination from `r table.densities[1, 5]` bits to `r table.densities[2, 5] ` bits.
+
+This value for the information conditional on age and sex can be added to the information for discrimination obtained with a logistic regression model using age and sex to obtain the total information for discrimination that would be obtained in the population  to give a total of WHAT??.
+PAUL HOW MUCH IS THAT THIS IS NOT CLEAR
+
+Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence favouring case over <!--thi figure should go into a supplement--> control status from the model based on the extended variable set.
+This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
+
+# Discussion  <!--helen still to edit -->
+## Multifactorial risk
+In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.
+Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged.
+Being in a residential care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis.
+
+In contrast reports from the U.S and elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020] we found no evidence of an increased risk in non-White minorities.
+This suggests that ethnic differences found elsewhere are not genetically driven but reflect other social factors.
+In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by the health care worker status [@niedzwiedz_ethnic_2020;@ho_modifiable_2020 ].
+
+Media reports of apparently otherwise healthy youths succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event.
+However we found that almost half of cases under 40 years had at least one condition listed by public health agencies as increasing risk.
+In those that did not have a listed condition  having any hospital admission  in the past five years and having any prescription in the past year  was much more common  among cases that in the background population of the same age.
+
+We chose to focus initially on the contribution of  conditions already proposed as high risk.
+We confirmed that these are associated with  substantially increased risk .
+However our systematic analysis has shown that there are many many  many other conditions considered in the same univariate way are also  associated with severe COVID-19 disease.
+
+## Possible drug effects
+A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year.
+This association was only partly explained by associations with listed conditions.
+Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
+In an accompanying paper we explore these associations more fully.
+
+## Methodological strengths and weaknesses
+Most of the reports of disease associations with COVID-19 have been case series descriptions there are few prospective cohort or case control analyses with appropriate population comparators.
+With this matched case control design using incidence density sampling, we have been able to estimate rate ratios conditional on age and sex.
+An unpublished analysis from England explored the association of similar set of risk conditions with in-hospital COVID-19 deaths, but did not systematically evaluate the rest of the medical record including prescription records.
+Although we have records of encashment of prescriptions, we do not at present have access to other the primary care data, which would contain additional information such as body mass index.
+Hospital discharge diagnoses however are coded to ICD-10 by trained coders.
+A further limitation is generalisability especially of sociodemographic factors to other locations.
+Also we limited our case control study to test-positive cases.
+This excludes deaths attributed to COVID-19 where no testing was carried out.
+At present there are xxx such deaths in Scotland almost all aged >> xxx.
+Others have argued that some COVID-19 deaths may have been attributed to other causes early in the epidemic.
+These aspects may mean some relative risks e,g that for residential care home status have been underestimated.
+However in the younger age bands, it is likely that most severe or fatal cases will have reached hospital and been tested.
+
+## Relevance to policy
+As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.
+The "stratify and shield" policy option, in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on accurate risk discrimination.
+It is also to be expected that as awareness grows of how risk varies between individuals, individuals will seek to learn whatever they can about their own level of risk.
+A key implication of our results is that risk of severe or fatal disease is multifactorial and that the ninefold rate ratio associated with a 20-year difference in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological that impair mobility.
+A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will not be an efficient way to allocate resources to those who most need it.
+Assigning all persons with the current list of risk conditions will be too insensitive and too non-specific - a third of 60-75 years old in the background population have at least one of the listed  conditions we examined.
+Half of the younger people with severe disease don't have any of them.
+Therefore it is necessary to construct a risk classifier that optimizes sensitivity and specificity based on the information available.
+Our results show that this is possible in principle, though more work is required to construct a practical risk classifier for Scotland, let alone to develop and validate classifiers for use in other populations.
+
+## Conclusion
+In conclusion the epidemiology of severe COVID-19 disease is becoming clearer. There is not just one small set of conditions that contribute to risk but many. ......
+
+# References {-}
+
+<div id="refs"></div>
+
+```{r plotrates, echo=FALSE, fig.cap="\\label{fig:rates}Incidence of severe and fatal COVID-19 in Scotland by age and sex: generalized additive models fitted to severe and fatal cases for males nd females separately", out.width="0.8\\textwidth"}
+
+breaks.gam <- c("1/20,000", "1/10,000", "1/5,000",
+                "1/2000", "1/1000", "1/500",
+                "1/200", "1/100")
+theme_set(
+    theme_gray(base_size = 14)
+)
+
+ggplot(data=gam, aes(x=Age, y=value, colour=Sex, linetype=Status)) +
+    geom_line() +
+    scale_x_continuous(name="Age", limits=c(20, 90),
+                       breaks=seq(20, 80, by=20),
+                       labels=seq(20, 80, by=20)) + 
+    scale_y_continuous(name="Risk (logit scale)",
+                       breaks=car::logit(c(5E-5, 1E-4, 2E-4,
+                                           5E-4, 1E-3, 2E-3,
+                                           5E-3, 1E-2)),
+                       labels=breaks.gam,
+                       limits=car::logit(c(2E-5, 5E-3)))
 ```
 
-## Associations of comorbid conditions with severe disease
-In univariate logistic regression models, both Type 1 and Type 2 diabetes were associated with severe disease (Table \ref{tab:conditions}.
-There were strong associations with circulatory disease, respiratory disease, kidney disease and neurological disease.
-No association of severe disease with cystic fibrosis was detected.
-A possible explanation for this may be that this group was effectively shielded from infection in accordance with official guidance.
+\newpage
 
-```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
+```{r plotwdists, echo=FALSE, fig.cap="\\label{fig:wevid}Cross-validation of model chosen by stepwise regression using extended variable set: class-conditional distributions of weight of evidence", out.width="0.8\\textwidth"}
 
-pvalue <- pvalue.latex(univariate.conditions[, 5])
-ci <- or.ci(univariate.conditions[, 1], univariate.conditions[, 3]) 
-table.conditions.aug <- data.frame(table.conditions,
-                                   ci, 
-                                   pvalue)
-colnames(table.conditions.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
-rownames(table.conditions.aug) <- replace.names(rownames(table.conditions.aug))
-
-pander(table.conditions.aug, 
-       digits=c(0, 0, 3, 3, 3, 1),
-       round=c(0, 0, 3, 3, 3, 10),
-	   split.tables="Inf",
-       split.cells=c(20, 20, 20, 7),
-       caption="\\label{tab:conditions}Univariate association of severe disease with comorbid conditions")
-
-```
-
-Further analyses to adjust for co-morbidity ...
-
-### Counts of co-morbid conditions
-
-Table xx shows the distribution of numbers of the co-morbid conditions listed by WHO or CDC or PHE by age band in severe cases versus their controls. 
-Across each age band there were people with severe COVID-19 who did not appear to have any of these co-morbid conditions.
-
-## Associations of prescribed drugs with severe disease 
-Table \ref{tab:drugs} shows that severe disease was associated with use of antihypertensive agents, anticoagulants, and lipid-lowering agents.
-Use of ACE inhibitors or angiotensin receptor blockers was not associated with higher risk than use of other classes of antihypertensive agent.
-Use of non-steroidal anti-inflammatory agents was weakly associated with severe disease.
-Use of hydroxychloroquine was associated with severe disease.
-
-Further analyses to adjust for co-morbidity ...
-
-```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
-
-pvalue <- pvalue.latex(univariate.drugs[, 5])
-ci <- or.ci(univariate.drugs[, 1], univariate.drugs[, 3]) 
-table.drugs.aug <- data.frame(table.drugs,
-                                   ci, 
-                                   pvalue)
-colnames(table.drugs.aug) <- c("Controls", "Cases", "Rate ratio (95% CI)", "p-value")
-rownames(table.drugs.aug) <- replace.names(rownames(table.drugs.aug))
-
-pander(table.drugs.aug, 
-       digits=c(0, 0, 3, 3, 3, 1),
-       round=c(0, 0, 3, 3, 3, 10),
-	   split.tables="Inf",
-       split.cells=c(20, 20, 20, 7), 
-       caption="\\label{tab:drugs}Univariate association of severe disease with drugs prescribed in last year")
-
-```
-
-## Examining those with no apparent co-morbid conditions
-
-A more in depth focus on severe cases at younger ages
-
-As shown in table  among severe case below age 50  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.
-Among severe case below age 50-70  xxx% had either xxx or xxx or xxx but in xxx% none of these comorbid conditions were found.
-However in those with none of these co morbid conditions xx% were on   drugs for xxx xxx and xxx.
-
-## Multivariate analysis of risk factors for severe disease
-
-A stepwise regression procedure was carried out with all demographic, morbidity and drug exposure variables, using alternating forward-backward selection.
-The `r nrow(stepwise.all)` variables retained are shown in Table \ref{tab:stepwise}
-
-```{r model, echo=FALSE, warning=FALSE, message=FALSE}
-
-table.stepwise <- as.data.frame(stepwise.all[, c(1, 5)])
-table.stepwise[, 1] <- exp(table.stepwise[, 1])
-colnames(table.stepwise) <- c("Rate ratio", "p.value")
-rownames(table.stepwise) <- replace.names(rownames(table.stepwise))
-
-table.stepwise$p.value <- pvalue.latex(table.stepwise$p.value)
-
-pander(table.stepwise, table.style="multiline",
-       # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
-	   digits=c(3, 1), 
-	   split.table="Inf",
-       caption="\\label{tab:stepwise}Stepwise regression: variables retained in model for severe disease")
-
-```
-
-## Predictive performance of a risk classifier
-The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.
-Table \ref{tab:wevid} shows that the C-statistic was `r summary(all.densities)[3]` and the expected information for discrimination was  `r summary(all.densities)[5]` bits.
-
-show incremental comparison of different models ...
-
-```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}
-
-pander(summary(all.densities), table.style="multiline",
-       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
-       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )
-
-```
-
-```{r plotwdists, echo=FALSE, fig.cap="Cross-validation of model chosen by stepwise regression: distributions of weight of evidence", out.width="0.8\\textwidth"}
-
-plotWdists(all.densities) + theme_grey(base_size = 18) + 
+plotWdists(full.densities) + theme_grey(base_size = 18) + 
 	theme(legend.title=element_blank())
@@
 
-# Discussion
+\newpage
 
-What is the explanation of the 14-fold higher incidence of severe disease in those who have encashed at least one prescription in the last year?
+```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
+options(knitr.kable.NA = '')
 
-# References {-}
+knitr::kable(table.demog.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="demog",
+             digits=c(0, 0, 0, 200), 
+             col.names=c(colnames(table.demog.aug)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Univariate associations of severe disease with demographic factors") %>%
+    column_spec(column=1, width="3cm") %>%
+    group_rows("Ethnicity", 1, 3) %>%
+    group_rows("SIMD quintile", 4, 8)
 
-<div id="refs"></div>
+```
+
+```{r agegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.agegr) <- replace.names(rownames(table.agegr))
+knitr::kable(table.agegr,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="agegr",
+             align=rep("c", 8), 
+             caption="Associations of severe disease with risk factors, by age group") %>%
+    column_spec(column=1, width="3.5cm") %>%
+    group_rows("Listed condition", 3, nrow(table.agegr) - 1) %>%
+    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
+                       "60-74 years"=2, "75+ years"=2)) %>%
+    landscape()
+
+```
+
+```{r listed.conditions.allages, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.agegr.all,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.conditions.allages",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(table.agegr.all)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions over all age groups") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+
+```{r listed.conditions.lt60, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(tables.agegr[[1]],
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.conditions.lt60",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(tables.agegr[[1]])[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+```{r listed.condition.60to74, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(tables.agegr[[2]],
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.conditions.60to74",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(tables.agegr[[2]])[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+```{r listed.condition.75plus, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(tables.agegr[[3]],
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.conditions.ge75",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(tables.agegr[[3]])[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+
+```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.conditions.aug) <- gsub("_", " ", rownames(table.conditions.aug))
+knitr::kable(table.conditions.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="conditions",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(table.conditions.aug)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with hospital diagnoses in last 5 years, in those without listed conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+
+```
+
+```{r icdchapter2, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.icdchapter2,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="icdchapter2",
+             col.names=c(colnames(table.icdchapter2)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 4), 
+             caption="Associations of severe disease with hospital diagnoses in ICD chapter 2, in those without listed conditions") %>%
+    column_spec(column=1, width="5cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2)) %>%
+    landscape()
+
+```
+
+```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.drugs.aug) <- gsub("_", " ", rownames(table.drugs.aug))
+knitr::kable(table.drugs.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="drugs",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(table.drugs.aug)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in those without listed conditions") %>%
+    column_spec(column=1, width="4.5cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+```
+
+```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.bnfchapter1,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="bnfchapter1",
+             digits=c(0, 0, 0, 200 , 0, 200), 
+             col.names=c(colnames(table.bnfchapter1)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 1, in those without listed conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+```
+
+```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.protonpump,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="protonpump",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(table.drugs.aug)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Univariate associations of severe disease with proton pump inhibitors in all cases and controls, by age group") %>%
+    column_spec(column=1, width="2cm")
+```
+
+```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.bnfchapter2,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="bnfchapter2",
+             col.names=c(colnames(table.bnfchapter2)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 2, in those without listed conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    group_rows("", 1, nrow(table.bnfchapter2)) %>%
+    landscape()
+
+```
+
+```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.bnfchapter4,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="bnfchapter4",
+             col.names=c(colnames(table.bnfchapter4)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 9, in those without listed conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    group_rows("", 1, nrow(table.bnfchapter4)) %>%
+    landscape()
+
+```
+
+```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.densities) <- c("Care home + listed conditions",
+                               "Extended variable set")
+
+knitr::kable(table.densities, 
+             booktabs=TRUE,
+             escape=FALSE,
+              row.names=TRUE,
+             label="xvalid",
+             align=rep("c", 6),
+             caption="Prediction of severe COVID-19: cross-validation of models chosen by stepwise regression") %>%
+    column_spec(column=1, width="2.5cm") %>%
+    column_spec(column=2:7, width="1.5cm")
+
+```
+
+\newpage
+
+# Appendix
+## Supplementary tables
+
+```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
+
+options(knitr.kable.NA = '')
+
+knitr::kable(table.ethnicsmr.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmr",
+             digits=c(0, 0, 0, 200), 
+             col.names=c(colnames(table.ethnicsmr.aug)[1:2],
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with ethnicity classified in hospital records") %>%
+    column_spec(column=1, width="3cm")
+
+```
+
+```{r model, echo=FALSE, warning=FALSE, message=FALSE}
+
+stepwise.full <- as.data.frame(stepwise.full[, c(1, 5)])
+stepwise.full[, 1] <- round(stepwise.full[, 1], 2)
+stepwise.full[, 2] <- signif(stepwise.full[, 2], 1)
+rownames(stepwise.full) <- replace.names(rownames(stepwise.full))
+rownames(stepwise.full) <- gsub("\\_", " ", rownames(stepwise.full))
+
+stepwise.full[, 2] <- pvalue.latex(stepwise.full[, 2])
+
+knitr::kable(stepwise.full, 
+             booktabs=TRUE,
+             escape=FALSE,
+             col.names=c("log rate ratio", "p-value"), 
+             label="stepwise",
+             align=rep("c", 2),
+             caption="Stepwise regression: variables retained in model for severe disease") %>%
+    group_rows("Diabetes", 3, 5)
+
+```
+
+<!--
+
+For each of these chapters, we examined univariate and multivariate associations with severe disease at the level of paragraph or subparagraph of the BNF classification, restricting the analysis to those without one of the listed conditions.
+
+Table \ref{tab:bnfchapter1} shows that prescriptions of most classes of drugs for gastrointestinal conditions were associated with severe COVID-19.
+By far the most widely prescribed drug class in this age group was proton pump inhibitors; 15% of controls and 36% of cases had encashed at least one prescription, with a univariate rate ratio of 3.2.
+In a multivariate analysis, the drug class most strongly associated with severe disease was proton pump inhibitors.
+
+Table \ref{tab:protonpump} shows the association of severe disease with proton pump inhibitors in all cases and controls by age group.
+The univariate rate ratio varied from 12 in those aged under 60 years to 2.8 in those aged 75 years and over.
+
+Table \ref{tab:bnfchapter2} shows the associations with classes of drugs prescribed for the cardiovascular system.
+Most classes of antihypertensive drug were associated with increased risks in univariate analysis, but there was no evidence that the risks associated with prescription of ACE inhibitors or angiotensin-II receptor antagonists differed from those for other drugs.
+Prescription of anticoagulants was associated with increased risk of severe disease.
+presumably reflecting the risks associated with the underlying cardiovascular conditions.
+In a multivariate analysis the strongest association of severe disease was with loop diuretics.
+Table \ref{tab:bnfchapter4} shows the associations with classes of drugs prescribed for the central nervous system.
+The strongest association with severe disease in a multivariate analysis was with prescription of non-opiate analgesics.
+
+A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year.
+This association was only partly explained by associations with listed conditions.
+Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
+
+Examination of the subchapters within this chapter showed that this association was largely explained by prescriptions of proton pump inhibitors, which had been encashed by 15% of controls.
+Over all age groups the rate ratio associated with use of proton pump inhibitors was 2.8, but in those aged less than 40 years, in whom serious underlying disease is presumaby less common, the rate ratio was 12.
+This association is not explained by association any underlying gastrointestinal disease that leads to hospital admission, by coprescribing with dual antiplatelet agents in people who have had a myocardial infarction or a vascular procedure.
+As proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.
+However as there are no prescription charges in Scotland, long-term users of these expensive drugs are likely to seek to obtain them on prescription.
+
+As proton pump inhibitors are commonly co-prescribed with NSAIDS, this is relevant to evaluating possible explanations for the association of proton pump inhibitors with NSAIDs. (CHECK).
+
+As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
+Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016].
+35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013].
+
+-->

### 2020-05-12 casecontrol.R

@@ -13,6 +13,8 @@ library(reshape2)
 library(readxl)
 library(DescTools)
 library(icd.data)
+library(gam)
+
 
 source("helperfunctions.R")
         
@@ -20,16 +22,13 @@ cc.all <- readRDS("./data/CC_linked_ANON_20200501.rds")
 
 diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 records ? excluded
 procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
+scotpop <- read_excel("../Scotland_midyearpop_est2019.xlsx")
+
+## we have only 7 digits on scrips, giving resolution to subpara level only
 scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
-bnfcodes <- read_excel("./data/BNF_Code_Information.xlsx", sheet=4)
-colnames(bnfcodes) <- c("chaptername", "chapternum", "sectionname", "sectioncode")
-    
-bnfchapters <- bnfcodes[, 1:2]
-bnfchapters <- base::unique(bnfchapters)
-#colnames(bnfchapters) <- c("fullname", "chapternum")
-truncate.at <- StrPos(bnfchapters$chaptername, " |,|-") -1
-truncate.at[is.na(truncate.at)] <- nchar(bnfchapters$chaptername[is.na(truncate.at)])
-bnfchapters$shortname <- substr(bnfchapters$chaptername, 1, truncate.at) 
+scrips$paracode <- as.integer(substr(scrips$bnf_paragraph_code, 1, 6))
+
+source("bnfcodes.R")
 
 names(cc.all) <- gsub("CASE_NO", "stratum", names(cc.all))
 names(cc.all) <- gsub("^SEX$", "sex", names(cc.all))
@@ -38,7 +37,7 @@ names(cc.all) <- gsub("^ethnic$", "ethnic.old", names(cc.all))
 names(cc.all) <- gsub("^CAREHOME$", "care.home", names(cc.all))
 names(cc.all) <- gsub("^simd$", "SIMD.quintile", names(cc.all))
 
-## exclude controls already dead on date of testing case they were matched to
+## exclude controls already dead on date of test of case they were matched to
 controls.deceased <- with(cc.all, CASE==0 &
                                   !is.na(Date.Death) &
                                   Date.Death <= SPECDATE)
[…]

### 2020-05-13 Covid_ethnicity_Scotland.Rmd
ethnicity report

---
title: "Briefing paper on Ethnicity and COVID-19 status in Scotland"

author: Modelling & Research Cell, Health Protection Scotland

date: "`r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}

output:
  pdf_document

# rmarkdown::render("Covid_ethnicity_Scotland.Rmd")

---

# Background
There have been anecdotal reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.  A preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95%CI 2.92-5.12); RR 2.11 (95%CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively)

Understanding whether such inequity in risk of COVID-19 by ethnicity pertains in Scotland and whether disease severity is worse is relevant to early anticipation of health care needs, triaging algorithms being used by NHS Hubs, and shielding policies.

# Methods 
All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems. This system records CHI but not ethnicity.

A matched set of 10 controls per case matched for age, sex and GP practice was identified using the CHI database as part of the REACT-COVID19 case control study led by the Modelling & Research cell.

## Coding of ethnicity
Names of cases and controls were available to the CHI team. They used the ONOMAP software to assign predicted ethnicity to all cases and controls. This was output in anonymised format to the Modelling & Research Cell linkage team. ONOMAP is a software package developed by Mateos at UCL which used people's first and family names to estimate underlying ethnic group. It has been developed using algorithms built from the Worldnames database that covered both names and ethnic group in a large number of countries. ONOMAP assigns both predicted geographical region of origin and ethnicity. ONOMAP has previously been reported to have variables sensitivity for assigning non white ethnic groups 17-89%. ([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were allocated to ethnic groups. Names classified "Muslim" in the ONOMAP classification were assigned to the South Asian group, as Scotland most people with Muslim names are South Asian.  The ONOMAP classification assigns a high proportion of names in Scotland to the category "Celtic" which does not distinguish Irish from Scottish names so that White Irish cannot be differentiated from White British.

Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of mandatory ethnic group in recent years. For all cases and controls the linkage team extracted i) the most recent ii) the most frequent ethnic group assignation from these two datasets  with a 10 year lookback period. Unknown or missing codes were excluded.  Concordance between the most recent SMR ethnic group   and the most frequent  SMR ethnic group was 98%.

A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded.  (Clark & Buchanan ISD 2017)
The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.

# Results

## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records

```{r ethnic, echo=FALSE, warning=FALSE, message=FALSE}

pander(table.ethnic, missing=".",
       split.tables=Inf,
       split.cells=rep(16, 6), 
       caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)")

```

**Amanda see line below for example of how to embed numbers from R in text**

Table \ref{tab:crosstab} shows that the ONOMAP name classification assigns to the same as self-reported category `r testpositives.ethnic[1, 1]`

of those who self-report their ethnicity as White, 92% of those who self-report as South Asian, 71% of those who self-report as Chinese, but only 21% of those who self-report as Black.

In all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.  Analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records.

## Relation of severity of COVID-19 to ethnicity among test-positive cases

```{r testpositives.ethnic, echo=FALSE, warning=FALSE, message=FALSE}

pander(testpositives.ethnic, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")

```

```{r testpositives.ethnic.smr, echo=FALSE, warning=FALSE, message=FALSE}

pander(testpositives.ethnic.smr, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")

```

## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population

```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.demog.aug) <- gsub("_", " ", rownames(table.demog.aug))
knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```

As shown above the numbers of persons testing positive from non-White ethnic groups is low. There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites. The frequency of severe disease was slightly lower in South Asans but with these small numbers this estimate is unstable.

# Conclusion
Although the numbers are small, there is no evidence that the main ethnic minorities in Scotland are at higher risk of severe COVID-19 than the  White  population. The point estimates for severe disease are lower in the main minority ethnic groups than in Whites.  However power to detect anything other than very large differences is limited as yet given the low numbers of non-White persons tested positive.  Therefore we will continue to monitor the situation.

Recording of ethnicity in hospital admission does not provide complete coverage of the population at risk of COVID-19.  The coding is also incomplete with substantial missingness.  Merging ethnicity data from the Scottish Morbidity Record with name based classification would introduce differential misclassification of ethnicity by health status.  ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.  Access to primary care data would likely increase the completeness with which ethnicity is assigned but governance issues need resolution.

A possible explanation for reports from England that people from minority ethnic groups were over-represented among cases of severe and fatal disease may be that these groups make up a high proportion of front-line health care workers and care givers in England, who were heavily exposed to infectious cases.  In Scotland a "healthy migrant effect", such that those who migrate tend to be fitter than those who do not migrate, may explain the low rates of severe disease in minority groups.

### 2020-05-13 pharmaco.Rmd
first draft of pharmacoepi paper

---
title: 'Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf")

---

```{r functions, echo=FALSE}

source("helperfunctions.R")

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(knitr)
library(kableExtra)

```

# Abstract

_Background_-- We previously  reported that across all ages having any hospital prescription  in the past year was strongly and highly significantly associated with severe COVID-19 independently of  high risk conditions. Whilst this may reflect associated condition effects it strength was sufficient to suggest direct drug effects on disease. Here we sought to further understand this phenomenon by  systematically testing associations with prescribing history with adjustment for relevant confounding by condition.

_Methods_ -- We linked the records of those tested positive for SARS-C0V-2 viral RNA in the national database that captures  virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data. Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test. We selected 10 sex, age and general practice based controls per test-positive case. For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year.  Filtering on which BNF chapters showed  association we  further tested for associations at   BNF section and subsection level. Conditional logistic regression was used to estimate odds ratios and to adjust for potential confounders.

_Findings_ -- Among those without conditions listed as risk  conditions by the National Health Service  (Diabetes, heart disease,  asthma or COAD, chronic liver disease, chronic renal failure , neurological including dementia, immune disorders or suppression)  severe cases were  highly significantly more likely than controls  to have had prescriptions in the past year from each BNF chapter. The strongest  independent associations were seen with Gastro-intestinal,  INSERT OR cardiovascular INSERT OR , Central nervous  INSERT OR  and Nutrition and Blood systems INSERT OR . When each of these were further disaggregated to the next level, key findings included a significantly greater risk of severe COVID-19 associated with proton pump inhibitor use independently of co-prescription of non-steroidal anti-inflammatory or anti-platelet agents.  ACE-inhibitors INSERT and ARBs INSERT  showed  weaker associations  with severe disease than loop diuretics  INSERT or calcium channel blockers INSERT  and  all these associations  INSERT were / were not significant on adjust for any circulatory disorder beyond heart disease. Antipsychotic drugs showed increased risk INSERT OR that was/ was not independent of psychotic  conditions. Oral iron supplements showed associations INSERT  but these were/ were not independent of prior neoplasm or renal failure INSERT.    The OR associated with any prior prescription was reduced from xx-xxx on adjusting for these specific sub-chapters ???

*Conclusions* -- These prescribing  data emphasise the multimorbidity nature of severe COVID-19 disease beyond the widely accepted risk conditions. The strong independent association with PPIs , clearly observable even in those < 40 years, combined with  prior data that this drug class is associated with gastro-enterological infections and pneumonia and the frequency of prescription of these drugs requires urgent evaluation in other datasets.

\newpage

# Background
A striking preliminary finding of the REACT-SCOT study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year. The rate ratio associated with this variable varied from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.  In this paper we investigate possible explanations of these associations explanations of this finding

# Methods

The methods of the REACT-SCOT study are described in detail in the accompanying paper.

## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.

Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensured complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material. In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.

## Statistical methods
For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.

## Information governance
This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.

# Results

We partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.  
Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.

** redo this table to show how much the association with scrip.any is reduced

The BNF chapter with the strongest association with severe disease was Chapter 1 (gastrointestinal).  Other chapters associated with severe disease were Chapter 2 (cardiovascular) and Chapter 4 (central nervous system).

## Drugs for gastro-intestinal system: BNF chapter 1
Table \ref{bnfchapter1} shows that encashed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19.  Of these by far the most commonly prescribed drug classes were proton pump inhibitors and osmotic laxatives.

Table \ref{protonpump} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".

## Drugs for cardiovascular system: BNF chapter 2
Table \ref{bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics and calcium-channel blockers.  All commonly-used drug classes used to treat hypertension were associated with moderately elevated risk of severed disease, with rate ratios from ranging from about 1.4 to 2.2.  The rate ratios associated with ACE inhibitors and angiotensin-II receptor blockers were in the middle of this range. Prescriptions of anticoagulant, antiplatelet and lipid-lowering drugs were associated with severe disease in univariate analyses, but these association were weaker in a multivariate analysis.

# Discussion

## Possible drug effects
.

## Methodological strengths and weaknesses

## Conclusion

# References {-}

<div id="refs"></div>

```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.drugs.aug) <- gsub("_", " ", rownames(table.drugs.aug))
knitr::kable(table.drugs.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="drugs",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.drugs.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in those without listed conditions") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter1,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter1",
             col.names=c(clean.header(colnames(table.bnfchapter1)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 1, in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.protonpump,
             escape=FALSE, 
             booktabs=TRUE,
             label="protonpump",
             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
			 "Rate ratio (95\\% CI)", "p-value",
                        "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with proton pump inhibitors by age groups, unadjusted and adjusted for NSAIDs, antiplatelet drugs and hospital diagnosis of a disorder of esophagus, stomach or duodenum") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Unadjusted"=2, "Adjusted"=2)) %>%
    landscape()
```

```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter2,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter2",
             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 2, in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    group_rows("", 1, nrow(table.bnfchapter2)) %>%
    landscape()

```

```{r bnfchapter10, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter10,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter10",
             col.names=c(clean.header(colnames(table.bnfchapter10)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 10 in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    group_rows("", 1, nrow(table.bnfchapter10)) %>%
    landscape()

```

\newpage

# Appendix
## Supplementary tables

### 2020-05-13 casecontrol.Rmd
multiple updates

@@
 
-table.densities <- rbind(summary(listed.densities),
+table.densities <- rbind(summary(demog.densities),
+                         summary(listed.densities),
                          summary(full.densities))[, 1:6]
@@
 In the case-control analysis, the odds ratios were
-
 `r table.agegr.all[1, 3]` for residence in a care home status, 
@@
 
-Overall but INSERT xx% of severe cases versus INSERT xx% of controls had at least one of these INSERT ( xx versus xx% in those aged under 40 years).
+Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 2]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years).
 The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
@@
 
-In comparison with a baseline model based on care home residence only, adding listed conditions increased the information for discrimination to  INSERT 0.9 bits.
-Adding indicator variables for ICD-10 and BNF chapters increased the information for discrimination further to  INSERT 1.5 bits.
-From a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
+With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
+The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed condtions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
 
-*Conclusions* --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and the vast majority of younger persons developing severe disease have an underlying medical condition.
+_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and most younger persons developing severe disease have an underlying medical condition.
 The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.
@@
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland.
-Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
-and cause of death were obtained from linkage to the National Register of Scotland.
+Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
 
-We classified test-positive cases into three groups:
-
-1.
-Severe or fatal COVID-19, defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test regardless of the cause of death given on the death certificate.
-
-2. Hospitalised but not severe as defined above
-
-3. Not hospitalised
-
-By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
+Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
+By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensured complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
 
@@
 For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
-Note that this is an incidence density sampling design i.e.
-a control who becomes a case at a later point is correctly retained in the control set.
-With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.
+As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
+With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.
 
 ## Demographic data
-Socioeconomic status was scored using the postcode based indicator  the Scottish Index of Multiple Deprivation using the postcode in the CHI database.
-Ethnic group was captured in two ways: self designated ethnicity  was derived from past records in the hospital admissions and outpatient attendances  (Scottish Morbidity Records) using  categories based on those used in the national census.
-However since this is incomplete  we also used the a name classification algorithm ONOMAP [@lakha_name_2011] with the names in the CHI database.
-This is able to distinguish South Asian and Chinese groups but not Black-Caribbean people from the White group.
-Residential care home status was available via the CHI database.
+Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.
+Ethnicity was assigned based on applying a name classification algorithm ONOMAP [@lakha_name_2011] to the names in the CHI database.
+This classification is able to distinguish South Asian and Chinese groups with fair accuracy but does not distinguish those who identify as Black or Black-Caribbean from those who identify as White.
+For those cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census.
 
 ## Morbidity and drug prescribing
-For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
+For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
 British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
@@
 
-We first extracted history of a specific list of conditions and drug classes that have been  included as  risk conditions for COVID-19 by  public health agencies https://www.nhs.uk/conditions/coronavirus-covid-19/people-at-higher-risk-from-coronavirus/.
+We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
 These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
-The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.
+The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material.
 In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.
-We did not  specifically  tabulate  the conditions that have been listed as making people " clinically extremely vulnerable"  such as cystic fibrosis or leukaemia  since there is no  clinical equipoise  regarding risk in such persons and since they will be underrepresented in the case data as they have been in complete shielding since early in the epidemic.
+We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic.
 
 ## Statistical methods
-We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020.
-(CHECK)  The univariate association of  sociodemographic variables, pre-defined conditions and discharge and prescribing history  with severe COVID-19 were modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit` We examined the associations by age band and overall.
+To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.
+For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.
 Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.
-To construct a risk prediction models, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.
+To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.
 The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.
@@
 ## Information governance
-This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data.
+This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
 Datasets were de-identified before analysis.
@@
 Still needed 
-
 - A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code
 - Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
--  Run table 2 analyses for all subjects combined 
-- what is currently tables 3,4,5  and 7  could be put into appendix and the table reference in text changed as appropriate ?
-- tables currently 9,11,12 should be removed as will go in pharmacoepi  paper  --->
+--->
 ## Incidence and mortality from severe COVID-19 in the Scottish population
-Logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.
 Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.
-The relationship of mortality to age was almost exactly linear on a logit scale.
-For fatal and non-fatal cases combined, the mortality differentials between males and females widened in middle age before narrowing again after about age 75 years.
+The relationship of mortality to age was almost exactly linear on a logit scale, and the lines for males and females were almost exactly parallel.
+For severe cases as defined in this study, the incidence appears to rise more slowly from age 50 to 70 years in females than in males, implying that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in this age range.
+This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.
 In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.
-The odds ratio associated  with male sex was is INSERT  for all severe disease and  INSERT  for fatal disease.
-As shown in Table INSERT  almost all of the severe cases  death over aged 75years died  but  the majority of severe cases under aged 60 survived.
+The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.
+As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.
 
@@
 ## Associations with listed conditions
-Overall  INSERT xx of cases and xx of controls had at least one of the listed conditions.
-The  prevalence of the listed conditions in cases and controls by age band is shown in Table  INSERT .
-Almost half of cases aged under 40 years had at least one listed condition, compared with only 9% (43 / 470) of controls.
-In those 75+ years INSERT xxx 80% of cases had at least one listed condition but  INSERT xx% of controls also did.
+Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.
+The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
+`r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.
+In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 6])` of the cases and `r gsub("\\\\", "", table.agegr[13, 5])` of the controls had at least one listed condition.
 
@@
 All of the listed conditions were significantly associated with severe disease  in each age band.
-Diabetes was associated with a 6.6.
-fold rate of severe COVID-19 under age 60 years.
+Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.
 Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.
@@
 In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-In a multivariate analysis of all chapters, the most strongly associated chapter-level diagnosis was Chapter 2 (neoplasms).
-This was further disaggregated with subchapters in as shown in Supplement  Table \ref{tab:icdchapter2}.
-Unexpectedly, the strongest association was with the subchapter "Malignant Neoplasms of Digestive Organs".
-Several other subchapters corresponding to cancer sites were also associated with severe disease.
+Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters: ..
+note associations with previous pneumonia and stroke.
 
@@
 Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.
-The rate ratio associated with this variable varies from INSERT 9.2 in those aged under 60 years to INSERT 42.2 in those aged 75 years and over.
-In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to  INSERT3.8 and INSERT11.9 respectively.
-It is notable that about one third of those over 75 had not encashed a prescription in the previous year.
+The rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
+In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.
 
@@
 Observed and predicted case status were compared over all test folds.
-Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[1, 3]` to `r table.densities[2, 3]` and the expected information for discrimination from `r table.densities[1, 5]` bits to `r table.densities[2, 5] ` bits.
+Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.
 
-This value for the information conditional on age and sex can be added to the information for discrimination obtained with a logistic regression model using age and sex to obtain the total information for discrimination that would be obtained in the population  to give a total of WHAT??.
-PAUL HOW MUCH IS THAT THIS IS NOT CLEAR
+This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier in the population as that would be obtained in the population to give a total of `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
 
-Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence favouring case over <!--thi figure should go into a supplement--> control status from the model based on the extended variable set.
+Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set.
 This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
+Figure \ref{fig:roc}, which shows the corresponding receiver operating characteristic (ROC) curve, encodes exactly the same information in a different form.
+The likelihood ratio $2^W$ at any given sensitivity and specificity is the gradient of the model-based ROC curve.
 
@@
 
+```{r plotroc, echo=FALSE, fig.cap="\\label{fig:roc}Cross-validation of model chosen by stepwise regression using extended variable set: receiver operating characteristic curve", out.width="0.8\\textwidth"}
+
+plotroc(full.densities) +
+    scale_color_manual(values=c("orange", "purple")) + 
+    coord_fixed(ratio=1) +
+    theme(axis.title=element_text(size=18)) +
+    theme(legend.title=element_blank()) + 
+    theme(legend.position=c(0.8, 0.2)) + 
+    scale_x_reverse(limits=c(1, 0), expand=c(0, 0)) +
+    scale_y_continuous(limits=c(0, 1), expand=c(0, 0)) +
+    xlab("Specificity") +
+    ylab("Sensitivity (reverse scale)")
+
+```
+
 \newpage
@@
 
-knitr::kable(table.demog.aug,
+knitr::kable(table.demog.aug[, 1:4],
              escape=FALSE, 
@@
              col.names=c(colnames(table.demog.aug)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
                          "Rate ratio (95\\% CI)", "p-value"),
@@
              caption="Univariate associations of severe disease with demographic factors") %>%
-    column_spec(column=1, width="3cm") %>%
+    column_spec(column=1, width="2.5cm") %>%
+    column_spec(column=2:4, width="2.5cm") %>%
     group_rows("Ethnicity", 1, 3) %>%
@@
              caption="Associations of severe disease with listed conditions over all age groups") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-
-```
-
-```{r listed.conditions.lt60, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(tables.agegr[[1]],
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="listed.conditions.lt60",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(tables.agegr[[1]])[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-
-```
-```{r listed.condition.60to74, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(tables.agegr[[2]],
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="listed.conditions.60to74",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(tables.agegr[[2]])[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-
-```
-```{r listed.condition.75plus, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(tables.agegr[[3]],
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="listed.conditions.ge75",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(tables.agegr[[3]])[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
     column_spec(column=1, width="4cm") %>%
@@
 
-```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter1,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter1",
-             digits=c(0, 0, 0, 200 , 0, 200), 
-             col.names=c(colnames(table.bnfchapter1)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 1, in those without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-```
-
-```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.protonpump,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="protonpump",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(table.drugs.aug)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Univariate associations of severe disease with proton pump inhibitors in all cases and controls, by age group") %>%
-    column_spec(column=1, width="2cm")
-```
-
-```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter2,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter2",
-             col.names=c(colnames(table.bnfchapter2)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 2, in those without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter2)) %>%
-    landscape()
-
-```
-
-```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter4,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter4",
-             col.names=c(colnames(table.bnfchapter4)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 9, in those without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter4)) %>%
-    landscape()
-
-```
-
 ```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.densities) <- c("Care home + listed conditions",
+rownames(table.densities) <- c("Demographic only",
+                               "Demographic + listed conditions",
                                "Extended variable set")
@@
 
-```{r model, echo=FALSE, warning=FALSE, message=FALSE}
+```{r listed.conditions.lt60, echo=FALSE, warning=FALSE, message=FALSE}
 
-stepwise.full <- as.data.frame(stepwise.full[, c(1, 5)])
-stepwise.full[, 1] <- round(stepwise.full[, 1], 2)
-stepwise.full[, 2] <- signif(stepwise.full[, 2], 1)
-rownames(stepwise.full) <- replace.names(rownames(stepwise.full))
-rownames(stepwise.full) <- gsub("\\_", " ", rownames(stepwise.full))
+knitr::kable(tables.agegr[[1]],
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.conditions.lt60",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(tables.agegr[[1]])[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
 
-stepwise.full[, 2] <- pvalue.latex(stepwise.full[, 2])
+```
+```{r listed.condition.60to74, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(stepwise.full, 
+knitr::kable(tables.agegr[[2]],
+             escape=FALSE, 
              booktabs=TRUE,
-             escape=FALSE,
-             col.names=c("log rate ratio", "p-value"), 
-             label="stepwise",
-             align=rep("c", 2),
-             caption="Stepwise regression: variables retained in model for severe disease") %>%
-    group_rows("Diabetes", 3, 5)
+             label="listed.conditions.60to74",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(tables.agegr[[2]])[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
 
@@
 
-<!--
+```{r listed.condition.75plus, echo=FALSE, warning=FALSE, message=FALSE}
 
-For each of these chapters, we examined univariate and multivariate associations with severe disease at the level of paragraph or subparagraph of the BNF classification, restricting the analysis to those without one of the listed conditions.
+knitr::kable(tables.agegr[[3]],
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.conditions.ge75",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(colnames(tables.agegr[[3]])[1:2],
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
 
-Table \ref{tab:bnfchapter1} shows that prescriptions of most classes of drugs for gastrointestinal conditions were associated with severe COVID-19.
-By far the most widely prescribed drug class in this age group was proton pump inhibitors; 15% of controls and 36% of cases had encashed at least one prescription, with a univariate rate ratio of 3.2.
-In a multivariate analysis, the drug class most strongly associated with severe disease was proton pump inhibitors.
+```
 
-Table \ref{tab:protonpump} shows the association of severe disease with proton pump inhibitors in all cases and controls by age group.
-The univariate rate ratio varied from 12 in those aged under 60 years to 2.8 in those aged 75 years and over.
+```{r icdsubchapters, echo=FALSE, warning=FALSE, message=FALSE}
+clean.header <- function(x) {
+    x <- gsub("\\.\\.", " (", x)
+    x <- gsub("\\.",  ")", x)
+    return(x)
+}
 
-Table \ref{tab:bnfchapter2} shows the associations with classes of drugs prescribed for the cardiovascular system.
-Most classes of antihypertensive drug were associated with increased risks in univariate analysis, but there was no evidence that the risks associated with prescription of ACE inhibitors or angiotensin-II receptor antagonists differed from those for other drugs.
-Prescription of anticoagulants was associated with increased risk of severe disease.
-presumably reflecting the risks associated with the underlying cardiovascular conditions.
-In a multivariate analysis the strongest association of severe disease was with loop diuretics.
-Table \ref{tab:bnfchapter4} shows the associations with classes of drugs prescribed for the central nervous system.
-The strongest association with severe disease in a multivariate analysis was with prescription of non-opiate analgesics.
+knitr::kable(table.icdsubchapters,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="icdsubchapters",
+             col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 20 cases and controls") %>%
+    column_spec(column=1, width="13cm") %>%
+    column_spec(column=2:3, width="1.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    landscape()
 
-A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year.
-This association was only partly explained by associations with listed conditions.
-Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
+```
 
-Examination of the subchapters within this chapter showed that this association was largely explained by prescriptions of proton pump inhibitors, which had been encashed by 15% of controls.
-Over all age groups the rate ratio associated with use of proton pump inhibitors was 2.8, but in those aged less than 40 years, in whom serious underlying disease is presumaby less common, the rate ratio was 12.
-This association is not explained by association any underlying gastrointestinal disease that leads to hospital admission, by coprescribing with dual antiplatelet agents in people who have had a myocardial infarction or a vascular procedure.
-As proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.
-However as there are no prescription charges in Scotland, long-term users of these expensive drugs are likely to seek to obtain them on prescription.
+```{r model, echo=FALSE, warning=FALSE, message=FALSE}
 
-As proton pump inhibitors are commonly co-prescribed with NSAIDS, this is relevant to evaluating possible explanations for the association of proton pump inhibitors with NSAIDs. (CHECK).
+retained.full <- as.data.frame(stepwise.full[, c(1, 5)])
+retained.full[, 1] <- round(retained.full[, 1], 2)
+retained.full[, 2] <- signif(retained.full[, 2], 1)
+rownames(retained.full) <- replace.names(rownames(retained.full))
+rownames(retained.full) <- gsub("\\_", " ", rownames(retained.full))
 
-As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
-Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016].
-35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013].
+retained.full[, 2] <- pvalue.latex(retained.full[, 2])
 
--->
+knitr::kable(retained.full, 
+             booktabs=TRUE,
+             escape=FALSE,
+             label="stepwise",
+             col.names=c("log rate ratio", "p-value"), 
+             align=rep("c", 2),
+             caption="Stepwise regression: variables retained in model for severe disease") %>%
+    group_rows("Diabetes", 6, 8)
+
+```

### 2020-05-13 bnfcodes.R

@@ -13,16 +13,16 @@
 ## 1001030C0 is hydroxychloroquine
 ## 1001030U0 is methotrexate
 
-bnfcodes <- read_excel("./data/BNF_Code_Information.xlsx", sheet=4)
+bnfcodes <- read_excel("./BNF_Code_Information.xlsx", sheet=4)
 colnames(bnfcodes) <- c("chaptername", "chapternum", "sectionname", "sectioncode")
 
-bnfparacodes <- read_excel("./data/BNF_Code_Information.xlsx", sheet=3)
+bnfparacodes <- read_excel("./BNF_Code_Information.xlsx", sheet=3)
 colnames(bnfparacodes) <- c("chaptername", "chapternum",
                             "sectionname", "sectioncode",
                             "paraname", "paracode")
 bnfparacodes$paracode <- as.integer(bnfparacodes$paracode)
 
-bnfsubparacodes <- read_excel("./data/BNF_Code_Information.xlsx", sheet=1)[, 1:8]
+bnfsubparacodes <- read_excel("./BNF_Code_Information.xlsx", sheet=1)[, 1:8]
 colnames(bnfsubparacodes) <- c("chaptername", "chapternum",
                                "sectionname", "sectioncode",
                                "paraname", "paracode",

### 2020-05-13 casecontrol.R

@@ -15,14 +15,13 @@ library(DescTools)
 library(icd.data)
 library(gam)
 
-
 source("helperfunctions.R")
         
 cc.all <- readRDS("./data/CC_linked_ANON_20200501.rds")
 
 diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 records ? excluded
 procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
-scotpop <- read_excel("../Scotland_midyearpop_est2019.xlsx")
+scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
 
 ## we have only 7 digits on scrips, giving resolution to subpara level only
 scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
@@ -148,7 +147,7 @@ cc.all <- within(cc.all, diabetes.any <- relevel(diabetes.any, ref="Not diabetic
 cc.all$dm.type <- 
   as.factor(car::recode(cc.all$dm.type, 
                         "0='Not diabetic'; 1='Type 1'; 2='Type 2';
-                         3:hi='Other diabetes type'"))
+                         3:100='Other diabetes type'; 101:102='Type 1'; 202:203='Type 2'"))
 cc.all <- within(cc.all, dm.type <- relevel(dm.type, ref="Not diabetic"))
 
 ########################################################################
@@ -245,7 +244,7 @@ cc.severe$num.icdchapters <-
     as.factor(car::recode(cc.severe$num.icdchapters,
                           "0='No discharge records'; 1:2='1-2 ICD-10 chapters'; 3:4='3-4 chapters'; 5:hi='5 or more chapters'")
                          )
-cc.all <- within(cc.severe, num.icdchapters <- relevel(num.icdchapters, ref="No discharge records"))
+cc.severe <- within(cc.severe, num.icdchapters <- relevel(num.icdchapters, ref="No discharge records"))
 
 ############ extract predefined disease categories #################
 
@@ -630,6 +629,19 @@ table.icdchapter11 <- tabulate.icdchapter(chnum=11, data=cc.notlisted)
 
 table.icdchapter7 <-  tabulate.icdchapter(chnum=7, data=cc.notlisted)
 
+tabulate.icdchapter(chnum=8, data=cc.notlisted)
+
+table.icdsubchapters <- NULL
+for(i in 1:20) {
+    table.icdsubchapters <-
[…]

### 2020-05-13 helperfunctions.R

@@ -1,6 +1,12 @@
 
 ## helperfunctions for COVID analyses
 
+clean.header <- function(x) {
+    x <- gsub("\\.\\.", " (", x)
+    x <- gsub("\\.",  ")", x)
+    return(x)
+}
+
 tabulate.freqs.regressions <- function(varnames, outcome="CASE", data) {
     ## FIXME: first run should identify sparse variables to be dropped from varnames
     table.freqs <- univariate.tabulate(varnames=varnames, outcome=outcome, data=data,
@@ -90,7 +96,7 @@ multivariate.clogit <- function(varnames, outcome="CASE", data, add.reflevel=FAL
     return(multivariate.table)
 }
 
-tabulate.bnfsection <- function(chnum, outcome="CASE", data=cc.severe) {
+tabulate.bnfsection <- function(chnum, outcome="CASE", data=cc.severe, minrowsum=20) {
     ## get sectioncode in wide format, one col per subchapter
     scrips.ch.wide <- reshape2::dcast(scrips[scrips$chapternum==chnum, ],
                                          ANON_ID ~ sectioncode, fun.aggregate=length,
@@ -114,7 +120,7 @@ tabulate.bnfsection <- function(chnum, outcome="CASE", data=cc.severe) {
     bnf.ch <- colnames(cc.bnf.ch)[-(1:3)]
     ## FIXME: regressions should be fixed to use only rows retained by univariate.tabulate
     table.bnf.ch <- univariate.tabulate(varnames=bnf.ch, outcome=outcome, data=cc.bnf.ch,
-                                        drop.sparserows=TRUE)
+                                        drop.sparserows=TRUE, minrowsum=20)
     if(nrow(table.bnf.ch) >= 1) { 
         bnf.ch <- rownames(table.bnf.ch)  
         subsectioncodes <- as.integer(gsub("section.", "", rownames(table.bnf.ch)))
@@ -282,7 +288,7 @@ tabulate.bnfchemicals <- function(chnum, outcome="CASE", data=cc.severe) {
     }
 }
 
-tabulate.icdchapter <- function(chnum, data=cc.severe) {
+tabulate.icdchapter <- function(chnum, data=cc.severe, minrowsum=20) {
     ## get chapternum in wide format, one col per subchapter
     diagnoses.ch.wide <- reshape2::dcast(diagnoses[diagnoses$chapter==chnum, ],
[…]

### 2020-05-14 casecontrol.Rmd
Corrected author names.

@@
     affiliation: HPS
-  - name: Stuart McGurnaghan
+  - name: Stuart J McGurnaghan
     affiliation: IGMM
@@
     affiliation: HPS 
-  - name: Tom Caparrotta 
+  - name: Thomas M Caparrotta 
     affiliation: IGMM

### 2020-05-14 Covid_ethnicity_Scotland.Rmd
updated for ethnicity report

@@
 output:
-  pdf_document
+  pdf_document:
+    keep_tex: true
 
@@
 
-A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded.
-(Clark & Buchanan ISD 2017)
+A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded (Clark & Buchanan ISD 2017).
 The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.
@@
 # Results
-
 ## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records
@@
 
-**Amanda see line below for example of how to embed numbers from R in text**
-
-Table \ref{tab:crosstab} shows that the ONOMAP name classification assigns to the same as self-reported category `r testpositives.ethnic[1, 1]`
-
-of those who self-report their ethnicity as White, 92% of those who self-report as South Asian, 71% of those who self-report as Chinese, but only 21% of those who self-report as Black.
+Table \ref{tab:crosstab} shows that the ONOMAP name classification assigns to the same as self-reported category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[5, 5]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.
 
@@
 Analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records.
+
+## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population
+
+```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
+knitr::kable(table.severe.demog,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="onomap",
+             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
+    column_spec(column=1, width="4.5cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+```
+
+```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
+knitr::kable(table.ethnicsmr.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmr",
+             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    column_spec(column=1, width="4.5cm")
+```
+
+As shown above the numbers of persons testing positive from non-White ethnic groups is low.
+There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites.
+The frequency of severe disease was lower in South Asans but with these small numbers this estimate is unstable.
+
+## Relation of hospitalised COVID-19 to ethnicity in the Scottish population
+
+```{r onomap.hosp, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
+knitr::kable(table.demog.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="onomap.hosp",
+             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
+    column_spec(column=1, width="4.5cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape()
+```
+
+```{r ethnicsmr.hosp, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
+knitr::kable(table.hospitalized.demog.ethnicsmr,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmr.hosp",
+             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    column_spec(column=1, width="4.5cm")
+```
+
+# Conclusion
+Although the numbers are small, it is possible to establish that South Asians in Scotland are at lower risk of severe disease than the White population.
+
+For the Black population, reliable classification of ethnicity is restricted to those who have an SMR record, and the numbers of cases and controls in this group are too small for rate ratios to be reliably estimated.
+
+ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.
+Recording of ethnicity in hospital admission distinguishes these groups but does not provide complete coverage of the population at risk of COVID-19.
+The coding is also incomplete with substantial missingness.
+We have not merged ethnicity data from the Scottish Morbidity Record with name based classification as this would introduce differential misclassification of ethnicity by health status.
+Access to primary care data would likely increase the completeness with which ethnicity is assigned but governance issues have not been resolved.
+
+A possible explanation for reports from England that people from minority ethnic groups were over-represented among cases of severe and fatal disease may be that these groups make up a high proportion of front-line health care workers and care givers in England, who were heavily exposed to infectious cases.
+In Scotland a "healthy migrant effect", such that those who migrate tend to be fitter than those who do not migrate, may explain the low rates of severe disease in minority groups.
+
+\newpage
+
+# Supplementary tables
+These tables including test-positive individuals who were not hospitalized are included by request.
+They should be interpreted with caution, as selection for testing is influenced by many factors including health care worker status.
 
@@
        split.tables=Inf,
-       split.cells=rep(25, 3), 
        caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")
@@
 
-## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population
+```{r testpositives.carehome, echo=FALSE, warning=FALSE, message=FALSE}
 
-```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}
+pander(testpositives.carehome, missing=".",
+       split.tables=Inf,
+       caption="Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")
 
-rownames(table.demog.aug) <- gsub("_", " ", rownames(table.demog.aug))
-knitr::kable(table.demog.aug,
+```
+
+```{r testpositives.healthboard, echo=FALSE, warning=FALSE, message=FALSE}
+
+pander(testpositives.healthboard, missing=".",
+       split.tables=Inf,
+       caption="Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+
+```
+
+## Relation of test-positive status to ethnicity in the Scottish population
+
+```{r onomap.testpos, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.testpositives.demog) <- gsub("_", " ", rownames(table.testpositives.demog))
+knitr::kable(table.testpositives.demog,
              escape=FALSE, 
              booktabs=TRUE,
-             label="onomap",
+             label="testpos.onomap",
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
+             col.names=c(clean.header(colnames(table.testpositives.demog)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
+             caption="Associations of test-positive status with ethnicity based on name classification") %>%
     column_spec(column=1, width="4.5cm") %>%
@@
 
-```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
+```{r ethnicsmr.testpos, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
+rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
 knitr::kable(table.ethnicsmr.aug,
@@
              booktabs=TRUE,
-             label="ethnicsmr",
+             label="testpos.ethnicsmr",
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
+             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value"),
              align=rep("c", 6), 
-             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
     column_spec(column=1, width="4.5cm")
 ```
-
-As shown above the numbers of persons testing positive from non-White ethnic groups is low.
-There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites.
-The frequency of severe disease was slightly lower in South Asans but with these small numbers this estimate is unstable.
-
-# Conclusion
-Although the numbers are small, there is no evidence that the main ethnic minorities in Scotland are at higher risk of severe COVID-19 than the  White  population.
-The point estimates for severe disease are lower in the main minority ethnic groups than in Whites.
-However power to detect anything other than very large differences is limited as yet given the low numbers of non-White persons tested positive.
-Therefore we will continue to monitor the situation.
-
-Recording of ethnicity in hospital admission does not provide complete coverage of the population at risk of COVID-19.
-The coding is also incomplete with substantial missingness.
-Merging ethnicity data from the Scottish Morbidity Record with name based classification would introduce differential misclassification of ethnicity by health status.
-ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.
-Access to primary care data would likely increase the completeness with which ethnicity is assigned but governance issues need resolution.
-
-A possible explanation for reports from England that people from minority ethnic groups were over-represented among cases of severe and fatal disease may be that these groups make up a high proportion of front-line health care workers and care givers in England, who were heavily exposed to infectious cases.
-In Scotland a "healthy migrant effect", such that those who migrate tend to be fitter than those who do not migrate, may explain the low rates of severe disease in minority groups.

### 2020-05-14 casecontrol.Rmd
updated for ethnicity report

--- 2020-05-13 casecontrol.Rmd
@@
 ---
-title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population'
+title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]'
 author: 
@@
 ```{r functions, echo=FALSE}
-
-source("helperfunctions.R")
+library(knitr)
+library(kableExtra)
 
@@
 
-library(knitr)
-library(kableExtra)
+source("helperfunctions.R")
 
@@
 
+\newpage
+
 # Abstract
 
-_Background_-- Why some people infected with Sars-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.
+_Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.
 Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.
 
-_Methods_ -- We linked the records of those tested positive for SARS-C0V-2 viral RNA in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data.
+_Methods_ -- We linked the records of those with a positive nucleic acid test  for SARS-CoV-2 in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data.
 Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test.
@@
 
-_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
+_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
 In the case-control analysis, the odds ratios were
@@
 `r table.agegr.all[8, 3]` for chronic renal failure, 
-`r table.agegr.all[5, 3]` for ischemic heart disease
+`r table.agegr.all[5, 3]` for ischemic heart disease,
+`r table.agegr.all[6, 3]` for other heart disease, 
 `r table.agegr.all[9, 3]` for neurological disease, 
 `r table.agegr.all[10, 3]` for chronic liver disease and 
-`r table.agegr.all[11, 3]` for immune deficiency or suppresssion
+`r table.agegr.all[11, 3]` for immune deficiency or suppresssion.
 
-Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 2]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years).
+Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years.
 The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
@@
 With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
-The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed condtions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
+The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
 
@@
 A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.
-The strong association with prescribing history we have observed may reflect co-morbidity or adverse drug effects and we are investigating this further.
 
@@
 # Background
-It is clear that  there is wide susceptibility to risk of severe COVID-19 disease  with older age being a predominant factor.
-Case series from many countries have  suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected.
-However there are also many anecdotal reports of apparently healthy young persons succumbing to disease.
-Apart from co-morbidities the potential  role of ethnicity in disease severity has been highlighted.
-However quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator.
-In particular a systematic evaluation of the past medical history in those with and without severe disease has not been reported.
+Case series from many countries have suggested that in those with severe COVID-19 the prevalence of diabetes and cardiovascular disease is higher than expected.
+For example in a large UK series the commonest co-morbidities were cardiac disease, diabetes, chronic pulmonary disease and asthma [@docherty_features_2020].
+However there are also anecdotal reports of apparently healthy young persons succumbing to disease [@mcgoogan_why_2020].
 
-Here we  used extensive health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID -19 disease.
-We defined  severe disease as viral RNA-confirmed diagnosis of COVID-19 as  that required critical care support or was associated with death within 28 days of testing.
-We focused on severe disease it is of most  concern but also  because it can be completely ascertained  whereas  testing policies means that   complete ascertainment of less severe disease is not possible in most countries.
+The quantification of the risk associated with  characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator  [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].
+Two recent UK analyses have  had population comparators and these have reported  increased relative risks for a range of co-morbidities including diabetes, asthma and heart disease and for non-white ethnicity[@niedzwiedz_ethnic_2020;@the_opensafely_collaborative_opensafely_2020].
+A systematic evaluation of the past medical history in those with and without severe disease has not as yet been reported.
+
+We undertook a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID-19 disease.
 
@@
 With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.
+The odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
 
@@
 In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.
-We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic.
+We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia  or organ transplantation as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic.
+This group also includes those with cancer in receipt of chemotherapy or radiotherapy.
 
@@
 Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.
+Where the sample of cases and controls is restricted, this will generate strata that contain no cases.
+These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
+
 To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.
-The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.
+The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.
+The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier has been explained elsewhere [@mckeigue_quantifying_2019].
+Especially relevant to this study is that the contributions of independent predictors are additive on the scale of $\Lambda$.
+We can thus add the information for discrimination conditional on age and sex (estimated from this matched case-control study) to the information for discrimination estimated from a logistic regression model on age and sex in the general population.
 
 ## Information governance
-This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
-Datasets were de-identified before analysis.
+This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147 . Datasets were de-identified before analysis.
 
 # Results
-<!---
-Still needed 
-- A statement on "In national mortality data, the n and % of those certified with COVID-19 as underlying case of death had/ did not have a positive test record"  or sensivitivty analysis if sharon provides a case control set of the deaths with ue07 code
-- Figure - ?? graph of population attributable risk fraction of listed conditions by age ?? not sure if this is worth the trouble ...
---->
 ## Incidence and mortality from severe COVID-19 in the Scottish population
 Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.
-The relationship of mortality to age was almost exactly linear on a logit scale, and the lines for males and females were almost exactly parallel.
-For severe cases as defined in this study, the incidence appears to rise more slowly from age 50 to 70 years in females than in males, implying that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in this age range.
-This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.
-In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(logistic.coeffs[3, 2]), 2)` for fatal disease.
+The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.
+For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.
+In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(10*exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
 The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.
@@
 `r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.
-In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 6])` of the cases and `r gsub("\\\\", "", table.agegr[13, 5])` of the controls had at least one listed condition.
+In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 8])` of the cases and `r gsub("\\\\", "", table.agegr[13, 7])` of the controls had at least one listed condition.
 
@@
 ## Systematic analysis of diagnoses associated with severe disease
-The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored.
-The association of hospital diagnoses of each ICD-10 chapter with severe COVID-19, restricted to those without any of the listed conditions.
+The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored by systematically testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.
 These results are shown in Table \ref{tab:conditions}.
-In univariate analyses, almost all ICD chapters with at least one diagnosis, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters: ..
-note associations with previous pneumonia and stroke.
+In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
+The strongest associations were seen with neoplasms and with eye disease.
+Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.
+The key point of this table shows that there are many subchapter diagnoses that are significantly more frequent in severe COVID-19 cases than controls.
+Of particular note is that prior hospitalisations with infections, pneumonia and acute respiratory diseases, circulatory disorders other than heart disease, including cerebrovascular accidents.
+Without filtering on the aforementioned criteria there are of course additional differences in counts of rarer diagnoses in cases and controls are apparent.
 
@@
 Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.
-The rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
+The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
 In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and  `r tables.agegr[[3]][2, 5]`   About one third of those over 75 had not encashed a prescription in the previous year.
@@
 ## Construction of a multivariate risk prediction model 
-Predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.
+To obtain a summary metric of the additional information there is in the hospitalisation and prescribing data predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.
 
@@
 
-# Discussion  <!--helen still to edit -->
-## Multifactorial risk
+# Discussion  
+### Sociodemographic features
 In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.
 Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged.
-Being in a residential care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis.
+Regarding sex differences we noted that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in the 50-70 year age range.
+This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.
+The effect of matching on general practice which will partly match for socioeconomic disadvantage will reduce the precision of the effect of the socioeconomic indicator (SIMD) in our analysis but will not bias it in either direction.
+Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold adjusted for the listed conditions.
+This eight-fold risk likely reflects both the spread of the epidemic in care homes and some residual confounding by frailty.
 
-In contrast reports from the U.S and elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020] we found no evidence of an increased risk in non-White minorities.
-This suggests that ethnic differences found elsewhere are not genetically driven but reflect other social factors.
-In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by the health care worker status [@niedzwiedz_ethnic_2020;@ho_modifiable_2020 ].
+In contrast to reports from the US and England [@the_opensafely_collaborative_opensafely_2020] we found no evidence of increased risk in non-White minorities.
+This suggests that ethnic differences reported in other countries are not genetically driven but reflect other social factors that differ between regions in their association with ethnicity.
+We note that the proportion of health care workers that are of non-British nationality varies across the UK from [23% in London to 7% in Scotland](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/internationalmigration/articles/internationalmigrationandthehealthcareworkforce/2019-08-15).
+In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk further[@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
+In the OpenSAFELY prospective cohort study using medical records in England compared to those recorded as White, Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively adjusted for comorbidities and socioeconomic status, However as acknowledged by the authors, data on health care worker status and other social factors would be needed to explore this further.
+We are in the process of obtaining more detailed data on health care worker status in Scotland.
 
-Media reports of apparently otherwise healthy youths succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event.
-However we found that almost half of cases under 40 years had at least one condition listed by public health agencies as increasing risk.
-In those that did not have a listed condition  having any hospital admission  in the past five years and having any prescription in the past year  was much more common  among cases that in the background population of the same age.
+### Co-morbidities
+We have confirmed the increased risks associated with the listed conditions. These were based on the list of " moderate risk " conditions highlighted by the NHS and other agencies [@nhs_whos_2020].
 
-We chose to focus initially on the contribution of  conditions already proposed as high risk.
-We confirmed that these are associated with  substantially increased risk .
-However our systematic analysis has shown that there are many many  many other conditions considered in the same univariate way are also  associated with severe COVID-19 disease.
+The risk associated with these conditions various by age - for example the odds ratio associated with diabetes is much higher at younger ages.
+The 2.7 fold risk in all ages combined is commensurate with that reported in UK Biobank and in the OpenSAFELY studies.
+We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
+Of note other heart disease is more strongly associated than ischaemic heart disease.
+This category includes many conditions such as atrial fibrillation, cardiomyopathies and heart failure.
+However overall 26% of cases did not have any of these conditions.
+Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as many people under current chemotherapy and radiotherapy are included in the shielding at present.
+We have not focussed on evaluating the risk associated with conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding.
+A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather very many associated diagnoses.
+We note however that past hospitalisation for influenza and respiratory tract infections is an important risk factor.
+
+Media reports of apparently healthy young people succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event.
+However we found that almost half of cases under 40 years had at least one of the listed conditions In those that did not have a listed condition having any hospital admission in the past five years and having any prescription in the past year was much more common among cases that in the background population of the same age.
+Thus the vast majority of younger persons succumbing to severe disease are not people with no prior health conditions.
 
@@
 A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year.
-This association was only partly explained by associations with listed conditions.
-Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
-In an accompanying paper we explore these associations more fully.
+This association was only partly explained by differing prescription levels associations with the listed conditions.
+Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19 .
+Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.
+It is likely that most drug associations will simply reflect co-morbidities for which they are prescribed.
+However we note that there have been proposals of directly adverse effects of drugs such as angiotensin converting enzyme inhibitors [@fang_are_2020] although in a case control study from Italy differences in usage between cases with COVID-19 and controls were attributable to co-morbidities in a case control analysis [@mancia_reninangiotensinaldosterone_2020]
+
+Nonetheless the strong association with encashed of at least one prescription warrants a more detailed investigation.
+In an accompanying paper we explore these associations more fully since they require a detailed consideration of confounding by indication before and directly causal relationship can be asserted.
+
+## Relevance to policy
+As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.
+The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on accurate risk discrimination.
+So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.
+
+It is to be expected that as awareness grows of how risk varies between individuals, individuals will seek to learn whatever they can about their own level of risk.
+A key implication of our results is that risk of severe or fatal disease is multifactorial and that the ninefold rate ratio associated with a 20-year difference in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological conditions that impair mobility.
+A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will not be an efficient way to allocate resources to those who most need it.
+Assigning all persons with the current list of risk conditions will be too insensitive and too non-specific - a third of 60-75 years old in the background population have at least one of the listed conditions we examined.
+Half of the younger people with severe disease do not have any of them.
+Therefore it is necessary to construct a risk classifier that optimizes sensitivity and specificity based on the information available.
+Our results show that this is possible in principle, though more work is required to construct a practical risk classifier for Scotland, let alone to develop and validate classifiers for use in other populations.
 
@@
 Also we limited our case control study to test-positive cases.
-This excludes deaths attributed to COVID-19 where no testing was carried out.
-At present there are xxx such deaths in Scotland almost all aged >> xxx.
-Others have argued that some COVID-19 deaths may have been attributed to other causes early in the epidemic.
-These aspects may mean some relative risks e,g that for residential care home status have been underestimated.
-However in the younger age bands, it is likely that most severe or fatal cases will have reached hospital and been tested.
-
-## Relevance to policy
-As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.
-The "stratify and shield" policy option, in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on accurate risk discrimination.
-It is also to be expected that as awareness grows of how risk varies between individuals, individuals will seek to learn whatever they can about their own level of risk.
-A key implication of our results is that risk of severe or fatal disease is multifactorial and that the ninefold rate ratio associated with a 20-year difference in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological that impair mobility.
-A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will not be an efficient way to allocate resources to those who most need it.
-Assigning all persons with the current list of risk conditions will be too insensitive and too non-specific - a third of 60-75 years old in the background population have at least one of the listed  conditions we examined.
-Half of the younger people with severe disease don't have any of them.
-Therefore it is necessary to construct a risk classifier that optimizes sensitivity and specificity based on the information available.
-Our results show that this is possible in principle, though more work is required to construct a practical risk classifier for Scotland, let alone to develop and validate classifiers for use in other populations.
+This excludes deaths where COVID-19 is mentioned as being of suspected or probable relevance to the death but where no testing was carried out.
+Up to 13 May 2020 an additional 1200 such deaths had been reported by the [National Registers of Scotland.](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland) Future iterations of this  analysis will include sensitivity analyses of the extent to which the results are changed by including these deaths, but the data are not yet available.
+Apart from residential care home status we anticipate that the ORs for most other risk factors will be similar or less as there will be some misclassification of cause in these deaths.
 
 ## Conclusion
-In conclusion the epidemiology of severe COVID-19 disease is becoming clearer. There is not just one small set of conditions that contribute to risk but many. ......
+In conclusion the epidemiology of severe COVID-19 disease is becoming clearer  and we confirm here the importance of sociodemographic factors and  chronic conditions such as diabetes, asthma, circulatory disease and others.
+However there is not just one small set of conditions that contribute to risk but many conditions as evidenced by associations across with past medical and prescribing history across many body systems.
+As countries attempt to emerge from lockdown whist protecting the most vulnerable multivariate classifiers rather than crude rule-based approaches will be needed to define those most at risk.
 
@@
 
-```{r plotrates, echo=FALSE, fig.cap="\\label{fig:rates}Incidence of severe and fatal COVID-19 in Scotland by age and sex: generalized additive models fitted to severe and fatal cases for males nd females separately", out.width="0.8\\textwidth"}
+```{r plotrates, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:rates}Incidence of severe and fatal COVID-19 in Scotland by age and sex: generalized additive models fitted to severe and fatal cases for males and females separately", out.width="0.8\\textwidth"}
 
@@
 
-```{r plotwdists, echo=FALSE, fig.cap="\\label{fig:wevid}Cross-validation of model chosen by stepwise regression using extended variable set: class-conditional distributions of weight of evidence", out.width="0.8\\textwidth"}
+```{r plotwdists, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:wevid}Cross-validation of model chosen by stepwise regression using extended variable set: class-conditional distributions of weight of evidence", out.width="0.8\\textwidth"}
 
@@
              digits=c(0, 0, 0, 200), 
-             col.names=c(colnames(table.demog.aug)[1:2],
+             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value"),
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(table.agegr.all)[1:2],
+             col.names=c(clean.header(colnames(table.agegr.all)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(table.conditions.aug)[1:2],
+             col.names=c(clean.header(colnames(table.conditions.aug)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-
-```
-
-```{r icdchapter2, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.icdchapter2,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="icdchapter2",
-             col.names=c(colnames(table.icdchapter2)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 4), 
-             caption="Associations of severe disease with hospital diagnoses in ICD chapter 2, in those without listed conditions") %>%
-    column_spec(column=1, width="5cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2)) %>%
     landscape()
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(table.drugs.aug)[1:2],
+             col.names=c(clean.header(colnames(table.drugs.aug)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(tables.agegr[[3]])[1:2],
+             col.names=c(clean.header(colnames(tables.agegr[[3]])[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              booktabs=TRUE,
+			 longtable=TRUE, 
              label="icdsubchapters",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 20 cases and controls") %>%
+             caption="Associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
     column_spec(column=1, width="13cm") %>%
@@
 retained.full[, 2] <- signif(retained.full[, 2], 1)
+
+rownames(retained.full) <- gsub("\\.quintile", "\\.quintile ", rownames(retained.full))
+rownames(retained.full) <- gsub("([a-z])1", "\\1", rownames(retained.full))
+rownames(retained.full) <- gsub("care.home", "", rownames(retained.full))
+rownames(retained.full) <- gsub("dm\\.type", "", rownames(retained.full))
+
 rownames(retained.full) <- replace.names(rownames(retained.full))

### 2020-05-14 pharmaco.Rmd
updated for ethnicity report

@@
 
-## Case definition 
+## Case ascertainment and sampling of controls 
 The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
@@
 Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
+Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
 
-Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensured complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
-
-## Matched controls
 For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
@@
 For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
-British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
 
-We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
+The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits) and subparagraph (1 digit), and chemical substance (2 characters).
+For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other"
+
+We derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
 These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
 The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.
-In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.
 
@@
 # Results
+As residence in a care home was the strongest risk factor for severe COVID-19, we first examined the association with enchashment of any prescription after stratifying by care home status.
 
 We partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.
-Table \ref{tab:drugs} shows these associations.  In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
+Table \ref{tab:drugs} shows these associations.
 
-** redo this table to show how much the association with scrip.any is reduced
+In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
+Table XX examines how much the effect of any prescription is reduced by adding an indicator variable for each chapter, one at a time.
+The chapter drug classes with the largest contributions to the any prescription effect were Chapter 4 (Central Nervous System), Chapter 1 (Gastrointestinal System) and Chapter 9 (Nutrition), The codes grouped under Chapter 14+ (Other) are mostly for dressings and appliances,  presumably reflecting a general association with frailty.
+This led us to examine associations with specific drug classes, grouped at BNF paragraph or subparagraph level, within each of these three chapters.
 
-The BNF chapter with the strongest association with severe disease was Chapter 1 (gastrointestinal).
-Other chapters associated with severe disease were Chapter 2 (cardiovascular) and Chapter 4 (central nervous system).
+## Drugs acting on the central nervous system: BNF chapter 4
+Table \ref{tab:bnfchapter4} shows that encashed prescriptions of most classes of drugs acting on the central nervous system were associated with severe COVID-19.
+The most commonly prescribed drug class, and the strongest independent effect in a multivariate analysis, was non-opiate analgesics and compound preparations.
+This class of drugs includes paracetamol and various compound preparations that are marketed for relief of upper respiratory symptoms, headache, influenza, and other conditions.
 
 ## Drugs for gastro-intestinal system: BNF chapter 1
-Table \ref{bnfchapter1} shows that encashed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19.
+Table \ref{tab:bnfchapter1} shows that encashed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19.
 Of these by far the most commonly prescribed drug classes were proton pump inhibitors and osmotic laxatives.
 
-Table \ref{protonpump} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
+Table \ref{tab:protonpump} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
+
+## Drugs for Nutrition and Blood: BNF chapter 9
 
 ## Drugs for cardiovascular system: BNF chapter 2
-Table \ref{bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19.
+Table \ref{tab:bnfchapter2} lists associations with drugs for the cardiovascular system, grouped by BNF subparagraph.
+This shows that prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19.
 In a multivariate analysis the strongest associations were with loop diuretics and calcium-channel blockers.
@@
 # Discussion 
+## Association with prescriptions for symptoms
 
-## Possible drug effects
-.
+Association with non-opiate analgesics, laxatives, oral iron
+
+Patients who consult with minor symptoms, leading to prescriptions for these symptoms without a clear medical indication, are at high risk.  This suggests that one or more of the drugs
+
+## Proton pump inhibitors
+The association with proton pump inhibitors is not explained by association any underlying gastrointestinal disease that leads to hospital admission, or by coprescribing with non-steroidal antiinflammatory or antiplatelet agents.
+As proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.
+However as there are no prescription charges in Scotland, long-term users of these expensive drugs are likely to seek to obtain them on prescription.
+
+As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
+Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016].
+35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013].
+
+## Cardiovascular drugs
+No evidence of any effect of ACE inhibitors or A-II receptor blockers beyond that associated with treated hypertension.
 
@@
 
+We have prescription encashments, not just prescriptions.  Drug coding by BNF codes is more systematic than SNOMED
+
+We don't have all the morbidity data available in the primary care record but we have hospital diagnoses which are ICD-coded.
+
 ## Conclusion
+Stay away from doctors if you can, don't take pills without a recognized medical indication
 
@@
 <div id="refs"></div>
+
+```{r carehome, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.scrip.any.carehome,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="carehome",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(clean.header(colnames(table.scrip.any.carehome)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with encashment of any prescription in the last year, by care home status") %>%
+    column_spec(column=1, width="4cm")
+```
 
@@
 
+```{r explained.chapter, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.anyscrip.chapter) <- gsub("\\_", " ", rownames(table.anyscrip.chapter))
+knitr::kable(table.anyscrip.chapter,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="explained.chapter",
+             digits=c(3, 3, 2), 
+             align=rep("r", 3), 
+             caption="Effect of an indicator variable for each BNF chapter on the size of effect associated with encashment of any prescription") %>%
+    column_spec(column=1, width="4.5cm") 
+```
+
 ```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 1, in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those without listed conditions") %>%
     column_spec(column=1, width="4cm") %>%
@@
 
-```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
+```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.bnfchapter2,
+knitr::kable(table.bnfchapter4,
              escape=FALSE, 
              booktabs=TRUE,
-             label="bnfchapter2",
-             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
+             label="bnfchapter4",
+             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 2, in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 4 in those without listed conditions") %>%
     column_spec(column=1, width="4cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter2)) %>%
+    group_rows("", 1, nrow(table.bnfchapter10)) %>%
     landscape()
@@
 
-```{r bnfchapter10, echo=FALSE, warning=FALSE, message=FALSE}
+```{r bnfchapter9, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.bnfchapter10,
+knitr::kable(table.bnfchapter9,
              escape=FALSE, 
              booktabs=TRUE,
-             label="bnfchapter10",
-             col.names=c(clean.header(colnames(table.bnfchapter10)[1:2]),
+             label="bnfchapter9",
+             col.names=c(clean.header(colnames(table.bnfchapter9)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 10 in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 9 in those without listed conditions") %>%
     column_spec(column=1, width="4cm") %>%
@@
 ## Supplementary tables
+
+```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.bnfchapter2,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="bnfchapter2",
+             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those without listed conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    group_rows("", 1, nrow(table.bnfchapter2)) %>%
+    landscape()
+
+```

### 2020-05-14 casecontrol.R

@@ -8,12 +8,13 @@ library(rmarkdown)
 library(pander)
 library(ggplot2)
 library(doParallel)
-registerDoParallel(cores=2)
 library(reshape2)
 library(readxl)
 library(DescTools)
 library(icd.data)
 library(gam)
+
+registerDoParallel(cores=2)
 
 source("helperfunctions.R")
         
@@ -146,18 +147,51 @@ cc.all$diabetes.any <- as.factor(car::recode(cc.all$dm.type,
 cc.all <- within(cc.all, diabetes.any <- relevel(diabetes.any, ref="Not diabetic"))
 cc.all$dm.type <- 
   as.factor(car::recode(cc.all$dm.type, 
-                        "0='Not diabetic'; 1='Type 1'; 2='Type 2';
-                         3:100='Other diabetes type'; 101:102='Type 1'; 202:203='Type 2'"))
+                        "c(0, 10)='Not diabetic';
+                         c(1, 101, 102)='Type 1';
+                         c(2, 202, 203)='Type 2'; 
+                         3:9='Other diabetes type';
+                         11:100='Other diabetes type'"))
+
 cc.all <- within(cc.all, dm.type <- relevel(dm.type, ref="Not diabetic"))
 
 ########################################################################
 
 ## tabulate ethnicity by case group
 testpositives.ethnic <- paste.colpercent(with(cc.all[cc.all$CASE==1, ],
-                                              table(ethnic5, casegroup)), 1)
+                                              table(ethnic4, casegroup)), 1)
 testpositives.ethnic.smr <- paste.colpercent(with(cc.all[cc.all$CASE==1, ],
                                                   table(ethnic5.smr, casegroup)), 1)
 
+testpositives.carehome <- paste.colpercent(with(cc.all[cc.all$CASE==1, ],
+                                                table(ethnic4, care.home)), 0)
+
+testpositives.healthboard <- t(paste.colpercent(with(cc.all[cc.all$CASE==1, ],
+                                                table(ethnic4, HBRES_NAME)), 0))
+
+table.severe.demog <-
+    tabulate.freqs.regressions(varnames=c("ethnic4", "care.home", "SIMD.quintile"),
+                               data=cc.severe)
+
+
+table.testpositives.demog <-
+    tabulate.freqs.regressions(varnames=c("ethnic4", "care.home", "SIMD.quintile"),
[…]

### 2020-05-15 Covid_ethnicity_Scotland.Rmd
fixed mislabelling in ethnicity report

@@
 # Methods 
-All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems. This system records CHI but not ethnicity.
+The REACT-SCOT case-control study was set up to investigate risk factors for severe COVID-19 and to lay the basis for developing risk stratification algorithms.
+All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems.
+This system records CHI but not ethnicity.
 
@@
 ONOMAP assigns both predicted geographical region of origin and ethnicity.
-ONOMAP has previously been reported to have variables sensitivity for assigning non white ethnic groups 17-89%.
-([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were allocated to ethnic groups.
+ONOMAP has previously been reported to have variable sensitivity for assigning non-White ethnic groups.
+([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were used to assign ethnicity.
 Names classified "Muslim" in the ONOMAP classification were assigned to the South Asian group, as Scotland most people with Muslim names are South Asian.
@@
 The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.
+The Black category combines those who self-identify as Black British, Black Caribbean, Black African and Black Other.
+
+The rate ratios shown are based on conditional logistic regression.
+With the incidence density sampling used in this study, the odds ratios in a conditional logistic regression are equal to rate ratios.
+The unconditional odds ratios calculated from the numbers of cases and controls in the tables are not the same as the conditional odds ratios.
 
@@
 ## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records
-
-```{r ethnic, echo=FALSE, warning=FALSE, message=FALSE}
-
-pander(table.ethnic, missing=".",
-       split.tables=Inf,
-       split.cells=rep(16, 6), 
-       caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)")
-
-```
 
@@
 ## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population
+Tables \ref{tab:testpos.ethnicsmr}.
+\ref{tab:ethnicsmr.hosp} and \ref{tab:ethnicsmr} compare cases and controls with ethnicity based on SMR classificaiton for all test positive cases, all hospitalized cases, and severe cases.
 
-```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}
+Table \ref{tab:testpos.onomap}, \ref{tab:onomap.hosp}, \ref{tab:onomap} show the same comparisons with ethnicity based on ONOMAP name classification
 
-rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
-knitr::kable(table.severe.demog,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="onomap",
-             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
-    column_spec(column=1, width="4.5cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-```
+There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites.
+The rate of severe disease in South Asians was about half the rate in Whites.
 
-```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
+Tables \ref{tab:severity.smr} and \ref{tab:severity} show the distribution over severity categories by ethnicity
 
-rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
-knitr::kable(table.ethnicsmr.aug,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="ethnicsmr",
-             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
-    column_spec(column=1, width="4.5cm")
-```
-
-As shown above the numbers of persons testing positive from non-White ethnic groups is low.
-There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites.
-The frequency of severe disease was lower in South Asans but with these small numbers this estimate is unstable.
-
-## Relation of hospitalised COVID-19 to ethnicity in the Scottish population
-
-```{r onomap.hosp, echo=FALSE, warning=FALSE, message=FALSE}
-
-rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
-knitr::kable(table.demog.aug,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="onomap.hosp",
-             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
-    column_spec(column=1, width="4.5cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
-```
-
-```{r ethnicsmr.hosp, echo=FALSE, warning=FALSE, message=FALSE}
-
-rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
-knitr::kable(table.hospitalized.demog.ethnicsmr,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="ethnicsmr.hosp",
-             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
-    column_spec(column=1, width="4.5cm")
-```
+Tables \ref{tab:carehome} and \ref{tab:board} show the distribution of care home status and health board.
 
@@
 
-# Supplementary tables
-These tables including test-positive individuals who were not hospitalized are included by request.
+# Tables
+Tables comparing all test-positive individuals with controls are included by request.
 They should be interpreted with caution, as selection for testing is influenced by many factors including health care worker status.
+Similar considerations apply to comparisons of disease severity among those who test positive.
 
-## Relation of severity of COVID-19 to ethnicity among test-positive cases
+```{r ethnic, echo=FALSE, warning=FALSE, message=FALSE}
 
-```{r testpositives.ethnic, echo=FALSE, warning=FALSE, message=FALSE}
-
-pander(testpositives.ethnic, missing=".",
+pander(table.ethnic, missing=".",
        split.tables=Inf,
-       split.cells=rep(25, 3), 
-       caption="Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+       split.cells=rep(16, 6), 
+       caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)")
 
 ```
+
+```{r ethnicsmr.testpos, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
+knitr::kable(table.testpositives.demog.ethnicsmr,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="testpos.ethnicsmr",
+             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm")  %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+```
+
+```{r ethnicsmr.hosp, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
+knitr::kable(table.hospitalized.demog.ethnicsmr,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmr.hosp",
+             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+
+```
+
+```{r ethnic5smr, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.ethnic5smr) <- gsub("_", " ", rownames(table.ethnic5smr))
+knitr::kable(table.ethnic5smr,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmr",
+             col.names=c(clean.header(colnames(table.ethnic5smr)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
 
@@
        split.tables=Inf,
-       caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")
+       caption="\\label{tab:severity.smr}Severity of disease in test-positive cases, by ethnicity assigned in SMR records")
 
@@
 
-```{r testpositives.carehome, echo=FALSE, warning=FALSE, message=FALSE}
-
-pander(testpositives.carehome, missing=".",
-       split.tables=Inf,
-       caption="Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")
-
-```
-
-```{r testpositives.healthboard, echo=FALSE, warning=FALSE, message=FALSE}
-
-pander(testpositives.healthboard, missing=".",
-       split.tables=Inf,
-       caption="Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")
-
-```
-
-## Relation of test-positive status to ethnicity in the Scottish population
+\newpage
 
@@
              caption="Associations of test-positive status with ethnicity based on name classification") %>%
-    column_spec(column=1, width="4.5cm") %>%
+    column_spec(column=1, width="2cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    column_spec(column=2, width="1.95cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+
 ```
 
-```{r ethnicsmr.testpos, echo=FALSE, warning=FALSE, message=FALSE}
+```{r onomap.hosp, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
-knitr::kable(table.ethnicsmr.aug,
+rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
+knitr::kable(table.demog.aug,
              escape=FALSE, 
              booktabs=TRUE,
-             label="testpos.ethnicsmr",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
+             label="onomap.hosp",
+             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
                          "Rate ratio (95\\% CI)", "p-value"),
              align=rep("c", 6), 
-             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
-    column_spec(column=1, width="4.5cm")
+             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    column_spec(column=2, width="1.95cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+
 ```
+
+```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
+knitr::kable(table.severe.demog,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="onomap",
+             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6),
+             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+
+```
+
+\newpage
+
+```{r testpositives.ethnic, echo=FALSE, warning=FALSE, message=FALSE}
+
+pander(testpositives.ethnic, missing=".",
+       split.tables=Inf,
+       split.cells=rep(25, 3), 
+       caption="\\label{tab:severity}Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+
+```
+
+```{r testpositives.carehome, echo=FALSE, warning=FALSE, message=FALSE}
+
+pander(testpositives.carehome, missing=".",
+       split.tables=Inf,
+       caption="\\label{tab:carehome}Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+
+```
+
+\newpage
+
+```{r testpositives.healthboard, echo=FALSE, warning=FALSE, message=FALSE}
+
+pander(testpositives.healthboard, missing=".",
+       split.tables=Inf,
+       caption="\\label{tab:board}Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+
+```

### 2020-05-15 casecontrol.Rmd
fixed mislabelling in ethnicity report

@@
 For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.
-In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(10*exp(logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
+In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
 The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.

### 2020-05-15 pharmaco.Rmd
fixed mislabelling in ethnicity report

@@
 
-_Background_-- We previously  reported that across all ages having any hospital prescription  in the past year was strongly and highly significantly associated with severe COVID-19 independently of  high risk conditions.
-Whilst this may reflect associated condition effects it strength was sufficient to suggest direct drug effects on disease.
-Here we sought to further understand this phenomenon by  systematically testing associations with prescribing history with adjustment for relevant confounding by condition.
+_Background_-- We previously reported from a case-control study that encashment of any prescription from primary care in the past year was strongly associated with severe COVID-19, independently of the presence of conditions designated as high risk.
+The objective of this study was to distinguish between three possible explanations for this unexpected association: confounding by other high-risk conditions, an effect associated with consultations for symptoms not necessarily associated with high-risk conditions, and a direct effect of exposure to one or more classes of drug on risk of severe disease.
 
-_Methods_ -- We linked the records of those tested positive for SARS-C0V-2 viral RNA in the national database that captures  virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data.
-Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test.
-We selected 10 sex, age and general practice based controls per test-positive case.
-For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year.
-Filtering on which BNF chapters showed  association we  further tested for associations at   BNF section and subsection level.
-Conditional logistic regression was used to estimate odds ratios and to adjust for potential confounders.
+_Methods_ -- Severe cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database who entered intensive care or died within 28 days.
+We selected 10 controls per case from the national population register, matched for sex, age and general practice.
+For all cases and controls we extracted International Classification of Disease (ICD-10) codes from the previous five years of hospital discharge records and British National Formulary (BNF) drug codes of encashed primary care issued prescriptions in the past year.
+Rate ratios were estimated by conditional logistic regression.
 
-_Findings_ -- Among those without conditions listed as risk  conditions by the National Health Service  (Diabetes, heart disease,  asthma or COAD, chronic liver disease, chronic renal failure , neurological including dementia, immune disorders or suppression)  severe cases were  highly significantly more likely than controls  to have had prescriptions in the past year from each BNF chapter.
-The strongest  independent associations were seen with Gastro-intestinal,  INSERT OR cardiovascular INSERT OR , Central nervous  INSERT OR  and Nutrition and Blood systems INSERT OR .
-When each of these were further disaggregated to the next level, key findings included a significantly greater risk of severe COVID-19 associated with proton pump inhibitor use independently of co-prescription of non-steroidal anti-inflammatory or anti-platelet agents.
-ACE-inhibitors INSERT and ARBs INSERT  showed  weaker associations  with severe disease than loop diuretics  INSERT or calcium channel blockers INSERT  and  all these associations  INSERT were / were not significant on adjust for any circulatory disorder beyond heart disease.
-Antipsychotic drugs showed increased risk INSERT OR that was/ was not independent of psychotic  conditions.
-Oral iron supplements showed associations INSERT  but these were/ were not independent of prior neoplasm or renal failure INSERT.
-The OR associated with any prior prescription was reduced from xx-xxx on adjusting for these specific sub-chapters ???
+_Findings_ -- The rate ratio associated with encashment of at least one prescription in the last year was elevated `r table.scrip.any.carehome[1, 3]` in those not resident in a care home but not in care home residents `r table.scrip.any.carehome[2, 3]`, of whom `r table.scrip.any.carehome[2, 1]` of controls had encashed a prescription in the last year.
+On restriction to those not resident in a care home, the rate ratio associated with encashment of a prescription was elevated in those without any of the designated high-risk conditions `r table.scrip.any.nocare.listed[1, 3]`, but not in those with any designated condition `r table.scrip.any.nocare.listed[2, 3]` of whom `r table.scrip.any.nocare.listed[2, 1]` of controls had encashed a prescription in the last year.
 
-*Conclusions* -- These prescribing  data emphasise the multimorbidity nature of severe COVID-19 disease beyond the widely accepted risk conditions.
-The strong independent association with PPIs , clearly observable even in those < 40 years, combined with  prior data that this drug class is associated with gastro-enterological infections and pneumonia and the frequency of prescription of these drugs requires urgent evaluation in other datasets.
+Among those without any of the designated high-risk conditions (diabetes, heart disease, obstructive airway disease, chronic liver disease, chronic renal failure, neurological including dementia, immune deficiency / suppression) severe cases were highly significantly more likely than controls to have had prescriptions in the past year from each BNF chapter.
+The strongest associations (rate ratio, 95% CI) were seen with chapters for drugs acting on the gastro-intestinal system `r table.drugs.nocare.notlisted[1, 3]`, cardiovascular system `r table.drugs.nocare.notlisted[2, 3]`, central nervous system `r table.drugs.nocare.notlisted[4, 3]` and "Nutrition and Blood" `r table.drugs.nocare.notlisted[9, 3]`.
+When each of these were further disaggregated to subchapter or paragraph level, key findings included a significantly greater risk of severe COVID-19 associated with proton pump inhibitor use with rate ratios varying from `r table.protonpump[4, 3]` in those aged 75 years and over to `r table.protonpump[1, 3]` in those aged under 40 years.
+This association was only slightly reduced by adjustment for for hospital diagnoses of conditions affecting the upper gastrointestinal tract and for co-prescribing of non-steroidal anti-inflammatory or anti-platelet agents.
+However many other drug classes associated with prescribing for non-specific symptoms were also associated with severe COVID-19: these included osmotic laxatives non-opioid analgesics, oral iron supplements and antipsychotic drugs.
+
+Among drugs prescribed for cardiovascular disease, only loop diuretics and calcium channel blockers were independently associated with risk of severe COVID-19 in a multivariate analysis.
+Associations with ACE inhibitors and angiotensin-II receptor blockers were similar to those for other antihypertensive agents.
+
+*Conclusions* -- Risk of severe COVID-19 is strongly associated with encashment of prescriptions for drugs that are widely prescribed for minor symptoms.
+It remains to be established whether this effect is accounted for by association with characteristics of patients who consult for these symptoms, or by direct drug effects.
+Of the drug classes that might possibly account for an association with prescribing, by far the strongest candidate is proton pump inhibitors, which are prescribed to about 30% of those aged over 60 years and are associated with a threefold elevation of risk of severe disease in this age range.
+As increased risk of infection is an established adverse effect of these drugs, a causal effect is plausible.
+This may be relevant to high rates of severe disease reported in other populations where this class of drugs is widely overprescribed.
+Until further studies are available, it is advisable to avoid using these drugs without a recognized medical indication.
 
@@
 In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` About one third of those over 75 had not encashed a prescription in the previous year.
-In this paper we investigate possible explanations of these associations explanations of this finding
+The objective of this study was to distinguish between three possible explanations for this unexpected association: confounding by other high-risk conditions, an effect associated with consultations for symptoms not necessarily associated with high-risk conditions, and a direct effect of exposure to one or more classes of drug on risk of severe disease.
 
 # Methods
-
-The methods of the REACT-SCOT study are described in detail in the accompanying paper.
+The design of the REACT-SCOT case-control study is described in detail in the accompanying paper.
 
@@
 
+# Data analysis 
 We derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
@@
 The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.
-
-## Statistical methods
-For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.
+Rate ratios were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
 
@@
 # Results
+## Encashment of any prescription by care home status and presence of any listed condition
 As residence in a care home was the strongest risk factor for severe COVID-19, we first examined the association with enchashment of any prescription after stratifying by care home status.
+Table \ref{tab:carehome} shows that the association was restricted to those not resident in a care home.
 
-We partitioned the "Any prescription" variable into indicator variables for each chapter of the British National Formulary, in which drugs are grouped by broad indication, and restricted the analysis to those without one of the listed conditions.
+As a next step, we examined the association with encashment of any prescription after stratifying by the presence or absence of a listed condition, excluding those resident in a care home.
+Table \ref{tab:listed.any} shows that the association was restricted to those without a listed condition
+
+## Partitioning by BNF chapter
+We next partitioned the "any prescription" variable into indicator variables for each chapter of the British National Formulary.
+The chapter level groups drugs by the system on which they act.
+We restricted this analysis to those without any listed condition and not resident in a care home.
 Table \ref{tab:drugs} shows these associations.
@@
 In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
-Table XX examines how much the effect of any prescription is reduced by adding an indicator variable for each chapter, one at a time.
-The chapter drug classes with the largest contributions to the any prescription effect were Chapter 4 (Central Nervous System), Chapter 1 (Gastrointestinal System) and Chapter 9 (Nutrition), The codes grouped under Chapter 14+ (Other) are mostly for dressings and appliances,  presumably reflecting a general association with frailty.
+
+Table \ref{tab:explained.chapter} examines how much the effect of any prescription is reduced by adding an indicator variable for each chapter, one at a time.
+The chapter indicators that explained the highest fractions of the "any prescription" effect were Chapter 1 (Gastrointestinal System) , Chapter 2 (Cardiovascular System), Chapter 4 (Central Nervous System), and Chapter 9 (Nutrition), The codes grouped under Chapter 14+ (Other) are mostly for dressings and appliances, presumably reflecting a general association with frailty.
 This led us to examine associations with specific drug classes, grouped at BNF paragraph or subparagraph level, within each of these three chapters.
 
-## Drugs acting on the central nervous system: BNF chapter 4
-Table \ref{tab:bnfchapter4} shows that encashed prescriptions of most classes of drugs acting on the central nervous system were associated with severe COVID-19.
-The most commonly prescribed drug class, and the strongest independent effect in a multivariate analysis, was non-opiate analgesics and compound preparations.
-This class of drugs includes paracetamol and various compound preparations that are marketed for relief of upper respiratory symptoms, headache, influenza, and other conditions.
+## Associations with severe COVID-19 at section, paragraph and subparagraph level
 
@@
 Table \ref{tab:bnfchapter1} shows that encashed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19.
-Of these by far the most commonly prescribed drug classes were proton pump inhibitors and osmotic laxatives.
-
-Table \ref{tab:protonpump} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
-
-## Drugs for Nutrition and Blood: BNF chapter 9
+Of these by far the most commonly prescribed drug classes was proton pump inhibitors.
+Osmotic laxatives were also associated with severe disease, but were far less commonly prescribed.
 
@@
 
+## Drugs acting on the central nervous system: BNF chapter 4
+Table \ref{tab:bnfchapter4} shows that encashed prescriptions of most classes of drugs acting on the central nervous system were associated with severe COVID-19.
+The most commonly prescribed drug class, and the strongest independent effect in a multivariate analysis, was non-opiate analgesics and compound preparations.
+This class of drugs includes paracetamol and various compound preparations that are marketed for relief of upper respiratory symptoms, headache, influenza, and other conditions.
+Other drug classes associated with severe COVID-19 in a multivariate analysis were antipsychotic agents and opioid analgesics.
+
+## Drugs for Nutrition and Blood: BNF chapter 9
+Table \ref{tab:bnfchapter9} shows that prescriptions of several classes of nutritional supplements were associated with severe disease: these included oral iron supplements, drugs used in megaloblastic anemias, enteral nutrition products, and Vitamin D.
+
+## Association with proton pump inhibitors
+As drugs for the gastrointestinal system accounted for the largest fraction of the effect of "any prescription", and proton pump inhibitors were the most commonly prescribe drug in this class, we examined the association with proton pump inhibitors more closely.
+Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
+The association was strongest in those aged under 40 years, and was reduced only slightly by adjusting for covariates.
+
 # Discussion 
@@
 
-Association with non-opiate analgesics, laxatives, oral iron
+The list of prescribed drug classes that are associated with elevated risk of severe COVID-19 correspond closely to those identified by clinical pharmacologists and pharmacists as widely overprescribed drug classes: these include proton pump inhibitors, laxatives, analgesics, antipsychotic agents and nutritional supplements.
 
-Patients who consult with minor symptoms, leading to prescriptions for these symptoms without a clear medical indication, are at high risk.  This suggests that one or more of the drugs
+Patients who consult with minor symptoms, leading to prescriptions of these drugs without a clear medical indication, are at high risk.
+The problem is to determine whether this association is accounted for by characteristics that underlie these consultations for minor symptoms, or by direct drug effects.
+
+Because this case-control study is matched on primary care practice, variation of over-prescribing between practices does not contribute to the conditional odds ratios.
 
@@
 
+In Szechuan, China, PPIs were prescribed to 37% of inpatients in a tertiary referral centre in 2016 [@luo_changes_2018].
+A Spanish drug-utilization study showed that 55% of hospital inpatients were presribed a PPI at discharge [@ramirez_overuse_2010].
+
 ## Cardiovascular drugs
-No evidence of any effect of ACE inhibitors or A-II receptor blockers beyond that associated with treated hypertension.
+There is evidence of any effect of ACE inhibitors or angiotensin-II receptor blockers beyond the associated with treated hypertension.
 
@@
 
-We don't have all the morbidity data available in the primary care record but we have hospital diagnoses which are ICD-coded.
+We don't have all the morbidity data available in the primary care record but we have hospital diagnoses which are ICD-coded by trained coders.
 
 ## Conclusion
-Stay away from doctors if you can, don't take pills without a recognized medical indication
+It is not possible to establish unequivocally whether the association with prescription encashment is accounted for by patient characteristics associated with prescribing for minor symptoms, or by a direct effect of any specific class of drugs.
+However of the drug classes that are associated with increased risk that is not explained by underlying high-risk conditions, proton pump inhibitors are the strongest candidate for a causal effect.
+Until further studies are available, it is advisable to avoid using drugs in this class without a recognized medical indication.
 
@@
 
+```{r listed, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.scrip.any.nocare.listed,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="listed.any",
+             col.names=c(clean.header(colnames(table.scrip.any.nocare.listed)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with encashment of any prescription in the last year, by presence of a listed condition, excluding care home residents") %>%
+    column_spec(column=1, width="4cm")
+```
+
 ```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.drugs.aug) <- gsub("_", " ", rownames(table.drugs.aug))
-knitr::kable(table.drugs.aug,
+rownames(table.drugs.nocare.notlisted) <- gsub("_", " ", rownames(table.drugs.nocare.notlisted))
+knitr::kable(table.drugs.nocare.notlisted,
              escape=FALSE, 
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.drugs.aug)[1:2]),
+             col.names=c(clean.header(colnames(table.drugs.nocare.notlisted)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in those not resident in care homes and without listed conditions") %>%
     column_spec(column=1, width="4.5cm") %>%
@@
              label="explained.chapter",
-             digits=c(3, 3, 2), 
-             align=rep("r", 3), 
-             caption="Effect of an indicator variable for each BNF chapter on the size of effect associated with encashment of any prescription") %>%
-    column_spec(column=1, width="4.5cm") 
+             col.names=c("Coefficient", "Rate.ratio", "Proportion explained by covariate"), 
+             digits=c(2, 2, 2), 
+             caption="Effect of an indicator variable for each BNF chapter on the size of effect associated with encashment of any prescription, in those not in care homes and without listed conditions") %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="2cm") %>%
+    add_header_above(c(" "=1, "Adjusted effect"=2))
+
 ```
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those not resident in a care home and without without listed conditions") %>%
     column_spec(column=1, width="4cm") %>%
@@
 
-```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.protonpump,
+knitr::kable(table.bnfchapter2,
              escape=FALSE, 
              booktabs=TRUE,
-             label="protonpump",
-             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
+             label="bnfchapter2",
+             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with proton pump inhibitors by age groups, unadjusted and adjusted for NSAIDs, antiplatelet drugs and hospital diagnosis of a disorder of esophagus, stomach or duodenum") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those without listed conditions") %>%
     column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Unadjusted"=2, "Adjusted"=2)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
     landscape()
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter10)) %>%
+    group_rows("", 1, nrow(table.bnfchapter4)) %>%
     landscape()
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter10)) %>%
+    group_rows("", 1, nrow(table.bnfchapter9)) %>%
+    landscape()
+
+```
+
+```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.protonpump,
+             escape=FALSE,
+             label="protonpump.adjusted", 
+             booktabs=TRUE,
+             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             caption="Associations of severe disease with proton pump inhibitor by age group, with and without adjustment for covariates") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
     landscape()
@@
 ## Supplementary tables
-
-```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter2,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter2",
-             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter2)) %>%
-    landscape()
-
-```

### 2020-05-15 casecontrol.R

@@ -155,43 +155,8 @@ cc.all$dm.type <-
 
 cc.all <- within(cc.all, dm.type <- relevel(dm.type, ref="Not diabetic"))
 
-########################################################################
+###############################################################################
 
-## tabulate ethnicity by case group
-testpositives.ethnic <- paste.colpercent(with(cc.all[cc.all$CASE==1, ],
-                                              table(ethnic4, casegroup)), 1)
-testpositives.ethnic.smr <- paste.colpercent(with(cc.all[cc.all$CASE==1, ],
-                                                  table(ethnic5.smr, casegroup)), 1)
-
-testpositives.carehome <- paste.colpercent(with(cc.all[cc.all$CASE==1, ],
-                                                table(ethnic4, care.home)), 0)
-
-testpositives.healthboard <- t(paste.colpercent(with(cc.all[cc.all$CASE==1, ],
-                                                table(ethnic4, HBRES_NAME)), 0))
-
-table.severe.demog <-
-    tabulate.freqs.regressions(varnames=c("ethnic4", "care.home", "SIMD.quintile"),
-                               data=cc.severe)
-
-
-table.testpositives.demog <-
-    tabulate.freqs.regressions(varnames=c("ethnic4", "care.home", "SIMD.quintile"),
-                               data=cc.all)
-table.testpositives.demog.ethnicsmr <-
-    tabulate.freqs.regressions(varnames=c("ethnic5.smr", "care.home", "SIMD.quintile"),
-                               data=cc.all[!is.na(cc.all$ethnic5.smr), ])
-
-table.hospitalized.demog <-
-    tabulate.freqs.regressions(varnames=c("ethnic4", "care.home", "SIMD.quintile"),
-                               data=cc.all[cc.all$casegroup=="Hospitalized, not severe" |
-                                           cc.all$casegroup=="Critical care or fatal", ])
-
-table.hospitalized.demog.ethnicsmr <-
-    tabulate.freqs.regressions(varnames=c("ethnic5.smr", "care.home", "SIMD.quintile"),
-                               data=cc.all[(cc.all$casegroup=="Hospitalized, not severe" |
-                                            cc.all$casegroup=="Critical care or fatal") &
-                                           !is.na(cc.all$ethnic5.smr), ])
- 
[…]

### 2020-05-15 Covid_ethnicity_Scotland.Rmd
updates to ethnicity report

@@
   \usepackage{array}
-  \usepackage{booktabs}
   \usepackage{longtable}
@@
   keep_tex: true
+  latex_engine: lualatex
 
@@
 There have been anecdotal reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.
-A preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95%CI 2.92-5.12); RR 2.11 (95%CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively)
+A preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95% CI 2.92-5.12); RR 2.11 (95% CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively)
 
@@
 The REACT-SCOT case-control study was set up to investigate risk factors for severe COVID-19 and to lay the basis for developing risk stratification algorithms.
-All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems.
+All persons to date  who tested positive  for SARS-CoV-2 RNA were identified using the ECOSS virology laboratory systems.
 This system records CHI but not ethnicity.
@@
 
-## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population
+## Relation of test -positive, hospitalised and severe / fatal COVID-19 to ethnicity in the Scottish population
 Tables \ref{tab:testpos.ethnicsmr}.
-\ref{tab:ethnicsmr.hosp} and \ref{tab:ethnicsmr} compare cases and controls with ethnicity based on SMR classificaiton for all test positive cases, all hospitalized cases, and severe cases.
+\ref{tab:ethnicsmr.hosp} and \ref{tab:ethnicsmr} compare cases and controls with ethnicity based on SMR classification for all test positive cases, all hospitalized cases, and severe cases.
+Severe cases are those who had ICU care or died within 28 days of a positive test.
+The first column of rate ratios are univariate.
+Those in the second column are further adjusted for SIMD and care home status.
 
-Table \ref{tab:testpos.onomap}, \ref{tab:onomap.hosp}, \ref{tab:onomap} show the same comparisons with ethnicity based on ONOMAP name classification
+Table \ref{tab:testpos.onomap}, \ref{tab:onomap.hosp}, \ref{tab:onomap} show the same comparisons with ethnicity based on ONOMAP name classification.
 
-There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites.
-The rate of severe disease in South Asians was about half the rate in Whites.
+SMR based ethnicity was available in 23833 Controls  and 4230 test-positive cases.
+In contrast name based classification was available in 113894 Controls and 11432 test positive cases.
+Thus in the ethnic groups, such as South Asians,  for whom the named based classification has good performance  the named based estimates are more stable and the confidence intervals on the rate ratios are narrower.
+
+Using SMR based ethnicity there is no significant difference in test positivity rate by ethnicity.
+Although rate ratios are above 1 for black ethnicity the  absolute numbers are sparse, just 15 test positives and the confidence intervals are very wide.
+For those of South Asian ethnicity using SMR based classification the rate ratio is  also above 1 with a wide confidence interval.
+Using the named based classification for which there are much more data  for those of South Asian ethnicity the rate ratio is  actually well below 1  and this is statistically significant.
+For those of Chinese  ethnicity the name based classification showed lower test positive rates also.
+
+There is no evidence that the frequency of  hospitalised COVID-19 disease or severe disease warranting ICU or resulting in death within 28 days among those who test positive is any higher in non-White groups than in Whites.
+Using the SMR based classification there are no differences by ethnic group in hospitalised COVID-19  or in severe COVID-19.
+Using the named based classification for which there are more data the  rate ratios for the non-White ethnic groups are lower than Whites and this is statistically significant in those of South Asian ethnicity.
 
@@
 # Conclusion
-Although the numbers are small, it is possible to establish that South Asians in Scotland are at lower risk of severe disease than the White population.
+There is no evidence that those of non-white ethnicity are at higher risk of COVID-19 disease that Whites at this stage in this epidemic in Scotland.
+Rates of hospitalised and severe disease were lower in those of South Asian ethnicity than Whites.
+For other non-White ethnic group the numbers of cases are so small that all estimates are unstable.
 
-For the Black population, reliable classification of ethnicity is restricted to those who have an SMR record, and the numbers of cases and controls in this group are too small for rate ratios to be reliably estimated.
-
+For the Black population, reliable classification of ethnicity is restricted to those who have an SMR record because name based classification performs poorly for this ethnic group.
 ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.
+The numbers of cases and controls in this group are too small for rate ratios to be reliably estimated so whilst we do not observe any elevations in rates the data are too sparse.
 Recording of ethnicity in hospital admission distinguishes these groups but does not provide complete coverage of the population at risk of COVID-19.
 The coding is also incomplete with substantial missingness.
+
 We have not merged ethnicity data from the Scottish Morbidity Record with name based classification as this would introduce differential misclassification of ethnicity by health status.
@@
 
-A possible explanation for reports from England that people from minority ethnic groups were over-represented among cases of severe and fatal disease may be that these groups make up a high proportion of front-line health care workers and care givers in England, who were heavily exposed to infectious cases.
-In Scotland a "healthy migrant effect", such that those who migrate tend to be fitter than those who do not migrate, may explain the low rates of severe disease in minority groups.
+We will continue to regularly report on the ethnicity distribution  as the epidemic progresses.
+In particular in further data extracts we hope to include those NRS reported deaths with COVID-19 emergency code on the certificate but where no positive test has been recorded.
+Further data extracts will in due course also capture the lighthouse testing data that are as yet not included in the ECOSS database.
 
@@
 
-pander(table.ethnic, missing=".",
-       split.tables=Inf,
-       split.cells=rep(16, 6), 
-       caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)")
+knitr::kable(table.ethnic, 
+             escape=FALSE, 
+             booktabs=TRUE,
+             caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)") %>%
+    add_header_above(c(" "=1, "SMR records"= 6)) %>%
+    pack_rows("ONOMAP", 1, 5) %>%
+    kable_styling(latex_options="hold_position")
 
@@
 
-rownames(table.testpositives.demog.ethnicsmr) <- gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
+rownames(table.testpositives.demog.ethnicsmr) <-
+    gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
 knitr::kable(table.testpositives.demog.ethnicsmr,
@@
     kable_styling(latex_options="hold_position") %>% 
-    column_spec(column=1, width="2cm")  %>%
-    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=1, width="2.7cm")  %>%
+    column_spec(column=2, width="2cm")  %>%
     column_spec(column=3, width="1.85cm")  %>%
@@
 
-rownames(table.hospitalized.demog.ethnicsmr) <- gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
+rownames(table.hospitalized.demog.ethnicsmr) <-
+    gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
 knitr::kable(table.hospitalized.demog.ethnicsmr,
@@
     kable_styling(latex_options="hold_position") %>% 
-    column_spec(column=1, width="2cm") %>%
-    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=1, width="2.7cm") %>%
+    column_spec(column=2, width="2cm")  %>%
     column_spec(column=3, width="1.85cm")  %>%
@@
     column_spec(column=7, width="1.5cm")
+```
 
@@
 
-pander(testpositives.ethnic.smr, missing=".",
-       split.tables=Inf,
-       caption="\\label{tab:severity.smr}Severity of disease in test-positive cases, by ethnicity assigned in SMR records")
+knitr::kable(testpositives.ethnic.smr, 
+           escape=FALSE, 
+             booktabs=TRUE,
+       caption="\\label{tab:severity.smr}Severity of disease in test-positive cases, by ethnicity assigned in SMR records") %>%
+    kable_styling(latex_options="hold_position")
 
@@
              label="testpos.onomap",
-             digits=c(0, 0, 0, 200, 0, 200), 
              col.names=c(clean.header(colnames(table.testpositives.demog)[1:2]),
@@
              caption="Associations of test-positive status with ethnicity based on name classification") %>%
-    column_spec(column=1, width="2cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
     kable_styling(latex_options="hold_position") %>% 
-    column_spec(column=1, width="2cm") %>%
+    column_spec(column=1, width="2.7cm") %>%
     column_spec(column=2, width="1.95cm")  %>%
@@
     kable_styling(latex_options="hold_position") %>% 
-    column_spec(column=1, width="2cm") %>%
-    column_spec(column=2, width="1.95cm")  %>%
+    column_spec(column=1, width="2.7cm") %>%
+    column_spec(column=2, width="2cm")  %>%
     column_spec(column=3, width="1.85cm")  %>%
@@
 
-\newpage
-
 ```{r testpositives.ethnic, echo=FALSE, warning=FALSE, message=FALSE}
 
-pander(testpositives.ethnic, missing=".",
-       split.tables=Inf,
-       split.cells=rep(25, 3), 
-       caption="\\label{tab:severity}Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+knitr::kable(testpositives.ethnic, 
+           escape=FALSE, 
+             booktabs=TRUE,
+       caption="\\label{tab:severity}Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
+    kable_styling(latex_options="hold_position")
 
@@
 
-pander(testpositives.carehome, missing=".",
-       split.tables=Inf,
-       caption="\\label{tab:carehome}Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+knitr::kable(testpositives.carehome, 
+           escape=FALSE, 
+             booktabs=TRUE,
+       caption="\\label{tab:carehome}Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
+    kable_styling(latex_options="hold_position")
 
@@
 
-pander(testpositives.healthboard, missing=".",
-       split.tables=Inf,
-       caption="\\label{tab:board}Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification")
+knitr::kable(testpositives.healthboard, 
+             escape=FALSE, 
+             booktabs=TRUE,
+             caption="\\label{tab:board}Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
+    kable_styling(latex_options="hold_position")
 

### 2020-05-15 stepwise.R

@@ -1,3 +1,38 @@
+
+## this doesn't work when wrapped as a function 
+cv.predict <- function(nfold, cv.data, lower.formula, upper.formula) {
+    
+    cv.predicted <- NULL
+    for(i in 1:nfold) {
+        test <- cv.data$test.fold == i
+        test.data  <- cv.data[test, ]
+        train.data <- cv.data[!test, ]
+        cat(length(which(test)), "observations in test fold", i, "\n")
+        
+        x <- t(with(test.data, table(CASE, stratum)))
+        numcontrols.stratum <- x[, 1]
+        numcases.stratum <- x[, 2]
+        a <- table(numcases.stratum, exclude=NULL)
+        b <- table(numcontrols.stratum, exclude=NULL)
+        if(length(a) > 1) stop("not all strata contain single case")
+        
+        start.model <- clogit(formula=upper.formula, data=train.data)
+        stepwise.model <- step(start.model,
+                               scope=list(lower=lower.formula, upper=upper.formula),
+                               direction="both", trace=-1, method="approximate")
+        
+        ## normalize within each stratum
+        unnorm.p <- predict(object=stepwise.model, newdata=test.data,
+                            na.action="na.pass", 
+                            type="risk", reference="sample")
+        norm.predicted <- normalize.predictions(unnorm.p=unnorm.p,
+                                                stratum=test.data$stratum,
+                                                y=test.data$CASE)
+        cv.predicted <- rbind(cv.predicted, norm.predicted)
+    }
+    return(cv.predicted)
+}
+
 ## stepwise regressions for casecontrol.R
 
 lower.varnames <- "care.home"
@@ -161,7 +196,7 @@ if(stepwise) {
     }
 ###################################
 
-    full <- cv.predicted
+    full.predicted <- cv.predicted
     full.densities <- with(full.predicted,
                            Wdensities(y, posterior.p, prior.p,
                                       recalibrate=FALSE))

### 2020-05-17 Covid_ethnicity_Scotland.Rmd
fixed ethnicity table

@@
 rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
-knitr::kable(table.demog.aug,
+knitr::kable(table.hospitalized.demog,
              escape=FALSE, 

### 2020-05-17 casecontrol.Rmd
fixed ethnicity table

@@
 As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
-With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.
-The odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
 
@@
 ## Morbidity and drug prescribing
-For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
-British National Formulary  (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
+For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge.
+British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive.
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
@@
 These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
+With incidence density sampling, odds ratios in conditional logistic regression models are equivalent to rate ratios.
+These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
 
@@
 The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.
-The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier has been explained elsewhere [@mckeigue_quantifying_2019].
-Especially relevant to this study is that the contributions of independent predictors are additive on the scale of $\Lambda$.
-We can thus add the information for discrimination conditional on age and sex (estimated from this matched case-control study) to the information for discrimination estimated from a logistic regression model on age and sex in the general population.
+The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier, and especially to quantify increments in predictive performance, has been explained elsewhere [@mckeigue_quantifying_2019].
+The contributions of independent predictors are additive on the scale of $\Lambda$.
+In this study we have estimated the information for discrimination provided by age and sex (from a logistic regression model in the general population), and the information for discrimination provided by other risk factors conditional on age and sex (from conditional logistic regression models in this case-control study matched for age and sex).
+We can add these to obtain the total information for discrimination that would be obtained with a classifier in the general population.
+As described elsewhere, this can be used to estimate the proportion of those destined to develop severe or fatal disease who would be correctly classified as high risk under a shielding policy [@mckeigue_evaluation_2020].
 
@@
 
-This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier in the population as that would be obtained in the population to give a total of `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
+This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier that would be obtained in the population to give a total of `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
 
 Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set.
+The model-based distributions have been adjusted slightly to make them mathematically consistent as described elsewhere [@mckeigue_quantifying_2019].
 This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
-Figure \ref{fig:roc}, which shows the corresponding receiver operating characteristic (ROC) curve, encodes exactly the same information in a different form.
+Figure \ref{fig:roc}, which shows the corresponding receiver operating characteristic (ROC) curve, encodes exactly the same information as Figure \ref{fig:wevid} in a different form.
 The likelihood ratio $2^W$ at any given sensitivity and specificity is the gradient of the model-based ROC curve.
@@
 
-## Conclusion
+# Conclusion
 In conclusion the epidemiology of severe COVID-19 disease is becoming clearer  and we confirm here the importance of sociodemographic factors and  chronic conditions such as diabetes, asthma, circulatory disease and others.
@@
 
-# References {-}
+# References
 
@@
              label="demog",
-             digits=c(0, 0, 0, 200), 
              col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
              caption="Univariate associations of severe disease with demographic factors") %>%
-    column_spec(column=1, width="2.5cm") %>%
-    column_spec(column=2:4, width="2.5cm") %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="1.8cm") %>%
+    column_spec(column=4, width="2.5cm") %>%
     group_rows("Ethnicity", 1, 3) %>%
@@
 rownames(table.agegr) <- replace.names(rownames(table.agegr))
+table.agegr[1, c(1, 3, 5, 7)] <- " "
 knitr::kable(table.agegr,
@@
              col.names=c(clean.header(colnames(table.agegr.all)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(clean.header(colnames(table.conditions.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
-             caption="Associations of severe disease with hospital diagnoses in last 5 years, in those without listed conditions") %>%
+             caption="Associations of severe disease with hospital diagnoses in last 5 years, in those without any listed condition") %>%
     column_spec(column=1, width="4cm") %>%
@@
              col.names=c(clean.header(colnames(table.drugs.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in those without any listed condition") %>%
     column_spec(column=1, width="4.5cm") %>%
@@
                                "Extended variable set")
+table.densities[, 6] <- round(table.densities[, 6] - table.densities[1, 6], 1)
 
@@
              digits=c(0, 0, 0, 200), 
-             col.names=c(colnames(table.ethnicsmr.aug)[1:2],
-                         "Rate ratio (95\\% CI)", "p-value"),
+             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(colnames(tables.agegr[[1]])[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(colnames(tables.agegr[[2]])[1:2],
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(clean.header(colnames(tables.agegr[[3]])[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              label="stepwise",
-             col.names=c("log rate ratio", "p-value"), 
+             col.names=c("log rate ratio", "$p$-value"), 
              align=rep("c", 2),

### 2020-05-17 pharmaco.Rmd
updates to ethnicity report
fixed ethnicity table

@@
 When each of these were further disaggregated to subchapter or paragraph level, key findings included a significantly greater risk of severe COVID-19 associated with proton pump inhibitor use with rate ratios varying from `r table.protonpump[4, 3]` in those aged 75 years and over to `r table.protonpump[1, 3]` in those aged under 40 years.
-This association was only slightly reduced by adjustment for for hospital diagnoses of conditions affecting the upper gastrointestinal tract and for co-prescribing of non-steroidal anti-inflammatory or anti-platelet agents.
 However many other drug classes associated with prescribing for non-specific symptoms were also associated with severe COVID-19: these included osmotic laxatives non-opioid analgesics, oral iron supplements and antipsychotic drugs.
@@
 Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
-The association was strongest in those aged under 40 years, and was reduced only slightly by adjusting for covariates.
+The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates, especially prescription of nonopiate analgesics.
+
+Table \ref{tab:analgesic.adjusted) shows the univariate and multivariate associations of severe COVID-19 with prescription of nonopioid analgesics .
 
@@
 As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
-Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016].
+Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].
 35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013].
@@
              col.names=c(clean.header(colnames(table.scrip.any.carehome)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
              caption="Associations of severe disease with encashment of any prescription in the last year, by care home status") %>%
+     kable_styling(latex_options="hold_position") %>%
    column_spec(column=1, width="4cm")
@@
              col.names=c(clean.header(colnames(table.scrip.any.nocare.listed)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
              caption="Associations of severe disease with encashment of any prescription in the last year, by presence of a listed condition, excluding care home residents") %>%
+    kable_styling(latex_options="hold_position") %>%
     column_spec(column=1, width="4cm")
@@
              col.names=c(clean.header(colnames(table.drugs.nocare.notlisted)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
     landscape()
@@
     column_spec(column=4, width="2cm") %>%
+    kable_styling(latex_options="hold_position") %>%
     add_header_above(c(" "=1, "Adjusted effect"=2))
@@
              col.names=c(clean.header(colnames(table.bnfchapter1)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
              col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
     group_rows("", 1, nrow(table.bnfchapter4)) %>%
+    kable_styling(latex_options="hold_position") %>%
     landscape()
@@
              col.names=c(clean.header(colnames(table.bnfchapter9)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
@@
     group_rows("", 1, nrow(table.bnfchapter9)) %>%
+    kable_styling(latex_options="hold_position") %>%
     landscape()
@@
              col.names=c(clean.header(colnames(table.protonpump)[1:2]),
-                         "Rate ratio (95\\% CI)", "p-value",
-                         "Rate ratio (95\\% CI)", "p-value"),
-             caption="Associations of severe disease with proton pump inhibitor by age group, with and without adjustment for covariates") %>%
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe disease with prescription of a proton pump inhibitor by age group, with and without adjustment for hospital diagnosis of any upper GI tract disorder, and prescription of NSAIDs,  antiplatelet agent, and nonopioid analgesics, excluding care home residents") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
     landscape()
@@
 
-\newpage
+```{r analgesic, echo=FALSE, warning=FALSE, message=FALSE}
 
-# Appendix
-## Supplementary tables
+knitr::kable(table.nonopioid.analgesic,
+             escape=FALSE,
+             label="analgesic.adjusted", 
+             booktabs=TRUE,
+             col.names=c(clean.header(colnames(table.nonopioid.analgesic)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe disease with prescription of a nonopioid analgesic by age group, with and without adjustment for hospital diagnosis of any neoplasm, and prescription of  NSAIDs, antiplatelet agents and proton pump inhibitors, excluding care home residents") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+
+```

### 2020-05-17 casecontrol.R

@@ -84,10 +84,12 @@ cc.all <- within(cc.all, ethnic4.smr <- relevel(as.factor(ethnic4.smr), ref="Whi
 ## recode ONOMAP ethnicity to 4 categories: White, South Asian, Chinese, Other
 cc.all$ethnic4 <- car::recode(cc.all$ethnic5, "'Black'='Other'")
 cc.all <- within(cc.all, ethnic4 <- relevel(as.factor(ethnic4), ref="White"))
+cc.all$ethnic4 <- factor(cc.all$ethnic4, levels=levels(cc.all$ethnic4)[c(1, 4, 2, 3)])
 
 ## recode ONOMAP ethnicity to 3 categories: White, South Asian, Other
 cc.all$ethnic3 <- car::recode(cc.all$ethnic4, "'Chinese'='Other'")
 cc.all <- within(cc.all, ethnic3 <- relevel(as.factor(ethnic3), ref="White"))
+cc.all$ethnic3 <- factor(cc.all$ethnic3, levels=levels(cc.all$ethnic3)[c(1, 3, 2)])
 
 ####################################################################
 
@@ -154,6 +156,7 @@ cc.all$dm.type <-
                          11:100='Other diabetes type'"))
 
 cc.all <- within(cc.all, dm.type <- relevel(dm.type, ref="Not diabetic"))
+cc.all$dm.type <- factor(cc.all$dm.type, levels=levels(cc.all$dm.type)[c(1, 3, 4, 2)])
 
 ###############################################################################
 
@@ -737,39 +740,17 @@ table.anyscrip.chapter$prop.explained <- round(with(table.anyscrip.chapter,
 
 ## tabulate para or subpara codes in BNF chapters of interest
 
-###### start with chapter 1
 table.bnfchapter1 <- tabulate.bnfparas(chnum=1, data=cc.nocare.notlisted)
+table.bnfchapter2 <- tabulate.bnfsubparas(chnum=2, data=cc.nocare.notlisted)
+table.bnfchapter4 <- tabulate.bnfsubparas(chnum=4, data=cc.nocare.notlisted)
+table.bnfchapter9 <- tabulate.bnfsubparas(chnum=9, data=cc.nocare.notlisted)
+table.bnfchapter10 <- tabulate.bnfsubparas(chnum=10, data=cc.nocare.notlisted)
+
+## fix to tabulate BNF chemical substance
 
 ############# proton pump #########################
 
-## tabulate associations with and without covariate adjustment, excluding care home
-
-table.protonpump <- NULL
[…]

### 2020-05-17 helperfunctions.R

@@ -140,7 +140,7 @@ tabulate.bnfsection <- function(chnum, outcome="CASE", data=cc.severe, minrowsum
     }
 }
 
-tabulate.bnfparas <- function(chnum, outcome="CASE", data=cc.severe) {
+tabulate.bnfparas <- function(chnum, outcome="CASE", data=cc.severe, minrowsum=20) {
     ## get sectioncode in wide format, one col per subchapter
     scrips.ch.wide <- reshape2::dcast(scrips[scrips$chapternum==chnum, ],
                                          ANON_ID ~ paracode, fun.aggregate=length,
@@ -163,7 +163,7 @@ tabulate.bnfparas <- function(chnum, outcome="CASE", data=cc.severe) {
     
     bnf.ch <- colnames(cc.bnf.ch)[-(1:3)]
     table.bnf.ch <- univariate.tabulate(varnames=bnf.ch, outcome="CASE", data=cc.bnf.ch,
-                                        drop.sparserows=TRUE)
+                                        drop.sparserows=TRUE, minrowsum=minrowsum)
     ## regressions fixed to use only rows retained by univariate.tabulate
     if(nrow(table.bnf.ch) >= 1) { 
         bnf.ch <- rownames(table.bnf.ch)  
@@ -188,7 +188,7 @@ tabulate.bnfparas <- function(chnum, outcome="CASE", data=cc.severe) {
     }
 }
 
-tabulate.bnfsubparas <- function(chnum, outcome="CASE", data=cc.severe) {
+tabulate.bnfsubparas <- function(chnum, outcome="CASE", data=cc.severe, minrowsum=20) {
     ## get subparacode in wide format, one col per subchapter
     ## subpara code is misnamed bnf_paragraph_code
     scrips.ch.wide <- reshape2::dcast(scrips[scrips$chapternum==chnum, ],
@@ -212,7 +212,7 @@ tabulate.bnfsubparas <- function(chnum, outcome="CASE", data=cc.severe) {
     
     bnf.ch <- colnames(cc.bnf.ch)[-(1:3)]
     table.bnf.ch <- univariate.tabulate(varnames=bnf.ch, outcome="CASE", data=cc.bnf.ch,
-                                        drop.sparserows=TRUE)
+                                        drop.sparserows=TRUE, minrowsum=minrowsum)
     ## regressions fixed to use only rows retained by univariate.tabulate
     if(nrow(table.bnf.ch) >= 1) { 
         bnf.ch <- rownames(table.bnf.ch)

### 2020-05-20 Covid_ethnicity_Scotland.Rmd
updated for pharmacoepi, incidencemortality.R refactored as separate file

@@
     kable_styling(latex_options="hold_position") %>% 
-    column_spec(column=1, width="2cm") %>%
+    column_spec(column=1, width="2.7cm") %>%
     column_spec(column=2, width="1.85cm")  %>%
@@
     column_spec(column=5, width="1.5cm") %>%
-    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=6, width="2.8cm") %>%
     column_spec(column=7, width="1.5cm")

### 2020-05-20 casecontrol.Rmd
updated for pharmacoepi, incidencemortality.R refactored as separate file

@@
 _Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.
-Here we  used sociodemographic data and extensive previous hospitalisation and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand factors associated with severity.
+The objectives of this study were to identify risk factors for severe disease and to lay the basis for risk stratification based on demographic data and health records.
 
-_Methods_ -- We linked the records of those with a positive nucleic acid test  for SARS-CoV-2 in the national database that captures virology testing in all NHS laboratories, to  hospital admissions, intensive care and mortality data.
-Severe COVID-19 was defined as disease requiring intensive care support or associated with death within 28 days of a positive test.
-We selected 10 sex, age and general practice based controls per test-positive case.
-For all cases and controls we extracted the International Classification of Disease (ICD)-10 codes from the previous five years of hospital discharge records  and the British National Formulary (BNF)  drug codes of  encashed  primary care issued prescriptions in the past year.
-We evaluated the risk of severe COVID-19 associated with  a pre-defined  set of  conditions using conditional logistic regression.
-We examined how much predictive performance  for severe COVID-19  a risk classifier built  and tested with four-fold cross validation based on these conditions could achieve and whether this could be improved using information from the rest of the hospitalisation and prescribing data.
+_Methods_ -- The design was a matched case-control study.
+Test-positive cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database and from national mortality data.
+Those who had been admitted to intensive care or had died within 28 days of a positive test were classified as severe cases.
+For each case, ten controls matched for sex, age and primary care practice were selected from the population register.
+For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from encashed prescriptions issued in primary care care during the past year and linked to the national diabetes register.
+Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.
+The performance of a predictive model for severe COVID-19 was estimated by cross-validation.
 
 _Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
-In the case-control analysis, the odds ratios were
-`r table.agegr.all[1, 3]` for residence in a care home status, 
+In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]` .
+Rate ratios (95% CIs) for conditions previously designated as conferring high risk were
 `r table.agegr.all[4, 3]` for diabetes, 
@@
 `r table.agegr.all[10, 3]` for chronic liver disease and 
-`r table.agegr.all[11, 3]` for immune deficiency or suppresssion.
+`r table.agegr.all[11, 3]` for immune deficiency or suppression.
 
-Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions: `r gsub("\\\\", "", table.agegr[13, 2])` versus `r gsub("\\\\", "", table.agegr[13, 1])` in those aged under 40 years.
-The odds ratio for severe disease associated with encashment of at least one prescription drug in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
-These associations with severe disease were present in those without any of the listed conditions, and were distributed over most BNF chapters and the associations with hospital diagnoses over most ICD-10 chapters.
+Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions.
+In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[13, 2])` in cases and `r gsub("\\\\", "", table.agegr[13, 1])` in controls.
+The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
+These associations remained after excluding those with any of the listed conditions.
+The associations with hospital diagnoses were distributed over most chapters of the ICD-10 and the associations with prescribed drugs were distributed over most chapters of the British National Formulary.
 
 With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
-The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
+The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included the listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
 
-_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history and most younger persons developing severe disease have an underlying medical condition.
-The comorbidities associated with increased risk are not limited to the current lists of risk conditions highlighted by public health agencies.
+_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history; most younger persons developing severe disease have an underlying medical condition.
+The comorbidities associated with increased risk are not limited to the lists of risk conditions designated by public health agencies.
 A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.
@@
 
-The quantification of the risk associated with  characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator  [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].
-Two recent UK analyses have  had population comparators and these have reported  increased relative risks for a range of co-morbidities including diabetes, asthma and heart disease and for non-white ethnicity[@niedzwiedz_ethnic_2020;@the_opensafely_collaborative_opensafely_2020].
-A systematic evaluation of the past medical history in those with and without severe disease has not as yet been reported.
+Quantification of the risk associated with characteristics and co-morbidities has been limited by the lack of comparisons with the background population [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].
+Two recent studies in the UK have included population comparators and have reported associations with co-morbidities including diabetes, asthma and heart disease.
+[@niedzwiedz_ethnic_2020; @the_opensafely_collaborative_opensafely_2020].
+A systematic comparison of medical records in those with and without severe disease has not yet been reported.
 
-We undertook a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the background population to understand sociodemographic and comorbidity factors associated with severe COVID-19 disease.
+We report a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population.
+The objectives were to identify risk factors and to lay the basis for risk stratification based on a predictive model.
 
@@
 The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
-The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.
+The CHI number is a unique identifier used in all care systems in Scotland.
 All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
-Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01.
-Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland.
+Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and from the Scottish Morbidity Record 01 which is updated weekly.
+Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all critical care units in Scotland.
 Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
@@
 Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensured complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.
+By restricting the case definition to cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from critical care.
 
@@
 Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.
-Ethnicity was assigned based on applying a name classification algorithm ONOMAP [@lakha_name_2011] to the names in the CHI database.
-This classification is able to distinguish South Asian and Chinese groups with fair accuracy but does not distinguish those who identify as Black or Black-Caribbean from those who identify as White.
-For those cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census.
+Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.
+This classification is able to distinguish South Asians, the largest non-White minority in Scotland, from those who identify as White with fair accuracy but misclassifies most of those who identify as Black or Black-Caribbean (a much smaller minority in Scotland) as White.
+<!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census.
+--->
 
@@
 
-We first extracted history of a specific list of conditions and drug classes that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
+We began by scoring a specific list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+These listed conditions were assigned based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes.
 The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material.
-In addition to these pre-specified groupings of codes we then undertook a systematic evaluation of the association of all ICD-10 codes and BNF codes to chapter level.
-We did not specifically tabulate the conditions that have been listed as making people " clinically extremely vulnerable" such as cystic fibrosis or leukaemia  or organ transplantation as there is no clinical equipoise regarding risk in such persons and as it is expected that they will be underrepresented in the case data as they have been advised to shield themselves completely since early in the epidemic.
-This group also includes those with cancer in receipt of chemotherapy or radiotherapy.
+In addition to these pre-specified groupings of codes we tabulated associations with all ICD-10 codes and BNF codes to chapter level.
+A
+separate list of conditions has been used to designate "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
+We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.
 
@@
 These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
-With incidence density sampling, odds ratios in conditional logistic regression models are equivalent to rate ratios.
+With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.
 These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
@@
 The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier, and especially to quantify increments in predictive performance, has been explained elsewhere [@mckeigue_quantifying_2019].
-The contributions of independent predictors are additive on the scale of $\Lambda$.
+A key advantage of using $\Lambda$ is that the contributions of independent predictors are additive on the scale of $\Lambda$.
+The class-conditional distributions of weights of evidence in cases and controls encode exactly the same information as the receiver operating characteristic curve, when these distributions are adjusted for mathematical consistency as described elsewhere [@mckeigue_quantifying_2019] to obtain the model-based distributions.
+The likelihood ratio at any given sensitivity and specificity is the gradient of the model-based ROC curve.
+
 In this study we have estimated the information for discrimination provided by age and sex (from a logistic regression model in the general population), and the information for discrimination provided by other risk factors conditional on age and sex (from conditional logistic regression models in this case-control study matched for age and sex).
@@
 The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.
-For severe cases as defined in this study, the sex differential is narrow up to about age 50, implying that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but widens from about age 40 onwards.
-In models including both age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
+For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.
+This implies that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical care admissions to total fatalities is higher in men.
+In models that included  age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
 The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.
@@
 Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.
-Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.
+Residence in a care home was by far the strongest risk factor for severe disease.
+Higher risk of severe disease was also associated with socioeconomic deprivation.
 In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.
 Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.
-Although the numbers are small, there is no evidence that the Black minority is at higher risk of severe disease.
+The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.
 
@@
 
-As shown  in the age band specific univariate logistic regression models in tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75}.
-Here we collapsed the two younger age bands to <60 years.
-All of the listed conditions were significantly associated with severe disease  in each age band.
+Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 age bands combined.
+All listed conditions were associated with severe disease in each age band.
 Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.
-Of note is that "other heart disease" was more  strongly associated with severe COVID-19 than ischaemic heart disease.
+The associations with severe COVID-19 were stronger for the broad category "other heart disease" than for ischaemic heart disease.
 
 Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} also show multivariate rate ratios by age group.
-The key point of this  analysis is to demonstrate that at all ages there was a very strong  and highly significant  association of having had any prescription in the last year  and of having had any admission in the past five years even adjusted for all the listed conditions and care home residence.
+These table show that even after adjusting for care home residence and listed conditions, encashment of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease.
 
 ## Systematic analysis of diagnoses associated with severe disease
-The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was further explored by systematically testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.
+The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.
 These results are shown in Table \ref{tab:conditions}.
 In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-The strongest associations were seen with neoplasms and with eye disease.
+The strongest residual associations were seen with neoplasms.
 Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.
-The key point of this table shows that there are many subchapter diagnoses that are significantly more frequent in severe COVID-19 cases than controls.
-Of particular note is that prior hospitalisations with infections, pneumonia and acute respiratory diseases, circulatory disorders other than heart disease, including cerebrovascular accidents.
-Without filtering on the aforementioned criteria there are of course additional differences in counts of rarer diagnoses in cases and controls are apparent.
+This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.
+We note that past hospital diagnoses of infections, pneumonia and acute respiratory diseases are strongly associated with severe COVID-19.
+Associations with cardiovascular disease are not limited to heart disease but are also present for stroke and other circulatory disorders that are not designated as risk conditions.
 
 ## Associations of prescribed drugs with severe disease 
-Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of any prescription in the last year.
+Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of at least one prescription in the last year.
 The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
@@
 ## Construction of a multivariate risk prediction model 
-To obtain a summary metric of the additional information there is in the hospitalisation and prescribing data predictive models were constructed from two sets of variables: a baseline set consisting only of care home residence plus the listed conditions, and an extended set that augmented the baseline set with an indicator variable for ICD-10 chapter level admissions, and indicator variables for prescriptions in each of the BNF chapters.
+To evaluate the incremental contributions of the listed conditions, and residual information in hospitalisation and prescription records data after assigning these conditions, predictive models were constructed from three sets of variables: a baseline set consisting only of demographic variables, a set that included indicator variables for each listed condition, and an extended set that included demographic + listed conditions + indicator variables for hospital diagnoses in each ICD-10 chapter and prescriptions in each BNF chapter.
 
-For each of variable set, a stepwise regression procedure was carried out using alternating forward-backward selection.
+For each variable set, a stepwise regression procedure was carried out using alternating forward-backward selection.
 The variables retained with each variable set are shown in Table \ref{tab:stepwise} .
@@
 This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
-Figure \ref{fig:roc}, which shows the corresponding receiver operating characteristic (ROC) curve, encodes exactly the same information as Figure \ref{fig:wevid} in a different form.
-The likelihood ratio $2^W$ at any given sensitivity and specificity is the gradient of the model-based ROC curve.
+Figure \ref{fig:roc} shows the corresponding receiver operating characteristic (ROC) curve, which encodes the same information as Figure \ref{fig:wevid} but does not allow the separation of cases and controls by the risk classifier to be easily visualized.
 
@@
 The effect of matching on general practice which will partly match for socioeconomic disadvantage will reduce the precision of the effect of the socioeconomic indicator (SIMD) in our analysis but will not bias it in either direction.
-Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold adjusted for the listed conditions.
-This eight-fold risk likely reflects both the spread of the epidemic in care homes and some residual confounding by frailty.
+Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold by adjustment for listed conditions.
+This eight-fold risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.
 
-In contrast to reports from the US and England [@the_opensafely_collaborative_opensafely_2020] we found no evidence of increased risk in non-White minorities.
-This suggests that ethnic differences reported in other countries are not genetically driven but reflect other social factors that differ between regions in their association with ethnicity.
+In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity, and other non-White minorities, are at lower risk than Whites.
+This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.
+In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
 We note that the proportion of health care workers that are of non-British nationality varies across the UK from [23% in London to 7% in Scotland](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/internationalmigration/articles/internationalmigrationandthehealthcareworkforce/2019-08-15).
-In a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk further[@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-In the OpenSAFELY prospective cohort study using medical records in England compared to those recorded as White, Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively adjusted for comorbidities and socioeconomic status, However as acknowledged by the authors, data on health care worker status and other social factors would be needed to explore this further.
-We are in the process of obtaining more detailed data on health care worker status in Scotland.
 
 ### Co-morbidities
-We have confirmed the increased risks associated with the listed conditions. These were based on the list of " moderate risk " conditions highlighted by the NHS and other agencies [@nhs_whos_2020].
-
-The risk associated with these conditions various by age - for example the odds ratio associated with diabetes is much higher at younger ages.
-The 2.7 fold risk in all ages combined is commensurate with that reported in UK Biobank and in the OpenSAFELY studies.
+We have confirmed that the " moderate risk " conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.
+However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages.
+The 2.7 fold rate ratio in all ages combined is similar to that reported in UK Biobank and in the OpenSAFELY studies.
 We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
 Of note other heart disease is more strongly associated than ischaemic heart disease.
-This category includes many conditions such as atrial fibrillation, cardiomyopathies and heart failure.
-However overall 26% of cases did not have any of these conditions.
-Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as many people under current chemotherapy and radiotherapy are included in the shielding at present.
-We have not focussed on evaluating the risk associated with conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding.
-A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather very many associated diagnoses.
-We note however that past hospitalisation for influenza and respiratory tract infections is an important risk factor.
+This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure.
+However overall 26% of cases did not have any of these listed conditions.
+Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as in the current dataset we cannot separately identify those who are currently receiving chemotherapy or radiotherapy for whom shielding is advised.
+We have not attempted to estimate the risk associated with these conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding rather than on the risk to those exposed to the epidemic.
+A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather many associated diagnoses.
+Past hospital admission for influenza or respiratory tract infection, though not usually designated as a high-risk condition, is an important and easily-scored risk factor for severe COVID-19.
 
-Media reports of apparently healthy young people succumbing to severe COVID-19 can lead to an impression that severe COVID-19 is a random event.
-However we found that almost half of cases under 40 years had at least one of the listed conditions In those that did not have a listed condition having any hospital admission in the past five years and having any prescription in the past year was much more common among cases that in the background population of the same age.
-Thus the vast majority of younger persons succumbing to severe disease are not people with no prior health conditions.
+Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.
+However we found that almost half of cases under 40 years had at least one of the listed conditions.
+Among those who did not have a listed condition, the proportions who had had at least one hospital admission in the past five years and had encashed at least one prescription in the past year where much higher among cases than in controls.
+Thus even among younger persons, most cases of severe disease are attributable to pre-existing health conditions.
 
 ## Possible drug effects
-A striking finding of this study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year.
-This association was only partly explained by differing prescription levels associations with the listed conditions.
+A striking finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
 Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19 .
 Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.
-It is likely that most drug associations will simply reflect co-morbidities for which they are prescribed.
-However we note that there have been proposals of directly adverse effects of drugs such as angiotensin converting enzyme inhibitors [@fang_are_2020] although in a case control study from Italy differences in usage between cases with COVID-19 and controls were attributable to co-morbidities in a case control analysis [@mancia_reninangiotensinaldosterone_2020]
-
-Nonetheless the strong association with encashed of at least one prescription warrants a more detailed investigation.
-In an accompanying paper we explore these associations more fully since they require a detailed consideration of confounding by indication before and directly causal relationship can be asserted.
+Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, causal effects of drug classes cannot be ruled out.
+These association, and possible explanations, are explored in an accompanying paper.
 
@@
 A key implication of our results is that risk of severe or fatal disease is multifactorial and that the ninefold rate ratio associated with a 20-year difference in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological conditions that impair mobility.
-A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will not be an efficient way to allocate resources to those who most need it.
-Assigning all persons with the current list of risk conditions will be too insensitive and too non-specific - a third of 60-75 years old in the background population have at least one of the listed conditions we examined.
-Half of the younger people with severe disease do not have any of them.
-Therefore it is necessary to construct a risk classifier that optimizes sensitivity and specificity based on the information available.
-Our results show that this is possible in principle, though more work is required to construct a practical risk classifier for Scotland, let alone to develop and validate classifiers for use in other populations.
+A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one-third of those aged 60-74 year olds in the population have at least one of the listed conditions we examined.
+It will also exclude many people at high risk because they have multiple risk factors.
+The only way to optimize risk classification so as to ensure equity with respect to risk is to construct a classifier that uses all available information to assign a risk score.
+Our results show that this is possible in principle: in Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.
 
 ## Methodological strengths and weaknesses
-Most of the reports of disease associations with COVID-19 have been case series descriptions there are few prospective cohort or case control analyses with appropriate population comparators.
+Most reports of disease associations with COVID-19 have been case series.
+There have been few reports based on evaluating these associations in the population through cohort or case-control studies.
 With this matched case control design using incidence density sampling, we have been able to estimate rate ratios conditional on age and sex.
 An unpublished analysis from England explored the association of similar set of risk conditions with in-hospital COVID-19 deaths, but did not systematically evaluate the rest of the medical record including prescription records.
-Although we have records of encashment of prescriptions, we do not at present have access to other the primary care data, which would contain additional information such as body mass index.
-Hospital discharge diagnoses however are coded to ICD-10 by trained coders.
-A further limitation is generalisability especially of sociodemographic factors to other locations.
-Also we limited our case control study to test-positive cases.
-This excludes deaths where COVID-19 is mentioned as being of suspected or probable relevance to the death but where no testing was carried out.
-Up to 13 May 2020 an additional 1200 such deaths had been reported by the [National Registers of Scotland.](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland) Future iterations of this  analysis will include sensitivity analyses of the extent to which the results are changed by including these deaths, but the data are not yet available.
-Apart from residential care home status we anticipate that the ORs for most other risk factors will be similar or less as there will be some misclassification of cause in these deaths.
+Although we have records of encashment of prescriptions, we do not at present have access to other primary care data, which would contain additional information on morbidity and measurements such as body mass index.
+A strength of our study however is that hospital discharge diagnoses are coded to ICD-10 by trained coders, in contrast to the coding systems used in primary care databases that do not map to recognized disease classifications.
+Associations with ethnicity and other sociodemographic factors are not necessarily generalizable from Scotland to other populations.
+
+This case-control study is limited to test-positive cases, excluding deaths where no testing was carried out but COVID-19 was mentioned on the death certificate as an underlying or contributing cause.
+Up to 13 May 2020 an additional 1200 such deaths had been reported by the [National Registers of Scotland.](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland) Future versions of this analysis will include sensitivity analyses of the extent to which the results are changed by including these deaths, but the data are not yet available.
+Apart from residential care home status, we do not expect most other risk factors to differ markedly between those who died from COVID-19 without being tested and those who died after testing.
 
 # Conclusion
-In conclusion the epidemiology of severe COVID-19 disease is becoming clearer  and we confirm here the importance of sociodemographic factors and  chronic conditions such as diabetes, asthma, circulatory disease and others.
-However there is not just one small set of conditions that contribute to risk but many conditions as evidenced by associations across with past medical and prescribing history across many body systems.
+This study confirms that risk of severe COVID-19 is associated with sociodemographic factors and with chronic conditions such as diabetes, asthma, circulatory disease and others.
+However the associations with pre-existing disease are not just with a small set of conditions that contribute to risk, but with many conditions as demonstrated by associations with past medical and prescribing history in relation to multiple physiological systems.
 As countries attempt to emerge from lockdown whist protecting the most vulnerable multivariate classifiers rather than crude rule-based approaches will be needed to define those most at risk.
@@
     column_spec(column=1, width="3.5cm") %>%
-    group_rows("Listed condition", 3, nrow(table.agegr) - 1) %>%
+    group_rows("Listed condition", 5, nrow(table.agegr) - 1) %>%
     add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
@@
              caption="Associations of severe disease with listed conditions over all age groups") %>%
+    group_rows("Listed condition", 4, nrow(table.agegr.all)) %>%
     column_spec(column=1, width="4cm") %>%
@@
 options(knitr.kable.NA = '')
-
+if(FALSE) {
 knitr::kable(table.ethnicsmr.aug,
@@
     column_spec(column=1, width="3cm")
-
+}
 ```
@@
              caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
+    group_rows("Listed condition", 4, nrow(tables.agegr[[1]])) %>%
     column_spec(column=1, width="4cm") %>%
@@
              caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
+    group_rows("Listed condition", 4, nrow(tables.agegr[[2]])) %>%
     column_spec(column=1, width="4cm") %>%
@@
              caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
+    group_rows("Listed condition", 4, nrow(tables.agegr[[3]])) %>%
     column_spec(column=1, width="4cm") %>%

### 2020-05-20 pharmaco.Rmd
updated for pharmacoepi, incidencemortality.R refactored as separate file

@@
 
+\newpage
+
 # Abstract
 
-_Background_-- We previously reported from a case-control study that encashment of any prescription from primary care in the past year was strongly associated with severe COVID-19, independently of the presence of conditions designated as high risk.
-The objective of this study was to distinguish between three possible explanations for this unexpected association: confounding by other high-risk conditions, an effect associated with consultations for symptoms not necessarily associated with high-risk conditions, and a direct effect of exposure to one or more classes of drug on risk of severe disease.
+_Background_-- We previously reported that dispensing of at least one prescription from primary care in the past year was strongly associated with severe COVID-19 in Scotland, independently of past diagnoses of conditions designated as high risk.
+The objective of this study was to investigate possible explanations for this association.
+.
 
 _Methods_ -- Severe cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database who entered intensive care or died within 28 days.
-We selected 10 controls per case from the national population register, matched for sex, age and general practice.
-For all cases and controls we extracted International Classification of Disease (ICD-10) codes from the previous five years of hospital discharge records and British National Formulary (BNF) drug codes of encashed primary care issued prescriptions in the past year.
+From the national population register ten controls per case were selected, matched for sex, age and general practice.
+International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from encashed primary care issued prescriptions in the past year.
 Rate ratios were estimated by conditional logistic regression.
 
-_Findings_ -- The rate ratio associated with encashment of at least one prescription in the last year was elevated `r table.scrip.any.carehome[1, 3]` in those not resident in a care home but not in care home residents `r table.scrip.any.carehome[2, 3]`, of whom `r table.scrip.any.carehome[2, 1]` of controls had encashed a prescription in the last year.
-On restriction to those not resident in a care home, the rate ratio associated with encashment of a prescription was elevated in those without any of the designated high-risk conditions `r table.scrip.any.nocare.listed[1, 3]`, but not in those with any designated condition `r table.scrip.any.nocare.listed[2, 3]` of whom `r table.scrip.any.nocare.listed[2, 1]` of controls had encashed a prescription in the last year.
+_Findings_ -- In those not resident in a care home and without past diagnoses of any conditions designated as high risk, the rate ratio (95% CI) associated with dispensing of any prescription in the last year was `r table.scrip.any.nocare.listed[1, 3]`.
+A backward selection procedure showed that two chapters of the BNF -- drugs for the gastrointestinal system (Chapter 1) and the cardiovascular system (Chapter 2) -- accounted for `r round(100 * stepwise.drop[13, 2] / stepwise.drop[1, 2])`% of this effect.
+A further backward selection of  drug codes in these two chapters identified proton pump inhibitors as the drug class with the strongest independent association with severe COVID-19.
+The rate ratio for severe COVID-19 associated with with proton pump inhibitor use varied from `r table.protonpump[4, 3]` in those aged 75 years and over to `r table.protonpump[1, 3]` in those aged under 40 years.
+Adjustment for hospital diagnoses of disorders of the esophagus, stomach and duodenum, and prescribing of antiplatelet or nonsteroidal anti-inflammatory agents reduced these effect sizes only slightly.
 
-Among those without any of the designated high-risk conditions (diabetes, heart disease, obstructive airway disease, chronic liver disease, chronic renal failure, neurological including dementia, immune deficiency / suppression) severe cases were highly significantly more likely than controls to have had prescriptions in the past year from each BNF chapter.
-The strongest associations (rate ratio, 95% CI) were seen with chapters for drugs acting on the gastro-intestinal system `r table.drugs.nocare.notlisted[1, 3]`, cardiovascular system `r table.drugs.nocare.notlisted[2, 3]`, central nervous system `r table.drugs.nocare.notlisted[4, 3]` and "Nutrition and Blood" `r table.drugs.nocare.notlisted[9, 3]`.
-When each of these were further disaggregated to subchapter or paragraph level, key findings included a significantly greater risk of severe COVID-19 associated with proton pump inhibitor use with rate ratios varying from `r table.protonpump[4, 3]` in those aged 75 years and over to `r table.protonpump[1, 3]` in those aged under 40 years.
-However many other drug classes associated with prescribing for non-specific symptoms were also associated with severe COVID-19: these included osmotic laxatives non-opioid analgesics, oral iron supplements and antipsychotic drugs.
-
-Among drugs prescribed for cardiovascular disease, only loop diuretics and calcium channel blockers were independently associated with risk of severe COVID-19 in a multivariate analysis.
-Associations with ACE inhibitors and angiotensin-II receptor blockers were similar to those for other antihypertensive agents.
-
-*Conclusions* -- Risk of severe COVID-19 is strongly associated with encashment of prescriptions for drugs that are widely prescribed for minor symptoms.
-It remains to be established whether this effect is accounted for by association with characteristics of patients who consult for these symptoms, or by direct drug effects.
-Of the drug classes that might possibly account for an association with prescribing, by far the strongest candidate is proton pump inhibitors, which are prescribed to about 30% of those aged over 60 years and are associated with a threefold elevation of risk of severe disease in this age range.
-As increased risk of infection is an established adverse effect of these drugs, a causal effect is plausible.
-This may be relevant to high rates of severe disease reported in other populations where this class of drugs is widely overprescribed.
-Until further studies are available, it is advisable to avoid using these drugs without a recognized medical indication.
+Other drug classes in Chapter 1, including laxatives and agents influencing motility, were associated with higher rates of severe COVID-19 but only proton pump inhibitors and osmotic laxatives were independently associated with disease in a multivariate analysis.
+Among drugs prescribed for cardiovascular disease, only loop diuretics and calcium channel blockers were independently associated with severe COVID-19 in a multivariate analysis.
+Associations of disease with ACE inhibitors and angiotensin-II receptor blockers were similar to those with other antihypertensive agents.
+                                                                                               *Conclusions* -- A systematic search for drug classes that underlie the association of severe COVID-19 with dispensing of any prescription has identified proton pump inhibitors as the drug class making the largest independent contribution to this association.
+The association with proton pump inhibitors is not explained by adjusting for covariates.
+As severe disease is associated with other drug classes such as laxatives that are commonly prescribed for minor symptoms, confounding by factors associated with these symptoms cannot be excluded.
+However as increased risk of community-acquired pneumonia is an established adverse effect of this class of drugs, a causal effect is plausible.
+In the Scottish population where 28\% of controls have used proton pump inhibitors and the rate ratio is 2.8, one-third of severe cases could be attributable to use of these drugs.
+This may be relevant to high rates of severe COVID-19 reported in other populations where proton pump inhibitors are overprescribed.
+Until the effects of proton pump inhibitors on susceptibility to severe COVID-19 have been investigated further, use of these drugs without a recognized medical indication is inadvisable.
 
@@
 # Background
-A striking preliminary finding of the REACT-SCOT study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year.
+A striking preliminary finding of the REACT-SCOT study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
 The rate ratio associated with this variable varied from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
@@
 The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
-The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.
 All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
@@
 
-For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
-As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
-With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.
+For each test-positive case, the CHI database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
+With this incidence density sampling design, odds ratios in conditional logistic regression models are equivalent to rate ratios.
 
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
-
 British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
-The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits) and subparagraph (1 digit), and chemical substance (2 characters).
-For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other"
+The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
+For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other".
+Most of these codes are for dressings and appliances rather than drugs.
 
@@
 Table \ref{tab:carehome} shows that the association was restricted to those not resident in a care home.
+In the next step, residents of care homes were excluded and the association with dispensing of any prescription was examined after stratifying by the presence or absence of a listed condition.
+Table \ref{tab:listed.any} shows that the association of severe COVID-19 with dispensed prescriptions was restricted to those without a listed condition.
 
-As a next step, we examined the association with encashment of any prescription after stratifying by the presence or absence of a listed condition, excluding those resident in a care home.
-Table \ref{tab:listed.any} shows that the association was restricted to those without a listed condition
+To investigate which drug classes contribute to this association, the analysis was restricted to those not resident in care homes and without a diagnosis of a listed condition.
+We undertook a backward stepwise variable selection procedure, based on BNF chapters which group drug classes by the body systems on which they act.
+Starting with the "any prescription" variable based on all 14 BNF chapters, models that dropped each of these chapters, retaining the other 13, were compared and the model with highest log-likelihood was retained.
+This procedure was repeated until the any prescription variable was based on a single remaining chapter.
+Table \ref{tab:stepwise.drop} shows the results of this process.
+Only the last five chapters decreased the log-likelihood when dropped: in order of deselection, these were chapters 6 (endocrine) 4 (nervous system, 10 (musculoskeletal), 1 (gastrointestinal) and 2 (cardiovascular).
 
-## Partitioning by BNF chapter
-We next partitioned the "any prescription" variable into indicator variables for each chapter of the British National Formulary.
-The chapter level groups drugs by the system on which they act.
-We restricted this analysis to those without any listed condition and not resident in a care home.
-Table \ref{tab:drugs} shows these associations.
+To investigate which drug classes in Chapters 1 and 2 contribute to the "any dispensed prescription effect, we repeated the backward stepwise variable selection procedure on all BNF subparagraph codes in these two chapters.
+Subparagraph codes correspond generally to drugs classified by their mode of action.
+Of the 48 subparagraph codes in Chapters 1 and 2, only the last ten retained by the stepwise variable selection decreased the log-likelihood when dropped.
+The results of the stepwise variable selection for these last ten are shown in Table \ref{tab:stepwise.subparas}.
+The cardiovascular drug classes retained at this stage were oral anticoagulants, angiotensin-II receptor antagonists, angiotensin-converting enzyme inhibitors, loop diuretics and calcium-channel blockers.
+The gastrointestinal drug classes retained at this stage were drugs altering gut motility, osmotic laxatives and proton pump inhibitors.
+The last variable retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.
 
-In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
+This led us to examine further the association of proton pump inhibitors with severe disease, without restricting the analysis to those not resident in care homes and without a diagnosis of a listed condition.
+Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
+The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.
 
-Table \ref{tab:explained.chapter} examines how much the effect of any prescription is reduced by adding an indicator variable for each chapter, one at a time.
-The chapter indicators that explained the highest fractions of the "any prescription" effect were Chapter 1 (Gastrointestinal System) , Chapter 2 (Cardiovascular System), Chapter 4 (Central Nervous System), and Chapter 9 (Nutrition), The codes grouped under Chapter 14+ (Other) are mostly for dressings and appliances, presumably reflecting a general association with frailty.
-This led us to examine associations with specific drug classes, grouped at BNF paragraph or subparagraph level, within each of these three chapters.
+Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.
+Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.
 
-## Associations with severe COVID-19 at section, paragraph and subparagraph level
-
-## Drugs for gastro-intestinal system: BNF chapter 1
-Table \ref{tab:bnfchapter1} shows that encashed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19.
-Of these by far the most commonly prescribed drug classes was proton pump inhibitors.
-Osmotic laxatives were also associated with severe disease, but were far less commonly prescribed.
-
-## Drugs for cardiovascular system: BNF chapter 2
-Table \ref{tab:bnfchapter2} lists associations with drugs for the cardiovascular system, grouped by BNF subparagraph.
-This shows that prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19.
+ Table \ref{tab:bnfchapter2} shows that We also examined the univariate and multivariate associations of prescribed drugs for the cardiovascular system (BNF chapter 2) with severe COVID-19.
+lists associations with drug classes (BNF subparagraphs) in this chapter.
+Prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19.
 In a multivariate analysis the strongest associations were with loop diuretics and calcium-channel blockers.
@@
 
-## Drugs acting on the central nervous system: BNF chapter 4
-Table \ref{tab:bnfchapter4} shows that encashed prescriptions of most classes of drugs acting on the central nervous system were associated with severe COVID-19.
-The most commonly prescribed drug class, and the strongest independent effect in a multivariate analysis, was non-opiate analgesics and compound preparations.
-This class of drugs includes paracetamol and various compound preparations that are marketed for relief of upper respiratory symptoms, headache, influenza, and other conditions.
-Other drug classes associated with severe COVID-19 in a multivariate analysis were antipsychotic agents and opioid analgesics.
+# Discussion 
+We have shown that the association of dispensed drug prescriptions with severe COVID-19 remains after excluding people who are at high risk for reasons that are associated with higher rates of prescribing; residence in care homes or a hospital diagnosis of a designated high-risk condition.
+We have shown further that this most of this association is accounted for by drugs prescribed for the gastrointestinal system and drugs prescribed for the cardiovascular system.
+There are smaller independent contributions from drugs prescribed for the nervous and musculoskeletal systems.
 
-## Drugs for Nutrition and Blood: BNF chapter 9
-Table \ref{tab:bnfchapter9} shows that prescriptions of several classes of nutritional supplements were associated with severe disease: these included oral iron supplements, drugs used in megaloblastic anemias, enteral nutrition products, and Vitamin D.
+Of the drugs prescribed for the cardiovascular system, the strongest independent effect is that of loop diuretics.
+This is likely to reflect an underlying association of heart failure with high risk of severe or fatal COVID-19.
+Prescription of ACE inhibitors or angiotensin-II receptor blockers is associated with elevated risk of severe disease, but the rate ratios are no higher than for other antihypertensive agents.
+This is likely to reflect an underlying assocaition of hypertension with severe disease.
+A more detailed study of the relation of COVID-19 to cardiovascular disease and drugs prescribed for cardiovascular disease will be reported elsewhere.
 
-## Association with proton pump inhibitors
-As drugs for the gastrointestinal system accounted for the largest fraction of the effect of "any prescription", and proton pump inhibitors were the most commonly prescribe drug in this class, we examined the association with proton pump inhibitors more closely.
-Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
-The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates, especially prescription of nonopiate analgesics.
+Of all the prescribed drug classes in Chapters 1 and 2 of the BNF, the strongest association was with proton pump inhibitors.
+This association with proton pump inhibitors is not explained by association with hospital diagnoses of diseases of the upper gastrointestinal tract, or by coprescribing with non-steroidal anti-inflammatory or antiplatelet agents.
+The highest rate ratios associated with this exposure are in those aged under 40.
 
-Table \ref{tab:analgesic.adjusted) shows the univariate and multivariate associations of severe COVID-19 with prescription of nonopioid analgesics .
+The possible effect of proton pump inhibitors on risk of severe COVID-19 was not one of the hypotheses that we specified in advance when setting up this case-control study.
+With hindsight, perhaps it should have been.
+In a systematic review of adverse effects of long-term use of proton pump inhibitors, all eight studies included showed elevated risk of community-acquired pneumonia [@islam_adverse_2018].
+In the largest of these studies, based on the UK General Practice Research Database from 1987 to 2002, elevated risk was restricted to proton pump inhibitor therapy started no more than 30 days before the date of diagnosis of pneumonia [@sarkar_proton-pump_2008].
+As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
 
-# Discussion 
-## Association with prescriptions for symptoms
+We note that in epidemiological experience rate ratios of 5 or more, as seen in those aged under 40 years in this study, usually have a causal explanation.
+However other drugs prescribed for the gastrointestinal system are associated with increased risk of severe disease, at least in univariate analyses.
+These include drugs such as laxatives that, like proton pump inhibitors, are used as indicators of prescribing quality in the National Therapeutic Index [macbride-stewart_national_2019].
+It is not possible at this stage to exclude the possibility that the association with proton pump inhibitors could be accounted for by underlying factors that lead to consultation for minor symptoms and prescribing of these drugs without a clear medical indication.
+In the UK and other countries proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].
+A study in Aberdeen reported that 35% of patients admitted to acute geriatric units had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of co-morbidity [@jarchow-macdonald_prescribing_2013].
 
-The list of prescribed drug classes that are associated with elevated risk of severe COVID-19 correspond closely to those identified by clinical pharmacologists and pharmacists as widely overprescribed drug classes: these include proton pump inhibitors, laxatives, analgesics, antipsychotic agents and nutritional supplements.
-
-Patients who consult with minor symptoms, leading to prescriptions of these drugs without a clear medical indication, are at high risk.
-The problem is to determine whether this association is accounted for by characteristics that underlie these consultations for minor symptoms, or by direct drug effects.
-
-Because this case-control study is matched on primary care practice, variation of over-prescribing between practices does not contribute to the conditional odds ratios.
-
-## Proton pump inhibitors
-The association with proton pump inhibitors is not explained by association any underlying gastrointestinal disease that leads to hospital admission, or by coprescribing with non-steroidal antiinflammatory or antiplatelet agents.
-As proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.
-However as there are no prescription charges in Scotland, long-term users of these expensive drugs are likely to seek to obtain them on prescription.
-
-As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
-Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].
-35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013].
-
-In Szechuan, China, PPIs were prescribed to 37% of inpatients in a tertiary referral centre in 2016 [@luo_changes_2018].
-A Spanish drug-utilization study showed that 55% of hospital inpatients were presribed a PPI at discharge [@ramirez_overuse_2010].
-
-## Cardiovascular drugs
-There is evidence of any effect of ACE inhibitors or angiotensin-II receptor blockers beyond the associated with treated hypertension.
+We can calculate that in the Scottish population where 28\% of controls have used proton pump inhibitors and the rate ratio is 2.8, one-third of severe cases could be attributable to prescribing of these drugs if the association is causal.
+This does not take into account possible over-the-counter purchases.
+This could be an undIt is possibly relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors had previously been reported.
+A study in Szechuan, China, reported that proton pump inhibitors were prescribed to 37% of inpatients in a tertiary referral centre in 2016 [@luo_changes_2018].
+In Italy, a survey of nursing homses found that 46% of residents had been prescribed a proton pump inhibitor [@pasina_overuse_2020].
+A Spanish drug-utilization study reported that 55% of hospital inpatients were presribed a PPI at discharge [@ramirez_overuse_2010].
 
 ## Methodological strengths and weaknesses
+A strength of this study is that it is based on dispensed prescriptions, rather than the prescriptions written in primary care.
+However as proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.
+Unless this this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
 
-We have prescription encashments, not just prescriptions.  Drug coding by BNF codes is more systematic than SNOMED
+A limitation of the data linkage undertaken by Health Protection Scotland is that we cannot construct longitudinal records of drug exposure with dose.
+However we have been able to do this in a separate study of all individuals with diabetes in Scotland, for whom we have constructed a curated database of drug exposure.
 
-We don't have all the morbidity data available in the primary care record but we have hospital diagnoses which are ICD-coded by trained coders.
+Although we do not have morbidity data from primary care, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
+This diagnostic coding is more relevant to disease classification than the SNOMED diagnostic codes used in primary care databases.
+
+Because this case-control study is matched on primary care practice, variation of prescribing between practices cannot confound the conditional odds ratios, though it is possible that variation between practitioners within the same practice could contribute to a confounding effect.
 
 ## Conclusion
-It is not possible to establish unequivocally whether the association with prescription encashment is accounted for by patient characteristics associated with prescribing for minor symptoms, or by a direct effect of any specific class of drugs.
-However of the drug classes that are associated with increased risk that is not explained by underlying high-risk conditions, proton pump inhibitors are the strongest candidate for a causal effect.
-Until further studies are available, it is advisable to avoid using drugs in this class without a recognized medical indication.
+Although some of the association of severe COVID-19 with dispensing of prescriptions can be accounted for underlying cardiovascular diseases that are not captured by hospital discharge diagnoses, there remains a residual association with prescription of proton pump inhibitors that is not easily explained by co-morbidity.
+Increased risk of community-acquired pneumonia is an established adverse effect of the use of these drugs, We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.
 
@@
 <div id="refs"></div>
+
+\newpage
 
@@
      kable_styling(latex_options="hold_position") %>%
-   column_spec(column=1, width="4cm")
+   column_spec(column=1, width="3cm")
 ```
@@
 
-```{r explained.chapter, echo=FALSE, warning=FALSE, message=FALSE}
+\newpage
 
-rownames(table.anyscrip.chapter) <- gsub("\\_", " ", rownames(table.anyscrip.chapter))
-knitr::kable(table.anyscrip.chapter,
+```{r stepwise.chapters, echo=FALSE, warning=FALSE, message=FALSE}
+
+stepwise.drop[, 1] <- gsub("\\_", " ", stepwise.drop[, 1])
+
+knitr::kable(stepwise.drop,
              escape=FALSE, 
              booktabs=TRUE,
-             label="explained.chapter",
-             col.names=c("Coefficient", "Rate.ratio", "Proportion explained by covariate"), 
-             digits=c(2, 2, 2), 
-             caption="Effect of an indicator variable for each BNF chapter on the size of effect associated with encashment of any prescription, in those not in care homes and without listed conditions") %>%
-    column_spec(column=1, width="3cm") %>%
+             label="stepwise.drop",
+             col.names=c("Dropped BNF chapter", "Log rate ratio before dropping",
+                         "Change in log-likelihood when dropped"), 
+             digits=c(0, 2, 1), 
+             caption="Stepwise dropping of BNF chapters from the 'any dispensed prescription' variable: dataset restricted to those not in care homes and without listed conditions") %>%
+    column_spec(column=1, width="4cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="3cm") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r stepwise.subparas, echo=FALSE, warning=FALSE, message=FALSE}
+
+stepwise.drop.subparas[, 1] <- gsub("subpara\\.", "", stepwise.drop.subparas[, 1])
+
+knitr::kable(stepwise.drop.subparas[-(1:29), ], 
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="stepwise.subparas",
+             row.names=FALSE,
+             col.names=c("Dropped drug class", "Log rate ratio before dropping",
+                         "Change in log-likelihood when dropped"), 
+             digits=c(0, 2, 1), 
+             caption="Stepwise dropping of last ten drug classes from a variable defined as 'any dispensed prescription from BNF chapters 1 or 2': dataset restricted to those not in care homes and without listed conditions") %>%
+    column_spec(column=1, width="9cm") %>%
     column_spec(column=2, width="2cm") %>%
     column_spec(column=3, width="2cm") %>%
-    column_spec(column=4, width="2cm") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.protonpump,
+             escape=FALSE,
+             label="protonpump.adjusted", 
+             booktabs=TRUE,
+             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe disease with prescription of a proton pump inhibitor in those not resident in a care home by age group, with and without adjustment for covariates: hospital diagnosis of GI tract disorder, prescriptions of NSAIDs,  antiplatelet agents") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
-    add_header_above(c(" "=1, "Adjusted effect"=2))
+    landscape()
 
@@
 ```
-
-```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter4,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter4",
-             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 4 in those without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter4)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```
-
-```{r bnfchapter9, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter9,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter9",
-             col.names=c(clean.header(colnames(table.bnfchapter9)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 9 in those without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    group_rows("", 1, nrow(table.bnfchapter9)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```
-
-```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.protonpump,
-             escape=FALSE,
-             label="protonpump.adjusted", 
-             booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with prescription of a proton pump inhibitor by age group, with and without adjustment for hospital diagnosis of any upper GI tract disorder, and prescription of NSAIDs,  antiplatelet agent, and nonopioid analgesics, excluding care home residents") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```
-
-```{r analgesic, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.nonopioid.analgesic,
-             escape=FALSE,
-             label="analgesic.adjusted", 
-             booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.nonopioid.analgesic)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with prescription of a nonopioid analgesic by age group, with and without adjustment for hospital diagnosis of any neoplasm, and prescription of  NSAIDs, antiplatelet agents and proton pump inhibitors, excluding care home residents") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```

### 2020-05-20 helperfunctions.R

@@ -1,6 +1,127 @@
 
 ## helperfunctions for COVID analyses
 
+select.union <- function(x, y, stratum) {
+    ## x is matrix of indicator variables
+    ## select column of x to drop that maximizes loglik of clogit model using all other cols of x
+    loglik.dropcol <- numeric(ncol(x))
+    x.u <- as.integer(rowSums(x) > 0)
+    model.full <- summary(clogit(y ~ x.u + strata(stratum)))
+    beta.full <- model.full$coefficients[1, 1]
+    loglik.full <- model.full$loglik[2]
+    for(j in 1:ncol(x)) {
+        x.u <- as.integer(rowSums(x[, -j, drop=FALSE]) > 0)
+        loglik.dropcol[j] <- summary(clogit(y ~ x.u + strata(stratum)))$loglik[2]
+    }
+
+    j.max <- which.max(loglik.dropcol)
+    return(data.frame(col.drop=j.max,
+                      name=colnames(x)[j.max],
+                      beta.full=beta.full,
+                      loglik.full=loglik.full))
+}
+
+stepwise.union.dropcols <- function(x, y, stratum) {
+    x.drop <- x
+    stepwise.drop <- NULL
+    for(j in 1:(ncol(x) - 1)) {
+        select.drop <- select.union(x=x.drop, y=y, stratum=stratum)
+        stepwise.drop <- rbind(stepwise.drop, select.drop)
+        x.drop <- x.drop[, -select.drop$col.drop, drop=FALSE]
+    }
+    ## add row for last variable retained
+    last.name <- colnames(x)[!colnames(x) %in% stepwise.drop$name]
+    last.col <- match(last.name, colnames(x))
+    last.model <- summary(clogit(y ~ x[, last.col] + strata(stratum)))
+    stepwise.drop <- rbind(stepwise.drop,
+                           data.frame(col.drop=NA,
+                                      name=last.name,
+                                      beta.full=last.model$coefficients[1, 1],
+                                      loglik.full=last.model$loglik[2]))
+    stepwise.drop$loglik.full <- stepwise.drop$loglik.full - stepwise.drop$loglik.full[1]
+    stepwise.drop$loglik.full <- c(stepwise.drop$loglik.full[-1], NA)
+    return(stepwise.drop[, -1])
+}
+
+lookup.names <- data.frame(varname=c("deathwithin28", "scrip.any", "diag.any", "care.home",
+                                     "emerg", "icu.hdu.ccu", "inpat",
+                                     "protonpump",
[…]

### 2020-05-20 incidencemortality.R

@@ -1,0 +1,86 @@
+####### incidence and mortality using national population estimates ######################
+
+case.freqs <- with(cc.severe[cc.severe$CASE==1, ], table(AGE, sex, exclude=NULL))
+case.freqs <- data.frame(Age=as.integer(rownames(case.freqs)),
+                         Females=as.integer(case.freqs[, 1]),
+                         Males=as.integer(case.freqs[, 2]))
+case.long <- reshape2::melt(case.freqs, id="Age")
+colnames(case.long) <- c("Age", "Sex", "Cases")
+
+death.freqs <- with(cc.severe[cc.severe$fatalcase==1, ], table(AGE, sex, exclude=NULL))
+death.freqs <- data.frame(Age=as.integer(rownames(death.freqs)),
+                         Females=as.integer(death.freqs[, 1]),
+                         Males=as.integer(death.freqs[, 2]))
+death.long <- reshape2::melt(death.freqs, id="Age")
+colnames(death.long) <- c("Age", "Sex", "Deaths")
+
+scotpop.long <- reshape2::melt(scotpop[, -2], id="Age")
+colnames(scotpop.long) <- c("Age", "Sex", "Population")
+
+discrim <- merge(scotpop.long, case.long, by=c("Age", "Sex"), all.x=TRUE)
+discrim <- merge(discrim, death.long, by=c("Age", "Sex"), all.x=TRUE)
+discrim$Cases[is.na(discrim$Cases)] <- 0
+discrim$Deaths[is.na(discrim$Deaths)] <- 0
+
+discrim$Sex <- as.factor(discrim$Sex)
+
+discrim$Noncases <- discrim$Population - discrim$Cases
+y.cases <- cbind(as.integer(discrim$Cases), as.integer(discrim$Noncases))
+
+discrim$Survivors <- discrim$Population - discrim$Deaths
+y.deaths <- cbind(as.integer(discrim$Deaths), as.integer(discrim$Survivors))
+
+
+cases.model <- glm(formula=y.cases ~ Sex + Age, family="binomial", data=discrim)
+deaths.model <- glm(formula=y.deaths ~ Sex + Age, family="binomial", data=discrim)
+
+cases.model.coeffs <- summary(cases.model)$coefficients
+deaths.model.coeffs <- summary(deaths.model)$coefficients
+logistic.coeffs <- data.frame(severecase=cases.model.coeffs[, 1],
+                              death=deaths.model.coeffs[, 1])
+
+male <- discrim$Sex=="Males"
[…]

### 2020-05-21 pharmaco.Rmd
extra diabetes cases from linkage included

@@
 From the national population register ten controls per case were selected, matched for sex, age and general practice.
-International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from encashed primary care issued prescriptions in the past year.
+International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary in the past nine months.
 Rate ratios were estimated by conditional logistic regression.
 
-_Findings_ -- In those not resident in a care home and without past diagnoses of any conditions designated as high risk, the rate ratio (95% CI) associated with dispensing of any prescription in the last year was `r table.scrip.any.nocare.listed[1, 3]`.
+_Findings_ -- In those not resident in a care home and without past diagnoses of any conditions designated as high risk, the rate ratio (95% CI) associated with dispensing of any prescription in the last year was `r table.scrip.cats3[3, 3]`.
 A backward selection procedure showed that two chapters of the BNF -- drugs for the gastrointestinal system (Chapter 1) and the cardiovascular system (Chapter 2) -- accounted for `r round(100 * stepwise.drop[13, 2] / stepwise.drop[1, 2])`% of this effect.
@@
 # Results
-## Encashment of any prescription by care home status and presence of any listed condition
-As residence in a care home was the strongest risk factor for severe COVID-19, we first examined the association with enchashment of any prescription after stratifying by care home status.
-Table \ref{tab:carehome} shows that the association was restricted to those not resident in a care home.
-In the next step, residents of care homes were excluded and the association with dispensing of any prescription was examined after stratifying by the presence or absence of a listed condition.
-Table \ref{tab:listed.any} shows that the association of severe COVID-19 with dispensed prescriptions was restricted to those without a listed condition.
-
-To investigate which drug classes contribute to this association, the analysis was restricted to those not resident in care homes and without a diagnosis of a listed condition.
+## Dispensing of any prescription by care home status and presence of any listed condition
+Table \ref{tab:scrip.cats3} shows that at least one prescription had been dispensed for `r table.scrip.cats3[1, 1]` of controls who were resident in care homes and `r table.scrip.cats3[2, 1]` of controls a diagnosis of a listed condition who were living independently.
+The association of severe disease with dispensing of at least one prescription was restricted to those not resident in a care home and without a diagnosis of a listed condition, in whom the rate ratio (95% CI) was `r table.scrip.cats3[3, 3]`.
+To investigate which drug classes contribute to this association, the analysis was therefore restricted to this group.
 We undertook a backward stepwise variable selection procedure, based on BNF chapters which group drug classes by the body systems on which they act.
@@
 Table \ref{tab:stepwise.drop} shows the results of this process.
-Only the last five chapters decreased the log-likelihood when dropped: in order of deselection, these were chapters 6 (endocrine) 4 (nervous system, 10 (musculoskeletal), 1 (gastrointestinal) and 2 (cardiovascular).
+Only the last four chapters decreased the log-likelihood when dropped: in order of deselection, these were chapters 10 (musculoskeletal), 9 (nutrition and blood), 1 (gastrointestinal) and 2 (cardiovascular).
 
-To investigate which drug classes in Chapters 1 and 2 contribute to the "any dispensed prescription effect, we repeated the backward stepwise variable selection procedure on all BNF subparagraph codes in these two chapters.
+To investigate which drug classes in Chapters 1 and 2 contribute to the "any dispensed prescription" effect, we repeated the backward stepwise variable selection procedure on all BNF subparagraph codes in these two chapters.
 Subparagraph codes correspond generally to drugs classified by their mode of action.
@@
 In a systematic review of adverse effects of long-term use of proton pump inhibitors, all eight studies included showed elevated risk of community-acquired pneumonia [@islam_adverse_2018].
-In the largest of these studies, based on the UK General Practice Research Database from 1987 to 2002, elevated risk was restricted to proton pump inhibitor therapy started no more than 30 days before the date of diagnosis of pneumonia [@sarkar_proton-pump_2008].
 As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
+In the largest study, based on the UK General Practice Research Database from 1987 to 2002, elevated risk was restricted to those for whom prescribing of proton pump inhibitors had been initiated no more than 30 days before the date of diagnosis of pneumonia [@sarkar_proton-pump_2008].
+A possible explanation for this is that at least in the UK, only those in whom prescribing of these drugs has recently been initiated are likely to be taking them continuously.
 
@@
 However other drugs prescribed for the gastrointestinal system are associated with increased risk of severe disease, at least in univariate analyses.
-These include drugs such as laxatives that, like proton pump inhibitors, are used as indicators of prescribing quality in the National Therapeutic Index [macbride-stewart_national_2019].
+These include drugs such as laxatives that, like proton pump inhibitors, are used as indicators of prescribing quality in the National Therapeutic Index [@macbride-stewart_national_2019].
 It is not possible at this stage to exclude the possibility that the association with proton pump inhibitors could be accounted for by underlying factors that lead to consultation for minor symptoms and prescribing of these drugs without a clear medical indication.
@@
 We can calculate that in the Scottish population where 28\% of controls have used proton pump inhibitors and the rate ratio is 2.8, one-third of severe cases could be attributable to prescribing of these drugs if the association is causal.
-This does not take into account possible over-the-counter purchases.
-This could be an undIt is possibly relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors had previously been reported.
+Although the rate ratio associated with use at the time of exposure to coronavirus may be underestimated because all those with a prescription in the last nine months are clasified as exposed, the attributable risk fraction will not be underestimated, at least not for prescribed use of these drugs.
+However this does not take into account possible over-the-counter purchases.
+
+It is possibly relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors had previously been reported.
 A study in Szechuan, China, reported that proton pump inhibitors were prescribed to 37% of inpatients in a tertiary referral centre in 2016 [@luo_changes_2018].
@@
 ## Methodological strengths and weaknesses
-A strength of this study is that it is based on dispensed prescriptions, rather than the prescriptions written in primary care.
-However as proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.
+A strength of this study is that it is based on dispensed prescriptions, rather than issued  prescriptions.
+However as proton pump inhibitors are available over the counter, our study of dispensed prescriptions does not capture all exposure to this class of drugs.
 Unless this this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
@@
 
-```{r carehome, echo=FALSE, warning=FALSE, message=FALSE}
+```{r cats3, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.scrip.any.carehome,
+knitr::kable(table.scrip.cats3,
              escape=FALSE, 
              booktabs=TRUE,
-             label="carehome",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.scrip.any.carehome)[1:2]),
+             label="scrip.cats3",
+             col.names=c(clean.header(colnames(table.scrip.cats3)[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 6), 
-             caption="Associations of severe disease with encashment of any prescription in the last year, by care home status") %>%
-     kable_styling(latex_options="hold_position") %>%
-   column_spec(column=1, width="3cm")
-```
-
-```{r listed, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.scrip.any.nocare.listed,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="listed.any",
-             col.names=c(clean.header(colnames(table.scrip.any.nocare.listed)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with encashment of any prescription in the last year, by presence of a listed condition, excluding care home residents") %>%
+             caption="Associations of severe disease with encashment of any prescription in the last year, by residence status and diagnosis of a listed condition") %>%
      kable_styling(latex_options="hold_position") %>%
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with prescription of a proton pump inhibitor in those not resident in a care home by age group, with and without adjustment for covariates: hospital diagnosis of GI tract disorder, prescriptions of NSAIDs,  antiplatelet agents") %>%
+             caption="Associations of severe disease with prescription of a proton pump inhibitor, with and without adjustment for covariates: care home residence, hospital diagnosis of upper GI tract disorder, prescriptions of NSAIDs,  antiplatelet agents") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those not resident in a care home and without without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those not resident in a care home and without a listed condition") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those without listed conditions") %>%
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those not resident in a care home and without a listed condition") %>%
     column_spec(column=1, width="4cm") %>%
@@
 ```
+
+```{r bnfchapter9, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.bnfchapter9,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="bnfchapter9",
+             col.names=c(clean.header(colnames(table.bnfchapter9)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 9 in those not resident in a care home and without a listed condition") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+```
+
+```{r bnfchapter10, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.bnfchapter10,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="bnfchapter10",
+             col.names=c(clean.header(colnames(table.bnfchapter10)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with prescribed drugs in BNF chapter 10 in those not resident in a care home and without a listed condition") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+```

### 2020-05-21 casecontrol.R

@@ -38,7 +38,6 @@ scrips$sectioncode <- as.integer(substr(scrips$bnf_paragraph_code, 1, 4))
 scrips$chapternum[is.na(scrips$chapternum)] <- 14
 scrips$chapternum[scrips$chapternum > 14] <- 14
 
-
 icdchapters <- data.frame(names(icd10_chapters),
                              t(matrix(as.character(unlist(icd10_chapters)), nrow=2)))
 colnames(icdchapters) <- c("name", "start", "end")
@@ -173,23 +172,43 @@ cc.all$casegroup <- car::recode(cc.all$casegroup,
 cc.all$casegroup <- as.factor(cc.all$casegroup)
 cc.all <- within(cc.all, casegroup <- relevel(casegroup, ref="Critical care or fatal"))
 
-## recode diabetes type
+########################################################################################
+###  diabetes based on combining Sci-Diabetes records with ICD codes and drugs 
+## add in BNF codes 6.1 for diabetes drugs and
+## E10 to E14 for diabetes
+
+## diabetes.any includes primary immunodeficiency and secondary immunosuppression
+ids.icd.diabetes <- unique(diagnoses$ANON_ID[grep("^E1[0-4]", diagnoses$ICD10)])
+
+## 802 other immunomodulating drugs
+## Methotrexate and chloroquine appear in musculoskeletal chapter 
+ids.bnf.diabetes <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 601])
+
+ids.diabetes.extra <- unique(c(ids.icd.diabetes, ids.bnf.diabetes))
+
+# recode diabetes type
 cc.all$dm.type <- as.integer(cc.all$dm.type)
 ## missing recoded as zero
 cc.all$dm.type[is.na(cc.all$dm.type)] <- 0
-cc.all$diabetes.any <- as.factor(car::recode(cc.all$dm.type, 
-                                             "0='Not diabetic'; 1:hi='Diabetic'"))
-cc.all <- within(cc.all, diabetes.any <- relevel(diabetes.any, ref="Not diabetic"))
+## code diagnoses detected from discharges or BNF codes as unknown type
+## we could classify those not on insulin as definite Type 2
+cc.all$dm.type[cc.all$dm.type & cc.all$ANON_ID %in% ids.diabetes.extra] <- 3
+
 cc.all$dm.type <- 
   as.factor(car::recode(cc.all$dm.type, 
                         "c(0, 10)='Not diabetic';
[…]

### 2020-05-21 comorbidity.R

@@ -28,7 +28,7 @@ cc.severe$heart.other.any <- as.factor(as.integer(cc.severe$ANON_ID %in% ids.hea
 
 ############# chronic kidney disease ##########################
 
-## this includes CKD stage 3
+## this includes CKD stage 4
 ids.icd.ckd <- unique(diagnoses$ANON_ID[grep("^N18[45]|^Z49[0-2]|^Z94[02]",
                                              diagnoses$ICD10)])
 
@@ -87,18 +87,6 @@ ids.bnf.immune <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 802 |
 ids.immune.any <- unique(c(ids.icd.immune, ids.bnf.immune))
 cc.severe$immune.any <- as.factor(as.integer(cc.severe$ANON_ID %in% ids.immune.any))
 
-
-#######################################################################################
-###  diabetes based on combining Sci-Diabetes records with ICD codes and drugs ############
-
-## add in BNF codes 6.1 for diabetes drugs and
-## E10 to E14 for diabetes
-
-###### seven listed conditions,  should add diabetes.any ###############################
-
-listed.conditions <- c("IHD.any", "heart.other.any", "oad.any",
-                       "ckd.any", "neuro.any", "liver.any", "immune.any")
-
 ############# neoplasms ################
 
 ids.icd.neoplasm <- unique(diagnoses$ANON_ID[grep("^C[0-9]|^D[0-4]", diagnoses$ICD10)])
@@ -106,8 +94,7 @@ ids.bnf.neoplasm <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 801]
 ids.neoplasm.any <- unique(c(ids.icd.neoplasm, ids.bnf.neoplasm))
 cc.severe$neoplasm.any <- as.factor(as.integer(cc.severe$ANON_ID %in% ids.neoplasm.any))
 
-
-###### disorders of esophagus, stomach and duodenum ################################
+###### disorders of esophagus, stomach and duodenum ############################
 
 ids.esoph.stomach.duod <-  unique(diagnoses$ANON_ID[grep("^K2[0-9]|^K3[01]", diagnoses$ICD10)])
 cc.severe$esoph.stomach.duod <-

### 2020-05-22 casecontrol.Rmd
fixed to keep memory usage down, also to add obsolete SMR ethnicity codes

@@
 ---
-title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population [**Draft, not to be quoted**]'
+title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population '
 author: 
-  - name: Paul M McKeigue
+  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
     email: paul.mckeigue@ed.ac.uk
-    affiliation: Usher
+    affiliation: HPS
   - name: Amanda Weir 
@@
     affiliation: IGMM
+  - name: Alison Smith-Palmer 
+    affiliation: HPS
+  - name: David Goldberg
+    affiliation: HPS
   - name: Sharon Hutchinson 
     affiliation: GCU
-  - name: Helen M Colhoun
-    affiliation: IGMM
+  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
+    affiliation: HPS
+  - name: Public Health Scotland COVID-19 Health Protection Study Group
+    affiliation: HPS
 address: 
@@
 NL - Clinical Senior Lecturer in Critical Care
+
   - code: IGMM
@@
 TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.
+
+  - code: HPS
+    address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health
-   - code: HPS
-     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
+
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health
+
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.
+
 geometry: margin=1in
@@
 bibliography: ../covid.bib
-date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
+
 header-includes: |
@@
     keep_tex: true
-#    toc: true
-#    toc_depth: 2
+
 csl: ../plos.csl
 urlcolor: red
-
-# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol.pdf")
-
 ---
@@
 library(kableExtra)
+
+# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol.pdf")
 
@@
 _Methods_ -- The design was a matched case-control study.
-Test-positive cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database and from national mortality data.
-Those who had been admitted to intensive care or had died within 28 days of a positive test were classified as severe cases.
+Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database.
+Those who had been admitted to an intensive care unit or had died within 28 days of a positive test were classified as severe cases.
 For each case, ten controls matched for sex, age and primary care practice were selected from the population register.
@@
 Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.
-The performance of a predictive model for severe COVID-19 was estimated by cross-validation.
+The performance of predictive models for severe COVID-19 was estimated by cross-validation.
 
 _Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
-In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]` .
-Rate ratios (95% CIs) for conditions previously designated as conferring high risk were
-`r table.agegr.all[4, 3]` for diabetes, 
-`r table.agegr.all[8, 3]` for chronic renal failure, 
-`r table.agegr.all[5, 3]` for ischemic heart disease,
-`r table.agegr.all[6, 3]` for other heart disease, 
-`r table.agegr.all[9, 3]` for neurological disease, 
-`r table.agegr.all[10, 3]` for chronic liver disease and 
-`r table.agegr.all[11, 3]` for immune deficiency or suppression.
+In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
+Univariate rate ratios (95% CIs) for eight conditions listed by public health agencies as conferring high risk were `r table.agegr.all[7, 3]` for Type 2 diabetes,
+`r table.agegr.all[8, 3]` for ischemic heart disease,
+`r table.agegr.all[9, 3]` for other heart disease, 
+`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
+`r table.agegr.all[11, 3]` for chronic kidney disease, 
+`r table.agegr.all[12, 3]` for neurological disease, 
+`r table.agegr.all[13, 3]` for chronic liver disease and 
+`r table.agegr.all[14, 3]` for immune deficiency or suppression.
 
-Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of these listed conditions.
-In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[13, 2])` in cases and `r gsub("\\\\", "", table.agegr[13, 1])` in controls.
+`r freqs.all[16, 2]` of the severe cases and `r freqs.all[16, 1]` of the controls had at least one of these listed conditions.
+In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[16, 2])` in cases and `r gsub("\\\\", "", table.agegr[16, 1])` in controls.
 The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
 These associations remained after excluding those with any of the listed conditions.
-The associations with hospital diagnoses were distributed over most chapters of the ICD-10 and the associations with prescribed drugs were distributed over most chapters of the British National Formulary.
 
 With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
-The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, `r table.densities[2, 5]` bits for a model that included the listed conditions, and `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
+The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, increasing to `r table.densities[2, 5]` bits for a model that included the listed conditions, and to `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
 
@@
 
-We report a systematic examination of the health care record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population.
+We report a systematic examination of the hospitalisation record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population.
 The objectives were to identify risk factors and to lay the basis for risk stratification based on a predictive model.
@@
 ## Case definition 
-The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
-The CHI number is a unique identifier used in all care systems in Scotland.
-All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
-Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and from the Scottish Morbidity Record 01 which is updated weekly.
-Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all critical care units in Scotland.
-Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
+The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.
+Individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained.
+Using the Community Health Index (CHI) identifier contained in ECOSS (the CHI number is a unique identifier used in all care systems in Scotland) linkage to other datasets was carried out.
+Hospital admissions from the time of testing were obtained from the RAPID database a daily return of current hospitalisations each day.
+Admissions to intensive care units (ICUs) care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all admissions to all adult ICUs and to combined ICU/high dependency units ( hereafter abbreviated as ICU) across Scotland and has returned a daily census of patients in ICU from the beginning of the COVID-19 epidemic.
+Death registrations up to `r gsub("^0", "", format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y"))` were obtained from linkage to the National Register of Scotland.
 
-Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-By restricting the case definition to cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from critical care.
+Severe or fatal COVID-19 was defined by a record of entering ICU in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
+By restricting the case definition to cases that were fatal or received ICU care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without ICU care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from ICU care.
 
 ## Matched controls
-For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
+For each test-positive case, we ascertained ten matched controls of the same sex, one-year age band and registered with the same primary care practice who were alive on the date of the first test in the case using the Community Health Index (CHI) database.
 As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
@@
 ## Demographic data
-Socioeconomic status was scored in the CHI database as the Scottish Index of Multiple Deprivation, an indicator based on postal code.
+Residence in a care home was ascertained from the CHI database.
+Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation, an indicator based on postal code.
 Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.
-This classification is able to distinguish South Asians, the largest non-White minority in Scotland, from those who identify as White with fair accuracy but misclassifies most of those who identify as Black or Black-Caribbean (a much smaller minority in Scotland) as White.
-<!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report or by hospital staff and based on the categories used in the Census.
+Cross-tabulation of a subset of records for which ethnicity assigned by patients or health care staff had been recorded in the Scottish Morbidity Record showed that this algorithm correctly classified 90% of those who reported South Asian ethnicity (1.6% of the  Scottish population in 2011) but misclassified most of those who identified as Caribbean or Black (0.1% of the population in 2011) as White.
+<!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report and based on the categories used in the Census.
 --->
@@
 For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge.
-British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive.
-Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
+Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric hospital and pregnancy units returns are not captured in SMR01.
+British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
+The BNF groups drugs by 2-digit chapter codes.
+For this analysis prescription codes from chapters 14 and above, mostly for dressings and appliances but also including vaccines were grouped as "Other".
 
-We began by scoring a specific list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-These listed conditions were assigned based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes.
-The ICD-10 codes and drug codes used to define each variable are listed in the R script provided as Supplementary Material.
-In addition to these pre-specified groupings of codes we tabulated associations with all ICD-10 codes and BNF codes to chapter level.
-A
-separate list of conditions has been used to designate "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
+We began by scoring a specific list of eight conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+A separate list of conditions designates "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
 We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.
+
+The eight listed conditions were scored based on diagnostic codes in any hospital discharge record during the last five years, or encashed prescription of a drug for which the only indications are in that group of diagnostic codes.
+The R script included as supplementary material contains the derivations of these variables from ICD-10 codes and BNF drug codes.
+Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.
+Cases of diabetes in the last update of the register were supplemented with new cases identified through discharge codes and drug codes.
+These new cases were coded as type Other/Unknown.
 
@@
 To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.
-For the case-control study, estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.
+To assess whether allowing for a non-linear relationship of the logit of risk to age would improve the fit, we also fitted generalized additive models, implemented in the R function `gam::gam`, with default smoothing function.
+
+For the case-control study,  all estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.
 Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.
-Where the sample of cases and controls is restricted, this will generate strata that contain no cases.
-These strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
+Where an exclusion criterion such as having a pre-defined condition was applied to cases  this was also applied to controls as otherwise subsequent association estimates would be incorrect.
+Where the sample of cases and controls is restricted, this will generate strata that contain no cases but these strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
 With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.
 These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
+Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios.
 
 To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.
-The performance of the risk prediction model as a classifier was examined by 4-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination $\Lambda$ [@mckeigue_quantifying_2019] and (more conventionally) as the C-statistic.
-The rationale for using $\Lambda$ to quantify the predictive performance of a binary classifier, and especially to quantify increments in predictive performance, has been explained elsewhere [@mckeigue_quantifying_2019].
-A key advantage of using $\Lambda$ is that the contributions of independent predictors are additive on the scale of $\Lambda$.
-The class-conditional distributions of weights of evidence in cases and controls encode exactly the same information as the receiver operating characteristic curve, when these distributions are adjusted for mathematical consistency as described elsewhere [@mckeigue_quantifying_2019] to obtain the model-based distributions.
-The likelihood ratio at any given sensitivity and specificity is the gradient of the model-based ROC curve.
-
-In this study we have estimated the information for discrimination provided by age and sex (from a logistic regression model in the general population), and the information for discrimination provided by other risk factors conditional on age and sex (from conditional logistic regression models in this case-control study matched for age and sex).
-We can add these to obtain the total information for discrimination that would be obtained with a classifier in the general population.
-As described elsewhere, this can be used to estimate the proportion of those destined to develop severe or fatal disease who would be correctly classified as high risk under a shielding policy [@mckeigue_evaluation_2020].
+The performance of the risk prediction model in classifying cases versus non-cases of severe COVID-19 was examined by cross-validation.
+We calculated the performance calculated over all test folds using the C-statistic but also using the "expected information for discrimination" $\Lambda$ expressed in bits [@mckeigue_quantifying_2019].
+The use of bits (logarithms to base 2) to quantify information is standard in information theory: one bit can be defined as the quantity of information that halves the hypothesis space.
+Although readers may be unfamiliar with the expected information for discrimination, it has several properties that make it more useful than the C-statistic for quantifying the performance of a risk prediction model.
+[@mckeigue_quantifying_2019] .
+A key advantage of using $\Lambda$ is that contributions of independent predictors can be added.
+Thus in this study we can add the predictive information from a logistic model of age and sex in the general population to the predictive information provided by other risk factors from the case-control study matched for age and sex.
 
 ## Information governance
-This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147 . Datasets were de-identified before analysis.
+This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.
 
@@
 Figure \ref{fig:rates} shows the relationships of incidence and mortality rates to age for each sex separately.
-The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for males and females are almost exactly parallel.
+The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for male and female mortality are almost parallel.
 For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.
-This implies that at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical care admissions to total fatalities is higher in men.
+This implies that at younger ages the ratio of ICU admissions to total fatalities is higher in women than in men, but that at later ages the ratio of ICU admissions to total fatalities is higher in men.
 In models that included age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
 The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.
-As shown in Table \ref{tab:agegr} almost all severe cases over aged 75 years died but most severe cases under aged 60 survived.
+As shown in Table \ref{tab:agegr} almost all severe cases aged over 75 years died but most severe cases aged under 60 years survived.
 
@@
 Higher risk of severe disease was also associated with socioeconomic deprivation.
-In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.
-Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.
-The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.
+In comparison with those classified as White, South Asians were at lower risk of severe disease.
+<!---Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.
+The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.--->
 
@@
 The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
-`r gsub("\\\\", "", table.agegr[13, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[13, 1])` of the controls.
-In those aged 75+ years `r gsub("\\\\", "", table.agegr[13, 8])` of the cases and `r gsub("\\\\", "", table.agegr[13, 7])` of the controls had at least one listed condition.
+`r gsub("\\\\", "", table.agegr[16, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[16, 1])` of the controls.
+In those aged 75+ years `r gsub("\\\\", "", table.agegr[16, 8])` of the cases and `r gsub("\\\\", "", table.agegr[16, 7])` of the controls had at least one listed condition.
 
@@
 All listed conditions were associated with severe disease in each age band.
-Diabetes was associated with a 6.6 fold rate of severe COVID-19 under age 60 years.
-The associations with severe COVID-19 were stronger for the broad category "other heart disease" than for ischaemic heart disease.
-
-Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} also show multivariate rate ratios by age group.
-These table show that even after adjusting for care home residence and listed conditions, encashment of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease.
+In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][6, 3]` for Type 1 diabetes and `r tables.agegr[[1]][7, 3]` for Type 2 diabetes.
+For all age groups combined, the rate ratio was `r tables.agegr[[1]][8, 3]` for ischemic heart disease and `r tables.agegr[[1]][9, 3]` for the broad category "other heart disease".
+The multivariate analyses shown in Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that in each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
 
 ## Systematic analysis of diagnoses associated with severe disease
-The approximately two-fold risk of with severe COVID-19 associated with prior hospital admission independently of other conditions and factors was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.
+The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.
 These results are shown in Table \ref{tab:conditions}.
-In univariate analyses, almost all ICD chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-The strongest residual associations were seen with neoplasms.
-Table \ref{tab:icdsubchapters} extracts univariate associations with ICD subchapters where the p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.
+In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
+The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (Mental) and 10 (musculoskeletal).
+Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for whcih the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.
+This was restricted to those without any listed conditions.
 This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.
@@
 In a multivariate analysis of all chapters, these associations were much weaker.
-The BNF chapter with the strongest association with severe disease was Chapter 1 (gastrointestinal).
-Other chapters associated with severe disease were Chapter 2 (cardiovascular) and Chapter 4 (central nervous system).
+The BNF chapters with the strongest association with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular)).
+Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
 
@@
 
-This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier that would be obtained in the population to give a total of `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
+This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier that would be obtained in the population as `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
 
-Figure \ref{fig:wevid} shows the class-conditional distributions of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set.
-The model-based distributions have been adjusted slightly to make them mathematically consistent as described elsewhere [@mckeigue_quantifying_2019].
+Figure \ref{fig:wevid} shows the distribution of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set with a footnote explaining how $\Lambda$ is derived.
 This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
-Figure \ref{fig:roc} shows the corresponding receiver operating characteristic (ROC) curve, which encodes the same information as Figure \ref{fig:wevid} but does not allow the separation of cases and controls by the risk classifier to be easily visualized.
+Figure \ref{fig:roc} shows the receiver operating characteristic curve with a footnote explaining its derivation from the distribution of the weights of evidence.
 
 # Discussion  
-### Sociodemographic features
-In this analysis of sociodemographic factors and past medical history we found a large number of factors to contribute to risk of severe COVID-19 disease.
-Apart from the very steep rise in risk with age, severe disease is more likely in men, and those who are more socioeconomically disadvantaged.
-Regarding sex differences we noted that the ratio of non-fatal cases who receive critical care to fatal cases is lower in females than in males in the 50-70 year age range.
-This may reflect either an underlying divergence in case fatality, or that women are less likely than men to be transferred to critical care if they have a good chance of surviving without critical care.
-The effect of matching on general practice which will partly match for socioeconomic disadvantage will reduce the precision of the effect of the socioeconomic indicator (SIMD) in our analysis but will not bias it in either direction.
-Residence in a care home was associated with a 15-fold increased risk of severe COVID-19 in this age matched analysis and this was reduced to eight-fold by adjustment for listed conditions.
-This eight-fold risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.
+### Sociodemographic factors
+This analysis shows that, as expected, risk for severe COVID-19 is associated with age, male sex and socioeconomic deprivation.
+The slope of the relationship of severe disease (on the scale of log odds) to age is less steep than the slope of the relationship of fatal disease to age.
+Another way of expressing this is that the fatality rate of severe disease increases with age.
+Although the relationships of mortality to age are linear and parallel for men and women, the fatality rate of severe disease is lower in women than in men up to age 50 and higher in women than in men between ages 50 and 70.
+This may reflect either variation in the biology of the disease, or differential selection for entry to intensive care.
+Residence in a care home was associated with a `r round(as.numeric(substr(table.agegr.all[1, 3], 1, 4)))`-fold increased rate of severe COVID-19 in this age matched analysis, reduced to `r round(as.numeric(substr(table.agegr.all[1, 5], 1, 4)))`-fold by adjustment for listed conditions.
+This excess risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.
 
-In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity, and other non-White minorities, are at lower risk than Whites.
+In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity are at lower risk than Whites.
 This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.
 In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-We note that the proportion of health care workers that are of non-British nationality varies across the UK from [23% in London to 7% in Scotland](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/internationalmigration/articles/internationalmigrationandthehealthcareworkforce/2019-08-15).
 
@@
 However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages.
-The 2.7 fold rate ratio in all ages combined is similar to that reported in UK Biobank and in the OpenSAFELY studies.
+The rate ratios of `r substr(table.agegr.all[6, 3], 1, 4)` for Type 1 diabetes and `r substr(table.agegr.all[7, 3], 1, 4)` for Type 2 diabetes are similar to those reported in UK Biobank and in the OpenSAFELY studies.
 We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
@@
 A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather many associated diagnoses.
-Past hospital admission for influenza or respiratory tract infection, though not usually designated as a high-risk condition, is an important and easily-scored risk factor for severe COVID-19.
+We note that past hospital admission for influenza or respiratory tract infection, not currently designated as an indicator of high risk is a strong and easily-scored risk factor for severe COVID-19.
+Others have reported that previous episodes of severe sepsis are associated with increased risk of subsequent sepsis episodes [@prescott_hospitalization_2015; @shen_risk_2016]
 
@@
 However we found that almost half of cases under 40 years had at least one of the listed conditions.
-Among those who did not have a listed condition, the proportions who had had at least one hospital admission in the past five years and had encashed at least one prescription in the past year where much higher among cases than in controls.
-Thus even among younger persons, most cases of severe disease are attributable to pre-existing health conditions.
+Among the remainder who did not have a listed condition, the proportions who had had at least one hospital admission in the past five years and had encashed at least one prescription in the past year were much higher among cases than in controls.
+Thus even among younger persons the vast majority developing  severe disease have  some pre-existing health condition.
 
-## Possible drug effects
 A striking finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
-Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19 .
+Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
 Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.
 Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, causal effects of drug classes cannot be ruled out.
-These association, and possible explanations, are explored in an accompanying paper.
+These associations are explored in an accompanying paper.
 
@@
 As lockdown restrictions are eased, there is general agreement that vulnerable individuals will require shielding, even if the restart of the epidemic can be slowed or suppressed by mass testing, contact tracing and isolation of those who test positive.
-The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on accurate risk discrimination.
+The "stratify and shield" policy option [@mckeigue_evaluation_2020], in which high-risk individuals comprising up to 15% of the population are shielded for a defined period while the epidemic is allowed to run relatively quickly in low-risk individuals until population-level immunity is attained, depends critically on informative risk discrimination.
 So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.
 
-It is to be expected that as awareness grows of how risk varies between individuals, individuals will seek to learn whatever they can about their own level of risk.
-A key implication of our results is that risk of severe or fatal disease is multifactorial and that the ninefold rate ratio associated with a 20-year difference in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological conditions that impair mobility.
+As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.
+A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 2)` associated with a 20-year increase in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological conditions that impair mobility.
 A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one-third of those aged 60-74 year olds in the population have at least one of the listed conditions we examined.
@@
 
-This case-control study is limited to test-positive cases, excluding deaths where no testing was carried out but COVID-19 was mentioned on the death certificate as an underlying or contributing cause.
-Up to 13 May 2020 an additional 1200 such deaths had been reported by the [National Registers of Scotland.](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland) Future versions of this analysis will include sensitivity analyses of the extent to which the results are changed by including these deaths, but the data are not yet available.
+This case-control study is limited to test-positive cases, excluding deaths with no record of a positive test where COVID-19 was mentioned on the death certificate as an underlying or contributing cause.
+Up to 13 May 2020 an additional 1200 such deaths had been reported by the [National Records of Scotland.](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland) Future versions of this analysis will include sensitivity analyses of the extent to which the results are changed by including these deaths, but the data are not yet available.
 Apart from residential care home status, we do not expect most other risk factors to differ markedly between those who died from COVID-19 without being tested and those who died after testing.
@@
 However the associations with pre-existing disease are not just with a small set of conditions that contribute to risk, but with many conditions as demonstrated by associations with past medical and prescribing history in relation to multiple physiological systems.
-As countries attempt to emerge from lockdown whist protecting the most vulnerable multivariate classifiers rather than crude rule-based approaches will be needed to define those most at risk.
+As countries attempt to emerge from lockdown whist protecting vulnerable individuals, multivariate classifiers rather than crude rule-based approaches will be needed to define those most at risk of developing severe disease.
+
+# Acknowledgements
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland and the Public Health Scotland Terminology Services.
+
+## Public Health Scotland COVID-19 Health Protection Study Group
+Alice Whettlock$^1$, Alison Smith-Palmer$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, Dave Yirrell$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Jim McMenamim$^1$, Joseph Jasperse$^1$,  Josie Murray$^1$,  Karen Heatlie$^1$,  Lindsay Mathie$^1$,  Lorraine Donaldson$^1$, Martin Paton$^1$,  Martin Reid$^1$,  Melissa Llano$^1$,  Michael Lockhart$^1$,  Michelle Murphy-Hall $^1$,  Paul Smith$^1$,  Ros Hall$^1$,  Ross Cameron$^1$,  Susan Brownlie$^1$,  Adam Gaffney$^2$,  Aynsley Milne$^2$,  Christopher Sullivan$^2$,  Edward McArdle$^2$,  Elaine Glass$^2$,  Johanna Young$^2$,  William Malcolm$^2$,  Kate Templeton$^3$,  Nazir Lone$^4$,  Rory Gunson$^4$
+
+1  Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+
+2  NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+
+3  NHS Lothian, Royal Infirmary of Edinburgh, 51 Little France Crescent, Old Dalkeith Road, Edinburgh EH16 4SA.
+
+4  Usher Institute, University of Edinburgh, Teviot Place, Edinburgh, EH8 9AG.
+
+# Conflicts of interest
+The authors declare no conflicts of interest.
+
+# Supplementary material
+The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
 
@@
 
-plotWdists(full.densities) + theme_grey(base_size = 18) + 
+plotWdists(densities=full.densities, distlabels=c("Crude", "Adjusted")) +
+                                         theme_grey(base_size = 18) +
                                          theme(legend.title=element_blank())
@@
 
-```{r plotroc, echo=FALSE, fig.cap="\\label{fig:roc}Cross-validation of model chosen by stepwise regression using extended variable set: receiver operating characteristic curve", out.width="0.8\\textwidth"}
+## Footnote for Figure \ref{fig:wevid} 
+For each individuals, the risk prediction model outputs the posterior probability of being a case, which can also be expressed as the posterior odds.
+Dividing the posterior odds by the prior odds gives the likelihood ratio favouring case over non-case status for an individual.
+The weight of evidence $W$ is the logarithm of this ratio.
+The sampling distributions of $W$ in cases and controls are plotted in Figure \ref{fig:wevid}.
+The further apart these curves are, the better the predictive performance.
+The expected information for discrimination $\Lambda$ is the average of the mean of the distribution of $W$ in cases and minus 1 times the mean of the distribution of $W$ in controls.
+The distributions have been adjusted by taking a weighted average to make them mathematically consistent [@mckeigue_quantifying_2019].
+
+\newpage
+
+```{r plotroc, echo=FALSE, fig.pos='H', fig.cap="\\label{fig:roc}Cross-validation of model chosen by stepwise regression using extended variable set: receiver operating characteristic curve", out.width="0.8\\textwidth"}
 
@@
 ```
+
+## Footnote for Figure \ref{fig:roc}
+The crude receiver operator characteristic curve (ROC) is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and noncases.
+The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the score evaluated over all possible such pairs in the dataset.
+When the ROC curve is computed from the adjusted distributions of the weight of evidence W, the curve is always concave downwards, as required for mathematical consistency.
 
@@
     group_rows("Ethnicity", 1, 3) %>%
-    group_rows("SIMD quintile", 4, 8)
+    group_rows("SIMD quintile", 4, 8) %>%
+    kable_styling(latex_options="hold_position")
 
@@
                        "60-74 years"=2, "75+ years"=2)) %>%
-    landscape()
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
 
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
 
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape()
+    landscape() %>%
+    kable_styling(latex_options="hold_position") 
 ```
@@
     column_spec(column=1, width="2.5cm") %>%
-    column_spec(column=2:7, width="1.5cm")
+    column_spec(column=2:7, width="1.5cm") %>%
+    kable_styling(latex_options="hold_position")
 
@@
              caption="Associations of severe disease with ethnicity classified in hospital records") %>%
-    column_spec(column=1, width="3cm")
+    column_spec(column=1, width="3cm") %>%
+    kable_styling(latex_options="hold_position") 
 }
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
+             caption="Univariate associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
     column_spec(column=1, width="13cm") %>%
@@
     column_spec(column=5, width="1.5cm") %>%
-    landscape()
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
 
@@
              caption="Stepwise regression: variables retained in model for severe disease") %>%
-    group_rows("Diabetes", 6, 8)
+    group_rows("SIMD - quintile 1 as reference", 2, 5) %>%
+    group_rows("Diabetes - non-diabetic as reference", 6, 8) %>%
+    kable_styling(latex_options="hold_position")
 

### 2020-05-22 pharmaco.Rmd
fixed to keep memory usage down, also to add obsolete SMR ethnicity codes

@@
 author: 
-  - name: Paul M McKeigue
+  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
     email: paul.mckeigue@ed.ac.uk
-    affiliation: Usher
+    affiliation: HPS
   - name: Amanda Weir 
@@
     affiliation: GCU
-  - name: Helen M Colhoun
-    affiliation: IGMM
+  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
+    affiliation: HPS
+  - name: Public Health Scotland COVID-19 Health Protection Study Group
+    affiliation: HPS
 address: 
@@
     keep_tex: true
-#    toc: true
-#    toc_depth: 2
 csl: ../plos.csl
@@
 ```{r functions, echo=FALSE}
+library(knitr)
+library(kableExtra)
 
@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
-
-library(knitr)
-library(kableExtra)
 
@@
 The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
-All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
+All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
 Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01.
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland.
-Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
+Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
 Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
@@
 In a multivariate analysis the strongest associations were with loop diuretics and calcium-channel blockers.
-All commonly-used drug classes used to treat hypertension were associated with moderately elevated risk of severed disease, with rate ratios from ranging from about 1.4 to 2.2.
+All commonly-used drug classes used to treat hypertension were associated with moderately elevated risk of severe disease, with rate ratios from ranging from about 1.4 to 2.2.
 The rate ratios associated with ACE inhibitors and angiotensin-II receptor blockers were in the middle of this range.
@@
 Increased risk of community-acquired pneumonia is an established adverse effect of the use of these drugs, We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.
+
+# Acknowledgements
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Registers of Scotland and the Public Health Scotland Terminology Services.
+
+## Public Health Scotland COVID-19 Health Protection Study Group
+AliceÂ Whettlock$^1$, Alison Smith-Palmer$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, Dave Yirrell$^1$, David Henderson$^1$, DiogoÂ Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, JimÂ McMenamim$^1$, Joseph Jasperse$^1$,  Josie Murray$^1$,  KarenÂ Heatlie$^1$,  Lindsay Mathie$^1$,  Lorraine Donaldson$^1$, Martin Paton$^1$,  Martin Reid$^1$,  Melissa Llano$^1$,  Michael Lockhart$^1$,  Michelle Murphy-HallÂ $^1$,  Paul Smith$^1$,  Ros Hall$^1$,  Ross Cameron$^1$,  Susan Brownlie$^1$,  Adam Gaffney$^2$,  Aynsley Milne$^2$,  Christopher Sullivan$^2$,  Edward McArdle$^2$,  Elaine Glass$^2$,  Johanna Young$^2$,  William Malcolm$^2$,  Kate Templeton$^3$,  Nazir Lone$^4$,  Rory Gunson$^4$
+
+1  Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+
+2  NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+
+3  NHS Lothian, Royal Infirmary of Edinburgh, 51 Little France Crescent, Old Dalkeith Road, Edinburgh EH16 4SA.
+
+4  Usher Institute, University of Edinburgh, Teviot Place, Edinburgh, EH8 9AG.
+
+## Conflicts of interest
+HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
+HC receives or has recently received non-binding research support from Pfizer, AstraZeneca and Novo-Nordisk.
+
+## Supplementary material
+The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
 

### 2020-05-22 casecontrol.R

@@ -1,4 +1,5 @@
 ## analysis script for case-control study
+rm(list=ls())
 
 library(car)
 library(survival)
@@ -17,16 +18,46 @@ library(gam)
 registerDoParallel(cores=2)
 
 source("helperfunctions.R")
-        
-#cc.all <- readRDS("./data/CC_linked_ANON_20200501.rds")
-cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
 
-diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 records ? excluded
-procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
+
 scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
 
-## we have only 7 digits on scrips, giving resolution to subpara level only
-scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
+old <- FALSE
+if(old) {
+    cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
+    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 records ? excluded
+    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
+    scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
+} else {
+    cc.all <- readRDS("./data/CC_linked_ANON_20200515.rds")
+    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
+    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds") 
+    scrips <- readRDS("./data/CC_PIS_x15_ANON_20200515.rds")[, c("ANON_ID",
+                                                           "bnf_paragraph_code",
+                                                           "bnf_paragraph_description")] 
+    ## other files for diagnoses, procedures and scrips within the time limit
+    ## data/CC_SMR01_ICD10_25_ANON_20200515.rds
+    ## data/CC_SMR01_OPCS4_MAIN.25_ANON_20200501.rds
+    ## data/CC_PIS_15_ANON_20200515.rds
+}
+
+###################################################################
+## scrips should be 2 tables to save space
+## one record per scrip
+scripvars <- c("ANON_ID", "dispensed_date", "bnf_paragraph_code",
[…]

### 2020-05-22 helperfunctions.R

@@ -169,8 +169,8 @@ multivariate.clogit <- function(varnames, outcome="CASE", data, add.reflevel=FAL
     multivariate.formula <- as.formula(paste(outcome, "~",
                                   paste(varnames, collapse=" + "),
                                   "+ strata(stratum)"))
-    data.nonmissing <- nonmissing.obs(data, varnames)
-    multivariate.model <- clogit(formula=multivariate.formula, data=data.nonmissing)
+    nonmissing <- nonmissing.obs(data, varnames)
+    multivariate.model <- clogit(formula=multivariate.formula, data=data[nonmissing, ])
     multivariate.coeffs <- summary(multivariate.model)$coefficients
     multivariate.table <- NULL
    
@@ -180,33 +180,33 @@ multivariate.clogit <- function(varnames, outcome="CASE", data, add.reflevel=FAL
         ## is length(levels) -1 for factor variables
         numrows.i <- 1
         ## set number of rows in multivariate.coeffs for factor variable as num levels -1 
-        if(with(data.nonmissing, is.factor(eval(str2expression(varnames[i]))))) {
+        if(with(data[nonmissing, ], is.factor(eval(str2expression(varnames[i]))))) {
             numrows.i <-
-                with(data.nonmissing, length(levels(eval(str2expression(varnames[i]))))) - 1
+                with(data[nonmissing, ], length(levels(eval(str2expression(varnames[i]))))) - 1
         }
         ## if variable is factor, > 2 levels and add.reflevel
-        if(with(data.nonmissing, is.factor(eval(str2expression(varnames[i])))) &
-           with(data.nonmissing, length(levels(eval(str2expression(varnames[i]))))) > 2 &
+        if(with(data[nonmissing, ], is.factor(eval(str2expression(varnames[i])))) &
+           with(data[nonmissing, ], length(levels(eval(str2expression(varnames[i]))))) > 2 &
            add.reflevel) {
             ## add empty line to multivariate.table
             multivariate.table <- rbind(multivariate.table,
                                         rep(NA, ncol(multivariate.coeffs)))
             ## label this empty line as reference level of factor
             rownames(multivariate.table)[nrow(multivariate.table)] <-
-                 with(data.nonmissing, levels(eval(str2expression(varnames[i])))[1])
+                 with(data[nonmissing, ], levels(eval(str2expression(varnames[i])))[1])
         }
         ## label rows of multivariate.coeffs that will be added
[…]

### 2020-05-22 stepwise.R

@@ -53,12 +53,12 @@ full.formula <- as.formula(paste0("CASE ~ ",
                             paste(full.varnames, collapse="+"), 
                             " + strata(stratum)"))
 
-cc.nonmissing <- nonmissing.obs(cc.severe, full.varnames)
+nonmissing <- nonmissing.obs(cc.severe, full.varnames)
 
 ## fit models
-demog.model <- clogit(formula=demog.formula, data=cc.nonmissing)
-listed.model <- clogit(formula=listed.formula, data=cc.nonmissing)
-full.model <- clogit(formula=full.formula, data=cc.nonmissing)
+demog.model <- clogit(formula=demog.formula, data=cc.severe[nonmissing, ])
+listed.model <- clogit(formula=listed.formula, data=cc.severe[nonmissing, ])
+full.model <- clogit(formula=full.formula, data=cc.severe[nonmissing, ])
 
 if(stepwise) {
     ## stepwise for full variable set
@@ -71,13 +71,11 @@ if(stepwise) {
 
 ################ cross-validation of stepwise regression ######
    
-    cc.nonmissing <- nonmissing.obs(cc.severe, full.varnames)
-
-    test.folds <- testfolds.bystratum(stratum=cc.nonmissing$stratum,
-                                      y=cc.nonmissing$CASE, nfold=nfold)
+    test.folds <- testfolds.bystratum(stratum=cc.severe[nonmissing, ]$stratum,
+                                      y=cc.severe[nonmissing, ]$CASE, nfold=nfold)
     ## test.folds has one row per stratum, one column per fold containing indicator vars
     ## merge by stratum adds single column test.fold
-    cv.data <- merge(cc.nonmissing, test.folds, by="stratum")
+    cv.data <- merge(cc.severe[nonmissing, ], test.folds, by="stratum")
     cv.data <- cv.data[, c("test.fold", "CASE", "stratum", full.varnames)]
     cat("cv.data has", nrow(cv.data), "rows\n")
 

### 2020-05-23 casecontrol.Rmd
fixed coding of severity groups with icu and adm28

@@
 ---
-title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors in COVID-19 in Scotland (REACT-SCOT): a case-control study based on the Scottish population '
+title: 'Rapid Epidemiological Analysis of Comorbidities and Treatments as risk factors for COVID-19 in Scotland (REACT-SCOT): a population-based case-control study'
 author: 
@@
 
-## Information governance
-This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.
-
 # Results
@@
 
-## Association with demographic factors
+## Risk factors
 Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.
@@
 Higher risk of severe disease was also associated with socioeconomic deprivation.
-In comparison with those classified as White, South Asians were at lower risk of severe disease.
-<!---Table \ref{tab:ethnicsmr} shows a separate analysis restricted to those with ethnicity recorded in hospital records, which allows those who identify as Black to be distinguished from those who identify as White.
-The numbers of cases and controls classified as Black are too small for the risk in this group to be estimated reliably.--->
-
-## Associations with listed conditions
-Overall `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.
+Associations with ethnicity are shown for the full dataset based on name classification.
+and separately for the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record.
+Over all age groups, `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.
 The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
@@
 
-Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 age bands combined.
+Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 year age bands combined.
 All listed conditions were associated with severe disease in each age band.
@@
 The multivariate analyses shown in Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that in each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
+Among those aged under 40 years, `r table.scripordiag[2, 2]` of the cases and `r table.scripordiag[2, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
+Among those aged under 60 years, `r table.scripordiag.fatal[2, 2]` of the fatal cases and `r table.scripordiag.fatal[2, 1]` of their matched controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
 
@@
 In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (Mental) and 10 (musculoskeletal).
-Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for whcih the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.
+The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (mental) and 10 (musculoskeletal).
+Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for which the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.
 This was restricted to those without any listed conditions.
 This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.
-We note that past hospital diagnoses of infections, pneumonia and acute respiratory diseases are strongly associated with severe COVID-19.
-Associations with cardiovascular disease are not limited to heart disease but are also present for stroke and other circulatory disorders that are not designated as risk conditions.
+Past hospital diagnoses of infections, pneumonia and acute respiratory diseases were strongly associated with severe COVID-19.
+Cardiovascular diagnoses associated with COVID-19 were not limited to heart disease but included also stroke and other circulatory disorders that are not designated as risk conditions.
 
@@
 This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.
-In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
+<!---In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
+Comment - not relevant to South Asians --->
 
@@
 However we found that almost half of cases under 40 years had at least one of the listed conditions.
-Among the remainder who did not have a listed condition, the proportions who had had at least one hospital admission in the past five years and had encashed at least one prescription in the past year were much higher among cases than in controls.
+Among the remainder who did not have a listed condition, the proportions with at least one hospital admission in the past five years and at least one dispensed prescription in the past year were much higher among cases than in controls.
 Thus even among younger persons the vast majority developing severe disease have some pre-existing health condition.
@@
 
-# Acknowledgements
+# Declarations
+## Information governance
+This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.
+
+## Conflicts of interest
+The authors declare no conflicts of interest.
+
+## Acknowledgements
 We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland and the Public Health Scotland Terminology Services.
@@
 
-# Conflicts of interest
-The authors declare no conflicts of interest.
-
-# Supplementary material
+## Supplementary material
 The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
@@
 plotroc(full.densities) +
-    scale_color_manual(values=c("orange", "purple")) + 
+    scale_color_manual(values=c("orange", "purple"), labels=c("Crude", "Adjusted")) + 
     coord_fixed(ratio=1) +
@@
 
-knitr::kable(table.demog.aug[, 1:4],
+table.demog.append <- table.ethnicsmr.aug
+colnames(table.demog.append) <- colnames(table.demog.aug)[1:4]
+table.demog.append[, 4] <- as.character(table.demog.append[, 4])
+table.demog.aug[, 4] <- as.character(table.demog.aug[, 4])
+
+table.demog.append <- data.frame(varname=c(rownames(table.demog.aug),
+                                           rownames(table.ethnicsmr.aug)),
+                                 rbind(table.demog.aug[, 1:4], table.demog.append))
+
+knitr::kable(table.demog.append[, 1:5],
              escape=FALSE, 
@@
              label="demog",
-             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
+             row.names=FALSE,
+             col.names=c("", clean.header(colnames(table.demog.aug)[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
+             align="lcccc", 
              caption="Univariate associations of severe disease with demographic factors") %>%
@@
     column_spec(column=4, width="2.5cm") %>%
-    group_rows("Ethnicity", 1, 3) %>%
+    group_rows("Ethnicity based on name classification", 1, 3) %>%
     group_rows("SIMD quintile", 4, 8) %>%
+    group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13,
+               hline_before=TRUE, latex_gap_space="10pt") %>% 
     kable_styling(latex_options="hold_position")
@@
 ```
+
+```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
+if(FALSE) {
+knitr::kable(table.ethnicsmr.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmr",
+             digits=c(0, 0, 0, 200), 
+             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with ethnicity classified in hospital records") %>%
+    column_spec(column=1, width="3cm") %>%
+    kable_styling(latex_options="hold_position") 
+}
+```
+\newpage
 
@@
 
-# Appendix
-## Supplementary tables
-
-```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
-
-options(knitr.kable.NA = '')
-if(FALSE) {
-knitr::kable(table.ethnicsmr.aug,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="ethnicsmr",
-             digits=c(0, 0, 0, 200), 
-             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with ethnicity classified in hospital records") %>%
-    column_spec(column=1, width="3cm") %>%
-    kable_styling(latex_options="hold_position") 
-}
-```
+# Supplementary tables
 

### 2020-05-23 pharmaco.Rmd
fixed coding of severity groups with icu and adm28

@@
     email: paul.mckeigue@ed.ac.uk
+    affiliation: HPS
+  - name: Sharon Kennedy
     affiliation: HPS
@@
     affiliation: IGMM
-  - name: Sharon Kennedy
-    affiliation: HPS
   - name: David McAllister 
@@
 The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
-All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.
+All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained for this study.
 Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01.
@@
 ## Morbidity and drug prescribing
-For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
+For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
-British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
+British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding those dispensed less than 25 days before testing positive for SARS-CoV-2.
 The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
@@
 Most of these codes are for dressings and appliances rather than drugs.
+Defined daily doses (DDDs) were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.
+Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD)  the time in
 
@@
 Rate ratios were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
-
-## Information governance
-This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
-Datasets were de-identified before analysis.
 
@@
 The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.
+Table \ref{tab:dose.protonpump} shows how the effect size varies with the dose of proton pump inibitor, calculated as average defined daily doses dispensed per day since 1 June 2019.
+There is a strong dose-response relationship in the two younger age groups, but not in those aged over 75 years.
 
@@
 
+# Declarations
+## Information governance
+This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
+Datasets were de-identified before analysis.
+
+## Conflicts of interest
+HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
+HC receives or has recently received non-binding research support from Pfizer, AstraZeneca and Novo-Nordisk.
+
 # Acknowledgements
@@
 
-## Conflicts of interest
-HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
-HC receives or has recently received non-binding research support from Pfizer, AstraZeneca and Novo-Nordisk.
-
 ## Supplementary material
@@
 ```{r cats3, echo=FALSE, warning=FALSE, message=FALSE}
+options(knitr.kable.NA = '')
 
@@
 
+```{r dose.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.dose.protonpump,
+             escape=FALSE,
+             label="dose.protonpump", 
+             booktabs=TRUE,
+             row.names=FALSE,
+             col.names=c("Average DDDs/day", "Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe disease with average dose of proton pump inhibitor (daily defined doses/day) dispensed since 1 June 2019, by age group") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    group_rows("Ages 0-59 years", 1, 5) %>%
+    group_rows("Ages 60-74 years", 6, 10) %>%
+    group_rows("Ages 75+ years", 11, 15) %>%
+    landscape()
+
+```
+
 ```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}

### 2020-05-23 casecontrol.R

@@ -14,15 +14,18 @@ library(readxl)
 library(DescTools)
 library(icd.data)
 library(gam)
+library(plyr) # ? needed to access .() function
+library(dplyr)
 
 registerDoParallel(cores=2)
 
 source("helperfunctions.R")
 
-
 scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
 
+old <- TRUE
 old <- FALSE
+
 if(old) {
     cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
     diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 records ? excluded
@@ -35,36 +38,43 @@ if(old) {
     scrips <- readRDS("./data/CC_PIS_x15_ANON_20200515.rds")[, c("ANON_ID",
                                                            "bnf_paragraph_code",
                                                            "bnf_paragraph_description")] 
+    ## scrips should be 2 tables to save space
+    ## one record per scrip
+    scripvars <- c("ANON_ID", "dispensed_date", "bnf_paragraph_code",
+                   "formulation_code",
+                   "item_strength", "item_strength_uom", "item_code",
+                   "num_items", "quantity")
+    ## lookup table for subpara code and item code
+    drugvars <- c("bnf_paragraph_code", "bnf_paragraph_description",
+                  "item_code", "approved_name")
     ## other files for diagnoses, procedures and scrips within the time limit
     ## data/CC_SMR01_ICD10_25_ANON_20200515.rds
     ## data/CC_SMR01_OPCS4_MAIN.25_ANON_20200501.rds
     ## data/CC_PIS_15_ANON_20200515.rds
+
+    scrips.protonpump <-
+        readRDS("./data/CC_PIS_x15_ANON_20200515.rds") %>%
+        subset(bnf_paragraph_code=="0103050")
 }
 
 ###################################################################
-## scrips should be 2 tables to save space
-## one record per scrip
-scripvars <- c("ANON_ID", "dispensed_date", "bnf_paragraph_code",
-               "formulation_code",
-               "item_strength", "item_strength_uom", "item_code",
[…]

### 2020-05-23 ethnic_assign.R

@@ -49,14 +49,14 @@
 #'98'='Refused/Not provided by patient';
 #'99'='Not Known'")
 
-## extra codes: 1[DEFGHJ] White, 3[ABC] South Asian, 3[E] Chinese, 3D Other Asian, 4[ABCE] presumably African
+## extra codes: 1[DEFGHJ] White, 3[ABC] South Asian, 3[E] Chinese, 3D Other Asian, 4[ABCE]  African
 
 ethnic5.smr <- character(length(ethnic.smr))
 ethnic5.smr[grep("^1[A-Z]$", ethnic.smr)] <- "White"
 ethnic5.smr[grep("^2[A-Z]$", ethnic.smr)] <- "Other"
 ethnic5.smr[grep("^3[ABCFGH]$", ethnic.smr)] <- "South Asian"
 ethnic5.smr[grep("^3[EJ]$", ethnic.smr)] <- "Chinese"
-ethnic5.smr[grep("^3D$", ethnic.smr)] <- "Other"
+ethnic5.smr[grep("^3[DZ]$", ethnic.smr)] <- "Other"
 ethnic5.smr[grep("^4[ABCDEY]$", ethnic.smr)] <- "Black"
 ethnic5.smr[grep("^5[ABCDY]$", ethnic.smr)] <- "Black"
 ethnic5.smr[grep("^6[AZ]$", ethnic.smr)] <- "Other"
@@ -66,7 +66,7 @@ ethnic5.smr <- as.factor(ethnic5.smr)
 ethnic5.smr <- factor(ethnic5.smr, levels=levels(ethnic5.smr)[c(5, 4, 2, 1, 3)])
 
 if(length(OnolyticsType) > 0) {
-    OnolyticsType <- recode(OnolyticsType,
+    OnolyticsType <- car::recode(OnolyticsType,
                             "'NOT FOUND'=NA; 'INTERNATIONAL'=NA; 'UNCLASSIFIED'=NA; 'VOID'=NA; 'VOID - FORENAME'=NA; 'VOID INITIAL'=NA")
     
     table(OnolyticsType[GeographicalArea=="SOUTH ASIA"])
@@ -102,6 +102,6 @@ if(length(OnolyticsType) > 0) {
     eth[OnolyticsType == "BLACK CARIBBEAN"] <- "Black Caribbean"
     
     ## reduce to 5 categories: White, South Asian, Chinese, Black, Other
[…]

### 2020-05-25 casecontrol.Rmd
final manuscript of paper 1

@@
 
+table.demog.append <- table.ethnicsmr.aug
+colnames(table.demog.append) <- colnames(table.demog.aug)[1:4]
+table.demog.append[, 4] <- as.character(table.demog.append[, 4])
+table.demog.aug[, 4] <- as.character(table.demog.aug[, 4])
+
+table.demog.append <- data.frame(varname=c(rownames(table.demog.aug),
+                                           rownames(table.ethnicsmr.aug)),
+                                 rbind(table.demog.aug[, 1:4], table.demog.append))
+
 table.densities <- rbind(summary(demog.densities),
@@
 Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database.
-Those who had been admitted to an intensive care unit or had died within 28 days of a positive test were classified as severe cases.
+Those who had been admitted to a critical care unit or had died within 28 days of a positive test were classified as severe cases.
 For each case, ten controls matched for sex, age and primary care practice were selected from the population register.
@@
 In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
-Univariate rate ratios (95% CIs) for eight conditions listed by public health agencies as conferring high risk were `r table.agegr.all[7, 3]` for Type 2 diabetes,
+Univariate rate ratios (95% CIs) for eight conditions listed by public health agencies as conferring high risk were `r table.agegr.all[6, 3]` for Type 2 diabetes,
 `r table.agegr.all[8, 3]` for ischemic heart disease,
@@
 
-`r freqs.all[16, 2]` of the severe cases and `r freqs.all[16, 1]` of the controls had at least one of these listed conditions.
-In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[16, 2])` in cases and `r gsub("\\\\", "", table.agegr[16, 1])` in controls.
+`r freqs.all[nrow(freqs.all), 2]` of the severe cases and `r freqs.all[nrow(freqs.all), 1]` of the controls had at least one of these listed conditions.
+In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[4, 2])` in cases and `r gsub("\\\\", "", table.agegr[4, 1])` in controls.
 The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
@@
 
-Severe or fatal COVID-19 was defined by a record of entering ICU in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-By restricting the case definition to cases that were fatal or received ICU care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without ICU care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from ICU care.
+Severe or fatal COVID-19 was defined by a record of entering critical care (intensive care, high dependencey unit or combined unit) in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
+By restricting the case definition to cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from critical care.
 
@@
 Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.
-Cross-tabulation of a subset of records for which ethnicity assigned by patients or health care staff had been recorded in the Scottish Morbidity Record showed that this algorithm correctly classified 90% of those who reported South Asian ethnicity (1.6% of the  Scottish population in 2011) but misclassified most of those who identified as Caribbean or Black (0.1% of the population in 2011) as White.
-<!--- For cases and controls who had a past record of hospital admission or outpatient attendance in the Scottish Morbidity Record, a separate classification of ethnicity was available, assigned by self-report and based on the categories used in the Census.
---->
+For 54% of cases and 28% of controls self-assigned ethnicity, based on the categories used in the Census, had been recorded in Scottish Morbidity Records (SMR).
+Cross-tabulation of `r sum.xtabulate` records for which both name classification and SMR records of ethnicity were available showed that the ONOMAP algorithm had sensitivity of `r round(SouthAsian.sensitivity)`% and specificity of `r round(SouthAsian.specificity, 2)`% for classifying South Asian ethnicity, but misclassified most of those who identified as African, Caribbean or Black.
 
@@
 We began by scoring a specific list of eight conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-A separate list of conditions designates "clinically extremely vulnerable" who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
+A separate list of conditions designates "clinically extremely vulnerable" individuals who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
 We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.
@@
 To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.
-To assess whether allowing for a non-linear relationship of the logit of risk to age would improve the fit, we also fitted generalized additive models, implemented in the R function `gam::gam`, with default smoothing function.
+To allow for possible non-linearity of the relationship of the logit of risk to age, we also fitted generalized additive models, implemented in the R function `gam::gam`, with default smoothing function.
 
@@
 With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.
-These odds ratios based on the conditional likelihood will differ from the unconditional odds ratios calculable from the frequencies of exposure in cases and controls [@breslow_estimation_1978]
-Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios.
+Odds ratios in a matched case control study are based on the conditional likelihood and  the unconditional odds ratios calculable from the frequencies of exposure in cases and controls will differ from these and should not be used [@breslow_estimation_1978].
+Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios but with less precision.[@breslow_estimation_1978]
 
@@
 The use of bits (logarithms to base 2) to quantify information is standard in information theory: one bit can be defined as the quantity of information that halves the hypothesis space.
-Although readers may be unfamiliar with the expected information for discrimination, it has several properties that make it more useful than the C-statistic for quantifying the performance of a risk prediction model.
-[@mckeigue_quantifying_2019] .
+Although readers may be unfamiliar with the expected information for discrimination $\Lambda$, it has several properties that make it more useful than the C-statistic for quantifying increments in the performance of a risk prediction model [@mckeigue_quantifying_2019] .
 A key advantage of using $\Lambda$ is that contributions of independent predictors can be added.
@@
 ## Risk factors
+### Sociodemographic factors
 Table \ref{tab:demog} shows univariate associations of demographic factors with severe disease.
@@
 Higher risk of severe disease was also associated with socioeconomic deprivation.
-Associations with ethnicity are shown for the full dataset based on name classification.
-and separately for the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record.
-Over all age groups, `r freqs.all[13, 2]` of severe cases and `r freqs.all[13, 1]` of controls had at least one of the listed conditions.
-The prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
-`r gsub("\\\\", "", table.agegr[16, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[16, 1])` of the controls.
-In those aged 75+ years `r gsub("\\\\", "", table.agegr[16, 8])` of the cases and `r gsub("\\\\", "", table.agegr[16, 7])` of the controls had at least one listed condition.
+Associations with ethnicity are shown for the full dataset based on name classification and separately for the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record.
+With Whites as reference category, the rate ratio (95% CI) associated with South Asian ethnicity was `r table.demog.append[2, 4]` based on name classification and `r table.demog.append[11, 4]`, based on the SMR records.
+The numbers of cases in other non-White ethnic groups were too sparse to tabulate separately.
+
+### Factors derived from hospitalisation and prescribing records
+Prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
+`r gsub("\\\\", "", table.agegr[nrow(table.agegr), 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[nrow(table.agegr), 1])` of the controls.
+In those aged 75+ years `r gsub("\\\\", "", table.agegr[nrow(table.agegr), 8])` of the cases and `r gsub("\\\\", "", table.agegr[nrow(table.agegr), 7])` of the controls had at least one listed condition.
+There was a much higher prevalence of at least one prescription in cases in controls even in those aged under under 40 years.
+Among those aged under 40 years, `r table.scripordiag[2, 2]` of the cases and `r table.scripordiag[2, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
+
+Over all age groups, `r freqs.all[nrow(freqs.all), 2]` of severe cases and `r freqs.all[nrow(freqs.all), 1]` of controls had at least one of the listed conditions.
+As shown in Table \ref{tab:listed.conditions.allages}, all the listed conditions were more frequent in cases than controls.
+The rate ratio associated with type 1 diabetes was higher than that for type 2 diabetes.
+The rate ratio was `r table.agegr.all[8, 3]` for ischemic heart disease compared to `r table.agegr.all[9, 3]` for the broad category "other heart disease".
+In multivariate analysis ischemic heart disease was not independently associated with severity whereas other heart disease remained strongly associated.
 
@@
 In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][6, 3]` for Type 1 diabetes and `r tables.agegr[[1]][7, 3]` for Type 2 diabetes.
-For all age groups combined, the rate ratio was `r tables.agegr[[1]][8, 3]` for ischemic heart disease and `r tables.agegr[[1]][9, 3]` for the broad category "other heart disease".
-The multivariate analyses shown in Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that in each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
-Among those aged under 40 years, `r table.scripordiag[2, 2]` of the cases and `r table.scripordiag[2, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
-Among those aged under 60 years, `r table.scripordiag.fatal[2, 2]` of the fatal cases and `r table.scripordiag.fatal[2, 1]` of their matched controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
+The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and within each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
+Table \ref{tab:fatal} shows that in each age group the proportion of cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.
 
 ## Systematic analysis of diagnoses associated with severe disease
-The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, restricted to those without any of the listed conditions.
+The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, among those without any of the listed conditions.
 These results are shown in Table \ref{tab:conditions}.
 In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
+Note that hospitalisations classified under the pregnancy chapter here derive from admission to non-obstetric units with pregnancy related medical conditions since in Scotland obstetric returns are not in the SMR01 dataset.
 The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (mental) and 10 (musculoskeletal).
-Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters, filtered to show only subchapters for which the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.
-This was restricted to those without any listed conditions.
-This table shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.
+Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.
+This table is filtered to show only subchapters for which the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.
+This shows that many diagnoses are associated with markedly higher risk of severe COVID-19.
 Past hospital diagnoses of infections, pneumonia and acute respiratory diseases were strongly associated with severe COVID-19.
@@
 Observed and predicted case status were compared over all test folds.
-Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.
+Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.
 
@@
 
-Figure \ref{fig:wevid} shows the distribution of the weight of evidence $W$ favouring case over control status from the model based on the extended variable set with a footnote explaining how $\Lambda$ is derived.
+Figure \ref{fig:wevid} shows the distribution of the weight of evidence favouring case over control status from the model based on the extended variable set with a footnote explaining how $\Lambda$ is derived.
 This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
@@
 
-In contrast to reports from England [@the_opensafely_collaborative_opensafely_2020] we found that people of South Asian ethnicity are at lower risk than Whites.
-This suggests that socioeconomic rather than genetic factors underlie at least some of the associations of risk with ethnicity reported in England.
-<!---In the OpenSAFELY cohort, those recorded as Black and Asian people were at 1.7 fold and 1.6 fold risk of death respectively compared with those recorded as White, after adjustment for comorbidities and socioeconomic status, However in a recent report based on the UK Biobank cohort, almost half of the fourfold risk ratio for hospitalised COVID associated with Black ethnicity was accounted for by health care worker status, with other social variables attenuating the risk ratio further [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-Comment - not relevant to South Asians --->
+Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now, and show no evidence that risk is higher in South Asians than in Whites.
+This contrasts with reports from England [@the_opensafely_collaborative_opensafely_2020] of risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, after adjustment for comorbidities and socioeconomic status.
+In a study based on the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated the crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
+In the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.
+Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average.
+As the relative socioeconomic position of these groups in Scotland is very different to that in England, it is to be expected that the relation of health status to ethnicity will also differ [@walsh_increasingly_2019].
+Future work may allow more complete assignment of ethnicity and disaggregation of broad categories based on continent of origin.
 
@@
 However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages.
-The rate ratios of `r substr(table.agegr.all[6, 3], 1, 4)` for Type 1 diabetes and `r substr(table.agegr.all[7, 3], 1, 4)` for Type 2 diabetes are similar to those reported in UK Biobank and in the OpenSAFELY studies.
+The rate ratios of `r substr(table.agegr.all[6, 3], 1, 4)` for Type 1 diabetes and `r substr(table.agegr.all[7, 3], 1, 4)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.
 We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
@@
 However overall 26% of cases did not have any of these listed conditions.
-Among cases and controls without these conditions not surprisingly neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as in the current dataset we cannot separately identify those who are currently receiving chemotherapy or radiotherapy for whom shielding is advised.
+Among cases and controls without these conditions, not surprisingly, neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as in the current dataset we cannot separately identify those who are currently receiving chemotherapy or radiotherapy for whom shielding is advised.
 We have not attempted to estimate the risk associated with these conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding rather than on the risk to those exposed to the epidemic.
-A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease but rather many associated diagnoses.
-We note that past hospital admission for influenza or respiratory tract infection, not currently designated as an indicator of high risk is a strong and easily-scored risk factor for severe COVID-19.
-Others have reported that previous episodes of severe sepsis are associated with increased risk of subsequent sepsis episodes [@prescott_hospitalization_2015; @shen_risk_2016]
+A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease in those without  listed conditions but rather many associated diagnoses.
 
@@
 However we found that almost half of cases under 40 years had at least one of the listed conditions.
-Among the remainder who did not have a listed condition, the proportions with at least one hospital admission in the past five years and at least one dispensed prescription in the past year were much higher among cases than in controls.
-Thus even among younger persons the vast majority developing severe disease have some pre-existing health condition.
+In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.
 
@@
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland and the Public Health Scotland Terminology Services.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the     HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories.
 
 ## Public Health Scotland COVID-19 Health Protection Study Group
-Alice Whettlock$^1$, Alison Smith-Palmer$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, Dave Yirrell$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Jim McMenamim$^1$, Joseph Jasperse$^1$,  Josie Murray$^1$,  Karen Heatlie$^1$,  Lindsay Mathie$^1$,  Lorraine Donaldson$^1$, Martin Paton$^1$,  Martin Reid$^1$,  Melissa Llano$^1$,  Michael Lockhart$^1$,  Michelle Murphy-Hall $^1$,  Paul Smith$^1$,  Ros Hall$^1$,  Ross Cameron$^1$,  Susan Brownlie$^1$,  Adam Gaffney$^2$,  Aynsley Milne$^2$,  Christopher Sullivan$^2$,  Edward McArdle$^2$,  Elaine Glass$^2$,  Johanna Young$^2$,  William Malcolm$^2$,  Kate Templeton$^3$,  Nazir Lone$^4$,  Rory Gunson$^4$
+Alice Whettlock$^1$,  Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Jim McMenamim$^1$, Joseph Jasperse$^1$,  Josie Murray$^1$,  Karen Heatlie$^1$,  Lindsay Mathie$^1$,  Lorraine Donaldson$^1$, Martin Paton$^1$,  Martin Reid$^1$,  Melissa Llano$^1$,    Michelle Murphy-Hall $^1$,  Paul Smith$^1$,  Ros Hall$^1$,  Ross Cameron$^1$,  Susan Brownlie$^1$,  Adam Gaffney$^2$,  Aynsley Milne$^2$,  Christopher Sullivan$^2$,  Edward McArdle$^2$,  Elaine Glass$^2$,  Johanna Young$^2$,  William Malcolm$^2$,  Jodie McCoubrey $^2$
 
@@
 2  NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
-
-3  NHS Lothian, Royal Infirmary of Edinburgh, 51 Little France Crescent, Old Dalkeith Road, Edinburgh EH16 4SA.
-
-4  Usher Institute, University of Edinburgh, Teviot Place, Edinburgh, EH8 9AG.
 
@@
 \newpage
+
+# Figures
 
@@
 ## Footnote for Figure \ref{fig:wevid} 
-For each individuals, the risk prediction model outputs the posterior probability of being a case, which can also be expressed as the posterior odds.
+For each individual, the risk prediction model outputs the posterior probability of being a case, which can also be expressed as the posterior odds.
 Dividing the posterior odds by the prior odds gives the likelihood ratio favouring case over non-case status for an individual.
 The weight of evidence $W$ is the logarithm of this ratio.
-The sampling distributions of $W$ in cases and controls are plotted in Figure \ref{fig:wevid}.
-The further apart these curves are, the better the predictive performance.
+The  distributions of $W$ in cases and controls in the test data are plotted in Figure \ref{fig:wevid}.
+For a classifier, the further apart these curves are, the better the predictive performance.
 The expected information for discrimination $\Lambda$ is the average of the mean of the distribution of $W$ in cases and minus 1 times the mean of the distribution of $W$ in controls.
@@
 ## Footnote for Figure \ref{fig:roc}
-The crude receiver operator characteristic curve (ROC) is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and noncases.
-The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the score evaluated over all possible such pairs in the dataset.
-When the ROC curve is computed from the adjusted distributions of the weight of evidence W, the curve is always concave downwards, as required for mathematical consistency.
+The crude receiver operator characteristic curve (ROC) is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and non-cases.
+The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the score, evaluated over all possible such pairs in the dataset.
 
@@
 
+# Tables
+
 ```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
 options(knitr.kable.NA = '')
-
-table.demog.append <- table.ethnicsmr.aug
-colnames(table.demog.append) <- colnames(table.demog.aug)[1:4]
-table.demog.append[, 4] <- as.character(table.demog.append[, 4])
-table.demog.aug[, 4] <- as.character(table.demog.aug[, 4])
-
-table.demog.append <- data.frame(varname=c(rownames(table.demog.aug),
-                                           rownames(table.ethnicsmr.aug)),
-                                 rbind(table.demog.aug[, 1:4], table.demog.append))
 
@@
 
-```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}
-if(FALSE) {
-knitr::kable(table.ethnicsmr.aug,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="ethnicsmr",
-             digits=c(0, 0, 0, 200), 
-             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with ethnicity classified in hospital records") %>%
-    column_spec(column=1, width="3cm") %>%
-    kable_styling(latex_options="hold_position") 
-}
-```
 \newpage
@@
 rownames(table.agegr) <- replace.names(rownames(table.agegr))
-table.agegr[1, c(1, 3, 5, 7)] <- " "
-knitr::kable(table.agegr,
+#table.agegr[1, c(1, 3, 5, 7)] <- " "
+knitr::kable(table.agegr[-7, ], # omit non-diabetic category
              escape=FALSE, 
@@
              align=rep("c", 8), 
-             caption="Associations of severe disease with risk factors, by age group") %>%
+             caption="Frequencies of risk factors in cases and controls, by age group") %>%
     column_spec(column=1, width="3.5cm") %>%
-    group_rows("Listed condition", 5, nrow(table.agegr) - 1) %>%
+    group_rows(" ", 7, nrow(table.agegr[-7, ])) %>%
     add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
                        "60-74 years"=2, "75+ years"=2)) %>%
+ #   kable_styling(latex_options="scale_down") %>%
     landscape() %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r scripordiag.fatal, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.scripordiag.fatal, 
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="fatal",
+             align=rep("c", 2), 
+             caption="Proportions of fatal cases and controls with no dispensed prescription or hospital diagnosis, by age group") %>%
     kable_styling(latex_options="hold_position")
@@
 
-knitr::kable(table.agegr.all,
+knitr::kable(table.agegr.all[-4, ],
              escape=FALSE, 
@@
              caption="Associations of severe disease with listed conditions over all age groups") %>%
-    group_rows("Listed condition", 4, nrow(table.agegr.all)) %>%
+    group_rows(" ", 4, nrow(table.agegr.all[-4, ])) %>%
     column_spec(column=1, width="4cm") %>%
@@
 
+colnames(table.densities) <- gsub("Model-based", "Adjusted", colnames(table.densities))
+
 knitr::kable(table.densities, 
@@
     column_spec(column=1, width="2.5cm") %>%
-    column_spec(column=2:7, width="1.5cm") %>%
+    column_spec(column=2:7, width="1.4cm") %>%
     kable_styling(latex_options="hold_position")
@@
 
-knitr::kable(tables.agegr[[1]],
+knitr::kable(tables.agegr[[1]][-4, ],
              escape=FALSE, 
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(tables.agegr[[1]])[1:2],
+             col.names=c(clean.header(colnames(tables.agegr[[1]])[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value",
@@
              caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
-    group_rows("Listed condition", 4, nrow(tables.agegr[[1]])) %>%
+    group_rows(" ", 4, nrow(tables.agegr[[1]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
@@
 ```
+
 ```{r listed.condition.60to74, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(tables.agegr[[2]],
+knitr::kable(tables.agegr[[2]][-4, ],
              escape=FALSE, 
@@
              digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(colnames(tables.agegr[[2]])[1:2],
+             col.names=c(clean.header(colnames(tables.agegr[[2]])[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value",
@@
              caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
-    group_rows("Listed condition", 4, nrow(tables.agegr[[2]])) %>%
+    group_rows(" ", 4, nrow(tables.agegr[[2]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
@@
 
-knitr::kable(tables.agegr[[3]],
+knitr::kable(tables.agegr[[3]][-4, ],
              escape=FALSE, 
@@
              caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
-    group_rows("Listed condition", 4, nrow(tables.agegr[[3]])) %>%
+    group_rows(" ", 4, nrow(tables.agegr[[3]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Univariate associations of severe disease with hospital diagnoses by ICD subchapters: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
+             caption="Univariate associations of severe disease with hospital diagnoses by ICD subchapters in those without any listed conditions: rows retained are those with p < 0.001 and at least 50 cases and controls") %>%
     column_spec(column=1, width="13cm") %>%

### 2020-05-25 pharmaco.Rmd
final manuscript of paper 1

@@
 A further backward selection of drug codes in these two chapters identified proton pump inhibitors as the drug class with the strongest independent association with severe COVID-19.
-The rate ratio for severe COVID-19 associated with with proton pump inhibitor use varied from `r table.protonpump[4, 3]` in those aged 75 years and over to `r table.protonpump[1, 3]` in those aged under 40 years.
+The rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[4, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 40 years.
 Adjustment for hospital diagnoses of disorders of the esophagus, stomach and duodenum, and prescribing of antiplatelet or nonsteroidal anti-inflammatory agents reduced these effect sizes only slightly.
@@
 As severe disease is associated with other drug classes such as laxatives that are commonly prescribed for minor symptoms, confounding by factors associated with these symptoms cannot be excluded.
-However as increased risk of community-acquired pneumonia is an established adverse effect of this class of drugs, a causal effect is plausible.
+However as increased risk of infection is an established adverse effect of this class of drugs, a causal effect is plausible.
 In the Scottish population where 28\% of controls have used proton pump inhibitors and the rate ratio is 2.8, one-third of severe cases could be attributable to use of these drugs.
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
-British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding those dispensed less than 25 days before testing positive for SARS-CoV-2.
+
+British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2.
+This ensures that only prescriptions dispensed before the onset of symptoms are included.
 The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
@@
 Defined daily doses (DDDs) were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.
-Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD)  the time in
+Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to 16 days before the date that the case (in a stratum consisting of case and matched controls) first tested.
 
-# Data analysis 
+# Statistical analysis 
 We derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
@@
 The last variable retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.
+This led us to examine further the association of proton pump inhibitors with severe disease, without restricting the analysis to those not resident in care homes and without a diagnosis of a listed condition.
 
-This led us to examine further the association of proton pump inhibitors with severe disease, without restricting the analysis to those not resident in care homes and without a diagnosis of a listed condition.
-Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".
+## Construction of a propensity score
+We began by constructing a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
+To construct this score, we used an independent dataset so that the score would not overfit the case-control dataset.
+As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a dataset of `nrow(cc.hosp)` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
+The propensity score was constructed on this training dataset by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
+The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
+As expected, the variables retained included NSAIDs and antiplatelet drugs, but also many other drug classes such as SSRIs that are not recognized medical indications for coprescribing of a proton pump inhibitor.
+The coefficients for age and sex were dropped before calculating the propensity score in the case-control dataset, because the case-control analysis is matched for age and sex.
+
+<!--- add something about how well the propensity score predicts ever-use of a PPI --->
+
+## Associations with proton pump inhibitor use
+Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for propensity score.
 The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.
-Table \ref{tab:dose.protonpump} shows how the effect size varies with the dose of proton pump inibitor, calculated as average defined daily doses dispensed per day since 1 June 2019.
-There is a strong dose-response relationship in the two younger age groups, but not in those aged over 75 years.
+
+Table \ref{tab:dose.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day since 1 June 2019.
+There is a strong dose-response relationship in the two younger age groups, but not in those
+aged over 75 years.
+
+Table \ref{tab:dosegr.protonpump} shows how the effect size varies with the dose of proton pump inibitor, grouped into categories.
+
+Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.
+time window.
+The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.
+
+Table \ref{tab:ppinames} compares the effects of the five drugs in this class.
+Omeprazole and lansoprazole were by far the most commonly prescribed drugs.
+The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day was higher for lansoprazole than for omeprazole.
+This suggests that the DDDs for lansoprazole and omeprazole (30 mg and 20 mg respectively) are not equivalent in potency.
 
@@
 
- Table \ref{tab:bnfchapter2} shows that We also examined the univariate and multivariate associations of prescribed drugs for the cardiovascular system (BNF chapter 2) with severe COVID-19.
-lists associations with drug classes (BNF subparagraphs) in this chapter.
+We also examined the univariate and multivariate associations of prescribed drugs for the cardiovascular system (BNF chapter 2) with severe COVID-19.
+Table \ref{tab:bnfchapter2} shows associations with drug classes (BNF subparagraphs) in this chapter.
 Prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19.
@@
 
+Tables \ref{tab:bnfchapter10} and \ref{tab:bnfchapter9} show associations with drug classes in BNF chapters 10 and 9.
+
 # Discussion 
 We have shown that the association of dispensed drug prescriptions with severe COVID-19 remains after excluding people who are at high risk for reasons that are associated with higher rates of prescribing; residence in care homes or a hospital diagnosis of a designated high-risk condition.
-We have shown further that this most of this association is accounted for by drugs prescribed for the gastrointestinal system and drugs prescribed for the cardiovascular system.
+We have shown further that this most of this association is accounted for by drugs prescribed for the gastrointestinal and cardiovascular systems.
 There are smaller independent contributions from drugs prescribed for the nervous and musculoskeletal systems.
@@
 
-It is possibly relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors had previously been reported.
-A study in Szechuan, China, reported that proton pump inhibitors were prescribed to 37% of inpatients in a tertiary referral centre in 2016 [@luo_changes_2018].
+It may be relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors has been reported previously.
+A study in China reported that proton pump inhibitors were prescribed to 37% of inpatients in a tertiary referral centre in Szechuan in 2016 [@luo_changes_2018].
 In Italy, a survey of nursing homses found that 46% of residents had been prescribed a proton pump inhibitor [@pasina_overuse_2020].
-A Spanish drug-utilization study reported that 55% of hospital inpatients were presribed a PPI at discharge [@ramirez_overuse_2010].
+A Spanish drug-utilization study reported that 55% of hospital inpatients were presribed a proton pump inhibitor at discharge [@ramirez_overuse_2010].
 
@@
 Unless this this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
-
-A limitation of the data linkage undertaken by Health Protection Scotland is that we cannot  construct longitudinal records of drug exposure with dose.
-However we have been able to do this in a separate study of all individuals with diabetes in Scotland, for whom we have constructed a curated database of drug exposure.
 
@@
 
-```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+```{r propensity, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.protonpump,
+knitr::kable(coeffs.propensity,
              escape=FALSE,
-             label="protonpump.adjusted", 
+             label="propensity", 
              booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
+             digits=c(0, 3), 
+             col.names=c("BNF subparagraph", "log odds ratio"),
+             caption="Propensity score for prescribing of a proton pump inhibitor, constructed in an independent dataset") %>%
+             kable_styling(latex_options="hold_position")
+
+```
+
+```{r everuse.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.dose.protonpump,
+             escape=FALSE,
+             label="everuse.protonpump", 
+             booktabs=TRUE,
+             col.names=c("Controls", "Cases",
                          "Rate ratio (95\\% CI)", "$p$-value",
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with prescription of a proton pump inhibitor, with and without adjustment for covariates: care home residence, hospital diagnosis of upper GI tract disorder, prescriptions of NSAIDs,  antiplatelet agents") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+             caption="Associations of severe disease with ever-use of proton pump inhibitor by age group, with and without adjustment for propensity score") %>%
+    add_header_above(c(" "=2, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
@@
              booktabs=TRUE,
-             row.names=FALSE,
+             col.names=c("Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe disease with average dispensed dose of proton pump inhibitor (DDDs/day) by age group, with and without adjustment for propensity score") %>%
+    add_header_above(c(" "=2, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+
+```
+
+```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.dosegr.protonpump,
+             escape=FALSE,
+             label="dosegr.protonpump", 
+             booktabs=TRUE,
+             row.names=FALSE, # dose levels in column 1
              col.names=c("Average DDDs/day", "Controls", "Cases",
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with average dose of proton pump inhibitor (daily defined doses/day) dispensed since 1 June 2019, by age group") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
+             caption="Associations of severe disease with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for propensity score") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
@@
     group_rows("Ages 75+ years", 11, 15) %>%
+    landscape()
+
+```
+
+```{r timewin.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.timewindow.protonpump,
+             escape=FALSE,
+             label="timewindow.protonpump", 
+             booktabs=TRUE,
+             col.names=c("Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window")) %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+
+```
+
+```{r ppinames, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.ppinames[, 1:4],
+             escape=FALSE,
+             label="ppinames", 
+             booktabs=TRUE,
+             col.names=c("Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Univariate associations of severe disease with dispensed average daily dose (DDDs/day) for each drug in the proton pump inhibitor class") %>%
+    add_header_above(c(" "=1, "Average daily dose (DDDs)"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
     landscape()

### 2020-05-25 casecontrol.R

@@ -14,7 +14,7 @@ library(readxl)
 library(DescTools)
 library(icd.data)
 library(gam)
-library(plyr) # ? needed to access .() function
+# library(plyr) # ? needed to access .() function
 library(dplyr)
 
 registerDoParallel(cores=2)
@@ -24,7 +24,7 @@ source("helperfunctions.R")
 scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
 
 old <- TRUE
-old <- FALSE
+#old <- FALSE
 
 if(old) {
     cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
@@ -136,17 +136,19 @@ cc.all$SIMD.quintile <- car::recode(cc.all$SIMD.quintile, "'Unknown'=NA")
 cc.all$scrip.any <- as.factor(as.integer(cc.all$ANON_ID %in% scrips$ANON_ID))
 cc.all$diag.any <- as.factor(as.integer(cc.all$ANON_ID %in% diagnoses$ANON_ID))
 
-#cc.all$emerg <- as.factor(cc.all$emerg.x25)
-#cc.all$icu.hdu.ccu <- as.factor(cc.all$icu.hdu.ccu.x25)
-#cc.all$inpat <- as.factor(cc.all$inpat.x25)
-
+cc.all$scripordiag <- as.integer(with(cc.all, as.integer(diag.any)==2 |
+                                                    as.integer(scrip.any)==2))
+                  
 cc.all$sex <- car::recode(as.factor(cc.all$sex), "1='Male'; 2='Female'")
 cc.all <- within(cc.all, sex <- relevel(sex, ref="Female"))
 
-#cc.all$agegr20 <- 20 * floor(0.05 * (cc.all$AGE - 15)) + 15
 cc.all$agegr20 <- as.factor(car::recode(as.integer(cc.all$AGE),
                               "0:39='0-39'; 40:59='40-59';
                                   60:74='60-74'; 75:hi='75 or more'"))
+
+cc.all$agegr3 <-
+    as.factor(car::recode(cc.all$AGE,
+                          "0:59='0-60 years'; 60:74='60-74 years'; 75:hi='75+ years'"))
 
 cc.all$care.home[cc.all$NURSINGHOME==1] <- 1
 cc.all$care.home <- as.factor(car::recode(cc.all$care.home, "0='Independent'; 1='Care/nursing home'"))
[…]

### 2020-05-25 propensity.R

@@ -1,0 +1,56 @@
+
+## calculate propensity score for proton pump using hospitalized not severe
+## select columns containing BNF subpara codes
+
+subpara.colnames <- grep("^subpara\\.", colnames(cc.hosp), value=TRUE)
+subpara.colnames <- grep("Proton pump", subpara.colnames, invert=TRUE, value=TRUE) 
+train.cols <- match(c("y.protonpump", "AGE", "sex", subpara.colnames), colnames(cc.hosp))
+
+train.data <- cc.hosp[, train.cols]
+
+colnames(train.data) <-
+    gsub("(subpara\\.[0-9]+)(.+)", "\\1", colnames(train.data))                
+
+#subparanames.orig <- colnames(cc.hosp)[subparacols]
+
+## clean up the column names so they can be used in a formula
+#colnames(cc.hosp)[subparacols] <- gsub(" |,|-|\\(|\\)|/", "\\.", #colnames(cc.hosp)[subparacols])
+
+## select subparas with frequency > 0.1
+
+subparanames <- grep("^subpara\\.", colnames(train.data), value=TRUE)
+subparanames.freqs <- apply(train.data[, subparanames], 2,
+                           function(x) sum(as.integer(x))) / nrow(train.data)
+
+keep <- subparanames.freqs > 0.05
+subparanames.kept <- subparanames[keep]
+
+lower.formula <- as.formula("y.protonpump ~ AGE + sex")
+upper.formula <- as.formula(paste("y.protonpump ~ AGE + sex +",
+                            paste(subparanames.kept, collapse=" + ")))
+
+full.model <- glm(formula=upper.formula, data=train.data, family="binomial")
+
+cat("Stepwise regression to construct propensity score for proton pump ...")
+stepwise.full <- step(full.model,
+                       scope=list(lower=lower.formula, upper=upper.formula),
+                      direction="both", method="approximate", trace=-1)
+cat("done\n")
+
+## get coeffs, dropping age and sex
+coeffs.full <- summary(stepwise.full)$coefficients[, 1][-(1:3)]
+
+names(coeffs.full) <- gsub("(.+)1$", "\\1", names(coeffs.full))
+
+names(coeffs.full) <- subpara.colnames[match(names(coeffs.full),
+                                             gsub("(subpara\\.[0-9]+)(.+)", "\\1",
+                                                  subpara.colnames))]
+
[…]

### 2020-05-29 casecontrol.Rmd
refactored pharmacoepi analyses as pharmaco.R

@@
   - name: Sharon Kennedy
+    affiliation: ISD
+  - name: David McAllister\textsuperscript{\getAff{Glasgow}} 
     affiliation: HPS
-  - name: David McAllister 
-    affiliation: Glasgow
-  - name: Chris Robertson 
-    affiliation: Strathclyde
+  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}} 
+    affiliation: HPS
   - name: Rachael Wood
@@
     affiliation: HPS
-  - name: Sharon Hutchinson 
-    affiliation: GCU
+  - name: Jim McMenamin
+    affiliation: HPS
+  - name: Colin Ramsay
+    affiliation: HPS
+  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
+    affiliation: HPS
   - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
-    affiliation: HPS
-  - name: Public Health Scotland COVID-19 Health Protection Study Group
     affiliation: HPS
@@
   - code: HPS
-    address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+    address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
 
   - code: Glasgow
-    address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health
+    address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Wellcome Trust Intermediate Clinical Fellow and Beit Fellow
 
@@
 csl: ../plos.csl
-urlcolor: red
+urlcolor: blue
 ---
+
+On behalf of Public Health Scotland COVID-19 Health Protection Study Group
 
@@
 _Methods_ -- The design was a matched case-control study.
-Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national database.
+Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national virology laboratory database.
 Those who had been admitted to a critical care unit or had died within 28 days of a positive test were classified as severe cases.
 For each case, ten controls matched for sex, age and primary care practice were selected from the population register.
-For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from encashed prescriptions issued in primary care care during the past year and linked to the national diabetes register.
+For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from prescriptions encashed during the past nine months and we linked to the national diabetes register.
 Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.
@@
 In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
-Univariate rate ratios (95% CIs) for eight conditions listed by public health agencies as conferring high risk were `r table.agegr.all[6, 3]` for Type 2 diabetes,
+Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were  `r table.agegr.all[5, 3]` for Type 1 diabetes,  `r table.agegr.all[6, 3]` for Type 2 diabetes,
 `r table.agegr.all[8, 3]` for ischemic heart disease,
@@
 In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[4, 2])` in cases and `r gsub("\\\\", "", table.agegr[4, 1])` in controls.
-The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`).
-These associations remained after excluding those with any of the listed conditions.
+The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`; these associations remained strong and significant after adjusting for the listed conditions.
+Most cases without listed conditions had either prescriptions or hospitalisations, and highly significant associations with severe disease were seen across many hospital diagnoses and drug categories.
 
-With a model based on age and sex alone, the expected information for discrimination was `r round(cases.Lambda.agesex, 2)` bits.
-The additional information for discrimination conditional on age and sex was `r table.densities[1, 5]` bits for a model based on demographic variables only, increasing to `r table.densities[2, 5]` bits for a model that included the listed conditions, and to `r table.densities[3, 5]` bits for a model that included indicator variables for hospital diagnoses and prescription encashments.
+A predictive  model based on age and sex alone had an expected information for discrimination of  `r round(cases.Lambda.agesex, 2)` bits.
+In the model conditional on age and sex further demographic variables provided another `r table.densities[1, 5]` bits,  and this increased to `r table.densities[2, 5]` bits  when listed conditions were included, and to `r table.densities[3, 5]` bits when indicator variables for hospital diagnoses and prescription encashments were also included.
 
@@
 Quantification of the risk associated with characteristics and co-morbidities has been limited by the lack of comparisons with the background population [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].
-Two recent studies in the UK have included population comparators and have reported associations with co-morbidities including diabetes, asthma and heart disease.
-[@niedzwiedz_ethnic_2020; @the_opensafely_collaborative_opensafely_2020].
-A systematic comparison of medical records in those with and without severe disease has not yet been reported.
+Two recent studies in the UK have included population comparators and have reported associations of in hospital test positive persons and COVID-19 death in hospital with co-morbidities including diabetes, asthma and heart disease [@niedzwiedz_ethnic_2020; @the_opensafely_collaborative_opensafely_2020].
+These studies  have focused on conditions presumptively listed by public health agencies as increasing risk for COVID-19 based on case series data.
 
-We report a systematic examination of the hospitalisation record and prescribing history among all people who have developed severe COVID-19 disease in Scotland compared to the general population.
-The objectives were to identify risk factors and to lay the basis for risk stratification based on a predictive model.
+Here we examine the frequency of sociodemographic factors and  these listed conditions in  all people with severe COVID-19 disease in  Scotland compared to matched controls from the general population.
+In those without listed conditions we report a systematic examination of the hospitalisation record and prescribing history in severe  COVID-19 cases  compared to controls.
+The objectives were to identify  risk factors for severe COVID-19 and to lay the basis for risk stratification based on a predictive model.
 
@@
 Hospital admissions from the time of testing were obtained from the RAPID database a daily return of current hospitalisations each day.
-Admissions to intensive care units (ICUs) care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers all admissions to all adult ICUs and to combined ICU/high dependency units ( hereafter abbreviated as ICU) across Scotland and has returned a daily census of patients in ICU from the beginning of the COVID-19 epidemic.
+Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers admissions to critical care [comprising adult intensive care units (ICUs), high dependency units (HDUs) and combined ICU / HDU units] across Scotland and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic.
 Death registrations up to `r gsub("^0", "", format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y"))` were obtained from linkage to the National Register of Scotland.
 
-Severe or fatal COVID-19 was defined by a record of entering critical care (intensive care, high dependencey unit or combined unit) in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-By restricting the case definition to cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, irrespective of any selection that may have excluded those assessed as unlikely to benefit from critical care.
+Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
+By restricting the case definition to those cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, whatever selection policies may have determined admission to hospital or entry to critical care.
 
@@
 Residence in a care home was ascertained from the CHI database.
-Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation, an indicator based on postal code.
+Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD), an indicator based on postal code.
 Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.
@@
 For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge.
-Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric hospital and pregnancy units returns are not captured in SMR01.
+Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric and maternity unit returns are not captured in SMR01.
 British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
@@
 
-We began by scoring a specific list of eight conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+We began by scoring a specific list of  conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
 A separate list of conditions designates "clinically extremely vulnerable" individuals who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.
-Cases of diabetes in the last update of the register were supplemented with new cases identified through discharge codes and drug codes.
-These new cases were coded as type Other/Unknown.
+Recent cases of diabetes since the last update of the register were supplemented with  a small number new cases identified through discharge codes and drug codes.
+These new cases were added into the diabetes type Other/Unknown category.
 
 ## Statistical methods
-To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which are available by one-year age group up to age 90 years.
+To estimate the relation of cumulative incidence and mortality from COVID-19 to age and sex, logistic regression models were fitted to the proportions of cases and non-cases in the Scottish population, using the estimated population of Scotland in mid-year 2019 which were available by one-year age group up to age 90 years.
 To allow for possible non-linearity of the relationship of the logit of risk to age, we also fitted generalized additive models, implemented in the R function `gam::gam`, with default smoothing function.
@@
 With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.
-Odds ratios in a matched case control study are based on the conditional likelihood and  the unconditional odds ratios calculable from the frequencies of exposure in cases and controls will differ from these and should not be used [@breslow_estimation_1978].
-Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios but with less precision.[@breslow_estimation_1978]
+Note that odds ratios in a matched case control study are based on the conditional likelihood and  the unconditional odds ratios calculable from the frequencies of exposure in cases and controls will differ from these and should not be used [@breslow_estimation_1978].
+Although matching on primary care practice will match to some extent for associated variables such as care home residence, socioeconomic disadvantage and prescribing practice, the effects of these variables are still estimated correctly by the conditional odds ratios but with less precision than in an unmatched study of the same size [@breslow_estimation_1978].
 
 To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.
-The performance of the risk prediction model in classifying cases versus non-cases of severe COVID-19 was examined by cross-validation.
+The performance of the risk prediction model in classifying cases versus non-cases of severe COVID-19 was examined by `r nfold`-fold cross-validation.
 We calculated the performance calculated over all test folds using the C-statistic but also using the "expected information for discrimination" $\Lambda$ expressed in bits [@mckeigue_quantifying_2019].
 The use of bits (logarithms to base 2) to quantify information is standard in information theory: one bit can be defined as the quantity of information that halves the hypothesis space.
-Although readers may be unfamiliar with the expected information for discrimination $\Lambda$, it has several properties that make it more useful than the C-statistic for quantifying increments in the performance of a risk prediction model [@mckeigue_quantifying_2019] .
+Although readers may be unfamiliar with the expected information for discrimination $\Lambda$, it has several properties that make it more useful than the C-statistic for quantifying increments in the performance of a risk prediction model [@mckeigue_quantifying_2019].
 A key advantage of using $\Lambda$ is that contributions of independent predictors can be added.
@@
 The relationship of mortality to age is almost exactly linear on a logit scale, and the lines for male and female mortality are almost parallel.
-For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.
-This implies that at younger ages the ratio of ICU admissions to total fatalities is higher in women than in men, but that at later ages the ratio of ICU admissions to total fatalities is higher in men.
 In models that included age and sex as covariates, the odds ratio associated with a 10-year increase in age was `r round(exp(10*logistic.coeffs[3, 1]), 2)` for all severe disease and `r round(exp(10*logistic.coeffs[3, 2]), 2)` for fatal disease.
 The odds ratio associated with male sex was `r round(exp(-logistic.coeffs[2, 1]), 2)` for all severe disease and `r round(exp(-logistic.coeffs[2, 2]), 2)` for fatal disease.
-As shown in Table \ref{tab:agegr} almost all severe cases aged over 75 years died but most severe cases aged under 60 years survived.
+For severe cases as defined in this study, the sex differential is narrow up to about age 50 but widens between ages 50 and 70 years.
+Thus at younger ages the ratio of critical care admissions to total fatalities is higher in women than in men, but that at later ages the ratio of critical admissions to total fatalities is higher in men.
 
@@
 Associations with ethnicity are shown for the full dataset based on name classification and separately for the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record.
-With Whites as reference category, the rate ratio (95% CI) associated with South Asian ethnicity was `r table.demog.append[2, 4]` based on name classification and `r table.demog.append[11, 4]`, based on the SMR records.
+With Whites as reference category, the rate ratio (95% CI) associated with South Asian ethnicity was `r table.demog.append[2, 4]` based on name classification and `r table.demog.append[11, 4]`, based on the subset with SMR records.
 The numbers of cases in other non-White ethnic groups were too sparse to tabulate separately.
@@
 Prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
-`r gsub("\\\\", "", table.agegr[nrow(table.agegr), 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[nrow(table.agegr), 1])` of the controls.
-In those aged 75+ years `r gsub("\\\\", "", table.agegr[nrow(table.agegr), 8])` of the cases and `r gsub("\\\\", "", table.agegr[nrow(table.agegr), 7])` of the controls had at least one listed condition.
-There was a much higher prevalence of at least one prescription in cases in controls even in those aged under under 40 years.
-Among those aged under 40 years, `r table.scripordiag[2, 2]` of the cases and `r table.scripordiag[2, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
+`r gsub("\\\\", "", table.agegr[4, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[4, 1])` of the controls.
+In those aged 75+ years `r gsub("\\\\", "", table.agegr[4, 8])` of the cases and `r gsub("\\\\", "", table.agegr[4, 7])` of the controls had at least one listed condition.
+The proportion with at least one dispensed prescription was much higher in cases than in controls in each age group.
+Among those aged under 40 years, `r table.agegr[6, 2]` of the cases and `r table.agegr[6, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
 
@@
 
-Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 year age bands combined.
+Supplementary Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} examine these associations by age group, with the 0-39 and 40-59 year age bands combined.
 All listed conditions were associated with severe disease in each age band.
-In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][6, 3]` for Type 1 diabetes and `r tables.agegr[[1]][7, 3]` for Type 2 diabetes.
+In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][5, 3]` for Type 1 diabetes and `r tables.agegr[[1]][6, 3]` for Type 2 diabetes.
 The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and within each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
-Table \ref{tab:fatal} shows that in each age group the proportion of cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.
+Table \ref{tab:fatal} shows that in each age group the proportion of fatal cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.
 
@@
 In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-Note that hospitalisations classified under the pregnancy chapter here derive from admission to non-obstetric units with pregnancy related medical conditions since in Scotland obstetric returns are not in the SMR01 dataset.
-The strongest residual associations were seen with diagnoses in ICD chapters 2 (neoplasms), 5 (mental) and 10 (musculoskeletal).
-Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.
-This table is filtered to show only subchapters for which the univariate p-value is <0.001 and there are at least 50 cases and controls with a diagnosis in this subchapter.
+Note that hospital diagnoses classified under the pregnancy chapter here are derived from admissions with pregnancy related medical conditions to non-obstetric units only, as obstetric returns are not in the SMR01 dataset.
+In a multivariate analysis the strongest associations were with diagnoses in ICD chapter 2 (neoplasms).
+Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.
+This table is filtered to show only subchapters for which the univariate p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.
 This shows that many diagnoses are associated with markedly higher risk of severe COVID-19.
@@
 ## Associations of prescribed drugs with severe disease 
-Tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of at least one prescription in the last year.
+As shown in Table \ref{tab:listed.conditions.allages} and supplementary tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} , the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of at least one prescription in the last year.
 The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
-In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` About one third of those over 75 had not encashed a prescription in the previous year.
+In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` respectively.
+About one third of controls aged over 75 had not encashed a prescription in the previous year.
 
@@
 In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
-In a multivariate analysis of all chapters, these associations were much weaker.
-The BNF chapters with the strongest association with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular)).
+In a multivariate analysis of all chapters, most of these associations were weaker.
+The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular)).
 Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
@@
 ## Construction of a multivariate risk prediction model 
-To evaluate the incremental contributions of the listed conditions, and residual information in hospitalisation and prescription records data after assigning these conditions, predictive models were constructed from three sets of variables: a baseline set consisting only of demographic variables, a set that included indicator variables for each listed condition, and an extended set that included demographic + listed conditions + indicator variables for hospital diagnoses in each ICD-10 chapter and prescriptions in each BNF chapter.
+To evaluate the contribution of the listed conditions to risk prediction, and the incremental contribution of other information in hospitalisation and prescription records after assigning these conditions, predictive models were constructed from three sets of variables: a baseline set consisting only of demographic variables, a set that included indicator variables for each listed condition, and an extended set that included demographic, variables, indicator variables for listed conditions and indicator variables for hospital diagnoses in each ICD-10 chapter and prescriptions in each BNF chapter.
 
 For each variable set, a stepwise regression procedure was carried out using alternating forward-backward selection.
-The variables retained with each variable set are shown in Table \ref{tab:stepwise} .
+The variables retained with each variable set are shown in Table \ref{tab:stepwise}.
+Coefficients for specific conditions here should not be interpreted as effect estimates, as global variables for any hospital diagnosis and any listed condition have been included in the model.
 The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation.
-Observed and predicted case status were compared over all test folds.
+Observed and predicted case status were compared within each stratum over all test folds.
 Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.
@@
 ### Sociodemographic factors
-This analysis shows that, as expected, risk for severe COVID-19 is associated with age, male sex and socioeconomic deprivation.
+This analysis confirms that risk for severe COVID-19 is associated with increasing age, male sex and socioeconomic deprivation.
 The slope of the relationship of severe disease (on the scale of log odds) to age is less steep than the slope of the relationship of fatal disease to age.
-Another way of expressing this is that the fatality rate of severe disease increases with age.
+<!---Another way of expressing this is that the fatality rate of severe disease increases with age.
 Although the relationships of mortality to age are linear and parallel for men and women, the fatality rate of severe disease is lower in women than in men up to age 50 and higher in women than in men between ages 50 and 70.
-This may reflect either variation in the biology of the disease, or differential selection for entry to intensive care.
-Residence in a care home was associated with a `r round(as.numeric(substr(table.agegr.all[1, 3], 1, 4)))`-fold increased rate of severe COVID-19 in this age matched analysis, reduced to `r round(as.numeric(substr(table.agegr.all[1, 5], 1, 4)))`-fold by adjustment for listed conditions.
+This may reflect either variation in the biology of the disease, or differential selection for entry to critical care.--->  Residence in a care home was associated with a `r round(as.numeric(substr(table.agegr.all[1, 3], 1, 4)))`-fold increased rate of severe COVID-19 in this age matched analysis, reduced to `r round(as.numeric(substr(table.agegr.all[1, 5], 1, 4)))`-fold by adjustment for listed conditions.
 This excess risk is likely to reflect both the spread of the epidemic in care homes and residual confounding by frailty.
 
-Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now, and show no evidence that risk is higher in South Asians than in Whites.
-This contrasts with reports from England [@the_opensafely_collaborative_opensafely_2020] of risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, after adjustment for comorbidities and socioeconomic status.
-In a study based on the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated the crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-In the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.
-Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average.
-As the relative socioeconomic position of these groups in Scotland is very different to that in England, it is to be expected that the relation of health status to ethnicity will also differ [@walsh_increasingly_2019].
+Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now.
+The only non-White ethnic group with any sizeable numbers is the South Asian category and we found no evidence of any elevation in risk in this group compared to Whites.
+Reports from England [@the_opensafely_collaborative_opensafely_2020] found elevation in risks for some non-White groups.
+In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.
+In a study of risk factors for hospitalized disease in the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated but did not fully explain the elevated crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
+The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity will also differ.
+For example in the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.
+Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average [@walsh_increasingly_2019].
 Future work may allow more complete assignment of ethnicity and disaggregation of broad categories based on continent of origin.
@@
 ### Co-morbidities
-We have confirmed that the " moderate risk " conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.
+We have confirmed that the moderate risk conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.
 However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages.
-The rate ratios of `r substr(table.agegr.all[6, 3], 1, 4)` for Type 1 diabetes and `r substr(table.agegr.all[7, 3], 1, 4)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.
+The rate ratios of `r round(as.numeric(substr(table.agegr.all[5, 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all[6, 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.
 We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
@@
 This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure.
-However overall 26% of cases did not have any of these listed conditions.
+Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all[nrow(freqs.all), 2])` of severe cases had at least one of these listed conditions.
 Among cases and controls without these conditions, not surprisingly, neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as in the current dataset we cannot separately identify those who are currently receiving chemotherapy or radiotherapy for whom shielding is advised.
 We have not attempted to estimate the risk associated with these conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding rather than on the risk to those exposed to the epidemic.
-A further systematic evaluation of past hospitalisation history did not reveal a sparse set of conditions underlying severe disease in those without  listed conditions but rather many associated diagnoses.
+In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.
 
 Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.
-However we found that almost half of cases under 40 years had at least one of the listed conditions.
+However we found that almost half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.
 In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.
@@
 Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.
-Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, causal effects of drug classes cannot be ruled out.
+Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.
 These associations are explored in an accompanying paper.
@@
 As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.
-A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 2)` associated with a 20-year increase in age is far stronger than that associated with conditions such as diabetes, asthma, and neurological conditions that impair mobility.
-A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one-third of those aged 60-74 year olds in the population have at least one of the listed conditions we examined.
-It will also exclude many people at high risk because they have multiple risk factors.
+A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 1)` associated with a 20-year increase in age is far stronger than that associated with common diseases such as Type 2 diabetes and asthma that are listed as conditions associated with high risk.
+A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.
+It will also exclude many people at high risk because they have multiple risk factors each of small effect.
 The only way to optimize risk classification so as to ensure equity with respect to risk is to construct a classifier that uses all available information to assign a risk score.
-Our results show that this is possible in principle: in Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.
+Our results show that this is possible in principle, though for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem.
+In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.
 
@@
 This case-control study is limited to test-positive cases, excluding deaths with no record of a positive test where COVID-19 was mentioned on the death certificate as an underlying or contributing cause.
-Up to 13 May 2020 an additional 1200 such deaths had been reported by the [National Records of Scotland.](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland) Future versions of this analysis will include sensitivity analyses of the extent to which the results are changed by including these deaths, but the data are not yet available.
+Up to 13 May 2020 an additional 1200 such deaths had been reported by the National Records of Scotland.
+Future analyses of this study will include sensitivity analyses of the extent to which the results are changed by including these deaths, but linked data are not yet available to us.
 Apart from residential care home status, we do not expect most other risk factors to differ markedly between those who died from COVID-19 without being tested and those who died after testing.
@@
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the     HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
 
 ## Public Health Scotland COVID-19 Health Protection Study Group
-Alice Whettlock$^1$,  Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Jim McMenamim$^1$, Joseph Jasperse$^1$,  Josie Murray$^1$,  Karen Heatlie$^1$,  Lindsay Mathie$^1$,  Lorraine Donaldson$^1$, Martin Paton$^1$,  Martin Reid$^1$,  Melissa Llano$^1$,    Michelle Murphy-Hall $^1$,  Paul Smith$^1$,  Ros Hall$^1$,  Ross Cameron$^1$,  Susan Brownlie$^1$,  Adam Gaffney$^2$,  Aynsley Milne$^2$,  Christopher Sullivan$^2$,  Edward McArdle$^2$,  Elaine Glass$^2$,  Johanna Young$^2$,  William Malcolm$^2$,  Jodie McCoubrey $^2$
+Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
 
@@
 <div id="refs"></div>
-
 \newpage
@@
 ## Footnote for Figure \ref{fig:roc}
-The crude receiver operator characteristic curve (ROC) is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and non-cases.
+The crude receiver operator characteristic (ROC) curve is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and non-cases.
 The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the score, evaluated over all possible such pairs in the dataset.
@@
     column_spec(column=4, width="2.5cm") %>%
-    group_rows("Ethnicity based on name classification", 1, 3) %>%
-    group_rows("SIMD quintile", 4, 8) %>%
-    group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13,
+    group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
+    group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
+    group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
                hline_before=TRUE, latex_gap_space="10pt") %>% 
@@
              caption="Frequencies of risk factors in cases and controls, by age group") %>%
-    column_spec(column=1, width="3.5cm") %>%
-    group_rows(" ", 7, nrow(table.agegr[-7, ])) %>%
+    column_spec(column=1, width="4cm") %>%
+    column_spec(column=2:9, width="2cm") %>%
     add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
@@
     landscape() %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r scripordiag.fatal, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.scripordiag.fatal, 
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="fatal",
-             align=rep("c", 2), 
-             caption="Proportions of fatal cases and controls with no dispensed prescription or hospital diagnosis, by age group") %>%
     kable_styling(latex_options="hold_position")
@@
              caption="Associations of severe disease with listed conditions over all age groups") %>%
-    group_rows(" ", 4, nrow(table.agegr.all[-4, ])) %>%
-    column_spec(column=1, width="4cm") %>%
+    # group_rows(" ", 4, nrow(table.agegr.all[-4, ])) %>%
+    column_spec(column=1, width="5cm") %>%
+    column_spec(column=2:7, width="2.5cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
     landscape() %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r scripordiag.fatal, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.scripordiag.fatal, 
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="fatal",
+             align=rep("c", 2), 
+             caption="Proportions of fatal cases and matched controls without and with a  dispensed prescription or hospital diagnosis, by age group") %>%
+   group_rows("Age <60", 1, 2) %>%
+   group_rows("Age 60-74", 3, 4) %>%
+   group_rows("Age 75+", 5, 6) %>%
    kable_styling(latex_options="hold_position")
@@
              caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
-    group_rows(" ", 4, nrow(tables.agegr[[1]]) - 1) %>%
+#   group_rows(" ", 4, nrow(tables.agegr[[1]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
+    column_spec(column=2:7, width="2.7cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
-    group_rows(" ", 4, nrow(tables.agegr[[2]]) - 1) %>%
+ #   group_rows(" ", 4, nrow(tables.agegr[[2]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
+    column_spec(column=2:7, width="2.7cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
-    group_rows(" ", 4, nrow(tables.agegr[[3]]) - 1) %>%
+#    group_rows(" ", 4, nrow(tables.agegr[[3]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
+   column_spec(column=2:7, width="2.7cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              caption="Stepwise regression: variables retained in model for severe disease") %>%
-    group_rows("SIMD - quintile 1 as reference", 2, 5) %>%
-    group_rows("Diabetes - non-diabetic as reference", 6, 8) %>%
+    group_rows("SIMD - quintile 1 as reference", 2, 5, bold=FALSE) %>%
+    group_rows("Diabetes - non-diabetic as reference", 6, 8, bold=FALSE) %>%
     kable_styling(latex_options="hold_position")

### 2020-05-29 pharmaco.Rmd
refactored pharmacoepi analyses as pharmaco.R

@@
   - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
-    affiliation: HPS
-  - name: Public Health Scotland COVID-19 Health Protection Study Group
     affiliation: HPS
@@
 
-```{r functions, echo=FALSE}
+On behalf of Public Health Scotland COVID-19 Health Protection Study Group
+
+```{r functions, echo=FALSE, message=FALSE}
 library(knitr)
@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
+options(knitr.kable.NA = '')
+
+stepwise.drop.subparas[, 1] <- gsub("subpara\\.", "", stepwise.drop.subparas[, 1])
+
+firstrow.neg <- which(stepwise.drop.subparas[, 3] < 0)[1]
+rows.print <- firstrow.neg:nrow(stepwise.drop.subparas)
 
@@
 
-_Background_-- We previously reported that dispensing of at least one prescription from primary care in the past year was strongly associated with severe COVID-19 in Scotland, independently of past diagnoses of conditions designated as high risk.
-The objective of this study was to investigate possible explanations for this association.
-.
+_Background_-- We previously reported that dispensing of at least one prescription from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as high risk (listed conditions).
+Here we conducted an initial investigation of possible explanations for this association which include  specific co-morbidity or  drug effects but also include more diffuse frailty or polypharmacy effects.
 
-_Methods_ -- Severe cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database who entered intensive care or died within 28 days.
+_Methods_ -- Test positive COVID-19 cases ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database.
 From the national population register ten controls per case were selected, matched for sex, age and general practice.
-International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary in the past nine months.
+Severe cases were those  who entered  a critical care unit or died within 28 days of testing.
+International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary in the past nine months and used to define  history of listed conditions.
+To examine if  any unexpected specific drug classes were likely to contribute to severe COVID19 risk, among  severe cases and controls without listed conditions  (INSERT n cases n controls) we undertook a backward selection procedure across all BNF chapters.
+Across the retained chapters we undertook a backward selection procedure at sub-paragraph level.
+This yielded proton pump inhibitors (PPI) as the drug class that most explained the association of a prescription in the past year with COVID-19.
+Accordingly we built a propensity score of PPI prescribing  across all 218 BNF subparagraph codes in a separate dataset of hospitalised non-severe cases and their controls.
+This score, which increases as polypharmacy increases, was then used as a covariate that allowed adjustment of the association of PPIs with severe COVID-19 across all severe cases and their controls for confounding by co-prescribing.
 Rate ratios were estimated by conditional logistic regression.
 
-_Findings_ -- In those not resident in a care home and without past diagnoses of any conditions designated as high risk, the rate ratio (95% CI) associated with dispensing of any prescription in the last year was `r table.scrip.cats3[3, 3]`.
-A backward selection procedure showed that two chapters of the BNF -- drugs for the gastrointestinal system (Chapter 1) and the cardiovascular system (Chapter 2) -- accounted for `r round(100 * stepwise.drop[13, 2] / stepwise.drop[1, 2])`% of this effect.
-A further backward selection of drug codes in these two chapters identified proton pump inhibitors as the drug class with the strongest independent association with severe COVID-19.
+*Findings* -- The risk of severe disease was strongly associated with the total number of drug classes dispensed in the last nine months.
+
+Across sub-paragraphs of these chapters identified proton pump inhibitors as the drug class with the strongest independent association with severe COVID-19.
 The rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[4, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 40 years.
-Adjustment for hospital diagnoses of disorders of the esophagus, stomach and duodenum, and prescribing of antiplatelet or nonsteroidal anti-inflammatory agents reduced these effect sizes only slightly.
+Adjustment for hospital diagnoses of disorders of the esophagus, stomach and duodenum reduced these effect sizes only slightly.
+In those under 60 years only, the risks ratios increased with increasing average dispensed daily dose.
+More recent exposure was not more strongly associated than earlier exposure.
+With adjustment for propensity score associations were no longer significant in those aged 75+ but remained strong INSERT in those aged under 60 years.
+Restricting to those with no other dispensed prescriptions, the rate ratio associated with ever-use of a proton pump inhibitor was `r table.protonpump.bynumdrugs[1, 3]`.
 
-Other drug classes in Chapter 1, including laxatives and agents influencing motility, were associated with higher rates of severe COVID-19 but only proton pump inhibitors and osmotic laxatives were independently associated with disease in a multivariate analysis.
-Among drugs prescribed for cardiovascular disease, only loop diuretics and calcium channel blockers were independently associated with severe COVID-19 in a multivariate analysis.
-Associations of disease with ACE inhibitors and angiotensin-II receptor blockers were similar to those with other antihypertensive agents.
 *Conclusions* -- A systematic search for drug classes that underlie the association of severe COVID-19 with dispensing of any prescription has identified proton pump inhibitors as the drug class making the largest independent contribution to this association.
-The association with proton pump inhibitors is not explained by adjusting for covariates.
-As severe disease is associated with other drug classes such as laxatives that are commonly prescribed for minor symptoms, confounding by factors associated with these symptoms cannot be excluded.
-However as increased risk of infection is an established adverse effect of this class of drugs, a causal effect is plausible.
-In the Scottish population where 28\% of controls have used proton pump inhibitors and the rate ratio is 2.8, one-third of severe cases could be attributable to use of these drugs.
-This may be relevant to high rates of severe COVID-19 reported in other populations where proton pump inhibitors are overprescribed.
-Until the effects of proton pump inhibitors on susceptibility to severe COVID-19 have been investigated further, use of these drugs without a recognized medical indication is inadvisable.
+Prescription of this drug class is recognised as a a sentinel for polypharmacy and for frailty and  the  data were consistent with this  explanation for the association in the older population.
+However the strength of the association in those under age 60, its independence of the propensity score and the observation of an association  in those not on any other drugs raise concern of a possible causal effect that requires further investigation.
+In the meantime   existing advice on the avoidance of  inappropriate and over-prescribing of PPIS should be reinforced.
 
@@
 # Background
+The REACT-SCOT  study is a matched case control study of COVID-19 cases versus controls in Scotland.
+In our first paper from this study we confirmed and quantified the risk ratios associated with conditions  such as diabetes, asthma and heart disease that had been  presumptively listed by the authorities as high risk for COVID-19 based on case series.
+We further showed that in those without listed conditions prior hospitalisation in the previous five years  and prescription history  in the previous nine months  were strongly associated with severe disease.
+We went on to show that the association with hospitalisation was not explained by just one or two other specific conditions but that many conditions of varying prevalence were more common in cases than controls.
+
 A striking preliminary finding of the REACT-SCOT study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
 The rate ratio associated with this variable varied from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
-In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` About one third of those over 75 had not encashed a prescription in the previous year.
-The objective of this study was to distinguish between three possible explanations for this unexpected association: confounding by other high-risk conditions, an effect associated with consultations for symptoms not necessarily associated with high-risk conditions, and a direct effect of exposure to one or more classes of drug on risk of severe disease.
+In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]`
+
+There are several potential explanations for such a finding;  there may be a sparse set of  drugs underlying this association or  there may be a large set of drugs contributing to it.
+These in turn may reflect  a i)sparse set of direct drug effects or ii) sparse set of underlying morbidities or a iii) larger set of direct drug effects  or  their interactions ( i.e.
+an effect due to the polypharmacy itself)  or a iv) larger set of co-morbidities ( i.e.
+that usually conceptualised as frailty) or even v) interactions between polypharmacy and frailty.
+
+The objective of this study was to conduct an initial analysis to discover if there is a a sparse set of direct drug effects underlying the finding of the strong association of any prescription in the past year and severe COVID-19.
+Further analyses will explore the role of polypharmacy and frailty.
 
@@
 ## Case ascertainment and sampling of controls 
-The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.
-All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained for this study.
-Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01.
-Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland.
+All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
+Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units.
 Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
-Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-
+Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
 For each test-positive case, the CHI database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
@@
 For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
-Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.
 
-British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2.
-This ensures that only prescriptions dispensed before the onset of symptoms are included.
+British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
 The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
-For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other".
-Most of these codes are for dressings and appliances rather than drugs.
-Defined daily doses (DDDs) were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.
-Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to 16 days before the date that the case (in a stratum consisting of case and matched controls) first tested.
+BNF subparagraph codes generally represent pharmacological drug classes.
+For this analysis prescription codes from chapters 14 and above, comprising mostly dressings and appliances but also vaccines, anaesthesia and other preparations were grouped as "Other".
+Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.
+Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to 16 days before the date that the case (in a stratum consisting of case and matched controls) first tested positive.
 
 # Statistical analysis 
-We derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.
+As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+For this analysis the list was extended to include non-cardiac circulatory disease as our analyses of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.
+These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that condition is the only indication.
+Diagnoses of diabetes were extracted from linkage to the national diabetes register.
 The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.
-Rate ratios were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
 
-# Results
-## Dispensing of any prescription by care home status and presence of any listed condition
-Table \ref{tab:scrip.cats3} shows that at least one prescription had been dispensed for `r table.scrip.cats3[1, 1]` of controls who were resident in care homes and `r table.scrip.cats3[2, 1]` of controls a diagnosis of a listed condition who were living independently.
-The association of severe disease with dispensing of at least one prescription was restricted to those not resident in a care home and without a diagnosis of a listed condition, in whom the rate ratio (95% CI) was `r table.scrip.cats3[3, 3]`.
-To investigate which drug classes contribute to this association, the analysis was therefore restricted to this group.
-We undertook a backward stepwise variable selection procedure, based on BNF chapters which group drug classes by the body systems on which they act.
-Starting with the "any prescription" variable based on all 14 BNF chapters, models that dropped each of these chapters, retaining the other 13, were compared and the model with highest log-likelihood was retained.
-This procedure was repeated until the any prescription variable was based on a single remaining chapter.
-Table \ref{tab:stepwise.drop} shows the results of this process.
-Only the last four chapters decreased the log-likelihood when dropped: in order of deselection, these were chapters 10 (musculoskeletal), 9 (nutrition and blood), 1 (gastrointestinal) and 2 (cardiovascular).
+To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we carried out a backward stepwise variable selection procedure. 
+Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.
+The model with the highest log-likelihood was used to select which drug class should be dropped.
+This procedure was repeated with the retained variables until only a single drug class remained.
 
-To investigate which drug classes in Chapters 1 and 2 contribute to the "any dispensed prescription" effect, we repeated the backward stepwise variable selection procedure on all BNF subparagraph codes in these two chapters.
-Subparagraph codes correspond generally to drugs classified by their mode of action.
-Of the 48 subparagraph codes in Chapters 1 and 2, only the last ten retained by the stepwise variable selection decreased the log-likelihood when dropped.
-The results of the stepwise variable selection for these last ten are shown in Table \ref{tab:stepwise.subparas}.
-The cardiovascular drug classes retained at this stage were oral anticoagulants, angiotensin-II receptor antagonists, angiotensin-converting enzyme inhibitors, loop diuretics and calcium-channel blockers.
-The gastrointestinal drug classes retained at this stage were drugs altering gut motility, osmotic laxatives and proton pump inhibitors.
-The last variable retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.
-This led us to examine further the association of proton pump inhibitors with severe disease, without restricting the analysis to those not resident in care homes and without a diagnosis of a listed condition.
-
-## Construction of a propensity score
-We began by constructing a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
+We constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
 To construct this score, we used an independent dataset so that the score would not overfit the case-control dataset.
@@
 The propensity score was constructed on this training dataset by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
-The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
-As expected, the variables retained included NSAIDs and antiplatelet drugs, but also many other drug classes such as SSRIs that are not recognized medical indications for coprescribing of a proton pump inhibitor.
-The coefficients for age and sex were dropped before calculating the propensity score in the case-control dataset, because the case-control analysis is matched for age and sex.
 
-<!--- add something about how well the propensity score predicts ever-use of a PPI --->
+Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
 
-## Associations with proton pump inhibitor use
+# Results
+## Relation of severe COVID-19 to number of drug classes dispensed. 
+Table \ref{tab:numdrugs.agegr} shows that severe COVID-19 was strongly related to the total number of different drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed since 1 June 2019.
+The strength of this association was present in all age groups.
+With no dispensed prescriptions as reference category, the rate ratios associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+
+Table \ref{tab:numdrugs.cats3} shows this relationship after stratifying by care home residence and diagnosis of at least one listed condition.
+This shows that the association of severe disease with number of drug classes dispensed was weak in care home residents, absent in those living independently with a diagnosed listed condition, and very strong in those who were living independently without a listed condition.
+In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.cats3[3, 11]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.cats3[7, 11]` in those with more than 12 drug classes dispensed.
+To examine this association further, we restricted the analysis to those not resident in a care home and without listed conditions so as to lessen confounding by underlying disease states and by frailty.
+
+As a next step, we partitioned the number of drug classes dispensed into totals for cardiovascular and non-cardiovascular drugs, on the basis that prescribing of multiple drugs for control of cardiovascular disease is supported by evidence-based guidelines, while prescribing of multiple drugs for other indications is more likely to represent inappropriate polypharmacy.
+Table \ref{numdrugs,cardiovasc} shows that numbers of drugs in both categories were associated with COVID-19.
+
+We next estimated univariate rate ratios for severe COVID-19 for indicator variables for dispensing of each of the `r length(subparacols)` BNF subparagraph codes.
+Table XX shows the results for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.
+These COVID-associated drug classes include many drug classes previously identified in Scotland as indicators of overprescribing and inappropriate polypharmacy.
+
+## Backwards selection to identify specific drug classes contributing to the association of severe COVID-19 with number of drug classes dispensed
+A backwards stepwise variable selection procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with number of drugs dispensed.
+To save computation, the stepwise procedure was initiated with the the subset `r length(subparacols.forstepwisedrop)` drug classes that showed univariate associations.
+The results of this procedure are shown in Table \ref{tab:stepwise.subparas}.
+Dropping the first `r firstrow.neg - 1` drug classes from the set of drug classes used to compute the number of dispensed drug classes increased the log-likelihood of the model.
+Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
+
+The only cardiovascular drug classes included among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
+Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.
+The last drug class retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.
+
+## Associations of severe COVID-19 with proton pump inhibitor use
+Testing for independent association of COVID-19 with exposure to proton pump inhibitors is challenging because these drugs are usually co-prescribed with other drugs, and because in this dataset the disease is strongly associated with polypharmacy.
+We used several approaches:
+
+* testing for a dose-response relationship,
+
+* examination of the relationship by age.
+
+* adjusting for prespecified covariates
+
+* comparison of dispensing between recent and less time windows.
+
+* construction of a propensity score using an independent dataset.
+
+For these analyses the dataset was not restricted to exclude care home residents or those with diagnosed with a listed condition.
+
+### Dose-response relationship
+Table \ref{tab:dosegr.protonpump} shows how the effect size varies with the dose of proton pump inibitor, grouped into categories.
+Table \ref{tab:ppinames} compares the effects of the five drugs in this class.
+Omeprazole and lansoprazole were by far the most commonly prescribed drugs.
+The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day was higher for lansoprazole than for omeprazole.
+This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.
+
+### Relationship by age
 Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for propensity score.
@@
 
-Table \ref{tab:dosegr.protonpump} shows how the effect size varies with the dose of proton pump inibitor, grouped into categories.
+### Adjusting for prespecified covariates
 
+### Time window comparison
 Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.
@@
 
-Table \ref{tab:ppinames} compares the effects of the five drugs in this class.
-Omeprazole and lansoprazole were by far the most commonly prescribed drugs.
-The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day was higher for lansoprazole than for omeprazole.
-This suggests that the DDDs for lansoprazole and omeprazole (30 mg and 20 mg respectively) are not equivalent in potency.
+## Construction of a propensity score
+Table \ref{tab:numdrugs.protonpump} shows the association of ever-use of proton pump inhibitors with severe disease, stratified by number of other drugs dispensed.
+This shows that the risk ratio associated with ever-use of proton pump inhibitors was XX in those who had no dispensed prescriptions of other drug classes.
 
+As a next step, we constructed a propensity score for ever-use of a proton pump inhibitor. 
+The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
+As expected, the variables retained included NSAIDs and antiplatelet drugs, but also many other drug classes such as tricyclic analgesics that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
+The coefficients for age and sex were dropped before calculating the propensity score in the case-control dataset, because the case-control analysis is matched for age and sex.
+
+<!--- show effect of each drug class in propensity score before and after adjustment for PPI --->
+
+## Associations with other drug classes
 Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.
@@
 A strength of this study is that it is based on dispensed prescriptions, rather than issued prescriptions.
-However as proton pump inhibitors are available over the counter, our study of dispensed prescriptions does not capture all exposure to this class of drugs.
+Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.
 Unless this this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
-
 Although we do not have morbidity data from primary care, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
 This diagnostic coding is more relevant to disease classification than the SNOMED diagnostic codes used in primary care databases.
-
 Because this case-control study is matched on primary care practice, variation of prescribing between practices cannot confound the conditional odds ratios, though it is possible that variation between practitioners within the same practice could contribute to a confounding effect.
@@
 ## Conclusion
-Although some of the association of severe COVID-19 with dispensing of prescriptions can be accounted for underlying cardiovascular diseases that are not captured by hospital discharge diagnoses, there remains a residual association with prescription of proton pump inhibitors that is not easily explained by co-morbidity.
-Increased risk of community-acquired pneumonia is an established adverse effect of the use of these drugs, We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.
+Severe COVID-19 is strongly associated with the number of drug classes dispensed as prescriptions during the last year.
+This effect is not easily explained by co-morbidity, and it is strongest in relatively young individuals without hospital diagnoses of conditions that confer high risk of disease.
+The profile of drug classes associated with COVID-19 includes many drug classes that are used as indicators of overprescribing and inappropriate polypharmacy.
+The single drug class with the greatest explanatory power is proton pump inhibitors.
+Although this study did not identify unequivocal evidence of causality for this drug class, increased risk of infection is an established adverse effect of the use of these drugs.
+We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.
 
@@
 
-# Acknowledgements
+## Acknowledgements
 We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Registers of Scotland and the Public Health Scotland Terminology Services.
@@
 
-```{r cats3, echo=FALSE, warning=FALSE, message=FALSE}
-options(knitr.kable.NA = '')
+# Tables
 
-knitr::kable(table.scrip.cats3,
+```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.numdrugsgr.agegr,
              escape=FALSE,
              booktabs=TRUE,
-             label="scrip.cats3",
-             col.names=c(clean.header(colnames(table.scrip.cats3)[1:2]),
+             label="numdrugs.agegr",
+             row.names=FALSE,
+             col.names=c("Number of drugs", "Controls", "Cases", 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with encashment of any prescription in the last year, by residence status and diagnosis of a listed condition") %>%
-     kable_styling(latex_options="hold_position") %>%
-   column_spec(column=1, width="4cm")
+             caption="Association of severe disease with number of drug classes (BNF subparagraph codes) dispensed, by age group") %>%
+    group_rows("Age 0-59", 1, 6) %>%
+    group_rows("Age 60-74", 7, 12) %>%
+    group_rows("Age 75+", 13, 18) %>%
+    kable_styling(latex_options="hold_position")
+
 ```
 
-```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
+```{r numdrugs.cats3, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.drugs.nocare.notlisted) <- gsub("_", " ", rownames(table.drugs.nocare.notlisted))
-knitr::kable(table.drugs.nocare.notlisted,
+knitr::kable(table.numdrugsgr.cats3,
              escape=FALSE,
              booktabs=TRUE,
-             label="drugs",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.drugs.nocare.notlisted)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
+             label="numdrugs.cats3",
+             row.names=FALSE,
+             col.names=c("Number of drugs", "Controls", "Cases", 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in those not resident in care homes and without listed conditions") %>%
-    column_spec(column=1, width="4.5cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
+             caption="Association of severe disease with number of drug classes (BNF subparagraph codes) dispensed, by residence/health status") %>%
+    group_rows("Care home residence", 1, 6) %>%
+    group_rows("Living independently, listed condition", 7, 12) %>%
+    group_rows("Living independently, no listed condition", 13, 18) %>%
+    kable_styling(latex_options="hold_position")
+
 ```
@@
 
-```{r stepwise.chapters, echo=FALSE, warning=FALSE, message=FALSE}
+```{r numdrugs.cardiovasc, echo=FALSE, warning=FALSE, message=FALSE}
 
-stepwise.drop[, 1] <- gsub("\\_", " ", stepwise.drop[, 1])
-
-knitr::kable(stepwise.drop,
+knitr::kable(table.jointcardiovasc,
              escape=FALSE,
              booktabs=TRUE,
-             label="stepwise.drop",
-             col.names=c("Dropped BNF chapter", "Log rate ratio before dropping",
-                         "Change in log-likelihood when dropped"), 
-             digits=c(0, 2, 1), 
-             caption="Stepwise dropping of BNF chapters from the 'any dispensed prescription' variable: dataset restricted to those not in care homes and without listed conditions") %>%
-    column_spec(column=1, width="4cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="3cm") %>%
+             label="numdrugs.cardiovasc",
+             row.names=FALSE,
+             col.names=c("Drugs category", "Controls", "Cases", 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Joint model for associations of numbers of cardiovascular and non-cardiovascular drug classes dispensed,  by age group") %>%
+    add_header_above(c(" "=1, "Mean number of drug classes"=2, " "=2)) %>%
+    group_rows("Age 0-59", 1, 2) %>%
+    group_rows("Age 60-74", 3, 4) %>%
+    group_rows("Age 75+", 5, 6) %>%
     kable_styling(latex_options="hold_position")
@@
 
+\newpage
+
 ```{r stepwise.subparas, echo=FALSE, warning=FALSE, message=FALSE}
 
-stepwise.drop.subparas[, 1] <- gsub("subpara\\.", "", stepwise.drop.subparas[, 1])
-
-knitr::kable(stepwise.drop.subparas[-(1:29), ], 
+knitr::kable(stepwise.drop.subparas[rows.print, ], 
              escape=FALSE, 
@@
              digits=c(0, 2, 1), 
-             caption="Stepwise dropping of last ten drug classes from a variable defined as 'any dispensed prescription from BNF chapters 1 or 2': dataset restricted to those not in care homes and without listed conditions") %>%
-    column_spec(column=1, width="9cm") %>%
+             caption=paste("Stepwise dropping of last", length(rows.print), "drug classes from the variable defined as number of drug classes dispensed: dataset restricted to those not in care homes and without listed conditions")) %>%
+    column_spec(column=1, width="8cm") %>%
     column_spec(column=2, width="2cm") %>%
@@
 
-```{r propensity, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(coeffs.propensity,
-             escape=FALSE,
-             label="propensity", 
-             booktabs=TRUE,
-             digits=c(0, 3), 
-             col.names=c("BNF subparagraph", "log odds ratio"),
-             caption="Propensity score for prescribing of a proton pump inhibitor, constructed in an independent dataset") %>%
-             kable_styling(latex_options="hold_position")
-
-```
-
 ```{r everuse.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.dose.protonpump,
+knitr::kable(table.everuse.protonpump,
              escape=FALSE,
@@
              booktabs=TRUE,
-             col.names=c("Controls", "Cases",
+             col.names=c(clean.header(colnames(table.everuse.protonpump)[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value",
@@
              booktabs=TRUE,
-             col.names=c("Controls", "Cases",
+             col.names=c(clean.header(colnames(table.dose.protonpump)[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value",
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with average dispensed dose of proton pump inhibitor (DDDs/day) by age group, with and without adjustment for propensity score") %>%
+             caption="Associations of severe disease with dose (DDDs/day) of proton pump inhibitor by age group, with and without adjustment for propensity score") %>%
     add_header_above(c(" "=2, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
@@
     landscape()
+
+```
+
+```{r fatal.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.fatal.protonpump,
+             escape=FALSE,
+             label="fatal.protonpump", 
+             booktabs=TRUE,
+             col.names=c("Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of fatal COVID-19 with ever-use of proton pump inhibitor  by age group") %>%
+    kable_styling(latex_options="hold_position")
 
@@
              caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window")) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
+	column_spec(column=1, width="4cm") %>%
+	column_spec(column=2:3, width="2cm") %>%
+	column_spec(column=4:5, width="2.5cm") %>%
+		kable_styling(latex_options="hold_position")
+
+```
+
+```{r numdrugs.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.numdrugs.protonpump,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="numdrugs.protonpump", 
+             col.names=clean.header(colnames(table.numdrugs.protonpump)), 
+             caption="Numbers of other drugs classes dispensed to non-users and ever-users of proton pump inhibitors, among controls") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r bynumdrugs, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.protonpump.bynumdrugs[, 1:4],
+             escape=FALSE,
+             booktabs=TRUE,
+             label="bynumdrugs", 
+             col.names=c(clean.header(colnames(table.protonpump.bynumdrugs)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+                caption="Association of severe disease with ever-use of proton pump inhibitor, by number of other drugs dispensed") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r propensity, echo=FALSE, warning=FALSE, message=FALSE}
+chapternums <- gsub("\\..*", "", coeffs.propensity[, 1])
+chapternums <- floor(as.integer(chapternums) / 100000)
+space.at <- which(diff(chapternums) >0)
+linesep.str <- rep('', nrow(coeffs.propensity))
+linesep.str[space.at] <- '\\addlinespace'
+
+knitr::kable(coeffs.propensity,
+             escape=FALSE,
+             label="propensity", 
+             booktabs=TRUE,
+             digits=c(0, 3), 
+             col.names=c("BNF subparagraph", "log odds ratio"),
+			 linesep=linesep.str,
+             caption="Coefficients of propensity score for prescribing of a proton pump inhibitor, constructed in an independent dataset") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r propensitygr, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.propensitygr[, 1:4],
+             escape=FALSE,
+             booktabs=TRUE,
+             label="propensitygr", 
+             col.names=c(clean.header(colnames(table.propensitygr)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+                caption="Association of severe disease with propensity score for prescribing of a proton pump inhibitor") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r bypropensity, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.protonpump.bypropensity[, 1:4],
+             escape=FALSE,
+             booktabs=TRUE,
+             label="bypropensity", 
+             col.names=c(clean.header(colnames(table.protonpump.bypropensity)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+                caption="Association of severe disease with ever-use of proton pump inhibitor, by propensity score") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r timewin.protonpump.propensity0, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.timewindow.protonpump.propensity0,
+             escape=FALSE,
+             label="timewindow.protonpump.propensity0", 
+             booktabs=TRUE,
+             col.names=c("Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window in those with propensity score of zero")) %>%
+	column_spec(column=1, width="4cm") %>%
+	column_spec(column=2:3, width="2cm") %>%
+	column_spec(column=4:5, width="2.5cm") %>%
+		kable_styling(latex_options="hold_position")
 
@@
     add_header_above(c(" "=1, "Average daily dose (DDDs)"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
+    kable_styling(latex_options="hold_position")
 

### 2020-05-29 casecontrol.R

@@ -14,21 +14,18 @@ library(readxl)
 library(DescTools)
 library(icd.data)
 library(gam)
-# library(plyr) # ? needed to access .() function
 library(dplyr)
 
 registerDoParallel(cores=2)
 
 source("helperfunctions.R")
 
-scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
-
 old <- TRUE
-#old <- FALSE
+old <- FALSE
 
 if(old) {
     cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
-    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") # 842 records ? excluded
+    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") 
     procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
     scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
 } else {
@@ -57,14 +54,14 @@ if(old) {
         subset(bnf_paragraph_code=="0103050")
 }
 
-###################################################################
+#################### add fields to scrips ########################################
 
 ## for now, just keep id, paragraph code, paragraph_description
+
 scrips$bnf_paragraph_description <- as.factor(scrips$bnf_paragraph_description)
 cat("scrips object uses", object.size(scrips) * 1E-6, "MB\n")
 
 ## we have 7 digits on scrips, giving resolution to subpara level only
-
 length(table(substr(scrips$bnf_paragraph_code, 1, 2))) # chapter
 length(table(substr(scrips$bnf_paragraph_code, 1, 4))) # chapter, section
 length(table(substr(scrips$bnf_paragraph_code, 1, 6)))  # chapter, section, paragraph
@@ -79,7 +76,7 @@ scrips$paracode <- as.integer(substr(scrips$bnf_paragraph_code, 1, 6))
 scrips$chapternum[is.na(scrips$chapternum)] <- 14
 scrips$chapternum[scrips$chapternum > 14] <- 14
 
-################################################################################
+################### short names for ICD chapters ########################
 
 icdchapters <- data.frame(names(icd10_chapters),
                              t(matrix(as.character(unlist(icd10_chapters)), nrow=2)))
[…]

### 2020-05-29 helperfunctions.R

@@ -2,15 +2,15 @@
 ## helperfunctions for COVID analyses
 
 select.union <- function(x, y, stratum) {
-    ## x is matrix of indicator variables
-    ## select column of x to drop that maximizes loglik of clogit model using all other cols of x
+    ## select j as column of x to drop that maximizes loglik of
+    ## clogit model using rowSums([x, -j])
     loglik.dropcol <- numeric(ncol(x))
-    x.u <- as.integer(rowSums(x) > 0)
+    x.u <- rowSums(x) 
     model.full <- summary(clogit(y ~ x.u + strata(stratum)))
     beta.full <- model.full$coefficients[1, 1]
     loglik.full <- model.full$loglik[2]
     for(j in 1:ncol(x)) {
-        x.u <- as.integer(rowSums(x[, -j, drop=FALSE]) > 0)
+        x.u <- rowSums(x[, -j, drop=FALSE])
         loglik.dropcol[j] <- summary(clogit(y ~ x.u + strata(stratum)))$loglik[2]
     }
 
@@ -481,7 +481,8 @@ tabulate.icdchapter <- function(chnum, data=cc.severe, minrowsum=20) {
                 
                 ## this should reformat pvalues
                 table.icd.ch.aug <- combine.tables2(ftable=table.icd.ch, utable=univariate.icd.ch)
-                table.icd.ch.aug[, 4] <- pvalue.latex(table.icd.ch.aug[, 4])
+                #browser("tabulateicd")
+                # table.icd.ch.aug[, 4] <- pvalue.latex(table.icd.ch.aug[, 4])
                 print(table.icd.ch.aug) # without this line the first two cols are dropped?
             } else {
                 return(NULL)
@@ -498,11 +499,11 @@ combine.tables3 <- function(ftable, utable, mtable)  {# returns single table fro
     
     u.ci <- or.ci(utable[, 1], utable[, 3]) 
     u.pvalue <- signif(utable[, 5], 1)
-    u.pvalue <- pvalue.latex(u.pvalue)
+    u.pvalue <- as.character(pvalue.latex(u.pvalue))
     
     mult.ci <- or.ci(mtable[, 1], mtable[, 3])
     mult.pvalue <- signif(mtable[, 5], 1)
-    mult.pvalue <- pvalue.latex(mult.pvalue)
+    mult.pvalue <- as.character(pvalue.latex(mult.pvalue))
      
[…]

### 2020-05-29 incidencemortality.R

@@ -1,13 +1,16 @@
+
+scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
+
+# load("casefreqs.RData") ## national cumulative cases and deaths by sex and one year age group
+
 ####### incidence and mortality using national population estimates ######################
 
-case.freqs <- with(cc.severe[cc.severe$CASE==1, ], table(AGE, sex, exclude=NULL))
 case.freqs <- data.frame(Age=as.integer(rownames(case.freqs)),
                          Females=as.integer(case.freqs[, 1]),
                          Males=as.integer(case.freqs[, 2]))
 case.long <- reshape2::melt(case.freqs, id="Age")
 colnames(case.long) <- c("Age", "Sex", "Cases")
 
-death.freqs <- with(cc.severe[cc.severe$fatalcase==1, ], table(AGE, sex, exclude=NULL))
 death.freqs <- data.frame(Age=as.integer(rownames(death.freqs)),
                          Females=as.integer(death.freqs[, 1]),
                          Males=as.integer(death.freqs[, 2]))
@@ -29,7 +32,6 @@ y.cases <- cbind(as.integer(discrim$Cases), as.integer(discrim$Noncases))
 
 discrim$Survivors <- discrim$Population - discrim$Deaths
 y.deaths <- cbind(as.integer(discrim$Deaths), as.integer(discrim$Survivors))
-
 
 cases.model <- glm(formula=y.cases ~ Sex + Age, family="binomial", data=discrim)
 deaths.model <- glm(formula=y.deaths ~ Sex + Age, family="binomial", data=discrim)

### 2020-05-29 pharmaco.R

@@ -1,0 +1,448 @@
+## pharmacoepi paper
+
+####### derived variables  ##########################
+
+
+##############################################################################
+
+## tabulate rate ratios for each num drug group, by agegr 
+
+table.numdrugsgr.agegr <- NULL
+for(agegr in levels(cc.severe$agegr3)) {
+    x <- tabulate.freqs.regressions(varnames="numdrugsgr",
+                                    data=cc.severe[cc.severe$agegr3==agegr, ])[, 1:4]
+    colnames(x)[1:2] <- c("controls", "cases")
+    table.numdrugsgr.agegr <- rbind(table.numdrugsgr.agegr, x)
+}
+table.numdrugsgr.agegr <- data.frame(numdrugsgr=rep(levels(cc.severe$numdrugsgr), 3),
+                               table.numdrugsgr.agegr)
+
+## tabulate rate ratios for each num drug group by cats3 (carehome, listed conditions)
+
+table.numdrugsgr.cats3 <- NULL
+for(cat in levels(cc.severe$cats3)) {
+    x <- tabulate.freqs.regressions(varnames="numdrugsgr",
+                                    data=cc.severe[cc.severe$cats3==cat, ])[, 1:4]
+    colnames(x)[1:2] <- c("controls", "cases")
+    table.numdrugsgr.cats3 <- rbind(table.numdrugsgr.cats3, x)
+}
+table.numdrugsgr.cats3 <- data.frame(numdrugsgr=rep(levels(cc.severe$numdrugsgr), 3),
+                                     table.numdrugsgr.cats3)
+
+## fit joint model for num cardiovascular and num non-cardiovascular drugs
+
+tj <- NULL
+for(agegr in levels(cc.severe$agegr3)) {
+    x <- summary(clogit(formula=CASE ~ numdrugs.cardiovasc + numdrugs.notcardiovasc + strata(stratum),
+                        data=cc.severe[nocare.notlisted & cc.severe$agegr3==agegr, ]))$coefficients
+    tj <- rbind(tj, x)
+}
+
+tj <- as.data.frame(tj)
+tj$u.ci <- or.ci(tj[, 1],
+                                    tj[, 3])
+tj$pvalue <- pvalue.latex(tj[, 5])
+tj$drugscat <- rep(c("Cardiovascular", "Non-cardiovascular"), 3)
+
+means.cardio <- with(cc.severe[nocare.notlisted, ], tapply(numdrugs.cardiovasc, list(CASE, agegr3), mean)) 
[…]

### 2020-06-01 casecontrol.Rmd
updated pharmacoepi analyses

@@
 
-_Background_-- Why some people infected with SARS-CoV-2 develop severe COVID-19 disease and others do not is poorly understood.
-The objectives of this study were to identify risk factors for severe disease and to lay the basis for risk stratification based on demographic data and health records.
+_Background_--   The objectives of this study were to identify risk factors for severe  COVID-19 and to lay the basis for risk stratification based on demographic data and health records.
 
 _Methods_ -- The design was a matched case-control study.
-Cases were ascertained as all those with a positive nucleic acid test for SARS-CoV-2 in the national virology laboratory database.
-Those who had been admitted to a critical care unit or had died within 28 days of a positive test were classified as severe cases.
-For each case, ten controls matched for sex, age and primary care practice were selected from the population register.
-For cases and controls we extracted all diagnostic codes from the past five years of hospital discharge records, all drug codes from prescriptions encashed during the past nine months and we linked to the national diabetes register.
-Rate ratios for associations with severe COVID-19 were estimated by conditional logistic regression.
-The performance of predictive models for severe COVID-19 was estimated by cross-validation.
+Severe cases were all those with a positive nucleic acid test for SARS-CoV-2 in the national database who had entered a critical care unit or died within 28 days of the first positive test.
+Ten controls per case matched for sex, age and primary care practice were selected from the population register.
+All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.
+Rate ratios for severe COVID-19 were estimated by conditional logistic regression.
 
-_Findings_ -- In a logistic regression based on the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
+_Findings_ -- In a logistic regression using the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
 In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
@@
 
-`r freqs.all[nrow(freqs.all), 2]` of the severe cases and `r freqs.all[nrow(freqs.all), 1]` of the controls had at least one of these listed conditions.
-In those aged under 40 years the frequencies of any listed condition were `r gsub("\\\\", "", table.agegr[4, 2])` in cases and `r gsub("\\\\", "", table.agegr[4, 1])` in controls.
-The odds ratio for severe disease associated with encashment of at least one prescription in the past year was `r table.agegr.all[2, 3]` and the odds ratio associated with at least one hospital admission in the past five years was `r table.agegr.all[3, 3]`; these associations remained strong and significant after adjusting for the listed conditions.
-Most cases without listed conditions had either prescriptions or hospitalisations, and highly significant associations with severe disease were seen across many hospital diagnoses and drug categories.
+`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all[nrow(freqs.all), 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all[nrow(freqs.all), 1])` of controls had at least one listed condition (`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr[4, 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr[4, 1])` of controls under age 40).
+Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all[2, 3]`] and `r table.agegr.all[3, 3]` respectively] even after adjusting for the listed conditions.
+In those without listed conditions significant associations with severe disease were seen across many hospital diagnoses and drug categories.
+Age and sex provided `r round(cases.Lambda.agesex, 2)` bits of information for discrimination.
+A model based on demographic variables, listed conditions, hospital diagnoses and prescriptions provided an additional `r table.densities[3, 5]` bits (C-statistic `r table.densities[3, 3]`).
 
-A predictive  model based on age and sex alone had an expected information for discrimination of  `r round(cases.Lambda.agesex, 2)` bits.
-In the model conditional on age and sex further demographic variables provided another `r table.densities[1, 5]` bits,  and this increased to `r table.densities[2, 5]` bits  when listed conditions were included, and to `r table.densities[3, 5]` bits when indicator variables for hospital diagnoses and prescription encashments were also included.
-
-_Conclusions_ --Along with older age and male sex, severe COVID-19 disease is strongly associated with past medical history; most younger persons developing severe disease have an underlying medical condition.
-The comorbidities associated with increased risk are not limited to the lists of risk conditions designated by public health agencies.
-A risk classifier that uses all the information available in medical and prescribing records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the SARS-CoV-2 epidemic is over.
+_Conclusions_ -- Along with older age and male sex, severe COVID-19 is strongly associated with past medical history across all age groups.
+Many comorbidities beyond the risk conditions designated by public health agencies contribute to this.
+A risk classifier that uses all the information available in health records, rather than only a limited set of conditions, will more accurately discriminate between low-risk and high-risk individuals who may require shielding until the epidemic is over.
 
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.
-Recent cases of diabetes since the last update of the register were supplemented with  a small number new cases identified through discharge codes and drug codes.
-These new cases were added into the diabetes type Other/Unknown category.
+Cases of diabetes diagnosed since the last update of the register were identified through discharge codes and drug codes, and assigned to the diabetes type Other/Unknown category.
 
@@
 In a multivariate analysis of all chapters, most of these associations were weaker.
-The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular)).
+The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular).
 Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
@@
 Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.
-However we found that almost half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.
+However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.
 In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.
@@
 rownames(table.agegr) <- replace.names(rownames(table.agegr))
-#table.agegr[1, c(1, 3, 5, 7)] <- " "
 knitr::kable(table.agegr[-7, ], # omit non-diabetic category
@@
                        "60-74 years"=2, "75+ years"=2)) %>%
- #   kable_styling(latex_options="scale_down") %>%
     landscape() %>%

### 2020-06-01 pharmaco.Rmd
updated pharmacoepi analyses

@@
 ---
-title: 'Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study'
+title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study'
 author: 
@@
   - name: Sharon Kennedy
-    affiliation: HPS
+    affiliation: ISD
   - name: Amanda Weir 
@@
     affiliation: HPS
-  - name: Stuart McGurnaghan
+  - name: Stuart J McGurnaghan
     affiliation: IGMM
-  - name: David McAllister 
-    affiliation: Glasgow
-  - name: Chris Robertson 
-    affiliation: Strathclyde
+  - name: David McAllister\textsuperscript{\getAff{Glasgow}} 
+    affiliation: HPS
+  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}} 
+    affiliation: HPS
   - name: Rachael Wood
@@
     affiliation: HPS 
-  - name: Tom Caparrotta 
+  - name: Thomas M Caparrotta 
     affiliation: IGMM
-  - name: Sharon Hutchinson 
-    affiliation: GCU
+  - name: Alison Smith-Palmer 
+    affiliation: HPS
+  - name: David Goldberg
+    affiliation: HPS
+  - name: Jim McMenamin
+    affiliation: HPS
+  - name: Colin Ramsay
+    affiliation: HPS
+  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
+    affiliation: HPS
   - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
@@
 TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.
+
+  - code: HPS
+    address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+
   - code: Glasgow
-     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
-   - code: HPS
-     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+    address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Wellcome Trust Intermediate Clinical Fellow and Beit Fellow
+
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
+
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health
+
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.
+
 geometry: margin=1in
@@
 
-_Methods_ -- Test positive COVID-19 cases ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database.
+_Methods_ -- Severe COVID-19 cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database and entered a critical care unit or died within 28 days of testing.
 From the national population register ten controls per case were selected, matched for sex, age and general practice.
-Severe cases were those  who entered  a critical care unit or died within 28 days of testing.
-International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary in the past nine months and used to define  history of listed conditions.
-To examine if  any unexpected specific drug classes were likely to contribute to severe COVID19 risk, among  severe cases and controls without listed conditions  (INSERT n cases n controls) we undertook a backward selection procedure across all BNF chapters.
-Across the retained chapters we undertook a backward selection procedure at sub-paragraph level.
-This yielded proton pump inhibitors (PPI) as the drug class that most explained the association of a prescription in the past year with COVID-19.
-Accordingly we built a propensity score of PPI prescribing  across all 218 BNF subparagraph codes in a separate dataset of hospitalised non-severe cases and their controls.
-This score, which increases as polypharmacy increases, was then used as a covariate that allowed adjustment of the association of PPIs with severe COVID-19 across all severe cases and their controls for confounding by co-prescribing.
+International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary care during the past nine months.
 Rate ratios were estimated by conditional logistic regression.
 
-*Findings* -- The risk of severe disease was strongly associated with the total number of drug classes dispensed in the last nine months.
+*Findings* -- The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
+This led us to examine associations with polypharmacy, defined as the total number of drug classes dispensed.
+The rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+The association with polypharmacy was strongest in those without past hospital admissions or diagnoses of conditions previously designated as conferring high risk.
+A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.
+Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.
+The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.
+However a comparison of time windows representing recent and less recent exposure in ever-users did not show a stronger association with recent exposure.
 
-Across sub-paragraphs of these chapters identified proton pump inhibitors as the drug class with the strongest independent association with severe COVID-19.
-The rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[4, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 40 years.
-Adjustment for hospital diagnoses of disorders of the esophagus, stomach and duodenum reduced these effect sizes only slightly.
-In those under 60 years only, the risks ratios increased with increasing average dispensed daily dose.
-More recent exposure was not more strongly associated than earlier exposure.
-With adjustment for propensity score associations were no longer significant in those aged 75+ but remained strong INSERT in those aged under 60 years.
-Restricting to those with no other dispensed prescriptions, the rate ratio associated with ever-use of a proton pump inhibitor was `r table.protonpump.bynumdrugs[1, 3]`.
-
-*Conclusions* -- A systematic search for drug classes that underlie the association of severe COVID-19 with dispensing of any prescription has identified proton pump inhibitors as the drug class making the largest independent contribution to this association.
-Prescription of this drug class is recognised as a a sentinel for polypharmacy and for frailty and  the  data were consistent with this  explanation for the association in the older population.
-However the strength of the association in those under age 60, its independence of the propensity score and the observation of an association  in those not on any other drugs raise concern of a possible causal effect that requires further investigation.
-In the meantime   existing advice on the avoidance of  inappropriate and over-prescribing of PPIS should be reinforced.
+*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy.
+and this association is not easily explained by co-morbidity.
+Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.
+As prescriptions of this drug class are a recognised indicator for overprescribing of drugs without clear medical indication, it is difficult to distinguish between causality and confounding as explanations for this association.
+However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation.
+In the meantime existing advice on reducing inappropriate polypharmacy and specifically limiting the prescribing of proton pump inhibitors should be reinforced.
 
@@
 # Background
-The REACT-SCOT  study is a matched case control study of COVID-19 cases versus controls in Scotland.
-In our first paper from this study we confirmed and quantified the risk ratios associated with conditions  such as diabetes, asthma and heart disease that had been  presumptively listed by the authorities as high risk for COVID-19 based on case series.
-We further showed that in those without listed conditions prior hospitalisation in the previous five years  and prescription history  in the previous nine months  were strongly associated with severe disease.
-We went on to show that the association with hospitalisation was not explained by just one or two other specific conditions but that many conditions of varying prevalence were more common in cases than controls.
+The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.
+In the first report of this study we reported a strong association of severe COVID-19 with having had at least one dispensed prescription in the past year.
+The univariate rate ratios associated with at least one dispensed prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.
+This association persisted after adjusting for care home residence, any hospital admission in the last five yeasrs and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
+In a multivariate analysis adjusting for these covariates, the rate ratios five years or residence in a care home, the rate ratios associated with any dispensed prescription were 5.0 in those aged under 40 and 11 in those aged 75 years and over.
+Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in almost all BNF chapters were associated with elevated rate ratios for COVID-19.
+In a multivariate analysis of all chapters, most of these associations were weaker, but prescriptions of at least one drug in chapters 1 (gastrointestinal system), 2 (nervous system) and 9 (nutrition and blood) remained strongly associated with COVID-19.
+We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of specific drugs or direct effects of polypharmacy on susceptibility could not be ruled out.
 
-A striking preliminary finding of the REACT-SCOT study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
-The rate ratio associated with this variable varied from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
-In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]`
+# Aim
+The primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
+Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.
+A further aim was to determine whether tthe effects underlying this association are sparse or diffuse.
+In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.
+In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.
 
-There are several potential explanations for such a finding;  there may be a sparse set of  drugs underlying this association or  there may be a large set of drugs contributing to it.
-These in turn may reflect  a i)sparse set of direct drug effects or ii) sparse set of underlying morbidities or a iii) larger set of direct drug effects  or  their interactions ( i.e.
-an effect due to the polypharmacy itself)  or a iv) larger set of co-morbidities ( i.e.
-that usually conceptualised as frailty) or even v) interactions between polypharmacy and frailty.
-
-The objective of this study was to conduct an initial analysis to discover if there is a a sparse set of direct drug effects underlying the finding of the strong association of any prescription in the past year and severe COVID-19.
-Further analyses will explore the role of polypharmacy and frailty.
+# Background
+The REACT-SCOT study is a matched case-control study of severe COVID-19 in Scotland.
+In the first report from this study an unexpected result was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
+The rate ratio associated with this variable varied from 10 in those aged under 60 years to 40
+in those aged 75 years and over.
+The objective of this study was to establish how much of this association could be explained by confounding by co-morbidity and other health-related factors that influece drug prescribing and independently affect susceptibility to severe COVID-19, and how much by direct effects of prescribed drugs.
+A secondary objective was to determine how much of the association could be explained by a sparse set of drug classes or other health-related factors, and how much was attributable to multifactorial causes, possibly including interactions between these factors.
 
@@
 Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-For each test-positive case, the CHI database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the date of test in the case using the Community Health Index (CHI) database.
-With this incidence density sampling design, odds ratios in conditional logistic regression models are equivalent to rate ratios.
+For each test-positive case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation, based on the postal code.
 
 ## Morbidity and drug prescribing
-For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge.
+For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.
 
-British National Formulary (BNF) drug codes were extracted from all prescriptions dispensed since 1 June 2019, excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
+British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
 The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
@@
 Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.
-Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to 16 days before the date that the case (in a stratum consisting of case and matched controls) first tested positive.
+Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.
 
 # Statistical analysis 
+Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
 As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
@@
 We constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
-To construct this score, we used an independent dataset so that the score would not overfit the case-control dataset.
-As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a dataset of `nrow(cc.hosp)` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
-The propensity score was constructed on this training dataset by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
-
-Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
+As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
+By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.
+The propensity score was constructed by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
+As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.
 
 # Results
+## Univariate associations of severe COVID-19 with drug classes
+We first examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
+Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.
+These COVID-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
+This led us to examine associations with polypharmacy, indexed by the number of different drug classes for which prescriptions had been dispensed.
+
 ## Relation of severe COVID-19 to number of drug classes dispensed. 
@@
 
-As a next step, we partitioned the number of drug classes dispensed into totals for cardiovascular and non-cardiovascular drugs, on the basis that prescribing of multiple drugs for control of cardiovascular disease is supported by evidence-based guidelines, while prescribing of multiple drugs for other indications is more likely to represent inappropriate polypharmacy.
-Table \ref{numdrugs,cardiovasc} shows that numbers of drugs in both categories were associated with COVID-19.
+## Backward selection to identify specific drug classes contributing to the association of severe COVID-19 with number of drug classes dispensed
+A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with number of drugs dispensed.
+To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.
+Dropping the first `r firstrow.neg - 1` drug classes from this subset increased the log-likelihood of the model.
+Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
+Table \ref{tab:stepwise.subparas} shows how the log-likelihood decreased (though the rate ratio increased) as these variables were dropped from the model.
 
-We next estimated univariate rate ratios for severe COVID-19 for indicator variables for dispensing of each of the `r length(subparacols)` BNF subparagraph codes.
-Table XX shows the results for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.
-These COVID-associated drug classes include many drug classes previously identified in Scotland as indicators of overprescribing and inappropriate polypharmacy.
-
-## Backwards selection to identify specific drug classes contributing to the association of severe COVID-19 with number of drug classes dispensed
-A backwards stepwise variable selection procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with number of drugs dispensed.
-To save computation, the stepwise procedure was initiated with the the subset `r length(subparacols.forstepwisedrop)` drug classes that showed univariate associations.
-The results of this procedure are shown in Table \ref{tab:stepwise.subparas}.
-Dropping the first `r firstrow.neg - 1` drug classes from the set of drug classes used to compute the number of dispensed drug classes increased the log-likelihood of the model.
-Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-
-The only cardiovascular drug classes included among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
+The only cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
 Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.
@@
 ## Associations of severe COVID-19 with proton pump inhibitor use
-Testing for independent association of COVID-19 with exposure to proton pump inhibitors is challenging because these drugs are usually co-prescribed with other drugs, and because in this dataset the disease is strongly associated with polypharmacy.
+This unexpected finding led us to investigate the association of COVID-19 with exposure to proton pump inhibitors in more depth.
+This is challenging because proton pump inhibitors are usually co-prescribed with other drugs.
 We used several approaches:
 
-* testing for a dose-response relationship,
+* Testing for a dose-response relationship
 
-* examination of the relationship by age.
+* Examination of the relationship by age.
 
-* adjusting for prespecified covariates
+* Adjusting for prespecified covariates: care home residence, SIMD quintile, index of hospital diagnosis of a disorder of the esophagus, stomach or duodenum, ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
 
-* comparison of dispensing between recent and less time windows.
+* Comparison in ever-exposed individuals between recent and less recent time windows of exposure.
 
-* construction of a propensity score using an independent dataset.
+* Stratification by propensity score for ever-use of a proton pump inhibitor.
 
-For these analyses the dataset was not restricted to exclude care home residents or those with diagnosed with a listed condition.
+For these analyses the dataset was not restricted to exclude care home residents or those  diagnosed with a listed condition.
 
 ### Dose-response relationship
-Table \ref{tab:dosegr.protonpump} shows how the effect size varies with the dose of proton pump inibitor, grouped into categories.
+Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inibitor, expressed as average DDDs/day over the period for which prescribing was observed.
 Table \ref{tab:ppinames} compares the effects of the five drugs in this class.
 Omeprazole and lansoprazole were by far the most commonly prescribed drugs.
-The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day was higher for lansoprazole than for omeprazole.
+The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day, was higher for lansoprazole than for omeprazole.
 This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.
@@
 ### Relationship by age
-Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for propensity score.
-The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates.
+Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for prespecified covariates.
+The association was strongest in those aged under 60 years, and was reduced by adjusting for covariates.
+Table \ref{tab:fatal.protonpump} shows the univariate associations of ever-use with fatal COVID-19 by age group.
 
-Table \ref{tab:dose.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day since 1 June 2019.
-There is a strong dose-response relationship in the two younger age groups, but not in those
-aged over 75 years.
-
-### Adjusting for prespecified covariates
+Table \ref{tab:dosegr.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day during the observation period.
+There is a strong dose-response relationship in those aged less than 75 years, but not in those aged over 75 years.
 
@@
 
-## Construction of a propensity score
-Table \ref{tab:numdrugs.protonpump} shows the association of ever-use of proton pump inhibitors with severe disease, stratified by number of other drugs dispensed.
-This shows that the risk ratio associated with ever-use of proton pump inhibitors was XX in those who had no dispensed prescriptions of other drug classes.
-
-As a next step, we constructed a propensity score for ever-use of a proton pump inhibitor. 
+### Construction of a propensity score
+A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section.
 The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
-As expected, the variables retained included NSAIDs and antiplatelet drugs, but also many other drug classes such as tricyclic analgesics that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
-The coefficients for age and sex were dropped before calculating the propensity score in the case-control dataset, because the case-control analysis is matched for age and sex.
-
-<!--- show effect of each drug class in propensity score before and after adjustment for PPI --->
+As expected, the variables retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
+Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, the association of COVID-19 with use of proton pump inhibitors remained strong only in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.
+<!--- show coefficient for COVID of each drug class in propensity score before and after adjustment for PPI?
+--->
 
 ## Associations with other drug classes
+Tables \ref{tab:bnfchapter1} and \ref{tab:bnfchapter4} show associations with drug classes in BNF chapters 1 (gastrointestinal) and 4 (nervous system).
+These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 even though the indications for which they are prescribed are not conditions recognized as conferring high risk for severe disease.
 Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.
 Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.
+Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19.
+In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.
+The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`;  Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.
 
-We also examined the univariate and multivariate associations of prescribed drugs for the cardiovascular system (BNF chapter 2) with severe COVID-19.
-Table \ref{tab:bnfchapter2} shows associations with drug classes (BNF subparagraphs) in this chapter.
-Prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19.
-In a multivariate analysis the strongest associations were with loop diuretics and calcium-channel blockers.
-All commonly-used drug classes used to treat hypertension were associated with moderately elevated risk of severe disease, with rate ratios from ranging from about 1.4 to 2.2.
-The rate ratios associated with ACE inhibitors and angiotensin-II receptor blockers were in the middle of this range.
-Prescriptions of anticoagulant, antiplatelet and lipid-lowering drugs were associated with severe disease in univariate analyses, but these association were weaker in a multivariate analysis.
-
-Tables \ref{tab:bnfchapter10} and \ref{tab:bnfchapter9} show associations with drug classes in BNF chapters 10 and 9.
+Tables \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 2 (cardiovascular) and 10 (musculoskeletal).
+These chapters were selected as of interest because possible effects of drugs in these chapters -- angiotensin-converting enzyme (ACE) inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulation [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
+Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and angiotensin-converting enzyme inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.
+Table \ref{tab:bnfchapter10} shows associations with drugs disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code.
+All antirheumatic drugs were associated with elevated rate ratios for COVID-19.
+The univariate rate ratio associated with hydroxychloroquine sulfate was `r table.bnfchapter10[grep("Hydroxychloroquine", rownames(table.bnfchapter10)), 3]`; adjustment for ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.hydroxychlor.adjusted[1], coeff.hydroxychlor.adjusted[3])`.
 
 # Discussion 
-We have shown that the association of dispensed drug prescriptions with severe COVID-19 remains after excluding people who are at high risk for reasons that are associated with higher rates of prescribing; residence in care homes or a hospital diagnosis of a designated high-risk condition.
-We have shown further that this most of this association is accounted for by drugs prescribed for the gastrointestinal and cardiovascular systems.
-There are smaller independent contributions from drugs prescribed for the nervous and musculoskeletal systems.
+We have shown that severe COVID-19 is associated with dispensed prescriptions of drugs of many different classes.  
+These include drugs acting on the the cardiovascular system, where prescribing of multiple drugs is supported by evidence-based guidelines, but also drugs acting on the gastrointestinal and central nervous systems that are widely used as indicators of overprescribing and inappropriate polypharmacy in the National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
+This led us to examine associations of COVID-19 with polypharmacy, defined as the number of drugs classes dispensed during the period of observation.
+We attempted to determine how much of this association is attributable to confounding by multimorbidity, and how much is attributable to adverse effects of drug prescribing.
+Without more data on health status from primary care records, the confounders cannot be accurately measured.
+We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with common conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
+The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than those associated with designated conditions.
+In epidemiological experience rate ratios of this magnitude usually have a causal explanation.
 
-Of the drugs prescribed for the cardiovascular system, the strongest independent effect is that of loop diuretics.
-This is likely to reflect an underlying association of heart failure with high risk of severe or fatal COVID-19.
-Prescription of ACE inhibitors or angiotensin-II receptor blockers is associated with elevated risk of severe disease, but the rate ratios are no higher than for other antihypertensive agents.
-This is likely to reflect an underlying assocaition of hypertension with severe disease.
+We have shown that a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with number of drug classes dispensed.
+The association of COVID-19 with dispensed prescriptions of proton pump inhibitors shows a clear dose-response effect.
+Adjustment for covariates pre-specified as likely to confound this assocation -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly.
+The rate ratio associated with exposure to proton pump inhibitors is highest in those aged under 40, where comorbid conditions and frailty are less likely to confound this association.
+The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptiblity to COVID-19: antidepressants, analgesics and nutritional supplements.
+When the exposure under study is not measured accurately, including such variables in the model as possible confounders may over-adjust the effect of that exposure under study towards the null.
+By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small.
+
+One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to investigate the relationship of adverse events to the time of exposure among ever-users of the drug under study, as in a case-crossover design.
+We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.
+However most users of these drugs had prescriptions dispensed in both time windows.
+Users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.
+It is also possible that use of proton pump inhibitors has a persistent effect on susceptibility to infection through alteration of the gut microbiome.
+
+It is thus not possible at this stage definitively to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
+Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.
+For now the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.
+If this correctly estimates a causal effect, then in the Scottish population where 26\% of controls have used proton pump inhibitors about 45% of severe cases are attributable to exposure to proton pump inhibitors.
+Although the rate ratio associated with use at the time of exposure to coronavirus may be underestimated by misclassification of all those with a prescription in the last nine months as exposed, the attributable risk fraction for prescribed use of these drugs will not be underestimated.
+However this does not take into account possible over-the-counter purchases.
+
+We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.
+The strongest independent effect of any class of drug acting on the cardiovascular system is that of loop diuretics.
+This is likely to reflect confounding by heart failure as one of the main indications for use of loop diuretics.
+For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].
 A more detailed study of the relation of COVID-19 to cardiovascular disease and drugs prescribed for cardiovascular disease will be reported elsewhere.
+The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
+Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including heart disease and co-prescribing of proton pump inhibitors.
+The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be accurately estimated.
+As this drug is commonly used in combination with non-steroidal anti-inflammatory agents with which proton pump inhibitors are frequently co-prescribed, we would expect it to be associated with elevated risk in univariate analyses.
 
-Of all the prescribed drug classes in Chapters 1 and 2 of the BNF, the strongest association was with proton pump inhibitors.
-This association with proton pump inhibitors is not explained by association with hospital diagnoses of diseases of the upper gastrointestinal tract, or by coprescribing with non-steroidal anti-inflammatory or antiplatelet agents.
-The highest rate ratios associated with this exposure are in those aged under 40.
-
-The possible effect of proton pump inhibitors on risk of severe COVID-19 was not one of the hypotheses that we specified in advance when setting up this case-control study.
+A possible effect of proton pump inhibitors on risk of severe COVID-19 was not one of the hypotheses that we specified in advance when setting up this case-control study.
 With hindsight, perhaps it should have been.
 In a systematic review of adverse effects of long-term use of proton pump inhibitors, all eight studies included showed elevated risk of community-acquired pneumonia [@islam_adverse_2018].
-As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
-In the largest study, based on the UK General Practice Research Database from 1987 to 2002, elevated risk was restricted to those for whom prescribing of proton pump inhibitors had been initiated no more than 30 days before the date of diagnosis of pneumonia [@sarkar_proton-pump_2008].
-A possible explanation for this is that at least in the UK, only those in whom prescribing of these drugs has recently been initiated are likely to be taking them continuously.
+As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk or severity of infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
 
-We note that in epidemiological experience rate ratios of 5 or more, as seen in those aged under 40 years in this study, usually have a causal explanation.
-However other drugs prescribed for the gastrointestinal system are associated with increased risk of severe disease, at least in univariate analyses.
-These include drugs such as laxatives that, like proton pump inhibitors, are used as indicators of prescribing quality in the National Therapeutic Index [@macbride-stewart_national_2019].
-It is not possible at this stage to exclude the possibility that the association with proton pump inhibitors could be accounted for by underlying factors that lead to consultation for minor symptoms and prescribing of these drugs without a clear medical indication.
-In the UK and other countries proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].
+In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].
 A study in Aberdeen reported that 35% of patients admitted to acute geriatric units had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of co-morbidity [@jarchow-macdonald_prescribing_2013].
-
-We can calculate that in the Scottish population where 28\% of controls have used proton pump inhibitors and the rate ratio is 2.8, one-third of severe cases could be attributable to prescribing of these drugs if the association is causal.
-Although the rate ratio associated with use at the time of exposure to coronavirus may be underestimated because all those with a prescription in the last nine months are clasified as exposed, the attributable risk fraction will not be underestimated, at least not for prescribed use of these drugs.
-However this does not take into account possible over-the-counter purchases.
-
 It may be relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors has been reported previously.
 A study in China reported that proton pump inhibitors were prescribed to 37% of inpatients in a tertiary referral centre in Szechuan in 2016 [@luo_changes_2018].
-In Italy, a survey of nursing homses found that 46% of residents had been prescribed a proton pump inhibitor [@pasina_overuse_2020].
+In Italy, a survey of nursing homes found that 46% of residents had been prescribed a proton pump inhibitor [@pasina_overuse_2020].
 A Spanish drug-utilization study reported that 55% of hospital inpatients were presribed a proton pump inhibitor at discharge [@ramirez_overuse_2010].
@@
 This diagnostic coding is more relevant to disease classification than the SNOMED diagnostic codes used in primary care databases.
-Because this case-control study is matched on primary care practice, variation of prescribing between practices cannot confound the conditional odds ratios, though it is possible that variation between practitioners within the same practice could contribute to a confounding effect.
+Although this case-control study is matched on primary care practice, in Scotland there is ample variation in high-risk prescribing between practitioners in the same practice [@guthrie_bad_2015], allowing study of the relationship of drug prescribing to COVID-19.
 
@@
 Datasets were de-identified before analysis.
+The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance.
 
@@
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Registers of Scotland and the Public Health Scotland Terminology Services.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
 
 ## Public Health Scotland COVID-19 Health Protection Study Group
-AliceÂ Whettlock$^1$, Alison Smith-Palmer$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, Dave Yirrell$^1$, David Henderson$^1$, DiogoÂ Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, JimÂ McMenamim$^1$, Joseph Jasperse$^1$,  Josie Murray$^1$,  KarenÂ Heatlie$^1$,  Lindsay Mathie$^1$,  Lorraine Donaldson$^1$, Martin Paton$^1$,  Martin Reid$^1$,  Melissa Llano$^1$,  Michael Lockhart$^1$,  Michelle Murphy-HallÂ $^1$,  Paul Smith$^1$,  Ros Hall$^1$,  Ross Cameron$^1$,  Susan Brownlie$^1$,  Adam Gaffney$^2$,  Aynsley Milne$^2$,  Christopher Sullivan$^2$,  Edward McArdle$^2$,  Elaine Glass$^2$,  Johanna Young$^2$,  William Malcolm$^2$,  Kate Templeton$^3$,  Nazir Lone$^4$,  Rory Gunson$^4$
+Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
 
@@
 
+```{r subpara.coeffs, echo=FALSE, warning=FALSE, message=FALSE}
+
+newch <- diff(as.integer(substr(subpara.coeffs$subpara, 1, 1))) !=0
+linesep <- ifelse(newch, '\\addlinespace', '')
+
+knitr::kable(subpara.coeffs,
+             escape=FALSE,
+             longtable=TRUE, 
+             booktabs=TRUE,
+             linesep = linesep, 
+             label="subpara.coeffs",
+             row.names=FALSE,
+col.names=c("BNF subparagraph", colnames(subpara.coeffs)[2:3],  
+            "Rate ratio", "$p$-value"),
+             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in a care home and without a listed condition, and filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
+    column_spec(column=1, width="12cm") %>%
+    column_spec(column=2, width="1.5cm") %>%
+    column_spec(column=3, width="1.5cm") %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+
+```
+
 ```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe disease with number of drug classes (BNF subparagraph codes) dispensed, by age group") %>%
+             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group") %>%
     group_rows("Age 0-59", 1, 6) %>%
@@
              row.names=FALSE,
-             col.names=c("Number of drugs", "Controls", "Cases", 
+             col.names=c("Number of drug classes", "Controls", "Cases", 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe disease with number of drug classes (BNF subparagraph codes) dispensed, by residence/health status") %>%
+             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by residence/health status") %>%
     group_rows("Care home residence", 1, 6) %>%
@@
 
-\newpage
+```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
 
-```{r numdrugs.cardiovasc, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.jointcardiovasc,
+knitr::kable(table.numdrugsgr.num.icdch,
              escape=FALSE,
              booktabs=TRUE,
-             label="numdrugs.cardiovasc",
+             label="numdrugs.num.icdch",
              row.names=FALSE,
-             col.names=c("Drugs category", "Controls", "Cases", 
+             col.names=c("Number of drug classes", "Controls", "Cases", 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Joint model for associations of numbers of cardiovascular and non-cardiovascular drug classes dispensed,  by age group") %>%
-    add_header_above(c(" "=1, "Mean number of drug classes"=2, " "=2)) %>%
-    group_rows("Age 0-59", 1, 2) %>%
-    group_rows("Age 60-74", 3, 4) %>%
-    group_rows("Age 75+", 5, 6) %>%
+             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by number of ICD-10 chapters represented in hospital diagnoses") %>%
+    group_rows("No admissions", 1, 6) %>%
+    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
+    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
     kable_styling(latex_options="hold_position")
@@
 ```{r stepwise.subparas, echo=FALSE, warning=FALSE, message=FALSE}
-
 knitr::kable(stepwise.drop.subparas[rows.print, ], 
@@
 
+```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.dosegr.protonpump,
+             escape=FALSE,
+             label="dosegr.protonpump", 
+             booktabs=TRUE,
+             row.names=FALSE, # dose levels in column 1
+             col.names=c("Average DDDs/day", "Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    group_rows("Ages 0-59 years", 1, 5) %>%
+    group_rows("Ages 60-74 years", 6, 10) %>%
+    group_rows("Ages 75+ years", 11, 15) %>%
+    landscape()
+
+```
+
+```{r ppinames, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.ppinames) <- tolower.exceptfirstchar(rownames(table.ppinames))
+
+knitr::kable(table.ppinames[, 1:4],
+             escape=FALSE,
+             label="ppinames", 
+             booktabs=TRUE,
+             col.names=c("Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Univariate associations of severe COVID-19 with dispensed average daily dose (DDDs/day) for each drug in the proton pump inhibitor class") %>%
+    add_header_above(c(" "=1, "Average daily dose (DDDs)"=2)) %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
 ```{r everuse.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with ever-use of proton pump inhibitor by age group, with and without adjustment for propensity score") %>%
-    add_header_above(c(" "=2, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```
-
-```{r dose.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.dose.protonpump,
-             escape=FALSE,
-             label="dose.protonpump", 
-             booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.dose.protonpump)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with dose (DDDs/day) of proton pump inhibitor by age group, with and without adjustment for propensity score") %>%
-    add_header_above(c(" "=2, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
+             caption="Associations of severe COVID-19 with ever-use of proton pump inhibitor by age group, with and without adjustment for prespecified covariates") %>%
+    add_header_above(c(" "=2, "Univariate"=2, "Adjusted for covariates"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
@@
 
-```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.dosegr.protonpump,
-             escape=FALSE,
-             label="dosegr.protonpump", 
-             booktabs=TRUE,
-             row.names=FALSE, # dose levels in column 1
-             col.names=c("Average DDDs/day", "Controls", "Cases",
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe disease with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for propensity score") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for propensity score"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    group_rows("Ages 0-59 years", 1, 5) %>%
-    group_rows("Ages 60-74 years", 6, 10) %>%
-    group_rows("Ages 75+ years", 11, 15) %>%
-    landscape()
-
-```
-
 ```{r timewin.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
@@
              caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window")) %>%
-	column_spec(column=1, width="4cm") %>%
+	column_spec(column=1, width="3cm") %>%
 	column_spec(column=2:3, width="2cm") %>%
-	column_spec(column=4:5, width="2.5cm") %>%
-		kable_styling(latex_options="hold_position")
-
-```
-
-```{r numdrugs.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.numdrugs.protonpump,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.protonpump", 
-             col.names=clean.header(colnames(table.numdrugs.protonpump)), 
-             caption="Numbers of other drugs classes dispensed to non-users and ever-users of proton pump inhibitors, among controls") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r bynumdrugs, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.protonpump.bynumdrugs[, 1:4],
-             escape=FALSE,
-             booktabs=TRUE,
-             label="bynumdrugs", 
-             col.names=c(clean.header(colnames(table.protonpump.bynumdrugs)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-                caption="Association of severe disease with ever-use of proton pump inhibitor, by number of other drugs dispensed") %>%
+	column_spec(column=4, width="2.5cm") %>%
+	column_spec(column=5, width="1.5cm") %>%
 		kable_styling(latex_options="hold_position")
@@
 
-```{r propensitygr, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.propensitygr[, 1:4],
-             escape=FALSE,
-             booktabs=TRUE,
-             label="propensitygr", 
-             col.names=c(clean.header(colnames(table.propensitygr)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-                caption="Association of severe disease with propensity score for prescribing of a proton pump inhibitor") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
 ```{r bypropensity, echo=FALSE, warning=FALSE, message=FALSE}
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-                caption="Association of severe disease with ever-use of proton pump inhibitor, by propensity score") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r timewin.protonpump.propensity0, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.timewindow.protonpump.propensity0,
-             escape=FALSE,
-             label="timewindow.protonpump.propensity0", 
-             booktabs=TRUE,
-             col.names=c("Controls", "Cases",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window in those with propensity score of zero")) %>%
-	column_spec(column=1, width="4cm") %>%
-	column_spec(column=2:3, width="2cm") %>%
-	column_spec(column=4:5, width="2.5cm") %>%
-		kable_styling(latex_options="hold_position")
-
-```
-
-```{r ppinames, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.ppinames[, 1:4],
-             escape=FALSE,
-             label="ppinames", 
-             booktabs=TRUE,
-             col.names=c("Controls", "Cases",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Univariate associations of severe disease with dispensed average daily dose (DDDs/day) for each drug in the proton pump inhibitor class") %>%
-    add_header_above(c(" "=1, "Average daily dose (DDDs)"=2)) %>%
+                caption="Association of severe COVID-19 with ever-use of proton pump inhibitor, by propensity score") %>%
     kable_styling(latex_options="hold_position")
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those not resident in a care home and without a listed condition") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 1 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those not resident in a care home and without a listed condition") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
 
-```{r bnfchapter9, echo=FALSE, warning=FALSE, message=FALSE}
+```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.bnfchapter9,
+knitr::kable(table.bnfchapter4,
              escape=FALSE, 
              booktabs=TRUE,
-             label="bnfchapter9",
-             col.names=c(clean.header(colnames(table.bnfchapter9)[1:2]),
+             label="bnfchapter4",
+             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value",
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 9 in those not resident in a care home and without a listed condition") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 4  by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
 ```{r bnfchapter10, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.bnfchapter10) <- tolower.exceptfirstchar(rownames(table.bnfchapter10))
+rownames(table.bnfchapter10) <- gsub("\\.", " ", rownames(table.bnfchapter10))
 
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with prescribed drugs in BNF chapter 10 in those not resident in a care home and without a listed condition") %>%
-    column_spec(column=1, width="4cm") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name, filtered to retain rows with at least 50 exposed individuals") %>%
+    column_spec(column=1, width="5cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
     landscape()
-```
+``````

### 2020-06-01 casecontrol.R

@@ -31,10 +31,11 @@ if(old) {
 } else {
     cc.all <- readRDS("./data/CC_linked_ANON_20200515.rds")
     diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
-    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds") 
-    scrips <- readRDS("./data/CC_PIS_x15_ANON_20200515.rds")[, c("ANON_ID",
-                                                           "bnf_paragraph_code",
-                                                           "bnf_paragraph_description")] 
+    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds")
+    scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
+    scrips <- readRDS(scrips.filename)[, c("ANON_ID", "bnf_paragraph_code",
+                                           "bnf_paragraph_description")] 
+
     ## scrips should be 2 tables to save space
     ## one record per scrip
     scripvars <- c("ANON_ID", "dispensed_date", "bnf_paragraph_code",
@@ -50,7 +51,7 @@ if(old) {
     ## data/CC_PIS_15_ANON_20200515.rds
 
     scrips.protonpump <-
-        readRDS("./data/CC_PIS_x15_ANON_20200515.rds") %>%
+        readRDS(scrips.filename) %>%
         subset(bnf_paragraph_code=="0103050")
 }
 
@@ -426,7 +427,19 @@ cc.all$antipsychotic <- as.factor(as.integer(cc.all$ANON_ID %in%
 
 ids.osmotic.laxative <- unique(scrips$ANON_ID[as.integer(substr(scrips$bnf_paragraph_code, 1, 6)) == 10604])
 cc.all$osmotic.laxative <- as.factor(as.integer(cc.all$ANON_ID %in%
-                                                   ids.osmotic.laxative))
+                                                ids.osmotic.laxative))
+
+ids.anticoagulant.any <-
+    unique(scrips$ANON_ID[scrips$bnf_paragraph_code == "0208010" |
+                          scrips$bnf_paragraph_code == "0208020"])
+cc.all$anticoagulant.any <- as.factor(as.integer(cc.all$ANON_ID %in%
+                                                ids.anticoagulant.any))
+
+ids.hydroxychloroquine <- readRDS(scrips.filename) %>%
[…]

### 2020-06-01 comorbidity.R

@@ -70,6 +70,7 @@ ids.icd.neuro <- unique(diagnoses$ANON_ID[grep("^F03|^G[1236789]", diagnoses$ICD
 ids.bnf.neuro <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 409 |
                                          as.integer(scrips$sectioncode) == 411])
 table(ids.bnf.neuro %in% ids.icd.neuro)
+## these drugs listed by HPS pharmacist as used for multiple sclerosis
 ## interferon beta 080204M, Glatiramer acetate 0802040U0, Natalizumab 0802040W0
 ## Dimethyl fumar 0802040AK, Teriflunomide 0802040AL, Alemtuzumab 0802030
 ## no records in scrips for these drugs

### 2020-06-01 helperfunctions.R

@@ -1,6 +1,12 @@
 
 ## helperfunctions for COVID analyses
 
+tolower.exceptfirstchar <- function(x) {
+    lowercase.str <- substr(tolower(x), 2, nchar(x))
+    x <- paste0(substr(x, 1, 1), lowercase.str)
+    return(x)
+}
+
 select.union <- function(x, y, stratum) {
     ## select j as column of x to drop that maximizes loglik of
     ## clogit model using rowSums([x, -j])
@@ -390,14 +396,16 @@ tabulate.bnfsubparas <- function(chnum, outcome="CASE", data=cc.severe, minrowsu
     }
 }
 
-tabulate.bnfchemicals <- function(chnum, outcome="CASE", data=cc.severe) {
-    ## get chemicalcode in wide format,one col per subchapter
-    ## FIXME: we do not have the chemical code,  is misnamed bnf_paragraph_code
-    scrips.ch.wide <- reshape2::dcast(scrips[scrips$chapternum==chnum, ],
-                                         ANON_ID ~ bnf_paragraph_code, fun.aggregate=length,
-                                         value.var="bnf_paragraph_code")
-    colnames(scrips.ch.wide)[-1] <- paste0("chemical.",
-                                              as.integer(colnames(scrips.ch.wide)[-1]))
+tabulate.bnfchemicals <- function(chnum, outcome="CASE", data=cc.severe, minrowsum=50) {
+    ## get drug names in wide format,one col per approved_name
+    chnum.str <- sprintf("%02d", chnum)
+    scrips.ch <- readRDS(scrips.filename) %>%
+        subset(substr(bnf_paragraph_code, 1, 2)==chnum.str)
+    scrips.ch.wide <- reshape2::dcast(scrips.ch,
+                                         ANON_ID ~ approved_name, fun.aggregate=length,
+                                         value.var="approved_name")
+    #colnames(scrips.ch.wide)[-1] <- paste0("chemical.",
+    #                                          as.integer(colnames(scrips.ch.wide)[-1]))
     ## drop rare chemicalgraphcodes
     scrips.ch.wide <- scrips.ch.wide[, colSums(scrips.ch.wide) > 20]
     
@@ -405,23 +413,22 @@ tabulate.bnfchemicals <- function(chnum, outcome="CASE", data=cc.severe) {
                        scrips.ch.wide,
[…]

### 2020-06-01 pharmaco.R

@@ -3,9 +3,77 @@
 ####### derived variables  ##########################
 
 
+
+########################################################
+
+## tabulate para or subpara codes in BNF chapters of interest
+table.bnfchapter1 <- tabulate.bnfparas(chnum=1, data=cc.severe, minrowsum=50)
+table.bnfchapter2 <- tabulate.bnfsubparas(chnum=2, data=cc.severe, minrowsum=50)
+table.bnfchapter4 <- tabulate.bnfsubparas(chnum=4, data=cc.severe, minrowsum=50)
+## chapter 10 has to be disaggregated to drug names as grouping of DMARDs is too broad
+table.bnfchapter10 <- tabulate.bnfchemicals(chnum=10, data=cc.severe, minrowsum=50)
+
+#############################
+
+coeff.anticoagulant.univariate <- summary(clogit(formula=CASE ~ anticoagulant.any + strata(stratum),
+               data=cc.severe))$coefficients[1, ]
+
+coeff.anticoagulant.adjusted <- summary(clogit(formula=CASE ~ anticoagulant.any +  heart.other.any + protonpump + strata(stratum),
+               data=cc.severe))$coefficients[1, ]
+
+############################
+
+coeff.hydroxychlor.univariate <- summary(clogit(formula=CASE ~ hydroxychloroquine + strata(stratum),
+               data=cc.severe))$coefficients
+
+coeff.hydroxychlor.adjusted <- summary(clogit(formula=CASE ~ hydroxychloroquine + protonpump + strata(stratum),
+               data=cc.severe))$coefficients[1, ]
+
+
 ##############################################################################
 
-## tabulate rate ratios for each num drug group, by agegr 
+## fit conditional logistic regression model for CASE ~ care.home + numdrugs
+cc.severe$numdrugs.sq <- cc.severe$numdrugs.subpara^2
+cc.severe.nonmissing <- cc.severe[nonmissing.obs(cc.severe, "SIMD.quintile"), ]
+
+multi.formula <- as.formula(paste("CASE ~ care.home + SIMD.quintile + diag.any +",
+                                  paste(listed.conditions, collapse="+"),
+                                  "+ numdrugsgr + numdrugs.cardiovasc + strata(stratum)"))
+
+model.numdrugs <- clogit(formula=multi.formula, data=cc.severe.nonmissing)
+print(summary(model.numdrugs)$coefficients)
+
+theta <- model.numdrugs$linear.predictors
[…]

### 2020-06-01 propensity.R

@@ -31,8 +31,6 @@ stepwise.full <- step(full.model,
                        scope=list(lower=lower.formula, upper=upper.formula),
                       direction="both", method="approximate", trace=-1)
 cat("done\n")
-rm(full.model)
-#rm(train.data)
 
 ## get coeffs, dropping age and sex
 coeffs.full <- summary(stepwise.full)$coefficients[, 1][-(1:3)]
@@ -51,3 +49,8 @@ propensity <- newdata.x %*% matrix(coeffs.full, ncol=1)
 
 coeffs.propensity <- data.frame(subpara=gsub("subpara\\.", "", names(coeffs.full)),
                                 coefficient=as.numeric(coeffs.full))
+
+rm(newdata.x)
+rm(full.model)
+rm(stepwise.full)
+rm(train.data)

### 2020-06-03 pharmaco.Rmd
updated pharmaco manuscript

@@
 
-_Background_-- We previously reported that dispensing of at least one prescription from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as high risk (listed conditions).
-Here we conducted an initial investigation of possible explanations for this association which include  specific co-morbidity or  drug effects but also include more diffuse frailty or polypharmacy effects.
+_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as high risk (listed conditions).
+We report a further investigation of possible explanations for this association, which include specific co-morbidity or drug effects but also more diffuse effects of frailty or polypharmacy.
 
@@
 
-*Findings* -- The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
-This led us to examine associations with polypharmacy, defined as the total number of drug classes dispensed.
-The rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+*Findings* -- We began by examining associations with polypharmacy, defined as total number of drug classes dispensed.
+The rate ratios (95% CIs) for severe COVID-19 associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
 The association with polypharmacy was strongest in those without past hospital admissions or diagnoses of conditions previously designated as conferring high risk.
+The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without clear medical indications.
 A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.
@@
 *Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy.
-and this association is not easily explained by co-morbidity.
+This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.
 Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.
-As prescriptions of this drug class are a recognised indicator for overprescribing of drugs without clear medical indication, it is difficult to distinguish between causality and confounding as explanations for this association.
-However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation.
-In the meantime existing advice on reducing inappropriate polypharmacy and specifically limiting the prescribing of proton pump inhibitors should be reinforced.
+As prescriptions of this drug class are a recognised indicator for overprescribing of drugs, it is difficult to distinguish between causality and confounding as explanations for this association.
+However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation in other settings.
+In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.
 
@@
 The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.
-In the first report of this study we reported a strong association of severe COVID-19 with having had at least one dispensed prescription in the past year.
-The univariate rate ratios associated with at least one dispensed prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.
+In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year.
+The univariate rate ratios associated with at least one prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.
 This association persisted after adjusting for care home residence, any hospital admission in the last five yeasrs and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
-In a multivariate analysis adjusting for these covariates, the rate ratios five years or residence in a care home, the rate ratios associated with any dispensed prescription were 5.0 in those aged under 40 and 11 in those aged 75 years and over.
 Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in almost all BNF chapters were associated with elevated rate ratios for COVID-19.
-In a multivariate analysis of all chapters, most of these associations were weaker, but prescriptions of at least one drug in chapters 1 (gastrointestinal system), 2 (nervous system) and 9 (nutrition and blood) remained strongly associated with COVID-19.
-We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of specific drugs or direct effects of polypharmacy on susceptibility could not be ruled out.
+We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of drugs on disease susceptibility could not be ruled out.
 
@@
 
-# Background
-The REACT-SCOT study is a matched case-control study of severe COVID-19 in Scotland.
-In the first report from this study an unexpected result was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
-The rate ratio associated with this variable varied from 10 in those aged under 60 years to 40
-in those aged 75 years and over.
-The objective of this study was to establish how much of this association could be explained by confounding by co-morbidity and other health-related factors that influece drug prescribing and independently affect susceptibility to severe COVID-19, and how much by direct effects of prescribed drugs.
-A secondary objective was to determine how much of the association could be explained by a sparse set of drug classes or other health-related factors, and how much was attributable to multifactorial causes, possibly including interactions between these factors.
-
 # Methods
 The design of the REACT-SCOT case-control study is described in detail in the accompanying paper.
-
-## Case ascertainment and sampling of controls 
 All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
@@
 For each test-positive case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
-Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation, based on the postal code.
+Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation (SIMD), based on the postal code.
 
@@
 For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.
-
 British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
@@
 As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-For this analysis the list was extended to include non-cardiac circulatory disease as our analyses of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.
-These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that condition is the only indication.
+For this analysis the list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.
+These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an dispensed prescription of a drug for which that condition is the only indication.
 Diagnoses of diabetes were extracted from linkage to the national diabetes register.
@@
 
-We constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
+To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
 As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
 By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.
-The propensity score was constructed by stepwise logistic regression with ever-use of a proton pump inhibitor as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
+The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
+`r nrow(coeffs.propensity)` drug classes were retained in the propensity score.
 As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.
@@
 # Results
+## Relation of severe COVID-19 to number of drug classes dispensed. 
+We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).
+Table \ref{tab:numdrugsgr.agegr} shows that risk of severe COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+
+Table \ref{tab:numdrugs.num.icdch} shows that this association was strongest in those who had no hospital diagnoses (because they had not been admitted to hospital).
+In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.
+Table \ref{tab:numdrugs.listed.any} shows that this association was strongest in those without a diagnosis of a listed condition.
+
+To study the relationship further, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.
+The rationale for this restriction was to reduce the effect of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and severe COVID-19.
+In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.
+Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.
+
 ## Univariate associations of severe COVID-19 with drug classes
-We first examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
-Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed for at least 10 individuals and were associated with severe disease at $p < 0.001$.
+We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
+Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.
 These COVID-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
-This led us to examine associations with polypharmacy, indexed by the number of different drug classes for which prescriptions had been dispensed.
-
-## Relation of severe COVID-19 to number of drug classes dispensed. 
-Table \ref{tab:numdrugs.agegr} shows that severe COVID-19 was strongly related to the total number of different drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed since 1 June 2019.
-The strength of this association was present in all age groups.
-With no dispensed prescriptions as reference category, the rate ratios associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
-
-Table \ref{tab:numdrugs.cats3} shows this relationship after stratifying by care home residence and diagnosis of at least one listed condition.
-This shows that the association of severe disease with number of drug classes dispensed was weak in care home residents, absent in those living independently with a diagnosed listed condition, and very strong in those who were living independently without a listed condition.
-In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.cats3[3, 11]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.cats3[7, 11]` in those with more than 12 drug classes dispensed.
-To examine this association further, we restricted the analysis to those not resident in a care home and without listed conditions so as to lessen confounding by underlying disease states and by frailty.
 
@@
 
-* Examination of the relationship by age.
-
 * Adjusting for prespecified covariates: care home residence, SIMD quintile, index of hospital diagnosis of a disorder of the esophagus, stomach or duodenum, ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
 
-* Comparison in ever-exposed individuals between recent and less recent time windows of exposure.
+* Comparison between recent and less recent time windows of exposure.
 
@@
 
-### Dose-response relationship
-Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inibitor, expressed as average DDDs/day over the period for which prescribing was observed.
-Table \ref{tab:ppinames} compares the effects of the five drugs in this class.
-Omeprazole and lansoprazole were by far the most commonly prescribed drugs.
-The effect size, expressed as rate ratio for unit increase in dose expressed as average DDDs/day, was higher for lansoprazole than for omeprazole.
+Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed.
+This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates.
+Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.
+The effect size for unit increase in dose expressed as average DDDs/day was higher for lansoprazole [rate ratio `r table.ppinames[2, 3]`] than for omeprazole rate [rate ratio `r table.ppinames[3, 3]`].
 This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.
 
-### Relationship by age
-Table \ref{tab:everuse.protonpump} shows the association with ever-use of proton pump inhibitors by age group, without and with adjustment for prespecified covariates.
-The association was strongest in those aged under 60 years, and was reduced by adjusting for covariates.
-Table \ref{tab:fatal.protonpump} shows the univariate associations of ever-use with fatal COVID-19 by age group.
-
-Table \ref{tab:dosegr.protonpump} shows the rate ratio per unit increase in dose of proton pump inibitor, calculated as average defined daily doses dispensed per day during the observation period.
-There is a strong dose-response relationship in those aged less than 75 years, but not in those aged over 75 years.
-
-### Time window comparison
 Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.
-time window.
 The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.
 
-### Construction of a propensity score
 A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section.
-The `r nrow(coeffs.propensity)` variables retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
-As expected, the variables retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
-Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, the association of COVID-19 with use of proton pump inhibitors remained strong only in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.
-<!--- show coefficient for COVID of each drug class in propensity score before and after adjustment for PPI?
---->
+The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
+As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
+Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.
 
 ## Associations with other drug classes
-Tables \ref{tab:bnfchapter1} and \ref{tab:bnfchapter4} show associations with drug classes in BNF chapters 1 (gastrointestinal) and 4 (nervous system).
-These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 even though the indications for which they are prescribed are not conditions recognized as conferring high risk for severe disease.
+Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).
+These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- angiotensin-converting enzyme (ACE) inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
+
 Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.
 Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.
+
 Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19.
 In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.
+Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and angiotensin-converting enzyme inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.
 The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.
 
-Tables \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 2 (cardiovascular) and 10 (musculoskeletal).
-These chapters were selected as of interest because possible effects of drugs in these chapters -- angiotensin-converting enzyme (ACE) inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulation [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
-Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and angiotensin-converting enzyme inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.
+Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with drugs widely used in care homes: antipsychotics, analgesics and drugs for dementia.
+
 Table \ref{tab:bnfchapter10} shows associations with drugs disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code.
@@
 We have shown that severe COVID-19 is associated with dispensed prescriptions of drugs of many different classes.
-These include drugs acting on the the cardiovascular system, where prescribing of multiple drugs is supported by evidence-based guidelines, but also drugs acting on the gastrointestinal and central nervous systems that are widely used as indicators of overprescribing and inappropriate polypharmacy in the National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
+These include drugs acting on the the cardiovascular system, where prescribing of multiple drugs is supported by evidence-based guidelines, but also drugs acting on the gastrointestinal and central nervous systems that are widely used as indicators of overprescribing and inappropriate polypharmacy in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
 This led us to examine associations of COVID-19 with polypharmacy, defined as the number of drugs classes dispensed during the period of observation.
@@
 We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with common conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
-The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than those associated with designated conditions.
-In epidemiological experience rate ratios of this magnitude usually have a causal explanation.
+The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of about 3 for severe COVID-19 that are associated with designated conditions.
 
@@
 Adjustment for covariates pre-specified as likely to confound this assocation -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly.
-The rate ratio associated with exposure to proton pump inhibitors is highest in those aged under 40, where comorbid conditions and frailty are less likely to confound this association.
+The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.
 The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptiblity to COVID-19: antidepressants, analgesics and nutritional supplements.
@@
 By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small.
-
-One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to investigate the relationship of adverse events to the time of exposure among ever-users of the drug under study, as in a case-crossover design.
+One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to investigate the relationship of adverse events to the time of exposure among ever-users of the drug under study.
 We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.
@@
 
-It is thus not possible at this stage definitively to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
+It is not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
 Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.
-For now the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.
-If this correctly estimates a causal effect, then in the Scottish population where 26\% of controls have used proton pump inhibitors about 45% of severe cases are attributable to exposure to proton pump inhibitors.
-Although the rate ratio associated with use at the time of exposure to coronavirus may be underestimated by misclassification of all those with a prescription in the last nine months as exposed, the attributable risk fraction for prescribed use of these drugs will not be underestimated.
-However this does not take into account possible over-the-counter purchases.
+If the effect is causal, the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.
+This would imply that in the Scottish population where 26\% of controls have used proton pump inhibitors about 45% of severe cases are attributable to exposure to proton pump inhibitors.
+This does not take into account possible over-the-counter purchases.
+
+As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk or severity of infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
+In a systematic review of adverse effects of long-term use of proton pump inhibitors, all studies included showed elevated risk of community-acquired pneumonia [@islam_adverse_2018].
+In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].
+It may be relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].
 
 We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.
+We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied in isolation from associations with the profile of all prescribed drugs.
 The strongest independent effect of any class of drug acting on the cardiovascular system is that of loop diuretics.
@@
 The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
-Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including heart disease and co-prescribing of proton pump inhibitors.
+Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.
 The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be accurately estimated.
 As this drug is commonly used in combination with non-steroidal anti-inflammatory agents with which proton pump inhibitors are frequently co-prescribed, we would expect it to be associated with elevated risk in univariate analyses.
-
-A possible effect of proton pump inhibitors on risk of severe COVID-19 was not one of the hypotheses that we specified in advance when setting up this case-control study.
-With hindsight, perhaps it should have been.
-In a systematic review of adverse effects of long-term use of proton pump inhibitors, all eight studies included showed elevated risk of community-acquired pneumonia [@islam_adverse_2018].
-As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk or severity of infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
-
-In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].
-A study in Aberdeen reported that 35% of patients admitted to acute geriatric units had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of co-morbidity [@jarchow-macdonald_prescribing_2013].
-It may be relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors has been reported previously.
-A study in China reported that proton pump inhibitors were prescribed to 37% of inpatients in a tertiary referral centre in Szechuan in 2016 [@luo_changes_2018].
-In Italy, a survey of nursing homes found that 46% of residents had been prescribed a proton pump inhibitor [@pasina_overuse_2020].
-A Spanish drug-utilization study reported that 55% of hospital inpatients were presribed a proton pump inhibitor at discharge [@ramirez_overuse_2010].
 
@@
 Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.
-Unless this this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
+Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
 Although we do not have morbidity data from primary care, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
@@
 
+```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::kable(table.numdrugsgr.agegr,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="numdrugsgr.agegr",
+             row.names=FALSE,
+             col.names=c("Number of drugs", colnames(table.numdrugsgr.agegr)[2:3], 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group") %>%
+    group_rows("Age 0-59", 1, 6) %>%
+    group_rows("Age 60-74", 7, 12) %>%
+    group_rows("Age 75+", 13, 18) %>%
+    kable_styling(latex_options="hold_position")
+```
+
+```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.numdrugsgr.num.icdch,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="numdrugs.num.icdch",
+             row.names=FALSE,
+             col.names=c("Number of drug classes", colnames(table.numdrugsgr.num.icdch)[2:3],  
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Association of severe COVID-19 with polypharmacy, by number of ICD-10 chapters represented in hospital diagnoses") %>%
+    column_spec(column=1, width="2cm") %>%
+    group_rows("No admissions", 1, 6) %>%
+    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
+    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r numdrugs.listed.any, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.numdrugsgr.listed.any,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="numdrugs.listed.any",
+             row.names=FALSE,
+             col.names=c("Number of drug classes", colnames(table.numdrugsgr.listed.any)[2:3], 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Association of severe COVID-19 with polypharmacy, by diagnosis of at least one listed condition") %>%
+    column_spec(column=1, width="2cm") %>%
+    group_rows("No listed condition", 1, 6) %>%
+    group_rows("Listed condition", 7, 12) %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
 ```{r subpara.coeffs, echo=FALSE, warning=FALSE, message=FALSE}
@@
             "Rate ratio", "$p$-value"),
-             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in a care home and without a listed condition, and filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
+             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in care homes and without a listed condition, filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
     column_spec(column=1, width="12cm") %>%
@@
     landscape()
-
-```
-
-```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.numdrugsgr.agegr,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.agegr",
-             row.names=FALSE,
-             col.names=c("Number of drugs", "Controls", "Cases", 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group") %>%
-    group_rows("Age 0-59", 1, 6) %>%
-    group_rows("Age 60-74", 7, 12) %>%
-    group_rows("Age 75+", 13, 18) %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r numdrugs.cats3, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.numdrugsgr.cats3,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.cats3",
-             row.names=FALSE,
-             col.names=c("Number of drug classes", "Controls", "Cases", 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by residence/health status") %>%
-    group_rows("Care home residence", 1, 6) %>%
-    group_rows("Living independently, listed condition", 7, 12) %>%
-    group_rows("Living independently, no listed condition", 13, 18) %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.numdrugsgr.num.icdch,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.num.icdch",
-             row.names=FALSE,
-             col.names=c("Number of drug classes", "Controls", "Cases", 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by number of ICD-10 chapters represented in hospital diagnoses") %>%
-    group_rows("No admissions", 1, 6) %>%
-    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
-    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
-    kable_styling(latex_options="hold_position")
 
@@
              col.names=c("Dropped drug class", "Log rate ratio before dropping",
-                         "Change in log-likelihood when dropped"), 
+                         "Log-likelihood of model when dropped"), 
              digits=c(0, 2, 1), 
-             caption=paste("Stepwise dropping of last", length(rows.print), "drug classes from the variable defined as number of drug classes dispensed: dataset restricted to those not in care homes and without listed conditions")) %>%
+             caption=paste("Stepwise dropping of last", length(rows.print), "drug classes from the polypharmacy variable: dataset restricted to those not resident in care homes and without a listed condition.
+Log-likelihood is difference from log-likelihood of full model with all drug classes included in the polypharmacy variable")) %>%
     column_spec(column=1, width="8cm") %>%
@@
              row.names=FALSE, # dose levels in column 1
-             col.names=c("Average DDDs/day", "Controls", "Cases",
+             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value",
@@
 
-```{r ppinames, echo=FALSE, warning=FALSE, message=FALSE}
-
-rownames(table.ppinames) <- tolower.exceptfirstchar(rownames(table.ppinames))
-
-knitr::kable(table.ppinames[, 1:4],
-             escape=FALSE,
-             label="ppinames", 
-             booktabs=TRUE,
-             col.names=c("Controls", "Cases",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Univariate associations of severe COVID-19 with dispensed average daily dose (DDDs/day) for each drug in the proton pump inhibitor class") %>%
-    add_header_above(c(" "=1, "Average daily dose (DDDs)"=2)) %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r everuse.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.everuse.protonpump,
-             escape=FALSE,
-             label="everuse.protonpump", 
-             booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.everuse.protonpump)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with ever-use of proton pump inhibitor by age group, with and without adjustment for prespecified covariates") %>%
-    add_header_above(c(" "=2, "Univariate"=2, "Adjusted for covariates"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```
-
-```{r fatal.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.fatal.protonpump,
-             escape=FALSE,
-             label="fatal.protonpump", 
-             booktabs=TRUE,
-             col.names=c("Controls", "Cases",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of fatal COVID-19 with ever-use of proton pump inhibitor  by age group") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
 ```{r timewin.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
@@
              booktabs=TRUE,
-             col.names=c("Controls", "Cases",
+             col.names=c(clean.header(colnames(table.timewindow.protonpump)[1:2]), 
                          "Rate ratio (95\\% CI)", "$p$-value"),

### 2020-06-03 pharmaco.R

@@ -13,6 +13,15 @@ table.bnfchapter4 <- tabulate.bnfsubparas(chnum=4, data=cc.severe, minrowsum=50)
 ## chapter 10 has to be disaggregated to drug names as grouping of DMARDs is too broad
 table.bnfchapter10 <- tabulate.bnfchemicals(chnum=10, data=cc.severe, minrowsum=50)
 
+nocare.drugfreqs <- sapply(cc.severe[cc.severe$care.home=="Independent" &
+                              cc.severe$CASE==0, subparacols],
+       function(x) table(x)[2] / sum(table(x))) %>% sort(decreasing=TRUE) %>% head(20)
+
+care.drugfreqs <- sapply(cc.severe[cc.severe$care.home=="Care/nursing home" &
+                              cc.severe$CASE==0, subparacols],
+       function(x) table(x)[2] / sum(table(x))) %>% sort(decreasing=TRUE) %>% head(20)
+
+
 #############################
 
 coeff.anticoagulant.univariate <- summary(clogit(formula=CASE ~ anticoagulant.any + strata(stratum),
@@ -28,6 +37,7 @@ coeff.hydroxychlor.univariate <- summary(clogit(formula=CASE ~ hydroxychloroquin
 
 coeff.hydroxychlor.adjusted <- summary(clogit(formula=CASE ~ hydroxychloroquine + protonpump + strata(stratum),
                data=cc.severe))$coefficients[1, ]
+
 
 
 ##############################################################################
@@ -70,7 +80,8 @@ for(numicd in levels(cc.severe$num.icdchapters.gr)) {
     table.numdrugsgr.num.icdch <- rbind(table.numdrugsgr.num.icdch, x)
 }
 table.numdrugsgr.num.icdch <- data.frame(numdrugsgr=rep(levels(cc.severe$numdrugsgr), 3),
-                               table.numdrugsgr.num.icdch)
+                                         table.numdrugsgr.num.icdch)
+colnames(table.numdrugsgr.num.icdch)[2:3] <- paste0(c("Controls (", "Cases ("), as.integer(table(cc.severe$CASE)), rep(")", 2))
 
 ####################################################################################
 ## tabulate rate ratios for each numdrugs group, by agegr 
@@ -84,18 +95,33 @@ for(agegr in levels(cc.severe$agegr3)) {
 }
 table.numdrugsgr.agegr <- data.frame(numdrugsgr=rep(levels(cc.severe$numdrugsgr), 3),
                                table.numdrugsgr.agegr)
[…]

### 2020-06-03 propensity.R

@@ -52,5 +52,4 @@ coeffs.propensity <- data.frame(subpara=gsub("subpara\\.", "", names(coeffs.full
 
 rm(newdata.x)
 rm(full.model)
-rm(stepwise.full)
 rm(train.data)

### 2020-06-04 pharmaco.Rmd
last version before shortening pharmacoepi paper for NEJM

@@
 author:
+
   - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
@@
 #  pdf_document:
-  rticles::plos_article:
+  rticles::nejm_article:
+    includes:
+      in_header: ../preamble.tex
     latex_engine: lualatex
@@
     keep_tex: true
-csl: ../plos.csl
+csl: ../../csl/styles/the-new-england-journal-of-medicine.csl
 urlcolor: red
@@
 
-_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as high risk (listed conditions).
+_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as increased risk (listed conditions).
 We report a further investigation of possible explanations for this association, which include specific co-morbidity or drug effects but also more diffuse effects of frailty or polypharmacy.
 
-_Methods_ -- Severe COVID-19 cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database and entered a critical care unit or died within 28 days of testing.
+_Methods_ -- Cases of severe COVID-19 were ascertained as all those who tested positive for SARS-CoV-2 nucleic acid in the national database and entered a critical care unit or died within 28 days of testing.
 From the national population register ten controls per case were selected, matched for sex, age and general practice.
-International Classification of Disease (ICD-10) codes were extracted from the past five years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary care during the past nine months.
+International Classification of Disease (ICD-10) codes were extracted from the past three years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary care during the past nine months.
 Rate ratios were estimated by conditional logistic regression.
 
-*Findings* -- We began by examining associations with polypharmacy, defined as total number of drug classes dispensed.
-The rate ratios (95% CIs) for severe COVID-19 associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
-The association with polypharmacy was strongest in those without past hospital admissions or diagnoses of conditions previously designated as conferring high risk.
-The drug classes associated with COVID-19 included widely-prescribed cardiovascular drugs and also multiple drug classes previously identified as indicators of overprescribing without clear medical indications.
+*Findings* -- We began by examining associations with polypharmacy, defined by total number of drug classes dispensed.
+With those who had no dispensed prescriptions as reference group, rate ratios (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+Over all age groups the association with polypharmacy was strongest in those without hospital admissions in the past three years: in this group the rate ratio associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.num.icdch[6, 4]`.
+The drug classes associated with COVID-19 included many drug classes previously identified as indicators of overprescribing.
 A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.
-Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for severe COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.
+Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.
 The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.
-However a comparison of time windows representing recent and less recent exposure in ever-users did not show a stronger association with recent exposure.
+However a comparison of time windows did not show a stronger association with recent exposure than with less recent exposure.
 
@@
 The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.
-In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year.
+In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
 The univariate rate ratios associated with at least one prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.
-This association persisted after adjusting for care home residence, any hospital admission in the last five yeasrs and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
-Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in almost all BNF chapters were associated with elevated rate ratios for COVID-19.
-We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by the indications for which these drugs were prescribed or by frailty especially in older persons, causal effects of drugs on disease susceptibility could not be ruled out.
+This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
+Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in drugs in almost all chapters were associated with elevated rate ratios for COVID-19.
+We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by co-morbidity or frailty, causal effects of drugs on disease susceptibility could not be ruled out.
 
-# Aim
-The primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
+Thus the primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
 Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.
-A further aim was to determine whether tthe effects underlying this association are sparse or diffuse.
+A further aim was to determine whether the effects underlying this association are sparse or diffuse.
 In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.
@@
 # Methods
-The design of the REACT-SCOT case-control study is described in detail in the accompanying paper.
-All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
+The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].
+In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units.
@@
 Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-For each test-positive case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
-Socioeconomic status was recorded as the quintile of the Scottish Index of Multiple Deprivation (SIMD), based on the postal code.
+The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).
 
 ## Morbidity and drug prescribing
-For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2.
+For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2.
 British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
@@
 BNF subparagraph codes generally represent pharmacological drug classes.
-For this analysis prescription codes from chapters 14 and above, comprising mostly dressings and appliances but also vaccines, anaesthesia and other preparations were grouped as "Other".
-Defined daily doses (DDDs) of each drug were obtained from the DDD/ATC Index of the WHO Collaborating Centre for Drug Statistics Methodology.
+For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Otherf".
+Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
 Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.
@@
 # Statistical analysis 
-Rate ratios for severe COVID-19 were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
-As described previously, we derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-For this analysis the list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.
+Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
+As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+ICD-10 and BNF codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565) For this analysis this list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.
 These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an dispensed prescription of a drug for which that condition is the only indication.
-Diagnoses of diabetes were extracted from linkage to the national diabetes register.
+Diagnoses of diabetes were extracted from linkage to the national diabetes register and supplemented by ICD-10 codes for diabetes and BNF codes for diabetes drugs for recently diagnosed persons not on the last extract of the register .
 The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.
+<!-- clone the github repo -->
 
@@
 To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
-As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, a independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
+As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, an independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
 By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.
@@
 # Results
-## Relation of severe COVID-19 to number of drug classes dispensed. 
+## Relation of severe COVID-19 to polypharmacy 
 We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).
-Table \ref{tab:numdrugsgr.agegr} shows that risk of severe COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+Table \ref{tab:numdrugsgr} shows that risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).
+In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.
 
-Table \ref{tab:numdrugs.num.icdch} shows that this association was strongest in those who had no hospital diagnoses (because they had not been admitted to hospital).
-In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.
-Table \ref{tab:numdrugs.listed.any} shows that this association was strongest in those without a diagnosis of a listed condition.
-
-To study the relationship further, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.
-The rationale for this restriction was to reduce the effect of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and severe COVID-19.
+To study further the relationship with polypharmacy, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.
+The rationale for this restriction was to reduce the frequency of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and COVID-19.
 In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.
+Supplementary Table \ref{tab:numdrugs.listed.any} shows that the association with COVID-19 was stronger in those without a diagnosis of a listed condition than in those with one or more listed conditions.
 Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.
 
-## Univariate associations of severe COVID-19 with drug classes
+## Associations of severe COVID-19 with drug classes
 We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
-Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.
-These COVID-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing without medical indications: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
+Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.
+These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
 
-## Backward selection to identify specific drug classes contributing to the association of severe COVID-19 with number of drug classes dispensed
-A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with number of drugs dispensed.
+A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.
 To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.
-Dropping the first `r firstrow.neg - 1` drug classes from this subset increased the log-likelihood of the model.
+Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
+Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.
+The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.
 Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-Table \ref{tab:stepwise.subparas} shows how the log-likelihood decreased (though the rate ratio increased) as these variables were dropped from the model.
-
-The only cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
+Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
 Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.
-The last drug class retained by backward selection, and thus the variable with largest independent explanatory power, was proton pump inhibitors.
+The last drug class retained by backward selection, and thus the variable with largest independent explanatory power for the effect of polypharmacy, was proton pump inhibitors.
 
 ## Associations of severe COVID-19 with proton pump inhibitor use
-This unexpected finding led us to investigate the association of COVID-19 with exposure to proton pump inhibitors in more depth.
-This is challenging because proton pump inhibitors are usually co-prescribed with other drugs.
+As the association of COVID-19 and exposure to proton pump inhibitors is not obviously explained by established co-morbidities, we explored this association in more depth.
+As this examined a specific exposure, rather than polypharmacy which may have different profiles in care home residents or those diagnosed with listed conditions, the observations were not restricted to exclude care home residents or those diagnosed with a listed condition.
+Distinguishing between causality and confounding is challenging because proton pump inhibitors are usually co-prescribed with other drugs.
 We used several approaches:
@@
 
-* Adjusting for prespecified covariates: care home residence, SIMD quintile, index of hospital diagnosis of a disorder of the esophagus, stomach or duodenum, ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+* Adjusting for prespecified covariates: care home residence, SIMD quintile, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
 
@@
 
-For these analyses the dataset was not restricted to exclude care home residents or those  diagnosed with a listed condition.
-
 Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed.
@@
 Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.
-The effect size for unit increase in dose expressed as average DDDs/day was higher for lansoprazole [rate ratio `r table.ppinames[2, 3]`] than for omeprazole rate [rate ratio `r table.ppinames[3, 3]`].
-This suggests that the DDDs for lansoprazole (30 mg) and omeprazole (20 mg) are not equivalent in potency.
+The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].
+cy.
 
@@
 A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section.
-The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Table \ref{tab:propensity}.
+The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Supplementary Table \ref{tab:propensity}.
 As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
@@
 ## Associations with other drug classes
-Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).
-These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- angiotensin-converting enzyme (ACE) inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
+Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).
+These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
 
@@
 In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.
-Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and angiotensin-converting enzyme inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.
+Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and ACE inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.
 The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.
@@
 Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with drugs widely used in care homes: antipsychotics, analgesics and drugs for dementia.
-
-Table \ref{tab:bnfchapter10} shows associations with drugs disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code.
+Table \ref{tab:bnfchapter10} shows associations with drugs for the musculoskeletal system disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code.
 All antirheumatic drugs were associated with elevated rate ratios for COVID-19.
@@
 # Discussion 
-We have shown that severe COVID-19 is associated with dispensed prescriptions of drugs of many different classes.
-These include drugs acting on the the cardiovascular system, where prescribing of multiple drugs is supported by evidence-based guidelines, but also drugs acting on the gastrointestinal and central nervous systems that are widely used as indicators of overprescribing and inappropriate polypharmacy in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
-This led us to examine associations of COVID-19 with polypharmacy, defined as the number of drugs classes dispensed during the period of observation.
-We attempted to determine how much of this association is attributable to confounding by multimorbidity, and how much is attributable to adverse effects of drug prescribing.
+We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.
+We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.
 Without more data on health status from primary care records, the confounders cannot be accurately measured.
-We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with common conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
-The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of about 3 for severe COVID-19 that are associated with designated conditions.
+We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
+The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.
+The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
 
-We have shown that a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with number of drug classes dispensed.
-The association of COVID-19 with dispensed prescriptions of proton pump inhibitors shows a clear dose-response effect.
-Adjustment for covariates pre-specified as likely to confound this assocation -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly.
+Although a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy, this procedure is likely to retain the drug classes that are most commonly prescribed.
+We have shown a clear dose-response relationship of dispensed average daily dose of proton pump inhibitors to risk of severe COVID-19.
+Adjustment for covariates pre-specified as likely to confound this association -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly.
 The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.
-The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptiblity to COVID-19: antidepressants, analgesics and nutritional supplements.
-When the exposure under study is not measured accurately, including such variables in the model as possible confounders may over-adjust the effect of that exposure under study towards the null.
+
+The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.
+Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.
+In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.
 By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small.
-One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to investigate the relationship of adverse events to the time of exposure among ever-users of the drug under study.
+One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.
 We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.
-However most users of these drugs had prescriptions dispensed in both time windows.
-Users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.
-It is also possible that use of proton pump inhibitors has a persistent effect on susceptibility to infection through alteration of the gut microbiome.
+However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.
 
-It is not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
+It is thus not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
 Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.
 If the effect is causal, the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.
-This would imply that in the Scottish population where 26\% of controls have used proton pump inhibitors about 45% of severe cases are attributable to exposure to proton pump inhibitors.
-This does not take into account possible over-the-counter purchases.
 
-As proton pump inhibitors are also associated with increased risk of enteric infections, it is plausible that they could increase the risk or severity of infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].
-In a systematic review of adverse effects of long-term use of proton pump inhibitors, all studies included showed elevated risk of community-acquired pneumonia [@islam_adverse_2018].
-In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].
-It may be relevant that in some other countries where COVID-19 epidemics have been especially severe, overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].
+As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through more persistent effects on the gut microbiome.
+Uncertainties about the effect of proton pump inhibitors on the risk of community acquired pneumonia [@othman_community_2016] are not necessarily relevant to effects on enteric infection.
+Proton pump inhibitors have been reported to affect the gut microbiome [@imhann_proton_2016].
+If effects of proton pump inhibitor use on susceptiblity to COVID-19 are mediated through alteration of the gut microbiome, these effects would not necessarily be detected in a comparison of time windows.
+ <!---barrier to viral infection, or through altering the microbiome allowing with Warnings been issued about increased risk of _Cl.
+difficile_ infection from use proton pump inhibitors.
+[ https://bnf.nice.org.uk/drug/omeprazole.html#importantSafetyInformations ] Given this possible association with increased risk of enteric infections --->.
+
+In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].
+Initiatives to improve prescribing quality have attempted to reduce this.
+It would be of interest to examine the association between severe COVID-19 and use of proton pump inhibitors in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].
 
 We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.
-We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied in isolation from associations with the profile of all prescribed drugs.
 The strongest independent effect of any class of drug acting on the cardiovascular system is that of loop diuretics.
@@
 Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
-Although we do not have morbidity data from primary care, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
+Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
 This diagnostic coding is more relevant to disease classification than the SNOMED diagnostic codes used in primary care databases.
@@
 ## Conclusion
-Severe COVID-19 is strongly associated with the number of drug classes dispensed as prescriptions during the last year.
-This effect is not easily explained by co-morbidity, and it is strongest in relatively young individuals without hospital diagnoses of conditions that confer high risk of disease.
-The profile of drug classes associated with COVID-19 includes many drug classes that are used as indicators of overprescribing and inappropriate polypharmacy.
-The single drug class with the greatest explanatory power is proton pump inhibitors.
-Although this study did not identify unequivocal evidence of causality for this drug class, increased risk of infection is an established adverse effect of the use of these drugs.
+Severe COVID-19 is strongly associated with polypharmacy.
+This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.
+We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs.
+The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.
+The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.
+Although this study did not identify unequivocal evidence of causality for the association with this drug class, effects on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible for these drugs.
 We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.
@@
 ## Information governance
-This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
+This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
 Datasets were de-identified before analysis.
-The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance.
+The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
 
@@
 HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
-HC receives or has recently received non-binding research support from Pfizer, AstraZeneca and Novo-Nordisk.
+HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.
+HC and PM own stock in Roche and Bayer.
 
@@
 
-# Tables
+```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the coefficient of the polypharmacy variable with that variable retained, and the log-likelihood when that drug class is dropped"}
+
+stepwise.asfactor <- stepwise.drop.subparas
+stepwise.asfactor$name <- gsub("\\.", " ", stepwise.asfactor$name)
+stepwise.asfactor$numrow <- nrow(stepwise.asfactor):1
+stepwise.asfactor$name <- factor(stepwise.asfactor$name, levels=rev(stepwise.asfactor$name))
+
+p.loglik <- ggplot(data=stepwise.asfactor, 
+                   aes(x=loglik.full, y=name)) + geom_bar(stat="identity", fill="gray") +
+    xlab("Log-likelihood")  +
+    theme(axis.title.y=element_blank())
+#          axis.text.y=element_blank())
+
+p.coeff <- ggplot(data=stepwise.asfactor, 
+                  aes(x=beta.full, y=name)) + geom_bar(stat="identity", fill="gray") +
+    xlab("Coefficient (log rate ratio)")
+
+                                        # gridExtra::grid.arrange(p.coeff, p.loglik, nrow=1, widths=c(0.5, 0.4))
+p.loglik
+
+if(FALSE) {
+knitr::kable(stepwise.drop.subparas[rows.print, ], 
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="stepwise.subparas",
+             row.names=FALSE,
+             col.names=c("Dropped drug class", "Log rate ratio before dropping",
+                         "Log-likelihood of model when dropped"), 
+             digits=c(0, 2, 1), 
+             caption=paste("Stepwise dropping of drug classes from the polypharmacy variable: dataset restricted to those not resident in care homes and without a listed condition")) %>%
+    column_spec(column=1, width="8cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    kable_styling(latex_options="hold_position")
+}            
+```
+
+\newpage
 
 ```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
-knitr::kable(table.numdrugsgr.agegr,
+
+table.numdrugsgr <- rbind(table.numdrugsgr.agegr, table.numdrugsgr.num.icdch)
+
+knitr::kable(table.numdrugsgr,
              escape=FALSE,
              booktabs=TRUE,
-             label="numdrugsgr.agegr",
+             label="numdrugsgr",
              row.names=FALSE,
-             col.names=c("Number of drugs", colnames(table.numdrugsgr.agegr)[2:3], 
+             col.names=c("Number of drug classes", colnames(table.numdrugsgr.agegr)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group") %>%
+             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group and by number of ICD chapters represented in hospital diagnoses") %>%
+    column_spec(column=1, width="2cm") %>%
     group_rows("Age 0-59", 1, 6) %>%
@@
     group_rows("Age 75+", 13, 18) %>%
+    group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
+    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
+    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
     kable_styling(latex_options="hold_position")
@@
 ```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
-
+if(FALSE) {
 knitr::kable(table.numdrugsgr.num.icdch,
@@
     kable_styling(latex_options="hold_position")
+}
+```
+
+```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.dosegr.protonpump,
+             escape=FALSE,
+             label="dosegr.protonpump", 
+             booktabs=TRUE,
+             row.names=FALSE, # dose levels in column 1
+             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    group_rows("Ages 0-59 years", 1, 5) %>%
+    group_rows("Ages 60-74 years", 6, 10) %>%
+    group_rows("Ages 75+ years", 11, 15) %>%
+    landscape()
 
 ```
+
+```{r timewin.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.timewindow.protonpump,
+             escape=FALSE,
+             label="timewindow.protonpump", 
+             booktabs=TRUE,
+             col.names=c(clean.header(colnames(table.timewindow.protonpump)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window")) %>%
+	column_spec(column=1, width="3cm") %>%
+	column_spec(column=2:3, width="2cm") %>%
+	column_spec(column=4, width="2.5cm") %>%
+	column_spec(column=5, width="1.5cm") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r bypropensity, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.protonpump.bypropensity[, 1:4],
+             escape=FALSE,
+             booktabs=TRUE,
+             label="bypropensity", 
+             col.names=c(clean.header(colnames(table.protonpump.bypropensity)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Association of severe COVID-19 with ever-use of proton pump inhibitor, by propensity score") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+\newpage
+\beginsupplement
+
+# Supplementary Tables
 
@@
 
-\newpage
-
-```{r stepwise.subparas, echo=FALSE, warning=FALSE, message=FALSE}
-knitr::kable(stepwise.drop.subparas[rows.print, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="stepwise.subparas",
-             row.names=FALSE,
-             col.names=c("Dropped drug class", "Log rate ratio before dropping",
-                         "Log-likelihood of model when dropped"), 
-             digits=c(0, 2, 1), 
-             caption=paste("Stepwise dropping of last", length(rows.print), "drug classes from the polypharmacy variable: dataset restricted to those not resident in care homes and without a listed condition.
-Log-likelihood is difference from log-likelihood of full model with all drug classes included in the polypharmacy variable")) %>%
-    column_spec(column=1, width="8cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.dosegr.protonpump,
-             escape=FALSE,
-             label="dosegr.protonpump", 
-             booktabs=TRUE,
-             row.names=FALSE, # dose levels in column 1
-             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    group_rows("Ages 0-59 years", 1, 5) %>%
-    group_rows("Ages 60-74 years", 6, 10) %>%
-    group_rows("Ages 75+ years", 11, 15) %>%
-    landscape()
-
-```
-
-```{r timewin.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.timewindow.protonpump,
-             escape=FALSE,
-             label="timewindow.protonpump", 
-             booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.timewindow.protonpump)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window")) %>%
-	column_spec(column=1, width="3cm") %>%
-	column_spec(column=2:3, width="2cm") %>%
-	column_spec(column=4, width="2.5cm") %>%
-	column_spec(column=5, width="1.5cm") %>%
-		kable_styling(latex_options="hold_position")
-
-```
-
 ```{r propensity, echo=FALSE, warning=FALSE, message=FALSE}
@@
              caption="Coefficients of propensity score for prescribing of a proton pump inhibitor, constructed in an independent dataset") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r bypropensity, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.protonpump.bypropensity[, 1:4],
-             escape=FALSE,
-             booktabs=TRUE,
-             label="bypropensity", 
-             col.names=c(clean.header(colnames(table.protonpump.bypropensity)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-                caption="Association of severe COVID-19 with ever-use of proton pump inhibitor, by propensity score") %>%
     kable_styling(latex_options="hold_position")

### 2020-06-04 casecontrol.R

@@ -288,10 +288,17 @@ ids.diabetes.extra <- unique(c(ids.icd.diabetes, ids.bnf.diabetes))
 cc.all$dm.type <- as.integer(cc.all$dm.type)
 ## missing recoded as zero
 cc.all$dm.type[is.na(cc.all$dm.type)] <- 0
+
 ## code diagnoses detected from discharges or BNF codes as unknown type
-## we could classify those not on insulin as definite Type 2
+## we could classify those not on insulin as definite Type 2 but Helen says no
 cc.all$dm.type[cc.all$dm.type==0 & cc.all$ANON_ID %in% ids.diabetes.extra] <- 3
 
+## add in extra cases notified directly from SCI-Diabetes register, without assignment
+## of diabetes type from SDRN database
+if(!old) {
+    cc.all$dm.type[cc.all$dm.type==0 & cc.all$diab.reg==1] <- 3
+}
+    
 cc.all$dm.type <- 
   as.factor(car::recode(cc.all$dm.type, 
                         "c(0, 10)='Not diabetic';

### 2020-06-05 casecontrol.Rmd
last version before shortening pharmacoepi paper for NEJM
Fixed to drop controls no longer current

@@
 # Tables
-
 ```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
-knitr::kable(table.icdsubchapters,
+knitr::kable(table.icdsubchapters[grep("times", table.icdsubchapters$u.pvalue), ],
              escape=FALSE, 

### 2020-06-05 pharmaco.Rmd
Fixed to drop controls no longer current

@@
 ---
-title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study'
+title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)'
 author:
@@
 Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive [and had not embarked -- to be checked in the CHI database by SW].
 Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).
@@
 We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.
-The strongest independent effect of any class of drug acting on the cardiovascular system is that of loop diuretics.
-This is likely to reflect confounding by heart failure as one of the main indications for use of loop diuretics.
+We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs.
 For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].
-A more detailed study of the relation of COVID-19 to cardiovascular disease and drugs prescribed for cardiovascular disease will be reported elsewhere.
 The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
 Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.
-The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be accurately estimated.
-As this drug is commonly used in combination with non-steroidal anti-inflammatory agents with which proton pump inhibitors are frequently co-prescribed, we would expect it to be associated with elevated risk in univariate analyses.
+The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.
 
@@
 Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
-This diagnostic coding is more relevant to disease classification than the SNOMED diagnostic codes used in primary care databases.
-Although this case-control study is matched on primary care practice, in Scotland there is ample variation in high-risk prescribing between practitioners in the same practice [@guthrie_bad_2015], allowing study of the relationship of drug prescribing to COVID-19.
 
@@
 This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.
-We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs.
 The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.
 The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.
-Although this study did not identify unequivocal evidence of causality for the association with this drug class, effects on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible for these drugs.
-We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and should advise against their use without a recognized medical indication.
+Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.
+We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors and on the avoidance of inappropriate polypharmacy.
 

### 2020-06-05 pharmaco_nejm.Rmd
Fixed to drop controls no longer current

--- 2020-06-05 pharmaco.Rmd
@@
 
-# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf")
+# rmarkdown::render("pharmaco_nejm.Rmd", output_file="pharmaco.pdf")
 
@@
 # Background
-The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.
+The REACT-SCOT study is a matched case-control study of severe COVID-19 in Scotland.
 In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
@@
 This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
-Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in drugs in almost all chapters were associated with elevated rate ratios for COVID-19.
-We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by co-morbidity or frailty, causal effects of drugs on disease susceptibility could not be ruled out.
-
-Thus the primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
-Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.
-A further aim was to determine whether the effects underlying this association are sparse or diffuse.
-In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.
-In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.
+The objective of this study was to investigate possible explanations for this association: confounding by co-morbidity  factors, or direct effects of prescribed drugs.
 
@@
 The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].
-In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
+In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units.
 Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
-Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive [and had not embarked -- to be checked in the CHI database by SW].
-Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).
+Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test.
+For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+For this study based on ascertainment of test-positive , up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.
 
@@
 For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2.
-British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
-The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
-BNF subparagraph codes generally represent pharmacological drug classes.
-For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Otherf".
-Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
-Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.
+British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2.
+The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.
+For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".
+Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology and used to calculate average daily doses were calculated for each individual as the sum of (strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).
+As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+ICD-10 diagnostic and BNF drug codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).
+For this analysis this list was extended to include non-cardiac circulatory disease as initial analysis of this case-control study had shown this to be a risk factor for COVID-19.
+Diagnoses of diabetes were extracted from linkage to the national diabetes register.
 
@@
 Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
-As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-ICD-10 and BNF codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565) For this analysis this list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.
-These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an dispensed prescription of a drug for which that condition is the only indication.
-Diagnoses of diabetes were extracted from linkage to the national diabetes register and supplemented by ICD-10 codes for diabetes and BNF codes for diabetes drugs for recently diagnosed persons not on the last extract of the register .
-The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material.
-<!-- clone the github repo -->
-
-To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we carried out a backward stepwise variable selection procedure. 
+To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure.
 Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.
@@
 As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, an independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
-By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.
 The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
@@
 # Results
-## Relation of severe COVID-19 to polypharmacy 
-We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).
+We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the observation period.
 Table \ref{tab:numdrugsgr} shows that risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
@@
 
-To study further the relationship with polypharmacy, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.
-The rationale for this restriction was to reduce the frequency of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and COVID-19.
-In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.
+We next investigated whether specific drug classes might underlie this association with polypharmacy.
+To reduce confounding by multimorbidity and frailty, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.
 Supplementary Table \ref{tab:numdrugs.listed.any} shows that the association with COVID-19 was stronger in those without a diagnosis of a listed condition than in those with one or more listed conditions.
-Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.
-
-## Associations of severe COVID-19 with drug classes
-We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
+We examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
 Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.
@@
 Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.
-The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.
+The drug classes dropped at this stage included some of the most widely-prescribed drug classes that were associated with COVID-19 in univariate analyses, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.
 Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
-Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.
+The `r length(rows.print)` drug classes retained at this stage included loop diuretics, calcium channel blockers, and non-opioid analgesics.
 The last drug class retained by backward selection, and thus the variable with largest independent explanatory power for the effect of polypharmacy, was proton pump inhibitors.
 
-## Associations of severe COVID-19 with proton pump inhibitor use
-As the association of COVID-19 and exposure to proton pump inhibitors is not obviously explained by established co-morbidities, we explored this association in more depth.
-As this examined a specific exposure, rather than polypharmacy which may have different profiles in care home residents or those diagnosed with listed conditions, the observations were not restricted to exclude care home residents or those diagnosed with a listed condition.
-Distinguishing between causality and confounding is challenging because proton pump inhibitors are usually co-prescribed with other drugs.
-We used several approaches:
+As the association with proton pump inhibitors was not obviously attributable to confounding by conditions designated as risk conditions for COVID-19, we explored it in more depth.
+For these analyses the dataset was not restricted to exclude care home residents or those diagnosed with a listed condition as we were examining the effects of a specific drug class rather than polypharmacy.
+We used several approaches to distinguish between causality and confounding:
 
@@
 
-* Adjusting for prespecified covariates: care home residence, SIMD quintile, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+* Adjusting for prespecified covariates: care home residence, quintile of index of socioeconomic deprivation, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
 
@@
 The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].
-cy.
-
 Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.
@@
 As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
-Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.
+Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained only in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.
 
 ## Associations with other drug classes
-Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).
-These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
+Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal), selected  because drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
 
 Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.
-Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.
-
 Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19.
@@
 We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.
-Without more data on health status from primary care records, the confounders cannot be accurately measured.
-We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
-The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.
+The association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
+The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the rate ratios of 2 to 5 for severe COVID-19 associated with designated conditions in this population .
+
 The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
-
 Although a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy, this procedure is likely to retain the drug classes that are most commonly prescribed.
@@
 
-The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.
-Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.
-In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.
+Where the exposure under study cannot be measured accurately, as in this study that relies on dispensed prescriptions, rather than direct recording of use including over the counter purchasing, adjusting for a propensity score may over-adjust the effect of the exposure under study towards the null.
 By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small.
@@
 It is thus not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
-Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.
-If the effect is causal, the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.
-
-As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through more persistent effects on the gut microbiome.
-Uncertainties about the effect of proton pump inhibitors on the risk of community acquired pneumonia [@othman_community_2016] are not necessarily relevant to effects on enteric infection.
-Proton pump inhibitors have been reported to affect the gut microbiome [@imhann_proton_2016].
-If effects of proton pump inhibitor use on susceptiblity to COVID-19 are mediated through alteration of the gut microbiome, these effects would not necessarily be detected in a comparison of time windows.
- <!---barrier to viral infection, or through altering the microbiome allowing with Warnings been issued about increased risk of _Cl.
-difficile_ infection from use proton pump inhibitors.
-[ https://bnf.nice.org.uk/drug/omeprazole.html#importantSafetyInformations ] Given this possible association with increased risk of enteric infections --->.
-
-In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].
-Initiatives to improve prescribing quality have attempted to reduce this.
+Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy may be able to resolve this.
+As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through effects on the gut microbiome that have been demonstrated [@imhann_proton_2016].
+In Scotland and other countries proton pump inhibitors are widely prescribed, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].
 It would be of interest to examine the association between severe COVID-19 and use of proton pump inhibitors in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].
@@
 The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
-Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.
-The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.
+We found no evidence that prior use of anticoagulants protects against severed disease, but the association is heavily confounded by factors including diagnosed heart disease and co-prescribing of proton pump inhibitors.
+The observed association with hydroxychloroquine is similar to that for other anti-rheumatic drugs, but the numbers exposed to this drug were too small for this association to be estimated reliably.
 
 ## Methodological strengths and weaknesses
-A strength of this study is that it is based on dispensed prescriptions, rather than issued prescriptions.
+A strength of this study is that it is based on dispensed rather than issued prescriptions.
 Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.
-Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
-Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
+Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 by trained coders.
 
@@
 Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.
-We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors and on the avoidance of inappropriate polypharmacy.
+We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and more generally on the avoidance of unnecessary polypharmacy.
 
 # Declarations
-## Information governance
+## Information governance and data availability
 This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
@@
 Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
+The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
 
@@
 ## Supplementary material
-The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
 
@@
 
-```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the coefficient of the polypharmacy variable with that variable retained, and the log-likelihood when that drug class is dropped"}
+```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.width=5, fig.cap="\\label{fig:stepwise}Stepwise dropping of drug classes from the polypharmacy variable: each row shows the coefficient of the polypharmacy variable with that variable retained, and the log-likelihood when that drug class is dropped"}
 

### 2020-06-05 casecontrol.R

@@ -21,13 +21,14 @@ registerDoParallel(cores=2)
 source("helperfunctions.R")
 
 old <- TRUE
-#old <- FALSE
+old <- FALSE
 
 if(old) {
     cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
     diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") 
     procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
     scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
+    controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-05.rds")
 } else {
     cc.all <- readRDS("./data/CC_linked_ANON_20200515.rds")
     diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
@@ -35,6 +36,7 @@ if(old) {
     scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
     scrips <- readRDS(scrips.filename)[, c("ANON_ID", "bnf_paragraph_code",
                                            "bnf_paragraph_description")] 
+    controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-15.rds")
 
     ## scrips should be 2 tables to save space
     ## one record per scrip
@@ -116,6 +118,26 @@ names(cc.all) <- gsub("^age$", "AGE", names(cc.all))
 
 ## HAI is based on the ECDC definition of nosocomial infection
 
+
+cc.all$scrip.any <- as.factor(as.integer(cc.all$ANON_ID %in% scrips$ANON_ID))
+cc.all$diag.any <- as.factor(as.integer(cc.all$ANON_ID %in% diagnoses$ANON_ID))
+
+## exclude controls not observable
+
+controls.status <- controls.status[, c("ANON_ID", "CHI.EXTENDED.STATUS", "CHI.DATE.OF.DEATH",
+                                   "CHI.Explanation")]
+cc.all <- merge(cc.all, controls.status, by="ANON_ID", all.x=TRUE)
+
+print(with(cc.all, table(CHI.Explanation, scrip.any)))
+print(with(cc.all, table(CHI.Explanation, diag.any)))
+
+browser("status")
+
+cc.all <- subset(cc.all, subset=is.na(CHI.Explanation) |
+                             CHI.Explanation=="Current and no history" |
[…]

### 2020-06-05 stepwise.R

@@ -113,6 +113,7 @@ if(stepwise) {
     }
 ###################################
     demog.predicted <- cv.predicted
+    demog.predicted <- demog.predicted[demog.predicted$prior.p < 1, ]
     demog.densities <- with(demog.predicted,
                             Wdensities(y, posterior.p, prior.p,
                                        recalibrate=FALSE))
@@ -153,6 +154,7 @@ if(stepwise) {
     }
 ###################################
     listed.predicted <- cv.predicted
+    listed.predicted <- listed.predicted[listed.predicted$prior.p < 1, ]
     listed.densities <- with(listed.predicted,
                             Wdensities(y, posterior.p, prior.p,
                                        recalibrate=FALSE))
@@ -195,6 +197,7 @@ if(stepwise) {
 ###################################
 
     full.predicted <- cv.predicted
+    full.predicted <- full.predicted[full.predicted$prior.p < 1, ]
     full.densities <- with(full.predicted,
                            Wdensities(y, posterior.p, prior.p,
                                       recalibrate=FALSE))

### 2020-06-09 pharmaco.Rmd
fixed to include NRS deaths in severe case def if they were in critical care

@@
 
-_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was strongly associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as increased risk (listed conditions).
+_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as increased risk (listed conditions).
 We report a further investigation of possible explanations for this association, which include specific co-morbidity or drug effects but also more diffuse effects of frailty or polypharmacy.
@@
 In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
-The univariate rate ratios associated with at least one prescription varied from 10 in those aged under 60 years to 40 in those aged 75 years and over.
+The univariate rate ratios associated with at least one prescription varied from 4.4 in those aged under 60 years to 3.5 in those aged 75 years and over.
 This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
@@
 
-Thus the primary aim of this study was to explore the strong association of severe COVID-19 with dispensing of at least one prescription in the past year.
+The primary aim of this study was to investigate the association of severe COVID-19 with drug prescribing.
 Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.
@@
 We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).
-Table \ref{tab:numdrugsgr} shows that risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+
+Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescription.
+
 The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).
 In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.
+
+Risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
 
@@
 In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.
-Supplementary Table \ref{tab:numdrugs.listed.any} shows that the association with COVID-19 was stronger in those without a diagnosis of a listed condition than in those with one or more listed conditions.
 Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.
@@
 
+```{r numdrugs.listed.any, echo=FALSE, warning=FALSE, message=FALSE}
+
+freqs <- with(cc.severe, table(listed.any, CASE))
+
+knitr::kable(table.numdrugsgr.listed.any,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="numdrugs.listed.any",
+             row.names=FALSE,
+             col.names=c("Number of drug classes", colnames(table.numdrugsgr.listed.any)[2:3], 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Association of severe COVID-19 with polypharmacy, by diagnosis of at least one listed condition") %>%
+    column_spec(column=1, width="2cm") %>%
+    group_rows(paste0("No listed condition (",
+                      freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
+               1, 6) %>%
+    group_rows(paste0("Listed condition (",
+                      freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
+               7, 12) %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
 ```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
 
+if(FALSE) {
 table.numdrugsgr <- rbind(table.numdrugsgr.agegr, table.numdrugsgr.num.icdch)
@@
     kable_styling(latex_options="hold_position")
+}
+
 ```
@@
 ```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
-if(FALSE) {
 knitr::kable(table.numdrugsgr.num.icdch,
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with polypharmacy, by number of ICD-10 chapters represented in hospital diagnoses") %>%
+             caption="Association of severe COVID-19 with polypharmacy in those without listed conditions, by number of ICD-10 chapters represented in hospital diagnoses") %>%
     column_spec(column=1, width="2cm") %>%
@@
     kable_styling(latex_options="hold_position")
-}
 ```
@@
 # Supplementary Tables
-
-```{r numdrugs.listed.any, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.numdrugsgr.listed.any,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.listed.any",
-             row.names=FALSE,
-             col.names=c("Number of drug classes", colnames(table.numdrugsgr.listed.any)[2:3], 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with polypharmacy, by diagnosis of at least one listed condition") %>%
-    column_spec(column=1, width="2cm") %>%
-    group_rows("No listed condition", 1, 6) %>%
-    group_rows("Listed condition", 7, 12) %>%
-    kable_styling(latex_options="hold_position")
-
-```
 

### 2020-06-09 casecontrol.R

@@ -18,7 +18,7 @@ library(dplyr)
 
 registerDoParallel(cores=2)
 
-bsource("helperfunctions.R")
+source("helperfunctions.R")
 
 old <- TRUE
 old <- FALSE # uncomment to use 15 May linkage
@@ -27,9 +27,7 @@ nrs <- TRUE
 nrs <- FALSE # uncomment to exclude NRS-only deaths
 
 stepwise <- TRUE   
-#stepwise <- FALSE ## uncomment to save time if old version still valid
-
-
+stepwise <- FALSE ## uncomment to save time if old version still valid
 
 if(old) {
     cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
@@ -231,13 +229,15 @@ cc.all$group[cc.all$CASE==1 & cc.all$nrs_covid_case==1] <- "D"
 cc.all$group[cc.all$CASE==1 & cc.all$nrs_covid_case==0 &
              cc.all$adm28==0 & cc.all$inhosp==0] <- "C"
 
-## Cases tested in hospit
+## Cases tested in hospital
 cc.all$group[cc.all$CASE==1 & cc.all$nrs_covid_case==0 &
              (cc.all$adm28 > 0 | cc.all$inhosp > 0)] <- "B"
 table(cc.all$deathwithin28, cc.all$group, exclude=NULL)
 
-cc.all$group[cc.all$CASE==1 & cc.all$nrs_covid_case==0 &
-                 (cc.all$icu==1 | cc.all$hdu==1 | cc.all$deathwithin28==1)]  <- "A"
+## severe cases defined as CASE=1 AND (critical care or (death but not extra NRS death))
+cc.all$group[cc.all$CASE==1 &
+             ((cc.all$icu==1 | cc.all$hdu==1) |
+             (cc.all$deathwithin28==1 & cc.all$nrs_covid_case==0))]  <- "A"
 
 print(table(cc.all$CASE, cc.all$group, exclude=NULL))
 
@@ -248,6 +248,8 @@ cc.all <- merge(cc.all, casegroups, by=c("stratum"), all.x=T)
 table(cc.all$CASE, cc.all$casegroup)
 with(cc.all[cc.all$CASE==1, ], table(casegroup, deathwithin28, exclude=NULL))
 
+## all test-positive cases dying within 28 days are fatal cases -- excludes
+## deaths ascertained through NRS that were not test-positive 
[…]

### 2020-06-09 helperfunctions.R

@@ -7,18 +7,31 @@ tolower.exceptfirstchar <- function(x) {
     return(x)
 }
 
-select.union <- function(x, y, stratum) {
+select.union <- function(x, y, covariates, stratum) {
     ## select j as column of x to drop that maximizes loglik of
     ## clogit model using rowSums([x, -j])
     ## returns coeff and loglik of full model without a dropped column,
     loglik.dropcol <- numeric(ncol(x))
-    x.u <- rowSums(x) 
-    model.full <- summary(clogit(y ~ x.u + strata(stratum)))
+    x.u <- rowSums(x)
+    covariates <- as.data.frame(covariates)
+    names(covariates) <- paste0("c", 1:ncol(covariates)) 
+    cl.formula <- as.formula(paste("y ~ x.u", ifelse(is.null(covariates), "", "+"),
+                                   paste(names(covariates), collapse=" + "),
+                                   "+ strata(stratum)"))
+    cl.data <- data.frame(y=y, x.u=x.u, covariates, stratum=stratum)
+    
+    #browser("select")
+    model.full <- summary(clogit(formula=cl.formula, data=cl.data))
     beta.full <- model.full$coefficients[1, 1]
     loglik.full <- model.full$loglik[2]
     for(j in 1:ncol(x)) {
         x.u <- rowSums(x[, -j, drop=FALSE])
-        loglik.dropcol[j] <- summary(clogit(y ~ x.u + strata(stratum)))$loglik[2]
+        cl.formula <- as.formula(paste("y ~ x.u", ifelse(is.null(covariates), "", "+"),
+                                       paste(names(covariates), collapse=" + "),
+                                       "+ strata(stratum)"))
+        cl.data <- data.frame(y=y, x.u=x.u, covariates, stratum=stratum)
+        
+        loglik.dropcol[j] <- summary(clogit(formula=cl.formula, data=cl.data))$loglik[2]
     }
 
     j.max <- which.max(loglik.dropcol)
@@ -28,11 +41,11 @@ select.union <- function(x, y, stratum) {
                       loglik.full=loglik.full))
 }
 
-stepwise.union.dropcols <- function(x, y, stratum) {
+stepwise.union.dropcols <- function(x, y, covariates=NULL, stratum) {
     x.drop <- x
     stepwise.drop <- NULL
     for(j in 1:(ncol(x) - 1)) {
-        select.drop <- select.union(x=x.drop, y=y, stratum=stratum)
[…]

### 2020-06-09 incidencemortality.R

@@ -5,16 +5,16 @@ scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
 
 ####### incidence and mortality using national population estimates ######################
 
-case.freqs <- data.frame(Age=as.integer(rownames(case.freqs)),
-                         Females=as.integer(case.freqs[, 1]),
-                         Males=as.integer(case.freqs[, 2]))
-case.long <- reshape2::melt(case.freqs, id="Age")
+narrow.case.freqs <- data.frame(Age=as.integer(rownames(narrow.case.freqs)),
+                         Females=as.integer(narrow.case.freqs[, 1]),
+                         Males=as.integer(narrow.case.freqs[, 2]))
+case.long <- reshape2::melt(narrow.case.freqs, id="Age")
 colnames(case.long) <- c("Age", "Sex", "Cases")
 
-death.freqs <- data.frame(Age=as.integer(rownames(death.freqs)),
-                         Females=as.integer(death.freqs[, 1]),
-                         Males=as.integer(death.freqs[, 2]))
-death.long <- reshape2::melt(death.freqs, id="Age")
+narrow.death.freqs <- data.frame(Age=as.integer(rownames(narrow.death.freqs)),
+                         Females=as.integer(narrow.death.freqs[, 1]),
+                         Males=as.integer(narrow.death.freqs[, 2]))
+death.long <- reshape2::melt(narrow.death.freqs, id="Age")
 colnames(death.long) <- c("Age", "Sex", "Deaths")
 
 scotpop.long <- reshape2::melt(scotpop[, -2], id="Age")

### 2020-06-09 pharmaco.R

@@ -77,13 +77,13 @@ pander(summary(numdrugs.densities), table.style="multiline",
 table.numdrugsgr.num.icdch <- NULL
 for(numicd in levels(cc.severe$num.icdchapters.gr)) {
     x <- tabulate.freqs.regressions(varnames="numdrugsgr",
-                                    data=cc.severe[cc.severe$num.icdchapters.gr==numicd, ])[, 1:4]
+                                    data=cc.severe[notlisted & cc.severe$num.icdchapters.gr==numicd, ])[, 1:4]
     colnames(x)[1:2] <- c("controls", "cases")
     table.numdrugsgr.num.icdch <- rbind(table.numdrugsgr.num.icdch, x)
 }
 table.numdrugsgr.num.icdch <- data.frame(numdrugsgr=rep(levels(cc.severe$numdrugsgr), 3),
                                          table.numdrugsgr.num.icdch)
-colnames(table.numdrugsgr.num.icdch)[2:3] <- paste0(c("Controls (", "Cases ("), as.integer(table(cc.severe$CASE)), rep(")", 2))
+colnames(table.numdrugsgr.num.icdch)[2:3] <- paste0(c("Controls (", "Cases ("), as.integer(table(cc.severe[notlisted, ]$CASE)), rep(")", 2))
 
 ####################################################################################
 ## tabulate rate ratios for each numdrugs group, by agegr 
@@ -165,7 +165,7 @@ for(col in subparacols) {
                              cc.severe[nocare.notlisted, "CASE"])
     
     if(nrow(freqs.cc) > 1 & ncol(freqs.cc) > 1) {
-        if(sum(freqs.cc[2, ]) >= 50) { # if at least 10 in each cell
+        if(sum(freqs.cc[2, ]) >= 20) { # if at least 10 in each cell
             coeffs <- summary(clogit(formula=y ~ x + strata(stratum)))$coefficients
             if(coeffs[1, 5] < 0.001) {
                 subpara.coeffs <- rbind(subpara.coeffs,
@@ -190,17 +190,21 @@ print(subpara.coeffs)
 
 
 #############################################################################
+restrict <- notlisted & cc.severe$num.icdchapters < 3
 
 subparacols.forstepwisedrop <- subparacols[match(subparacols.keep, subparacols)] ## reduces to 61 subpara codes
 
[…]

### 2020-06-10 Covid_ethnicity_Scotland_PartB.Rmd
Ethnicity report and code written by Amanda

---
title: "Briefing paper on Ethnicity and COVID-19 status in Scotland"

author: Modelling & Research Cell, Public Health Scotland

date: "`r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{longtable}
  \usepackage{lscape}

output:
  pdf_document:
  keep_tex: true
  latex_engine: lualatex

# rmarkdown::render("Covid_ethnicity_Scotland.Rmd")

---

```{r settings, echo = FALSE, results='asis', warning=FALSE, error=FALSE}

# GLobal R options for this document
options(knitr.table.format="latex")
options(knitr.kable.NA='')

```

# Part B Tables
In this paper we attempt carry out a more detailed disaggregation of ethnicity presented in the "Briefing paper on Ethnicity and COVID-19 status in Scotland".

SMR recodes:-
1L - White Polish.
3B/3C/3F/3H - Pakistani/Bangladeshi.
3A/3G - Indian.
4A/4E/5C - Caribbean.
4B/4D/4Y - African.

ONOMAP "OnolyticsType" recodes:-
White - White and White Polish.
Black - Black African and Black Caribbean.
Bangladeshi - Muslim South Asian.
Muslim Indian - Muslim South Asian.
Pakistani - Muslim South Asian.
Pakistani Kashmir - Muslim South Asian.
Muslim - Muslim South Asian.

```{r ethnicB, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.ethnicB, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)") %>%
  add_header_above(c(" "=1, "SMR records"= 10)) %>%
  pack_rows("ONOMAP", 1, 5) %>%
  landscape() %>%
  kable_styling(latex_options="hold_position", font_size=8)

```

```{r ethnicsmr.testposB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.testpositives.demog.ethnicsmrB) <-
  gsub("_", " ", rownames(table.testpositives.demog.ethnicsmrB))
knitr::kable(table.testpositives.demog.ethnicsmrB,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.ethnicsmr",
             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmrB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
  kable_styling(latex_options="hold_position") %>% 
  column_spec(column=1, width="2.7cm")  %>%
  column_spec(column=2, width="2cm")  %>%
  column_spec(column=3, width="1.85cm")  %>%
  column_spec(column=4, width="2.5cm") %>%
  column_spec(column=5, width="1.5cm") %>%
  column_spec(column=6, width="2.5cm") %>%
  column_spec(column=7, width="1.5cm")
```

```{r ethnicsmr.hospB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.hospitalized.demog.ethnicsmrB) <-
  gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmrB))
knitr::kable(table.hospitalized.demog.ethnicsmrB,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmrB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
  kable_styling(latex_options="hold_position") %>% 
  column_spec(column=1, width="2.7cm") %>%
  column_spec(column=2, width="2cm")  %>%
  column_spec(column=3, width="1.85cm")  %>%
  column_spec(column=4, width="2.5cm") %>%
  column_spec(column=5, width="1.5cm") %>%
  column_spec(column=6, width="2.5cm") %>%
  column_spec(column=7, width="1.5cm")

```

```{r ethnic5smrB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic9smrB) <- gsub("_", " ", rownames(table.ethnic9smrB))
knitr::kable(table.ethnic9smrB,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnic9smrB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
  kable_styling(latex_options="hold_position") %>% 
  column_spec(column=1, width="2.7cm") %>%
  column_spec(column=2, width="1.85cm")  %>%
  column_spec(column=3, width="1.85cm")  %>%
  column_spec(column=4, width="2.5cm") %>%
  column_spec(column=5, width="1.5cm") %>%
  column_spec(column=6, width="2.8cm") %>%
  column_spec(column=7, width="1.5cm")
```

```{r ethnic9smrNRSB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic9smrNRSB) <- gsub("_", " ", rownames(table.ethnic9smrNRSB))
knitr::kable(table.ethnic9smrNRSB,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmrNRSB",
             col.names=c(clean.header(colnames(table.ethnic9smrNRSB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease OR a COVID-19 death with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

```{r testpositives.ethnic.smrB, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.ethnic.smrB, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:severity.smr}Severity of disease in test-positive cases, by ethnicity assigned in SMR records") %>%
  kable_styling(latex_options="hold_position")

```

\newpage

```{r onomap.testposB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.testpositives.demogB) <- gsub("_", " ", rownames(table.testpositives.demogB))
knitr::kable(table.testpositives.demogB,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.onomap",
             col.names=c(clean.header(colnames(table.testpositives.demogB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity based on name classification") %>%
  add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
  kable_styling(latex_options="hold_position") %>% 
  column_spec(column=1, width="2.7cm") %>%
  column_spec(column=2, width="1.95cm")  %>%
  column_spec(column=3, width="1.85cm")  %>%
  column_spec(column=4, width="2.5cm") %>%
  column_spec(column=5, width="1.5cm") %>%
  column_spec(column=6, width="2.5cm") %>%
  column_spec(column=7, width="1.5cm")

```

```{r onomap.hospB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.hospitalized.demogB) <- gsub("_", " ", rownames(table.hospitalized.demogB))
knitr::kable(table.hospitalized.demogB,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demogB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
  add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
  kable_styling(latex_options="hold_position") %>% 
  column_spec(column=1, width="2.7cm") %>%
  column_spec(column=2, width="2cm")  %>%
  column_spec(column=3, width="1.85cm")  %>%
  column_spec(column=4, width="2.5cm") %>%
  column_spec(column=5, width="1.5cm") %>%
  column_spec(column=6, width="2.5cm") %>%
  column_spec(column=7, width="1.5cm")

```

```{r onomapB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.severe.demogB) <- gsub("_", " ", rownames(table.severe.demogB))
knitr::kable(table.severe.demogB,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demogB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
  kable_styling(latex_options="hold_position") %>% 
  column_spec(column=1, width="2cm") %>%
  add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
  column_spec(column=2, width="1.85cm")  %>%
  column_spec(column=3, width="1.85cm")  %>%
  column_spec(column=4, width="2.5cm") %>%
  column_spec(column=5, width="1.5cm") %>%
  column_spec(column=6, width="2.5cm") %>%
  column_spec(column=7, width="1.5cm")

```

```{r onomapNRSB, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.severe.demogNRSB) <- gsub("_", " ", rownames(table.severe.demogNRSB))
knitr::kable(table.severe.demogNRSB,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomapNRSB",
             col.names=c(clean.header(colnames(table.severe.demogNRSB)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease OR any COVID-19 death with demographic variables including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```

```{r testpositives.ethnicB, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.ethnicB, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:severity}Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
  kable_styling(latex_options="hold_position")

```

```{r testpositives.carehomeB, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.carehomeB, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:carehome}Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
  kable_styling(latex_options="hold_position")

```

\newpage

```{r testpositives.healthboardB, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.healthboardB, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:board}Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
  landscape() %>%
  kable_styling(latex_options="hold_position")

```

### 2020-06-10 Covid_ethnicity_Scotlandhc.Rmd
Ethnicity report and code written by Amanda

---
title: "Briefing paper on Ethnicity and COVID-19 status in Scotland"

author: Modelling & Research Cell, Public Health Scotland

date: "`r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{longtable}
  \usepackage{lscape}

output:
  pdf_document:
  keep_tex: true
  latex_engine: lualatex

# rmarkdown::render("Covid_ethnicity_Scotland.Rmd")

---

# Background
There have been reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.   For example a preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95% CI 2.92-5.12); RR 2.11 (95% CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively).

Understanding whether such inequity in risk of COVID-19 by ethnicity pertains in Scotland and whether disease severity is worse is relevant to early anticipation of health care needs, triaging algorithms being used by NHS Hubs, and shielding policies.

The purpose of this report was to  i) enumerate the numbers of cases of  COVID-19 of non-White ethnicity ii) to examine whether there is any evidence of ethnic differences in COVID-19 disease in Scotland acknowledging that the size of non-White ethnic minority populations is small in Scotland and that ethnicity is not known with completeness.

# Methods 
The REACT-SCOT case-control study was set up to investigate risk factors for severe COVID-19 and to lay the basis for developing risk stratification algorithms. All persons to date  who tested positive  for SARS-CoV-2 nucleic acid were identified using the ECOSS virology laboratory systems. This system records CHI but not ethnicity.

Any additional persons with a registered death with a COVID-19 code ( suspected or probable ) were also identified.

A matched set of 10 controls per case matched for age, sex and GP practice was identified using the CHI database as part of the REACT-COVID19 case control study led by the Modelling & Research cell.

Data were extracted for  this analysis on 15th May 2020. They do not include the Lighthouse data as we do not have access to this through ECOSS as yet.

## Coding of ethnicity
Names of cases and controls were available to the CHI team. They used the ONOMAP software to assign predicted ethnicity to all cases and controls. This was output in anonymised format to the Modelling & Research Cell linkage team. ONOMAP is a software package developed by Mateos at UCL which used people's first and family names to estimate underlying ethnic group. It has been developed using algorithms built from the Worldnames database that covered both names and ethnic group in a large number of countries. ONOMAP assigns both predicted geographical region of origin and onolytic type. ONOMAP has previously been reported to have variable sensitivity for assigning non-White ethnic groups. ([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were used to assign ethnicity. Names classified "Muslim" in the ONOMAP classification were assigned to the South Asian group, as Scotland most people with Muslim names are South Asian.  The ONOMAP classification assigns a high proportion of names in Scotland to the category "Celtic" which does not distinguish Irish from Scottish names so that White Irish cannot be differentiated from White British. A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded (Clark & Buchanan ISD 2017).

Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of  recording of self assigned ethnic group in recent years.   Census categories are used. However there remains substantial missingness.   For all cases and controls the linkage team extracted i) the most recent ethnic group assignation from these two datasets  with a 10 year lookback period. Unknown or missing codes were excluded.  Concordance between the most recent SMR ethnic group   and the most frequent  SMR ethnic group  recorded was 98%.

The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other.  The Black category combines those who self-identify as Black British, Black Caribbean, Black African and Black Other.

The rate ratios shown are based on conditional logistic regression.  With the incidence density sampling used in this study, the odds ratios in a conditional logistic regression are equal to rate ratios.  The unconditional odds ratios calculated from the numbers of cases and controls in the tables are not the same as the conditional odds ratios and should not be used.

As other sociodemographic factors may vary by ethnicity we further adjusted analyses for residential care home status and Scottish Index of Multiple Deprivation (SIMD).

# Results
SMR based ethnicity was unavailable for approximately 25% of cases and 50% of controls. The ONOMAP approach  assigned ethnicity  in >99% of persons.  Given the substantial and differential missingness of SMR based ethnicity the principle value of showing the  SMR data here is to  inform on the likely absolute  numbers of cases   of non-White ethnicity rather than for the purposes of estimating rate ratios.

## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records

Table \ref{tab:crosstab} shows the agreement between the ONOMAP  (name based) assigned ethnicity and the self assigned SMR ethnicity.  The name classification assigns to the same as  SMR category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[5, 5]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.  The proportion with  non-White ethnicity  is low for all groups. Thus  although the sensitivity of ONOMAP for assigning SMR South Asian ethnicity is >99% the positive predictive value of ONOMAP assigned South Asian ethnicity is just 60%.

Clearly analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records.  *Please note that we do not consider the rate ratios here to be reliable because of the differential missingness.* Therefore in all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.

## Relation of test -positive, hospitalised and severe / fatal COVID-19 to ethnicity in the Scottish population
The tables below compare cases and controls  first with ethnicity based on SMR classification for

i)all test positive cases ii) all hospitalized or severe  test positive cases, iii) severe  test positive cases iv) severe test positive cases OR a COVID-19 death registration,    Severe test positive cases are those who  entered a critical care unit  or died within 28 days of a positive test.  Note that full follow up to 28 days has not yet completed for  the recent cases. The first column of rate ratios are univariate. Those in the second column are further adjusted for SIMD and care home status.

Subsequent tables show the same comparisons with ethnicity based on ONOMAP name classification. Tables \ref{tab:severity.smr} and \ref{tab:severity} show the distribution over severity categories by ethnicity. Tables \ref{tab:carehome} and \ref{tab:board} show the distribution of care home status and health board.  Note that  p -values  that are so extreme that they are less than the threshold for printing a floating point number are output as 0.

Tables comparing all test-positive individuals with controls and hospitalised test  positives are included by request.   However they should be interpreted with caution, as selection for testing is influenced by many factors that are changing through the epidemic as policy changes.  Similar considerations apply to proportion with severe  disease  among those who test positive.   **We  therefore strongly advise that the most useful data  for examining ethnic differences are those for severe test positive disease  and registered deaths as they will have least selection effects.**

There was no evidence of elevation in risk of  "severe  test positive disease" or "severe test positive disease or death"  in  any non-White ethnic group using  either the ONOMAP  or SMR  approach.

Using the named based classification for which there are much more data  for those of South Asian ethnicity the rate ratio was below 1  and was statistically significant.  A further disaggregation of the name based classification into Muslim South Asian and Non-Muslim South Asian ( data not presented) showed risk ratios of similar size below 1 for both subgroups.  For those of Chinese  ethnicity the name based classification showed lower test positive rates also.

Adjusting for SIMD and residential care home status  did not affect these estimates.

There were increased risk ratios for non-White ethnic groups for test positivity  using SMR  - however note our caveats on interpretation of   this - both because of differential missingness in SMR based ethnicity and because of selection effects on test positive levels.

# Conclusion

With current data obtaining a complete picture of the ethnic distribution of  people with COVID-19 disease in Scotland is difficult.  Historical medical records are incomplete and differentially so in their recording of ethnicity.  The numbers of  people in non-White ethnic groups in our population is relatively low compared with England and not known with certainty. Our attempt to improve completeness of ethnicity assignation is of some use for those of South Asian and Chinese origin but remains imperfect and  is not useful for those of Black ethnicity.

Here for transparency we have  nonetheless presented the results of analyses we specified in advance that we would undertake.  However there are important caveats on interpreting these data as we have outlined above.

The absolute numbers shown by ethnicity are useful in providing some indication of the likely  upper bound on numbers of cases in non-White ethnic groups.

We do not find evidence at this stage of elevations in risk of severe COVID-19 disease in non-White ethnic groups compared to White.

Further approaches need to be explored to improve ascertainment of ethnicity. One possibility may be to train a new  Scotland specific name based classifier against the SMR  based data and then apply that.

\newpage

```{r settings, echo = FALSE, results='asis', warning=FALSE, error=FALSE}

# GLobal R options for this document
options(knitr.table.format="latex")
options(knitr.kable.NA='')

```

# Tables

```{r ethnic, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.ethnic, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)") %>%
    add_header_above(c(" "=1, "SMR records"= 6)) %>%
    pack_rows("ONOMAP", 1, 5) %>%
    kable_styling(latex_options="hold_position")

```

```{r ethnicsmr.testpos, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.testpositives.demog.ethnicsmr) <-
    gsub("_", " ", rownames(table.testpositives.demog.ethnicsmr))
knitr::kable(table.testpositives.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.ethnicsmr",
             col.names=c(clean.header(colnames(table.testpositives.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm")  %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")
```

```{r ethnicsmr.hosp, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.hospitalized.demog.ethnicsmr) <-
    gsub("_", " ", rownames(table.hospitalized.demog.ethnicsmr))
knitr::kable(table.hospitalized.demog.ethnicsmr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog.ethnicsmr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized or severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```

```{r ethnic5smr, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic5smr) <- gsub("_", " ", rownames(table.ethnic5smr))
knitr::kable(table.ethnic5smr,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             col.names=c(clean.header(colnames(table.ethnic5smr)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

```{r ethnic5smrNRS, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnic5smrNRS) <- gsub("_", " ", rownames(table.ethnic5smrNRS))
knitr::kable(table.ethnic5smrNRS,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmrNRS",
             col.names=c(clean.header(colnames(table.ethnic5smrNRS)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease OR a COVID-19 death with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.8cm") %>%
    column_spec(column=7, width="1.5cm")
```

```{r testpositives.ethnic.smr, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.ethnic.smr, 
           escape=FALSE, 
             booktabs=TRUE,
       caption="\\label{tab:severity.smr}Severity of disease in test-positive cases, by ethnicity assigned in SMR records") %>%
    kable_styling(latex_options="hold_position")

```

\newpage

```{r onomap.testpos, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.testpositives.demog) <- gsub("_", " ", rownames(table.testpositives.demog))
knitr::kable(table.testpositives.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="testpos.onomap",
             col.names=c(clean.header(colnames(table.testpositives.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of test-positive status with ethnicity based on name classification") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="1.95cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```

```{r onomap.hosp, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.hospitalized.demog) <- gsub("_", " ", rownames(table.hospitalized.demog))
knitr::kable(table.hospitalized.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap.hosp",
             col.names=c(clean.header(colnames(table.hospitalized.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of hospitalized or severe disease with demographic variables including ethnicity based on name classification") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2.7cm") %>%
    column_spec(column=2, width="2cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```

```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.severe.demog) <- gsub("_", " ", rownames(table.severe.demog))
knitr::kable(table.severe.demog,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             col.names=c(clean.header(colnames(table.severe.demog)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```

```{r onomapNRS, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.severe.demogNRS) <- gsub("_", " ", rownames(table.severe.demogNRS))
knitr::kable(table.severe.demogNRS,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomapNRS",
             col.names=c(clean.header(colnames(table.severe.demogNRS)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6),
             caption="Associations of severe disease OR any COVID-19 death with demographic variables including ethnicity based on name classification") %>%
    kable_styling(latex_options="hold_position") %>% 
    column_spec(column=1, width="2cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    column_spec(column=2, width="1.85cm")  %>%
    column_spec(column=3, width="1.85cm")  %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.5cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="1.5cm")

```

```{r testpositives.ethnic, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.ethnic, 
           escape=FALSE, 
             booktabs=TRUE,
       caption="\\label{tab:severity}Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
    kable_styling(latex_options="hold_position")

```

```{r testpositives.carehome, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.carehome, 
           escape=FALSE, 
             booktabs=TRUE,
       caption="\\label{tab:carehome}Care home status in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
    kable_styling(latex_options="hold_position")

```

\newpage

```{r testpositives.healthboard, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(testpositives.healthboard, 
             escape=FALSE, 
             booktabs=TRUE,
             caption="\\label{tab:board}Health board of residence in test-positive cases, by ethnicity assigned from ONOMAP name classification") %>%
    kable_styling(latex_options="hold_position")

```

### 2020-06-12 casecontrol.Rmd
fixed to install required packages

@@
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland.
-HC - Axa Chair in Medical Informatics and Epidemiology.
+HC - AXA Chair in Medical Informatics and Epidemiology.
 TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology.
@@
 linenumbers: true
-bibliography: ../covid.bib
+bibliography: ./covid.bib
 
@@
   rticles::plos_article:
+    includes:
+      in_header: ./preamble.tex
     latex_engine: lualatex
@@
 
-csl: ../plos.csl
+always_allow_html: true
+csl: ./plos.csl
 urlcolor: blue
@@
 library(kableExtra)
+library(bookdown)
 
@@
 
+options(knitr.table.format = "latex") # adds tab: prefix to labels
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
@@
 
+table.ethnicsmr.missing <- paste.colpercent(with(cc.severe, table(is.na(ethnic5.smr), CASE)))
+
+table.notlisted.scrip <- paste.colpercent(with(cc.severe[notlisted, ], table(scrip.any, CASE)))
+table.notlisted.diag <- paste.colpercent(with(cc.severe[notlisted, ], table(diag.any, CASE)))
+table.notlisted.scripordiag <- paste.colpercent(with(cc.severe[notlisted, ], table(scripordiag, CASE)))
+
+table.lt60.notlisted.scripordiag <- paste.colpercent(with(cc.severe[cc.severe$AGE < 60 & notlisted, ], table(scripordiag, CASE)))
+
 ```
@@
 Severe cases were all those with a positive nucleic acid test for SARS-CoV-2 in the national database who had entered a critical care unit or died within 28 days of the first positive test.
-Ten controls per case matched for sex, age and primary care practice were selected from the population register.
+Seven controls per case matched for sex, age and primary care practice were selected from the population register.
 All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.
@@
 
-_Findings_ -- In a logistic regression using the age-sex distribution of the national population, the odds ratios were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
+_Findings_ -- There were `r length(which(cc.severe$CASE==1))` severe cases.
+In a logistic regression using the age-sex distribution of the national population, the odds ratios for severe disease were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
 In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
@@
 The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.
-Individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained.
+All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained for this study.
 Using the Community Health Index (CHI) identifier contained in ECOSS (the CHI number is a unique identifier used in all care systems in Scotland) linkage to other datasets was carried out.
@@
 For each test-positive case, we ascertained ten matched controls of the same sex, one-year age band and registered with the same primary care practice who were alive on the date of the first test in the case using the Community Health Index (CHI) database.
+After removing embarkations there were seven controls per case.
 As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
@@
 Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.
-For 54% of cases and 28% of controls self-assigned ethnicity, based on the categories used in the Census, had been recorded in Scottish Morbidity Records (SMR).
+For `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls and `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases self-assigned ethnicity, based on the categories used in the Census, had been recorded in Scottish Morbidity Records (SMR).
 Cross-tabulation of `r sum.xtabulate` records for which both name classification and SMR records of ethnicity were available showed that the ONOMAP algorithm had sensitivity of `r round(SouthAsian.sensitivity)`% and specificity of `r round(SouthAsian.specificity, 2)`% for classifying South Asian ethnicity, but misclassified most of those who identified as African, Caribbean or Black.
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.
-Cases of diabetes diagnosed since the last update of the register were identified through discharge codes and drug codes, and assigned to the diabetes type Other/Unknown category.
 
@@
 Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.
-Where an exclusion criterion such as having a pre-defined condition was applied to cases  this was also applied to controls as otherwise subsequent association estimates would be incorrect.
-Where the sample of cases and controls is restricted, this will generate strata that contain no cases but these strata will be ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
+Restriction of cases and controls, for instance to exclude those with any listed condition, may generate strata that do not contain at least one case and at least one control, but these strata are ignored by the conditional logistic regression model as they do not contribute to the conditional likelihood.
 With incidence density sampling, the odds ratios in conditional logistic regression models are equivalent to rate ratios.
@@
 In those aged 75+ years `r gsub("\\\\", "", table.agegr[4, 8])` of the cases and `r gsub("\\\\", "", table.agegr[4, 7])` of the controls had at least one listed condition.
-The proportion with at least one dispensed prescription was much higher in cases than in controls in each age group.
 Among those aged under 40 years, `r table.agegr[6, 2]` of the cases and `r table.agegr[6, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
+Differences in prescription rates  between cases an controls narrowed with increasing age.
 
 Over all age groups, `r freqs.all[nrow(freqs.all), 2]` of severe cases and `r freqs.all[nrow(freqs.all), 1]` of controls had at least one of the listed conditions.
-As shown in Table \ref{tab:listed.conditions.allages}, all the listed conditions were more frequent in cases than controls.
+As shown in Table \ref{tab:listed.conditions.allages}, all the listed conditions were more frequent in cases than controls except for immune conditions in the 75+ age group.
 The rate ratio associated with type 1 diabetes was higher than that for type 2 diabetes.
@@
 In multivariate analysis ischemic heart disease was not independently associated with severity whereas other heart disease remained strongly associated.
+In those without a listed condition `r table.notlisted.scripordiag[2, 2]` of the cases and `r table.notlisted.scripordiag[2, 1]` of the controls had either a recent admission or a prescription.
+In those aged under 60 years without a listed condition, `r table.lt60.notlisted.scripordiag[2, 2]` of the cases and `r table.lt60.notlisted.scripordiag[2, 1]` had either a recent admission or a prescription.
 
@@
 In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][5, 3]` for Type 1 diabetes and `r tables.agegr[[1]][6, 3]` for Type 2 diabetes.
-The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and within each age group dispensing of any prescription in the past year and any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
+The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and in each age group  group  any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
+Dispensing of any prescription in the past year was associated with severe disease  in multivariate analyses in the two younger age bands.
 Table \ref{tab:fatal} shows that in each age group the proportion of fatal cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.
+
+In a sensitivity analysis in which we also included deaths registered with mention of COVID-19 on the death certificate in the definition of severe cases (Table \ref{tab:agegr.withNRS}) the same pattern of differences in prior admissions and prescribing history and in listed conditions between cases and controls was found, with the difference in residential are home status being somewhat greater.
+Such deaths were not included in our primary outcome definition as misclassification rates for COVID assignation as cause are unknown.
 
@@
 These results are shown in Table \ref{tab:conditions}.
-In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
+In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 7 (eye) Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
 Note that hospital diagnoses classified under the pregnancy chapter here are derived from admissions with pregnancy related medical conditions to non-obstetric units only, as obstetric returns are not in the SMR01 dataset.
-In a multivariate analysis the strongest associations were with diagnoses in ICD chapter 2 (neoplasms).
+In a multivariate analysis the most significant association was with diagnoses in ICD chapter 2 (neoplasms).
 Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.
 This table is filtered to show only subchapters for which the univariate p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.
-This shows that many diagnoses are associated with markedly higher risk of severe COVID-19.
-Past hospital diagnoses of infections, pneumonia and acute respiratory diseases were strongly associated with severe COVID-19.
-Cardiovascular diagnoses associated with COVID-19 were not limited to heart disease but included also stroke and other circulatory disorders that are not designated as risk conditions.
+This shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.
+<!-- check why cerebrovascular disease does not come out here -->
 
 ## Associations of prescribed drugs with severe disease 
-As shown in Table \ref{tab:listed.conditions.allages} and supplementary tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} , the strongest risk factor for severe disease, apart from residence in a care home, is the encashment of at least one prescription in the last year.
+As shown in Table \ref{tab:listed.conditions.allages} and supplementary tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} ,  encashment of at least one prescription in the last year was associated with severe disease.
 The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
 In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` respectively.
-About one third of controls aged over 75 had not encashed a prescription in the previous year.
 
@@
 In a multivariate analysis of all chapters, most of these associations were weaker.
-The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and 2 (cardiovascular).
-Other chapters associated with severe disease were chapters 4 (central nervous system), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
+The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and chapters 4 (central nervous system).
+Other chapters associated with severe disease were 2 (cardiovascular),  5 (infections), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
 
@@
 Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now.
-The only non-White ethnic group with any sizeable numbers is the South Asian category and we found no evidence of any elevation in risk in this group compared to Whites.
+The only non-White ethnic group with any sizeable numbers is the South Asian category and we found no clear evidence of any elevation in risk in this group compared to Whites.
 Reports from England [@the_opensafely_collaborative_opensafely_2020] found elevation in risks for some non-White groups.
@@
 Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.
-However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were much higher in cases than in controls.
+However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were higher in cases than in controls.
 In all age groups, very few of the fatal cases had not had either a hospital admission in the past five years or a dispensed prescription in the past year.
 
-A striking finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
+An important finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
 Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
-Also associated were those in the cardiovascular, nervous system and nutritional and blood chapters.
+Also associated were those in the  nervous system,  cardiovascular, and nutritional and blood chapters.
 Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.
@@
 As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.
-A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 1)` associated with a 20-year increase in age is far stronger than that associated with common diseases such as Type 2 diabetes and asthma that are listed as conditions associated with high risk.
+A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 1)` associated with a 20-year increase in age is stronger than that associated with common diseases such as Type 2 diabetes and asthma that are listed as conditions associated with high risk.
 A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.
@@
 Associations with ethnicity and other sociodemographic factors are not necessarily generalizable from Scotland to other populations.
-
-This case-control study is limited to test-positive cases, excluding deaths with no record of a positive test where COVID-19 was mentioned on the death certificate as an underlying or contributing cause.
-Up to 13 May 2020 an additional 1200 such deaths had been reported by the National Records of Scotland.
-Future analyses of this study will include sensitivity analyses of the extent to which the results are changed by including these deaths, but linked data are not yet available to us.
-Apart from residential care home status, we do not expect most other risk factors to differ markedly between those who died from COVID-19 without being tested and those who died after testing.
 
@@
     column_spec(column=4, width="2.5cm") %>%
-    group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
-    group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
-    group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
+    kableExtra::group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
+    kableExtra::group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
+    kableExtra::group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
                hline_before=TRUE, latex_gap_space="10pt") %>% 
@@
 rownames(table.agegr) <- replace.names(rownames(table.agegr))
-knitr::kable(table.agegr[-7, ], # omit non-diabetic category
+table.agegr.print <- table.agegr[c(1:4, 6, 8:nrow(table.agegr)), ] 
+knitr::kable(table.agegr.print, 
              escape=FALSE, 
@@
              label="agegr",
+			 linesep=c(rep('', 4), '\\addlinespace', rep('', 9)), 
              align=rep("c", 8), 
@@
     column_spec(column=1, width="4cm") %>%
-    column_spec(column=2:9, width="2cm") %>%
+    column_spec(column=2:3, width="1.6cm") %>%
+    column_spec(column=4:7, width="1.8cm") %>%
+    column_spec(column=8:9, width="2cm") %>%
     add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
@@
              caption="Associations of severe disease with listed conditions over all age groups") %>%
-    # group_rows(" ", 4, nrow(table.agegr.all[-4, ])) %>%
+    # kableExtra::group_rows(" ", 4, nrow(table.agegr.all[-4, ])) %>%
     column_spec(column=1, width="5cm") %>%
@@
              caption="Proportions of fatal cases and matched controls without and with a  dispensed prescription or hospital diagnosis, by age group") %>%
-   group_rows("Age <60", 1, 2) %>%
-   group_rows("Age 60-74", 3, 4) %>%
-   group_rows("Age 75+", 5, 6) %>%
+   kableExtra::group_rows("Age <60", 1, 2) %>%
+   kableExtra::group_rows("Age 60-74", 3, 4) %>%
+   kableExtra::group_rows("Age 75+", 5, 6) %>%
    kable_styling(latex_options="hold_position")
@@
 \newpage
+\beginsupplement
 
@@
              caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
-#   group_rows(" ", 4, nrow(tables.agegr[[1]]) - 1) %>%
+#   kableExtra::group_rows(" ", 4, nrow(tables.agegr[[1]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
@@
              caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
- #   group_rows(" ", 4, nrow(tables.agegr[[2]]) - 1) %>%
+ #   kableExtra::group_rows(" ", 4, nrow(tables.agegr[[2]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
@@
              caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
-#    group_rows(" ", 4, nrow(tables.agegr[[3]]) - 1) %>%
+#    kableExtra::group_rows(" ", 4, nrow(tables.agegr[[3]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
@@
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r agegr.withNRS, echo=FALSE, warning=FALSE, message=FALSE}
+
+table.agegr.withNRS <- readRDS(table.agegr, file="table.agegr.withNRS.rds")
+
+rownames(table.agegr.withNRS) <- replace.names(rownames(table.agegr.withNRS))
+table.agegr.withNRS.print <- table.agegr.withNRS[c(1:4, 6, 8:nrow(table.agegr.withNRS)), ] 
+knitr::kable(table.agegr.withNRS.print, 
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="agegr.withNRS",
+			 linesep=c(rep('', 4), '\\addlinespace', rep('', 9)), 
+             align=rep("c", 8), 
+             caption="Frequencies of risk factors in cases and controls when deaths with mention of COVID-19 on the death certificate were included as cases, by age group") %>%
+    column_spec(column=1, width="4cm") %>%
+    column_spec(column=2:3, width="1.6cm") %>%
+    column_spec(column=4:7, width="1.8cm") %>%
+    column_spec(column=8:9, width="2cm") %>%
+    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
+                       "60-74 years"=2, "75+ years"=2)) %>%
     landscape() %>%
@@
              caption="Stepwise regression: variables retained in model for severe disease") %>%
-    group_rows("SIMD - quintile 1 as reference", 2, 5, bold=FALSE) %>%
-    group_rows("Diabetes - non-diabetic as reference", 6, 8, bold=FALSE) %>%
+    kableExtra::group_rows("SIMD - quintile 1 as reference", 2, 5, bold=FALSE) %>%
+    kableExtra::group_rows("Diabetes - non-diabetic as reference", 6, 8, bold=FALSE) %>%
     kable_styling(latex_options="hold_position")

### 2020-06-12 pharmaco.Rmd
fixed to install required packages

@@
 author: 
-
   - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
@@
 linenumbers: true
-bibliography: ../covid.bib
+bibliography: ./covid.bib
 date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
@@
     includes:
-      in_header: ../preamble.tex
+      in_header: ./preamble.tex
     latex_engine: lualatex
@@
     keep_tex: true
-csl: ../../csl/styles/the-new-england-journal-of-medicine.csl
+csl: ../csl/styles/the-new-england-journal-of-medicine.csl
 urlcolor: red
@@
 
-*Findings* -- We began by examining associations with polypharmacy, defined by total number of drug classes dispensed.
-With those who had no dispensed prescriptions as reference group, rate ratios (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
-Over all age groups the association with polypharmacy was strongest in those without hospital admissions in the past three years: in this group the rate ratio associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.num.icdch[6, 4]`.
-The drug classes associated with COVID-19 included many drug classes previously identified as indicators of overprescribing.
-A backwards elimination procedure identified proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy.
-Further analysis of this association showed a dose-response relationship with the average daily dose dispensed and that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.
-The association remained after adjusting for covariates prespecified as likely confounders, and in those not prescribed any of the `r nrow(coeffs.propensity)` drug classes used to construct a propensity score.
-However a comparison of time windows did not show a stronger association with recent exposure than with less recent exposure.
+*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed: with those who had no dispensed prescriptions as reference group, the rate ratio (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes was XX.
+In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was XX.
+The regression coefficients for number of cardiovascular drug classes and number of other drug classes in a joint model were XX and XX respectively; thus the association with polypharmacy was specifically with non-cardiovascular drugs.
+A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
+Further analysis of the association with proton pump inhibitors showed a dose-response relationship with the average daily dose dispensed, that the association remained that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.
 
@@
 This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.
-Of all drug classes, proton pump inhibitors make the largest independent contribution to this association.
-As prescriptions of this drug class are a recognised indicator for overprescribing of drugs, it is difficult to distinguish between causality and confounding as explanations for this association.
-However the strength of the association in younger age groups and in those not on other drugs with which proton pump inhibitors are widely co-prescribed raise concern about a possible causal effect that requires further investigation in other settings.
+Of all drug classes, analgesics and proton pump inhibitors make the largest independent contribution to this association.
+As prescriptions of these drug classes are recognised indicators for overprescribing, it is difficult to distinguish between causality and confounding as explanations for this association.
+This requires further investigation in other settings.
 In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.
@@
 
-The primary aim of this study was to investigate the association of severe COVID-19 with drug prescribing.
+The objective of this study was to investigate the association of severe COVID-19 with drug prescribing.
 Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.
@@
 Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.
-The last drug class retained by backward selection, and thus the variable with largest independent explanatory power for the effect of polypharmacy, was proton pump inhibitors.
+The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.
+
+## Associations of severe COVID-19 with analgesic use
 
@@
     column_spec(column=1, width="2cm") %>%
-    group_rows(paste0("No listed condition (",
+    kableExtra::group_rows(paste0("No listed condition (",
                       freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
                1, 6) %>%
-    group_rows(paste0("Listed condition (",
+    kableExtra::group_rows(paste0("Listed condition (",
                       freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
                7, 12) %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+```{r numdrugs.cardio, echo=FALSE, warning=FALSE, message=FALSE}
+
+colnames(table.jointcardiovasc)[c(1, 4:5)] <- c("Drug category", "Odds ratio (95\\% CI)",
+                                                "$p$-value")
+
+knitr::kable(table.jointcardiovasc,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="numdrugs.cardio",
+             row.names=FALSE,
+             caption="Multivariate regression of severe COVID-19 on number of cardiovascular and number of noncardiovascular drug classes in dispensed prescriptions, among those without listed conditions") %>%
+    column_spec(column=1, width="2.5cm") %>%
+    add_header_above(c(" "=1, "Mean number of drug classes"=2)) %>%
     kable_styling(latex_options="hold_position")
@@
     column_spec(column=1, width="2cm") %>%
-    group_rows("Age 0-59", 1, 6) %>%
-    group_rows("Age 60-74", 7, 12) %>%
-    group_rows("Age 75+", 13, 18) %>%
-    group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
-    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
-    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
+    kableExtra::group_rows("Age 0-59", 1, 6) %>%
+    kableExtra::group_rows("Age 60-74", 7, 12) %>%
+    kableExtra::group_rows("Age 75+", 13, 18) %>%
+    kableExtra::group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
+    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
+    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
     kable_styling(latex_options="hold_position")
@@
     column_spec(column=1, width="2cm") %>%
-    group_rows("No admissions", 1, 6) %>%
-    group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
-    group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
+    kableExtra::group_rows("No admissions", 1, 6) %>%
+    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
+    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
     kable_styling(latex_options="hold_position")
@@
     kable_styling(latex_options="hold_position") %>%
-    group_rows("Ages 0-59 years", 1, 5) %>%
-    group_rows("Ages 60-74 years", 6, 10) %>%
-    group_rows("Ages 75+ years", 11, 15) %>%
+    kableExtra::group_rows("Ages 0-59 years", 1, 5) %>%
+    kableExtra::group_rows("Ages 60-74 years", 6, 10) %>%
+    kableExtra::group_rows("Ages 75+ years", 11, 15) %>%
     landscape()

### 2020-06-12 casecontrol.R

@@ -1,6 +1,33 @@
 ## analysis script for case-control study
 rm(list=ls())
 
+## required system packages: libcurl4-openssl-dev, pandoc, pandoc-citeproc
+##   libssl-dev libxml2-dev
+## texlive-full or at least texlive-latex-extra, texlive-luatex
+
+## install all required R packages
+list.of.packages = c("car", 
+                     "survival", 
+                     "MASS", 
+                     "wevid", 
+                     "rmarkdown",
+                     "bookdown",
+                     "rticles",
+                     "kableExtra",
+                     "pander", 
+                     "ggplot2", 
+                     "doParallel", 
+                     "reshape2", 
+                     "readxl", 
+                     "DescTools", 
+                     "icd.data", 
+                     "gam", 
+                     "dplyr")
+
+new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
+if(length(new.packages) > 0) {install.packages(new.packages)}
+lapply(list.of.packages, require, character.only=T)
+
 library(car)
 library(survival)
 library(MASS)
@@ -141,7 +168,6 @@ print(with(cc.all, table(CHI.Explanation, diag.any)))
 cc.all <- subset(cc.all, subset=is.na(CHI.Explanation) |
                              CHI.Explanation=="Current and no history" |
                              CHI.Explanation=="Current with history")
-
 
 ## exclude controls already dead on date of test of case they were matched to
 controls.deceased <- with(cc.all, CASE==0 &

### 2020-06-12 pharmaco.R

@@ -5,7 +5,42 @@
 
 
 ########################################################
+## tabulate rate ratios for each num drug group by listed.any
+## this shows that the polypharmacy association is only in those without listed conditions
 
+table.numdrugsgr.listed.any <- NULL
+for(cat in levels(cc.severe$listed.any)) {
+    x <- tabulate.freqs.regressions(varnames="numdrugsgr",
+                                    data=cc.severe[cc.severe$listed.any==cat, ])[, 1:4]
+    colnames(x)[1:2] <- c("controls", "cases")
+    table.numdrugsgr.listed.any <- rbind(table.numdrugsgr.listed.any, x)
+}
+table.numdrugsgr.listed.any <- data.frame(numdrugsgr=rep(levels(cc.severe$numdrugsgr), 2),
+                                     table.numdrugsgr.listed.any)
+colnames(table.numdrugsgr.listed.any)[2:3] <- paste0(c("Controls (", "Cases ("), as.integer(table(cc.severe$CASE)), rep(")", 2))
+
+########################################################################
+
+## in those without listed conditions, fit joint model for num cardiovascular and num non-cardiovascular drugs
+
+tj <- summary(clogit(formula=CASE ~ numdrugs.cardiovasc + numdrugs.notcardiovasc + strata(stratum),
+                        data=cc.severe[notlisted, ]))$coefficients
+
+tj <- as.data.frame(tj)
+tj$u.ci <- or.ci(tj[, 1], tj[, 3])
+tj$pvalue <- pvalue.latex(tj[, 5])
+means.cardio <- with(cc.severe[notlisted, ], tapply(numdrugs.cardiovasc, CASE, mean)) 
+means.notcardio <- with(cc.severe[notlisted, ], tapply(numdrugs.notcardiovasc, CASE, mean))
+
+means.numdrugs <- rbind(means.cardio, means.notcardio)
+drugcat <- c("Cardiovascular", "Other")
+
+table.jointcardiovasc <- data.frame(drugcat, round(means.numdrugs, 1),
+                                    tj[, c("u.ci", "pvalue")])
+colnames(table.jointcardiovasc)[2:3] <- paste0(c("Controls (", "Cases ("),
+                                               table(cc.severe[notlisted, ]$CASE), rep(")", 2))
+
+#################################################################
[…]

### 2020-06-13 pharmaco.Rmd
Updated to use a data.table object for case-control dataset

@@
 
-*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed: with those who had no dispensed prescriptions as reference group, the rate ratio (95% CIs) for COVID-19 associated with dispensing of 12 or more drug classes was XX.
-In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was XX.
-The regression coefficients for number of cardiovascular drug classes and number of other drug classes in a joint model were XX and XX respectively; thus the association with polypharmacy was specifically with non-cardiovascular drugs.
+*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed.
+With those who had no dispensed prescriptions as reference group, the rate ratio (95% CI) for COVID-19 associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.listed.any[6, 4]` in those without listed conditions.
+In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.
+The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
 A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
-Further analysis of the association with proton pump inhibitors showed a dose-response relationship with the average daily dose dispensed, that the association remained that the association was strongest in younger age groups: the rate ratio for COVID-19 associated with ever-use of a proton pump inhibitor varied from `r table.everuse.protonpump[3, 3]` in those aged 75 years and over to `r table.everuse.protonpump[1, 3]` in those aged under 60 years.
+Further analysis of the association with proton pump inhibitors showed a dose-response relationship that was strongest in younger age groups: the rate ratio associated with average dispensed dose of 2 or more DDDs per day was `r table.dosegr.protonpump[5, 4]` in those under 60 years and `r table.dosegr.protonpump[15, 4]` in those aged under 60 years.
 
@@
 
-Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescription.
-
+Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions.
+<!---
 The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).
@@
 Risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
+--->
 
-To study further the relationship with polypharmacy, we restricted the analysis to those not resident in a care home and without a diagnosis of a listed condition.
-The rationale for this restriction was to reduce the frequency of factors including multi-morbidity and frailty which might be expected to confound the association between polypharmacy and COVID-19.
-In people with listed conditions, we might expect prescribing of multiple drugs in accordance with evidence-based guidelines for management of these conditions.
-Residents in care homes are at high risk of severe COVID-19 for reasons unrelated to polypharmacy, and the profile of drugs most commonly prescribed to people in care homes differs from the profile of drugs most commonly prescribed to people living independently.
+To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.
+As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.
+In a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
 
 ## Associations of severe COVID-19 with drug classes
-We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19.
+We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.
 Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.
@@
 Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.
-The drug classes dropped at this stage included some of the most widely-prescribed drug classes that showed univariate associations with COVID-19 in Table \ref{tab:subpara.coeffs}, including lipid-lowering drugs, non-steroidal antiinflammatory agents and angiotensin-converting enzyme (ACE) inhibitors.
 Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-Cardiovascular drugs retained among these `r length(rows.print)` classes were loop diuretics and calcium channel blockers.
-Four classes of drugs acting on the central nervous system were retained: other antidepressant drugs (excludes tricyclics and selective serotonin reuptake inhibitors), drugs used in nausea and vertigo, antipsychotic drugs, and non-opioid analgesics.
 The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.
 
-## Associations of severe COVID-19 with analgesic use
-
-## Associations of severe COVID-19 with proton pump inhibitor use
-As the association of COVID-19 and exposure to proton pump inhibitors is not obviously explained by established co-morbidities, we explored this association in more depth.
-As this examined a specific exposure, rather than polypharmacy which may have different profiles in care home residents or those diagnosed with listed conditions, the observations were not restricted to exclude care home residents or those diagnosed with a listed condition.
-Distinguishing between causality and confounding is challenging because proton pump inhibitors are usually co-prescribed with other drugs.
-We used several approaches:
+The associations with these drug classes were examined in more depth.  For studies of these specific exposures the observations were not restricted to exclude those with listed conditions. 
+ Distinguishing between causality and confounding is challenging because these drug classes  are usually co-prescribed with other drugs.  We used several approaches:
 
@@
 
-* Stratification by propensity score for ever-use of a proton pump inhibitor.
+* Stratification by propensity score.
+
+## Associations of severe COVID-19 with analgesic use
+
+## Associations of severe COVID-19 with proton pump inhibitor use
 

### 2020-06-13 casecontrol.R

@@ -22,7 +22,8 @@ list.of.packages = c("car",
                      "DescTools", 
                      "icd.data", 
                      "gam", 
-                     "dplyr")
+                     "dplyr",
+                     "data.table")
 
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
 if(length(new.packages) > 0) {install.packages(new.packages)}
@@ -41,7 +42,7 @@ library(readxl)
 library(DescTools)
 library(icd.data)
 library(gam)
-library(dplyr)
+library(data.table)
 
 registerDoParallel(cores=2)
 
@@ -63,7 +64,7 @@ if(old) {
     scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
     controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-05.rds")
 } else {
-    cc.all <- readRDS("./data/CC_linked_ANON_20200515.rds")
+    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"))
     diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
     procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds")
     scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
@@ -612,16 +613,16 @@ cc.severe <- merge(cc.severe, scrips.wide, by="ANON_ID", all.x=TRUE)
 
 bnfcols <- grep("^BNF", colnames(cc.severe))
 for(j in bnfcols) {
-    cc.severe[, j][is.na(cc.severe[, j])] <- 0
-    cc.severe[, j][cc.severe[, j] > 1] <- 1
-    cc.severe[, j] <- as.factor(cc.severe[, j])
+    cc.severe[[j]][is.na(cc.severe[[j]])] <- 0
+    cc.severe[[j]][cc.severe[[j]] > 1] <- 1
+    cc.severe[[j]] <- as.factor(cc.severe[[j]])
 }
 
 ## merge BNF chapters, one variable per subpara
 chnums = 1:13
 cc.severe <- merge.bnfsubparas(chnums=chnums, data=cc.severe)
 subparacols <- grep("^subpara\\.", names(cc.severe))
-x <- apply(cc.severe[, subparacols], 2, as.character)
[…]

### 2020-06-13 helperfunctions.R

@@ -1,4 +1,3 @@
-
 ## helperfunctions for COVID analyses
 
 tolower.exceptfirstchar <- function(x) {
@@ -20,7 +19,6 @@ select.union <- function(x, y, covariates, stratum) {
                                    "+ strata(stratum)"))
     cl.data <- data.frame(y=y, x.u=x.u, covariates, stratum=stratum)
     
-    #browser("select")
     model.full <- summary(clogit(formula=cl.formula, data=cl.data))
     beta.full <- model.full$coefficients[1, 1]
     loglik.full <- model.full$loglik[2]
@@ -263,9 +261,9 @@ tabulate.bnfsection <- function(chnum, outcome="CASE", data=cc.severe, minrowsum
     ## now fix colnames and set missing to 0
     bnfcols <- grep("^section.", colnames(cc.bnf.ch))
     for(j in bnfcols) {
-        cc.bnf.ch[, j][is.na(cc.bnf.ch[, j])] <- 0
-        cc.bnf.ch[, j][cc.bnf.ch[, j] > 1] <- 1
-        cc.bnf.ch[, j] <- as.factor(cc.bnf.ch[, j])
+        cc.bnf.ch[[j]][is.na(cc.bnf.ch[[j]])] <- 0
+        cc.bnf.ch[[j]][cc.bnf.ch[[j]] > 1] <- 1
+        cc.bnf.ch[[j]] <- as.factor(cc.bnf.ch[[j]])
     }
     
     bnf.ch <- colnames(cc.bnf.ch)[-(1:3)]
@@ -307,9 +305,9 @@ tabulate.bnfparas <- function(chnum, outcome="CASE", data=cc.severe, minrowsum=2
     ## now fix colnames and set missing to 0
     bnfcols <- grep("^para.", colnames(cc.bnf.ch))
     for(j in bnfcols) {
-        cc.bnf.ch[, j][is.na(cc.bnf.ch[, j])] <- 0
-        cc.bnf.ch[, j][cc.bnf.ch[, j] > 1] <- 1
-        cc.bnf.ch[, j] <- as.factor(cc.bnf.ch[, j])
+        cc.bnf.ch[[j]][is.na(cc.bnf.ch[[j]])] <- 0
+        cc.bnf.ch[[j]][cc.bnf.ch[[j]] > 1] <- 1
+        cc.bnf.ch[[j]] <- as.factor(cc.bnf.ch[[j]])
     }
     
     bnf.ch <- colnames(cc.bnf.ch)[-(1:3)]
[…]

### 2020-06-13 propensity.R

@@ -6,7 +6,7 @@ subpara.colnames <- grep("^subpara\\.", colnames(cc.hosp), value=TRUE)
 subpara.colnames <- grep("Proton pump", subpara.colnames, invert=TRUE, value=TRUE) 
 train.cols <- match(c("y.protonpump", "AGE", "sex", subpara.colnames), colnames(cc.hosp))
 
-train.data <- cc.hosp[, train.cols]
+train.data <- subset(cc.hosp, select=train.cols)
 
 colnames(train.data) <-
     gsub("(subpara\\.[0-9]+)(.+)", "\\1", colnames(train.data))                
@@ -14,7 +14,7 @@ colnames(train.data) <-
 ## select subparas with frequency > 0.1
 
 subparanames <- grep("^subpara\\.", colnames(train.data), value=TRUE)
-subparanames.freqs <- apply(train.data[, subparanames], 2,
+subparanames.freqs <- apply(subset(train.data, select=subparanames), 2,
                            function(x) sum(as.integer(x))) / nrow(train.data)
 
 keep <- subparanames.freqs > 0.05
@@ -41,7 +41,7 @@ names(coeffs.full) <- subpara.colnames[match(names(coeffs.full),
                                              gsub("(subpara\\.[0-9]+)(.+)", "\\1",
                                                   subpara.colnames))]
 
-newdata.x <- cc.severe[, match(names(coeffs.full), colnames(cc.severe))]
+newdata.x <- subset(cc.severe, select=match(names(coeffs.full), colnames(cc.severe)))
 newdata.x <- as.matrix(newdata.x)
 ## convert from character to numeric
 newdata.x <- matrix(as.numeric(newdata.x), nrow=nrow(newdata.x))

### 2020-06-15 pharmaco.Rmd
refactored to create new file drugs.R

@@
 Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions.
-<!---
-The association of COVID-19 with polypharmacy was strongest in those who had no hospital diagnoses in the previous three years (because they had not been admitted to hospital).
-In this group, with the no dispensed prescriptions category as baseline, the rate ratio ranged from `r table.numdrugsgr.num.icdch[2, 4]` in those with 1 to 3 drug classes dispensed to `r table.numdrugsgr.num.icdch[6, 4]` in those with more than 12 drug classes dispensed.
-
-Risk of COVID-19 was strongly related to polypharmacy in all age bands: with no dispensed prescriptions as baseline category, the rate ratios (95% CIs) associated with dispensing of 12 or more drug classes were `r table.numdrugsgr.agegr[6, 4]` in those aged less than 60 years, `r table.numdrugsgr.agegr[12, 4]` in those aged 60-74 years, and `r table.numdrugsgr.agegr[18, 4]` in those aged 75 years and over.
---->
 
@@
 
+```{r dosegr.opiate, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.dosegr.opiate,
+             escape=FALSE,
+             label="dosegr.opiate", 
+             booktabs=TRUE,
+             row.names=TRUE, 
+             col.names=c(clean.header(colnames(table.dosegr.opiate)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Associations of severe COVID-19 with dispensed average daily dose (MME) of opiate grouped into categoriesb") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kableExtra::group_rows("Average dispensed daily dose (MME) of opiates", 1, 4) %>%
+        kable_styling(latex_options="hold_position") %>%
+        landscape()
+
+```
+
 ```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}

### 2020-06-15 bnfcodes.R

@@ -40,3 +40,10 @@ bnfchapters$shortname <- substr(bnfchapters$chaptername, 1, truncate.at)
 bnfchapters$chapternum[bnfchapters$chapternum > 14] <- 14
 bnfchapters$shortname[bnfchapters$chapternum==14] <- "Other" 
 bnfchapters$shortname[bnfchapters$chapternum==4] <- "Nervous" 
+
+## chemical codes
+bnfchemicalcodes <- read_excel("./BNF_Code_Information.xlsx", sheet=1)[, 1:10]
+colnames(bnfchemicalcodes) <- c(colnames(bnfsubparacodes),
+                                "chemicalname", "chemicalcode")
+# bnfchemicalcodes <- subset(bnfchemicalcodes, subset=!duplicated(chemicalcode))
+

### 2020-06-15 casecontrol.R

@@ -57,61 +57,17 @@ nrs <- FALSE # uncomment to exclude NRS-only deaths
 stepwise <- TRUE   
 stepwise <- FALSE ## uncomment to save time if old version still valid
 
-if(old) {
-    cc.all <- readRDS("./data/CC_linked_ANON_20200501 (2).rds")
-    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200501.rds") 
-    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200501.rds")
-    scrips <- readRDS("./data/CC_PIS_x15_ANON_20200501.rds")
-    controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-05.rds")
-} else {
-    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"))
-    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
-    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds")
-    scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
-    scrips <- readRDS(scrips.filename)[, c("ANON_ID", "bnf_paragraph_code",
-                                           "bnf_paragraph_description")] 
-    controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-15.rds")
-
-    ## scrips should be 2 tables to save space
-    ## one record per scrip
-    scripvars <- c("ANON_ID", "dispensed_date", "bnf_paragraph_code",
-                   "formulation_code",
-                   "item_strength", "item_strength_uom", "item_code",
-                   "num_items", "quantity")
-    ## lookup table for subpara code and item code
-    drugvars <- c("bnf_paragraph_code", "bnf_paragraph_description",
-                  "item_code", "approved_name")
-    ## other files for diagnoses, procedures and scrips within the time limit
-    ## data/CC_SMR01_ICD10_25_ANON_20200515.rds
-    ## data/CC_SMR01_OPCS4_MAIN.25_ANON_20200501.rds
-    ## data/CC_PIS_15_ANON_20200515.rds
-
-    scrips.protonpump <-
-        readRDS(scrips.filename) %>%
-        subset(bnf_paragraph_code=="0103050")
-}
-
-#################### add fields to scrips ########################################
-
[…]

### 2020-06-15 pharmaco.R

@@ -1,28 +1,5 @@
 ## pharmacoepi paper
 
-####### derived variables  ##########################
-
-conversionMME <- 
-"
-Opioid 	Conversion Factor 
-Codeine 	0.15
-Fentanyl	2.4
-Hydrocodone 	1
-Hydromorphone 	4
-
-Morphine 	1
-Oxycodone 	1.5
-Oxymorphone 	3
-
-Methadone hydrochloride
-
-Morphine sulphate
-
-Pentazocine hydrochloride
-Pentazocine lactate
-Pethidine hydrochloride
-Nalbuphine hydrochloride
-"
 
 ########################################################
 ## tabulate rate ratios for each num drug group by listed.any
@@ -228,6 +205,35 @@ print(stepwise.drop.subparas)
 table.drugs.nocare.notlisted <- tabulate.freqs.regressions(varnames=drugs, 
                                                            data=cc.severe[nocare.notlisted, ])
 
+#### opiates  #####################################
+####### tabulate dose-response effect by opiate dose group ###############################
+
+opiates.varnames <- c("dosegr.opiate", "care.home", "neoplasm.any")
+
+table.dosegr.opiate <- tabulate.freqs.regressions(varnames=opiates.varnames,
+                                                  data=cc.severe)
+
+#### opiate effect by time window #######################
+
+cc.severe$opiate.exposure.nonrecent <- as.integer(cc.severe$opiateMME.interval2 > 0 | cc.severe$opiateMME.interval3 > 0)
+cc.severe$opiate.exposure.recent <- as.integer(cc.severe$opiateMME.interval1 > 0)
+
+cc.severe$opiate.exposurecat <- integer(nrow(cc.severe))
+cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==0 & cc.severe$opiate.exposure.recent==0] <- 0
+cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==1 & cc.severe$opiate.exposure.recent==0] <- 1
+cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==0 & cc.severe$opiate.exposure.recent==1] <- 2
+cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==1 & cc.severe$opiate.exposure.recent==1] <- 3
+
[…]

### 2020-06-16 pharmaco.Rmd
last commit before new linkage

@@
 
+```{r timewin.opiate, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.timewindow.opiate,
+             escape=FALSE,
+             label="timewindow.opiate", 
+             booktabs=TRUE,
+             col.names=c(clean.header(colnames(table.timewindow.opiate)[1:2]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption=paste0("Associations of severe disease with dispensed prescriptions for opiate analgesics by ", TW, "-day time window")) %>%
+	column_spec(column=1, width="3cm") %>%
+	column_spec(column=2:3, width="2cm") %>%
+	column_spec(column=4, width="2.5cm") %>%
+	column_spec(column=5, width="1.5cm") %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
 ```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}

### 2020-06-16 CC_update.R

@@ -50,8 +50,6 @@ SelectMatchedControls <- function(cases, popextract, first.stratum.number) {
   ret.pop.lt.10$is.case <- FALSE
   ret.table <- rbind(ret.table, ret.pop.lt.10)
   
-  
-  
   # For >=10*n[k] in each mkey in popextract then randomly sample 10 * n[k] without replacement
   pop.m.gt.10 <- pop.m[(num.cases > 0) & (num.pop.mkey >= 10 * num.cases)]
   

### 2020-06-16 Covid_CC_linkage_Part1_server.R

@@ -151,7 +151,6 @@ length(unique(NRS.wk$CHI))
 nrow(NRS.wk)
 # will be same unless people have >1 death record
 
-
 ## previous coded deaths
 
 # odbcPreviewObject(SMRAconnection, table="ANALYSIS.GRO_DEATHS_C", rowLimit=0)

### 2020-06-16 drugs.R

@@ -1,3 +1,43 @@
+
+## move to main script
+cc.all$dispensing.days <- as.integer(cc.all$SPECDATE - as.Date("2019-06-01"))
+
+getdaysbeforetest <- function(scrips.in, specdates) {
+    ## merge SPECDATE into scrips table and add columns daysbefore, dispensingdays, intervalTWday
+    ## checked that minimum value is 16 days -- the 15-day cutoff has been applied to the scrips table
+    scrips <- merge(scrips.in, specdates, by="ANON_ID", all.x=TRUE) 
+    scrips$daysbefore <- as.integer(scrips$SPECDATE - scrips$dispensed_date)
+    scrips$dispensing.days <- as.integer(scrips$SPECDATE - 15 - 
+                                         as.Date("2019-06-01"))
+    scrips$intervalTWday <- ceiling((scrips$daysbefore - 15) / TW)
+    scrips$intervalTWday[scrips$intervalTWday > 3] <- 3
+    return(scrips)
+}
+
+getdoseTWday.wide <- function(scrips.in, varname.prefix="dose.") {
+    doseTWday <-  reshape2::dcast(data=scrips.in,
+                                  formula=ANON_ID + intervalTWday ~ approved_name,
+                                  value.var="dose",
+                                  fun.aggregate=sum)
+    ## strictly, each interval should be divided by the number of days in that interval for that individual
+    ## for now, use TW days
+    for(j in 1:5) { # loop over approved names to divide by DDD x TW 
+        doseTWday[, j + 2] <- doseTWday[, j + 2] / TW
+    }
+    doseTWday$doseTWday.all <- rowSums(doseTWday[, -(1:2)])
+    ## drop columns for individual drugs, and cast again to get one column for each interval
+    doseTWday <- doseTWday[, c("ANON_ID", "intervalTWday", "doseTWday.all")]
+    doseTWday <- doseTWday[!is.na(doseTWday$intervalTWday), ]
+    doseTWday.wide <-  reshape2::dcast(data=doseTWday,
+                                       formula=ANON_ID ~ intervalTWday,
+                                       value.var="doseTWday.all",
+                                       fun.aggregate=sum)
+    colnames(doseTWday.wide)[-1] <-
+        paste0(varname.prefix,
+               c("interval1", "interval2", "interval3"))
+    return(doseTWday.wide)
+}
+
 ## reading scrips
 
[…]

### 2020-06-19 casecontrol.Rmd
fixed miscoding of broad definition of fatal cases

@@
 options(knitr.kable.NA = '')
-
+if(FALSE) {
 knitr::kable(table.demog.append[, 1:5],
@@
     kable_styling(latex_options="hold_position") 
+}
 

### 2020-06-19 casecontrol_script.Rmd
fixed miscoding of broad definition of fatal cases

---
title: 'REACT-SCOT analysis script report'
author: Paul McKeigue

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
  pdf_document

always_allow_html: true
urlcolor: blue
---

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

# rmarkdown::render("casecontrol_script.Rmd", output_file="casecontrol_script.pdf")

options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

source("helperfunctions.R")

```

\newpage

\newpage

# Tables

```{r death28, echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')

knitr::kable(with(cc.all[cc.all$CASE==1, ],
                  paste.colpercent(table(deathwithin28, unitcategory))), 
             escape=FALSE, 
             booktabs=TRUE,
             row.names=FALSE,
             caption="Death within 28 days by hospitalization group") %>%
    kable_styling(latex_options="hold_position")

```

```{r casegroup, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table(cc.all$CASE, cc.all$group, exclude=NULL),
             escape=FALSE, 
             booktabs=TRUE,
             row.names=FALSE,
             caption="Case groups") %>%
    kable_styling(latex_options="hold_position")

```

```{r NRS.cases, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(with(subset(cc.all, CASE==1 & nrs_covid_case==0), table(criticalcare, deathwithin28, exclude=NULL)),
             caption="CASE x NRS COVID case")

knitr::kable(with(subset(cc.all, CASE==1), table(criticalcare, nrs_covid_case, exclude=NULL)),
             caption="Critical care x NRS COVID case")

knitr::kable(with(subset(cc.all, CASE==1), table(nrs_covid_case, deathwithin28, exclude=NULL)),
             caption="NRS covid case x death within 28 days")

```

### 2020-06-19 CC_update.R

@@ -9,7 +9,7 @@
 #****************************************
 
 
-library(dplyr)
+
 library(odbc)
 library(lubridate)
 library(tidylog)
@@ -19,7 +19,9 @@ library(stringdist)
 library(glue)
 library(phsmethods)
 library(stringr)
-# install.packages("remotes")
+library(Hmisc)
+library(dplyr)
+ #install.packages("Hmisc")
 # remotes::install_github("Health-SocialCare-Scotland/phsmethods")
 
 
@@ -49,6 +51,8 @@ SelectMatchedControls <- function(cases, popextract, first.stratum.number) {
   ret.pop.lt.10 <- pop.m[(num.cases > 0) & (num.pop.mkey < 10 * num.cases), .(upi, mkey)]
   ret.pop.lt.10$is.case <- FALSE
   ret.table <- rbind(ret.table, ret.pop.lt.10)
+  
+  
   
   # For >=10*n[k] in each mkey in popextract then randomly sample 10 * n[k] without replacement
   pop.m.gt.10 <- pop.m[(num.cases > 0) & (num.pop.mkey >= 10 * num.cases)]
@@ -109,13 +113,15 @@ ECOSS_deduped2 <-ECOSS_deduped1 %>%
   mutate(DateOB=as.character(DateOB)) %>%
   mutate(DateOB=str_replace_all(DateOB, "-", "")) %>% 
     mutate(PATID=as.character(PATID)) %>% 
-  select(CHI,PATID,Forename,Surname,Sex,DateOB,PostCode)
+  arrange(PATID) %>% 
+  select(PATID,CHI,Forename,Surname,Sex,DateOB,PostCode)
 
 
-fwrite(ECOSS_deduped2,"/chi/(1) Project Folders/Case Control/CCseed080620_2.csv")
+fwrite(ECOSS_deduped2,"/chi/(1) Project Folders/Case Control/CCseed080620_3.csv")
 
 #file from indexer after seeding
-new_indexer_result<-fread("/chi/(1) Project Folders/Case Control/File_3118_UPI.csv")
+new_indexer_result<-fread("/chi/(1) Project Folders/Case Control/seededjunefile/File_3124_UPI.csv")
+
 #new_indexer_result<-fread("/chi/(1) Project Folders/Case Control/File_3093_UPI.csv")
[…]

### 2020-06-19 casecontrol.R

@@ -44,31 +44,35 @@ library(icd.data)
 library(gam)
 library(data.table)
 
-registerDoParallel(cores=2)
+registerDoParallel(cores=4)
 
 source("helperfunctions.R")
 
 old <- TRUE
-old <- FALSE # uncomment to use 15 May linkage
+#old <- FALSE # uncomment to use 8 June linkage
 
-nrs <- TRUE
-nrs <- FALSE # uncomment to exclude NRS-only deaths
+include.nrs.only <- TRUE
+include.nrs.only <- FALSE # uncomment to exclude NRS-only deaths
 
 stepwise <- TRUE   
 stepwise <- FALSE ## uncomment to save time if old version still valid
 
-cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"))
-diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
-procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds")
-scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
-scrips <- readRDS(scrips.filename)[, c("ANON_ID", "bnf_paragraph_code",
-                                       "bnf_paragraph_description")] 
-controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-15.rds")
-## other files for diagnoses, procedures and scrips within the time limit
-## data/CC_SMR01_ICD10_25_ANON_20200515.rds
-## data/CC_SMR01_OPCS4_MAIN.25_ANON_20200501.rds
-## data/CC_PIS_15_ANON_20200515.rds
+if(old) {
+    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"))
+    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
+    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_20200515.rds")
+    scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
+    controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-15.rds")
+} else {
+    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_2020-06-08.rds"))
[…]

### 2020-06-19 incidencemortality.R

@@ -1,3 +1,8 @@
+logit <- function(x) log(x) - log(1 - x) 
+
+## https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/mid-year-population-estimates/mid-2019
+
+## https://www.nrscotland.gov.uk/files//statistics/population-estimates/mid-19/mid-year-pop-est-2019-figures.xlsx
 
 scotpop <- read_excel("./Scotland_midyearpop_est2019.xlsx")
 
@@ -52,21 +57,21 @@ gam.model.MaleCases<- gam::gam(formula=y.cases[male, ] ~ s(Age), family=binomial
 gam.model.FemaleCases <- gam::gam(formula=y.cases[female, ] ~ s(Age), family=binomial("logit"),
                                   data=discrim[female, ])
 
-gam.male <- data.frame(Cases=car::logit(gam.model.MaleCases$fitted.values),
-                       Deaths=car::logit(gam.model.MaleDeaths$fitted.values),
+gam.male <- data.frame(Cases=logit(gam.model.MaleCases$fitted.values),
+                       Deaths=logit(gam.model.MaleDeaths$fitted.values),
                        Age=discrim$Age[male])
 gam.male.long <- reshape2::melt(data=gam.male, id="Age")
 colnames(gam.male.long)[2] <- "Status"
 gam.male.long$Sex <- "Males"
 
-gam.female <- data.frame(Cases=car::logit(gam.model.FemaleCases$fitted.values),
-                       Deaths=car::logit(gam.model.FemaleDeaths$fitted.values),
+gam.female <- data.frame(Cases=logit(gam.model.FemaleCases$fitted.values),
+                       Deaths=logit(gam.model.FemaleDeaths$fitted.values),
                        Age=discrim$Age[female])
 gam.female.long <- reshape2::melt(data=gam.female, id="Age")
 colnames(gam.female.long)[2] <- "Status"
 gam.female.long$Sex <- "Females"
 gam <- rbind(gam.male.long, gam.female.long)
-     
+
 ###############################################################
 
 logodds.posterior <- predict(object=cases.model, newdata=discrim, type="link")

### 2020-06-19 casecontrol.R
parallelized stepwise variable selection

@@ -44,7 +44,9 @@ library(icd.data)
 library(gam)
 library(data.table)
 
-registerDoParallel(cores=4)
+## FORK cluster points to a shared memory space
+cl <- makeCluster(10, type="FORK")  
+registerDoParallel(cores=10)
 
 source("helperfunctions.R")
 
@@ -442,7 +444,7 @@ rm(cc.hosp)
 ########## restrict to severe cases and matched controls ###################### 
 
 cat("Restricting to severe cases and matched controls\n")
-cc.severe <- cc.all[cc.all$casegroup=="A", ]
+cc.severe <- subset(cc.all, subset=casegroup=="A")
 
 ############ FIXME: with enough memory, this can be done on cc.all ###
 

### 2020-06-19 helperfunctions.R

@@ -631,19 +631,42 @@ replace.names <- function(varnames) {
     return(names)
 }
 
-traintest.fold <- function(i, cv.data,
+## cannot find train.data 
+cv.predict <- function(nfold, cv.data, lower.formula, upper.formula) {
+    cv.predicted <- NULL
+    ## maybe more memory-efficient not to pass cv.data as arg
+    cv.predicted <-
+        foreach(foldnum=1:nfold,
+                .combine=rbind,
+                .inorder=FALSE) %dopar% traintest.fold(foldnum=foldnum,
+                                                       cv.data=cv.data,
+                                                       lower.formula=lower.formula,
+                                                       upper.formula=upper.formula)
+    return(cv.predicted)
+}
+
+traintest.fold <- function(foldnum, cv.data,
                            lower.formula, upper.formula) { # stepwise clogit on training fold, predict on test fold
-    test.data  <- cv.data[cv.data$test.fold == i, ]
-    train.data <- cv.data[cv.data$test.fold != i, ]
-    start.model <- clogit(formula=upper.formula, data=train.data)
-    cat("train.data dimensions", dim(train.data), "\n")
+    test <- cv.data$test.fold == foldnum
+    train.data <- subset(cv.data, subset=!test)
+    ## use do.call to refer to the dataset in the calling environment
+    ## otherwise step cannot find train.data
+    start.model <- do.call(clogit, args=list(formula=upper.formula, data=train.data))
     stepwise.model <- step(start.model,
                            scope=list(lower=lower.formula, upper=upper.formula),
                            direction="both", trace=-1, method="approximate")
-    ## normalize within each stratum
+    
+    test.data  <- subset(cv.data, subset=test)
+    x <- t(with(test.data, table(CASE, stratum)))
+    numcontrols.stratum <- x[, 1]
+    numcases.stratum <- x[, 2]
+    a <- table(numcases.stratum, exclude=NULL)
+    b <- table(numcontrols.stratum, exclude=NULL)
+    if(length(a) > 1) stop("not all strata contain single case")
     unnorm.p <- predict(object=stepwise.model, newdata=test.data,
                         na.action="na.pass", 
                         type="risk", reference="sample")
+    ## normalize within each stratum
     norm.predicted <- normalize.predictions(unnorm.p=unnorm.p,
                                             stratum=test.data$stratum,
                                             y=test.data$CASE)

### 2020-06-19 stepwise.R

@@ -1,38 +1,3 @@
-
-## this doesn't work when wrapped as a function 
-cv.predict <- function(nfold, cv.data, lower.formula, upper.formula) {
-    
-    cv.predicted <- NULL
-    for(i in 1:nfold) {
-        test <- cv.data$test.fold == i
-        test.data  <- cv.data[test, ]
-        train.data <- cv.data[!test, ]
-        cat(length(which(test)), "observations in test fold", i, "\n")
-        
-        x <- t(with(test.data, table(CASE, stratum)))
-        numcontrols.stratum <- x[, 1]
-        numcases.stratum <- x[, 2]
-        a <- table(numcases.stratum, exclude=NULL)
-        b <- table(numcontrols.stratum, exclude=NULL)
-        if(length(a) > 1) stop("not all strata contain single case")
-        
-        start.model <- clogit(formula=upper.formula, data=train.data)
-        stepwise.model <- step(start.model,
-                               scope=list(lower=lower.formula, upper=upper.formula),
-                               direction="both", trace=-1, method="approximate")
-        
-        ## normalize within each stratum
-        unnorm.p <- predict(object=stepwise.model, newdata=test.data,
-                            na.action="na.pass", 
-                            type="risk", reference="sample")
-        norm.predicted <- normalize.predictions(unnorm.p=unnorm.p,
-                                                stratum=test.data$stratum,
-                                                y=test.data$CASE)
-        cv.predicted <- rbind(cv.predicted, norm.predicted)
-    }
-    return(cv.predicted)
-}
-
 ## stepwise regressions for casecontrol.R
 
 lower.varnames <- "care.home"
@@ -53,66 +18,33 @@ full.formula <- as.formula(paste0("CASE ~ ",
                             paste(full.varnames, collapse="+"), 
                             " + strata(stratum)"))
 
-nonmissing <- nonmissing.obs(cc.severe, full.varnames)
-
 ## fit models
-demog.model <- clogit(formula=demog.formula, data=cc.severe[nonmissing, ])
-listed.model <- clogit(formula=listed.formula, data=cc.severe[nonmissing, ])
-full.model <- clogit(formula=full.formula, data=cc.severe[nonmissing, ])
+nonmissing <- nonmissing.obs(cc.severe, full.varnames)
+cc.nonmissing <- subset(cc.severe, subset=nonmissing)
 
-if(stepwise) {
-    ## stepwise for full variable set
[…]

### 2020-06-22 pharmaco.Rmd
updated to use new linkage from 8 June

@@
 The univariate rate ratios associated with at least one prescription varied from 4.4 in those aged under 60 years to 3.5 in those aged 75 years and over.
-This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
-Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in drugs in almost all chapters were associated with elevated rate ratios for COVID-19.
-We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by co-morbidity or frailty, causal effects of drugs on disease susceptibility could not be ruled out.
-
-The objective of this study was to investigate the association of severe COVID-19 with drug prescribing.
-Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.
-A further aim was to determine whether the effects underlying this association are sparse or diffuse.
-In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.
-In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.
+This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
+The objective of this study was to investigate possible explanations for this association: confounding by co-morbidity  factors, or direct effects of prescribed drugs.
 
 # Methods
+# Methods
 The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].
-In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
+In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units.
 Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
-Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive [and had not embarked -- to be checked in the CHI database by SW].
-Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).
+Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
+
+For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
+For this study based on ascertainment of test-positive , up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.
 
 ## Morbidity and drug prescribing
-For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2.
-British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.
-The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).
-BNF subparagraph codes generally represent pharmacological drug classes.
-For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Otherf".
-Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
-Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.
+For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry.
+Diagnoses of diabetes were extracted from linkage to the national diabetes register.
+British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2.
+The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.
+For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".
+Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
+Morphine milligram equivalents (MMEs) for opiates were obtained from the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthetists.
+Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).
+
+As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
+ICD-10 diagnostic and BNF drug codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).
 
@@
 This procedure was repeated with the retained variables until only a single drug class remained.
-
-To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.
-As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, an independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls.
-By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.
-The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription.
-`r nrow(coeffs.propensity)` drug classes were retained in the propensity score.
-As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.
 
@@
 These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
-
-A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.
-To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.
+A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.
 Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
@@
 
-* Stratification by propensity score.
+## Associations of severe COVID-19 with analgesic use
+Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.
+This showed a dose-response relationship.
+With those who had not been dispensed opiates as baseline, the univariate rate ratio was XX in those with average dose daily of more than 50 mg morphine equivalent, reduced to YY on adjusting for prespecified covariates.
 
-## Associations of severe COVID-19 with analgesic use
+Table \ref{tab:timewindow.opiate shows that the rate ratio associated with exposure in the most recent 120-day time window was higher than that associated with exposure in a less recent time window (from 1 June 2019 to 121 days before the specimen date)
 
 ## Associations of severe COVID-19 with proton pump inhibitor use
-
 Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed.
@@
 The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.
-
-A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section.
-The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Supplementary Table \ref{tab:propensity}.
-As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor.
-Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.
 
@@
 
-The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.
-Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.
+<!---The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.
+---> Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.
 In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.
@@
 ## Methodological strengths and weaknesses
-A strength of this study is that it is based on dispensed prescriptions, rather than issued prescriptions.
+A strength of this study is that it is based on dispensed rather than issued prescriptions.
 Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.
-Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.
-Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.
+Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 by trained coders.
 
@@
 Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.
-We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors and on the avoidance of inappropriate polypharmacy.
+We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and more generally on the avoidance of unnecessary polypharmacy.
 
@@
 The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [github](https://github.com/pmckeigue/covid-scotland_public).
 Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
@@
 
-```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}
-
-if(FALSE) {
-table.numdrugsgr <- rbind(table.numdrugsgr.agegr, table.numdrugsgr.num.icdch)
-
-knitr::kable(table.numdrugsgr,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugsgr",
-             row.names=FALSE,
-             col.names=c("Number of drug classes", colnames(table.numdrugsgr.agegr)[2:3], 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group and by number of ICD chapters represented in hospital diagnoses") %>%
-    column_spec(column=1, width="2cm") %>%
-    kableExtra::group_rows("Age 0-59", 1, 6) %>%
-    kableExtra::group_rows("Age 60-74", 7, 12) %>%
-    kableExtra::group_rows("Age 75+", 13, 18) %>%
-    kableExtra::group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
-    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
-    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
-    kable_styling(latex_options="hold_position")
-}
-
-```
-
-```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
-knitr::kable(table.numdrugsgr.num.icdch,
-             escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.num.icdch",
-             row.names=FALSE,
-             col.names=c("Number of drug classes", colnames(table.numdrugsgr.num.icdch)[2:3],  
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with polypharmacy in those without listed conditions, by number of ICD-10 chapters represented in hospital diagnoses") %>%
-    column_spec(column=1, width="2cm") %>%
-    kableExtra::group_rows("No admissions", 1, 6) %>%
-    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
-    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
-    kable_styling(latex_options="hold_position")
-```
-
 ```{r dosegr.opiate, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
-```{r bypropensity, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.protonpump.bypropensity[, 1:4],
-             escape=FALSE,
-             booktabs=TRUE,
-             label="bypropensity", 
-             col.names=c(clean.header(colnames(table.protonpump.bypropensity)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with ever-use of proton pump inhibitor, by propensity score") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
 \newpage
@@
              caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in care homes and without a listed condition, filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
-    column_spec(column=1, width="12cm") %>%
+    column_spec(column=1, width="10cm") %>%
     column_spec(column=2, width="1.5cm") %>%
@@
     landscape()
-
-```
-
-```{r propensity, echo=FALSE, warning=FALSE, message=FALSE}
-chapternums <- gsub("\\..*", "", coeffs.propensity[, 1])
-chapternums <- floor(as.integer(chapternums) / 100000)
-space.at <- which(diff(chapternums) >0)
-linesep.str <- rep('', nrow(coeffs.propensity))
-linesep.str[space.at] <- '\\addlinespace'
-
-knitr::kable(coeffs.propensity,
-             escape=FALSE,
-             label="propensity", 
-             booktabs=TRUE,
-             digits=c(0, 3), 
-             col.names=c("BNF subparagraph", "log odds ratio"),
-			 linesep=linesep.str,
-             caption="Coefficients of propensity score for prescribing of a proton pump inhibitor, constructed in an independent dataset") %>%
-    kable_styling(latex_options="hold_position")
 

### 2020-06-22 casecontrol.R

@@ -1,5 +1,11 @@
 ## analysis script for case-control study
+if(exists("cl")) {
+    parallel::stopCluster(cl)
+    showConnections(all = TRUE)
+    closeAllConnections()
+}
 rm(list=ls())
+gc()
 
 ## required system packages: libcurl4-openssl-dev, pandoc, pandoc-citeproc
 ##   libssl-dev libxml2-dev
@@ -17,7 +23,7 @@ list.of.packages = c("car",
                      "pander", 
                      "ggplot2", 
                      "doParallel", 
-                     "reshape2", 
+                     "reshape2",  # deprecated
                      "readxl", 
                      "DescTools", 
                      "icd.data", 
@@ -44,17 +50,14 @@ library(icd.data)
 library(gam)
 library(data.table)
 
-## FORK cluster points to a shared memory space
-cl <- makeCluster(10, type="FORK")  
-registerDoParallel(cores=10)
 
 source("helperfunctions.R")
 
 old <- TRUE
-#old <- FALSE # uncomment to use 8 June linkage
+old <- FALSE # uncomment to use 8 June linkage
 
 include.nrs.only <- TRUE
-include.nrs.only <- FALSE # uncomment to exclude NRS-only deaths
+#include.nrs.only <- FALSE # uncomment to exclude NRS-only deaths
 
 stepwise <- TRUE   
 stepwise <- FALSE ## uncomment to save time if old version still valid
@@ -66,15 +69,17 @@ if(old) {
     scrips.filename <- "./data/CC_PIS_x15_ANON_20200515.rds" 
     controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-15.rds")
 } else {
-    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_2020-06-08.rds"))
-    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_2020-06-08.rds")
-    procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_2020-06-08.rds")
-    scrips.filename <- "./data/CC_PIS_x15_ANON_2020-06-08.rds" 
+    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_2020-06-18.rds"), key="ANON_ID")
+    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_2020-06-18.rds")
[…]

### 2020-06-22 comorbidity.R

@@ -103,7 +103,10 @@ cc.severe$immune.any <- as.factor(as.integer(cc.severe$ANON_ID %in% ids.immune.a
 ids.icd.neoplasm <- unique(diagnoses$ANON_ID[grep("^C[0-9]|^D[0-4]", diagnoses$ICD10)])
 ids.bnf.neoplasm <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 801])
 ids.neoplasm.any <- unique(c(ids.icd.neoplasm, ids.bnf.neoplasm))
-cc.severe$neoplasm.any <- as.factor(as.integer(cc.severe$ANON_ID %in% ids.neoplasm.any))
+
+cc.severe <- mutate(cc.severe,
+                    neoplasm.any = as.factor(as.integer(ANON_ID %in% ids.neoplasm.any |
+                                             can.reg==1)))
 
 ###### disorders of esophagus, stomach and duodenum ############################
 

### 2020-06-22 drugs.R

@@ -1,14 +1,11 @@
 
-## move to main script
-cc.all$dispensing.days <- as.integer(cc.all$SPECDATE - as.Date("2019-06-01"))
-
 getdaysbeforetest <- function(scrips.in, specdates) {
-    ## merge SPECDATE into scrips table and add columns daysbefore, dispensingdays, intervalTWday
+    ## merge SPECDATE and dispensing.days into scrips table and add columns daysbefore, intervalTWday
     ## checked that minimum value is 16 days -- the 15-day cutoff has been applied to the scrips table
-    scrips <- merge(scrips.in, specdates, by="ANON_ID", all.x=TRUE) 
+    setkey(scrips.in, ANON_ID)
+    setkey(specdates, ANON_ID)
+    scrips <- specdates[scrips.in]
     scrips$daysbefore <- as.integer(scrips$SPECDATE - scrips$dispensed_date)
-    scrips$dispensing.days <- as.integer(scrips$SPECDATE - 15 - 
-                                         as.Date("2019-06-01"))
     scrips$intervalTWday <- ceiling((scrips$daysbefore - 15) / TW)
     scrips$intervalTWday[scrips$intervalTWday > 3] <- 3
     return(scrips)
@@ -118,7 +115,7 @@ scrips.opioid$mode <- car::recode(scrips.opioid$formulation_code,
 
 table(scrips.opioid$approved_name, scrips.opioid$mode)
 ## only buprenorphine and fentanyl are used as patch formulations
-print(table(scrips.opioid$approved_name, scrips.opioid$mode))
+#print(table(scrips.opioid$approved_name, scrips.opioid$mode))
 
 ## item strength missing for patch formulations
 subset(scrips.opioid, subset=mode=="Patch") %>% summary()
@@ -208,8 +205,8 @@ scrips.opiates.names$factor.patch <- NA
 scrips.opiates.names$factor.patch[scrips.opiates.names$approved_name=="BUPRENORPHINE"] <- 24 * 7
 scrips.opiates.names$factor.patch[scrips.opiates.names$approved_name=="FENTANYL"] <- 180 * 3
 
-print(subset(scrips.opiates.names,
-             select=c("approved_name", "numscrips", "factor.oral", "factor.patch")))
+#print(subset(scrips.opiates.names,
+#             select=c("approved_name", "numscrips", "factor.oral", "factor.patch")))
 
 scrips.opioid <- merge(scrips.opioid,
[…]

### 2020-06-22 ethnic_assign.R

@@ -1,9 +1,5 @@
 ######## coding ethnicity ##############################
 
-## dataset is named cc.all
-
-## this script should be edited as a function that will work with any dataset, with user-specified names for source variables and output variables
-
 ## Source variables: 
 
 ## ethnic.smr - raw SMR categories
@@ -15,10 +11,6 @@
 ## this script
 
 ## (1) recodes ETHNIC_smr to specify Chinese as separate category, then collapses to 5 categories: White, South Asian, Chinese, Black, Other
-
-## (2) derives a new variable "eth" from Onomap types
-
-## (3) collapses eth to eth5, which has five categories as above
 
 ## We cannot combine SMR coding with Onomap coding because this will introduce non-differential missclassification of ethnicity between cases and controls, especially for those categorised as Black.
 
@@ -51,23 +43,26 @@
 
 ## extra codes: 1[DEFGHJ] White, 3[ABC] South Asian, 3[E] Chinese, 3D Other Asian, 4[ABCE]  African
 
-ethnic5.smr <- character(length(ethnic.smr))
-ethnic5.smr[grep("^1[A-Z]$", ethnic.smr)] <- "White"
-ethnic5.smr[grep("^2[A-Z]$", ethnic.smr)] <- "Other"
-ethnic5.smr[grep("^3[ABCFGH]$", ethnic.smr)] <- "South Asian"
-ethnic5.smr[grep("^3[EJ]$", ethnic.smr)] <- "Chinese"
-ethnic5.smr[grep("^3[DZ]$", ethnic.smr)] <- "Other"
-ethnic5.smr[grep("^4[ABCDEY]$", ethnic.smr)] <- "Black"
-ethnic5.smr[grep("^5[ABCDY]$", ethnic.smr)] <- "Black"
-ethnic5.smr[grep("^6[AZ]$", ethnic.smr)] <- "Other"
-ethnic5.smr[grep("^9", ethnic.smr)] <- NA
-ethnic5.smr[ethnic5.smr==""] <- NA
-ethnic5.smr <- as.factor(ethnic5.smr)
-ethnic5.smr <- factor(ethnic5.smr, levels=levels(ethnic5.smr)[c(5, 4, 2, 1, 3)])
+collapseto5.ethnicsmr <- function(ethnic.smr) {
[…]

### 2020-06-22 pharmaco.R

@@ -1,6 +1,5 @@
 ## pharmacoepi paper
 
-
 ########################################################
 ## tabulate rate ratios for each num drug group by listed.any
 ## this shows that the polypharmacy association is only in those without listed conditions
@@ -220,21 +219,7 @@ table.dosegr.compound.opiates <- tabulate.freqs.regressions(varnames=compound.op
 
 
 #### opiate effect by time window #######################
-
-cc.severe$opiate.exposure.nonrecent <- as.integer(cc.severe$opiateMME.interval2 > 0 | cc.severe$opiateMME.interval3 > 0)
-cc.severe$opiate.exposure.recent <- as.integer(cc.severe$opiateMME.interval1 > 0)
-
-cc.severe$opiate.exposurecat <- integer(nrow(cc.severe))
-cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==0 & cc.severe$opiate.exposure.recent==0] <- 0
-cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==1 & cc.severe$opiate.exposure.recent==0] <- 1
-cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==0 & cc.severe$opiate.exposure.recent==1] <- 2
-cc.severe$opiate.exposurecat[cc.severe$opiate.exposure.nonrecent==1 & cc.severe$opiate.exposure.recent==1] <- 3
-
-cc.severe$opiate.exposurecat <- as.factor(car::recode(cc.severe$opiate.exposurecat,
-                                "0='No prescriptions'; 1='Non-recent only'; 2='Recent only';
-                                 3='Prescriptions in both time windows'"))
-cc.severe$opiate.exposurecat <- factor(cc.severe$opiate.exposurecat,
-                                levels=levels(cc.severe$opiate.exposurecat)[c(1, 2, 4, 3)])
+## excluding neoplasms
 
 table.timewindow.opiate <-
     tabulate.freqs.regressions(varnames="opiate.exposurecat",
@@ -388,31 +373,10 @@ rownames(x) <- "All"
 colnames(table.everuse.protonpump.numdrugs)[1:2] <- colnames(x)[1:2]
 table.everuse.protonpump.numdrugs <- rbind(table.everuse.protonpump.numdrugs, x)
 
-#### time window #######################
-
-cc.severe$ppi.exposure.nonrecent <- as.integer(cc.severe$DDD.interval2 > 0 | cc.severe$DDD.interval3 > 0)
[…]

### 2020-06-22 stepwise.R

@@ -19,14 +19,28 @@ full.formula <- as.formula(paste0("CASE ~ ",
                             " + strata(stratum)"))
 
 ## fit models
+
+## FIXME - for now, drop strata containing 2 or more cases
+## should fix calculation of test log-likelihood to handle these strata
+## also fix calculation of C-statistic conditional on marginal freqs
+severe.numcases.strata <- tapply(cc.severe$CASE, cc.severe$stratum, sum)
+strata.onecase <-
+    as.integer(names(severe.numcases.strata)[as.integer(severe.numcases.strata) == 1])
+keep <- cc.severe$stratum %in% strata.onecase
 nonmissing <- nonmissing.obs(cc.severe, full.varnames)
-cc.nonmissing <- subset(cc.severe, subset=nonmissing)
+cc.nonmissing <- subset(cc.severe, subset=nonmissing & keep)
 
 demog.model <- clogit(formula=demog.formula, data=cc.nonmissing)
 listed.model <- clogit(formula=listed.formula, data=cc.nonmissing)
 full.model <- clogit(formula=full.formula, data=cc.nonmissing)
 
-if(stepwise) { # stepwise for full variable set
+if(stepwise) {
+
+    ## FORK cluster points to a shared memory space
+    cl <- makeCluster(8, type="FORK")  
+    registerDoParallel(cl, cores=8)
+
+    ## stepwise for full variable set
     stepwise.full <- step(full.model,
                           scope=list(lower=lower.formula, upper=full.formula),
                           direction="both", method="approximate", trace=-1)
@@ -84,6 +98,11 @@ if(stepwise) { # stepwise for full variable set
 
     }
 
+    
+    parallel::stopCluster(cl)
+    showConnections(all = TRUE)
+    closeAllConnections()
+
 } else {
     if(old) {
         load("./data/stepwise_15May.RData")
@@ -91,4 +110,3 @@ if(stepwise) { # stepwise for full variable set
          load("./data/stepwise.RData")
     }
 }
-

### 2020-06-23 casecontrol.Rmd
parallelized stepwise variable selection
debuggedtoworkwith15Maydataset

--- 2020-06-12 casecontrol.Rmd
@@
 options(knitr.kable.NA = '')
-
 knitr::kable(table.demog.append[, 1:5],
@@
 ```{r agegr.withNRS, echo=FALSE, warning=FALSE, message=FALSE}
-
 table.agegr.withNRS <- readRDS(table.agegr, file="table.agegr.withNRS.rds")

### 2020-06-23 pharmaco.Rmd
debuggedtoworkwith15Maydataset

@@
 
-*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed.
-With those who had no dispensed prescriptions as reference group, the rate ratio (95% CI) for COVID-19 associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.listed.any[6, 4]` in those without listed conditions.
+*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus no dispensed prescriptions was `r table.numdrugsgr.listed.any[6, 4]`.
 In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.
 The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
-A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
-Further analysis of the association with proton pump inhibitors showed a dose-response relationship that was strongest in younger age groups: the rate ratio associated with average dispensed dose of 2 or more DDDs per day was `r table.dosegr.protonpump[5, 4]` in those under 60 years and `r table.dosegr.protonpump[15, 4]` in those aged under 60 years.
+A backwards elimination procedure identified proton pump inhibitors and opiate analgesics as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
+Further analysis of these associations showed dose-response relationships that were strongest in younger age groups and reduced only slightly by adjusting for prespecified covariates.
+Stratification by time window of exposure showed that the rate ratios were higher for those with prescriptions dispensed only in the last 120 days than for those with prescriptions dispensed only in an earlier time window.
 
@@
 This requires further investigation in other settings.
-In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.
+In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of opiates and proton pump inhibitors should be reinforced.
 
@@
 Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).
+The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items, including compound analgesics.
 
@@
 Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.
+The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and nonopioid or compound analgesics.
 
-The associations with these drug classes were examined in more depth.  For studies of these specific exposures the observations were not restricted to exclude those with listed conditions. 
- Distinguishing between causality and confounding is challenging because these drug classes  are usually co-prescribed with other drugs.  We used several approaches:
+The associations with these drug classes were examined in more depth.
+For studies of these specific exposures the observations were not restricted to exclude those with listed conditions.
+Distinguishing between causality and confounding is challenging because these drug classes are usually co-prescribed with other drugs.
+We used several approaches:
 
-* Testing for a dose-response relationship
+* Testing for a dose-response relationship and stratifying by age group
 
-* Adjusting for prespecified covariates: care home residence, SIMD quintile, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+* Adjusting for prespecified covariates: for proton pump inhibitors these covariates were specified as hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+For opiate use any history of neoplasm was included as a covariate.
+Care home residence and SIMD quintile were included as covariates for both these drug classes.
 
-* Comparison between recent and less recent time windows of exposure.
+* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing
 
-## Associations of severe COVID-19 with analgesic use
+## Associations of severe COVID-19 with opiate use
 Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.
@@
 The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].
-cy.
 
 Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.
-The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.
+The rate ratio was significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.
 
@@
 The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.
+
 The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
+These indicators of overprescribing include proton pump inhibitors and analgesics, identifed by backwards elimination as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
+Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.
+The time of exposure to a drug does not necessarily correspond to the time intervals for which the prescription was issued.
+Because proton pump inhibitors are available over the counter, dispensed prescriptions do not capture all exposure.
 
-Although a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy, this procedure is likely to retain the drug classes that are most commonly prescribed.
-We have shown a clear dose-response relationship of dispensed average daily dose of proton pump inhibitors to risk of severe COVID-19.
-Adjustment for covariates pre-specified as likely to confound this association -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly.
-The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.
+For both opiates and proton pump inhibitors, four criteria provide modest evidence favouring causal explanations over confounding.
+There are strong dose-response relationships of COVID-19 to dispensed average daily dose.
+Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly.
+The effect sizes are larger in younger individuals, in whom the prevalence of comorbid conditions and frailty is lower than in older individuals.
+Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
+We emphasize however that this   evidence is not conclusive.
+It may be possible to resolve this through studies in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.
+In Scotland proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018], but overprescribing of opiates is still relatively uncommon.
 
-<!---The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.
----> Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.
-In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.
-By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small.
-One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.
-We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.
-However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.
-
-It is thus not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.
-Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.
-If the effect is causal, the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.
-
-As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through more persistent effects on the gut microbiome.
-Uncertainties about the effect of proton pump inhibitors on the risk of community acquired pneumonia [@othman_community_2016] are not necessarily relevant to effects on enteric infection.
-Proton pump inhibitors have been reported to affect the gut microbiome [@imhann_proton_2016].
-If effects of proton pump inhibitor use on susceptiblity to COVID-19 are mediated through alteration of the gut microbiome, these effects would not necessarily be detected in a comparison of time windows.
- <!---barrier to viral infection, or through altering the microbiome allowing with Warnings been issued about increased risk of _Cl.
-difficile_ infection from use proton pump inhibitors.
-[ https://bnf.nice.org.uk/drug/omeprazole.html#importantSafetyInformations ] Given this possible association with increased risk of enteric infections --->.
-
-In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].
-Initiatives to improve prescribing quality have attempted to reduce this.
-It would be of interest to examine the association between severe COVID-19 and use of proton pump inhibitors in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].
+The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration before 2012 [@edelman_association_2019].
+As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection.
 
 We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.
-We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs.
+We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing.
 For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].
@@
 ## Methodological strengths and weaknesses
+Although we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints, we have diagnoses from hospital discharge records that have been coded to ICD-10 by trained coders.
 A strength of this study is that it is based on dispensed rather than issued prescriptions.
-Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.
-Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 by trained coders.
 
@@
 
-```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the coefficient of the polypharmacy variable with that variable retained, and the log-likelihood when that drug class is dropped"}
+```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood when that drug class is dropped from a model with care home as covariate"}
 
@@
              booktabs=TRUE,
-             row.names=TRUE, 
-             col.names=c(clean.header(colnames(table.dosegr.opiate)[1:2]), 
+             row.names=FALSE, # dose levels in column 1
+             col.names=c("Average MME/day", colnames(table.dosegr.opiate)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value",
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose (MME) of opiate grouped into categoriesb") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    kableExtra::group_rows("Average dispensed daily dose (MME) of opiates", 1, 4) %>%
+             caption="Associations of severe COVID-19 with dispensed average daily dose of opiate (MME) grouped into categories, with and without adjustment for prespecified covariates") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
+    kableExtra::group_rows("Ages 0-74 years", 1, 4) %>%
+    kableExtra::group_rows("Ages 75+ years", 5, 8) %>%
     landscape()
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption=paste0("Associations of severe disease with dispensed prescriptions for opiate analgesics by ", TW, "-day time window")) %>%
+             caption=paste0("Associations of severe disease with dispensed prescriptions for opiate-containing analgesics by ", TW, "-day time window, in those without a history of neoplasm")) %>%
 	column_spec(column=1, width="3cm") %>%
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
+             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs) grouped into categories, with and without adjustment for prespecified covariates") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
-    kableExtra::group_rows("Ages 0-59 years", 1, 5) %>%
-    kableExtra::group_rows("Ages 60-74 years", 6, 10) %>%
-    kableExtra::group_rows("Ages 75+ years", 11, 15) %>%
+    kableExtra::group_rows("Ages 0-74 years", 1, 5) %>%
+    kableExtra::group_rows("Ages 75+ years", 6, 10) %>%
     landscape()

### 2020-06-23 casecontrol.R

@@ -31,6 +31,7 @@ list.of.packages = c("car",
                      "dplyr",
                      "data.table")
 
+
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
 if(length(new.packages) > 0) {install.packages(new.packages)}
 lapply(list.of.packages, require, character.only=T)
@@ -50,6 +51,7 @@ library(icd.data)
 library(gam)
 library(data.table)
 
+Rprof(tmp <- tempfile())
 
 source("helperfunctions.R")
 
@@ -178,6 +180,10 @@ cc.all$agegr3 <-
     as.factor(car::recode(cc.all$AGE,
                           "0:59='0-60 years'; 60:74='60-74 years'; 75:hi='75+ years'"))
 
+cc.all$agegr2 <-
+    as.factor(car::recode(cc.all$AGE,
+                          "0:74='0-74 years'; 75:hi='75+ years'"))
+
 cc.all$care.home[cc.all$NURSINGHOME==1] <- 1
 cc.all$care.home <- as.factor(car::recode(cc.all$care.home, "0='Independent'; 1='Care/nursing home'"))
 cc.all[, care.home := relevel(care.home, ref="Independent")]
@@ -193,10 +199,11 @@ if(old) {
     cc.all$testpositive.case <- cc.all$CASE==1 &
         (cc.all$ECOSS_POSITIVE=="Positive" |
          cc.all$ECOSS_POSITIVE=="Negative" | 
-         cc.all$ECOSS_POSITIVE=="No result")  
+         cc.all$ECOSS_POSITIVE=="No result")
+    with(subset(cc.all, CASE==1), table(ECOSS_POSITIVE, covid_cod))
+    with(cc.all[cc.all$CASE==1, ], table(ECOSS_POSITIVE, adm28, exclude=NULL))
 }
 
-with(subset(cc.all, CASE==1), table(ECOSS_POSITIVE, covid_cod))
 
 ## all cases have nonmissing SPECDATE
 ## controls should be assigned same SPECDATE as the case they were matched to
@@ -225,7 +232,6 @@ cc.all$criticalcare <- cc.all$icu==1 | cc.all$hdu==1
 cc.all <- mutate(cc.all, adm28 = adm28 & testpositive.case)
 
 with(cc.all[cc.all$CASE==1, ], table(inhosp, adm28, exclude=NULL))
[…]

### 2020-06-23 drugs.R

@@ -82,6 +82,7 @@ scrips$paracode <- as.integer(substr(scrips$bnf_paragraph_code, 1, 6))
 scrips$chapternum[is.na(scrips$chapternum)] <- 14
 scrips$chapternum[scrips$chapternum > 14] <- 14
 
+if(!old) {
 ## separate tables for prescriptions of proton pump, opioids, nonopioids
 
 scrips.protonpump <-
@@ -450,6 +451,7 @@ opiate.exposure.cat <- factor(opiate.exposure.cat,
 ## data.table behaves strangely when type of an existing column is changed
 cc.all <- mutate(cc.all, opiate.exposurecat = opiate.exposure.cat)
 
+}
 
 ########### other drugs of interest coded as binary ####################
 

### 2020-06-23 helperfunctions.R

@@ -208,35 +208,33 @@ multivariate.clogit <- function(varnames, outcome="CASE", data, add.reflevel=FAL
         ## is length(levels) -1 for factor variables
         numrows.i <- 1
         ## set number of rows in multivariate.coeffs for factor variable as num levels -1 
-        if(with(data[nonmissing, ], is.factor(eval(str2expression(varnames[i]))))) {
-            numrows.i <-
-                with(data[nonmissing, ], length(levels(eval(str2expression(varnames[i]))))) - 1
+        if(is.factor(data[nonmissing, ][[varnames[i]]])) {
+            numrows.i <- length(levels(data[nonmissing, ][[varnames[i]]])) - 1
         }
-        ## if variable is factor, > 2 levels and add.reflevel
-        if(with(data[nonmissing, ], is.factor(eval(str2expression(varnames[i])))) &
-           with(data[nonmissing, ], length(levels(eval(str2expression(varnames[i]))))) > 2 &
+        ## if variable is factor with > 2 levels and add.reflevel,
+        ## add line for reference level
+        if(is.factor(data[nonmissing, ][[varnames[i]]]) &
+           length(levels(data[nonmissing, ][[varnames[i]]])) > 2 &
            add.reflevel) {
             ## add empty line to multivariate.table
             multivariate.table <- rbind(multivariate.table,
                                         rep(NA, ncol(multivariate.coeffs)))
             ## label this empty line as reference level of factor
             rownames(multivariate.table)[nrow(multivariate.table)] <-
-                 with(data[nonmissing, ], levels(eval(str2expression(varnames[i])))[1])
+                levels(data[nonmissing, ][[varnames[i]]])[1]
         }
         ## label rows of multivariate.coeffs that will be added
         ## this step is run irrespective of add.reflevel
         ## if variable is factor with >2 levels
-        if(with(data[nonmissing, ], is.factor(eval(str2expression(varnames[i])))) &
-           with(data[nonmissing, ], length(levels(eval(str2expression(varnames[i]))))) > 2) {
+        if(is.factor(data[nonmissing, ][[varnames[i]]]) &
+           length(levels(data[nonmissing, ][[varnames[i]]])) > 2) {
             ## label rows with levels
             rownames(multivariate.coeffs)[1:numrows.i] <-
-                with(data[nonmissing, ], levels(eval(str2expression(varnames[i])))[-1])
[…]

### 2020-06-23 pharmaco.R

@@ -36,116 +36,8 @@ table.jointcardiovasc <- data.frame(drugcat, round(means.numdrugs, 1),
 colnames(table.jointcardiovasc)[2:3] <- paste0(c("Controls (", "Cases ("),
                                                table(cc.severe[notlisted, ]$CASE), rep(")", 2))
 
-#################################################################
-## tabulate para or subpara codes in BNF chapters of interest
-table.bnfchapter1 <- tabulate.bnfparas(chnum=1, data=cc.severe, minrowsum=50)
-table.bnfchapter2 <- tabulate.bnfsubparas(chnum=2, data=cc.severe, minrowsum=50)
-table.bnfchapter4 <- tabulate.bnfsubparas(chnum=4, data=cc.severe, minrowsum=50)
-## chapter 10 has to be disaggregated to drug names as grouping of DMARDs is too broad
-table.bnfchapter10 <- tabulate.bnfchemicals(chnum=10, data=cc.severe, minrowsum=50)
-
-nocare.drugfreqs <-
-    sapply(subset(cc.severe,
-                  subset=care.home=="Independent" & cc.severe$CASE==0,
-                  select=subparacols),
-           function(x) table(x)[2] / sum(table(x))) %>% sort(decreasing=TRUE) %>% head(20)
-
-care.drugfreqs <-
-    sapply(subset(cc.severe,
-                  subset=care.home=="Care/nursing home" & cc.severe$CASE==0,
-                  select=subparacols),
-           function(x) table(x)[2] / sum(table(x))) %>% sort(decreasing=TRUE) %>% head(20)
-
-#############################
-
-coeff.anticoagulant.univariate <- summary(clogit(formula=CASE ~ anticoagulant.any + strata(stratum),
-               data=cc.severe))$coefficients[1, ]
-
-coeff.anticoagulant.adjusted <- summary(clogit(formula=CASE ~ anticoagulant.any +  heart.other.any + protonpump + strata(stratum),
-               data=cc.severe))$coefficients[1, ]
-
-############################
-
-coeff.hydroxychlor.univariate <- summary(clogit(formula=CASE ~ hydroxychloroquine + strata(stratum),
-               data=cc.severe))$coefficients
-
-coeff.hydroxychlor.adjusted <- summary(clogit(formula=CASE ~ hydroxychloroquine + protonpump + strata(stratum),
-               data=cc.severe))$coefficients[1, ]
-
[…]

### 2020-06-23 stepwise.R

@@ -3,7 +3,7 @@
 lower.varnames <- "care.home"
 demog.varnames <- c("care.home", "SIMD.quintile")
 listed.varnames <- c(demog.varnames, "dm.type", listed.conditions)
-full.varnames <- c(listed.varnames, "diag.any", "scrip.any", drugs)
+full.varnames <- c(listed.varnames, "diag.any", "scrip.any", bnf.chapternames)
 
 lower.formula <- as.formula(paste0("CASE ~ ",
                             paste(lower.varnames, collapse="+"), 
@@ -97,8 +97,7 @@ if(stepwise) {
          file="./data/stepwise.RData")
 
     }
-
-    
+  
     parallel::stopCluster(cl)
     showConnections(all = TRUE)
     closeAllConnections()

### 2020-06-27 pharmaco.Rmd
separate script for paper 1 tables

@@
 
-# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf")
+# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n")
+
+# abstract limit 250 words
 
@@
 
-_Background_-- We previously reported that dispensing of prescriptions from primary care in the past year was associated with severe COVID-19 in Scotland, even in those without past diagnoses of conditions designated as increased risk (listed conditions).
-We report a further investigation of possible explanations for this association, which include specific co-morbidity or drug effects but also more diffuse effects of frailty or polypharmacy.
+_Background_ -- Dispensing of at least one prescription in the past year is a risk factor fo severe COVID-19 in Scotland, even in those without diagnoses of conditions designated as conferring increased risk.
 
-_Methods_ -- Cases of severe COVID-19 were ascertained as all those who tested positive for SARS-CoV-2 nucleic acid in the national database and entered a critical care unit or died within 28 days of testing.
-From the national population register ten controls per case were selected, matched for sex, age and general practice.
-International Classification of Disease (ICD-10) codes were extracted from the past three years of hospital discharge records and British National Formulary (BNF) drug codes from dispensed prescriptions issued in primary care during the past nine months.
-Rate ratios were estimated by conditional logistic regression.
+_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the 4320 cases of severe COVID-19 in Scotland since the start of the epidemic.
+Cases and controls were linked to records of hospital discharges in the last three years and dispensed prescriptions issued in primary care since June 2019.
 
-*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus no dispensed prescriptions was `r table.numdrugsgr.listed.any[6, 4]`.
-In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.
-The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
-A backwards elimination procedure identified proton pump inhibitors and opiate analgesics as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
-Further analysis of these associations showed dose-response relationships that were strongest in younger age groups and reduced only slightly by adjusting for prespecified covariates.
-Stratification by time window of exposure showed that the rate ratios were higher for those with prescriptions dispensed only in the last 120 days than for those with prescriptions dispensed only in an earlier time window.
+*Results* -- In those without a diagnosed risk condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.numdrugsgr.listed.any[">12", 4]`.
+Cardiovascular drugs did not contribute to this association: the largest independent contributors were proton pump inhibitors (PPI) and analgesics.
+For both PPI and opiate-containing analgesics, there were dose response relationships to severe COVID-19 that were strongest in those aged under 75 years: rate ratios were `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for $\ge 2$ DDDs/day of PPI and `r table.dosegr.opiate["0-74 years: >50", 4]` for $\ge 50$ MME opiates, The association was stronger with recent than with non-recent exposure.
 
-*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy.
-This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.
-Of all drug classes, analgesics and proton pump inhibitors make the largest independent contribution to this association.
-As prescriptions of these drug classes are recognised indicators for overprescribing, it is difficult to distinguish between causality and confounding as explanations for this association.
-This requires further investigation in other settings.
-In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of opiates and proton pump inhibitors should be reinforced.
+*Conclusions* -- In Scotland severe COVID-19 is associated with polypharmacy and this is not easily explained by co-morbidity.
+Although the evidence for causality of proton pump inhibitors and opiates is not conclusive, these results reinforce existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.
 
@@
 # Background
-The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.
-In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
-The univariate rate ratios associated with at least one prescription varied from 4.4 in those aged under 60 years to 3.5 in those aged 75 years and over.
-This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).
-The objective of this study was to investigate possible explanations for this association: confounding by co-morbidity  factors, or direct effects of prescribed drugs.
+In the initial analysis of REACT-SCOT, a matched case control study of severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
+The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.
+This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated by public health agencies as conferring vulnerability to COVID-19 (hereafter listed conditions).
+The objective of this study was to investigate this association.
 
-# Methods
 # Methods
@@
 For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
-For this study based on ascertainment of test-positive , up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.
 
@@
 Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
-Morphine milligram equivalents (MMEs) for opiates were obtained from the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthetists.
+Morphine milligram equivalents (MMEs) for opiates were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).
 Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).
-The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items, including compound analgesics.
+The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items.
 
-As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-ICD-10 diagnostic and BNF drug codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).
+As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic obstructive airway disease, chronic kidney disease, disabling neurological disease, liver disease and immunodeficiency or immunosuppression.
+ICD-10 diagnostic and BNF drug codes used to derive these conditions are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).
+Socioeconomic status was encoded as the quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).
 
@@
 We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).
-
-Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without a listed condition, but that in those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions.
-
+Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without any listed condition.
+In those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions.
 To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.
@@
 
-## Associations of severe COVID-19 with drug classes
+## Associations with drug classes
 We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.
@@
 Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and nonopioid or compound analgesics.
+The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (subparagraph code 0407010).
+The items grouped in this subparagraph include opiate-containing compound preparations.
+`r paste.colpercent(with(subset(cc.severe, CASE==0), table(nonopioid.analgesic, opioid.analgesic)))[2, 2]` of those with a dispensed prescription of a drug in subparagraph 0407020 (opiate analgesics) also had a prescription in subparagraph 0207010.
 
-The associations with these drug classes were examined in more depth.
-For studies of these specific exposures the observations were not restricted to exclude those with listed conditions.
+To examine the associations with these drug classes in more depth, the total dispensed doses of proton pump inhibitors (as DDDs) of proton pump inhibitors and opiate (as MME, summed over subparagraphs 0407010 and 0407020) were calculated.
 Distinguishing between causality and confounding is challenging because these drug classes are usually co-prescribed with other drugs.
@@
 
-* Adjusting for prespecified covariates: for proton pump inhibitors these covariates were specified as hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
-For opiate use any history of neoplasm was included as a covariate.
-Care home residence and SIMD quintile were included as covariates for both these drug classes.
+* Adjusting for prespecified covariates and restricting the analysis to those without any  listed condition.
 
-* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing
+* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing.
 
-## Associations of severe COVID-19 with opiate use
-Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.
-This showed a dose-response relationship.
-With those who had not been dispensed opiates as baseline, the univariate rate ratio was XX in those with average dose daily of more than 50 mg morphine equivalent, reduced to YY on adjusting for prespecified covariates.
+## Associations with opiate use
+Table \ref{tab:dosegr} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.
+This showed a dose-response relationship, strongest in those aged less than 75 years.
+With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opiate.notlisted[">50", 4]` and `r table.dosegr.opiate.notlisted[">50", 6]` respectively.
+Table \ref{tab:timewindow} shows that the rate ratio associated with opiate dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)
 
-Table \ref{tab:timewindow.opiate shows that the rate ratio associated with exposure in the most recent 120-day time window was higher than that associated with exposure in a less recent time window (from 1 June 2019 to 121 days before the specimen date)
-
-## Associations of severe COVID-19 with proton pump inhibitor use
-Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed.
-This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates.
-Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.
-The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].
-
-Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.
-The rate ratio was significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.
+## Associations with proton pump inhibitor use
+Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for COVID-19 increases with the dose of proton pump inhibitor.
+The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile and any history of neoplasm.
+This dose-response relationship was absent in those aged 75 years and over.
+In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 2 DDDs/day MME /day were `r table.dosegr.protonpump.notlisted["2 or more", 4]` and `r table.dosegr.protonpump.notlisted["2 or more", 6]` respectively.
+Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.
 
@@
 We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.
-We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.
-Without more data on health status from primary care records, the confounders cannot be accurately measured.
-We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.
-The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.
+in individuals without conditions designated as conferring high risk.
+The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017].
+We attempted to investigate how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of specific drugs.
+A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.
+Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10, and that drug exposure is based on dispensed rather than issued prescriptions.
 
-The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
-These indicators of overprescribing include proton pump inhibitors and analgesics, identifed by backwards elimination as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.
+The drug classes that are strongly associated with severe COVID-19 include proton pump inhibitors and analgesics, widely recognized to be indicators of overprescribing as in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
+As these are among the most commonly prescribed drug classes, it is to be expected that they are selected by elimination as having the largest independent contributions to the association of COVID-19 with polypharmacy.
 Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.
-The time of exposure to a drug does not necessarily correspond to the time intervals for which the prescription was issued.
 Because proton pump inhibitors are available over the counter, dispensed prescriptions do not capture all exposure.
 
-For both opiates and proton pump inhibitors, four criteria provide modest evidence favouring causal explanations over confounding.
+For both opiates and proton pump inhibitors, four criteria provide moderate but not conclusive evidence favouring causal explanations over confounding.
 There are strong dose-response relationships of COVID-19 to dispensed average daily dose.
 Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly.
-The effect sizes are larger in younger individuals, in whom the prevalence of comorbid conditions and frailty is lower than in older individuals.
+The effect sizes are larger in younger individuals and when the analysis is restricted to those without any of the designated risk conditions for COVID-19.
 Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
-We emphasize however that this   evidence is not conclusive.
-It may be possible to resolve this through studies in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.
-In Scotland proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018], but overprescribing of opiates is still relatively uncommon.
+It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.
 
-The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration before 2012 [@edelman_association_2019].
-As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection.
+The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration from 2000-2012 [@edelman_association_2019].
+As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase susceptibility to severe infection.
 
-We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest.
 We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing.
@@
 The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.
-
-## Methodological strengths and weaknesses
-Although we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints, we have diagnoses from hospital discharge records that have been coded to ICD-10 by trained coders.
-A strength of this study is that it is based on dispensed rather than issued prescriptions.
 
@@
 ## Information governance
-This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland.
+This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.
 Datasets were de-identified before analysis.
-The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
-All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [github](https://github.com/pmckeigue/covid-scotland_public).
+The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
 Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
@@
 
-```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood when that drug class is dropped from a model with care home as covariate"}
+```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood of a conditional logistic regression of severe COVID-19 on care home and polypharmacy as covariates when that drug class is dropped, in  those without listed conditions."}
 
@@
 
-```{r numdrugs.cardio, echo=FALSE, warning=FALSE, message=FALSE}
+```{r dosegr, echo=FALSE, warning=FALSE, message=FALSE}
 
-colnames(table.jointcardiovasc)[c(1, 4:5)] <- c("Drug category", "Odds ratio (95\\% CI)",
-                                                "$p$-value")
+colnames(table.dosegr.protonpump) <- colnames(table.dosegr.opiate)
+table.dosegr <- rbind(table.dosegr.opiate, table.dosegr.protonpump)
 
-knitr::kable(table.jointcardiovasc,
+knitr::kable(table.dosegr,
              escape=FALSE,
-             booktabs=TRUE,
-             label="numdrugs.cardio",
-             row.names=FALSE,
-             caption="Multivariate regression of severe COVID-19 on number of cardiovascular and number of noncardiovascular drug classes in dispensed prescriptions, among those without listed conditions") %>%
-    column_spec(column=1, width="2.5cm") %>%
-    add_header_above(c(" "=1, "Mean number of drug classes"=2)) %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r dosegr.opiate, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.dosegr.opiate,
-             escape=FALSE,
-             label="dosegr.opiate", 
+             label="dosegr", 
              booktabs=TRUE,
              row.names=FALSE, # dose levels in column 1
-             col.names=c("Average MME/day", colnames(table.dosegr.opiate)[2:3], 
+             col.names=c("Average dose", colnames(table.dosegr.opiate)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value",
@@
     add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
+	column_spec(column=1, width="3cm") %>%
     kable_styling(latex_options="hold_position") %>%
+    kableExtra::group_rows(group_label="Opiate-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
+    kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 16) %>%
     kableExtra::group_rows("Ages 0-74 years", 1, 4) %>%
     kableExtra::group_rows("Ages 75+ years", 5, 8) %>%
+    kableExtra::group_rows("Ages 0-74 years", 9, 13) %>%
+    kableExtra::group_rows("Ages 75+ years", 14, 18) %>%
     landscape()
@@
 
-```{r timewin.opiate, echo=FALSE, warning=FALSE, message=FALSE}
+```{r timewin, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.timewindow.opiate,
+colnames(table.timewindow.protonpump) <- colnames(table.timewindow.opiate)
+table.timewindow <- rbind(table.timewindow.opiate, table.timewindow.protonpump)
+table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opiate), 2),
+                               table.timewindow)
+
+knitr::kable(table.timewindow,
              escape=FALSE,
-             label="timewindow.opiate", 
+             label="timewindow", 
              booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.timewindow.opiate)[1:2]), 
+             row.names=FALSE, 
+             col.names=c("Time window", clean.header(colnames(table.timewindow.opiate)[1:2]), 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption=paste0("Associations of severe disease with dispensed prescriptions for opiate-containing analgesics by ", TW, "-day time window, in those without a history of neoplasm")) %>%
+             caption=paste0("Associations of severe disease with dispensed prescriptions by ", TW, "-day time window")) %>%
 	column_spec(column=1, width="3cm") %>%
@@
 	column_spec(column=5, width="1.5cm") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r dosegr.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.dosegr.protonpump,
-             escape=FALSE,
-             label="dosegr.protonpump", 
-             booktabs=TRUE,
-             row.names=FALSE, # dose levels in column 1
-             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs) grouped into categories, with and without adjustment for prespecified covariates") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
-    kableExtra::group_rows("Ages 0-74 years", 1, 5) %>%
-    kableExtra::group_rows("Ages 75+ years", 6, 10) %>%
-    landscape()
-
-```
-
-```{r timewin.protonpump, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.timewindow.protonpump,
-             escape=FALSE,
-             label="timewindow.protonpump", 
-             booktabs=TRUE,
-             col.names=c(clean.header(colnames(table.timewindow.protonpump)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption=paste0("Associations of severe disease with dispensed prescriptions for proton pump inhibitor by ", TW, "-day time window")) %>%
-	column_spec(column=1, width="3cm") %>%
-	column_spec(column=2:3, width="2cm") %>%
-	column_spec(column=4, width="2.5cm") %>%
-	column_spec(column=5, width="1.5cm") %>%
-    kable_styling(latex_options="hold_position")
-
+    kableExtra::group_rows(group_label="Opiate-containing analgesics", 1, 4) %>%
+    kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
 ```
@@
             "Rate ratio", "$p$-value"),
-             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in care homes and without a listed condition, filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
+             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code in those without a listed condition, filtered to show only drug classes with at least 50 exposed individuals and  $p < 0.001$: ")) %>%
     column_spec(column=1, width="10cm") %>%

### 2020-06-27 casecontrol.R

@@ -31,7 +31,6 @@ list.of.packages = c("car",
                      "dplyr",
                      "data.table")
 
-
 new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
 if(length(new.packages) > 0) {install.packages(new.packages)}
 lapply(list.of.packages, require, character.only=T)
@@ -64,15 +63,53 @@ include.nrs.only <- TRUE
 stepwise <- TRUE   
 stepwise <- FALSE ## uncomment to save time if old version still valid
 
+recode.dmtype <- function(x) {
+    r <- as.factor(car::recode(x, 
+                        "c(0, 10, 17, 97)='Not diabetic';
+                         c(1, 101, 102)='Type 1 diabetes';
+                         c(2, 202, 203)='Type 2 diabetes'; 
+                         3:9='Other/unknown type';
+                         11:16='Other/unknown type';
+                         18:96='Other/unknown type';
+                         98:100='Other/unknown type'"))
+    r <- factor(r, levels=levels(r)[c(1, 3, 4, 2)])
+    return(r)
+}
+
+cases.old <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"), key="ANON_ID")
+cases.new <- as.data.table(readRDS("./data/CC_linked_ANON_2020-06-18.rds"), key="ANON_ID")
+cases.old <- subset(cases.old, CASE==1)
+cases.new <- subset(cases.new, is.case)
+
+setnames(cases.old, "age", "AGE")
+
+setnames(cases.new, "SPECIMENDATE", "SPECDATE")
+setnames(cases.new, "AgeYear", "AGE")
+
+cases.old$dm.type <- recode.dmtype(cases.old$dm.type)
+cases.new$dm.type <- recode.dmtype(cases.new$dm.type)
+
+with(subset(cases.old, AGE<50), table(dead28 | covid_cod, icu==1 | hdu==1))
+with(subset(cases.new, AGE<50), table(dead28 | covid_cod, icu==1 | hdu==1))
+
+##################################################################
+     
 if(old) {
-    cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"))
-    diagnoses <- readRDS("./data/CC_SMR01_ICD10_x25_ANON_20200515.rds")
[…]

### 2020-07-01 pharmaco.Rmd

@@
 
-_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the 4320 cases of severe COVID-19 in Scotland since the start of the epidemic.
-Cases and controls were linked to records of hospital discharges in the last three years and dispensed prescriptions issued in primary care since June 2019.
+_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic.
+Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care since June 2019.
 
@@
 For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
-For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
 
@@
 To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure.
-Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.
+Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models with care home status and number of drug classes dispensed as covariates, in which each drug class was dropped from the total number of drug classes.
 The model with the highest log-likelihood was used to select which drug class should be dropped.
@@
 As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.
-In a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
+In a model with two covariates -- number of cardiovascular drug classes and number of other drug classes -- the rate ratios per drug class for severe COVID-19 were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
 
@@
 We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.
-Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.
+Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 50 cases and controls and were associated with severe disease at $p < 0.001$.
 These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
-A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.
+A backward stepwise variable elimination procedure as described above was carried out to identify which of these `r length(subparacols.forstepwisedrop)` drug classes contributed most to the association of COVID-19 with polypharmacy.
 Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
@@
 Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (subparagraph code 0407010).
-The items grouped in this subparagraph include opiate-containing compound preparations.
-`r paste.colpercent(with(subset(cc.severe, CASE==0), table(nonopioid.analgesic, opioid.analgesic)))[2, 2]` of those with a dispensed prescription of a drug in subparagraph 0407020 (opiate analgesics) also had a prescription in subparagraph 0207010.
+The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (BNF subparagraph code 0407010).
 
-To examine the associations with these drug classes in more depth, the total dispensed doses of proton pump inhibitors (as DDDs) of proton pump inhibitors and opiate (as MME, summed over subparagraphs 0407010 and 0407020) were calculated.
-Distinguishing between causality and confounding is challenging because these drug classes are usually co-prescribed with other drugs.
+The items grouped under subparagraph 0407010 include opiate-containing compound preparations and also aspirin-containing preparations with which proton pump inhibitors are commonly co-prescribed.
+To examine the associations with these drug classes in more depth, the total dispensed daily dose of opiate as MME, summed over subparagraphs 0407010 and 0407020 (opiate analgesics) was calculated.
+The total dispensed daily dose of proton pump inhibitors was calculated as DDDs.
+Distinguishing between causality and confounding is challenging because opiate-containing drugs and proton pump inhibitores are usually co-prescribed with other drugs.
 We used several approaches:
@@
 ## Associations with opiate use
-Table \ref{tab:dosegr} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.
+Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily opiate dose (MMEs) over the observation period.
 This showed a dose-response relationship, strongest in those aged less than 75 years.
-With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm.
 In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opiate.notlisted[">50", 4]` and `r table.dosegr.opiate.notlisted[">50", 6]` respectively.
@@
 ## Associations with proton pump inhibitor use
-Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for COVID-19 increases with the dose of proton pump inhibitor.
-The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile and any history of neoplasm.
+Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for severe COVID-19 increases with the dose of proton pump inhibitor.
+The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
 This dose-response relationship was absent in those aged 75 years and over.
@@
 
-Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, osmotic laxatives, and aminosalicylates were independently associated with disease in a multivariate analysis.
-Proton pump inhibitors were far more commonly prescribed than other drug classes in this chapter.
+Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, antimotility and laxatives were strongly and independently associated with disease in a multivariate analysis.
+Proton pump inhibitors were prescribed far more commonly than any other drug class in this chapter.
 
-Table \ref{tab:bnfchapter2} shows that prescriptions of most classes of drugs for the cardiovascular system were associated with elevated rate ratios for severe COVID-19.
-In a multivariate analysis the strongest associations were with loop diuretics, anticoagulants and antiplatelet drugs.
-Of drug classes commonly used to treat hypertension, potassium-sparing diuretics, beta-blockers and ACE inhibitors were associated with increased risk of COVID-19 in univariate analyses, but these associations were weak or absent in a multivariate analysis.
-The univariate rate ratio associated with oral anticoagulants was `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]`; Adjustment for other (non-ischaemic) heart disease and ever-use of a proton pump inhibitor reduced this to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])`.
+Table \ref{tab:bnfchapter2} shows associations with drugs for the cardiovascular system.
+Prescriptions of loop diuretics and anticoagulants were associated with elevated rate ratios for severe COVID-19 in univariate and multivariate analyses.
+Over all age groups combined, the univariate rate ratio associated with oral anticoagulants was reduced from `r table.bnfchapter2[grep("Oral anticoagulants", rownames(table.bnfchapter2)), 3]` to `r or.ci(coeff.anticoagulant.adjusted[1], coeff.anticoagulant.adjusted[3])` by adjustment for ischaemic heart disese, other heart disease and ever-use of a proton pump inhibitor.
+Of drug classes commonly used to treat hypertension, ACE inhibitors and angiotensin II receptor antagonists were associated with reduced risk of COVID-19.
 
-Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with drugs widely used in care homes: antipsychotics, analgesics and drugs for dementia.
+Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with antipsychotics, analgesics and drugs for dementia.
 Table \ref{tab:bnfchapter10} shows associations with drugs for the musculoskeletal system disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code.
@@
 # Discussion 
-We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.
-in individuals without conditions designated as conferring high risk.
+We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation, in individuals without conditions designated as conferring high risk.
 The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017].
@@
 This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.
-The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.
-The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.
-Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.
-We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and more generally on the avoidance of unnecessary polypharmacy.
+The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing, althoug proton pump inhibitors and analgesics have the greatest explanatory power for the association of COVID-19 with polypharmacy.
+Although this study did not identify unequivocal evidence of causality for opiates and proton pump inhibitors, effects of these drug classes on susceptibility to COVID-19 are biologically plausible.
+We recommend that public health agencies should reinforce existing guidelines on the overprescribing of these drug classes, and more generally on the avoidance of unnecessary polypharmacy.
 
@@
 HC and PM own stock in Roche and Bayer.
+SH received honoraria from Gilead.
 
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose of opiate (MME) grouped into categories, with and without adjustment for prespecified covariates") %>%
+             caption="Associations of severe COVID-19 with dispensed average daily dose of opiates and proton pump inhibitors, with and without adjustment for prespecified covariates") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
@@
     kableExtra::group_rows(group_label="Opiate-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
-    kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 16) %>%
-    kableExtra::group_rows("Ages 0-74 years", 1, 4) %>%
-    kableExtra::group_rows("Ages 75+ years", 5, 8) %>%
-    kableExtra::group_rows("Ages 0-74 years", 9, 13) %>%
-    kableExtra::group_rows("Ages 75+ years", 14, 18) %>%
+    kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 18) %>%
+    kableExtra::group_rows("  Ages 0-74 years", 1, 4) %>%
+    kableExtra::group_rows("  Ages 75+ years", 5, 8) %>%
+    kableExtra::group_rows("  Ages 0-74 years", 9, 13) %>%
+    kableExtra::group_rows("  Ages 75+ years", 14, 18) %>%
     landscape()
@@
 
-colnames(table.timewindow.protonpump) <- colnames(table.timewindow.opiate)
+colnames(table.timewindow.opiate) <- colnames(table.timewindow.protonpump)
 table.timewindow <- rbind(table.timewindow.opiate, table.timewindow.protonpump)
@@
              row.names=FALSE, 
-             col.names=c("Time window", clean.header(colnames(table.timewindow.opiate)[1:2]), 
+             col.names=c("Time window", clean.header(colnames(table.timewindow.protonpump)[1:2]), 
                          "Rate ratio (95\\% CI)", "$p$-value"),
@@
     kable_styling(latex_options="hold_position") %>%
-    kableExtra::group_rows(group_label="Opiate-containing analgesics", 1, 4) %>%
+    kableExtra::group_rows(group_label="Opiate-containing analgesics (restricted to those without history of neoplasm)", 1, 4) %>%
     kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 1 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 1 by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 4  by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 4  by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="5cm") %>%

### 2020-07-01 Covid_ethnicity_Scotland_reportPartB.Rmd
Part B of ethnicity report - disaggregated ethnicity groupings.

--- 2020-06-10 Covid_ethnicity_Scotland_PartB.Rmd
@@
 In this paper we attempt carry out a more detailed disaggregation of ethnicity presented in the "Briefing paper on Ethnicity and COVID-19 status in Scotland".
+
+Where there is evidence of an elevated risk of "severe test positive disease" or "severe test positive or death" among those of South Asian ethnicity in Part A of this report, the observed rate ratios are slightly higher among those of Indian ethnicity compared to those of Pakistani/Bangladeshi using the disaggregated SMR approach.
+For those with Black ethnicity in Part A of the report, the disaggregated data show that those classed as Black Caribbean have an increased risk of "severe test positive disease" and "severe test positive or death", as do Black Africans, but numbers are very small in these groups and therefore estimates are more unstable.
+
+When applying the disaggregated ONOMAP approach, we see that those with Muslim South Asian ethnicity are at an increased risk of "severe test positive disease" or "severe test positive or death" and not those classed as Other South Asian ethnicity.
 
@@
              align=rep("c", 6), 
-             caption="Associations of hospitalized disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+             caption="Associations of hospitalized or severe disease among test positives with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
   kable_styling(latex_options="hold_position") %>% 
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+             caption="Associations of severe disease among test positives with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
   kable_styling(latex_options="hold_position") %>% 
@@
 
-rownames(table.hospitalized.demogB) <- gsub("_", " ", rownames(table.hospitalized.demogB))
-knitr::kable(table.hospitalized.demogB,
+rownames(table.hospitalized.demogB7) <- gsub("_", " ", rownames(table.hospitalized.demogB7))
+knitr::kable(table.hospitalized.demogB7,
              escape=FALSE, 
@@
              label="onomap.hosp",
-             col.names=c(clean.header(colnames(table.hospitalized.demogB)[1:2]),
+             col.names=c(clean.header(colnames(table.hospitalized.demogB7)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6), 
-             caption="Associations of hospitalized disease with demographic variables including ethnicity based on name classification") %>%
+             caption="Associations of hospitalized or severe disease among test positives with demographic variables including ethnicity based on name classification") %>%
   add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
@@
 
-rownames(table.severe.demogB) <- gsub("_", " ", rownames(table.severe.demogB))
-knitr::kable(table.severe.demogB,
+#rownames(table.severe.demogB) <- gsub("_", " ", rownames(table.severe.demogB))
+#knitr::kable(table.severe.demogB,
+#             escape=FALSE, 
+#             booktabs=TRUE,
+#             label="onomapB",
+#             col.names=c(clean.header(colnames(table.severe.demogB)[1:2]),
+#                         "Rate ratio (95\\% CI)", "p-value",
+#                         "Rate ratio (95\\% CI)", "p-value"),
+#             align=rep("c", 6),
+#             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
+#  kable_styling(latex_options="hold_position") %>% 
+#  column_spec(column=1, width="2cm") %>%
+#  add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+#  column_spec(column=2, width="1.85cm")  %>%
+#  column_spec(column=3, width="1.85cm")  %>%
+#  column_spec(column=4, width="2.5cm") %>%
+#  column_spec(column=5, width="1.5cm") %>%
+#  column_spec(column=6, width="2.5cm") %>%
+#  column_spec(column=7, width="1.5cm")
+
+```
+
+```{r onomapB7, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.severe.demogB7) <- gsub("_", " ", rownames(table.severe.demogB7))
+knitr::kable(table.severe.demogB7,
              escape=FALSE, 
              booktabs=TRUE,
-             label="onomap",
-             col.names=c(clean.header(colnames(table.severe.demogB)[1:2]),
+             label="onomapB7",
+             col.names=c(clean.header(colnames(table.severe.demogB7)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
              align=rep("c", 6),
-             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
+             caption="Associations of severe disease among test positives with demographic variables including ethnicity based on name classification") %>%
   kable_styling(latex_options="hold_position") %>% 
@@
 
-rownames(table.severe.demogNRSB) <- gsub("_", " ", rownames(table.severe.demogNRSB))
-knitr::kable(table.severe.demogNRSB,
+#rownames(table.severe.demogNRSB) <- gsub("_", " ", rownames(table.severe.demogNRSB))
+#knitr::kable(table.severe.demogNRSB,
+#             escape=FALSE, 
+#             booktabs=TRUE,
+#             label="onomapNRSB",
+#             col.names=c(clean.header(colnames(table.severe.demogNRSB)[1:2]),
+#                         "Rate ratio (95\\% CI)", "p-value",
+#                         "Rate ratio (95\\% CI)", "p-value"),
+#             align=rep("c", 6),
+#             caption="Associations of severe disease OR any COVID-19 death with demographic variables including ethnicity based on name #classification") %>%
+#    kable_styling(latex_options="hold_position") %>% 
+#    column_spec(column=1, width="2cm") %>%
+#    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+#    column_spec(column=2, width="1.85cm")  %>%
+#    column_spec(column=3, width="1.85cm")  %>%
+#    column_spec(column=4, width="2.5cm") %>%
+#    column_spec(column=5, width="1.5cm") %>%
+#    column_spec(column=6, width="2.5cm") %>%
+#    column_spec(column=7, width="1.5cm")
+
+```
+
+```{r onomapNRSB7, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.severe.demogNRSB7) <- gsub("_", " ", rownames(table.severe.demogNRSB7))
+knitr::kable(table.severe.demogNRSB7,
              escape=FALSE, 
              booktabs=TRUE,
-             label="onomapNRSB",
-             col.names=c(clean.header(colnames(table.severe.demogNRSB)[1:2]),
+             label="onomapNRSB7",
+             col.names=c(clean.header(colnames(table.severe.demogNRSB7)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",

### 2020-07-01 Covid_ethnicity_Scotland_reportPartB.Rmd
Updated to remove disaggregation of Black ethnicity in onomap tables.

@@
 
-Where there is evidence of an elevated risk of "severe test positive disease" or "severe test positive or death" among those of South Asian ethnicity in Part A of this report, the observed rate ratios are slightly higher among those of Indian ethnicity compared to those of Pakistani/Bangladeshi using the disaggregated SMR approach.
-For those with Black ethnicity in Part A of the report, the disaggregated data show that those classed as Black Caribbean have an increased risk of "severe test positive disease" and "severe test positive or death", as do Black Africans, but numbers are very small in these groups and therefore estimates are more unstable.
-
-When applying the disaggregated ONOMAP approach, we see that those with Muslim South Asian ethnicity are at an increased risk of "severe test positive disease" or "severe test positive or death" and not those classed as Other South Asian ethnicity.
+The following tables further disaggregate the ethnicity distributions - the coding to these disaggregated groups is as follows;
 
@@
 White - White and White Polish.
-Black - Black African and Black Caribbean.
+Black - Other
 Bangladeshi - Muslim South Asian.
@@
 
-rownames(table.hospitalized.demogB7) <- gsub("_", " ", rownames(table.hospitalized.demogB7))
-knitr::kable(table.hospitalized.demogB7,
+rownames(table.hospitalized.demogB6) <- gsub("_", " ", rownames(table.hospitalized.demogB6))
+knitr::kable(table.hospitalized.demogB6,
              escape=FALSE, 
@@
              label="onomap.hosp",
-             col.names=c(clean.header(colnames(table.hospitalized.demogB7)[1:2]),
+             col.names=c(clean.header(colnames(table.hospitalized.demogB6)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
 
-```{r onomapB7, echo=FALSE, warning=FALSE, message=FALSE}
+```{r onomapB6, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.severe.demogB7) <- gsub("_", " ", rownames(table.severe.demogB7))
-knitr::kable(table.severe.demogB7,
+rownames(table.severe.demogB6) <- gsub("_", " ", rownames(table.severe.demogB6))
+knitr::kable(table.severe.demogB6,
              escape=FALSE, 
              booktabs=TRUE,
-             label="onomapB7",
-             col.names=c(clean.header(colnames(table.severe.demogB7)[1:2]),
+             label="onomapB6",
+             col.names=c(clean.header(colnames(table.severe.demogB6)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",
@@
 
-```{r onomapNRSB7, echo=FALSE, warning=FALSE, message=FALSE}
+```{r onomapNRSB6, echo=FALSE, warning=FALSE, message=FALSE}
 
-rownames(table.severe.demogNRSB7) <- gsub("_", " ", rownames(table.severe.demogNRSB7))
-knitr::kable(table.severe.demogNRSB7,
+rownames(table.severe.demogNRSB6) <- gsub("_", " ", rownames(table.severe.demogNRSB6))
+knitr::kable(table.severe.demogNRSB6,
              escape=FALSE, 
              booktabs=TRUE,
-             label="onomapNRSB7",
-             col.names=c(clean.header(colnames(table.severe.demogNRSB7)[1:2]),
+             label="onomapNRSB6",
+             col.names=c(clean.header(colnames(table.severe.demogNRSB6)[1:2]),
                          "Rate ratio (95\\% CI)", "p-value",

### 2020-07-03 casecontrol_script.Rmd

@@
 knitr::kable(with(cc.all[cc.all$CASE==1, ],
-                  paste.colpercent(table(deathwithin28, unitcategory))), 
+                  paste.colpercent(table(deathwithin28, casegroup))), 
              escape=FALSE, 
@@
              row.names=FALSE,
-             caption="Death within 28 days by hospitalization group") %>%
+             caption="Death within 28 days by case group") %>%
     kable_styling(latex_options="hold_position")
@@
 
-knitr::kable(table(cc.all$CASE, cc.all$group, exclude=NULL),
+knitr::kable(table(cc.all$CASE, cc.all$casegroup, exclude=NULL),
              escape=FALSE, 
@@
 
-```{r NRS.cases, echo=FALSE, warning=FALSE, message=FALSE}
+```{r criticalcare, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(with(subset(cc.all, CASE==1 & nrs_covid_case==0), table(criticalcare, deathwithin28, exclude=NULL)),
-             caption="CASE x NRS COVID case")
-
-knitr::kable(with(subset(cc.all, CASE==1), table(criticalcare, nrs_covid_case, exclude=NULL)),
-             caption="Critical care x NRS COVID case")
-
-knitr::kable(with(subset(cc.all, CASE==1), table(nrs_covid_case, deathwithin28, exclude=NULL)),
-             caption="NRS covid case x death within 28 days")
+knitr::kable(with(subset(cc.all, CASE==1), table(criticalcare, deathwithin28, exclude=NULL)),
+             caption="critical care x deathwithin28")
 

### 2020-07-03 Covid_ethnicity_Scotland_report.Rmd

--- 2020-06-10 Covid_ethnicity_Scotlandhc.Rmd
@@
 There have been reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.
-For example a preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95% CI 2.92-5.12); RR 2.11 (95% CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively).
+For example, a preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95% CI 2.92-5.12); RR 2.11 (95% CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively).
 
@@
 
-Any additional persons with a registered death with a COVID-19 code ( suspected or probable ) were also identified.
+Any additional persons with a registered death with a COVID-19 code (suspected or probable) were also identified.
 
@@
 
-Data were extracted for  this analysis on 15th May 2020. They do not include the Lighthouse data as we do not have access to this through ECOSS as yet.
+Data were extracted for this analysis on 8th June 2020. They do not include the Lighthouse data as we do not have access to this through ECOSS as yet.
 
@@
 
-Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of  recording of self assigned ethnic group in recent years.
+Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of recording of self-assigned ethnic group in recent years.
 Census categories are used.
-However there remains substantial missingness.
+However, there remains substantial missingness.
 For all cases and controls the linkage team extracted i) the most recent ethnic group assignation from these two datasets with a 10 year lookback period.
@@
 
-As other sociodemographic factors may vary by ethnicity we further adjusted analyses for residential care home status and Scottish Index of Multiple Deprivation (SIMD).
+As other sociodemographic factors may vary by ethnicity we further adjusted analyses for residential care home status, Scottish Index of Multiple Deprivation (SIMD) and diabetes.
 
 # Results
-SMR based ethnicity was unavailable for approximately 25% of cases and 50% of controls.
-The ONOMAP approach  assigned ethnicity  in >99% of persons.
-Given the substantial and differential missingness of SMR based ethnicity the principle value of showing the  SMR data here is to  inform on the likely absolute  numbers of cases   of non-White ethnicity rather than for the purposes of estimating rate ratios.
+SMR based ethnicity was unavailable for approximately 25% of cases and 35% of controls. The ONOMAP approach  assigned ethnicity  in >99% of persons.
 
@@
 
-Table \ref{tab:crosstab} shows the agreement between the ONOMAP  (name based) assigned ethnicity and the self assigned SMR ethnicity.
-The name classification assigns to the same as  SMR category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[5, 5]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.
+Table \ref{tab:crosstab} shows the agreement between the ONOMAP  (name based) assigned ethnicity and the self-assigned SMR ethnicity.
+The name classification assigns to the same as  SMR category `r table.ethnic[1, 1]` of those who self-report their ethnicity as White, `r table.ethnic[4, 4]` of those who self-report as South Asian, `r table.ethnic[3, 3]` of those who self-report as Chinese, but only `r table.ethnic[2, 2]` of those who self-report as Black.
 The proportion with  non-White ethnicity  is low for all groups.
-Thus  although the sensitivity of ONOMAP for assigning SMR South Asian ethnicity is >99% the positive predictive value of ONOMAP assigned South Asian ethnicity is just 60%.
+Thus although the sensitivity of ONOMAP for assigning SMR South Asian ethnicity is >99% the positive predictive value of ONOMAP assigned South Asian ethnicity is just 65%.
 
 Clearly analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records.
-*Please note that we do not consider the rate ratios here to be reliable because of the differential missingness.* Therefore in all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.
+However, the differential missingness between cases and controls means the rate ratios here must be interpreted with caution.
+Therefore, in all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.
 
@@
 
-i)all test positive cases ii) all hospitalized or severe  test positive cases, iii) severe  test positive cases iv) severe test positive cases OR a COVID-19 death registration,    Severe test positive cases are those who  entered a critical care unit  or died within 28 days of a positive test.
+i)all test positive cases ii) all hospitalized or severe  test positive cases, iii) severe  test positive cases iv) severe test positive cases OR a COVID-19 death registration.
+Severe test positive cases are those who  entered a critical care unit  or died within 28 days of a positive test.
 Note that full follow up to 28 days has not yet completed for  the recent cases.
@@
 Those in the second column are further adjusted for SIMD and care home status.
+Tables \ref{tab:ethnicsmrDB} and \ref{tab:ethnicsmrNRSDB} include an additional adjustment for Diabetes.
 
@@
 Tables \ref{tab:severity.smr} and \ref{tab:severity} show the distribution over severity categories by ethnicity.
+Tables \ref{tab:onomapDB} and \ref{tab:onomapNRSDB} include an additional adjustment for Diabetes.
 Tables \ref{tab:carehome} and \ref{tab:board} show the distribution of care home status and health board.
@@
 
-There was no evidence of elevation in risk of  "severe  test positive disease" or "severe test positive disease or death"  in  any non-White ethnic group using  either the ONOMAP  or SMR  approach.
+There is evidence of an elevated risk of  "severe test positive disease" for those of South Asian ethnicity using both the SMR and ONOMAP approach (univariate rate ratios of 1.58 and 1.45 respectively) with borderline significance.
+The rate ratios increase and are slightly more significant when SIMD and residential care home status are adjusted for (adjusted rate ratios of 1.68 and 1.59 using SMR and ONOMAP respectively).They are attenuated and lose significance when further adjustment is made for diabetes status (rate ratios of 1.44 and 1.38 using SMR and ONOMAP respectively).
 
-Using the named based classification for which there are much more data  for those of South Asian ethnicity the rate ratio was below 1  and was statistically significant.
-A further disaggregation of the name based classification into Muslim South Asian and Non-Muslim South Asian ( data not presented) showed risk ratios of similar size below 1 for both subgroups.
-For those of Chinese  ethnicity the name based classification showed lower test positive rates also.
+There is also some evidence of an elevation in risk of "severe test positive or death" among those of Black and South Asian ethnicity.
+The univariate rate ratio among those of Black ethnicity is 1.12 and increases to 1.38 after adjustment for SIMD and residential care home status.
+This effect is observed using the SMR approach since ONOMAP cannot be used to assign Black ethnicity as we describe above.
+Further adjustment for diabetes status reduces the rate ratio to 1.25.
 
-Adjusting for SIMD and residential care home status  did not affect these estimates.
+For those of South Asian ethnicity, the rate ratios associated with "severe test positive or death" using the SMR approach are lower than for "severe test positive disease" and are not significant (univariate rate ratio is 1.23 and rate ratio after adjustment for SIMD and residential care home status is 1.42).
+Using the named based classification, for those of South Asian ethnicity the univariate rate ratio is 1.16 and is not significant.
+Adjustment for SIMD and residential care home status increases the rate ratio to 1.4.
+Further adjustment for diabetes status reduces the rate ratios, using SMR and ONOMAP, to 1.24.
 
-There were increased risk ratios for non-White ethnic groups for test positivity  using SMR  - however note our caveats on interpretation of   this - both because of differential missingness in SMR based ethnicity and because of selection effects on test positive levels.
+Using the SMR approach, there are significant increased rate ratios for those of Black and South Asian ethnicity for test positivity.
+However note our caveats on interpretation of this - both because of differential missingness in SMR based ethnicity and because of selection effects on test positive levels.
+
+Significant increased rate ratios for "hospital or severe disease" were observed among those of South Asian ethnicity using both SMR and ONOMAP approaches.
+
+For those of Chinese ethnicity, lower test positivity was observed using the SMR and ONOMAP approaches.
+Further to this, using SMR data, there was also a reduced risk of "severe test positive disease" and "severe test positive or death" among this ethnic group.
 
@@
 
-We do not find evidence at this stage of elevations in risk of severe COVID-19 disease in non-White ethnic groups compared to White.
+There are borderline significant elevations in rates of having a positive test, being test-positive and hospitalised or being test-positive and severe (CCU or death) rates in those of South Asian origin.
+When deaths with a mention of COVID in those never tested positive are included as fatal cases there is no significant elevation in risk ratio in those of South Asian origin compared to Whites.
+Where elevated risk ratios are seen the are increased by adjusting for SIMD and residential care home status so lower socioeconomic status and care home status does not explain these elevations.
+Further adjustment for diabetes attenuated the risk ratio consistent with higher prevalence of diabetes contributing to the elevations in risk.
+
+For those of Black ethnic origin, there was evidence of elevation in risk of having a positive test and being test-positive and being hospitalised but not of severe disease or severe disease with additional death registrations in those never tested positive included.
+
+For those of Chinese origin here was evidence of a reduction in risk of having a positive test compared to whites.
+Point estimates for hospitalised and severe disease risk ratios were well below 1 but not significantly.
+Absolute numbers were very low for this ethnic group.
 
@@
              align=rep("c", 6), 
-             caption="Associations of hospitalized or severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+             caption="Associations of hospitalized or severe disease among test positives with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+             caption="Associations of severe disease among test positives with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2.7cm") %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.8cm") %>%
+    column_spec(column=7, width="1.5cm")
+```
+
+```{r ethnic5smrDB, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.ethnic5smrDB) <- gsub("_", " ", rownames(table.ethnic5smrDB))
+knitr::kable(table.ethnic5smrDB,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmrDB",
+             col.names=c(clean.header(colnames(table.ethnic5smrDB)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease among test positives with ethnicity (additional adjustment for Diabetes) in those with ethnicity recorded in Scottish Morbidity record") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              caption="Associations of severe disease OR a COVID-19 death with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2.7cm") %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.8cm") %>%
+    column_spec(column=7, width="1.5cm")
+```
+
+```{r ethnic5smrNRSDB, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.ethnic5smrNRSDB) <- gsub("_", " ", rownames(table.ethnic5smrNRSDB))
+knitr::kable(table.ethnic5smrNRSDB,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="ethnicsmrNRSDB",
+             col.names=c(clean.header(colnames(table.ethnic5smrNRSDB)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease OR a COVID-19 death with ethnicity (additional adjustment for Diabetes) in those with ethnicity recorded in Scottish Morbidity record") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
@@
              align=rep("c", 6), 
-             caption="Associations of hospitalized or severe disease with demographic variables including ethnicity based on name classification") %>%
+             caption="Associations of hospitalized or severe disease among test positives with demographic variables including ethnicity based on name classification") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2))  %>%
@@
              align=rep("c", 6),
-             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
+             caption="Associations of severe disease among test positives with demographic variables including ethnicity based on name classification") %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+
+```
+
+```{r onomapDB, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.severe.demogDB) <- gsub("_", " ", rownames(table.severe.demogDB))
+knitr::kable(table.severe.demogDB,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="onomapDB",
+             col.names=c(clean.header(colnames(table.severe.demogDB)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6),
+             caption="Associations of severe disease among test positives with demographic variables (additional adjustment for Diabetes) including ethnicity based on name classification") %>%
     kable_styling(latex_options="hold_position") %>% 
@@
              caption="Associations of severe disease OR any COVID-19 death with demographic variables including ethnicity based on name classification") %>%
+    kable_styling(latex_options="hold_position") %>% 
+    column_spec(column=1, width="2cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    column_spec(column=2, width="1.85cm")  %>%
+    column_spec(column=3, width="1.85cm")  %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="1.5cm")
+
+```
+
+```{r onomapNRSDB, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.severe.demogNRSDB) <- gsub("_", " ", rownames(table.severe.demogNRSDB))
+knitr::kable(table.severe.demogNRSDB,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="onomapNRSDB",
+             col.names=c(clean.header(colnames(table.severe.demogNRSDB)[1:2]),
+                         "Rate ratio (95\\% CI)", "p-value",
+                         "Rate ratio (95\\% CI)", "p-value"),
+             align=rep("c", 6),
+             caption="Associations of severe disease OR any COVID-19 death with demographic variables (additional adjustment for Diabetes) including ethnicity based on name classification") %>%
     kable_styling(latex_options="hold_position") %>% 

### 2020-07-05 casecontrol.Rmd
updates for revised PLoS Med manuscript -- fixes to helper functions

--- 2020-06-12 casecontrol.Rmd
@@
 
-# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol.pdf")
+# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol8June.pdf"); rmarkdown::render("response_reviewers.md")
 
@@
                                  rbind(table.demog.aug[, 1:4], table.demog.append))
+table.demog.append <- table.demog.append[-(1:3), ] # drop onomap classification
 
@@
 
+freqs.all <- as.data.frame(freqs.all)
+table.agegr <- as.data.frame(table.agegr)
+
 ```
@@
 
-_Methods_ -- The design was a matched case-control study.
-Severe cases were all those with a positive nucleic acid test for SARS-CoV-2 in the national database who had entered a critical care unit or died within 28 days of the first positive test.
-Seven controls per case matched for sex, age and primary care practice were selected from the population register.
+_Methods and Findings_ -- The design was a matched case-control study.
+Severe COVID-19 was defined as either a positive nucleic acid test for SARS-CoV-2 in the national database followed by entry to a critical care unit or death within 28 days, or a death certificate with COVID-19 as underlying cause.
+Up to ten controls per case matched for sex, age and primary care practice were selected from the population register.
 All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.
@@
 
-_Findings_ -- There were `r length(which(cc.severe$CASE==1))` severe cases.
+There were `r length(which(cc.severe$CASE==1))` severe cases.
 In a logistic regression using the age-sex distribution of the national population, the odds ratios for severe disease were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
 In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
-Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all[5, 3]` for Type 1 diabetes, `r table.agegr.all[6, 3]` for Type 2 diabetes,
-`r table.agegr.all[8, 3]` for ischemic heart disease,
-`r table.agegr.all[9, 3]` for other heart disease, 
-`r table.agegr.all[10, 3]` for chronic lower respiratory tract disease, 
-`r table.agegr.all[11, 3]` for chronic kidney disease, 
-`r table.agegr.all[12, 3]` for neurological disease, 
-`r table.agegr.all[13, 3]` for chronic liver disease and 
-`r table.agegr.all[14, 3]` for immune deficiency or suppression.
+Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all["Type 1 diabetes", 3]` for Type 1 diabetes, `r table.agegr.all["Type 2 diabetes", 3]` for Type 2 diabetes,
+`r table.agegr.all["Ischaemic heart disease", 3]` for ischemic heart disease,
+`r table.agegr.all["Other heart disease", 3]` for other heart disease, 
+`r table.agegr.all["Asthma or chronic airway disease", 3]` for chronic lower respiratory tract disease, 
+`r table.agegr.all["Chronic kidney disease or transplant recipient", 3]` for chronic kidney disease, 
+`r table.agegr.all["Neurological (except epilepsy) or dementia", 3]` for neurological disease, 
+`r table.agegr.all["Liver disease", 3]` for chronic liver disease and 
+`r table.agegr.all["Immune deficiency or suppression", 3]` for immune deficiency or suppression.
 
-`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all[nrow(freqs.all), 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all[nrow(freqs.all), 1])` of controls had at least one listed condition (`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr[4, 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr[4, 1])` of controls under age 40).
-Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all[2, 3]`] and `r table.agegr.all[3, 3]` respectively] even after adjusting for the listed conditions.
+`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all["listed.any", 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all["listed.any", 1])` of controls had at least one listed condition (`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr["Any listed condition", 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr["Any listed condition", 1])` of controls under age 40).
+Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all["Any prescription", 3]`] and `r table.agegr.all["Any admission", 3]` respectively] even after adjusting for the listed conditions.
 In those without listed conditions significant associations with severe disease were seen across many hospital diagnoses and drug categories.
@@
 A model based on demographic variables, listed conditions, hospital diagnoses and prescriptions provided an additional `r table.densities[3, 5]` bits (C-statistic `r table.densities[3, 3]`).
+A limitation of this study is that records from primary care were not available.
 
@@
 \newpage
+
+# Author summary
+Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill.
+It is The risk of severe or fatal illness is higher in older than in younger people, and is higher in people with conditions such as asthma and diabetes than in people without these conditions.
+Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.
+We show that the conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.
+However the strongest risk factors are age (and residence in a also asthma, the risk of in younger individuals the most frequent risk-a factor is asthma.
+This lays the basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves until the epidemic is over.
 
@@
 # Methods
-## Case definition 
-The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.
-All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` in ECOSS were ascertained for this study.
-Using the Community Health Index (CHI) identifier contained in ECOSS (the CHI number is a unique identifier used in all care systems in Scotland) linkage to other datasets was carried out.
-Hospital admissions from the time of testing were obtained from the RAPID database a daily return of current hospitalisations each day.
-Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that covers admissions to critical care [comprising adult intensive care units (ICUs), high dependency units (HDUs) and combined ICU / HDU units] across Scotland and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic.
-Death registrations up to `r gsub("^0", "", format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y"))` were obtained from linkage to the National Register of Scotland.
+The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+The pharmacoepidemiological analyses will be reported elsewhere.
 
-Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.
-By restricting the case definition to those cases that were fatal or received critical care, we ensured complete ascertainment of all test-positive cases that were severe enough to have been fatal without critical care, whatever selection policies may have determined admission to hospital or entry to critical care.
+## Case definition and selection of matched controls
+ All individuals testing positive for nucleic acid for SARS-CoV-2 were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database, which captures all virology testing in all NHS laboratories nationally.
+Linkage to other datasets was carried out using the Community Health Index (CHI) number, a unique identifier used in all care systems in Scotland.
+Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units and has returned a daily census of patients in critical care from the beginning of the COVID-19 epidemic.
+Death registrations were obtained from linkage to the National Register of Scotland.
+Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
+Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.
 
-## Matched controls
-For each test-positive case, we ascertained ten matched controls of the same sex, one-year age band and registered with the same primary care practice who were alive on the date of the first test in the case using the Community Health Index (CHI) database.
-After removing embarkations there were seven controls per case.
+For each case, the CHI database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive and resident in Scotland on the day as the first date that the case tested positive.
+For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
+To ensure that cases and controls were representative of the same population at risk, the 0.6% of cases who were recorded on the CHI database as no longer alive and resident in Scotland on the day that ECOSS recorded them as testing positive were also excluded.
 As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
 
@@
 Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD), an indicator based on postal code.
-Ethnicity was assigned based on applying a name classification algorithm (ONOMAP) [@lakha_name_2011] to the names in the CHI database.
 For `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls and `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases self-assigned ethnicity, based on the categories used in the Census, had been recorded in Scottish Morbidity Records (SMR).
-Cross-tabulation of `r sum.xtabulate` records for which both name classification and SMR records of ethnicity were available showed that the ONOMAP algorithm had sensitivity of `r round(SouthAsian.sensitivity)`% and specificity of `r round(SouthAsian.specificity, 2)`% for classifying South Asian ethnicity, but misclassified most of those who identified as African, Caribbean or Black.
 
@@
 Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric and maternity unit returns are not captured in SMR01.
-British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.
-The BNF groups drugs by 2-digit chapter codes.
-For this analysis prescription codes from chapters 14 and above, mostly for dressings and appliances but also including vaccines were grouped as "Other".
+British National Formulary (BNF) drug codes for dispensed prescriptions issued in primary care were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016].
+A cutoff date of 15 days before the incident date (date of testing positive for SARS-CoV-2, or 14 days before death for fatal cases without a positive test) was set, and prescriptions dispensed in a 240-day interval before this cutoff date were included.
+For this analysis prescription codes from BNF chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".
 
 We began by scoring a specific list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].
-A separate list of conditions designates "clinically extremely vulnerable" individuals who have been advised to shield themselves completely since early in the epidemic: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
+A separate list of conditions designates "clinically extremely vulnerable" individuals who were advised to shield themselves completely since 24 March 2020: this list includes solid organ transplant recipients, people receiving chemotherapy for cancer, and people with cystic fibrosis or leukaemia.
 We did not separately tabulate these conditions as we expected these individuals to be underrepresented among cases if shielding was adequate.
@@
 
-For the case-control study, all estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`.
+For the case-control study, all estimates of associations with severe COVID-19 were based on conditional logistic regression, implemented as Cox regression in the R function `survival::clogit` [@therneau_modeling_2000].
 Among those cases and controls without any of the pre-defined conditions we then further examined associations of ICD-10 and BNF chapter with severe COVID-19.
@@
 To construct risk prediction models, we used stepwise regression alternating between forward and backward steps to maximize the AIC, implemented in the R function `stats::step`.
+To evaluate the contribution of the listed conditions to risk prediction, and the incremental contribution of other information in hospitalisation and prescription records after assigning these conditions, predictive models were constructed from three sets of variables: a baseline set consisting only of demographic variables, a set that included indicator variables for each listed condition, and an extended set that included demographic, variables, indicator variables for listed conditions and indicator variables for hospital diagnoses in each ICD-10 chapter and prescriptions in each BNF chapter.
+
 The performance of the risk prediction model in classifying cases versus non-cases of severe COVID-19 was examined by `r nfold`-fold cross-validation.
@@
 # Results
+## Comparison of fatal and non-fatal cases
+Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.
+Among those who entered critical care, the median age and the proportion of males were higher in fatal than in non-fatal cases.
+Among fatal cases who did not enter critical care.
+the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.
+
 ## Incidence and mortality from severe COVID-19 in the Scottish population
@@
 Higher risk of severe disease was also associated with socioeconomic deprivation.
-Associations with ethnicity are shown for the full dataset based on name classification and separately for the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record.
-With Whites as reference category, the rate ratio (95% CI) associated with South Asian ethnicity was `r table.demog.append[2, 4]` based on name classification and `r table.demog.append[11, 4]`, based on the subset with SMR records.
-The numbers of cases in other non-White ethnic groups were too sparse to tabulate separately.
+For the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record, the rate ratio (95% CI) associated with South Asian ethnicity with White as reference category was `r table.demog.append["South Asian1", 4]` based on the subset with SMR records.
+There were vert few cases in other non-White ethnic groups.
 
@@
 Prevalence of the listed conditions in cases and controls by age band is shown in Table \ref{tab:agegr}.
-`r gsub("\\\\", "", table.agegr[4, 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr[4, 1])` of the controls.
-In those aged 75+ years `r gsub("\\\\", "", table.agegr[4, 8])` of the cases and `r gsub("\\\\", "", table.agegr[4, 7])` of the controls had at least one listed condition.
-Among those aged under 40 years, `r table.agegr[6, 2]` of the cases and `r table.agegr[6, 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last year.
-Differences in prescription rates  between cases an controls narrowed with increasing age.
+`r gsub("\\\\", "", table.agegr["Any listed condition", 2])` of the cases aged under 40 years had at least one listed condition, compared with only `r gsub("\\\\", "", table.agegr["Any listed condition", 1])` of the controls.
+In those aged 75+ years `r gsub("\\\\", "", table.agegr["Any listed condition", 8])` of the cases and `r gsub("\\\\", "", table.agegr["Any listed condition", 7])` of the controls had at least one listed condition.
+Among those aged under 40 years, `r table.agegr["Diagnosis or prescription", 2]` of the cases and `r table.agegr["Diagnosis or prescription", 1]` of the controls had either a hospital admission in the last five years or a dispensed prescription in the last 240 days.
+Differences in prescription rates between cases and controls narrowed with increasing age.
 
-Over all age groups, `r freqs.all[nrow(freqs.all), 2]` of severe cases and `r freqs.all[nrow(freqs.all), 1]` of controls had at least one of the listed conditions.
+Over all age groups, `r freqs.all["listed.any", 2]` of severe cases and `r freqs.all["listed.any", 1]` of controls had at least one of the listed conditions.
 As shown in Table \ref{tab:listed.conditions.allages}, all the listed conditions were more frequent in cases than controls except for immune conditions in the 75+ age group.
@@
 The rate ratio was `r table.agegr.all[8, 3]` for ischemic heart disease compared to `r table.agegr.all[9, 3]` for the broad category "other heart disease".
-In multivariate analysis ischemic heart disease was not independently associated with severity whereas other heart disease remained strongly associated.
+In multivariable analysis ischemic heart disease was not independently associated with severity whereas other heart disease remained strongly associated.
 In those without a listed condition `r table.notlisted.scripordiag[2, 2]` of the cases and `r table.notlisted.scripordiag[2, 1]` of the controls had either a recent admission or a prescription.
@@
 All listed conditions were associated with severe disease in each age band.
-In those aged under 60 years, the rate ratio was `r tables.agegr[[1]][5, 3]` for Type 1 diabetes and `r tables.agegr[[1]][6, 3]` for Type 2 diabetes.
-The multivariate analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and in each age group  group  any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
-Dispensing of any prescription in the past year was associated with severe disease  in multivariate analyses in the two younger age bands.
+In those aged under 60 years, the rate ratio was `r tables.agegr[[1]]["Type 1 diabetes", 3]` for Type 1 diabetes and `r tables.agegr[[1]]["Type 2 diabetes", 3]` for Type 2 diabetes.
+The multivariable analyses shown in Table \ref{tab:listed.conditions.allages} and \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} show that overall and in each age group  group  any admission to hospital in the past five years were strongly and independently associated with severe disease even after adjusting for care home residence and listed conditions.
+Dispensing of any prescription in the past year was associated with severe disease  in multivariable analyses in the two younger age bands.
 Table \ref{tab:fatal} shows that in each age group the proportion of fatal cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was very low.
-
-In a sensitivity analysis in which we also included deaths registered with mention of COVID-19 on the death certificate in the definition of severe cases (Table \ref{tab:agegr.withNRS}) the same pattern of differences in prior admissions and prescribing history and in listed conditions between cases and controls was found, with the difference in residential are home status being somewhat greater.
-Such deaths were not included in our primary outcome definition as misclassification rates for COVID assignation as cause are unknown.
 
@@
 The association of severe COVID-19 with prior hospital admission was examined further by testing for association of hospitalisations at each ICD-10 chapter level with severe COVID-19, among those without any of the listed conditions.
-These results are shown in Table \ref{tab:conditions}.
+These results are shown in Supplementary Table \ref{tab:conditions}.
 In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 7 (eye) Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-Note that hospital diagnoses classified under the pregnancy chapter here are derived from admissions with pregnancy related medical conditions to non-obstetric units only, as obstetric returns are not in the SMR01 dataset.
-In a multivariate analysis the most significant association was with diagnoses in ICD chapter 2 (neoplasms).
+In a multivariable analysis the strongest associations were with diagnoses in ICD chapters 4 (mental disorders) and 10 (respiratory).
 Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.
-This table is filtered to show only subchapters for which the univariate p-value is <0.001 and where there are at least 50 cases and controls with a diagnosis in this subchapter.
+This table is filtered to show only subchapters for which there are at least 50 cases and controls and the univariate p-value is <0.001.
 This shows that many subchapter diagnoses are associated with markedly higher risk of severe COVID-19.
-<!-- check why cerebrovascular disease does not come out here -->
 
@@
 As shown in Table \ref{tab:listed.conditions.allages} and supplementary tables \ref{tab:listed.conditions.lt60} to \ref{tab:listed.conditions.ge75} ,  encashment of at least one prescription in the last year was associated with severe disease.
-The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.
-In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` respectively.
+The univariate rate ratio associated with this variable varies from `r tables.agegr[[1]]["Any prescription", 3]` in those aged under 60 years to `r tables.agegr[[3]]["Any prescription", 3]` in those aged 75 years and over.
+In a multivariable analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]]["Any prescription", 5]` and `r tables.agegr[[3]]["Any prescription", 5]` respectively.
 
@@
 In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease.
-In a multivariate analysis of all chapters, most of these associations were weaker.
-The BNF chapters with the strongest independent associations with severe disease were chapters 1 (gastrointestinal) and chapters 4 (central nervous system).
-Other chapters associated with severe disease were 2 (cardiovascular),  5 (infections), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
+In a multivariable analysis of all chapters, the strongest independent associations with severe disease were with prescriptions in chapters 1 (gastrointestinal), 4 (central nervous system), 5 (infections), 9 (nutrition and blood) and 14+ (other, mostly dressings and appliances).
 
-## Construction of a multivariate risk prediction model 
-To evaluate the contribution of the listed conditions to risk prediction, and the incremental contribution of other information in hospitalisation and prescription records after assigning these conditions, predictive models were constructed from three sets of variables: a baseline set consisting only of demographic variables, a set that included indicator variables for each listed condition, and an extended set that included demographic, variables, indicator variables for listed conditions and indicator variables for hospital diagnoses in each ICD-10 chapter and prescriptions in each BNF chapter.
-
-For each variable set, a stepwise regression procedure was carried out using alternating forward-backward selection.
-The variables retained with each variable set are shown in Table \ref{tab:stepwise}.
+## Construction of a multivariable risk prediction model 
+ The variables retained from the extended variable set (demographic variables, listed conditions, hospital diagnoses in each ICD-10 chapter, prescriptions in each BNF chapter) are shown in Table \ref{tab:stepwise}.
 Coefficients for specific conditions here should not be interpreted as effect estimates, as global variables for any hospital diagnosis and any listed condition have been included in the model.
@@
 Observed and predicted case status were compared within each stratum over all test folds.
-Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5] ` bits.
+Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5]` bits.
+
+Figure \ref{fig:wevid} shows the distributions in cases and controls of the weight of evidence favouring case over control status from the model based on the extended variable set with a footnote explaining how $\Lambda$ is derived.
+This shows, as expected for a multifactorial classifier, that the distribution in controls is approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
+In cases there is a bimodal distribution, where the second mode represents care home residents.
+Figure \ref{fig:roc} shows the receiver operating characteristic curve with a footnote explaining its derivation from the distributions of the weights of evidence.
 
 This estimate of `r table.densities[3, 5] ` bits for the information conditional on age and sex obtained from the matched case-control study can be added to the information for discrimination `r round(cases.Lambda.agesex, 2)` bits obtained from the logistic regression on age and sex in the population using age and sex to estimate the total information for discrimination of a risk classifier that would be obtained in the population as `r table.densities[3, 5] + round(cases.Lambda.agesex, 2)` bits.
-
-Figure \ref{fig:wevid} shows the distribution of the weight of evidence favouring case over control status from the model based on the extended variable set with a footnote explaining how $\Lambda$ is derived.
-This shows, as expected for a multifactorial classifier, that the distributions are approximately Gaussian: there is no clear divide between high-risk and low-risk individuals of the same age and sex.
-Figure \ref{fig:roc} shows the receiver operating characteristic curve with a footnote explaining its derivation from the distribution of the weights of evidence.
 
@@
 
-Although the numbers of cases and controls of non-White ethnicity are small and the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White ethnic groups up to now.
-The only non-White ethnic group with any sizeable numbers is the South Asian category and we found no clear evidence of any elevation in risk in this group compared to Whites.
-Reports from England [@the_opensafely_collaborative_opensafely_2020] found elevation in risks for some non-White groups.
+Although the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White minority ethnic groups in Scotland, and do not show clear evidence of elevated risk in those who identify as South Asian or Black compared to those who identify as White.
+This contrasts with reports from England [@the_opensafely_collaborative_opensafely_2020] that have found elevated risks in   non-White groups.
 In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.
 In a study of risk factors for hospitalized disease in the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated but did not fully explain the elevated crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity will also differ.
+The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity may also differ.
 For example in the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.
@@
 However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages.
-The rate ratios of `r round(as.numeric(substr(table.agegr.all[5, 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all[6, 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.
+The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.
 We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
-Of note other heart disease is more strongly associated than ischaemic heart disease.
+An unexpected finding was that the risk associated with other forms of heart disease is higher than that associated with ischaemic heart disease.
 This category includes conditions such as atrial fibrillation, cardiomyopathies and heart failure.
-Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all[nrow(freqs.all), 2])` of severe cases had at least one of these listed conditions.
-Among cases and controls without these conditions, not surprisingly, neoplasms were associated with severe COVID-19; we had omitted it from the pre-specified list as in the current dataset we cannot separately identify those who are currently receiving chemotherapy or radiotherapy for whom shielding is advised.
-We have not attempted to estimate the risk associated with these conditions for which shielding is recommended, as the observed risk will depend on the adequacy of shielding rather than on the risk to those exposed to the epidemic.
+One of the highest rate ratios is that associated with chronic kidney disease.
+Prevention of nosocomial transmission in dialysis units may help to reduce this risk.
+Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions.
+In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system.
+We hope to explore this in a separate study that will include linkage to records of shielding advice.
 In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.
@@
 
-An important finding of this study was the strong association of severe COVID-19 with having encashed at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
-Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that the strongest association was with prescription of Chapter 1 drugs, prescribed for gastrointestinal conditions, which are not generally listed as risk factors for severe COVID-19.
-Also associated were those in the  nervous system,  cardiovascular, and nutritional and blood chapters.
-Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.
-These associations are explored in an accompanying paper.
+An important finding of this study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
+Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.
+Although it is likely that most associations of severe COVID-19 with drug prescribing are attributable to the indications for which these drugs were prescribed, or to more diffuse frailty especially in older persons, causal effects of drugs or direct effects of polypharmacy on susceptibility cannot be ruled out.
+These associations are explored in a separate paper.
 
@@
 So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.
+Although for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.
+Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.
+We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population [@mckeigue_evaluation_2020]
 
 As awareness grows of how risk varies between individuals, individuals will seek information about their own level of risk.
-A key implication of our results is that risk of severe or fatal disease is multifactorial and that the rate ratio of `r round(exp(20 * logistic.coeffs[3, 1]), 1)` associated with a 20-year increase in age is stronger than that associated with common diseases such as Type 2 diabetes and asthma that are listed as conditions associated with high risk.
+A key implication of our results is that risk of severe or fatal disease is multifactorial.
+The rate ratio of `r round(exp(10 * logistic.coeffs[3, 1]), 1)` associated with a 10-year increase in age is stronger than the rate ratios associated with common diseases such as asthma or Type 2 diabetes that are listed as conditions associated with high risk.
 A corollary of this is that a crude classification based on assigning all persons with a listed condition to a group for whom shielding is recommended will have poor specificity, as one quarter of those aged 60-74 years in the population have at least one of the listed conditions we examined.
 It will also exclude many people at high risk because they have multiple risk factors each of small effect.
-The only way to optimize risk classification so as to ensure equity with respect to risk is to construct a classifier that uses all available information to assign a risk score.
-Our results show that this is possible in principle, though for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem.
+A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study.
+Thus the rate ratio of `r rr <- round(as.numeric(gsub(" .+", "", table.agegr.all["Type 1 diabetes", 3])), 1); rr` associated with Type 1 diabetes equates to an increase of `r round(log(rr) / logistic.coeffs[3, 1], 1)` years in COVID age.
 In Scotland it is technically possible to use existing electronic health records to calculate a risk score for every individual in the population, though more work would be required to develop this as a basis for official advice and individual decisions.
@@
 However the associations with pre-existing disease are not just with a small set of conditions that contribute to risk, but with many conditions as demonstrated by associations with past medical and prescribing history in relation to multiple physiological systems.
-As countries attempt to emerge from lockdown whist protecting vulnerable individuals, multivariate classifiers rather than crude rule-based approaches will be needed to define those most at risk of developing severe disease.
+As countries attempt to emerge from lockdown whist protecting vulnerable individuals, multivariable classifiers rather than crude rule-based approaches will be needed to define those most at risk of developing severe disease.
 
@@
 ## Information governance
-This study was conducted under approvals from the  Privacy Advisory Committee ref 44/13  and Public Benefit Privacy Protection amendment 1617-0147. Datasets were de-identified before analysis.
+This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.
+Datasets were de-identified before analysis.
+
+## Protocol registration
+The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+
+## Data availability
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
+Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
 
@@
     scale_y_continuous(limits=c(0, 1), expand=c(0, 0)) +
-    xlab("Specificity") +
-    ylab("Sensitivity (reverse scale)")
+    xlab("Specificity (reverse scale)") +
+    ylab("Sensitivity")
 
@@
 ## Footnote for Figure \ref{fig:roc}
-The crude receiver operator characteristic (ROC) curve is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and non-cases.
-The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the score, evaluated over all possible such pairs in the dataset.
+The receiver operator characteristic (ROC) curve is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and non-cases.
+Using the adjusted distributions from Figure \ref{wevid} gives a curve that is concave downwards.
+The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the risk score, evaluated over all possible such pairs in the dataset.
 
@@
 options(knitr.kable.NA = '')
+
+nonmissing.ethnic <- with(cc.severe[!is.na(ethnic4.smr)], table(CASE))
 
@@
     column_spec(column=4, width="2.5cm") %>%
-    kableExtra::group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
-    kableExtra::group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
-    kableExtra::group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
+    kableExtra::group_rows("SIMD quintile", 1, 5, bold=FALSE) %>%
+        kableExtra::group_rows(paste0("Ethnicity based on Scottish Morbidity Record (",
+                                      nonmissing.ethnic[1], " controls, ",
+                                      nonmissing.ethnic[2], " cases)"), 7, 10, bold=FALSE,
                hline_before=TRUE, latex_gap_space="10pt") %>% 
@@
              label="listed.conditions.allages",
+			 linesep=c(rep('', 2), '\\addlinespace', rep('', 9)), 
              digits=c(0, 0, 0, 200, 0, 200), 
@@
              caption="Associations of severe disease with listed conditions over all age groups") %>%
-    # kableExtra::group_rows(" ", 4, nrow(table.agegr.all[-4, ])) %>%
     column_spec(column=1, width="5cm") %>%
     column_spec(column=2:7, width="2.5cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
     landscape() %>%
@@
 
-```
-
-```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
-
-rownames(table.conditions.aug) <- gsub("_", " ", rownames(table.conditions.aug))
-knitr::kable(table.conditions.aug,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="conditions",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.conditions.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe disease with hospital diagnoses in last 5 years, in those without any listed condition") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape() %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
-
-rownames(table.drugs.aug) <- gsub("_", " ", rownames(table.drugs.aug))
-knitr::kable(table.drugs.aug,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="drugs",
-             digits=c(0, 0, 0, 200, 0, 200), 
-             col.names=c(clean.header(colnames(table.drugs.aug)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6),
-             caption="Associations of severe disease with prescribed drugs in those without any listed condition") %>%
-    column_spec(column=1, width="4.5cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    landscape() %>%
-    kable_styling(latex_options="hold_position") 
 ```
@@
 
+```{r casegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.casegr,
+             escape=FALSE, 
+             booktabs=TRUE,
+			 longtable=TRUE, 
+             label="casegr",
+             col.names=paste(c("Critical care, non-fatal", "Critical care, fatal",
+                               "No critical care, fatal", "No critical care, fatal"),
+                             paste0(rep("(", 4),
+                                    with(cc.all, table(severe.casegr)), rep(")", 4))), 
+             caption="Comparison of severe non-fatal and fatal cases, by test positive status and entry to critical care") %>%
+    add_header_above(c(" "=1, "Test-positive"=3, "No positive test"=1)) %>%
+    add_header_above(c(" "=1, "Non-fatal"=1, "Fatal"=3)) %>%
+    column_spec(column=2:5, width="3cm") %>%
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
 ```{r listed.conditions.lt60, echo=FALSE, warning=FALSE, message=FALSE}
@@
              label="listed.conditions.lt60",
+			 linesep=c(rep('', 2), '\\addlinespace', rep('', 9)), 
              digits=c(0, 0, 0, 200, 0, 200), 
@@
              caption="Associations of severe disease with listed conditions in those aged less than 60") %>%
-#   kableExtra::group_rows(" ", 4, nrow(tables.agegr[[1]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
     column_spec(column=2:7, width="2.7cm") %>%
-   add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+   add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
     landscape() %>%
@@
              label="listed.conditions.60to74",
+			 linesep=c(rep('', 2), '\\addlinespace', rep('', 9)), 
              digits=c(0, 0, 0, 200, 0, 200), 
@@
              caption="Associations of severe disease with listed conditions in those aged 60-74 years") %>%
- #   kableExtra::group_rows(" ", 4, nrow(tables.agegr[[2]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
     column_spec(column=2:7, width="2.7cm") %>%
-   add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+   add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
     landscape() %>%
@@
              label="listed.conditions.ge75",
+			 linesep=c(rep('', 2), '\\addlinespace', rep('', 9)), 
              digits=c(0, 0, 0, 200, 0, 200), 
@@
              caption="Associations of severe disease with listed conditions in those aged 75 years and over") %>%
-#    kableExtra::group_rows(" ", 4, nrow(tables.agegr[[3]]) - 1) %>%
     column_spec(column=1, width="4cm") %>%
    column_spec(column=2:7, width="2.7cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
     landscape() %>%
@@
 
-```{r agegr.withNRS, echo=FALSE, warning=FALSE, message=FALSE}
+```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
 
-table.agegr.withNRS <- readRDS(table.agegr, file="table.agegr.withNRS.rds")
-
-rownames(table.agegr.withNRS) <- replace.names(rownames(table.agegr.withNRS))
-table.agegr.withNRS.print <- table.agegr.withNRS[c(1:4, 6, 8:nrow(table.agegr.withNRS)), ] 
-knitr::kable(table.agegr.withNRS.print, 
+rownames(table.conditions.aug) <- gsub("_", " ", rownames(table.conditions.aug))
+knitr::kable(table.conditions.aug,
              escape=FALSE, 
              booktabs=TRUE,
-             label="agegr.withNRS",
-			 linesep=c(rep('', 4), '\\addlinespace', rep('', 9)), 
-             align=rep("c", 8), 
-             caption="Frequencies of risk factors in cases and controls when deaths with mention of COVID-19 on the death certificate were included as cases, by age group") %>%
+             label="conditions",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(clean.header(colnames(table.conditions.aug)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe disease with hospital diagnoses in last 5 years, in those without any listed condition") %>%
     column_spec(column=1, width="4cm") %>%
-    column_spec(column=2:3, width="1.6cm") %>%
-    column_spec(column=4:7, width="1.8cm") %>%
-    column_spec(column=8:9, width="2cm") %>%
-    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
-                       "60-74 years"=2, "75+ years"=2)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
     landscape() %>%
@@
 
+```
+
+```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.drugs.aug) <- gsub("_", " ", rownames(table.drugs.aug))
+knitr::kable(table.drugs.aug,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="drugs",
+             digits=c(0, 0, 0, 200, 0, 200), 
+             col.names=c(clean.header(colnames(table.drugs.aug)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 6),
+             caption="Associations of severe disease with prescribed drugs in those without any listed condition") %>%
+    column_spec(column=1, width="4.5cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
+    landscape() %>%
+    kable_styling(latex_options="hold_position") 
 ```
@@
              label="icdsubchapters",
+			 linesep=c('', '\\addlinespace',
+                       rep('', 2), '\\addlinespace', # neoplasma
+                       '\\addlinespace', # metabolic
+                       rep('', 2), '\\addlinespace', # nervous
+                       '', '\\addlinespace', # respiratory
+                       '\\addlinespace', # skin
+                       '', '\\addlinespace', # urinary
+                       rep('', 8)), 
              col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),

### 2020-07-05 pharmaco.Rmd
updates for revised PLoS Med manuscript -- fixes to helper functions

@@
 ---
-title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)'
+title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study'
 author: 
@@
     affiliation: HPS
+  - name: Bruce Guthrie
+    affiliation: Usher
   - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
@@
 
-_Background_ -- Dispensing of at least one prescription in the past year is a risk factor fo severe COVID-19 in Scotland, even in those without diagnoses of conditions designated as conferring increased risk.
+_Background_ -- Dispensing of at least one prescription in the past year is a risk factor for severe COVID-19 in Scotland, even in those without diagnoses of conditions designated as conferring increased risk.
 
 _Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic.
-Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care since June 2019.
+Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.
 
@@
 
-For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+For each case, the Community Health Index database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
 For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
-For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
 
 ## Morbidity and drug prescribing
-For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry.
+For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry.
 Diagnoses of diabetes were extracted from linkage to the national diabetes register.
-British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2.
-The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.
-For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".
-Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
+British National Formulary (BNF) drug codes for dispensed prescriptions issued in primary care were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016].
+A cutoff date of 15 days before the incident date (date of testing positive for SARS-CoV-2, or 14 days before death for fatal cases without a positive test) was set, and prescriptions dispensed in a 240-day interval before this cutoff date were included.
+For this analysis prescription codes from BNF chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".
+Defined daily doses (DDDs) of each proton pump inhibitor were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
 Morphine milligram equivalents (MMEs) for opiates were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).
@@
 This dose-response relationship was absent in those aged 75 years and over.
-In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 2 DDDs/day MME /day were `r table.dosegr.protonpump.notlisted["2 or more", 4]` and `r table.dosegr.protonpump.notlisted["2 or more", 6]` respectively.
+Among those in this age group without any listed condition, the unadjusted and adjusted rate ratios associated with more than 1.5 DDDs/day were `r table.dosegr.protonpump.notlisted.agelt75["1.5 or more", 4]` and `r table.dosegr.protonpump.notlisted.agelt75["1.5 or more", 6]` respectively.
 Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.
@@
 Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
-It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.
+It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates [@weiner_opioid_2017] has been reported previously.
 

### 2020-07-05 casecontrol.R

@@ -1,6 +1,6 @@
 ## analysis script for case-control study
 if(exists("cl")) {
-    parallel::stopCluster(cl)
+    # parallel::stopCluster(cl)
     showConnections(all = TRUE)
     closeAllConnections()
 }
@@ -23,7 +23,6 @@ list.of.packages = c("car",
                      "pander", 
                      "ggplot2", 
                      "doParallel", 
-                     # "reshape2",  # deprecated
                      "readxl", 
                      "DescTools", 
                      "icd.data", 
@@ -43,7 +42,7 @@ library(rmarkdown)
 library(pander)
 library(ggplot2)
 library(doParallel)
-library(reshape2)
+# library(reshape2)
 library(readxl)
 library(DescTools)
 library(icd.data)
@@ -61,9 +60,10 @@ include.nrs.only <- TRUE
 #include.nrs.only <- FALSE # uncomment to exclude NRS-only deaths
 
 stepwise <- TRUE   
-# stepwise <- FALSE ## uncomment to save time if old version still valid
+stepwise <- FALSE ## uncomment to save time if old version still valid
 
 fatal.predict <- TRUE
+fatal.predict <- FALSE
 
 recode.dmtype <- function(x) {
     r <- as.factor(car::recode(x, 
@@ -78,21 +78,6 @@ recode.dmtype <- function(x) {
     return(r)
 }
 
-cases.old <- as.data.table(readRDS("./data/CC_linked_ANON_20200515.rds"), key="ANON_ID")
-cases.new <- as.data.table(readRDS("./data/CC_linked_ANON_2020-06-18.rds"), key="ANON_ID")
-cases.old <- subset(cases.old, CASE==1)
-cases.new <- subset(cases.new, is.case)
-
-setnames(cases.old, "age", "AGE")
-
-setnames(cases.new, "SPECIMENDATE", "SPECDATE")
-setnames(cases.new, "AgeYear", "AGE")
-
-cases.old$dm.type <- recode.dmtype(cases.old$dm.type)
-cases.new$dm.type <- recode.dmtype(cases.new$dm.type)
-
-with(subset(cases.old, AGE<50), table(dead28 | covid_cod, icu==1 | hdu==1))
-with(subset(cases.new, AGE<50), table(dead28 | covid_cod, icu==1 | hdu==1))
 
 ##################################################################
      
@@ -103,7 +88,10 @@ if(old) {
[…]

### 2020-07-05 drugs.R

@@ -1,14 +1,11 @@
 
-getdaysbeforetest <- function(scrips.in, specdates) {
-    ## merge SPECDATE and dispensing.days into scrips table and add columns daysbefore, intervalTWday
-    ## checked that minimum value is 16 days -- the 15-day cutoff has been applied to the scrips table
-    setkey(scrips.in, ANON_ID)
-    setkey(specdates, ANON_ID)
-    scrips <- specdates[scrips.in]
-    scrips$daysbefore <- as.integer(scrips$SPECDATE - scrips$dispensed_date)
-    scrips$intervalTWday <- ceiling((scrips$daysbefore - 15) / TW)
-    scrips$intervalTWday[scrips$intervalTWday > 3] <- 3
-    return(scrips)
+get.timewindow <- function(scrips.in) {
+    ## add columns daysbefore, intervalTWday, dispensing.days
+    ## the 15-day cutoff has been applied to the scrips table
+    scrips.in$intervalTWday <- ceiling((scrips.in$daysbefore - 15) / TW)
+    scrips.in$intervalTWday[scrips.in$intervalTWday > 2] <- 2
+    scrips.in$dispensing.days <- 240  # same for all individuals
+    return(scrips.in)
 }
 
 getdoseTWday.wide <- function(scrips.in, varname.prefix="dose.") {
@@ -31,7 +28,7 @@ getdoseTWday.wide <- function(scrips.in, varname.prefix="dose.") {
                                        fun.aggregate=sum)
     colnames(doseTWday.wide)[-1] <-
         paste0(varname.prefix,
-               c("interval1", "interval2", "interval3"))
+               c("interval1", "interval2"))
     return(doseTWday.wide)
 }
 
@@ -110,6 +107,7 @@ scrips.opioid <-
     readRDS(scrips.filename) %>%
     subset(bnf_paragraph_code=="0407020")
 
+    
 scrips.opioid$mode <- car::recode(scrips.opioid$formulation_code,
                                   "c('CAPS', 'LOZ', 'SACH', 'SOLN', 'TABS')='Oral';
                                     'PATCH'='Patch'; 'INJ'='Injection';")
@@ -282,8 +280,7 @@ TW <- 120 # time window in days
 
 ####### proton pump exposure   ##################
 
-scrips.protonpump <- getdaysbeforetest(scrips.in=scrips.protonpump,
[…]

### 2020-07-05 helperfunctions.R

@@ -1,5 +1,27 @@
 ## helperfunctions for COVID analyses
 
+recode.indicator <- function(x) {
+    x[is.na(x)] <- 0
+    x[x > 1] <- 1
+    return(as.factor(x))
+}
+
+pnorm.extreme <- function(z, upper=TRUE) {
+    ## https://www.johndcook.com/blog/norm-dist-bounds/
+    if(upper) { # upper bound on log10 tail prob
+        c <- 8 / pi
+    } else { # lower bound
+        c = 4 
+    }
+    x <-  2 * sqrt(2 / pi) / (z + sqrt(z^2 + c))
+    ln_p = -0.5 * z^2 + log(x)
+    log10p <- ln_p / log(10)
+    exponent <- floor(log10p)
+    coeff <- 10^(log10p - exponent)
+    string <- paste0(round(coeff), "E", exponent)
+    return(string)
+}
+
 logit <- function(x) log(x) - log(1 - x) 
 
 invlogit <- function(x) 1 / (1 + exp(-x))
@@ -72,7 +94,8 @@ stepwise.union.dropcols <- function(x, y, covariates=NULL, stratum) {
     return(stepwise.drop[, -1])
 }
 
-lookup.names <- data.frame(varname=c("deathwithin28", "scrip.any", "diag.any", "care.home",
+lookup.names <- data.frame(varname=c("AGE", "sex",
+                                     "deathwithin28", "scrip.any", "diag.any", "care.home",
                                      "diag.other", "scripordiag", 
                                      "emerg", "icu.hdu.ccu", "inpat",
                                      "protonpump",
@@ -111,7 +134,8 @@ lookup.names <- data.frame(varname=c("deathwithin28", "scrip.any", "diag.any", "
                                      "statins6.bnf",
                                      "hydroxychloroquine6.bnf"
                                      ),
-                           longname=c("Death within 28 days of test",
+                           longname=c("Age", "Males",
+                                      "Death within 28 days of test",
                                       "Any prescription", "Any admission", "Care home",
                                       "No listed condition, other diagnosis",
                                       "Diagnosis or prescription",
[…]

### 2020-07-05 pharmaco.R

@@ -15,7 +15,6 @@ table.numdrugsgr.listed.any <- data.frame(numdrugsgr=rep(levels(cc.severe$numdru
                                      table.numdrugsgr.listed.any)
 colnames(table.numdrugsgr.listed.any)[2:3] <- paste0(c("Controls (", "Cases ("), as.integer(table(cc.severe$CASE)), rep(")", 2))
 
-
 ########################################################################
 ## in those without listed conditions, by age group
 
@@ -222,27 +221,35 @@ colnames(table.dosegr.protonpump)[2:3] <- paste0(c("Controls (", "Cases ("),
                                                  rep(")", 2))
 
 ## print(table.dosegr.protonpump)
-x <- summary(clogit(formula=CASE ~ DDDsgr + strata(stratum), 
-                    data=subset(cc.severe, notlisted)))$coefficients
+
+cc.severe$DDDsgr4 <- car::recode(cc.severe$DDDsgr,
+                                 "c('1.5', '2 or more')='1.5 or more'")
+
+withcovariates.formula <- as.formula(paste("CASE ~ DDDsgr4 +",
+                                           ppi.covariates.string, "+ strata(stratum)"))
+coeff.rows <- 1:3
+
+x <- summary(clogit(formula=CASE ~ DDDsgr4 + strata(stratum), 
+                    data=cc.severe[notlisted & AGE < 75]))$coefficients
 x <- as.data.frame(x)
 x$u.ci <- or.ci(x[, 1], x[, 3])
 x$u.pvalue <- pvalue.latex(x[, 5])
 x <- x[, c("u.ci", "u.pvalue")]
 y <- summary(clogit(formula=withcovariates.formula,
-                    data=subset(cc.severe, notlisted)))$coefficients[coeff.rows, ]
+                    data=cc.severe[notlisted & AGE < 75]))$coefficients[coeff.rows, ]
 y <- as.data.frame(y)
 y$m.ci <- or.ci(y[, 1], y[, 3])
 y$m.pvalue <- pvalue.latex(y[, 5])
 y <- y[, c("m.ci", "m.pvalue")]
 x <- data.frame(x, y)               
 x <- rbind(rep(NA, ncol(x)), x)
-freqs <- paste.colpercent(with(subset(cc.severe, notlisted), table(DDDsgr, CASE)))
+freqs <- paste.colpercent(with(cc.severe[notlisted & AGE < 75], table(DDDsgr4, CASE)))
[…]

### 2020-07-05 stepwise.R

@@ -60,7 +60,9 @@ if(stepwise) {
     ## test.folds has one row per stratum, one column per fold containing indicator vars
     ## merge by stratum adds single column test.fold
     cv.data <- merge(cc.nonmissing, test.folds, by="stratum")
-    cv.data <- subset(cv.data, select=c("test.fold", "CASE", "stratum", full.varnames))
+    select.cols <- c("test.fold", "CASE", "stratum", full.varnames)
+    ## .. prefix evaluates select.cols in the parent environment 
+    cv.data <- cv.data[, ..select.cols]
     cat("cv.data has", nrow(cv.data), "rows\n")
     
 ###########################################

### 2020-07-05 trainmodel.R

@@ -1,0 +1,143 @@
+## train model to predict fatal disease
+
+library(rstan)
+
+rstan.sampling <- function(model=rstan.model, data=stan.data, pars=pars,
+                           iter=1000, warmup=500,
+                           open_progress=FALSE,
+                           adapt_delta=0.8, 
+                           chains=4,
+                           seed=12345) {
+    
+    cat("sampling via rstan ... ")
+    samples.mc <- rstan::sampling(object=model, data=data,
+                                  pars=pars, 
+                           iter=iter, warmup=warmup,
+                           control=list(adapt_delta=adapt_delta),
+                           chains=chains, seed=seed,
+                           open_progress=open_progress)
+    cat("done\n")
+    return(samples.mc)
+}
+
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+qr <- TRUE
+
+clogit.model <- stan_model(file="~/hsstan/inst/stan/hs_clogit.stan")
+
+## restrict to fatal cases + matched controls, and complete data on covariates
+select.cols <- c("CASE", "stratum", "care.home", "diabetes.any",
+                 grep("^Ch\\.", colnames(cc.severe), value=TRUE))
+
+
+cc.nonmissing <- na.omit(cc.severe[fatal.casegroup==1], cols=select.cols)
+cc.nonmissing <- cc.nonmissing[, ..select.cols]
+covariate.names <- select.cols[-(1:2)]
+str(cc.nonmissing)
+cc.nonmissing <- cc.nonmissing[, lapply(.SD, as.numeric), by=row.names(cc.nonmissing)][, -1]
+str(cc.nonmissing)
+              
+## drop strata with num cases not equal to 1, or < 2 observations 
+cc.drop <- cc.nonmissing[CASE==1, .(.N), by = .(stratum)][N != 1, ]
+cc.nonmissing <- cc.nonmissing[!(stratum %in% cc.drop$stratum)]
+cc.drop <- cc.nonmissing[, .(.N), by = .(stratum)][N < 2, ]
+cc.nonmissing <- cc.nonmissing[!(stratum %in% cc.drop$stratum)]
+              
+table(cc.nonmissing[, .(.N), by = .(stratum)][, N])
+
+## renumber levels of stratum   
+cc.nonmissing[, stratum := as.integer(as.factor(as.integer(stratum)))]
[…]

### 2020-07-12 response_reviewers.Rmd
updated pharmaco scripts

---
title: "PMEDICINE-D-20-02483R1: authors' response to reviewers' comments"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
output: pdf_document

# rmarkdown::render("response_reviewers.md")

---

# Requests from the editors:

> Please ensure that the paper adheres to the PLOS Data Availability Policy

We have included the following data availability statement on the submission portal and in the manuscript:

The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on https://github.com/pmckeigue/covid-scotland_public

> Please structure the abstract using the PLOS Medicine headings (Background, Methods and Findings, Conclusions - "Methods and Findings" should be a single subsection heading).

Done

> In the last sentence of the Abstract Methods and Findings section, please include a brief note about any key limitation(s) of the study's methodology.

We have added a sentence to note that records from primary care were not available.

> At this stage, we ask that you include a short, non-technical Author Summary of your research to make findings accessible to a wide audience that includes both scientists and non-scientists. The Author Summary should immediately follow the Abstract in your revised manuscript. This text is subject to editorial change and should be distinct from the scientific abstract. Please see our author guidelines for more information: https://journals.plos.org/plosmedicine/s/revising-your-manuscript#loc-author-summary

Done

> Did your study have a prospective protocol or analysis plan? Please state this (either way) early in the Methods section.

We have inserted this sentence into the Methods: "The study followed a pre-specified protocol. Modifications after study started were that we aligned the list of conditions of interest to be consistent with those designated as moderate risk conditions by public health agencies, and extended the list of drug classes under study to include all drugs."

> The authors could consider using an appropriate reporting guideline (eg, RECORD - https://www.equator-network.org/reporting-guidelines/record/ - designed to support reporting of observational studies using routinely-collected data) to enhance reporting of methods and findings in their study. If the authors choose to use this please upload the completed RECORD (or other) checklist as supporting information with the revised paper.

RECORD checklist uploaded

# Note for all reviewers. 
As discussed with the editor in this revised manuscript we have used an updated dataset that includes all deaths certified with COVID-19 as underlying cause, irrespective of whether a positive nucleic acid test for SARS-CoV-2 was obtained. The data capture was extended from May 15th to June 8th.

This update has changed all the numbers in the manuscript, but none of the conclusions have changed.  Our reproducible research pipeline allows us to regenerate the entire manuscript from this updated dataset with a single keystroke.

Using this update deals with several points raised by reviewers, including the inclusion of deaths among those who did not test positive, and the incomplete recording of ethnicity.

# Reviewer #1:

> Full case ascertainment of severe disease arguably limited by LTC deaths that do not reach hospital not accounted for, and may bias the rate ratios described.

 As noted above, this updated dataset includes deaths certified with COVID-19 as underlying cause among people without a positive test for SARS-CoV-2.  These are tabulated separately in Supplementary Table 1.

> Challenges are the sheer number of univariate comparisons reported, and the limited  corrections for multiple comparisons in place.

By reporting exact p–values, we allow readers to make their own assessment of the strength of evidence for association allowing for the size of the prior hypothesis space (number of independent tests).  This practice is endorsed by Reviewer #2.

> Table 5 and 6 could be placed in the supplementary appendix, unless the purpose is to emphasize the non-specificity of the associations.

We have moved these tables to the supplementary appendix

> The presence of less robust associations in common conditions such as ischemic heart disease, and more robust associations in rare diseases like chronic kidney disease, is notable. This is mentioned in the discussion, but should be understood better, given the conclusions.

We have added a sentence to note the relevance of non-ischaemic heart disease and kidney disease to the involvement of these systems in severe COVID-19.

> A great deal of the Results should be moved to the Methods, such as the multivariable model construction, and the prescription association data.

We have moved the material describing the stepwise regression procedure to the Methods section

> Removing neoplasms from the prespecified list because of inability to discern who has already been shielded belies the fact that shielding is i) not 100% applied in populations; ii) not practiced in other regions, making the association between neoplasm and outcome an important one to consider for policy in other regions.

We have rewritten this sentence to clarify that neoplasms are not on the designated list of conditions that the National Health Service uses to classify "people at moderate risk", though primary neoplasms of blood or bone marrow and cancer treatments affecting the immune system are on the list used to classify "clinically extremely vulnerable" individuals who are advised to shield themselves.  We have added a sentence to the Discussion to state that we are unable to investigate adequately the risk associated with neoplasms in this study, but that we hope to investigate it in a further study that will include linkage to records of shielding advice.

> Abbreviation 'scrip' in Table 4 not universal, would use 'prescription'

We have made this substitution.

> Very hesitant, given the sample size, the inaccurate labelling problem, the exclusive examination of one ethnic group, and the limited generalizability, to include any reporting or conclusions on the ethnicity-based risk factors. Would suggest excluding altogether, and reporting on the need for more robust ethnicity-based data collection strategies

This updated dataset has more complete recording of self assigned ethnicity as the ascertained was widened to include more SMR datasets and longer lookback thus yielding data on 85% of cases and 74% of controls.  Therefore we have dropped the sensitivity analysis based on name classification which lacked accuracy as noted by the referee.  We now describe the numbers and rate ratios for South Asia, Black and other. We agree that the numbers are too small to make definitive conclusions. However if we do not include any data on ethnicity readers will query this given the amount of concern and publicity there has been about large ethnic differences in risk in the rest of the UK.   Therefore we have made a clear statement in discussion that the numbers are too small to provide definitive statements on risk ratios.

# Reviewer #2:

> Page 3 I understand (probably from Sheila Bird) that death registration in Scotland is much more timely than south of the border but is there any important bias due to delays? I might just give a little bit more detail here as death registration practices are not the same all over the world.

We have edited the Methods section to give the cutoff date for ascertainment of death registrations (rather than deaths) and to note that 94% of deaths among cases were registered within 5 days.

> Page 3 It took me a couple of attempts to see what the sentence starting ‘By restricting’ meant. How about ‘Using this definition means we include all severe cases even if they die without entering critical care.’  That might be too abbreviated so feel free to re–write.

We have rewritten this sentence for clarity as suggested

> Page 3 In my idiolect embarkation means boarding a ship. Does this mean emigration? Whatever it means did the authors really start with 2755 strata of size 10 + 1 and end up with them all of 7 + 1 as this seems to imply. I think I have misunderstood here. I also note that (Table 1) 19670 2755 = 7.14 not 7.

We have rewritten the section on selection of matched controls. In this updated dataset the exclusion of those no longer in Scotland was done before selection of matched controls, so we are closer to the original design target of ten controls per case

> Pages 4 and 5 I think it would be good to cite both gam and survival properly. It is the only credit package authors get. (Disclaimer: I am a package authors although not of these.)

We have added the citation given by the authors of the `survival` package.  For the `gam` package, the author does not give a specific citation.

> Page 5 If the incidence and mortality figures are a core part of the authors’ thesis then I would suggest making some comparison between this and the corresponding figures for all–cause mortality. I sense though that this is perhaps just background. Is there any reason to give overall figures and then just men?

Others have compared COVID-19 mortality with all-cause mortality: a full exploration of this important question is beyond the scope of this paper.

Our figure shows mortality and incidence for men and women separately -- it's possible that the lines for women were not visible to the reviewer.

> Page 6 If the authors are going, correctly, to do a sensitivity analysis for different ways of identifying ethnicity then should it not be flagged in the methods? > Page 6 Despite what it says here Table 1 does show the figures for Black as well as White and South Asian and the risk ratio for Black seems indistinguishable from that for South Asian.

We have deleted the sensitivity analyses regarding using name-based ethnicity classification as we now have more complete recording of  self assigned ethnicity  by doing longer and wider lookback in SMR records.  We now report for White, South Asian, Black and other categories.

> Page 6 The authors state here that Table 3 shows listed conditions broken down by age (so that we see that immune conditions were not more common in 75+). I do not see such a breakdown there.

We have corrected this table reference

> Page 7 The phrase ‘most significant association was with diagnoses in ICD chapter 2’ seems to equate the smallness of the p–value with importance since several of the other risk ratios in Table 5 are of comparable size or even greater than that for neoplasms.

We have rewritten this.

> Page 7 Viewing this from the perspective of a future researcher who finds something apparently different in their data I wonder whether the criteria for Table S5 are too restrictive. I can understand excluding rare conditions (the current limit is 50) but if you restrict by p–value people who get a similar risk ratio to the present authors will not be aware of the similarity if the current authors’ p–value was > 0.001.

We have filtered this table by p-value to keep the length of the table within reasonable limits.  We plan to make more detailed summary statistics available elsewhere.

> Page 8 One bit does not seem very much information to me and I find it hard to see how it would work in practice.

We have added a more detailed explanation of this under "Relevance to policy" in the Discussion -- the total information for discrimination is the sum of the information from the logistic regression on age and sex and the information from the conditional logistic regression on risk factors, conditioned on age and sex.

> Page 8 While it is welcome to see the authors challenge the notion that everybody whose family origins are in South Asia is the same I wonder how much sense this is going to make to people outside the UK or South Asia. International readers may not know how and why people from the disparate communities came to the UK nor why this issue seems so salient in the UK discourse about COVID–19.

Given the comments of Reviewer #1 we have shortened this part of the discussion making clearer the background concern from other studies in the UK but keeping our inference about the data in Scotland brief and noting the wide confidence intervals.

> Page 14 I assume the black dots are hidden behind the black curve?  Strictly speaking multivariable is meant throughout I think (Hidalgo and Goodman, 2013). Multivariate means multiple variables on the left hand side which is not, as far as I can see, true of the models used. I know this is very picky.

We have substituted the term multivariable for multivariate.

> The authors comment on page 8 that they have shown that the risk of a severe outcome increases with age. As far as the death component is concerned that was true of course before this particular virus appeared. According to the English Life Tables ELT17 the value of mx for a male my age was 0.025 and for a male ten years older than me 0.075 which is the same sort of increase with age as the authors are reporting for their severe outcome. What we do not know is what the excess risk is over and above the background risk and how that varies with age.

We have added an explanation in the Discussion of how rate ratios can be interpreted as increases in "Covid age".

-----------------------------------------------------------

# Reviewer #3

> It would be informative if the authors could break down the severe cases into those that were due to death within 28 days without admission to critical care units, those that were admitted to critical care units and survived, those admitted to critical care but perished. It would be interesting to analyze whether what proportion of severe cases among care home residents were death within 28 days without admission to critical care units. Providing the severe cases breakdown and a more in-depth discussion will further strengthen the conclusions.

We have added a supplementary table S1 to provide this breakdown of severe cases, and described what it shows in the first paragraph of Results.  As Reviewer #3 suggested, this shows clearly that almost all fatal cases in care home residents did not enter critical care.

> Since the authors concluded that many preexisting conditions were associated with severe COVID-19, may consider adding ANY comorbidity in Table 3.

The "any comorbidity" variable is implicit in the variable "any prescription or diagnosis".  Other than diabetes which was ascertained from a population register, all other comorbidities are based on dispensed prescriptions or hospital diagnoses.

### 2020-07-12 casecontrol.Rmd
updated pharmaco scripts

@@
   \def\quote{\origquote\itshape}
+  \usepackage{soul}
   \usepackage{graphicx}
@@
 
+delay <- with(cc.severe, table(DATE_OF_REGISTRATION - Date.Death))
+percent.withindays <- round(100 * cumsum(delay) / sum(delay))
+
 ```
@@
 Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.
-We show that the conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.
-However the strongest risk factors are age (and residence in a also asthma, the risk of in younger individuals the most frequent risk-a factor is asthma.
-This lays the basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves until the epidemic is over.
+We show that the strongest risk factor, apart from age, is residence in a care home.
+The conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.
+This lays a basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves when there cases in their neighbourhood.
 
@@
 # Methods
-The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
-The pharmacoepidemiological analyses will be reported elsewhere.
+\hl{The study followed a pre-specified protocol.
+Modifications after study started were that we aligned the list of conditions of interest to be consistent with those designated as moderate risk conditions by public health agencies, and extended the list of drug classes under study to include all drugs.}
 
@@
 Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
-Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.
+\hl{Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.}
 
@@
 As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
-For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths registered up to `r format(max(cc.severe$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
+Among fatal cases, `r percent.withindays[6]`% of deaths were registered within 5 days.
 
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.
+
+Identifiable data were extracted by the Public Health Scotland CHI database and linkage team and de-identified before provision to the analysis team.
 
@@
 # Results
-## Comparison of fatal and non-fatal cases
-Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.
-Among those who entered critical care, the median age and the proportion of males were higher in fatal than in non-fatal cases.
-Among fatal cases who did not enter critical care.
-the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.
-
 ## Incidence and mortality from severe COVID-19 in the Scottish population
@@
 Higher risk of severe disease was also associated with socioeconomic deprivation.
-For the subset of cases and controls in whom ethnicity had been recorded in the Scottish Morbidity Record, the rate ratio (95% CI) associated with South Asian ethnicity with White as reference category was `r table.demog.append["South Asian1", 4]` based on the subset with SMR records.
-There were vert few cases in other non-White ethnic groups.
+In the   `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases  and  `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls  in whom ethnicity had been recorded in the Scottish Morbidity Record, there were few non-White cases or controls and the confidence limits for the risk ratios for severe or fatal disease by ethnic group were wide.
 
@@
 Table \ref{tab:fatal} shows that in each age group the proportion of fatal cases who had not had either a hospital admission in the last five years or a dispensed prescription in the last year was 	very low.
+
+## Comparison of fatal and non-fatal cases
+
+Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.
+\hl{Severe cases who entered critical care were much younger than severe cases never entering critical care.
+The majority of severe cases from residential care homes never entered critical care.
+Among fatal cases who did not enter critical care.
+the distribution of age and other risk factors was similar in those with and without a positive test result, except that the proportion of care home residents was higher among those without a positive test result.
+Among  those entering critical care  the median age and the proportion of males  and prevalence or prior comorbidities were higher in fatal than in non-fatal cases.
+}
 
@@
 
-Although the assignment of ethnicity is incomplete, the results give some indication of the likely upper bound of the absolute numbers of severe cases in non-White minority ethnic groups in Scotland, and do not show clear evidence of elevated risk in those who identify as South Asian or Black compared to those who identify as White.
-This contrasts with reports from England [@the_opensafely_collaborative_opensafely_2020] that have found elevated risks in   non-White groups.
+Although the assignment of ethnicity is incomplete and the prevalence of non-White ethnic groups in Scotland is low, we show the data as they give some indication of the likely upper bound of the absolute numbers of severe cases in South Asian and Black ethnic groups in Scotland.
+This is relevant because substantial elevations in risk of hospitalised and fatal COVID-19 have been reported elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020; @niedzwiedz_ethnic_2020; @ho_modifiable_2020].
 In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.
-In a study of risk factors for hospitalized disease in the UK Biobank cohort, adjustment for health care worker status and other social variables attenuated but did not fully explain the elevated crude risk ratios associated with non-White ethnicity [@niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-The relative socioeconomic position of ethnic groups in Scotland is different to that in England, so it is plausible that the relation of health status to ethnicity may also differ.
-For example in the 2011 Scottish Census 1.6% of the population reported South Asian ethnicity.
-Among the 1.0% who identified as Pakistani or Bangladeshi the proportion living in the most deprived neighbourhoods was not higher than the national average [@walsh_increasingly_2019].
-Future work may allow more complete assignment of ethnicity and disaggregation of broad categories based on continent of origin.
+In Scotland however, although point estimates for rate ratios in those of South Asian and Black ethnicity compared to Whites are above 1, the confidence intervals are too wide for any definitive conclusions on ethnic differences.
 
@@
 Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions.
-In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system.
-We hope to explore this in a separate study that will include linkage to records of shielding advice.
-In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.
+\hl{ In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system.
+We hope to explore this in a separate study that will include linkage to records of shielding advice.} In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.
 
@@
 
-An important finding of this study was the strong association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
+A striking finding of this study was the association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
 Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.
@@
 Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.
-We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population [@mckeigue_evaluation_2020]
+\hl{We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population} [@mckeigue_evaluation_2020]
 
@@
 It will also exclude many people at high risk because they have multiple risk factors each of small effect.
-A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study.
+\hl{A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study}.
 Thus the rate ratio of `r rr <- round(as.numeric(gsub(" .+", "", table.agegr.all["Type 1 diabetes", 3])), 1); rr` associated with Type 1 diabetes equates to an increase of `r round(log(rr) / logistic.coeffs[3, 1], 1)` years in COVID age.
@@
 
-## Protocol registration
-The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+## Data availability
+The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public
 
-## Data availability
-All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
-Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
+## Funding
+
+No funding was received for this work.
 
@@
 2  NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
-
-## Supplementary material
-The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
 
@@
 
-ggplot(data=gam, aes(x=Age, y=value, colour=Sex, linetype=Status)) +
+p.incidence <- ggplot(data=gam, aes(x=Age, y=value, colour=Sex, linetype=Status)) +
     geom_line() +
@@
                        limits=car::logit(c(2E-5, 5E-3)))
+p.incidence
+
 ```
@@
 
-plotWdists(densities=full.densities, distlabels=c("Crude", "Adjusted")) +
+p.wdists <- plotWdists(densities=full.densities, distlabels=c("Crude", "Adjusted")) +
     theme_grey(base_size = 18) +
     theme(legend.title=element_blank())
+
+p.wdists
 
@@
 
-plotroc(full.densities) +
+p.roc <- plotroc(full.densities) +
     scale_color_manual(values=c("orange", "purple"), labels=c("Crude", "Adjusted")) + 
@@
 
+p.roc
+
+if(FALSE) {
+## Width: 789 – 2250 pixels (at 300 dpi).
+## Height maximum: 2625 pixels (at 300 dpi).
+## Resolution 300 – 600 dpi
+tiff("Figure1.tiff", res=300, width=1600, height=1200)
+p.incidence
+dev.off()
+tiff("Figure2.tiff", res=300, width=2100, height=1800)
+p.wdists
+dev.off()
+tiff("Figure3.tiff", res=300, width=2100, height=1800)
+p.roc
+dev.off()
+}
+
 ```
@@
 The receiver operator characteristic (ROC) curve is computed by calculating at each value of the risk score the sensitivity and specificity of a classifier that uses this value as the threshold for classifying cases and non-cases.
-Using the adjusted distributions from Figure \ref{wevid} gives a curve that is concave downwards.
+Using the adjusted distributions from Figure \ref{fig:wevid} gives a curve that is concave downwards.
 The C-statistic is the area under this curve, computed as the probability of correctly classifying a case/noncase pair using the risk score, evaluated over all possible such pairs in the dataset.
@@
                              paste0(rep("(", 4),
-                                    with(cc.all, table(severe.casegr)), rep(")", 4))), 
+                                    with(cc.severe, table(severe.casegr)), rep(")", 4))), 
              caption="Comparison of severe non-fatal and fatal cases, by test positive status and entry to critical care") %>%

### 2020-07-12 pharmaco.Rmd
updated pharmaco scripts

@@
 Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.
+Associations of prescribed drugs with severe COVID-19 were examined initially for all drug classes, and subsequently for a list of 16 drug classes postulated at the start of the epidemic as "medications compromising COVID": these included proton pump inhibitors, antipsychotics, and opioids.
 
 *Results* -- In those without a diagnosed risk condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.numdrugsgr.listed.any[">12", 4]`.
-Cardiovascular drugs did not contribute to this association: the largest independent contributors were proton pump inhibitors (PPI) and analgesics.
-For both PPI and opiate-containing analgesics, there were dose response relationships to severe COVID-19 that were strongest in those aged under 75 years: rate ratios were `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for $\ge 2$ DDDs/day of PPI and `r table.dosegr.opiate["0-74 years: >50", 4]` for $\ge 50$ MME opiates, The association was stronger with recent than with non-recent exposure.
+This association was specifically with non-cardiovascular drugs.
+Except for non-steroidal anti-inflammatory drugs, all 16 pre-specified drug classes were associated with increased risk of severe COVID-19.
+The largest effect was for antipsychotic agents: rate ratio `r table.laporte["402010.Antipsychotic drugs", 3]`.
+Drugs used in the treatment of nausea that are pharmacologically similar to antipsychotic agents were also associated with severe disease.
+Other drug classes with large effects included proton pump inhibitors (PPI), opioids and gabapentinoids.
+For both PPI and opioids, there were dose-response relationships to severe COVID-19: rate ratios in those aged under 75 years were `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for $\ge 2$ DDDs/day of PPI and `r table.dosegr.opioid["0-74 years: >50", 4]` for $\ge 50$ MME opioids.
+The association was stronger with recent than with non-recent exposure.
 
-*Conclusions* -- In Scotland severe COVID-19 is associated with polypharmacy and this is not easily explained by co-morbidity.
-Although the evidence for causality of proton pump inhibitors and opiates is not conclusive, these results reinforce existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.
+*Conclusions* -- Severe COVID-19 is associated with polypharmacy and with drugs that cause sedation, respiratory depression or dyskinesia, have anticholinergic effects or affect the gastrointestinal system.
+These associations are not easily explained by co-morbidity.
+Although the evidence for causality is not conclusive, these results reinforce existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.
 
@@
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units.
-Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.
+Death registrations were obtained from linkage to the National Register of Scotland.
 Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
@@
 For this analysis prescription codes from BNF chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".
+
+To define a prespecified hypothesis about drug classes postulated to increase the risk of severe COVID-19, we used a review by Laporte and Healy published on 2 April 2020 that listed several commonly used drug classes as associated with increased pneumonia risk and therefore of concern as "compromising COVID infections" [@laporte_midst_2020].
+Excluding immunosuppressive drugs that are criteria for shielding, the Laporte-Healy list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics and sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.
+
+For proton pump inhibitors and opioids, it was possible to use equivalent doses to calculate a total dose for all drugs in the class.
 Defined daily doses (DDDs) of each proton pump inhibitor were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
-Morphine milligram equivalents (MMEs) for opiates were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).
-Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).
-The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items.
+Morphine milligram equivalents (MMEs) for opioids were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).
+Average daily doses of proton pump inhibitors (as DDDs) and opioids (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by observation period.
+The dose of opioid was calculated as the sum over opioid-containing items in subparagraph 0407010 (nonopioid and compound preparations) and 0407020 (opioid analgesics).
 
@@
 Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
-To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure.
-Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models with care home status and number of drug classes dispensed as covariates, in which each drug class was dropped from the total number of drug classes.
-The model with the highest log-likelihood was used to select which drug class should be dropped.
-This procedure was repeated with the retained variables until only a single drug class remained.
+To distinguish between causality and confounding, we used several approaches:
+
+* Testing for consistency of association with drugs that have a similar mode of action across different indication groups
+
+* Testing for a dose-response relationship and stratifying by age group
+
+* Adjusting for prespecified covariates and restricting the analysis to those without any  listed condition.
+
+* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing.
 
@@
 
-## Associations with drug classes
-We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.
-Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 50 cases and controls and were associated with severe disease at $p < 0.001$.
-These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.
-A backward stepwise variable elimination procedure as described above was carried out to identify which of these `r length(subparacols.forstepwisedrop)` drug classes contributed most to the association of COVID-19 with polypharmacy.
-Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
-Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.
-Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.
-The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and "nonopioid analgesics and compound preparations" (BNF subparagraph code 0407010).
+## Associations with specific drug classes
+Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, in those without a listed condition and filtered to show only drug classes with at least 50 exposed individuals and p < 0.001.
+The drug classes associated include proton pump inhibitors, laxatives, a diverse range of drugs acting on the central nervous system, and nutritional supplements.
 
-The items grouped under subparagraph 0407010 include opiate-containing compound preparations and also aspirin-containing preparations with which proton pump inhibitors are commonly co-prescribed.
-To examine the associations with these drug classes in more depth, the total dispensed daily dose of opiate as MME, summed over subparagraphs 0407010 and 0407020 (opiate analgesics) was calculated.
-The total dispensed daily dose of proton pump inhibitors was calculated as DDDs.
-Distinguishing between causality and confounding is challenging because opiate-containing drugs and proton pump inhibitores are usually co-prescribed with other drugs.
-We used several approaches:
+Most of these drugs are usually co-prescribed with other drugs and are associated with polypharmacy.
+To study these associations further, we abandoned the hypothesis-free approac and resorted to examining the specific drug classes listed by Laporte and Healy.
+Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the Laporte-Healy list in those aged under 75 years.
+With the exception of non-steroidal anti-inflammatory drugs, all drug classes on this list were associated with increased risk in univariate analyses.
+All the drug classes listed by Laporte and Healy as having anticholinergic effects -- H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs -- were associated with increased risk.
 
-* Testing for a dose-response relationship and stratifying by age group
+In a multivariable regression, the drug classes with strongest evidence for independent association were proton pump inhibitors, antihistamines, antipsychotic drugs, and opioid analgesics.
+The highest rate ratio in both univariate and multivariate analyses was that associated with antipsychotic drugs.
 
-* Adjusting for prespecified covariates and restricting the analysis to those without any  listed condition.
+## Associations with antipsychotic agents and drugs used in the treatment of nausea
+As others have noted [@laporte_midst_2020], the chemical structures and modes of action of drugs used in the treatment of nausea and vertigo overlap with those of drugs used as antipsychotics.
+Table \ref{tab:antipsych} shows the univariate associations of severe COVID-19 with specific drugs classified in these two subparagraphs of the BNF.
+Across both groups of indications, phenothiazines and other drugs that are dopamine antagonists were strongly associated with increased rates of severe disease.
+The rate ratios associated with second-generation antipsychotics -- aripiprazole, olanzapine, quetiapine, and risperidone -- were as high as the rate ratios associated with phenothiazines.
 
-* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing.
+** reanalyse association with drugs after excluding those likely to be receiving palliative care. **
 
-## Associations with opiate use
-Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily opiate dose (MMEs) over the observation period.
+## Associations with opioids
+Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily opioid dose (MMEs) over the observation period.
 This showed a dose-response relationship, strongest in those aged less than 75 years.
-With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opiate["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opiate["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm.
-In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opiate.notlisted[">50", 4]` and `r table.dosegr.opiate.notlisted[">50", 6]` respectively.
-Table \ref{tab:timewindow} shows that the rate ratio associated with opiate dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)
+With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opioid["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opioid["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm.
+In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opioid.notlisted[">50", 4]` and `r table.dosegr.opioid.notlisted[">50", 6]` respectively.
+Table \ref{tab:timewindow} shows that the rate ratio associated with opioid dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)
 
-## Associations with proton pump inhibitor use
+## Associations with proton pump inhibitors
 Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for severe COVID-19 increases with the dose of proton pump inhibitor.
@@
 ## Associations with other drug classes
-Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2}, \ref{tab:bnfchapter4} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), 4 (nervous system) and 10 (musculoskeletal).
+Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), and 10 (musculoskeletal).
 These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
 
-Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, antimotility and laxatives were strongly and independently associated with disease in a multivariate analysis.
+Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, antimotility agents and laxatives were strongly and independently associated with disease in a multivariate analysis.
 Proton pump inhibitors were prescribed far more commonly than any other drug class in this chapter.
@@
 
-Table \ref{tab:bnfchapter4} shows that the strongest associations of COVID-19 with drugs for the nervous system were with antipsychotics, analgesics and drugs for dementia.
 Table \ref{tab:bnfchapter10} shows associations with drugs for the musculoskeletal system disaggregated by generic name, as the BNF groups all disease-modifying antirheumatic drugs under a single subparagraph code.
@@
 The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017].
-We attempted to investigate how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of specific drugs.
-A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.
-Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10, and that drug exposure is based on dispensed rather than issued prescriptions.
+Attempting to investigate associations with specific drugs with a hypothesis-free approach is difficult because many of the drug classes that are strongly associated with severe COVID-19, such as proton pump inhibitors, opioids and gabapentinoids, are indicators of overprescribing, recognized as such in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
 
-The drug classes that are strongly associated with severe COVID-19 include proton pump inhibitors and analgesics, widely recognized to be indicators of overprescribing as in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
-As these are among the most commonly prescribed drug classes, it is to be expected that they are selected by elimination as having the largest independent contributions to the association of COVID-19 with polypharmacy.
+For this reason we resorted to examining a pre-specified list of drugs postulated at the start of the epidemic to increase risk of severe COVID-19, based on previously described associations with pneumonia or activity on relevant pathways, especially anticholinergic agents [@laporte_midst_2020].
+We have shown that all the drugs originally listed are associated with increased risk, except for non-steroidal anti-inflammatory agents.
 Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.
-Because proton pump inhibitors are available over the counter, dispensed prescriptions do not capture all exposure.
 
-For both opiates and proton pump inhibitors, four criteria provide moderate but not conclusive evidence favouring causal explanations over confounding.
+The strongest association was with antipsychotic drugs.
+As the recognized indications for these drugs are not known to be associated with susceptibility to severe COVID-19, there is no obvious explanation for this association other than causality.
+Further evidence favouring causality is that drugs prescribed for nausea that have similar chemical structures and mode of action to antipsychotic drugs show similar associations with severe COVID-19.
+
+For both opioids and proton pump inhibitors, four criteria provide moderate but not conclusive evidence favouring causal explanations over confounding.
 There are strong dose-response relationships of COVID-19 to dispensed average daily dose.
@@
 Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
-It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates [@weiner_opioid_2017] has been reported previously.
+The dose-response relationship of opioid use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration from 2000-2012 [@edelman_association_2019].
+Although gabapentinoids are classified in the BNF under subparagraph 0408010 ("Control of epilepsy"), in Scotland they are widely used in combination with or as substitutes for opioid analgesics.
 
-The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration from 2000-2012 [@edelman_association_2019].
-As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase susceptibility to severe infection.
+The likely mechanisms of associations of these drugs with severe COVID-19 include sedation, respiratory depression, respiratory dyskinesia and anticholinergic effects.
+The evidence for these effects has been reviewed in detail elsewhere [@laporte_midst_2020] and we have little to add.
+We note that as SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors and other drugs acting on the gastrointestinal tract could increase susceptibility to severe infection.
+It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of drug classes such as proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opioids [@weiner_opioid_2017] has been reported previously.
+
+A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.
+Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10 (rather than the SNOMED CT codes used in primary care databases), and that drug exposure is based on dispensed rather than issued prescriptions.
 
@@
 This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.
-The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing, althoug proton pump inhibitors and analgesics have the greatest explanatory power for the association of COVID-19 with polypharmacy.
-Although this study did not identify unequivocal evidence of causality for opiates and proton pump inhibitors, effects of these drug classes on susceptibility to COVID-19 are biologically plausible.
+As a prediction of which drug classes would be associated with increased susceptibility to severe COVID-19, the Laporte-Healy list prepared at the start of the epidemic appears to be remarkably accurate.
+Most if not all of the drug classes on this list are widely overprescribed.
+The consistency of associations with drugs that have similar modes of action across different groups of indication, the dose-response and time window effects support causal explanations for at least some  these associations.
 We recommend that public health agencies should reinforce existing guidelines on the overprescribing of these drug classes, and more generally on the avoidance of unnecessary polypharmacy.
@@
 
-```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood of a conditional logistic regression of severe COVID-19 on care home and polypharmacy as covariates when that drug class is dropped, in  those without listed conditions."}
-
-stepwise.asfactor <- stepwise.drop.subparas
-stepwise.asfactor$name <- gsub("\\.", " ", stepwise.asfactor$name)
-stepwise.asfactor$numrow <- nrow(stepwise.asfactor):1
-stepwise.asfactor$name <- factor(stepwise.asfactor$name, levels=rev(stepwise.asfactor$name))
-
-p.loglik <- ggplot(data=stepwise.asfactor, 
-                   aes(x=loglik.full, y=name)) + geom_bar(stat="identity", fill="gray") +
-    xlab("Log-likelihood")  +
-    theme(axis.title.y=element_blank())
-#          axis.text.y=element_blank())
-
-p.coeff <- ggplot(data=stepwise.asfactor, 
-                  aes(x=beta.full, y=name)) + geom_bar(stat="identity", fill="gray") +
-    xlab("Coefficient (log rate ratio)")
-
-# gridExtra::grid.arrange(p.coeff, p.loglik, nrow=1, widths=c(0.5, 0.4))
-p.loglik
-
-if(FALSE) {
-knitr::kable(stepwise.drop.subparas[rows.print, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="stepwise.subparas",
-             row.names=FALSE,
-             col.names=c("Dropped drug class", "Log rate ratio before dropping",
-                         "Log-likelihood of model when dropped"), 
-             digits=c(0, 2, 1), 
-             caption=paste("Stepwise dropping of drug classes from the polypharmacy variable: dataset restricted to those not resident in care homes and without a listed condition")) %>%
-    column_spec(column=1, width="8cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-    kable_styling(latex_options="hold_position")
-}            
-```
-
 \newpage
@@
 
+```{r subpara.coeffs, echo=FALSE, warning=FALSE, message=FALSE}
+
+newch <- diff(as.integer(substr(subpara.coeffs$subpara, 1, 1))) !=0
+linesep <- ifelse(newch, '\\addlinespace', '')
+
+knitr::kable(subpara.coeffs,
+             escape=FALSE,
+             longtable=TRUE, 
+             booktabs=TRUE,
+             linesep = linesep, 
+             label="subpara.coeffs",
+             row.names=FALSE,
+col.names=c("BNF subparagraph", colnames(subpara.coeffs)[2:3],  
+            "Rate ratio", "$p$-value"),
+             caption=paste("Univariate rate ratios for severe COVID-19 in those without a listed condition associated with dispensing of each BNF subparagraph code, filtered to show only drug classes with at least 50 exposed individuals and  $p < 0.001$: ")) %>%
+    column_spec(column=1, width="10cm") %>%
+    column_spec(column=2, width="1.5cm") %>%
+    column_spec(column=3, width="1.5cm") %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+
+```
+
+```{r laporte, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.laporte,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="laporte",
+             col.names=c(clean.header(colnames(table.laporte)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value",
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 6), 
+             caption="Associations of severe COVID-19 in those aged under 75 years with drug classes listed by Laporte and Healy (2020)") %>%
+    column_spec(column=1, width="4cm") %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kable_styling(latex_options="hold_position") %>%
+    landscape()
+```
+
+```{r antipsych, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::kable(table.bnf.antipsych,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="antipsych",
+             col.names=c(clean.header(colnames(table.bnf.antipsych)[1:2]),
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             align=rep("c", 4), 
+             caption="Univariate associations of severe COVID-19 in those aged under 75 years with antipsychotic agents and drugs used in the treatment of nausea and vertigo") %>%
+    column_spec(column=1, width="4cm") %>%
+    kable_styling(latex_options="hold_position")
+```
+
 ```{r dosegr, echo=FALSE, warning=FALSE, message=FALSE}
 
-colnames(table.dosegr.protonpump) <- colnames(table.dosegr.opiate)
-table.dosegr <- rbind(table.dosegr.opiate, table.dosegr.protonpump)
+colnames(table.dosegr.protonpump) <- colnames(table.dosegr.opioid)
+table.dosegr <- rbind(table.dosegr.opioid, table.dosegr.protonpump)
 
@@
              row.names=FALSE, # dose levels in column 1
-             col.names=c("Average dose", colnames(table.dosegr.opiate)[2:3], 
+             col.names=c("Average dose", colnames(table.dosegr.opioid)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value",
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose of opiates and proton pump inhibitors, with and without adjustment for prespecified covariates") %>%
+             caption="Associations of severe COVID-19 with dispensed average daily dose of opioids and proton pump inhibitors, with and without adjustment for prespecified covariates") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
@@
     kable_styling(latex_options="hold_position") %>%
-    kableExtra::group_rows(group_label="Opiate-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
+    kableExtra::group_rows(group_label="Opioid-containing analgesics (mg morphine equivalent / day)", 1, 8) %>%
     kableExtra::group_rows(group_label="Proton pump inhibitors (defined daily doses / day)", 9, 18) %>%
@@
 
-colnames(table.timewindow.opiate) <- colnames(table.timewindow.protonpump)
-table.timewindow <- rbind(table.timewindow.opiate, table.timewindow.protonpump)
-table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opiate), 2),
+colnames(table.timewindow.opioid) <- colnames(table.timewindow.protonpump)
+table.timewindow <- rbind(table.timewindow.opioid, table.timewindow.protonpump)
+table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opioid), 2),
                                table.timewindow)
@@
     kable_styling(latex_options="hold_position") %>%
-    kableExtra::group_rows(group_label="Opiate-containing analgesics (restricted to those without history of neoplasm)", 1, 4) %>%
+    kableExtra::group_rows(group_label="Opioid-containing analgesics (restricted to those without history of neoplasm)", 1, 4) %>%
     kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
@@
 # Supplementary Tables
-
-```{r subpara.coeffs, echo=FALSE, warning=FALSE, message=FALSE}
-
-newch <- diff(as.integer(substr(subpara.coeffs$subpara, 1, 1))) !=0
-linesep <- ifelse(newch, '\\addlinespace', '')
-
-knitr::kable(subpara.coeffs,
-             escape=FALSE,
-             longtable=TRUE, 
-             booktabs=TRUE,
-             linesep = linesep, 
-             label="subpara.coeffs",
-             row.names=FALSE,
-col.names=c("BNF subparagraph", colnames(subpara.coeffs)[2:3],  
-            "Rate ratio", "$p$-value"),
-             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code in those without a listed condition, filtered to show only drug classes with at least 50 exposed individuals and  $p < 0.001$: ")) %>%
-    column_spec(column=1, width="10cm") %>%
-    column_spec(column=2, width="1.5cm") %>%
-    column_spec(column=3, width="1.5cm") %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-
-```
 
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 1 by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 in those aged under 75 years, with prescribed drugs in BNF chapter 1 by subparagraph code filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 in those aged under 75 years with prescribed drugs in BNF chapter 2 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 4  by subparagraph code in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 in those aged under 75 years with prescribed drugs in BNF chapter 4 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name in those aged under 75 years, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19  in those aged under 75 years with prescribed drugs in BNF chapter 10 by approved name, filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="5cm") %>%

### 2020-07-12 casecontrol.R

@@ -42,7 +42,6 @@ library(rmarkdown)
 library(pander)
 library(ggplot2)
 library(doParallel)
-# library(reshape2)
 library(readxl)
 library(DescTools)
 library(icd.data)
@@ -78,7 +77,6 @@ recode.dmtype <- function(x) {
     return(r)
 }
 
-
 ##################################################################
      
 if(old) {
@@ -108,7 +106,7 @@ if(old) {
     sicsag <- as.data.table(readRDS("./data/CC_SICSAG_ANON_2020-06-18.rds"), key="ANON_ID")
     scrips.filename <- "./data/scrips.last240days.rds"
     scrips.firsttime <- FALSE
-    if(scrips.firsttime) {
+    if(scrips.firsttime) { 
         scrips <- readRDS( "./data/CC_PIS_x15_ANON_2020-06-18.rds")
         scrips[, daysbefore := as.integer(SPECIMENDATE - dispensed_date)]
         ## restrict to last 240 days before cutoff of (specimendate - 15) 
@@ -116,9 +114,14 @@ if(old) {
         ## save this file so that other reads from it will have this restriction applied 
         saveRDS(scrips, scrips.filename)
     }
-    scrips <- as.data.table(readRDS(scrips.filename)[, c("ANON_ID", "daysbefore",
-                                           "bnf_paragraph_code",
-                                           "bnf_paragraph_description")], key="ANON_ID")
+    scrips <-
+        as.data.table(readRDS(scrips.filename)[, c("ANON_ID",
+                                                   "dispensed_date",
+                                                   "approved_name",
+                                                   "bnf_item_code", 
+                                                   "bnf_paragraph_code",
+                                                   "bnf_paragraph_description",
+                                                   "daysbefore")], key="ANON_ID")
 }
 
 ################### short names for ICD chapters ########################
@@ -300,7 +303,6 @@ with(cc.all[CASE==1], table(icu, hdu, exclude=NULL))
 ## only 854 in critical care -- must be wrong
 cc.all$criticalcare <- cc.all$icu==1 | cc.all$hdu==1
 
[…]

### 2020-07-12 clogit_hsstan.R

@@ -1,0 +1,74 @@
+## train model to predict fatal disease
+
+library(hsstan)
+library(data.table)
+library(parallel)
+
+cc.severe <- as.data.table(readRDS("data/cc.severe.rds"))
+                           
+## restrict to fatal cases + matched controls, and complete data on covariates
+select.cols <- c("CASE", "stratum", "care.home", "diabetes.any",
+                 grep("^Ch\\.", colnames(cc.severe), value=TRUE))
+covariates <- select.cols[-(1:2)][1:8]
+
+cc.nonmissing <- na.omit(cc.severe[fatal.casegroup==1], cols=select.cols)
+cc.nonmissing <- cc.nonmissing[, ..select.cols]
+covariate.names <- select.cols[-(1:2)]
+## convert all columns to numeric
+cc.nonmissing <- cc.nonmissing[, lapply(.SD, as.numeric), by=row.names(cc.nonmissing)][, -1]
+cc.nonmissing[, (covariate.names) := lapply(.SD, scale), .SDcols=covariate.names] 
+str(cc.nonmissing)
+ 
+## drop strata with num cases not equal to 1, or < 2 observations 
+cc.drop <- cc.nonmissing[CASE==1, .(.N), by = .(stratum)][N != 1, ]
+cc.nonmissing <- cc.nonmissing[!(stratum %in% cc.drop$stratum)]
+cc.drop <- cc.nonmissing[, .(.N), by = .(stratum)][N < 2, ]
+cc.nonmissing <- cc.nonmissing[!(stratum %in% cc.drop$stratum)]
+table(cc.nonmissing[, .(.N), by = .(stratum)][, N])
+## renumber levels of stratum   
+cc.nonmissing[, stratum := as.integer(as.factor(as.integer(stratum)))]
+setkey(cc.nonmissing, stratum)
+cc.nonmissing <- as.data.frame(cc.nonmissing)
+
+covs.model <- as.formula(paste("CASE ~", paste(covariates, collapse="+")))  
+
+penalized <- covariates[-1:2]
+iter <- 1000
+warmup <- 500
+regularized <- TRUE
+nu <- 1
+
+############################################################################
+
+hs.clogit <- hsstan(x=cc.nonmissing, covs.model=covs.model,
+                   penalized=penalized, family="cox",
+                   iter=iter, warmup=warmup,
+                   scale.u=2, regularized=TRUE, nu=ifelse(regularized, 1, 3),
+                   par.ratio=0.05, global.df=1, slab.scale=2, slab.df=4,
[…]

### 2020-07-12 drugs.R

@@ -32,8 +32,6 @@ getdoseTWday.wide <- function(scrips.in, varname.prefix="dose.") {
     return(doseTWday.wide)
 }
 
-## reading scrips
-
 scripvars <- c("ANON_ID", "dispensed_date", "bnf_paragraph_code",
                "formulation_code",
                "item_strength", "item_strength_uom", "item_code",
@@ -71,15 +69,16 @@ length(table(substr(scrips$bnf_paragraph_code, 1, 6)))  # chapter, section, para
 length(table(scrips$bnf_paragraph_code)) # 537 groups
 
 ## we need integer variables chapter, sectioncode, paracode for use with data.table::dcast
-scrips$chapternum <- as.integer(substr(scrips$bnf_paragraph_code, 1, 2))
-scrips$sectioncode <- as.integer(substr(scrips$bnf_paragraph_code, 1, 4))
-scrips$paracode <- as.integer(substr(scrips$bnf_paragraph_code, 1, 6))
+scrips[, chapternum := as.integer(substr(bnf_paragraph_code, 1, 2))]
+scrips[, sectioncode := as.integer(substr(bnf_paragraph_code, 1, 4))]
+scrips[, paracode := as.integer(substr(bnf_paragraph_code, 1, 6))]
 
 ## recode scrips$bnf.chapter values > 14 or NA to 14
-scrips$chapternum[is.na(scrips$chapternum)] <- 14
-scrips$chapternum[scrips$chapternum > 14] <- 14
+scrips[is.na(chapternum), chapternum := 14]
+scrips[chapternum > 14, chapternum := 14] 
 
 if(!old) {
+    
 ## separate tables for prescriptions of proton pump, opioids, nonopioids
 
 scrips.protonpump <-
@@ -106,7 +105,6 @@ scrips.protonpump <- merge(scrips.protonpump, DDD.protonpump, by="approved_name"
 scrips.opioid <-
     readRDS(scrips.filename) %>%
     subset(bnf_paragraph_code=="0407020")
-
     
 scrips.opioid$mode <- car::recode(scrips.opioid$formulation_code,
                                   "c('CAPS', 'LOZ', 'SACH', 'SOLN', 'TABS')='Oral';
@@ -200,8 +198,8 @@ scrips.opioid[scrips.opioid$approved_name=="PARACETAMOL WITH TRAMADOL HYDROCHLOR
 ## transdermal fentanyl: assume patch strength 50 microgram/h changed every 3 days, frequency not given, equivalent to 180 mg/day morphine
[…]

### 2020-07-12 helperfunctions.R

@@ -339,13 +339,9 @@ tabulate.bnfparas <- function(chnum, outcome="CASE", data=cc.severe, minrowsum=2
     scrips.para.wide <- as.data.table(scrips.para.wide, key="ANON_ID")
     cc.bnf.para <- scrips.para.wide[data[, c("ANON_ID", "CASE", "stratum")]]
     ## now fix colnames and set missing to 0
-    bnfcols <- grep("^para.", colnames(cc.bnf.para))
-    for(j in bnfcols) {
-        cc.bnf.para[[j]][is.na(cc.bnf.para[[j]])] <- 0
-        cc.bnf.para[[j]][cc.bnf.para[[j]] > 1] <- 1
-        cc.bnf.para[[j]] <- as.factor(cc.bnf.para[[j]])
-    }
-    
+    bnfcols <- as.integer(grep("^para.", colnames(cc.bnf.para)))
+    cc.bnf.para[, (bnfcols) := lapply(.SD, recode.indicator), .SDcols = bnfcols]
+
     bnf.para <- colnames(cc.bnf.para)[bnfcols]
     
     table.bnf.para <- univariate.tabulate(varnames=bnf.para, outcome="CASE", data=cc.bnf.para,
@@ -386,7 +382,6 @@ merge.bnfsubparas <- function(chnums, data) {
     colnames(scrips.subpara.wide)[-1] <- paste("subpara",
                                           as.integer(colnames(scrips.subpara.wide)[-1]),
                                           names.subparas, sep=".")
-    
     ## drop rare subparagraphcodes
     scrips.subpara.wide <- subset(scrips.subpara.wide, select=colSums(scrips.subpara.wide) > 20)
 
@@ -396,11 +391,6 @@ merge.bnfsubparas <- function(chnums, data) {
     bnfcols <- as.integer(grep("^subpara.", colnames(cc.bnf.subpara)))
     ## recode indicator variables
     cc.bnf.subpara[, (bnfcols) := lapply(.SD, recode.indicator), .SDcols = bnfcols]
-#   for(j in bnfcols) {
-#        cc.bnf.subpara[[j]][is.na(cc.bnf.subpara[[j]])] <- 0
-#        cc.bnf.subpara[[j]][cc.bnf.subpara[[j]] > 1] <- 1
-#        cc.bnf.subpara[[j]] <- as.factor(cc.bnf.subpara[[j]])
-#    }
     return(cc.bnf.subpara)
 }
 
[…]

### 2020-07-12 pharmaco.R

@@ -88,32 +88,6 @@ print(subpara.coeffs)
 
 #############################################################################
 
-colnames.drugclasses.forstepwise <- colnames(cc.severe)[subparacols.keep]
-
-table.subpara.coeffs <- tabulate.freqs.regressions(colnames.drugclasses.forstepwise,
-                                                   data=subset(cc.severe, restrict))
-
-## stepwise drop procedure over subparacols.keep
-
-## replace original subparas with nonopiate and anyopiate (scrips for compound preparations assigned to opiate)
-#colnames.drugclasses.forstepwise <-
-#    grep("\\.407020\\.", colnames.drugclasses.forstepwise, invert=TRUE, value=TRUE)
-
-subparacols.forstepwisedrop <- match(colnames.drugclasses.forstepwise,
-                                     colnames(cc.severe)) ## reduces to 61 columns
-
-x <- subset(cc.severe, subset=restrict, select=subparacols.forstepwisedrop)
-x <- matrix(as.integer(as.matrix(x)), nrow=nrow(x))
-colnames(x) <- colnames(cc.severe)[subparacols.forstepwisedrop]
-y <- subset(cc.severe, restrict)[["CASE"]]
-stratum <- subset(cc.severe, restrict)[["stratum"]]
-covariates.subparas <- subset(cc.severe, restrict)[["care.home"]] 
-cat("Stepwise drop procedure over BNF subpara codes adjusted for care home status...")
-stepwise.drop.subparas <- stepwise.union.dropcols(x=x, y=y, covariates=covariates.subparas, stratum=stratum)
-cat("done\n")
-
-print(stepwise.drop.subparas)
-
 ################################################################
 
 ## tabulate dose-response effects by age group
@@ -127,7 +101,7 @@ coeff.rows <- 1:3
 withcovariates.formula <- as.formula(paste("CASE ~ dosegr.opiate +",
                                            opiate.covariates.string, "+ strata(stratum)"))
 
-table.dosegr.opiate <- NULL
+table.dosegr.opioid <- NULL
 for(agegr in levels(cc.severe$agegr2)) {
     x <- summary(clogit(formula=CASE ~ dosegr.opiate + strata(stratum), 
[…]

### 2020-07-12 test_clogit.R

@@ -1,0 +1,48 @@
+library("extraDistr")
+
+data(diabetes, package="hsstan")
+
+map.varnames <- grep("map", names(diabetes), value=TRUE)
+P <- length(map.varnames)
+N <- nrow(diabetes)
+beta <- matrix(rnorm(P, sd=0.5), ncol=1)
+Xbeta <- as.matrix(diabetes[, map.varnames]) %*% beta
+S <- ceiling(N / 5)
+diabetes$stratum <- rep(1:S, each=5)[1:N]
+
+seed <- 12345
+
+diabetes$y.clogit <- numeric(N)
+for(s in 1:S) {
+    xbeta <- Xbeta[diabetes$stratum==s, ]
+    ycat <- extraDistr::rcat(n=1, prob=exp(xbeta) / sum(exp(xbeta)) )
+
+    ## for s = 1 this will assign y=1 to element ycat
+    diabetes$y.clogit[match(s, diabetes$stratum) - 1 + ycat] <- 1
+}
+
+penalized <- map.varnames[-(1:2)]
+regularized <- TRUE
+
+covs.model <- as.formula(paste("y.clogit ~", paste(map.varnames, collapse="+")))  
+
+hs.clogit <- hsstan(x=diabetes, covs.model=covs.model,
+                   penalized=penalized, family="cox",
+                   iter=1000, warmup=500,
+                   scale.u=2, regularized=TRUE, nu=ifelse(regularized, 1, 3),
+                   par.ratio=0.05, global.df=1, slab.scale=2, slab.df=4,
+                   qr=TRUE, seed=123, adapt.delta=0.9,
+                   keep.hs.pars=FALSE)
+
+print(hs.clogit)
+
+sampler.stats(hs.clogit)
+
+invlogit <- function(x) 1/(1+exp(-x))
+
+hs.clogit$family <- list(family="cox", linkinv=invlogit)
+
+sel.vars <- projsel(hs.clogit, start.from="map.2")
+
+print(sel.vars, digits=4)
+

### 2020-07-28 casecontrol.Rmd
updated pharmacoepi manuscript

@@
 As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.
-For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths registered up to `r format(max(cc.severe$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECIMENDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths registered up to `r format(max(cc.severe$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
 Among fatal cases, `r percent.withindays[6]`% of deaths were registered within 5 days.

### 2020-07-28 pharmaco.Rmd
updated pharmacoepi manuscript

@@
     keep_tex: true
+    md_extensions: -autolink_bare_uris
+
 csl: ../csl/styles/the-new-england-journal-of-medicine.csl
@@
 
-# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n")
+# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n"); system("make4ht -u pharmaco.tex");
 
-# abstract limit 250 words
+# abstract limit 350 words
 
@@
 
-On behalf of Public Health Scotland COVID-19 Health Protection Study Group
+On behalf of Public Health Scotland COVID-19 Health Protection Study Group.
 
@@
 
-stepwise.drop.subparas[, 1] <- gsub("subpara\\.", "", stepwise.drop.subparas[, 1])
-
-firstrow.neg <- which(stepwise.drop.subparas[, 3] < 0)[1]
-rows.print <- firstrow.neg:nrow(stepwise.drop.subparas)
-
 ```
@@
 
-_Background_ -- Dispensing of at least one prescription in the past year is a risk factor for severe COVID-19 in Scotland, even in those without diagnoses of conditions designated as conferring increased risk.
+*Objectives* -- To investigate the relation of severe COVID-19 to prior drug prescribing.
 
-_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic.
-Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.
-Associations of prescribed drugs with severe COVID-19 were examined initially for all drug classes, and subsequently for a list of 16 drug classes postulated at the start of the epidemic as "medications compromising COVID": these included proton pump inhibitors, antipsychotics, and opioids.
+*Design* -- Matched case-control study (REACT-SCOT) based on record linkage to hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.
 
-*Results* -- In those without a diagnosed risk condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.numdrugsgr.listed.any[">12", 4]`.
-This association was specifically with non-cardiovascular drugs.
-Except for non-steroidal anti-inflammatory drugs, all 16 pre-specified drug classes were associated with increased risk of severe COVID-19.
+*Setting* -- Scottish population.
+
+*Main outcome measure* -- Severe COVID-19, defined by entry to critical care or fatal outcome.
+
+*Participants* -- All `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic, with `r nrow(cc.severe[CASE==0])` controls matched for age, sex and primary care practice.
+
+*Results* -- Severe COVID-19 was strongly associated with the number of non-cardiovascular drug classes dispensed.
+This association was strongest in those not resident in care homes, in whom the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.notcv.numdrugsgr.care.home[">12", 4]`, and was not accounted for by treatment of conditions designated as conferring increased risk.
+Of 17 drug classes postulated at the start of the epidemic to be "medications compromising COVID", all were associated with increased risk of severe COVID-19.
 The largest effect was for antipsychotic agents: rate ratio `r table.laporte["402010.Antipsychotic drugs", 3]`.
-Drugs used in the treatment of nausea that are pharmacologically similar to antipsychotic agents were also associated with severe disease.
-Other drug classes with large effects included proton pump inhibitors (PPI), opioids and gabapentinoids.
-For both PPI and opioids, there were dose-response relationships to severe COVID-19: rate ratios in those aged under 75 years were `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for $\ge 2$ DDDs/day of PPI and `r table.dosegr.opioid["0-74 years: >50", 4]` for $\ge 50$ MME opioids.
-The association was stronger with recent than with non-recent exposure.
+Other drug classes with large effects included proton pump inhibitors (rate rato `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for >= 2 defined daily doses/day), opioids (`r table.dosegr.opioid["0-74 years: >50", 4]` for >= 50 mg morphine equivalent/day) and gabapentinoids.
+These associations persisted after adjusting for covariates, and were stronger with recent than with non-recent exposure.
 
@@
 These associations are not easily explained by co-morbidity.
-Although the evidence for causality is not conclusive, these results reinforce existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.
+Although the evidence for causality is not conclusive, these results support existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy as a potential means of reducing COVID-19 risk.
+
+*Registration* -- ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)
 
@@
 
+***
+*What is already known on this topic* -- Two previous studies have examined the relationship of severe COVID-19 to drugs for the cardiovascular system.
+This is the first systematic study of the relationship of severe COVID-19 to prior drug prescribing.
+
+*What this study adds* -- Severe COVID-19 is associated with polypharmacy and with drugs that cause sedation, respiratory depression or dyskinesia, have anticholinergic effects or affect the gastrointestinal system.
+These associations are not easily explained by co-morbidity.
+These results support earlier warnings that these drug classes that these drugs might increase susceptibility to COVID-19, and reinforce existing guidance on reducing overprescribing of these drug classes.
+
+***
+
 # Background
-In the initial analysis of REACT-SCOT, a matched case control study of severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
+In the initial analysis of REACT-SCOT, a matched case control study of  risk factors for severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
 The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.
@@
 The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].
-In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
+All record linkage studies using NHS data in Scotland, including this one, are governed by the Public Benefit and Privacy Panel for Health and Social Care, which includes patient and public representatives.
+All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
 Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units.
@@
 For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
-For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECIMENDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.
 
@@
 
-To define a prespecified hypothesis about drug classes postulated to increase the risk of severe COVID-19, we used a review by Laporte and Healy published on 2 April 2020 that listed several commonly used drug classes as associated with increased pneumonia risk and therefore of concern as "compromising COVID infections" [@laporte_midst_2020].
-Excluding immunosuppressive drugs that are criteria for shielding, the Laporte-Healy list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics and sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.
+We began by testing for association of severe COVID-19 with the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation.
+In accordance with earlier suggestions that prescribing of multiple cardiovascular drugs, which is supported by evidence-based guidelines, should be considered separately from putatively inappropriate polypharmacy [@appleton_cardiovascular_2014], we partitioned the number of drug classes dispensed into cardiovascular and other drugs.
+To define a prespecified hypothesis about drug classes postulated to increase the risk of severe COVID-19, we used a review by Laporte and Healy published on 2 April 2020 that listed 17 widely prescribed drug classes associated with increased pneumonia risk and therefore of concern as "medications compromising COVID" [@laporte_midst_2020].
+Excluding immunosuppressive drugs that are criteria for shielding, the Laporte-Healy list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics and sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, anti-epileptic drugs, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.
 
@@
 Defined daily doses (DDDs) of each proton pump inhibitor were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.
-Morphine milligram equivalents (MMEs) for opioids were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).
+Morphine milligram equivalents (MMEs) for opioids were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthestists.
 Average daily doses of proton pump inhibitors (as DDDs) and opioids (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by observation period.
@@
 
-* Adjusting for prespecified covariates and restricting the analysis to those without any  listed condition.
+* Adjusting for prespecified covariates.
 
@@
 ## Relation of severe COVID-19 to polypharmacy 
-We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive).
-Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without any listed condition.
-In those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions.
-To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.
-As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.
-In a model with two covariates -- number of cardiovascular drug classes and number of other drug classes -- the rate ratios per drug class for severe COVID-19 were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.
+Figure \ref{fig:notcv.numdrugs} shows that the rate of of severe COVID-19 increased steeply with the number of non-cardiovascular drug classes and decreased with the number of noncardiovascular drug classes dispensed.
+Table \ref{tab:notcv.numdrugs.care.home} shows that this association was restricted to those not resident in care homes.
+Supplementary Table \ref{tab:notcv.numdrugs.listed.any} shows that among those not resident in care homes, the association of severed COVID-19 with number of non-cardiovascular drug classes dispensed was present both in those without any diagnosed listed condition and in those with at least one diagnosed listed condition.
 
 ## Associations with specific drug classes
-Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, in those without a listed condition and filtered to show only drug classes with at least 50 exposed individuals and p < 0.001.
-The drug classes associated include proton pump inhibitors, laxatives, a diverse range of drugs acting on the central nervous system, and nutritional supplements.
+In order to identify specific drug classes associated with severe COVID-19, we examined thse associations among those not resident in a care home and without any listed condition, as most people diagnosed with listed conditions such as asthma and diabetes are on medication for these conditions.
+Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, filtered to show only drug classes with at least 50 exposed individuals and $p < 0.001$.
+The drug classes associated with severe COVID-19 include proton pump inhibitors, laxatives, multiple classes of drugs acting on the central nervous system, nutritional supplements, and non-steroidal anti-inflammatory drugs.
 
-Most of these drugs are usually co-prescribed with other drugs and are associated with polypharmacy.
-To study these associations further, we abandoned the hypothesis-free approac and resorted to examining the specific drug classes listed by Laporte and Healy.
-Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the Laporte-Healy list in those aged under 75 years.
-With the exception of non-steroidal anti-inflammatory drugs, all drug classes on this list were associated with increased risk in univariate analyses.
-All the drug classes listed by Laporte and Healy as having anticholinergic effects -- H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs -- were associated with increased risk.
+Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the Laporte-Healy list in those not resident in a care home.
+All drug classes on this list were associated with increased risk in univariate analyses (though the association with NSAIDS was stronger in Table \ref{tab:subpara.coeffs} where listed conditions were excluded.
+This included all the drug classes listed by Laporte and Healy as having anticholinergic effects: H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs.
 
-In a multivariable regression, the drug classes with strongest evidence for independent association were proton pump inhibitors, antihistamines, antipsychotic drugs, and opioid analgesics.
-The highest rate ratio in both univariate and multivariate analyses was that associated with antipsychotic drugs.
+In a multivariable regression, the strongest independent associations were with proton pump inhibitors, antihistamines, antipsychotic drugs, and opioid analgesics.
+In both univariate and multivariable analyses, the highest rate ratio was that associated with antipsychotic drugs.
+As others have noted [@laporte_midst_2020], the chemical structures and modes of action of drugs used in the treatment of nausea and vertigo overlap with those of antipsychotic drugs.
+Supplementary Table \ref{tab:antipsych} shows the univariate associations of severe COVID-19 with specific drugs classified in these two subparagraphs of the BNF.
+Across both groups of indications, phenothiazines and other drugs that are dopamine antagonists were strongly associated with increased rates of severe disease.
+Rate ratios were elevated both for phenothiazines and for second-generation antipsychotics: aripiprazole, olanzapine, risperidone and amrisulpride.
 
-## Associations with antipsychotic agents and drugs used in the treatment of nausea
-As others have noted [@laporte_midst_2020], the chemical structures and modes of action of drugs used in the treatment of nausea and vertigo overlap with those of drugs used as antipsychotics.
-Table \ref{tab:antipsych} shows the univariate associations of severe COVID-19 with specific drugs classified in these two subparagraphs of the BNF.
-Across both groups of indications, phenothiazines and other drugs that are dopamine antagonists were strongly associated with increased rates of severe disease.
-The rate ratios associated with second-generation antipsychotics -- aripiprazole, olanzapine, quetiapine, and risperidone -- were as high as the rate ratios associated with phenothiazines.
+## Time window analyses
+Supplementary Table \ref{tab:timewindow} shows for each drug class on the Laporte-Healy list the rate ratio associated with dispensing in the most recent 120-day time window, with dispensing only in the previous time window as reference category.
+Because most users of these drugs had dispensed prescriptions in both time windows, these analyses are based on relatively small numbers: the drug classes shown are restricted to those with at least 500 cases and controls exposed only in the earlier time window .
+For most of these drug classes shown, the rate ratio associated with recent exposure only is above 1, but only for opioid analgesics, and to some extent proton pump inhibitors and anxiolytics, does this association reach conventional levels of statistical significance.
 
-** reanalyse association with drugs after excluding those likely to be receiving palliative care. **
-
-## Associations with opioids
-Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily opioid dose (MMEs) over the observation period.
-This showed a dose-response relationship, strongest in those aged less than 75 years.
-With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opioid["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opioid["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm.
-In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opioid.notlisted[">50", 4]` and `r table.dosegr.opioid.notlisted[">50", 6]` respectively.
-Table \ref{tab:timewindow} shows that the rate ratio associated with opioid dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)
-
-## Associations with proton pump inhibitors
-Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for severe COVID-19 increases with the dose of proton pump inhibitor.
-The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
-This dose-response relationship was absent in those aged 75 years and over.
-Among those in this age group without any listed condition, the unadjusted and adjusted rate ratios associated with more than 1.5 DDDs/day were `r table.dosegr.protonpump.notlisted.agelt75["1.5 or more", 4]` and `r table.dosegr.protonpump.notlisted.agelt75["1.5 or more", 6]` respectively.
-Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.
+## Dose-response analyses
+Supplementary Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily doses of opioids (as MME) and proton pump inhibitors (as DDDs) over the 240-day observation period.
+For both these drug classes there were dose-response relationships, and for proton pump inhibitors this relationship was strongest in those aged less than 75 years.
+With unexposed as baseline, the univariate rate ratio associated with opioid use in this age group was `r table.dosegr.opioid["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opioid["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm.
+For proton pump inhibitors, the univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
 
 ## Associations with other drug classes
-Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), and 10 (musculoskeletal).
-These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
-
-Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, antimotility agents and laxatives were strongly and independently associated with disease in a multivariate analysis.
-Proton pump inhibitors were prescribed far more commonly than any other drug class in this chapter.
+Supplementary Tables \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 2 (cardiovascular) and 10 (musculoskeletal).
+These chapters were selected as of interest because because specific hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.
 
@@
 # Discussion 
+<!---
+• Statement of principal findings
+• Strengths and weaknesses of the study
+• Strengths and weaknesses in relation to other studies, discussing important differences in results
+• Meaning of the study: possible explanations and implications for clinicians and policymakers
+• Unanswered questions and future research
+--->
 We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation, in individuals without conditions designated as conferring high risk.
@@
 
-For this reason we resorted to examining a pre-specified list of drugs postulated at the start of the epidemic to increase risk of severe COVID-19, based on previously described associations with pneumonia or activity on relevant pathways, especially anticholinergic agents [@laporte_midst_2020].
-We have shown that all the drugs originally listed are associated with increased risk, except for non-steroidal anti-inflammatory agents.
+To narrow the hypothesis space we tested a pre-specified list of drugs postulated at the start of the epidemic to increase risk of severe COVID-19, based on previously described associations with pneumonia or activity on relevant pathways, especially anticholinergic agents [@laporte_midst_2020].
+We have shown that all the drugs originally listed are associated with increased risk.
 Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.
@@
 
-The likely mechanisms of associations of these drugs with severe COVID-19 include sedation, respiratory depression, respiratory dyskinesia and anticholinergic effects.
-The evidence for these effects has been reviewed in detail elsewhere [@laporte_midst_2020] and we have little to add.
+A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.
+Another limitation is that it is not possible to capture hospital prescribing data which includes biologic agents that have immunosuppressive effects.
+Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10 (rather than the SNOMED-CT codes used in primary care databases), and that drug exposure is based on dispensed rather than issued prescriptions.
+
+The mechanisms postulated by Laporte and Healy for drugs to increase risk of severe COVID-19 include sedation, respiratory depression, respiratory dyskinesia and anticholinergic effects  [@laporte_midst_2020].
 We note that as SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors and other drugs acting on the gastrointestinal tract could increase susceptibility to severe infection.
@@
 
-A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.
-Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10 (rather than the SNOMED CT codes used in primary care databases), and that drug exposure is based on dispensed rather than issued prescriptions.
-
 We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing.
-For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].
+Of those drug classes that are not on the Laporte-Healy list, there are strong univariate associations of severe COVID-10 with dispensing of antibiotics, laxatives and nutritional supplements.
+The associations with nutritional supplements is likely to be confounded by overprescribing.
+For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020; @de_abajo_use_2020] and indeed suggest some protective effect may be possible.
+To explore this more fully will require access to other datasets where measurements of blood pressure and other covariates are available.
 The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
-Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.
-The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.
+Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.
+Associations with disease-modifying antirheumatic drugs such as hydroxychloroquine are likely to be confounded by hospital-based prescribing of biologic anti-rheumatic drugs, which is not captured by record linkage in Scotland.
 
@@
 As a prediction of which drug classes would be associated with increased susceptibility to severe COVID-19, the Laporte-Healy list prepared at the start of the epidemic appears to be remarkably accurate.
-Most if not all of the drug classes on this list are widely overprescribed.
+Many of the drug classes on this list are recognized indicators of overprescribing.
 The consistency of associations with drugs that have similar modes of action across different groups of indication, the dose-response and time window effects support causal explanations for at least some these associations.
-We recommend that public health agencies should reinforce existing guidelines on the overprescribing of these drug classes, and more generally on the avoidance of unnecessary polypharmacy.
+We recommend that public health agencies should reinforce existing guidelines on avoidance of overprescribing of these drug classes and more generally on inappropriate polypharmacy.
 
 # Declarations
-## Information governance
+## Ethics approval and information governance
 This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.
 Datasets were de-identified before analysis.
-The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
-All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
+Dissemination of results will be managed by Public Health Scotland.
+
+## Data sharing
 Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
 
+## Role of the funding source
+This study was undertaken without any external funding.
+
+## Transparency declaration
+PM, as the manuscript's guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
+This manuscript has been generated directly from the source data by a reproducible research pipeline.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
+
+## Registration
+The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+
 ## Conflicts of interest
-HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
+HC receives research support and honoraria and is a member of advisory panels or speaker bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
 HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.
-HC and PM own stock in Roche and Bayer.
 SH received honoraria from Gilead.
@@
 ## Supplementary material
-The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available with this manuscript.
+The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available when the paper is published.
 
@@
 
-\newpage
+```{r plot.numdrugsgr, fig.pos="H", fig.cap="\\label{fig:notcv.numdrugs}Rate ratios (with standard errors) in a conditional logistic regression of severe COVID-19 on number of cardiovascular (BNF chapter 4) and non-cardiovascular drug classes dispensed", out.width="0.8\\textwidth", echo=FALSE, warning=FALSE, message=FALSE}
+
+p.numdrugs <- ggplot(aes(x=numdrugsgr, y=coeff, shape=Indication, colour=Indication), data=coeffs.data) +
+    geom_errorbar(aes(ymin=coeff-se, ymax=coeff+se), width=.1) +
+    geom_point(size=4) +
+    geom_line() +
+    scale_x_continuous(name="Number of drug classes dispensed", labels=levels(cc.severe$notcv.numdrugsgr)) + 
+    scale_y_continuous(name="Rate ratio for severe COVID-19",
+		minor_breaks=log(c(0.6, 0.8, 1, 1.5, 2, 3, 4, 5, 6, 7, 8, 9, 10, 12)), 
+  	breaks=log(c(0.6, 0.8, 1:5, 10)), 
+labels=as.character(c(0.6, 0.8, 1:5, 10))) + 
+theme(legend.position=c(0.75, 0.4)) + 
+    theme(text=element_text(size=20))
+
+p.numdrugs
+
+if(FALSE) {
+tiff("Figure1.tiff", res=300, width=1600, height=1200)
+p.numdrugs
+dev.off()
+}
+
+```
 
@@
 
-```{r numdrugs.listed.any, echo=FALSE, warning=FALSE, message=FALSE}
+```{r numdrugs.care.home, echo=FALSE, warning=FALSE, message=FALSE}
 
-freqs <- with(cc.severe, table(listed.any, CASE))
+freqs <- with(cc.severe, table(care.home, CASE))
 
-knitr::kable(table.numdrugsgr.listed.any,
+knitr::kable(table.notcv.numdrugsgr.care.home,
              escape=FALSE,
              booktabs=TRUE,
-             label="numdrugs.listed.any",
+             label="notcv.numdrugs.care.home",
              row.names=FALSE,
-             col.names=c("Number of drug classes", colnames(table.numdrugsgr.listed.any)[2:3], 
+             col.names=c("Number of drug classes", colnames(table.notcv.numdrugsgr.care.home)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Association of severe COVID-19 with polypharmacy, by diagnosis of at least one listed condition") %>%
+             caption="Association of severe COVID-19 with number of non-cardiovascular drug classes dispensed, by care home residence") %>%
     column_spec(column=1, width="2cm") %>%
-    kableExtra::group_rows(paste0("No listed condition (",
+    kableExtra::group_rows(paste0("Not resident in care home (",
                       freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
                1, 6) %>%
-    kableExtra::group_rows(paste0("Listed condition (",
+    kableExtra::group_rows(paste0("Care home residents (",
                       freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
@@
 ```
+\newpage
 
@@
 
-newch <- diff(as.integer(substr(subpara.coeffs$subpara, 1, 1))) !=0
+newch <- diff(as.integer(substr(subpara.coeffs$subpara, 1, 1))) != 0
 linesep <- ifelse(newch, '\\addlinespace', '')
+subpara.coeffs$subpara <- gsub("\\.", ": ", subpara.coeffs$subpara)
 
@@
              escape=FALSE,
-             longtable=TRUE, 
+             # longtable=TRUE, 
              booktabs=TRUE,
@@
             "Rate ratio", "$p$-value"),
-             caption=paste("Univariate rate ratios for severe COVID-19 in those without a listed condition associated with dispensing of each BNF subparagraph code, filtered to show only drug classes with at least 50 exposed individuals and  $p < 0.001$: ")) %>%
-    column_spec(column=1, width="10cm") %>%
+             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those without a listed condition and not resident in a care home, filtered to show only drug classes with at least 50 exposed individuals and  $p < 0.001$: ")) %>%
+    column_spec(column=1, width="6cm") %>%
     column_spec(column=2, width="1.5cm") %>%
     column_spec(column=3, width="1.5cm") %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
+    kable_styling(latex_options="hold_position")
 
@@
 
-knitr::kable(table.laporte,
+knitr::kable(table.laporte[-(1:5), ],
              escape=FALSE, 
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 in those aged under 75 years with drug classes listed by Laporte and Healy (2020)") %>%
+             caption="Associations of severe COVID-19 with drug classes listed by Laporte and Healy (2020), restricted to those not resident in a care home.
+Adjusted rate ratios are based on a joint model with all drug classes in the table and SIMD quintile as covariates") %>%
     column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
     kable_styling(latex_options="hold_position") %>%
@@
 
+\newpage
+\beginsupplement
+
+# Supplementary Tables
+
+```{r numdrugs.listed.any, echo=FALSE, warning=FALSE, message=FALSE}
+
+freqs <- with(cc.severe[nocare], table(listed.any, CASE))
+
+knitr::kable(table.notcv.numdrugsgr.listed.any,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="notcv.numdrugs.listed.any",
+             row.names=FALSE,
+             col.names=c("Number of drug classes", colnames(table.notcv.numdrugsgr.listed.any)[2:3], 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption="Association of severe COVID-19 with number of non-cardiovascular drug classes dispensed, in those not resident in a care home by diagnosis of at least one listed condition") %>%
+    column_spec(column=1, width="2cm") %>%
+    kableExtra::group_rows(paste0("No listed condition (",
+                      freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
+               1, 6) %>%
+    kableExtra::group_rows(paste0("At least one listed condition (",
+                      freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
+               7, 12) %>%
+    kable_styling(latex_options="hold_position")
+
+```
+\newpage
+
 ```{r antipsych, echo=FALSE, warning=FALSE, message=FALSE}
-knitr::kable(table.bnf.antipsych,
+
+rownames(table.bnfchapter4.chem) <- tolower.exceptfirstchar(rownames(table.bnfchapter4.chem))
+rownames(table.bnfchapter4.chem) <- gsub("\\.", " ", rownames(table.bnfchapter4.chem))
+
+knitr::kable(table.bnfchapter4.chem,
              escape=FALSE, 
@@
              label="antipsych",
-             col.names=c(clean.header(colnames(table.bnf.antipsych)[1:2]),
+             col.names=c(clean.header(colnames(table.bnfchapter4.chem)[1:2]),
                          "Rate ratio (95\\% CI)", "$p$-value"),
              align=rep("c", 4), 
-             caption="Univariate associations of severe COVID-19 in those aged under 75 years with antipsychotic agents and drugs used in the treatment of nausea and vertigo") %>%
-    column_spec(column=1, width="4cm") %>%
+             caption="Univariate associations of severe COVID-19 with drugs in four subparagraphs of BNF chapter 4, in those not resident in a care home and not treated for cancer in last year, filtered to retain rows with at least 20 exposed individuals") %>%
+    column_spec(column=1, width="5cm") %>%
+    kableExtra::group_rows("Antipsychotic drugs", 1, 7) %>%
+        kableExtra::group_rows("Drugs for nausea and vertigo", 8, 13) %>%
+        kableExtra::group_rows("Opioids", 14, 19) %>%
+        kableExtra::group_rows("Gabapentinoids", 20, 21) %>%
+        kableExtra::group_rows("Control of epilepsy", 22, 30) %>%
         kable_styling(latex_options="hold_position")
+
+```
+
+\newpage
+
+```{r timewin, echo=FALSE, warning=FALSE, message=FALSE}
+
+## drugclasses.timewindows contains names of retained drug classes
+
+knitr::kable(table.timewindows,
+             escape=FALSE,
+             label="timewindow", 
+             booktabs=TRUE,
+             row.names=FALSE, 
+             col.names=c("Exposure category", clean.header(colnames(table.timewindows)[2:3]), 
+                         "Rate ratio (95\\% CI)", "$p$-value"),
+             caption=paste0("Comparison of associations with exposure only in the most recent 120-day time window with exposure only in the previous 120-day time window (baseline category for rate ratios) for drug classes on the Laporte-Healy list with at least 500 cases and controls exposed only in the earlier time window")) %>%
+    kableExtra::group_rows("Proton pump inhibitors", 1, 2) %>%
+    kableExtra::group_rows("Antihistamines", 3, 4) %>%
+    kableExtra::group_rows("Anxiolytics", 5, 6) %>%
+    kableExtra::group_rows("Tricyclic and related antidepressant drugs", 7, 8) %>%
+    kableExtra::group_rows("Drugs used in nausea and vertigo", 9, 10) %>%
+    kableExtra::group_rows("Opioid analgesics", 11, 12) %>%
+    kableExtra::group_rows("Antimuscarinic drugs used in parkinsonism", 13, 14) %>%
+    kableExtra::group_rows("Drugs for urinary frequency enuresis and incontinence", 15, 16) %>%
+	column_spec(column=1, width="3cm") %>%
+	column_spec(column=2:3, width="2cm") %>%
+	column_spec(column=4, width="2.5cm") %>%
+	column_spec(column=5, width="1.5cm") %>%
+    kable_styling(latex_options="hold_position")
+
 ```
@@
 
-```{r timewin, echo=FALSE, warning=FALSE, message=FALSE}
-
-colnames(table.timewindow.opioid) <- colnames(table.timewindow.protonpump)
-table.timewindow <- rbind(table.timewindow.opioid, table.timewindow.protonpump)
-table.timewindow <- data.frame(timewin=rep(rownames(table.timewindow.opioid), 2),
-                               table.timewindow)
-
-knitr::kable(table.timewindow,
-             escape=FALSE,
-             label="timewindow", 
-             booktabs=TRUE,
-             row.names=FALSE, 
-             col.names=c("Time window", clean.header(colnames(table.timewindow.protonpump)[1:2]), 
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             caption=paste0("Associations of severe disease with dispensed prescriptions by ", TW, "-day time window")) %>%
-	column_spec(column=1, width="3cm") %>%
-	column_spec(column=2:3, width="2cm") %>%
-	column_spec(column=4, width="2.5cm") %>%
-	column_spec(column=5, width="1.5cm") %>%
-    kable_styling(latex_options="hold_position") %>%
-    kableExtra::group_rows(group_label="Opioid-containing analgesics (restricted to those without history of neoplasm)", 1, 4) %>%
-    kableExtra::group_rows(group_label="Proton pump inhibitors", 5, 8)
-```
-
-\newpage
-\beginsupplement
-
-# Supplementary Tables
-
-```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter1,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter1",
-             col.names=c(clean.header(colnames(table.bnfchapter1)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe COVID-19 in those aged under 75 years, with prescribed drugs in BNF chapter 1 by subparagraph code filtered to retain rows with at least 50 exposed individuals") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-```
-
 ```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 in those aged under 75 years with prescribed drugs in BNF chapter 2 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
-    column_spec(column=1, width="4cm") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
-    kable_styling(latex_options="hold_position") %>%
-    landscape()
-```
-
-```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.bnfchapter4,
-             escape=FALSE, 
-             booktabs=TRUE,
-             label="bnfchapter4",
-             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-             align=rep("c", 6), 
-             caption="Associations of severe COVID-19 in those aged under 75 years with prescribed drugs in BNF chapter 4 by subparagraph code, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19  in those aged under 75 years with prescribed drugs in BNF chapter 10 by approved name, filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19  in with prescribed drugs in BNF chapter 10 by approved name, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals") %>%
     column_spec(column=1, width="5cm") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
+    kableExtra::group_rows("Non-steroidal anti-inflammatory drugs (excluding topical preparations)", 1, 6) %>%
+    kableExtra::group_rows("Rheumatic disease suppressant drugs", 7, 8) %>%
+    kableExtra::group_rows("Gout and hyperuricaemia", 9, 10) %>%
+    kableExtra::group_rows("Skeletal muscle relaxants", 11, 11) %>%
     kable_styling(latex_options="hold_position") %>%

### 2020-07-28 casecontrol.R

@@ -93,11 +93,14 @@ if(old) {
     controls.status <- readRDS("./data/CC_CHI_CHECK_2020-05-15.rds")
 } else {
     cc.all <- as.data.table(readRDS("./data/CC_linked_ANON_2020-06-18.rds"), key="ANON_ID")
-    setnames(cc.all, "SPECIMENDATE", "SPECDATE")
+    # setnames(cc.all, "SPECIMENDATE", "SPECDATE")
     setnames(cc.all, "AgeYear", "AGE")
     
-    diagnoses <- as.data.table(readRDS("./data/CC_SMR01_ICD10_x25_ANON_2020-06-18.rds"),
+    #diagnoses <- as.data.table(readRDS("./data/CC_SMR01_ICD10_x25_ANON_2020-06-18.rds"),
+    diagnoses <- as.data.table(readRDS("./data/CC_SMR01_ICD10_x25_2020-06-18.rds"),
                                key="ANON_ID")
+    diagnoses <- cc.all[, .(ANON_ID, SPECIMENDATE)][diagnoses] 
+    
     procedures <- readRDS("./data/CC_SMR01_OPCS4_MAIN.x25_ANON_2020-06-18.rds")
     onomap <- as.data.table(readRDS("./data/ONOMAP_ANON_2020-06-18.rds"), key="ANON_ID")
     rapid <- readRDS("./data/CC_RAPID_ANON_2020-06-18.rds")
@@ -154,7 +157,6 @@ names(cc.all) <- gsub("imumune", "immune", names(cc.all))
 names(cc.all) <- gsub("^CAREHOME$", "care.home", names(cc.all))
 names(cc.all) <- gsub("^simd$", "SIMD.quintile", names(cc.all))
 names(cc.all) <- gsub("DATE_OF_DEATH", "Date.Death", names(cc.all))
-names(cc.all) <- gsub("SPECIMENDATE", "SPECDATE", names(cc.all))
 names(cc.all) <- gsub("^age$", "AGE", names(cc.all))
 names(cc.all) <- gsub("^AgeYear$", "AGE", names(cc.all))
 if(!old) {
@@ -164,10 +166,10 @@ cc.all$stratum <- as.integer(cc.all$stratum)
 
 ## merge SICSAG data and overwrite icu and hdu values incorrectly coded 0
 sicsag <- sicsag[covidICUorHDU==1 | covidICUorHDU==3] ## what are codes 4 and 5?
[…]

### 2020-07-28 clogit_hsstan.R

@@ -4,21 +4,31 @@ library(hsstan)
 library(data.table)
 library(parallel)
 
-cc.severe <- as.data.table(readRDS("data/cc.severe.rds"))
-                           
-## restrict to fatal cases + matched controls, and complete data on covariates
-select.cols <- c("CASE", "stratum", "care.home", "diabetes.any",
-                 grep("^Ch\\.", colnames(cc.severe), value=TRUE))
-covariates <- select.cols[-(1:2)][1:8]
+devtools::load_all("../hsstan")
 
-cc.nonmissing <- na.omit(cc.severe[fatal.casegroup==1], cols=select.cols)
-cc.nonmissing <- cc.nonmissing[, ..select.cols]
-covariate.names <- select.cols[-(1:2)]
-## convert all columns to numeric
-cc.nonmissing <- cc.nonmissing[, lapply(.SD, as.numeric), by=row.names(cc.nonmissing)][, -1]
-cc.nonmissing[, (covariate.names) := lapply(.SD, scale), .SDcols=covariate.names] 
-str(cc.nonmissing)
- 
+newrun <- TRUE
+
+cc.severe <- as.data.table(readRDS("data/cc.severe.rds"))
+
+drugclasses <-  grep("^subpara\\.", colnames(cc.severe), value=TRUE)
+## restrict to subparas with at least 50 exposed
+keep.drugclasses <- logical(length(drugclasses))
+for(j in 1:length(keep.drugclasses)) {
+    x <- length(which(cc.severe[[drugclasses[j]]]=="1"))
+    if(x >= 50) {
+        keep.drugclasses[j] <- TRUE
+    }
+}
+drugclasses <- drugclasses[keep.drugclasses]
+
+## restrict to fatal cases + matched controls, and complete data on covariates
+select.cols <- c("CASE", "stratum", "care.home",
+                 grep("^Ch\\.", colnames(cc.severe), value=TRUE),
+                 drugclasses)
+
+cc.nonmissing <- na.omit(cc.severe[fatal.casegroup==1, ..select.cols], cols=select.cols)
+rm(cc.severe)
+
 ## drop strata with num cases not equal to 1, or < 2 observations 
 cc.drop <- cc.nonmissing[CASE==1, .(.N), by = .(stratum)][N != 1, ]
 cc.nonmissing <- cc.nonmissing[!(stratum %in% cc.drop$stratum)]
[…]

### 2020-07-28 comorbidity.R

@@ -100,13 +100,19 @@ cc.severe$immune.any <- as.factor(as.integer(cc.severe$ANON_ID %in% ids.immune.a
 
 ############# neoplasms ################
 
-ids.icd.neoplasm <- unique(diagnoses$ANON_ID[grep("^C[0-9]|^D[0-4]", diagnoses$ICD10)])
-ids.bnf.neoplasm <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 801])
-ids.neoplasm.any <- unique(c(ids.icd.neoplasm, ids.bnf.neoplasm))
+ids.icd.neoplasm <- unique(diagnoses[grepl("^C[0-9]|^D[0-4]", ICD10), ANON_ID])
 
-cc.severe <- mutate(cc.severe,
-                    neoplasm.any = as.factor(as.integer(ANON_ID %in% ids.neoplasm.any |
-                                             can.reg==1)))
+ids.icd.neoplasm.lastyear <-
+    unique(diagnoses[as.integer(SPECIMENDATE - DISCHARGE_DATE) + 25 > 365 &
+                     grepl("^C[0-9]|^D[0-4]", ICD10), ANON_ID])
+ids.bnf.neoplasm <- unique(scrips$ANON_ID[as.integer(scrips$sectioncode) == 801])
+
+ids.neoplasm.any <- unique(c(ids.icd.neoplasm, ids.bnf.neoplasm))
+ids.neoplasm.lastyear <- unique(c(ids.icd.neoplasm.lastyear, ids.bnf.neoplasm))
+                                
+cc.severe[, neoplasm.any := as.factor(as.integer(ANON_ID %in% ids.neoplasm.any |
+                                             can.reg==1))]
+cc.severe[, neoplasm.lastyear := as.factor(as.integer(ANON_ID %in% ids.neoplasm.lastyear))]
 
 ###### disorders of esophagus, stomach and duodenum ############################
 

### 2020-07-28 drugs.R

@@ -1,4 +1,24 @@
 
+compare.timewindows <- function(scrips.subset, data, recent.cutoff=120) {
+    ids.recent <- unique(scrips.subset[daysbefore < recent.cutoff + 16, .(ANON_ID)]) 
+    ids.nonrecent <- unique(scrips.subset[daysbefore >= recent.cutoff + 16, .(ANON_ID)])
+    cc.test <- data[, .(ANON_ID, CASE, stratum)]
+    cc.test[, recent := ANON_ID %in% ids.recent$ANON_ID]
+    cc.test[, nonrecent := ANON_ID %in% ids.nonrecent$ANON_ID]
+    cc.test[, exposure.cat := as.integer(recent | nonrecent)]
+    cc.test[recent & !nonrecent, exposure.cat := 2]
+    cc.test[recent & nonrecent, exposure.cat := 3]
+    cc.test[, exposure.cat := as.factor(car::recode(exposure.cat, 
+                               "0='No prescriptions'; 1='Non-recent only';
+                                2='Recent only';
+                                3='Prescriptions in both time windows'"))]
+    cc.test[, exposure.cat := factor(exposure.cat,
+                                     levels=levels(exposure.cat)[c(2, 4, 3, 1)])]
+    freqs <- univariate.tabulate(varnames="exposure.cat", outcome="CASE", data=cc.test, drop.reflevel=FALSE)
+    coeffs <- univariate.clogit(varnames="exposure.cat", outcome="CASE", data=cc.test, add.reflevel=TRUE)
+    return(combine.tables2(freqs, coeffs))
+}
+
 get.timewindow <- function(scrips.in) {
     ## add columns daysbefore, intervalTWday, dispensing.days
     ## the 15-day cutoff has been applied to the scrips table
@@ -59,7 +79,7 @@ antihypertensive.classes <- c("vasodilator_ah6.bnf",
 
 ## for now, just keep id, paragraph code, paragraph_description
 
-scrips$bnf_paragraph_description <- as.factor(scrips$bnf_paragraph_description)
+scrips[, bnf_paragraph_description := as.factor(bnf_paragraph_description)]
 cat("scrips object uses", object.size(scrips) * 1E-6, "MB\n")
 
 ## we have 7 digits on scrips, giving resolution to subpara level only
@@ -78,9 +98,7 @@ scrips[is.na(chapternum), chapternum := 14]
 scrips[chapternum > 14, chapternum := 14] 
 
 if(!old) {
[…]

### 2020-07-28 helperfunctions.R

@@ -1,5 +1,50 @@
 ## helperfunctions for COVID analyses
 
+
+`clogistLoglike`  <-  function(n,m,x,beta) {
+    ## https://rdrr.io/cran/saws/src/R/clogistLoglike.R
+    ## for computational stability should use matrixStats::logSumExp()
+    ## m is vector of 0-1 indicators of failure
+    ## n is vector of 1s
+    M <- sum(m) 
+    N <- sum(n)
+    if (M==0)  return(0)
+    else if (M==N) return(0)
+    x <- as.matrix(x)
+    eta <-  x %*% beta
+    U <- exp(eta)
+    if (M==1) return(sum(eta*m) - log(sum(U*n)) )
+    if (M > N/2) {
+        ## for efficiency, keep loop part of calculation to minimum
+        ## by switching m and n-m, beta and -beta
+        m <- n - m
+        M <- N - M
+        U <- 1/U
+        eta <-  -eta
+    }
+    if (M==1) return(sum(eta*m) - log(sum(U*n)) )
+    B <- rep(1, N-M+1)
+    u <- rep(NA, N)
+    count <- 1
+    for (a in 1:length(n)){
+        u[count:(count + n[a] - 1)] <- U[a]
+        count <- count + n[a]
+    }
+    ## last 2 lines may be written more 
+    ## clearly (i.e. more like in Gail et al) BUT LESS EFFICIENTLY as:
+                                        #B <- matrix(0,M+1,N+1)
+                                        #B[1,] <- 1
+                                        #for (i in 1:M){
+                                        #for (j in i:(N-M+i)){
+                                        #B[i+1,j+1] <-  B[i+1,j]+u[j]*B[i,j]
+                                        #}
+                                        #} 
+                                        #sum(eta*m) - log(B[M+1,N+1])
+    for (i in 1:(M-1)) B <-  cumsum(B * u[i:(N-M+i)])
+    sum(eta * m) - log(sum(B * u[M:N]))
+}
+
+
 recode.indicator <- function(x) {
     x[is.na(x)] <- 0
     x[x > 1] <- 1
@@ -121,19 +166,9 @@ lookup.names <- data.frame(varname=c("AGE", "sex",
                                      "lung.cancer",
                                      "other.cancer",
                                      "immune.any",
-                                     "alpha_adrenoreceptor6.bnf",
-                                     "ace6.bnf",
-                                     "angio6.bnf",
-                                     "renin_angiotensin6.bnf",
[…]

### 2020-07-28 test_clogit.R

@@ -27,7 +27,7 @@ regularized <- TRUE
 covs.model <- as.formula(paste("y.clogit ~", paste(map.varnames, collapse="+")))  
 
 hs.clogit <- hsstan(x=diabetes, covs.model=covs.model,
-                   penalized=penalized, family="cox",
+                   penalized=penalized, family="clogit",
                    iter=1000, warmup=500,
                    scale.u=2, regularized=TRUE, nu=ifelse(regularized, 1, 3),
                    par.ratio=0.05, global.df=1, slab.scale=2, slab.df=4,

### 2020-11-03 casecontrol.Rmd
multiple updates of pharmaco and shielding manuscripts

@@
 
-# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol8June.pdf"); rmarkdown::render("response_reviewers.md")
+# rmarkdown::render("casecontrol.Rmd", output_file="casecontrol_supptables.pdf"); rmarkdown::render("response_reviewers.md")
 
@@
 
+pctonly <- function(x) gsub("([0-9]+ \\()(.*)(\\))", "\\2", x)
+
+rr.pval <- function(r, p) {
+    r <- gsub(",", " to", r)
+    p <- gsub("ensuremath\\{", "\\ensuremath\\{p=", p)
+    p <- gsub("times", "\\\\times", p)
+    gsub("\\)", paste0(", \\\\", p, "\\)"), r)
+}
+
+##rr.pval(table.agegr.all[1, 3], table.agegr.all[1, 4])
+
 ```
@@
 Severe COVID-19 was defined as either a positive nucleic acid test for SARS-CoV-2 in the national database followed by entry to a critical care unit or death within 28 days, or a death certificate with COVID-19 as underlying cause.
-Up to ten controls per case matched for sex, age and primary care practice were selected from the population register.
-All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past nine months were extracted.
+Up to ten controls per case matched for sex, age and primary care practice were selected from the national population register.
+For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECIMENDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths registered up to `r format(max(cc.severe$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==0])` controls and `r nrow(cc.severe[CASE==1])` cases of whom `r table.demog.append["Care home", 3]` were care home residents.
+All diagnostic codes from the past five years of hospitalisation records and all drug codes from prescriptions dispensed during the past 240 days were extracted.
 Rate ratios for severe COVID-19 were estimated by conditional logistic regression.
 
-There were `r length(which(cc.severe$CASE==1))` severe cases.
 In a logistic regression using the age-sex distribution of the national population, the odds ratios for severe disease were `r round(exp(10 * logistic.coeffs[3, 1]), 2)` for a 10-year increase in age and `r round(exp(-logistic.coeffs[2, 1]), 2)` for male sex.
-In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r table.agegr.all[1, 3]`.
-Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r table.agegr.all["Type 1 diabetes", 3]` for Type 1 diabetes, `r table.agegr.all["Type 2 diabetes", 3]` for Type 2 diabetes,
-`r table.agegr.all["Ischaemic heart disease", 3]` for ischemic heart disease,
-`r table.agegr.all["Other heart disease", 3]` for other heart disease, 
-`r table.agegr.all["Asthma or chronic airway disease", 3]` for chronic lower respiratory tract disease, 
-`r table.agegr.all["Chronic kidney disease or transplant recipient", 3]` for chronic kidney disease, 
-`r table.agegr.all["Neurological (except epilepsy) or dementia", 3]` for neurological disease, 
-`r table.agegr.all["Liver disease", 3]` for chronic liver disease and 
-`r table.agegr.all["Immune deficiency or suppression", 3]` for immune deficiency or suppression.
+In the case-control analysis, the strongest risk factor was residence in a care home, with rate ratio (95% CI) `r rr.pval(table.agegr.all[1, 3], table.agegr.all[1, 4])`.
+Univariate rate ratios (95% CIs) for conditions listed by public health agencies as conferring high risk were `r rr.pval(table.agegr.all["Type 1 diabetes", 3], table.agegr.all["Type 1 diabetes", 4])` for Type 1 diabetes, `r rr.pval(table.agegr.all["Type 2 diabetes", 3], table.agegr.all["Type 2 diabetes", 4])` for Type 2 diabetes,
+`r rr.pval(table.agegr.all["Ischaemic heart disease", 3], table.agegr.all["Ischaemic heart disease", 4])` for ischemic heart disease,
+`r rr.pval(table.agegr.all["Other heart disease", 3], table.agegr.all["Other heart disease", 4])` for other heart disease, 
+`r rr.pval(table.agegr.all["Asthma or chronic airway disease", 3], table.agegr.all["Asthma or chronic airway disease", 3])` for chronic lower respiratory tract disease, 
+`r rr.pval(table.agegr.all["Chronic kidney disease or transplant recipient", 3], table.agegr.all["Chronic kidney disease or transplant recipient", 4])` for chronic kidney disease, 
+`r rr.pval(table.agegr.all["Neurological (except epilepsy) or dementia", 3], table.agegr.all["Neurological (except epilepsy) or dementia", 4])` for neurological disease, 
+`r rr.pval(table.agegr.all["Liver disease", 3], table.agegr.all["Liver disease", 4])` for chronic liver disease and 
+`r rr.pval(table.agegr.all["Immune deficiency or suppression", 3], table.agegr.all["Immune deficiency or suppression", 4])` for immune deficiency or suppression.
 
-`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all["listed.any", 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", freqs.all["listed.any", 1])` of controls had at least one listed condition (`r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr["Any listed condition", 2])` of cases and `r gsub("([0-9]+ \\()(.*)(\\))", "\\2", table.agegr["Any listed condition", 1])` of controls under age 40).
+`r pctonly(freqs.all["listed.any", 2])` of cases and `r pctonly(freqs.all["listed.any", 1])` of controls had at least one listed condition (`r pctonly(table.agegr["Any listed condition", 2])` of cases and `r pctonly(table.agegr["Any listed condition", 1])` of controls under age 40).
 Severe disease was associated with encashment of at least one prescription in the past nine months and with at least one hospital admission in the past five years [rate ratios `r table.agegr.all["Any prescription", 3]`] and `r table.agegr.all["Any admission", 3]` respectively] even after adjusting for the listed conditions.
@@
 
-_Conclusions_ -- Along with older age and male sex, severe COVID-19 is strongly associated with past medical history across all age groups.
+_Conclusions_ -- We have shown that along with older age and male sex, severe COVID-19 is strongly associated with past medical history across all age groups.
 Many comorbidities beyond the risk conditions designated by public health agencies contribute to this.
@@
 # Author summary
-Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill.
-It is The risk of severe or fatal illness is higher in older than in younger people, and is higher in people with conditions such as asthma and diabetes than in people without these conditions.
-Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.
-We show that the strongest risk factor, apart from age, is residence in a care home.
-The conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but many other diagnoses, associated with frailty and poor health.
-This lays a basis for constructing risk scores based on electronic health records that can be used to advise people at high risk of severe disease to shield themselves when there cases in their neighbourhood.
+## Why was this study done? 
+* Most people infected with the SARS-CoV-2 coronavirus do not become seriously ill: risk of severe or fatal disease is associated with older age, male sex, and conditions designated by public health agencies including asthma, diabetes and heart disease.
+
+* Studies reported so far have focused on these "listed conditions", but have not examined medical records systematically to identify possible risk factors for severe COVID-19.
+
+* The objectives of this study were to identify risk factors for severe COVID-19 and to lay the basis for risk stratification based on electronic health records.
+
+## What did the researchers do and find? 
+* Using Scotland's capability for linking electronic health records, we report the first systematic study of the relation of severe or fatal COVID-19 to pre-existing health conditions and other risk factors.
+
+* Residents in care homes were 21 times more likely to develop severe disease than people of the same age and sex not living in care homes.
+
+* The conditions associated with increased risk include not only those already designated by public health agencies -- asthma, diabetes, heart disease, disabling neurological disease, kidney disease -- but other diagnoses that are associated with frailty and poor health such as strokes and a history of falls.
+
+* In those without any listed conditions, use of prescribed drugs acting on the digestive system or nervous system is associated with increased risk of severe COVID-19.
+
+## What do these findings mean? 
+* The risk to younger individuals without any recent history of hospital admission or use of prescription drugs is very low.
+
+* This study lays a basis for calculating a risk score based on electronic health records for every individual in the population, and using it to advise those at high risk of severe disease to shield themselves when there is a COVID-19 epidemic in their locality.
 
@@
 Case series from many countries have suggested that in those with severe COVID-19 the prevalence of diabetes and cardiovascular disease is higher than expected.
-For example in a large UK series the commonest co-morbidities were cardiac disease, diabetes, chronic pulmonary disease and asthma [@docherty_features_2020].
+For example in a large UK series the commonest co-morbidities were cardiac disease, diabetes, chronic pulmonary disease and asthma [@docherty_features_2020-1].
 However there are also anecdotal reports of apparently healthy young persons succumbing to disease [@mcgoogan_why_2020].
@@
 Quantification of the risk associated with characteristics and co-morbidities has been limited by the lack of comparisons with the background population [@zhou_clinical_2020;@grasselli_baseline_2020;@guan_clinical_2020].
-Two recent studies in the UK have included population comparators and have reported associations of in hospital test positive persons and COVID-19 death in hospital with co-morbidities including diabetes, asthma and heart disease [@niedzwiedz_ethnic_2020; @the_opensafely_collaborative_opensafely_2020].
+Two recent studies in the UK have included population comparators and have reported associations of in hospital test positive persons and COVID-19 death in hospital with co-morbidities including diabetes, asthma and heart disease [@niedzwiedz_ethnic_2020; @williamson_opensafely_2020].
 These studies  have focused on conditions presumptively listed by public health agencies as increasing risk for COVID-19 based on case series data.
@@
 # Methods
-\hl{The study followed a pre-specified protocol.
-Modifications after study started were that we aligned the list of conditions of interest to be consistent with those designated as moderate risk conditions by public health agencies, and extended the list of drug classes under study to include all drugs.}
+The protocol of the study dated 16 April 2020, together with all code used to prepare this manuscript is available in a public [repository](https://github.com/pmckeigue/covid-scotland_public/).
+We modified the original protocol to align the list of risk conditions to be consistent with those designated by public health agencies, and extended the list of drug classes under study to include all drugs.
+The study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).
+All record linkage studies using NHS data in Scotland are governed by the Public Benefit and Privacy Panel for Health and Social Care, which includes patient and public representatives.
+Identifiable data were extracted by the Public Health Scotland CHI database and linkage team and de-identified before provision to the analysis team.
 
@@
 Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
-\hl{Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.}
+Using this definition ensures ascertainment of all severe cases even if they die without testing positive or entering critical care, whatever selection policies may have limited entry to critical care.
 
@@
 For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2 and using all codes on the discharge.
-Diagnostic coding under ICD chapters 5 (Mental, Behavioural and Neurodevelopmental) and 15 (Pregnancy) is incomplete as most psychiatric and maternity unit returns are not captured in SMR01.
+Diagnostic coding under ICD chapters V (Mental, Behavioural and Neurodevelopmental) and XV (Pregnancy) is incomplete as most psychiatric and maternity unit returns are not captured in SMR01.
 British National Formulary (BNF) drug codes for dispensed prescriptions issued in primary care were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016].
@@
 Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other/Unknown.
-
-Identifiable data were extracted by the Public Health Scotland CHI database and linkage team and de-identified before provision to the analysis team.
 
@@
 Higher risk of severe disease was also associated with socioeconomic deprivation.
-In the   `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases  and  `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls  in whom ethnicity had been recorded in the Scottish Morbidity Record, there were few non-White cases or controls and the confidence limits for the risk ratios for severe or fatal disease by ethnic group were wide.
+In the   `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 2])` of cases  and  `r gsub("(.*\\()(.*)\\)", "\\2", table.ethnicsmr.missing[1, 1])` of controls  in whom ethnicity had been recorded in the Scottish Morbidity Record, there were few non-White individuals and the confidence limits for the rate ratios by ethnic group were wide.
 
@@
 ## Comparison of fatal and non-fatal cases
-
 Supplementary Table \ref{tab:casegr} shows a breakdown of severe cases by test-positive status, entry to critical care, and fatal versus non-fatal outcome.
-\hl{Severe cases who entered critical care were much younger than severe cases never entering critical care.
+Severe cases who entered critical care were much younger than severe cases never entering critical care.
 The majority of severe cases from residential care homes never entered critical care.
@@
 Among those entering critical care the median age and the proportion of males and prevalence or prior comorbidities were higher in fatal than in non-fatal cases.
-}
 
@@
 These results are shown in Supplementary Table \ref{tab:conditions}.
-In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters 7 (eye) Chapters 8 (ear) and Chapter 15 (pregnancy) were associated with increased risk of severe disease.
-In a multivariable analysis the strongest associations were with diagnoses in ICD chapters 4 (mental disorders) and 10 (respiratory).
+In univariate analyses, almost all ICD-10 chapters, with the exception of Chapters VII (eye), VIII (ear) and XV (pregnancy) were associated with increased risk of severe disease.
+In a multivariable analysis the strongest associations were with diagnoses in ICD chapters IV (mental disorders) and X (respiratory).
 Supplementary Table \ref{tab:icdsubchapters} extracts univariate associations with ICD-10 subchapters in those without any listed conditions.
@@
 Observed and predicted case status were compared within each stratum over all test folds.
-Table \ref{tab:xvalid} shows that using the extended set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5]` bits.
+Table \ref{tab:xvalid} shows that in comparison with using only demographic variables and listed conditions, using the extended variable set increased the C-statistic from `r table.densities[2, 3]` to `r table.densities[3, 3]` and the expected information for discrimination $\Lambda$ from `r table.densities[2, 5]` bits to `r table.densities[3, 5]` bits.
 
@@
 # Discussion  
-### Sociodemographic factors
+## Sociodemographic factors
 This analysis confirms that risk for severe COVID-19 is associated with increasing age, male sex and socioeconomic deprivation.
@@
 
-Although the assignment of ethnicity is incomplete and the prevalence of non-White ethnic groups in Scotland is low, we show the data as they give some indication of the likely upper bound of the absolute numbers of severe cases in South Asian and Black ethnic groups in Scotland.
-This is relevant because substantial elevations in risk of hospitalised and fatal COVID-19 have been reported elsewhere in the UK [@the_opensafely_collaborative_opensafely_2020; @niedzwiedz_ethnic_2020; @ho_modifiable_2020].
-In the OpenSAFELY study risk ratios for fatal COVID-19 of 1.7 in those recorded as Black and and 1.6 in those recorded as Asian, in comparison with those recorded as White, persisted after adjustment for comorbidities and socioeconomic status.
-In Scotland however, although point estimates for rate ratios in those of South Asian and Black ethnicity compared to Whites are above 1, the confidence intervals are too wide for any definitive conclusions on ethnic differences.
+As the proportion of the Scottish population that is of non-White ethnicity is low and the assignment of ethnicity in this dataset is incomplete, the confidence intervals for the rate ratios associated with South Asian and Black ethnicity are wide.
+Studies from England [@williamson_opensafely_2020; @niedzwiedz_ethnic_2020; @ho_modifiable_2020] have reported elevations in risk of hospitalised and fatal COVID-19 in non-White ethnic groups; in the OpenSAFELY study the risk ratios for fatal COVID-19 associated with Black and Asian ethnicity were 1.7 and 1.6 respectively.
+The confidence intervals in this study are compatible with the effect sizes estimated in England.
 
-### Co-morbidities
+## Co-morbidities
 We have confirmed that the moderate risk conditions designated by the NHS and other agencies [@nhs_whos_2020] are associated with increased risk of severe COVID-19.
-However the rate ratios associated with these conditions vary with age - for example the rate ratio associated with diabetes is higher at younger ages.
-The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in UK Biobank and in the OpenSAFELY studies.
+The rate ratios of `r round(as.numeric(substr(table.agegr.all["Type 1 diabetes", 3], 1, 4)), 1)` for Type 1 diabetes and `r round(as.numeric(substr(table.agegr.all["Type 2 diabetes", 3], 1, 4)), 1)` for Type 2 diabetes are broadly similar to those reported in the UK Biobank [@ho_modifiable_2020] and OpenSAFELY [@williamson_opensafely_2020] studies.
 We confirm the higher risk with asthma and chronic lung disease and liver disease reported in these and earlier studies.
+The rate ratios associated with these risk conditions vary with age: for example the rate ratio associated with diabetes is higher at younger ages.
 An unexpected finding was that the risk associated with other forms of heart disease is higher than that associated with ischaemic heart disease.
@@
 Over all age groups, `r gsub("(.*\\()(.*)\\)", "\\2", freqs.all["listed.any", 2])` of severe cases had at least one of these listed conditions.
-\hl{ In this dataset it is not possible to examine adequately the risk associated with neoplasms, as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system.
-We hope to explore this in a separate study that will include linkage to records of shielding advice.} In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.
+In this dataset it is not possible to examine adequately the risk associated with neoplasms as we cannot separately identify those who were advised to shield themselves because they had active neoplasms of lymphoid or hematopoietic tissue or were receiving treatments that affect the immune system.
+We plan to explore this in a separate study based on linkage to records of shielding advice.
+In patients without any listed conditions, further systematic evaluation of past hospitalisation history did not reveal a sparse set of underlying conditions; instead many diagnoses were associated with severe COVID-19.
 
-Media reports of apparently healthy young people succumbing to severe COVID-19 have disseminated the message that all are at risk of disease whatever their age or health status.
+Public health agencies [@ghebreyesus_who_2020] and media reports of apparently healthy young people succumbing to severe COVID-19 [@mcgoogan_why_2020] have disseminated the message that all are at risk of severe COVID-19 whatever their age or health status.
 However we found that half of cases under 40 years had at least one of the listed conditions and among those who did not have one of these conditions, the proportions who had at least one prior hospitalisation or dispensed prescription were higher in cases than in controls.
@@
 
-A striking finding of this study was the association of severe COVID-19 with dispensing of at least one prescription in the past year, only partly explained by higher rates of prescribing among those with listed conditions.
+A striking finding of this study was the association of severe COVID-19 with dispensing of at least one prescription in the 240-day interval preceding the cutoff of 15 days before diagnosis, only partly explained by higher rates of prescribing among those with listed conditions.
 Partitioning of this association between BNF chapters, which represent broad indication-based drug classes, showed that prescription of drugs for the gastrointestinal and central nervous systems, together with nutritional supplements, contributed to this association.
@@
 So too does the similarly named "segment and shield " option [@bunnik_segmentation_2020] which has the opposite objective of keeping transmissions low.
-Although for this preliminary study we have not used the full repertoire of machine learning methods available for this type of problem, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.
+Although in this preliminary study we have not used the full repertoire of machine learning methods available for constructing predictive models, we have shown that a model based on health records provides `r table.densities[3, 5]` bits of information for discrimination conditional on age and sex.
 Adding this to the `r round(cases.Lambda.agesex, 2)` bits provided by age and sex gives a total information for discrimination of `r round(table.densities[3, 5] + cases.Lambda.agesex, 1)` bits.
-\hl{We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population} [@mckeigue_evaluation_2020]
+We have shown elsewhere that this level of predictive performance would allow at least 80% of those at risk of severe or fatal disease to be allocated to a shielded group comprising no more than 15% of the population [@mckeigue_evaluation_2020].
 
@@
 It will also exclude many people at high risk because they have multiple risk factors each of small effect.
-\hl{A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study}.
+A more meaningful way to score risk for an individual would be to use all available information to calculate a "COVID age" as the age at which the average risk for someone of the same sex in the population equates to the risk for the individual under study.
 Thus the rate ratio of `r rr <- round(as.numeric(gsub(" .+", "", table.agegr.all["Type 1 diabetes", 3])), 1); rr` associated with Type 1 diabetes equates to an increase of `r round(log(rr) / logistic.coeffs[3, 1], 1)` years in COVID age.
@@
 With this matched case control design using incidence density sampling, we have been able to estimate rate ratios conditional on age and sex.
-An unpublished analysis from England explored the association of similar set of risk conditions with in-hospital COVID-19 deaths, but did not systematically evaluate the rest of the medical record including prescription records.
+The OpenSafely study has explored associations of a similar set of risk conditions with in-hospital COVID-19 deaths [@williamson_opensafely_2020], but has not yet reported a systematic evaluation of the rest of the medical record including prescription records.
 Although we have records of encashment of prescriptions, we do not at present have access to other primary care data, which would contain additional information on morbidity and measurements such as body mass index.
@@
 
+<!---
 # Declarations
@@
 The authors declare no conflicts of interest.
-
-## Acknowledgements
+--->
+# Acknowledgements
 We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
@@
 theme_set(
-    theme_gray(base_size = 14)
+    theme_gray(base_size = 12)
 )
@@
                        labels=seq(20, 80, by=20)) + 
+    scale_x_continuous(limits=c(20, 90), expand=c(0, 0)) +
     scale_y_continuous(name="Risk (logit scale)",
@@
                        labels=breaks.gam,
-                       limits=car::logit(c(2E-5, 5E-3)))
+                       limits=car::logit(c(2E-5, 5E-3))) +
+    theme(axis.title=element_text(size=12)) +
+    theme(legend.position=c(0.85, 0.3))
 p.incidence
@@
 p.wdists <- plotWdists(densities=full.densities, distlabels=c("Crude", "Adjusted")) +
-    theme_grey(base_size = 18) +
-    theme(legend.title=element_blank())
+    theme_grey(base_size = 16) +
+    theme(axis.title=element_text(size=14)) +
+    theme(legend.title=element_blank()) +
+    theme(legend.position=c(0.85, 0.75))
 
@@
     coord_fixed(ratio=1) +
-    theme(axis.title=element_text(size=18)) +
+    theme_grey(base_size = 16) +
+    theme(axis.title=element_text(size=14)) +
     theme(legend.title=element_blank()) + 
@@
     ylab("Sensitivity")
+p.roc
+```
 
-p.roc
+```{r tiff, echo=FALSE, warning=FALSE, message=FALSE}
 
 if(FALSE) {
-## Width: 789 – 2250 pixels (at 300 dpi).
 ## Height maximum: 2625 pixels (at 300 dpi).
 ## Resolution 300 – 600 dpi
-tiff("Figure1.tiff", res=300, width=1600, height=1200)
+tiff("Fig1.tiff", res=300, width=1800, height=1500)
 p.incidence
 dev.off()
-tiff("Figure2.tiff", res=300, width=2100, height=1800)
+tiff("Fig2.tiff", res=300, width=2100, height=1800)
 p.wdists
 dev.off()
-tiff("Figure3.tiff", res=300, width=2100, height=1800)
+tiff("Fig3.tiff", res=300, width=2100, height=1800)
 p.roc
@@
 
+\newpage
+
 ```{r infodiscrim, echo=FALSE, warning=FALSE, message=FALSE}
@@
 # Supplementary tables
-
-```{r casegr, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.casegr,
-             escape=FALSE, 
-             booktabs=TRUE,
-			 longtable=TRUE, 
-             label="casegr",
-             col.names=paste(c("Critical care, non-fatal", "Critical care, fatal",
-                               "No critical care, fatal", "No critical care, fatal"),
-                             paste0(rep("(", 4),
-                                    with(cc.severe, table(severe.casegr)), rep(")", 4))), 
-             caption="Comparison of severe non-fatal and fatal cases, by test positive status and entry to critical care") %>%
-    add_header_above(c(" "=1, "Test-positive"=3, "No positive test"=1)) %>%
-    add_header_above(c(" "=1, "Non-fatal"=1, "Fatal"=3)) %>%
-    column_spec(column=2:5, width="3cm") %>%
-    landscape() %>%
-    kable_styling(latex_options="hold_position")
-
-```
 
@@
 
+```{r casegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.casegr,
+             escape=FALSE, 
+             booktabs=TRUE,
+			 longtable=TRUE, 
+             label="casegr",
+             col.names=paste(c("Critical care, non-fatal", "Critical care, fatal",
+                               "No critical care, fatal", "No critical care, fatal"),
+                             paste0(rep("(", 4),
+                                    with(cc.severe, table(severe.casegr)), rep(")", 4))), 
+             caption="Comparison of severe non-fatal and fatal cases, by test positive status and entry to critical care") %>%
+    add_header_above(c(" "=1, "Test-positive"=3, "No positive test"=2)) %>%
+    add_header_above(c(" "=1, "Non-fatal"=1, "Fatal"=3)) %>%
+    column_spec(column=2:8, width="3cm") %>%
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
 ```{r conditions, echo=FALSE, warning=FALSE, message=FALSE}
@@
              align=rep("c", 6), 
-             caption="Associations of severe disease with hospital diagnoses in last 5 years, in those without any listed condition") %>%
+             caption="Associations of severe disease with hospital diagnoses by ICD chapter in last 5 years, in those without any listed condition") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6),
-             caption="Associations of severe disease with prescribed drugs in those without any listed condition") %>%
+             caption="Associations of severe disease with prescribed drugs by BNF chapter in those without any listed condition") %>%
     column_spec(column=1, width="4.5cm") %>%
@@
 
+rownames(table.icdsubchapters) <- gsub("_", " ", rownames(table.icdsubchapters))
+
 knitr::kable(table.icdsubchapters[grep("times", table.icdsubchapters$u.pvalue), ],
@@
              label="icdsubchapters",
-			 linesep=c('', '\\addlinespace',
-                       rep('', 2), '\\addlinespace', # neoplasma
-                       '\\addlinespace', # metabolic
-                       rep('', 2), '\\addlinespace', # nervous
-                       '', '\\addlinespace', # respiratory
-                       '\\addlinespace', # skin
-                       '', '\\addlinespace', # urinary
-                       rep('', 8)), 
+			 linesep=c('', '\\addlinespace', # 2 bacterial
+                       rep('', 3), '\\addlinespace', # 4 neoplasms
+                           '\\addlinespace', # 1 metabolic
+                       '', '\\addlinespace', # 2 nervous
+                           '\\addlinespace', # 1 circulatory
+                       '', '\\addlinespace', # 2 respiratory
+                           '\\addlinespace', # 1 skin
+                       '', '\\addlinespace', # 2 urinary
+                       rep('', 8)),  # 8 symptoms, injuries
              col.names=c(clean.header(colnames(table.icdsubchapters)[1:2]),
@@
     column_spec(column=1, width="13cm") %>%
-    column_spec(column=2:3, width="1.5cm") %>%
+    column_spec(column=2:3, width="1.6cm") %>%
     column_spec(column=5, width="1.5cm") %>%
@@
 
+z <- stepwise.full[, 4]
+pchar <- as.character(signif(stepwise.full[, 5], 1))
+pchar[pchar == "0"] <- pnorm.extreme(z[pchar == "0"])
+pchar <- as.character(pvalue.latex(pchar))
+
 retained.full <- as.data.frame(stepwise.full[, c(1, 5)])
 retained.full[, 1] <- round(retained.full[, 1], 2)
-retained.full[, 2] <- signif(retained.full[, 2], 1)
+retained.full[, 2] <- pchar
 
@@
 rownames(retained.full) <- gsub("dm\\.type", "", rownames(retained.full))
+rownames(retained.full) <- gsub("SIMD\\.quintile", "quintile", rownames(retained.full))
 
@@
 rownames(retained.full) <- gsub("\\_", " ", rownames(retained.full))
-
-retained.full[, 2] <- pvalue.latex(retained.full[, 2])
 

### 2020-11-03 herd.Rmd
multiple updates of pharmaco and shielding manuscripts

---
title: 'Understanding the COVID-19 epidemic'
author: Paul McKeigue

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
  pdf_document

always_allow_html: true
urlcolor: blue
---

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

# rmarkdown::render("herd.Rmd", output_file="herd.pdf")

options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

source("helperfunctions.R")

```

# The herd immunity threshold

This 1927 [paper](https://royalsocietypublishing.org/doi/pdf/10.1098/rspa.1927.0118) from the Scottish biochemist William Kermack and Lt Col Anderson McKendrick of the Indian Medical Service is the first derivation of the standard forumula for the herd immunity threshold.   They were careful to state a key assumption in the abstract:

> In the present communication discussion will be limited to the case in which all members of the community are initially equally susceptible to the disease,

Expressed in modern notation, the equation that they derived for the herd immunity threshold  immunity threshold $H$ can be written in terms of the basic reproduction number $R_0$ as:

\[
H = 1 - \frac{1}{R_0} 
\]

I'll try to give this a visual interpretation

With all uninfected individuals initially equally susceptible, the distribution of susceptibility can be represented by a spike at 1 representing those who have not been infected.

As the epidemic proceeds, the distribution of susceptibility becomes a mixture of a spike at zero (the recovered individuals who are now immune) and a spike at 1 (the remaining susceptible individuals). At time $t$, the average susceptibility $S\left( t \right)$ is the centre of mass (the first moment) of this distribution.

As the average susceptibility $S\left( t \right)$ falls, the effective reproduction number $R_e$  falls proportionately.

\[
R_e = S \left( t \right) R_0
\]

At the point where the average susceptibility $S\left( t \right)$ falls to $1 / R_0$, the effective reproduction number $R_e$ is 1 and the herd immunity threshold is reached.

If all individuals are equally susceptible, $S\left( t \right)$ falls to $1 / R_0 is simply the proportion of individuals who are still susceptible: this is $1 - H$

If susceptiblity varies between individuals,

# The Imperial College model

The basis for the imposition of a lockdown was the well-known paper by Ferguson et al posted on 16 March:

https://www.imperial.ac.uk/media/imperial-college/medicine/sph/ide/gida-fellowships/Imperial-College-COVID19-NPI-modelling-16-03-2020.pdf

>  In such scenarios, given an estimated $R_0$ of 2.4, we predict 81% of the GB and US populations would be infected over the course of the epidemic.

Using the standard formula, this value  of $R_0$ would give a herd immunity threshold of 58% so presumably the IC model was projecting an overshoot, as we would expect in a fast-spreading epidemic.  On this basis, assuming an infection fatality rate of about 1%, they predicted 500,000 deaths in Britain if no measures were taken to mitigate the epidemic.

# The epidemic curve
Early in the COVID-19 epidemic it was noted that in each country where the epidemic took hold, the numbers of cases and deaths peaked a few weeks after the first few cases or deaths, irrespective of what control measures were taken.

The first such analysis was by Michael Levitt, a computational biochemist at Stanford

Levitt 2020b) 1.The_Corona_Chronologies. Part I - China. Part II. Rest_of_World_Michael_Levitt,Stanford_r.pdf
https://www.dropbox.com/s/ovlsfr53horcmsn/1.The_Corona_Chronologies.%20Part%20I%20-
%20China.%20Part%20II.%20Rest_of_World_Michael_Levitt%2CStanford_r.pdf?dl=0
(14-Mar-20 China and World analysis).

Levitt [questioned](https://www.stanforddaily.com/2020/05/04/qa-nobel-laureate-says-covid-19-curve-could-be-naturally-self-flattening/) the conventional estimate of 60-80% for the herd immunity threshold

> You don’t actually have to infect everybody, depending on how fast the virus grows. Some people say 80% [of the population needs to develop antibodies], some people say 60%. I personally think it’s less than 30%. And some people are saying we’re never going to get herd immunity. I don’t think so.

Levitt and colleagues have recently posted an [update](https://www.medrxiv.org/content/10.1101/2020.06.26.20140814v2.full.pdf).

A similar [analysis](https://www.timesofisrael.com/the-end-of-exponential-growth-the-decline-in-the-spread-of-coronavirus/) was published in a newspaper by Isaac Ben-Israel at Tel Aviv University, who has had a leading role in the Israeli military-industrial complex.

> A similar pattern – rapid increase in infections to a peak in the sixth week, and decline from the eighth week – is common everywhere, regardless of response policies

As neither Levitt nor Ben-Israel had any track record in infectious disease modelling, their results were not taken seriously by most biomedical researchers.

# Studies of immunity: antibodies and T cell reactivity
There are two kinds of acquired immune response: antibodies, which are circulating proteins that recognize and bind to antigens; and T cell responses, which recognize antigens only when they are presented on the surface of a cell.

Many viruses are efficiently mopped up by circulating antibodies, but this may not be the case for SARS-CoV-2.  This [paper](https://www.pnas.org/content/117/21/11727) examines how the virus is able to hide its proteins from the immune system while it is circulating. But once a virus infects a cell and starts using the cell's machinery to make viral proteins, it cannot escape surveillance.  The cytosolic system will chop up samples of these proteins into peptide fragments (epitopes) and present them on the surface of the cell to be recognized by T cells.

# Effect of varying susceptibility on the herd immunity threshold
<!---
R Aguas, et al., Herd immunity thresholds for sars-cov-2 estimated from unfolding epidemics.
medRxiv 2020.07.23.20160762 (2020)  First version 24 July

MGM Gomes, et al., Individual variation in susceptibility or exposure to sars-cov-2 lowers the
herd immunity threshold. medRxiv 2020.04.27.20081893 (2020). First version 21 May
29

but Swedish paper cites it dated 12 May and was uploaded on 19 May

Gomes, M. G. M. & Montalbán, A. A SEIR model with variable susceptibility or
296 exposure to infection. arXiv (2020)
--->

If individuals vary in their susceptibility

Gabriela Gomes, a biomathematician at Strathclyde, showed in a preprint first uploaded on n that the standard equation is not even approximately correct for immunity induced by natural infection.

https://arxiv.org/pdf/2008.00098.pdf

This preprint fits her models to real data

https://www.medrxiv.org/content/10.1101/2020.07.23.20160762v1.full.pdf

The standard formula is correct for herd immunity induced by vaccination, but not for herd immunity generated by natural infection, where $R_0$ should be replaced by

\[
H = 1 - \frac{1}{R_0^{\frac{\alpha}{1 + \alpha}}} 
\]

where $\alpha$ is a non-negative number that encodes the narrowness (precision) of the range over which susceptibility varies in the population.  If all individuals have exactly the same susceptiblity and connectivity, the A value of how narrow is the range over which susceptimuch the dthe "precision" (inverse variance)  of the distriuti(inverse variance) of the represents the .   of susceptibility which is scaled to have mean 1. DD

Where susceptibility and infectiousness vary together, the equation is

\[
H = 1 - \frac{1}{R_0^{\frac{\alpha}{1 + \alpha}}} 
\]

This reduces to the standard formula only on the highly implausible assumption that $\alpha = \infty$ (implying that the distribution of susceptibility is a spike).

```{r percentiles, echo=FALSE}

alpha <- 1
k_t <- 1

x <- seq(0, 0.99, by=0.01)
S.x <- qgamma(x, shape=alpha, rate=alpha)

S_t.x <- (alpha / (alpha + k_t))^alpha * qgamma(x, shape=alpha, rate=alpha + kt)

plot(x, S)

```
# Implications of this result  
The Imperial and LSHTM projections for SAGE in March were based on a herd immunity threshold H of 58% (R_0 = 2.5).  Allowing for overshoot of the threshold, assuming an IFR of about 1%, and dismissing the possibility of shielding based on risk stratification they projected 500k deaths unless a lockdown was imposed.

Gomes's derivation shows that the value of H could be much less than 60% if we allow for susceptibility to infection or connectivity to vary between individuals.  She estimates H values of 10-20%, implying that we are already close to herd immunity (based on seroprevalence surveys and additional T cell immunity not detected by antibody tests).   Sunetra Gupta has made the same argument, but without the elegant maths to back it up.

This makes sense of what is now a consistent observation: in every country where the epidemic has taken hold, the number of deaths has peaked about 4 weeks after the first few deaths irrespective of whether there was a lockdown or not, even though the proportion infected is only about 10%.   It also ties in with recent work on pre-existing T cell responses based on cross-reactivity.

It's remarkable that such a simple and fundamental mathematical result should have remained undiscovered by all the people working on modelling of infectious disease over the last 40 years. 
\newpage

### 2020-11-03 pharmaco.Rmd
multiple updates of pharmaco and shielding manuscripts

@@
 ---
-title: 'Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study'
+title: 'Relation of severe COVID-19 to polypharmacy and prescribing of psychotropic drugs: the REACT-SCOT case-control study'
 author: 
@@
   \usepackage{lscape}
+  \usepackage{soul}
 output: 
@@
 
-csl: ../csl/styles/the-new-england-journal-of-medicine.csl
+csl: biomed-central.csl
 urlcolor: red
 
-# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n"); system("make4ht -u pharmaco.tex");
+# rmarkdown::render("pharmaco.Rmd", output_file="pharmacoBMED-D-20-02413.pdf");
+
+# rmarkdown::render("authors_response_BMED-D-20-02413.Rmd"); cat(system("./wc.manuscript.sh"), "\n");
 
@@
 library(kableExtra)
-
 source("helperfunctions.R")
@@
 
-*Objectives* -- To investigate the relation of severe COVID-19 to prior drug prescribing.
+*Background* -- The objective of this study was to investigate the relation of severe COVID-19 to prior drug prescribing.
 
-*Design* -- Matched case-control study (REACT-SCOT) based on record linkage to hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.
-
-*Setting* -- Scottish population.
-
-*Main outcome measure* -- Severe COVID-19, defined by entry to critical care or fatal outcome.
-
-*Participants* -- All `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic, with `r nrow(cc.severe[CASE==0])` controls matched for age, sex and primary care practice.
+*Methods* -- Severe cases were defined by entry to critical care or fatal outcome. 
+For this matched case-control study (REACT-SCOT), all  `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic were matched for age sex and primary care practice to `r nrow(cc.severe[CASE==0])` controls from the population register.
+Records were linked to hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.
 
 *Results* -- Severe COVID-19 was strongly associated with the number of non-cardiovascular drug classes dispensed.
-This association was strongest in those not resident in care homes, in whom the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.notcv.numdrugsgr.care.home[">12", 4]`, and was not accounted for by treatment of conditions designated as conferring increased risk.
-Of 17 drug classes postulated at the start of the epidemic to be "medications compromising COVID", all were associated with increased risk of severe COVID-19.
+This association was strongest in those not resident in care homes, in whom the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.notcv.numdrugsgr.care.home[">12", 4]`, and in those without any of the  conditions designated as conferring increased risk of COVID-19.
+Of 17 drug classes postulated at the start of the epidemic to be "medications compromising COVID", all were associated with increased risk of severe COVID-19 and these associations were present in those without any of the designated risk conditions.
+The fraction of cases in the population attributable to exposure to these drug classes was 38%.
 The largest effect was for antipsychotic agents: rate ratio `r table.laporte["402010.Antipsychotic drugs", 3]`.
@@
 These associations are not easily explained by co-morbidity.
-Although the evidence for causality is not conclusive, these results support existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy as a potential means of reducing COVID-19 risk.
+Measures to reduce the burden of mortality and morbidity from COVID-19 should include reinforcing existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.
 
@@
 
+# Keywords
+COVID-19, pharmacoepidemiology, antipsychotic agents, opioids, gabapentinoids, proton pump inhibitors, polypharmacy, overprescribing.
+
 \newpage
-
+<!---
 ***  
@@
 These results support earlier warnings that these drug classes that these drugs might increase susceptibility to COVID-19, and reinforce existing guidance on reducing overprescribing of these drug classes.
-
 ***
 
+Dr Andrew D Wiese, Vanderbilt University Medical Center, USA. andrew.d.wiese.1@vumc.org
+
+Dr GianLuca Trifiro, Erasmus University Medical Center, Netherlands. g.trifiro@erasmusmc.nl
+
+Dr E Jennifer Edelman, Yale School of Medicine, USA. ejennifer.edelman@yale.edu
+
+Professor Carl Heneghan, University of Oxford, England. carl.heneghan@phc.ox.ac.uk
+
+Professor David Healy: McMaster University, Canada healyd1@mcmaster.ca
+--->
+
 # Background
-In the initial analysis of REACT-SCOT, a matched case control study of  risk factors for severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].
+At the start of the COVID-19 epidemic, possible effects of prior use of medications on the risk of severe disease were widely discussed.
+Some of this discussion focused on drugs suggested to be relevant specifically to SARS-CoV-2, such as drugs acting on the renin-angiotensin system [@aronson_drugs_2020].
+Other discussion focused on drugs associated with increased risk of community-acquired pneumonia.
+A review published on 2 April 2020 listed 17 widely prescribed drug classes associated with increased pneumonia risk and therefore of concern as "medications compromising COVID" [@laporte_midst_2020].
+Excluding immunosuppressive drugs that are criteria for shielding, this list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics, sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, anti-epileptic drugs, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.
+
+In initial analysis of a  matched case control study of risk factors for severe COVID-19 in Scotland (REACT-SCOT), we reported a strong association of severe COVID-19 with dispensing of  at least one prescription in the past year [@mckeigue_rapid_2020].
 The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.
 This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated by public health agencies as conferring vulnerability to COVID-19 (hereafter listed conditions).
-The objective of this study was to investigate this association.
+The objective of this study was to investigate what drug classes underlie this association, and whether any causal effects underlie these associations of severe COVID-19 with dispensing of prescribed drugs.
 
@@
 The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].
-All record linkage studies using NHS data in Scotland, including this one, are governed by the Public Benefit and Privacy Panel for Health and Social Care, which includes patient and public representatives.
+\hl{A matched case-control design was chosen to keep computational requirements within the limits of the resources available.} Record linkage studies using NHS data in Scotland are governed by the Public Benefit and Privacy Panel for Health and Social Care which includes patient and public representatives.
 All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
@@
 Death registrations were obtained from linkage to the National Register of Scotland.
-Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
+Severe or fatal COVID-19 was defined as: (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
 
@@
 In accordance with earlier suggestions that prescribing of multiple cardiovascular drugs, which is supported by evidence-based guidelines, should be considered separately from putatively inappropriate polypharmacy [@appleton_cardiovascular_2014], we partitioned the number of drug classes dispensed into cardiovascular and other drugs.
-To define a prespecified hypothesis about drug classes postulated to increase the risk of severe COVID-19, we used a review by Laporte and Healy published on 2 April 2020 that listed 17 widely prescribed drug classes associated with increased pneumonia risk and therefore of concern as "medications compromising COVID" [@laporte_midst_2020].
-Excluding immunosuppressive drugs that are criteria for shielding, the Laporte-Healy list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics and sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, anti-epileptic drugs, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.
+We then undertook a systematic study of associations will all drugs classes, and subsequently a test of the drug classes postulated to be "medications compromising COVID" [@laporte_midst_2020].
 
@@
 Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
-To distinguish between causality and confounding, we used several approaches:
+To estimate the population attributable fraction -- the fraction of cases in the population that are attributable to the exposure if the association between exposure and disease is causal -- for a group of drugs, we fitted a model including these drug classes together with other covariates, and computed the difference \ensuremath{\eta} between the linear predictors from this model with drug exposures set to their observed values and the linear predictors from the same model with drug exposures set to zero.
+For a rare disease, $e^{\eta} / \left( 1 + e^{\eta} \right)$ approximates the population attributable risk fraction     [@bruzzi_estimating_1985].
+
+\hl{As we had previously shown that care home residence is associated with a rate ratio of more than 20 for severe COVID-19, and the profile of prescribing in care home residents is likely to be different from that in the general population, most analyses of the relation of severe COVID-19 to drug exposure were restricted to those not resident in care homes.}
+
+\hl{To control for confounding by co-morbidity, most analyses of association of severe COVID-19 used restriction to those without any of the designated risk conditions, rather than adjustment for covariates.
+The rationale for this was that electronic health records of drug prescribing and hospital admissions do not provide information on the severity of co-morbid conditions, but do contain enough information to assign the presence or absence of each of the designated risk conditions.
+For some specific drug classes, we were able to define a prespecified list of relevant covariates.}
+
+To distinguish between causality and confounding, we used several further approaches:
 
@@
 
-* Adjusting for prespecified covariates.
-
-* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing.
+* Comparison between recent and less recent time windows of dispensing.
+Because of limits on the computational resources available for this study, the extract of prescription data was limited to prescriptions encashed from 1 July 2019.
+An interval of 240 days, split into two equal windows of 120 days, was chosen so that there was no left-censoring.}
 
@@
 ## Relation of severe COVID-19 to polypharmacy 
-Figure \ref{fig:notcv.numdrugs} shows that the rate of of severe COVID-19 increased steeply with the number of non-cardiovascular drug classes and decreased with the number of noncardiovascular drug classes dispensed.
+Figure \ref{fig:notcv.numdrugs} shows that the rate of of severe COVID-19 increased steeply with the number of non-cardiovascular drug classes and decreased with the number of cardiovascular drug classes dispensed.
 Table \ref{tab:notcv.numdrugs.care.home} shows that this association was restricted to those not resident in care homes.
-Supplementary Table \ref{tab:notcv.numdrugs.listed.any} shows that among those not resident in care homes, the association of severed COVID-19 with number of non-cardiovascular drug classes dispensed was present both in those without any diagnosed listed condition and in those with at least one diagnosed listed condition.
+Among those not resident in care homes, the association of severe COVID-19 with the number of non-cardiovascular drug classes dispensed remained after stratifying by the presence or absence of at least one listed condition (Supplementary Table \ref{tab:notcv.numdrugs.listed.any}.
 
 ## Associations with specific drug classes
-In order to identify specific drug classes associated with severe COVID-19, we examined thse associations among those not resident in a care home and without any listed condition, as most people diagnosed with listed conditions such as asthma and diabetes are on medication for these conditions.
-Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, filtered to show only drug classes with at least 50 exposed individuals and $p < 0.001$.
+For the next step in the analysis, cases and controls were restricted to those not resident in a care home.
+Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, filtered to show only drug classes with at least 50 exposed individuals and \ensuremath{p < 0.001}.
 The drug classes associated with severe COVID-19 include proton pump inhibitors, laxatives, multiple classes of drugs acting on the central nervous system, nutritional supplements, and non-steroidal anti-inflammatory drugs.
 
-Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the Laporte-Healy list in those not resident in a care home.
+Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the prespecified list of "medications compromising COVID" [laporte_midst_2020], restricted to those not resident in a care home.
 All drug classes on this list were associated with increased risk in univariate analyses (though the association with NSAIDS was stronger in Table \ref{tab:subpara.coeffs} where listed conditions were excluded.
-This included all the drug classes listed by Laporte and Healy as having anticholinergic effects: H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs.
+This included all the drug classes listed by Laporte and Healy [@laporte_midst_2020] as having anticholinergic effects likely to increase risk of pneumonia: H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs.
+In a conditional logistic regression model with care home residence as covariate, the fraction of cases of severe COVID-19 attributable to exposure to drug classes on the Laporte-Healy list was estimated as 38%.
 
@@
 Attempting to investigate associations with specific drugs with a hypothesis-free approach is difficult because many of the drug classes that are strongly associated with severe COVID-19, such as proton pump inhibitors, opioids and gabapentinoids, are indicators of overprescribing, recognized as such in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].
-
 To narrow the hypothesis space we tested a pre-specified list of drugs postulated at the start of the epidemic to increase risk of severe COVID-19, based on previously described associations with pneumonia or activity on relevant pathways, especially anticholinergic agents [@laporte_midst_2020].
-We have shown that all the drugs originally listed are associated with increased risk.
-Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.
-
-The strongest association was with antipsychotic drugs.
-As the recognized indications for these drugs are not known to be associated with susceptibility to severe COVID-19, there is no obvious explanation for this association other than causality.
-Further evidence favouring causality is that drugs prescribed for nausea that have similar chemical structures and mode of action to antipsychotic drugs show similar associations with severe COVID-19.
-
-For both opioids and proton pump inhibitors, four criteria provide moderate but not conclusive evidence favouring causal explanations over confounding.
-There are strong dose-response relationships of COVID-19 to dispensed average daily dose.
-Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly.
-The effect sizes are larger in younger individuals and when the analysis is restricted to those without any of the designated risk conditions for COVID-19.
-Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
+We have shown that all the drugs originally listed are associated with increased risk, but the strongest associations are with antipsychotic drugs.
 The dose-response relationship of opioid use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration from 2000-2012 [@edelman_association_2019].
@@
 
+\hl{In observational studies based on electronic health records, distinguishing between causality and confounding as possible explanations for associations of adverse outcomes with drug exposure is challenging because such associations are confounded by the indications for which the drugs were prescribed (co-morbidities).
+Because the severity of these co-morbid conditions cannot usually be assessed accurately using electronic health records, control of confounding based on adjustment for baseline covariates is likely to be inadequate.
+We have used several approaches to overcome this limitation.
+
+First, we have examined associations between drug exposure and severe COVID-19 after restricting to those not diagnosed with designated risk conditions for COVID-19; this makes it unnecessary to adjust for the severity of these co-morbid conditions.
+Where these exposure-outcome associations are stronger in younger than in older individuals, as for opioids and proton pump inhibitors, this favours causality over confounding by co-morbidity as these co-morbid conditions are less common at younger ages.
+
+Second, we have used time-window analyses, in which the comparison is restricted to ever-exposed individuals and the rate ratios associated with use only in a recent time window are compared with rate ratios associated with use only in a non-recent time window.
+Such "self-controlled" designs, which eliminate confounding by time-invariant factors are a standard method in pharmacoepidemiology} [@hallas_use_2014].
+\hl{We have shown that among ever-exposed individuals the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
+
+Third, we have examined dose-response relationships for two drug classes -- opioids and proton pump inhibitors -- where it is possible to standardize dosages across different drugs in the same class.
+We have shown strong dose-response relationships of COVID-19 to dispensed average daily dose.
+Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly.
+
+Fourth, we can examine whether exposure-outcome associations are consistent for pharmacologically similar drugs that are prescribed for different indications.
+We have shown that the associations of severe COVID-19 with phenothiazines and other dopamine antagonists are similar whether these drugs are prescribed as antipsychotic agents or as drugs for nausea and vomiting.
+As these two indications are unrelated and there is no obvious prior hypothesis that they would both predispose to severe COVID-19, this result favours causality over confounding by indication.}
+
 A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.
-Another limitation is that it is not possible to capture hospital prescribing data which includes biologic agents that have immunosuppressive effects.
+Another limitation is that it is not yet possible in Scotland to capture hospital prescribing data which includes biologic agents that have immunosuppressive effects.
 Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10 (rather than the SNOMED-CT codes used in primary care databases), and that drug exposure is based on dispensed rather than issued prescriptions.
+A report from the Swedish national public health agency that compared prior drug prescribing rates in fatal cases of COVID-19 and with prescribing rates in the general population showed a similar pattern of drug classes associated with fatal disease, including antipsychotics, opioids and drugs for gastroesophageal reflux but did not include any control for covariates [@fastboom_lakemedelsanvandningen_2020].
 
 The mechanisms postulated by Laporte and Healy for drugs to increase risk of severe COVID-19 include sedation, respiratory depression, respiratory dyskinesia and anticholinergic effects [@laporte_midst_2020].
-We note that as SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors and other drugs acting on the gastrointestinal tract could increase susceptibility to severe infection.
-It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of drug classes such as proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opioids [@weiner_opioid_2017] has been reported previously.
+Dose-response relationships to risk of community-acquired pneumonia have been reported for opioids [@wiese_opioid_2018] and for antipsychotic drugs [@trifiro_association_2010].
+For proton pump inhibitors, association with risk of community-acquired pneumonia for recent but not long-term use has been reported in a large study [@sarkar_proton-pump_2008].
+However as  SARS-CoV-2 is at least partly an enteric infection [@shi_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors and other drugs acting on the gastrointestinal tract could increase susceptibility to severe COVID-19 even if they do not increase risk of pneumonia caused by other infectious agents.
+In Scotland overprescribing is monitored and to some extent controlled through the National Prescribing Quality Initiative.
+It may be relevant to investigate associations of COVID-19 with drug prescribing in other countries where COVID-19 epidemics have been especially severe and overprescribing of drug classes such as proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opioids [@weiner_opioid_2017] has been reported previously.
 
@@
 Of those drug classes that are not on the Laporte-Healy list, there are strong univariate associations of severe COVID-10 with dispensing of antibiotics, laxatives and nutritional supplements.
-The associations with nutritional supplements is likely to be confounded by overprescribing.
+The associations with nutritional supplements are likely to be confounded by overprescribing.
 For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020; @de_abajo_use_2020] and indeed suggest some protective effect may be possible.
 To explore this more fully will require access to other datasets where measurements of blood pressure and other covariates are available.
-The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
+The relation of prior anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].
 Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.
@@
 Many of the drug classes on this list are recognized indicators of overprescribing.
-The consistency of associations with drugs that have similar modes of action across different groups of indication, the dose-response and time window effects support causal explanations for at least some these associations.
+The consistency of associations with drugs that have similar modes of action across different groups of indication, the dose-response and time window effects support causal explanations for at least some of these associations.
 We recommend that public health agencies should reinforce existing guidelines on avoidance of overprescribing of these drug classes and more generally on inappropriate polypharmacy.
@@
 
-## Data sharing
-Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.
+## Consent to publish
+Not applicable
 
-## Role of the funding source
+## Availability of data and materials
+Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel for Health and Social Care (https://www.informationgovernance.scot.nhs.uk/pbpphsc/).
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
+
+## Competing interests
+HC receives research support and honoraria and is a member of advisory panels or speaker bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
+HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.
+SH received honoraria from Gilead.
+
+## Funding
 This study was undertaken without any external funding.
+
+## Authors' contributions
+All authors read and approved the final manuscript.
 
@@
 This manuscript has been generated directly from the source data by a reproducible research pipeline.
-All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
 
@@
 
-## Conflicts of interest
-HC receives research support and honoraria and is a member of advisory panels or speaker bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
-HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.
-SH received honoraria from Gilead.
-
 ## Acknowledgements
@@
 
-## Public Health Scotland COVID-19 Health Protection Study Group
+### Public Health Scotland COVID-19 Health Protection Study Group
 Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
@@
              caption="Associations of severe COVID-19 with drug classes listed by Laporte and Healy (2020), restricted to those not resident in a care home.
-Adjusted rate ratios are based on a joint model with all drug classes in the table and SIMD quintile as covariates") %>%
+Multivariate rate ratios are based on a joint model with all drug classes in the table and SIMD quintile as covariates") %>%
     column_spec(column=1, width="4cm") %>%
@@
                          "Rate ratio (95\\% CI)", "$p$-value"),
-             caption="Associations of severe COVID-19 with dispensed average daily dose of opioids and proton pump inhibitors, with and without adjustment for prespecified covariates") %>%
+             caption="Associations of severe COVID-19 with dispensed average daily dose of opioids and proton pump inhibitors, with and without adjustment for prespecified covariates.
+For opioids the prespecified covariates are care home residence, SIMD quintile and any history of neoplasm.
+For proton pump inhibitors the prespecified covariates are care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.") %>%
     add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals.
+Multivariate rate ratios are based on a joint model with all drug classes in the table as covariates") %>%
     column_spec(column=1, width="4cm") %>%
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19  in with prescribed drugs in BNF chapter 10 by approved name, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals") %>%
+             caption="Associations of severe COVID-19  in with prescribed drugs in BNF chapter 10 by approved name, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals.
+Multivariate rate ratios are based on a joint model that includes all drugs in the table as covariates") %>%
     column_spec(column=1, width="5cm") %>%

### 2020-11-03 shielding.Rmd
multiple updates of pharmaco and shielding manuscripts

---
title: "Risk of COVID-19 in clinically extremely vulnerable individuals advised to shield during the epidemic: the REACT-SCOT case-control study" 
author:
  - name: Paul M McKeigue
    affiliation: "HPS,Usher"
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow  
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Jim McMenamin  
    affiliation: HPS
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun
    affiliation: "HPS,IGMM,NHSFife"
  - name: Public Health Scotland COVID  -19 Health Protection Study Group
    affiliation: HPS

address: 
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: ISD
  address: Information Services Division, NHS Scotland, UK

header-includes:
  \usepackage{newunicodechar}
  \DeclareUnicodeCharacter{039B}{\ensuremath{\mit\Lambda}}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
  \usepackage[nolists, nomarkers]{endfloat}

always_allow_html: true
urlcolor: blue
output:
      bookdown::pdf_book:  
        base_format: rticles::elsevier_article
        keep_tex: true
        fig_caption: yes

geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"

bibliography: ./covid.bib
csl: ./biomed-central.csl
linenumbers: false
linkcolor: cyan

---

Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group (CLAGS)

Correspondence: Professor Helen Colhoun 07791652665/ helen.colhoun@igmm.ed.ac.uk

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
source("helperfunctions.R")

# rmarkdown::render("shielding.Rmd", output_file="shielding.pdf")

```

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test being recorded on March 1st. Extreme social distancing measures were soon imposed, as in many other countries, with subsequent relaxation now that the effective reproduction number of cases is below 1. The overall policy is one of suppression and if monitoring indicates a second wave of COVID-19, more stringent measures are likely to be be re-imposed.

Early in the epidemic it was realised that certain people would likely be extremely vulnerable to severe COVID-19 disease including those receiving certain immunosuppressant drugs or with other conditions.  Therefore Public Health Scotland  instituted a process of identifying such persons through their primary care physician and through searching of administrative databases  (see appendix for full protocol). Those considered eligible then received a letter detailing the need for self isolation and the support they would receive to achieve this. These letters were sent out over several weeks between xx-and xx.

An initial survey of ~12000 shielding eligible persons indicates that in general some degree of compliance was high but often partial and it varied across different socio-demographic groups.

Here we take the original listing of ~ 180,228 persons deemed eligible for shielding at the outset of the epidemic. We examine the cumulative incidence of any evidence of COVID-19 and any serious COVID-19. using a case control design of all serious COVID-19 cases and age and sex matched controls in Scotland we examine the odds of disease associated with being shielding eligible. We explore whether there is any evidence that shielding reduced risks in these persons.

Using data from the period in which shielding was then stopped ( 1 August 2020) and the next instruction issued to take extra  precautions we examine the number of cases that have occurred.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].

## Ascertainment of eligibility for shielding

To identify  all those nationally deemed eligible for shielding we took the list generated by PHS at the outset of the epidemic (see https://hpspubsrepo.blob.core.windows.net/hps-website/nss/3008/documents/1_covid-19-search-criteria-highest-risk-patients.pdf  for details) .

In brief the following groups were identified

* Solid organ transplant recipients

* People with specific cancers

* People with severe respiratory conditions: cystic fibrosis, severe asthma, severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)

* People with rare diseases

* People on immunosuppression therapies

* People who are pregnant with significant heart disease, congenital or acquired.

* People on renal dialysis and who have had a splenectomy were also added ** to which group?

* A 7th group for primary and secondary care clinicians to identify patients who should be advised to shield.

 As noted in the search criteria document cited above patients were identified via searches of hospital discharge and prescribing records, as well as other sources including regional cancer chemotherapy databases, NHS Blood and Transplant and individual specialist clinician lists.

The list contained the CHI number which enabled linkage to other datasets. it also contained# the reason for eligibility and the date on which the shielding letter was sent to the person concerned.

## COVID-19 case ascertainment and definition
For the total Scottish population "Evidence of  any detected COVID-19" was obtained as having had a positive  nucleic acid test  for SARS-CoV-2 in  the national virology laboratory database,  or having had a hospital discharge code for COVID-19 in those with no positive test, or a death with COVID-19 mention anywhere on the death certificate even where testing had not been  positive.  The databases used were  the Electronic Communication of Surveillance in Scotland (ECOSS) database  that captures all virology testing in all NHS laboratories nationally,  the National Records of Scotland (NRS) death registrations data, the RAPID
database which is a daily update of hospitalisation nationwide and Scottish Morbidity Records-01 which captures hospital discharges  and discharge diagnostic codes nationally.

Databases were queried from March 1st 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to the `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.

All these databases use the  Community Health Index (CHI) identifier. The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland. The CHI database is the master database for the CHI number and  contains several fields including age, sex, residential postcode, and residential care home status.

Among all those with any evidence of detected COVID-19 we then defined severe disease as that requiring critical care or  fatal.   All admissions to intensive (ICU), high dependency (HDU) or combined ICU/HDU care (hereafter critical care unit-CCU) were obtained by linkage using the CHI number  to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal COVID-19 was defined  based on their being a mention of  COVID-19 anywhere  on the death certificate regardless of prior testing or not, OR death within 28 days of testing positive  (or of the admission date for those SMR01 ascertained cases) regardless of whether COVID-19 was mentioned on the death certificate or not. Note that these definitions of COVID-19 deaths are the official ones used by the  National Records of Scotland and Public Health Scotland, respectively.

## Ascertainment of controls
For every case i.e. person with "any evidence of COVID-19" "as defined above, we selected 10 controls matched for sex, one year age band and attending the same GP practice  and alive on the same date of the index case , test, two weeks prior to index case admission or two weeks prior to death for cases ascertained through testing, SMR01 admission data, or NRS death data respectively.  The sampling frame for controls was the Community Health Care Index.

## Demographic data
For all those on the shielding list   and for all COVID-19 cases and controls i)socioeconomic status was evaluated using the postcode-based indicator the Scottish Index of Multiple Deprivation (SIMD) using the residential postcode in the CHI database [@fraser_scottish_2020]. Postcodes were allocated to which quintile of the socioeconomic distribution of Scottish residential datazones they belonged.  (ii) Ethnic group
was captured by linkage to prior SMR01 and SMR00 and SMR04  records   in which the
patients self assigns ethnicity according to the groups that are used in the
national census. iii) Residential care home status for those with diabetes was available via the
care home flag for postal address in the CHI database.

## Other morbidity and drug prescribing
For all those on the shielding list   and for all COVID-19 cases and controls, so as to ascertain other comorbidities the  general hospital admissions data in the  Scottish Morbidity Record 01 (SMR01),  and  all discharge codes over the past five years and Anatomical Therapeutic Chemical codes [@who_anatomical_2016] from Prescribing Information Systems (PIS) data for the past three years were extracted and used to define co-morbid conditions and prior drug exposures .

We first extracted history of a specific list of conditions and drug classes that have been included as moderate risk conditions for COVID-19 by public health agencies hereafter termed "listed conditions" [@nhs_whos_2020]. These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record, or a dispensed prescription in which that diagnostic code is the only indication (see Supplementary Materials for details).

## Statistical methods
\begin{itemize}

\item First we tabulated the shielding list by age-group  sex and reason  for shielding.

\item We then tabulated the \% ( and CI for the \%) of those on the shielding list with "Any evidence of COVID-19" as described above by age and sex   and reason for shielding as well as overall.

\item We also tabulated the \% of those on the shielding list with "Severe i.e. CCU treated or fatal COVID-19 " as described above, by age and sex   and reason for shielding as well as overall.

\item To compare risks in shielded persons before and after shielding was introduced For each shielded person we  marked up their longitudinal record  from Feb 7th with date of shielding letter, date of stop shielding letter (Aug 1st)  and any date of death with the record being censored at data of death. We then computed the number of cases of any evidence of COVID-19  per total  person days in pre shielded time, then shielded time , then post shielded time.  We repeated this for severe cases.

\item Among the shielding to understand whether there are other determinants of becoming a case, we compared the  \% with each of the  other listed conditions,  the \% with more than 1 listed condition, the number of prescribed ATC drug classes, the \% in  residential care home,  the  SIMD  distribution and ethnicity  distribution in those who became cases and those who did not.

\item To provide a comparator for the background population,  we used  the 2019 age band and sex - mid year estimates for the general population and the shielding status in cases to derive  the \% ( and 95\%CI) with any evidence of COVID-19 and severe COVID-19 in the same age and sex groups in the general population  (DAVID this is available from STUART). We then used logistic regression of the one year age band counts of cases and non cases in the shielded and general population to compute and OR associated with shielding.

\item The odds of "exposure "to eligibility for shielding in cases  versus  controls was also  quantified as the conditional odds ratio  from a conditional logistic regression.  We did this because it was easy to extend this regression was then extended to include SIMD, ethnicity, residential care home status,  the "listed "COVID-19 risk conditions.

\item Since the calculation in 4. above can be confounded by overall background trends in the Re to further evaluate if there is any evidence that shielding reduced risk, The odds of "exposure "to eligibility for shielding in cases  versus  controls  BY WEEK of epidemic was also quantified as the conditional odds ratio  from a conditional logistic regression.

\end{itemize}

Analyses were conducted using the R statistical package .

# Results

Table \ref{tab:shieldgroups} shows the frequency of shielding categories. The 6 women pregnant with heart disese were allocated to the "clinician identified" category.

Table \ref{tab:letterdates} shows a breakdown by date of sending letter

Table \ref{tab:afterletter} shows cases among shielded individuals classified by whether estimated date of infection was before or after the shielding letter.

Table \ref{tab:regression.allcases} shows the estimated rate ratios for any diagnosis of COVID-19

Table \ref{tab:regression.severecases} shows the estimated rate ratios for severe COVID-19.

# Discussion

## Summary of findings

## Comparison with previous literature

## Limitations and advantages 
Some key advantages of our study are the population coverage, the range of
outcomes assessed, the extensive prior risk factor and clinical history
data.

## Policy Implications

# Declarations
## Information governance

## Conflicts of interest
The authors declare no conflicts of interest.

## Acknowledgements
We thank all staff in critical care units who submitted data to the SICSAG
database, the Scottish Morbidity Record Data Team, the staff of the National
Register of Scotland, the Public Health Scotland Terminology Services, the HPS
COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology
Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.

## Funding 
There was no specific funder for this study.

## Public Health Scotland COVID-19 Health Protection Study Group
Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew 
Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin
Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo
Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme
Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade
Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$,
Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$,
Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross
Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$,
Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna
Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$

1 Health Protection Scotland (Public Health Scotland), Meridian Court, 5
Cadogan Street, Glasgow G2 6QE.

2 NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2
6QE.

# References
<div id="refs"></div>

# Tables

```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldgroups, 
escape=FALSE, 
             booktabs=TRUE,
			 label="shieldgroups",
             row.names=TRUE,
             caption="Shielded categories") %>%
    kable_styling(latex_options="hold_position")

```

```{r letterdates, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.date.letter,
             escape=FALSE, 
             booktabs=TRUE,
			 label="letterdates",
             row.names=TRUE,
             caption="Dates of sending letters") %>%
    kable_styling(latex_options="hold_position")

```

```{r afterletter, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.after.letter,
escape=FALSE,
             booktabs=TRUE,
			 label="afterletter",
             caption="Proportions of cases advised to shield who were diagnosed > 10 days after date of sending letter")

```

```{r regression.allcases, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shielded.allcases,
escape=FALSE,
             booktabs=TRUE,
			 label="regression.allcases",
   col.names=c(clean.header(colnames(table.shielded.allcases)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
            caption="Rate ratios for diagnosis associated with shielding categories, with and without adjustment for diagnosis > 10 days after sending letter") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
   column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="2cm") %>%
   column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="2cm") %>%
    kable_styling(latex_options="hold_position")

```

```{r regression.severecases, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shielded.severecases,
escape=FALSE,
             booktabs=TRUE,
			 label="regression.severecases",
   col.names=c(clean.header(colnames(table.shielded.severecases)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),

             caption="Rate ratios for severe COVID-19 associated with shielding categories, with and without adjustment for diagnosis > 10 days after sending letter") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
   column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="2cm") %>%
   column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="2cm") %>%
    kable_styling(latex_options="hold_position")

```

### 2021-01-09 authors_response2_BMED-D-20-02413.Rmd
updated shielding paper, fixed trainmodel.R

---
title: Authors' response to 3rd round of comments   on BMED-D-20-02413
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
output: 
  pdf_document

---

# Reviewer 2:

> 1. The authors present data in the newly added tables 1 and 2 and in the supplementary tables which are not completely in accordance with current GDPR. Data enabling identification of individual should be made unavailable.

These tables were reviewed for statistical disclosure control not only by the authors but also by officials of Public Health Scotland, our national public health agency.  We guess that the reviewer's concerns relate to frequency tables containing small numbers. For instance Table 1 shows that one of 483 individuals in Scotland who entered critical care for COVID-19 and survived had been diagnosed with liver disease.  If an attacker already knew the identity of an individual who had been diagnosed with liver disease and had entered critical care for COVID-19 and survived, they would be able to infer a match of this individual to this cell, but this would not provide the attacker with any new information about this individual.

The individual concerned would be able to identify himself/herself as matching this cell in the table, but again this would not reveal any new information to that individual.  The question of self-identification in a table cell was discussed during the statistical disclosure control review and determined to be acceptable

> 2. The authors should comment on the content of the added tables and how the characteristics of the cases may have affected the results. In particular, from Table 1 it may be interpreted that COVID-19/severely ill and in the end fatal cases above a certain age were not admitted to critical care, thus were potentially withheld critical treatment due to triage? The authors should add information in their manuscript whether there were  any phases of restricting access to hospitals and testing that would be relevant for reviewers or readers to know to interpret the results of the study.

For the REACT-SCOT study, we specified in advance a stringent definition of severe COVID-19 so that ascertainment of severe cases would not be affected by triage.  This was based on advice from a senior critical care specialist (Dr Lone, co-author of this study).  As our case definition  included all deaths certified with COVID-19 as underlying cause, triage of individuals who were assessed (on the basis of age and underlying conditions) as unlikely to benefit from critical care would have been ascertained through their death certificates.

> As the authors provide age stratified information in Table 1, age stratified analysis regarding polypharmacy could also be of relevance.  The authors should comment on and provide methodological adjustments how triage decisions could be separated from polypharmacy as being (another) risk factor for fatal outcomes?

We have added a supplementary table showing associations of severe COVID-19 with polypharmacy by age group

> 3. A comparison of the characteristics by residency is missing, meaning comparing characterstics in cases and controls not resident in care homes, and those resident in care homes.

We have added a supplementary table showing associations of care home residency with drug prescribing in the control group.

> 4. The adequacy of the choice of population controls without having been tested positively for COVID-19 should be discussed by the authors.

The design of this study is based on incidence density sampling in which matched controls were sampled from those who had not yet tested positive for COVID-19 on the day that the case was diagnosed. This ensures that the conditional odds ratios are interpretable as rate ratios. For each date that one or more cases were diagnosed, all those in the national population register who had tested positive up to that date were excluded from the sampling of controls. This is correct, on the assumption that those who have already tested positive are not at risk of severe disease.

> 5. The potential for underdiagnosis of CVD and other somatic disorders and unfavourable life-style in people with psychiatric disorders should be discussed as psychotropics were identified as risk factors.

The rate ratio of 4.1 associated with antipsychotic agents is considerably larger than the rate ratios associated with heart disease and most of the other listed conditions.

> 6. Table 4 misses information on SSRIs (BNF 403030)

As the caption states, this table was filtered to show only drug classes for which associations are significant at p<0.001.  We have added the rate ratio for SSRIs to the text

> 7. The authors refer to methods that would be available with the registration of the study in the ENCEPP register, EUPAS35558, Page 5 of the manuscript: "ICD-10 diagnostic and BNF drug codes used to derive these conditions are available with the ENCEPP registration", but the protocol was not available at ENCEPP, but will be 'when the study ends'. These should be provided in the current mansucript as well.

This is provided with the PLOS Medicine paper.

### 2021-01-09 authors_response3_BMED-D-20-02413.Rmd
updated shielding paper, fixed trainmodel.R

--- 2021-01-09 authors_response2_BMED-D-20-02413.Rmd
@@
 
-These tables were reviewed for statistical disclosure control not only by the authors but also by officials of Public Health Scotland, our national public health agency.
+These tables were reviewed for statistical disclosure control by officials of Public Health Scotland, our national public health agency.
 We guess that the reviewer's concerns relate to frequency tables containing small numbers.
 For instance Table 1 shows that one of 483 individuals in Scotland who entered critical care for COVID-19 and survived had been diagnosed with liver disease.
+**This does not enable identification of an individual**.
 If an attacker already knew the identity of an individual who had been diagnosed with liver disease and had entered critical care for COVID-19 and survived, they would be able to infer a match of this individual to this cell, but this would not provide the attacker with any new information about this individual.
@@
 
-For the REACT-SCOT study, we specified in advance a stringent definition of severe COVID-19 so that ascertainment of severe cases would not be affected by triage.
-This was based on advice from a senior critical care specialist (Dr Lone, co-author of this study).
-As our case definition  included all deaths certified with COVID-19 as underlying cause, triage of individuals who were assessed (on the basis of age and underlying conditions) as unlikely to benefit from critical care would have been ascertained through their death certificates.
+The definition of severe COVID-19 used for this study ensures that case ascertainment is not affected by triage of those assessed as unlikely (on the basis of age and underlying conditions) to benefit from critical care.
+As our case definition included all deaths certified with COVID-19 as underlying cause, these cases would have been ascertained through their death certificates.
+
+We have added a sentence (highlighted) to the first paragraph of the Methods section to make this point clear.
 
@@
 
-The design of this study is based on incidence density sampling in which matched controls were sampled from those who had not yet tested positive for COVID-19 on the day that the case was diagnosed.
-This ensures that the conditional odds ratios are interpretable as rate ratios.
+This study was based on incidence density sampling in which matched controls were sampled from those who had not yet tested positive for COVID-19 on the day that the case was diagnosed.
+With this design, the conditional odds ratios are interpretable as rate ratios (ratios of hazard rates) in the population at risk.
+(Greenland S, Thomas DC.
+On the need for the rare disease assumption in case-control studies.
+American Journal of Epidemiology 1982;116:547–553,  https://doi.org/10.1093/oxfordjournals.aje.a113439)
+
 For each date that one or more cases were diagnosed, all those in the national population register who had tested positive up to that date were excluded from the sampling of controls.
-This is correct, on the assumption that those who have already tested positive are not at risk of severe disease.
+This defines the population at risk of severe COVID-19 as all those who have not yet tested positive.
+For clinicians and public health officials this is the relevant definition of the population at risk, as population-wide screening of immunity status has not been conducted.
+
+We have added a sentence (highlighted) to the second paragraph of the Methods section to explain this technical point.
 
@@
 
-The rate ratio of 4.1 associated with antipsychotic agents is considerably larger than the rate ratios associated with heart disease and most of the other listed conditions.
+We have added the following sentence (highlighted) to the Discussion
+
+As the rate ratio of 4.1 associated with use of antipsychotic agents is larger than the rate ratios of about 2 associated with common risk conditons such as Type 2 diabetes and heart disease that we have previously reported from  this case-control study, it is unlikely that underdiagnosis of these risk conditions in users of antipsychotic agents can explain this association.
 
@@
 
-As the caption states, this table was filtered to show only drug classes for which associations are significant at p<0.001.  We have added the rate ratio for SSRIs to the text
+The rate ratio associated with SSRIs is given in Table 5, which includes all the drug classes listed by Laporte and Healy.
+
+SSRIs do not appear in Table 4 because (as the caption states) this table was restricted to those without listed conditions and filtered to show only drug classes for which associations are significant at p<0.001.
 
@@
 
-This is provided with the PLOS Medicine paper.
+The ICD-10 and BNF drug codes are available as a linked attachment on the ENCEPP registration page.  The link is at Section 19 "Other relevant documents":
+
+http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565
+
+This link has been added to the manuscript under "Availability of data and materials", and the ENCEPP registration has been updated to cite the PLOS Medicine paper as the source for the full study protocol.

### 2021-01-09 authors_response_BMED-D-20-02413.Rmd
updated shielding paper, fixed trainmodel.R

---
title: Authors' response to reviewers' comments on BMED-D-20-02413
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
output: 
  pdf_document

---

# Reviewer #1:

> 1.1 The main limitation is the inability to assess the causal effect of polypharmacy on the risk of severe COVID-19. In particular, the observed association between polypharmacy (and several drug classes) and severe COVID-19 could be explained by the underlying disease/clinical frailty.

The assessment of evidence for causality in observational studies is at the core of epidemiologic methods (Bradford Hill criteria).   We have added or rewritten several paragraphs (highlighted) in the Methods and  Discussion sections  to note that it is well recognized that in pharmacoepidemiologic studies based on electronic health records control for confounding based on adjustment for recorded covariates is often inadequate, and that special methods such as self-controlled (time window) analyses have been developed to overcome this.

We have used time window analyses, together with other approaches such as restriction (to those without listed conditions), dose-response analyses, and examining whether associations of disease with pharmacologically similar drugs are consistent across different indications for these drugs (specifically psychosis and nausea/vomiting).

Despite this, we are appropriately cautious: the Conclusion states only that "consistency, ... dose-response and time window effects support causal explanations for at least some of these associations.  The Abstract states only that "These associations are not easily explained by co-morbidity."

> 1.2 The manuscript is confused about diseases used as exclusion criteria and covariates used to adjust the analyses. Did the author use a comorbidity score to adjust the estimates?

We have rewritten a paragraph (highlighted) in the Methods section to make clear the rationale for using restriction (to those without listed conditions) rather than adjustment for covariates (or co-morbidity score) to control for confounding.

> 1.3 I suggest that a better description of the covariates used to adjust the estimates of interest should be given.

We have fixed the caption of each table that shows regression coefficients to list the covariates.  The matched case-control design ensures that all effect estimates are controlled for age, sex and primary care practice.

> 1.4 a table showing the distribution of covariates between cases and controls should be added at least in the supplementary material.

Tables showing these distributions, for all individuals combined and stratified by age group are given in the initial paper on risk stratification which is now in press at PLoS Medicine (copy edited proofs returned, expected to appear very shortly).  We have added a sentence to the Methods section to direct the reader to this.

> 1.5. I didn't understand the sentence "in individuals without conditions designated as conferring high risk".

This sentence seems clear and unambiguous to us, as the "conditions designated as conferring high risk" have been defined earlier. We shall seek the editors' guidance on this.

> 1.6. In the Results section, authors stated "We also excluded those diagnosed with listed conditions". The authors should clearly specify all the inclusion/exclusion criteria in the Methods section.

The list of designated conditions, with  links to the algorithms for assigning the presence of each condition and the R code that implements these algorithms, is given in the PLoS Medicine paper now in press.  We have not repeated material that is already published and available by open access, but we shall be guided by the editors on this.

> 1.7 Did the strategy that was used for testing for coronavirus change during the data collection?

Anyone who entered hospital with a suspected diagnosis of COVID-19 would have been tested during this period.   As explained in the PLoS Medicine paper now in press, the restriction of the case  definition for severe COVID-19 to those who entered critical care within 28 days of a positive test or had COVID-19 as underlying cause on their death certificate ensures that we ascertain all severe or fatal cases, whatever triage procedures may have been used to limit admission to hospital.

> 1.8 Why the authors chose a 240-day period to assess drug prescriptions?

Because of limits on the computational resources available during this emergency, the extract of prescription data was limited to prescriptions encashed from 1 July 2019.  240 days was chosen to give the same look-back period for all cases and controls without left-censoring.  We have added a few sentences sentence to the Methods to document this (highlighted)

> 1.9 A possible explanation of the higher risk associated with the use of antipsychotic drugs is that patients treated with these medicaments could take few precautions to minimize the risk of infection.

We show that elevated risk of severe COVID-19 is associated not only with antipsychotic agents but with dopamine antagonists used for nausea and vomiting that, as LaPorte and Healy pointed out, are pharmacologically similar to these antipsychotic agents.   We note in the Discussion  that this is evidence for causality rather than confounding by indication.  If the association with drugs in this class were attributable to confounding by indication, we would not expect to see similar associations with the outcome for drugs in the same class that are used for entirely different indications.

# Reviewer #2:

> 2.1 the rationale for choices and defintions are spread thoughout the manuscript, e.g. in the statistical section of the method section, result or even in the discussion section.

See response to point 2.5 below.

> 2.2 The authors provide no table of characteristics of the study population or other descriptive elements to get to know the study population, completeness of data or to evaluate the matching.

See response to point 2.5 below.

> 2.3 Although the study was registered at ENCEPP reporting according to STROBE or PE-RECORD could help clarifying many of the issues. Referring to another non peer-reviewed article or even the ENCEPP subnmission during the review process is not sufficient and helpful, but increases workload of the reviewer.

We have provided a STROBE checklist for case-control studies with the initial paper on risk stratification that is now in press in PLoS Medicine.  With this paper we have provided a PE-RECORD checklist for the additional paragraphs relevant to pharmacoepidemiologic studies.

> 2.4 The title is not aligned with the objectives.

The association of severe COVID-19 with psychotropic drugs is a key result that we wish to highlight in the title because it is important to clinical practice and to public health.  Testing this association was not a specific objective when the study was designed, though this association was predefined as a hypothesis by the Laport-Healy list of "medications compromising COVID".

> 2.5 Results: Includes no basic information on how many died or were in critical care, sex and age distribution, etc..  Why does the result section start out with mentioning non-cardiovascular drugs when the topic is polypharmacy and psychotropic drugs?

Tables giving numbers of deaths, numbers in critical care, and a breakdown by sex and age are given in the earlier article on risk factors.  This article is in press in PLoS Medicine and will be published shortly.  We prefer not to burden this paper with material already published and freely available elsewhere, but we shall be guided by the editors on this.

The steps in the analysis are clearly described in the Results section: (1) investigate association with polypharmacy; (2) partition polypharmacy variable into cardiovascular and non-cardiovascular components; (3) hypothesis-free tabulation of associations with non-cardiovascular drugs; (4) test the specific hypothesis of a list of "medications compromising COVID" predefined by others before the start of the study.

As each of these steps was determined by the results obtained in the previous step, we think that it would confuse the reader and would misrepresent how the study was done to describe them in the Methods section as if they had been specified in advance.  We think it is more straightforward and honest to describe how, in this study of a pandemic of a new disease, we adapted the protocol as appropriate to test the hypotheses generated during the study.

> 2.6. Resident care versus non-resident care not mentioned in the methods as a particular focus or covariate.

We have added a paragraph to the Methods (highlighted) to explain the rationale for stratifying by care home residence and for excluding care home residents from subsequent analyses.

> 2.7 Other parts of the results section should be put into the methods section, also the focus on the list  of drug classes of medications compromising COVID. Definition of polypharmacy or inappropriate medication/polypharmacy not provided.

We have moved material into the Methods section where possible, but we think it is still necessary to explain in the Results section how the results at each step determined the next analysis.

> 2.8 How can the authors make sure that the drugs were not proxies for underlying conditions rather than directly related to worse outcomes of COVID-19, confounding by indication?

See response to Reviewer 1 (point 1.1) above on the assessment of causality.

> 2.9 How can the authors conclude based on their findings that overprescribing of these (which ?) was present?

We do not make any judgement about whether overprescribing was present, but only refer to national guidelines in which prescribing of these drug classes is used as an indicator of overprescribing.

> 2.10 Objectives: Not aligned with title or objectives of the abstract. How did the authors assess causal effects of potential associations?

See response to Reviewer 1 (point 1.1) above on the assessment of causality.

> 2.11 Polypharmacy may have a 'inappropriate annotation' but may also simply mean that many drugs used at the same time, and are eventually required in case of multimorbidity beyond CVD. In the method section the authors treat cardiovascular drugs separately, though this may not automatically result in the remainder drugs to be considered inappropriate by definition. Thus defintion of inappropriate is missing here as well.

We have simply followed an earlier study (cited) in partitioning polypharmacy into cardiovascular and non-cardiovascular components as a proxy for appropriate and inappropriate prescribing. We are not in a position to define what is appropriate or inappropriate prescribing.

> 2.12 What was the rationale of investigating doses of selected drugs?

A dose-response relationship is one of the standard (Bradford Hill) criteria for assessing causality of an association between exposure and disease.  For opioids there is a reference method of standardizing doses of different drugs as Milligrammes Morphine Equivalent.  For proton pump inhibitors, which do not have appreciable off-target effects, the Defined Daily Dose can be used to standardize equivalent doses.  For the other classes of drug under study it is not possible to assign equivalent doses, as they have multiple off-target effects for instance anticholinergic effects.

> 2.13 Statistical analysis: Very enlightening section regarding the objectives and design choices. Although it may appear redundant, the information for design choices are needed to be provided in the preceding parts of the method section as well or mainly.

We have added a sentence (highlighted) to the Methods section to note that the matched case-control design (rather than a cohort consisting of the total population) was chosen to keep computational requirements within the limits of the resources available.

> 2.14 Results: Again the result section is missing informaion on the study population, fatal, non-fatal cases and provides reasons for design choices, e.g. exclusion/inclusion of care home residence, stratification, or specific statistical analyses; thus it appears that these choices were made retrospectively kind of 'on the go' rather than planned in the design stage of the study.

The rewritten paragraphs (highlighted) in the Statistical Methods subsection describe how these choices were made.  In studying a pandemic of a new disease, it is not feasible or appropriate to define in advance a rigid protocol for the analysis; for instance the 20-fold rate ratio associated with care home residence, not anticipated at the outset, necessitated  exclusion of care home residents from analyses that would otherwise be confounded by this very large effect size.

We emphasize that the core hypothesis under test: the list of "medications compromising COVID" was proposed by others before the study started.

> 2.15 Discussion: The discussion starts out with numbers/results not highlighted in the result section, proceeds with (finally) defining overprescribing and discussing extensively the hypothesis free (and on the other hand not so hypothesis free) approach and causality (as alluded in the objectives).

We have checked that all quantitative results in the Discussion are mentioned in the text of the the Results section.

We do not "define" overprescribing -- we only cite the National Prescribing Quality Initiative, which uses the levels of prescribing of proton pump inhibitors, opioids and gabapentinoids as indicators of overprescribing.

> 2.16 Though causality may be supported by appropriate methods applied in epidemiological studies and should be discussed, the mentioning of it in the objectives is beyond the scope of the possibilities of the presented study design.

See response to Reviewer #1 (point 1.1) on the assessment of causality in observational pharmacoepidemiological studies.

# Reviewer #3

> 3.1 covariates of the multivariate analysis should be clearly reported

See above -- all these results have been reported in the PLoS Medicine paper which is expected to be published shortly, and is available meanwhile as a preprint on medRxiv.

> 3.2 The values of renal function could mediate the relationship of some drugs and mortality

We do not have measurements of renal function in this dataset, but by excluding older individuals and excluding those diagnosed with chronic kidney disease we are able to minimize effects that would be observed only in those with impaired renal function.

> 3.3 Also body composition, eg. weight and height,  could be important

As noted in the Discussion section, a limitation of the study is that we do not have access to weight and height measurements, which would be recorded only in primary care records.

# Editorial Policies

> Where a mandatory Declarations section is not relevant to your study design or article type, please write "Not applicable" in these sections.

Done

> For the 'Availability of data and materials' section, please provide information about where the data supporting your findings can be found.  We encourage authors to deposit their datasets in publicly available repositories (where available and appropriate), or to be presented within the manuscript and/or additional supporting files.Please note that identifying/confidential patient data should not be shared.  Authors who do not wish to share their data must confirm this under this sub-heading and also provide their reasons.  For further guidance on how to format this section, please refer to BioMed Central's editorial policies page (see links below).

We have provided a link to the Public Benefit and Privacy Panel for Health and Social Care, to which requests for access to the data may be submitted.

### 2021-01-09 beamerkable.Rmd
updated shielding paper, fixed trainmodel.R

---
title: "beamer kable" 
author: PM 
output: 
  beamer_presentation:
    slide_level: 1
    includes:
      in_header: /mnt/phs/beamer_header.tex
    keep_tex: yes
---

# Title slide

# Table with kable

```{r kable, echo=FALSE, warning=FALSE, message=FALSE}

library(knitr)
library(kableExtra)

options(knitr.table.format = "latex") # adds tab: prefix to labels

knitr::kable(mtcars[1:4, 1:4], 
             escape=FALSE,
             booktabs=TRUE,
	label="regressionseverecasesnocare",
	caption="Rate ratios for severe COVID-19 in those not resident in care homes" 
	) %>% 
    add_header_above(c(" "=2, "Univariate"=2))

```
# more stuff

           format="latex", 
             escape=FALSE,
             booktabs=TRUE,
             label="regressionseverecasesnocare",
             caption="Rate ratios for severe COVID-19 in those not resident in care homes. ", 
             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
 #                        "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
#    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
#    column_spec(column=4, width="2.5cm") %>%
#    column_spec(column=5, width="2cm") %>%
#    column_spec(column=6, width="2.5cm") %>%
#    column_spec(column=7, width="2cm") %>%
    kableExtra::group_rows("Shielding eligibility category", 3, 8) %>%
#    kableExtra::group_rows("School-age children in household", 10, 12) %>%
#    kableExtra::group_rows("Adults in household", 13, 16) %>%
#    kableExtra::group_rows("SIMD quintile", 18, 22) %>%
#    kableExtra::group_rows("Occupation", 23, 26) %>%
    kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
    add_footnote(label=c("Percentages are column percentages for each variable",
                         "PF, patient-facing"), notation="none")

### 2021-01-09 coverletter.Rmd
updated shielding paper, fixed trainmodel.R

---
title: "PMEDICINE-D-20-02483: covering letter"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
output: pdf_document

# rmarkdown::render("coverletter.Rmd")

---

# PMEDICINE-D-20-02483R2 Decision - [EMID:48eda5e01d53d77d]

As advised by Pam Berkman we are resubmitting this accepted manuscript for technical editing
We include the PDF, the LaTex source, three figures, and a STROBE checklist.

# Responses to comments from editors

> 8 – Scottish ? national database?

Done

> Abstract – please include p values (here and throughout) for quantifiable data and where 95% Cis are given. In addition, please add summary demographic information to the abstract and the study dates also need to be quoted. Please begin the "Conclusions" subsection of the abstract "In this study, we observed that ... were associated" or similar.

We have done this.  The inclusion of p values, the reformatting of 95% CIs with extra spaces, and the addition of demographic information and study dates have taken the length of the abstract over 500 words.  To allow the submission to proceed we have deleted a few sentences from the version pasted in the online submission.  The full version of the abstract is that in the PDF and the LaTeX source.

> Author Summary, please look at published examples, as your summary needs to be reformatted. It should be in bullet form with 3 headings. This text is subject to editorial change and should be distinct from the scientific abstract. Please see our author guidelines for more information: https://journals.plos.org/plosmedicine/s/revising-your-manuscript#loc-author-summary

Done

> Methods – the mention of the pre specified plan: Please provide a call out to where that can be accessed.

Done - we have given the URL of our github repository

> Please remove “Declarations” on page 11 as these are pulled in automatically from EM in the metadata.

Done

> Please use sections and paragraphs in reporting checklists as page number scan change in formatting and revisions.

> At line 400, is this study available as a preprint that can be cited, for example?

Preprint link given

> In the reference list, where you cite preprints please add "[preprint]".

Done

### 2021-01-09 pharmaco.Rmd
updated shielding paper, fixed trainmodel.R

@@
   - name: Jim McMenamin
-    affiliation: HPS
-  - name: Colin Ramsay
     affiliation: HPS
@@
     md_extensions: -autolink_bare_uris
+always_allow_html: true
 
@@
 
-# rmarkdown::render("pharmaco.Rmd", output_file="pharmacoBMED-D-20-02413.pdf");
-
-# rmarkdown::render("authors_response_BMED-D-20-02413.Rmd"); cat(system("./wc.manuscript.sh"), "\n");
+# rmarkdown::render("pharmaco.Rmd", output_file="pharmacoBMED-D-20-02413_rev3.pdf"); rmarkdown::render("authors_response3_BMED-D-20-02413.Rmd"); cat(system("./wc.manuscript.sh"), "\n");
 
@@
 options(knitr.kable.NA = '')
+options(knitr.table.format = "latex") # adds tab: prefix to labels
 
@@
 *Results* -- Severe COVID-19 was strongly associated with the number of non-cardiovascular drug classes dispensed.
-This association was strongest in those not resident in care homes, in whom the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.notcv.numdrugsgr.care.home[">12", 4]`, and in those without any of the  conditions designated as conferring increased risk of COVID-19.
+This association was strongest in those not resident in a care home, in whom the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.notcv.numdrugsgr.care.home[">12", 4]`, and in those without any of the  conditions designated as conferring increased risk of COVID-19.
 Of 17 drug classes postulated at the start of the epidemic to be "medications compromising COVID", all were associated with increased risk of severe COVID-19 and these associations were present in those without any of the designated risk conditions.
 The fraction of cases in the population attributable to exposure to these drug classes was 38%.
-The largest effect was for antipsychotic agents: rate ratio `r table.laporte["402010.Antipsychotic drugs", 3]`.
+The largest effect was for antipsychotic agents: rate ratio `r table.laporte[grep("Antipsychotic drugs", rownames(table.laporte)), 3]`.
 Other drug classes with large effects included proton pump inhibitors (rate rato `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for >= 2 defined daily doses/day), opioids (`r table.dosegr.opioid["0-74 years: >50", 4]` for >= 50 mg morphine equivalent/day) and gabapentinoids.
@@
 
-In initial analysis of a  matched case control study of risk factors for severe COVID-19 in Scotland (REACT-SCOT), we reported a strong association of severe COVID-19 with dispensing of  at least one prescription in the past year [@mckeigue_rapid_2020].
+In initial analysis of a  matched case control study of risk factors for severe COVID-19 in Scotland (REACT-SCOT), we reported a strong association of severe COVID-19 with dispensing of  at least one prescription in the past year [@mckeigue_rapid_2020-1].
 The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.
@@
 # Methods
-The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].
-\hl{A matched case-control design was chosen to keep computational requirements within the limits of the resources available.} Record linkage studies using NHS data in Scotland are governed by the Public Benefit and Privacy Panel for Health and Social Care which includes patient and public representatives.
+The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+A matched case-control design was chosen to keep computational requirements within the limits of the resources available.
+Record linkage studies using NHS data in Scotland are governed by the Public Benefit and Privacy Panel for Health and Social Care which includes patient and public representatives.
 All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.
@@
 Severe or fatal COVID-19 was defined as: (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.
+This definition of severe COVID-19 used for this study ensures that case ascertainment is not affected by triage of those assessed as unlikely (on the basis of age and underlying conditions) to benefit from critical care.
 
-For each case, the Community Health Index database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.
+For each case, the Community Health Index database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive and had not yet tested positive on the date that the case first tested positive.
+With this incidence density sampling design, the conditional odds ratios are interpretable as rate ratios for severe COVID-19 associated with exposure in the population at risk (defined as all those who have not yet tested positive).
+
 For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.
@@
 As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic obstructive airway disease, chronic kidney disease, disabling neurological disease, liver disease and immunodeficiency or immunosuppression.
-ICD-10 diagnostic and BNF drug codes used to derive these conditions are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).
+ICD-10 diagnostic and BNF drug codes used to derive these conditions are available with the [ENCEPP registration](http://www.encepp.eu/encepp/viewResource.htm?id=35559).
 Socioeconomic status was encoded as the quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).
 
-# Statistical analysis 
+## Statistical analysis 
 Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.
@@
 
-\hl{As we had previously shown that care home residence is associated with a rate ratio of more than 20 for severe COVID-19, and the profile of prescribing in care home residents is likely to be different from that in the general population, most analyses of the relation of severe COVID-19 to drug exposure were restricted to those not resident in care homes.}
+As we had previously shown that care home residence is associated with a rate ratio of more than 20 for severe COVID-19, and the profile of prescribing in care home residents is likely to be different from that in the general population, most analyses of the relation of severe COVID-19 to drug exposure were restricted to those not resident in a care home.
 
-\hl{To control for confounding by co-morbidity, most analyses of association of severe COVID-19 used restriction to those without any of the designated risk conditions, rather than adjustment for covariates.
+To control for confounding by co-morbidity, most analyses of association of severe COVID-19 used restriction to those without any of the designated risk conditions, rather than adjustment for covariates.
 The rationale for this was that electronic health records of drug prescribing and hospital admissions do not provide information on the severity of co-morbid conditions, but do contain enough information to assign the presence or absence of each of the designated risk conditions.
-For some specific drug classes, we were able to define a prespecified list of relevant covariates.}
+For some specific drug classes, we were able to define a prespecified list of relevant covariates.
 
-To distinguish between causality and confounding, we used several further approaches:
+To distinguish between causality and confounding as alternative explanations for associations of severe COVID-19 with drug exposure, we used several further approaches:
 
@@
 Because of limits on the computational resources available for this study, the extract of prescription data was limited to prescriptions encashed from 1 July 2019.
-An interval of 240 days, split into two equal windows of 120 days, was chosen so that there was no left-censoring.}
+An interval of 240 days, split into two equal windows of 120 days, was chosen so that there was no left-censoring.
+The classic case-crossover design, which compares in cases only the frequencies of exposure in recent and nonrecent time windows, can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and nonrecent time windows [@maclure_should_2000].
+The conditional odds ratio can thus be estimated as the ratio of number of cases with recent exposure only to number of cases with nonrecent exposure only.
+The analysis that we report is a refinement of this discordant-pairs ratio estimate that takes advantage of the availability of a matched control group.
+We contrast not the numbers of cases with recent exposure only and nonrecent exposure only, but the conditional odds of case status between those with recent exposure only and those with nonrecent exposure only.
+This controls for any difference in the population frequencies of exposure between recent and nonrecent time windows.
 
 # Results
+Table \ref{tab:casegr} compares severe non-fatal and fatal cases by test-positive and entry to critical care.
+Table \ref{tab:agegr} shows frequencies of risk factors in cases and controls by age group.
+
 ## Relation of severe COVID-19 to polypharmacy 
 Figure \ref{fig:notcv.numdrugs} shows that the rate of of severe COVID-19 increased steeply with the number of non-cardiovascular drug classes and decreased with the number of cardiovascular drug classes dispensed.
-Table \ref{tab:notcv.numdrugs.care.home} shows that this association was restricted to those not resident in care homes.
-Among those not resident in care homes, the association of severe COVID-19 with the number of non-cardiovascular drug classes dispensed remained after stratifying by the presence or absence of at least one listed condition (Supplementary Table \ref{tab:notcv.numdrugs.listed.any}.
+Table \ref{tab:notcv.numdrugs.care.home} shows that this association was restricted to those not resident in a care home.
+Among those not resident in a care home, the association of severe COVID-19 with the number of non-cardiovascular drug classes dispensed remained after stratifying by the presence or absence of at least one listed condition (Supplementary Table \ref{tab:notcv.numdrugs.listed.any}).
+This association was stronger in those aged less than 75 years than in those aged 75 years and over (Supplementary Table \ref{tab:numdrugsagegr}).
 
 ## Associations with specific drug classes
+Supplementary Table \ref{tab:drugscarehome} shows that among controls aged over 75 years dispensed prescriptions of many drug classes were more frequent in care home residents than in those living independently.
 For the next step in the analysis, cases and controls were restricted to those not resident in a care home.
-Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, filtered to show only drug classes with at least 50 exposed individuals and \ensuremath{p < 0.001}.
+Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, filtered to show only drug classes with at least 50 exposed individuals and _p_ < 0.001.
 The drug classes associated with severe COVID-19 include proton pump inhibitors, laxatives, multiple classes of drugs acting on the central nervous system, nutritional supplements, and non-steroidal anti-inflammatory drugs.
 
-Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the prespecified list of "medications compromising COVID" [laporte_midst_2020], restricted to those not resident in a care home.
+Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the prespecified list of "medications compromising COVID" [@laporte_midst_2020], restricted to those not resident in a care home.
 All drug classes on this list were associated with increased risk in univariate analyses (though the association with NSAIDS was stronger in Table \ref{tab:subpara.coeffs} where listed conditions were excluded.
@@
 In a multivariable regression, the strongest independent associations were with proton pump inhibitors, antihistamines, antipsychotic drugs, and opioid analgesics.
-In both univariate and multivariable analyses, the highest rate ratio was that associated with antipsychotic drugs.
+In both univariate and multivariable analyses, the highest rate ratio was that associated with antipsychotic drugs: univariate rate ratio (95% CI) `r table.laporte[grep("Antipsychotic drugs", rownames(table.laporte)), 3]`.
+
 As others have noted [@laporte_midst_2020], the chemical structures and modes of action of drugs used in the treatment of nausea and vertigo overlap with those of antipsychotic drugs.
@@
 Because most users of these drugs had dispensed prescriptions in both time windows, these analyses are based on relatively small numbers: the drug classes shown are restricted to those with at least 500 cases and controls exposed only in the earlier time window .
-For most of these drug classes shown, the rate ratio associated with recent exposure only is above 1, but only for opioid analgesics, and to some extent proton pump inhibitors and anxiolytics, does this association reach conventional levels of statistical significance.
+For all of the drug classes shown, the rate ratio associated with recent exposure only is above 1, but only for opioid analgesics does this association reach stringent levels of statistical significance.
 
@@
 
-\hl{In observational studies based on electronic health records, distinguishing between causality and confounding as possible explanations for associations of adverse outcomes with drug exposure is challenging because such associations are confounded by the indications for which the drugs were prescribed (co-morbidities).
+In observational studies based on electronic health records, distinguishing between causality and confounding as possible explanations for associations of adverse outcomes with drug exposure is challenging because such associations are confounded by the indications for which the drugs were prescribed (co-morbidities).
 Because the severity of these co-morbid conditions cannot usually be assessed accurately using electronic health records, control of confounding based on adjustment for baseline covariates is likely to be inadequate.
@@
 First, we have examined associations between drug exposure and severe COVID-19 after restricting to those not diagnosed with designated risk conditions for COVID-19; this makes it unnecessary to adjust for the severity of these co-morbid conditions.
-Where these exposure-outcome associations are stronger in younger than in older individuals, as for opioids and proton pump inhibitors, this favours causality over confounding by co-morbidity as these co-morbid conditions are less common at younger ages.
+This use of restriction as a technique to assess evidence for causality does not affect the generalizability of the association of severe COVID-19 with polypharmacy, as this association is present in those with and without risk conditions.
+Where exposure-outcome associations are stronger in younger than in older individuals, as for opioids and proton pump inhibitors, this favours causality over confounding by co-morbidity as these co-morbid conditions are less common at younger ages.
 
 Second, we have used time-window analyses, in which the comparison is restricted to ever-exposed individuals and the rate ratios associated with use only in a recent time window are compared with rate ratios associated with use only in a non-recent time window.
-Such "self-controlled" designs, which eliminate confounding by time-invariant factors are a standard method in pharmacoepidemiology} [@hallas_use_2014].
-\hl{We have shown that among ever-exposed individuals the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
+Such "self-controlled" designs, which eliminate confounding by time-invariant factors are a standard method in pharmacoepidemiology [@hallas_use_2014].
+We have shown that among ever-exposed individuals the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.
 
@@
 We have shown that the associations of severe COVID-19 with phenothiazines and other dopamine antagonists are similar whether these drugs are prescribed as antipsychotic agents or as drugs for nausea and vomiting.
-As these two indications are unrelated and there is no obvious prior hypothesis that they would both predispose to severe COVID-19, this result favours causality over confounding by indication.}
+As these two indications are unrelated and there is no obvious prior hypothesis that they would both predispose to severe COVID-19, this result favours causality over confounding by indication.
+As the rate ratio of `r round(as.numeric(gsub("\\(.+", "", table.laporte[grep("Antipsychotic drugs", rownames(table.laporte)), 3])), 1)` associated with use of antipsychotic agents is larger than the rate ratios of about 2 associated with common risk conditons such as Type 2 diabetes and heart disease that we have previously reported from  this case-control study [@mckeigue_rapid_2020-1], it is unlikely that underdiagnosis of these risk conditions in users of antipsychotic agents can explain this association.
 
@@
 
-## Conclusion
-Severe COVID-19 is strongly associated with polypharmacy.
+# Conclusion
+Severe COVID-19 is strongly associated with polypharmacy in those not resident in a care home.
 This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.
@@
 
+# References {-}
+
+<div id="refs"></div>
+
 # Declarations
-## Ethics approval and information governance
+## Ethics approval and consent to participate
 This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.
 Datasets were de-identified before analysis.
-Dissemination of results will be managed by Public Health Scotland.
 
-## Consent to publish
+## Consent for publication
 Not applicable
@@
 All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).
+
+The ICD-10 and BNF drug codes used to define risk conditions are available on the ENCEPP registration page (http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565)
 
@@
 ## Authors' contributions
-All authors read and approved the final manuscript.
+PM, DA, CR, RW, JM, SH and HC provided substantial contributions to the conception or design and drafting of the manuscript.
+NL, ASP, DG, and JM provided substantial contributions to the data acquisition.
+SK, AW, JB, SM, and BG provided substantial contributions to the data analysis or interpretation.
+All authors contributed to revising the manuscript critically for important intellectual content and approved the final manuscript.
 
-## Transparency declaration
+# Acknowledgements
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
+
+## Public Health Scotland COVID-19 Health Protection Study Group
+Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
+
+1  Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+
+2  NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+
+<!---## Transparency declaration
 PM, as the manuscript's guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
@@
 
-## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
+--->
+# Supplementary information
 
-### Public Health Scotland COVID-19 Health Protection Study Group
-Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
+## Table \ref{tab:notcv.numdrugs.listed.any} 
+Association of severe COVID-19 with number of non-cardiovascular drug classes dispensed, in those not resident in a care home by diagnosis of at least one listed condition.
 
-1  Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+## Table \ref{tab:numdrugsagegr} 
+Association of severe COVID-19 with number of non-cardiovascular drug classes dispensed, in those not resident in a care home by age group .
 
-2  NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
+## Table \ref{tab:drugscarehome} 
+Univariate conditional odds ratios for dispensing of BNF subparagraph codes associated with care home status in controls aged 75 years, filtered to retain rows with at least 50 exposed care home residents.
 
-## Supplementary material
-The R and Rmarkdown scripts used to generate this article, which include the code used to derive variables, will be made available when the paper is published.
+## Table \ref{tab:antipsych} 
+Univariate associations of severe COVID-19 with drugs in four subparagraphs of BNF chapter 4, in those not resident in a care home and not treated for cancer in last year, filtered to retain rows with at least 20 exposed individuals.
 
-# References {-}
+## Table \ref{tab:timewindow}  
+Comparison of associations with exposure only in the most recent 120-day time window with exposure only in the previous 120-day time window (baseline category for rate ratios) for drug classes on the Laporte-Healy list with at least 500 cases and controls exposed only in the earlier time window.
 
-<div id="refs"></div>
+## Table \ref{tab:dosegr}  
+Associations of severe COVID-19 with dispensed average daily dose of opioids and proton pump inhibitors, with and without adjustment for prespecified covariates.
+For opioids the prespecified covariates are care home residence, SIMD quintile and any history of neoplasm.
+For proton pump inhibitors the prespecified covariates are care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.
+
+## Table \ref{tab:bnfchapter2} 
+Associations of severe COVID-19 with prescribed drugs in BNF chapter 2 by subparagraph code, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals.
+Multivariate rate ratios are based on a joint model with all drug classes in the table as covariates.
+
+## Table \ref{tab:bnfchapter10} 
+Associations of severe COVID-19  with prescribed drugs in BNF chapter 10 by approved name, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals.
+Multivariate rate ratios are based on a joint model that includes all drugs in the table as covariates.
 
 \newpage
+# Figures
 
@@
 }
+
+```
+\newpage
+
+```{r casegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.casegr,
+             escape=FALSE, 
+             booktabs=TRUE,
+			 longtable=TRUE, 
+             label="casegr",
+             col.names=paste(c("Critical care, non-fatal", "Critical care, fatal",
+                               "No critical care, fatal", "No critical care, fatal"),
+                             paste0(rep("(", 4),
+                                    with(cc.severe, table(severe.casegr)), rep(")", 4))), 
+             caption="Comparison of severe non-fatal and fatal cases, by test-positive status and entry to critical care", 
+             align=c("l", rep("c", 4))) %>%
+    add_header_above(c(" "=1, "Test-positive"=3, "No positive test"=1)) %>%
+    add_header_above(c(" "=1, "Non-fatal"=1, "Fatal"=3)) %>%
+    column_spec(column=2:5, width="3cm") %>%
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
+
+```
+
+\newpage
+
+```{r agegr, echo=FALSE, warning=FALSE, message=FALSE}
+
+rownames(table.agegr) <- replace.names(rownames(table.agegr))
+table.agegr.print <- table.agegr[c(1:4, 6, 8:nrow(table.agegr)), ] 
+knitr::kable(table.agegr.print, 
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="agegr",
+			 linesep=c(rep('', 4), '\\addlinespace', rep('', 9)), 
+             caption="Frequencies of risk factors in cases and controls, by age group",
+             align=rep("c", 8)) %>%
+    column_spec(column=1, width="4cm") %>%
+    column_spec(column=2:3, width="1.6cm") %>%
+    column_spec(column=4:7, width="1.8cm") %>%
+    column_spec(column=8:9, width="2cm") %>%
+    add_header_above(c(" "=1, "0-39 years"=2, "40-59 years"=2,
+                       "60-74 years"=2, "75+ years"=2)) %>%
+    landscape() %>%
+    kable_styling(latex_options="hold_position")
 
@@
             "Rate ratio", "$p$-value"),
-             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those without a listed condition and not resident in a care home, filtered to show only drug classes with at least 50 exposed individuals and  $p < 0.001$: ")) %>%
+             caption="Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those without a listed condition and not resident in a care home, filtered to show only drug classes with at least 50 exposed individuals and  $p$ < 0.001") %>%
     column_spec(column=1, width="6cm") %>%
@@
 
-freqs <- with(cc.severe[nocare], table(listed.any, CASE))
+freqs <- with(cc.severe[care.home=="Independent"], table(listed.any, CASE))
 
@@
              row.names=FALSE,
-             col.names=c("Number of drug classes", colnames(table.notcv.numdrugsgr.listed.any)[2:3], 
+             col.names=c("Number of drug classes",
+                         colnames(table.notcv.numdrugsgr.listed.any)[2:3], 
                          "Rate ratio (95\\% CI)", "$p$-value"),
@@
 \newpage
+
+```{r numdrugsgr.agegr2, echo=FALSE, warning=FALSE, message=FALSE}
+
+freqs <- with(cc.severe[care.home=="Independent"], table(agegr2, CASE))
+
+colnames(table.numdrugsgr.agegr2)[1] <- "Number of drug classes"
+colnames(table.numdrugsgr.agegr2)[4:5] <- c("Rate ratio (95\\% CI)", "$p$-value")
+
+knitr::kable(table.numdrugsgr.agegr2,
+             escape=FALSE, 
+             booktabs=TRUE,
+             row.names=FALSE,
+             label="numdrugsagegr",
+             align=c("l", "r", "r", "c", "c"), 
+             caption="Association of severe COVID-19 with number of non-cardiovascular drug classes dispensed, by age group in those not resident in a care home") %>%
+    column_spec(column=1, width="2cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="3.5cm") %>%
+    column_spec(column=5, width="2cm") %>%
+    kableExtra::group_rows(paste0("Age 0-74 years (",
+                      freqs[1, 1], " controls, ", freqs[1, 2], " cases)"),
+               1, 6) %>%
+        kableExtra::group_rows(paste0("Age 75 years and over (",
+                                      freqs[2, 1], " controls, ", freqs[2, 2], " cases)"),
+               7, 12) %>%
+     kable_styling(latex_options="hold_position")
+
+```
+
+```{r carehometable, echo=FALSE, warning=FALSE, message=FALSE}
+
+colnames(table.drugs.carehome) <- c(clean.header(colnames(table.drugs.carehome)[1:2]),
+                                    "Odds ratio (95\\% CI)", "$p$-value")
+colnames(table.drugs.carehome) <- gsub("X0", "Independent",
+                                       colnames(table.drugs.carehome))
+colnames(table.drugs.carehome) <- gsub("X1", "Care home resident",
+                                       colnames(table.drugs.carehome))
+rownames(table.drugs.carehome) <- gsub("\\.", " ", rownames(table.drugs.carehome))
+
+newch <- diff(as.integer(substr(rownames(table.drugs.carehome), 1, 1))) != 0
+linesep <- ifelse(newch, '\\addlinespace', '')
+
+knitr::kable(table.drugs.carehome,
+             escape=FALSE, 
+             booktabs=TRUE,
+             label="drugscarehome",
+             linesep=linesep,
+             align=rep("c", 4), 
+             caption="Univariate conditional odds ratios for dispensing of BNF subparagraph codes associated with care home status in controls aged 75 years, filtered to retain rows with at least 50 exposed care home residents") %>%
+    column_spec(column=1, width="5.5cm") %>%
+    column_spec(column=2, width="2.25cm") %>%
+    column_spec(column=3, width="2.25cm") %>%
+    column_spec(column=4, width="3cm") %>%
+    kable_styling(latex_options="hold_position")
+
+```
 
@@
              align=rep("c", 6), 
-             caption="Associations of severe COVID-19  in with prescribed drugs in BNF chapter 10 by approved name, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals.
+             caption="Associations of severe COVID-19 with prescribed drugs in BNF chapter 10 by approved name, restricted to those not resident in a care home and filtered to retain rows with at least 50 exposed individuals.
 Multivariate rate ratios are based on a joint model that includes all drugs in the table as covariates") %>%
@@
     landscape()
-``````
+```

### 2021-01-09 shielding.Rmd
updated shielding paper, fixed trainmodel.R

@@
 ---
-title: "Risk of COVID-19 in clinically extremely vulnerable individuals advised to shield during the epidemic: the REACT-SCOT case-control study" 
+title: "Relation of severe COVID-19 to household composition, recent hospital exposure, and advice to shield in clinically vulnerable and low-risk individuals in Scotland: REACT-SCOT  case-control studyDeterminants of rates of serious hypoglycaemic events in type 1 diabetes"
 author:
@@
     affiliation: "HPS,Usher"
+  - name: David McAllister 
+    affiliation: Glasgow 
   - name: David Caldwell 
@@
     affiliation: HPS
+  - name: Sam Colville
+    affiliation: HPShttps://t.co/ro5FIUsaWr?amp=1
   - name: Sharon Hutchinson 
     affiliation: GCU
-  - name: David McAllister 
-    affiliation: Glasgow  
   - name: Chris Robertson 
@@
     affiliation: Usher
-  - name: Janet Murray
-    affiliation: HPS 
   - name: Jim McMenamin 
-    affiliation: HPS
-  - name: Alison Smith-Palmer 
     affiliation: HPS
@@
   - name: Helen M Colhoun
+    email: helen.colhoun@igmm.ed.ac.uk (07791 652665)
     affiliation: "HPS,IGMM,NHSFife"
-  - name: Public Health Scotland COVID  -19 Health Protection Study Group
-    affiliation: HPS
 
 address: 
-- code: IGMM
-  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland.
-HC - AXA Chair in Medical Informatics and Epidemiology
 - code: Usher
@@
 NL - Clinical Senior Lecturer in Critical Care
+- code: IGMM
+  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland.
+HC - AXA Chair in Medical Informatics and Epidemiology
 - code: HPS
-  address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
 - code: Glasgow
@@
   \usepackage{newunicodechar}
-  \DeclareUnicodeCharacter{039B}{\ensuremath{\mit\Lambda}}
   \let\origquote\quote
@@
   \usepackage{float}
+  \floatplacement{figure}{H}
   \usepackage{array}
@@
   \usepackage{pdfpages}
-  \usepackage[nolists, nomarkers]{endfloat}
-
+output:
+  rticles::nejm_article:
+    includes:
+      in_header: ./preamble.tex
+    latex_engine: lualatex
+    keep_tex: true
+    fig_caption: yes
+#    number_sections: false
+    md_extensions: -autolink_bare_uris
+geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
+bibliography: ./covid.bib
+csl: ./biomed-central.csl
 always_allow_html: true
 urlcolor: blue
-output:
-      bookdown::pdf_book:  
-        base_format: rticles::elsevier_article
-        keep_tex: true
-        fig_caption: yes
-
-geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
-
-bibliography: ./covid.bib
-csl: ./biomed-central.csl
 linenumbers: false
@@
 
-Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group (CLAGS)
+On behalf of Public Health Scotland COVID-19 Health Protection Study Group
 
-Correspondence: Professor Helen Colhoun 07791652665/ helen.colhoun@igmm.ed.ac.uk
+Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group
 
@@
 library(kableExtra)
-library(bookdown)
 
@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(fig.width=6, fig.pos = "H", out.extra = "")
 source("helperfunctions.R")
@@
 
+dateformat <- function(x) { # format date and strip leading zero
+     gsub("^0", "", format(x, "%d %B %Y"))
+}
+
+format.estcipv <- function(x.ci, x.pval) {
+    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
+    x <- gsub(", ", " to ", x)
+    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
+    x <- gsub("times", "\\\\times", x)
+    return(x)
+}
+
+peakdate <- casedates.allgr[which.max(avdaily), date]
+maxdeaths <- casedates.allgr[which.max(avdaily), avdaily]
+deaths.quartermax.date <- casedates.allgr[date > peakdate & avdaily < 0.25 * maxdeaths,
+                                 date][1]
+
+peakdate.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), date]
+maxdeaths.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), avdaily]
+deaths.elig.quartermax.date <- casedates.byelig[listedgr3=="Eligible for shielding" &
+                                         date > peakdate.elig &
+                                         avdaily < 0.25 * maxdeaths.elig,
+                                         date][1]
+
 ```
 
+ \newpage
+
+# Abstract
+* _Objectives_ - to investigate the relation of severe COVID-19 to factors influencing exposure to SARS-CoV-2 and the implications of this for protecting vulnerable individuals through shielding and targeted vaccination.
+
+* _Design_ - (1) cohort study of all individuals identified as eligible for shielding; (2) matched case-control study (REACT-SCOT).
+
+* _Setting_ - population of Scotland from 1 March 2020 to 15 October 2020
+
+* _Participants_ - The cohort study comprised `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland.
+The case-control study included  all `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice.
+
+* _Interventions_ - Letters advising eligible individuals to shield were sent by Public Health Scotland from 3 April 2020 onwards.
+
+* _Main outcome measure_ - severe COVID-19 defined as cases that entered critical care or were  fatal.
+
+* _Results_ - `r nrow(shielded.full[!is.na(casegr2) & casegr2=="Severe"])` individuals eligible for shielding developed severe COVID-19.
+Although the incidence of severe disease fell from early April 2020 onwards in the Scottish population, the rate ratio associated with eligibility for shielding increased.
+With those without risk conditions as reference category, the univariate rate ratio for severe disease was
+`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
+in those with moderate risk conditions and
+`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
+in those eligible for shielding.
+The highest rate ratio for severe disease was `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` in solid organ transplant recipients.
+In all risk groups the strongest risk factors (in terms of deviance explained) for severe COVID-19 were recent hospital exposure with a rate ratio of `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 4])` and number of adults in the household, with a rate ratio of  `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 3], table.listedgr3.severecases.nocare["adultsgt1", 4])` for two or more adults compared with one.
+
+* _Conclusions_ -  Protection of vulnerable individuals requires control of transmission in hospital settings and prevention of transmission from other adults in the household.
+For solid organ transplant recipients, in whom vaccination is likely to be ineffective, vaccination of others in the household should not be ruled out as a means of reducing transmission.
+
 # Background  
-The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test being recorded on March 1st.
-Extreme social distancing measures were soon imposed, as in many other countries, with subsequent relaxation now that the effective reproduction number of cases is below 1.
-The overall policy is one of suppression and if monitoring indicates a second wave of COVID-19, more stringent measures are likely to be be re-imposed.
+The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020.
+Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.
+Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions and "diseases and conditions considered to be very high risk".
+Those with "very high risk" conditions were designated as clinically extremely vulnerable and eligible for shielding.
 
-Early in the epidemic it was realised that certain people would likely be extremely vulnerable to severe COVID-19 disease including those receiving certain immunosuppressant drugs or with other conditions.
-Therefore Public Health Scotland  instituted a process of identifying such persons through their primary care physician and through searching of administrative databases  (see appendix for full protocol).
-Those considered eligible then received a letter detailing the need for self isolation and the support they would receive to achieve this.
-These letters were sent out over several weeks between xx-and xx.
+The practicality of shielding vulnerable individuals, in the context of a population-wide lockdown where contacts have been reduced to a low level, depends on the extent to which their exposure to likely sources of infection either directly or through other household members can be reduced.
+It is now possble in Scotland to study this through record linkage to household composition, recent hospital exposure, and with relevant occupations including health care [@shah_risk_2020] and teaching [@mcallister_report_2020].
 
-An initial survey of ~12000 shielding eligible persons indicates that in general some degree of compliance was high but often partial and it varied across different socio-demographic groups.
-
-Here we take the original listing of ~ 180,228 persons deemed eligible for shielding at the outset of the epidemic.
-We examine the cumulative incidence of any evidence of COVID-19 and any serious COVID-19.
-using a case control design of all serious COVID-19 cases and age and sex matched controls in Scotland we examine the odds of disease associated with being shielding eligible.
-We explore whether there is any evidence that shielding reduced risks in these persons.
-
-Using data from the period in which shielding was then stopped ( 1 August 2020) and the next instruction issued to take extra  precautions we examine the number of cases that have occurred.
+The objectives of this study were to investigate the relation of severe COVID-19 to factors influencing exposure to SARS-CoV-2 and to examine the implications of this for protecting vulnerable individuals through shielding and targeted vaccination.
 
 # Methods
-The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+## Ascertainment of eligibility for shielding 
+On 21 March 2020 a [list](https://www.england.nhs.uk/coronavirus/wp-content/uploads/sites/52/2020/03/200321_GP-letter_Annex-2_FINAL_2.pdf) of "diseases and conditions considered to be very high risk" was issued by NHS England.
+Six categories were initially designated as eligible for shielding [@health_protection_scotland_search_2020]:
 
-## Ascertainment of eligibility for shielding
+1. Solid organ transplant recipients
 
-To identify  all those nationally deemed eligible for shielding we took the list generated by PHS at the outset of the epidemic (see https://hpspubsrepo.blob.core.windows.net/hps-website/nss/3008/documents/1_covid-19-search-criteria-highest-risk-patients.pdf  for details) .
+2. People with cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system
 
-In brief the following groups were identified
+3. People with severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
 
-* Solid organ transplant recipients
+4. People with rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease.
 
-* People with specific cancers
+5. People on immunosuppression therapies sufficient to increase risk of infection
 
-* People with severe respiratory conditions: cystic fibrosis, severe asthma, severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
+6. Women who were pregnant with heart disease
 
-* People with rare diseases
+An additional category comprised patients identified by clinicians as requiring shielding advice.
+On 11 May 2020 the list of eligible conditions was updated to include those on renal dialysis and those who had had a splenectomy.
 
-* People on immunosuppression therapies
+Public Health Scotland (PHS) began identifying patients in this category through health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries, and direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].
+Those identified as eligible were sent a letter advising them to self-isolate and offering support.
+The first batch of letters was sent on `r dateformat(batches$Date.Sent)[1]`.
 
-* People who are pregnant with significant heart disease, congenital or acquired.
+The cohort of `r nrow(shielded.full)` individuals comprised the list generated by Public Health Scotland at the outset of the epidemic, updated with those persons newly identified as eligible up to the end of July.
+Letters were sent on Fridays, beginning on `r dateformat(batches$Date.Sent[1])` an throughout the first wave until `r  dateformat(max(shielded.full$Date.Sent, na.rm=TRUE))` though advice to shield was rescinded on 1 August 2020.
 
-* People on renal dialysis and who have had a splenectomy were also added ** to which group?
-
-* A 7th group for primary and secondary care clinicians to identify patients who should be advised to shield.
-
- As noted in the search criteria document cited above patients were identified via searches of hospital discharge and prescribing records, as well as other sources including regional cancer chemotherapy databases, NHS Blood and Transplant and individual specialist clinician lists.
-
-The list contained the CHI number which enabled linkage to other datasets. it also contained# the reason for eligibility and the date on which the shielding letter was sent to the person concerned.
+Those identified as eligible were advised to "work from home if possible" and that employers should "support you to do this2 or "make sure suitable arrangements are in place so you can go to work."  Help was offered with deliveries of food and medicine, but no support was offered for others resident in the household eitheir to isolate with the vulnerable individual or to move out temporarily.
 
 ## COVID-19 case ascertainment and definition
-For the total Scottish population "Evidence of  any detected COVID-19" was obtained as having had a positive  nucleic acid test  for SARS-CoV-2 in  the national virology laboratory database,  or having had a hospital discharge code for COVID-19 in those with no positive test, or a death with COVID-19 mention anywhere on the death certificate even where testing had not been  positive.
-The databases used were  the Electronic Communication of Surveillance in Scotland (ECOSS) database  that captures all virology testing in all NHS laboratories nationally,  the National Records of Scotland (NRS) death registrations data, the RAPID
-database which is a daily update of hospitalisation nationwide and Scottish Morbidity Records-01 which captures hospital discharges  and discharge diagnostic codes nationally.
+Case ascertainment has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS  that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general /acute hospital discharges including day cases.
+All these databases use the Community Health Index (CHI) number as identifier.
+The CHI database includes age, sex, postcode and care home status, and can be queried to extract numbers of adults and children in the household.
 
-Databases were queried from March 1st 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to the `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.
+Cases of COVID-19 were ascertained as those with a positive  nucleic acid test for SARS-CoV-2 in ECOSS, a hospital discharge diagnosis of COVID-19 in SMR01, or a death registration with mention of COVID-19 anywhere on the death certificate.
+The presentation date was assigned as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those without a positive test result ascertained through hospital discharge records, and as fourteen days before the date of death for those without a positive test result ascertained through death certificates.
+Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.
 
-All these databases use the  Community Health Index (CHI) identifier.
-The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.
-The CHI database is the master database for the CHI number and  contains several fields including age, sex, residential postcode, and residential care home status.
+Severe COVID-19 was defined by entry to critical care within 21 days of presentation date, or fatal outcome.
+Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
+Fatal outcome was defined as death with  COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive, or death within 28 days of presentation date for cases ascertained through discharge records.
+This definition of severe cases was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
 
-Among all those with any evidence of detected COVID-19 we then defined severe disease as that requiring critical care or  fatal.
-All admissions to intensive (ICU), high dependency (HDU) or combined ICU/HDU care (hereafter critical care unit-CCU) were obtained by linkage using the CHI number  to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
-Fatal COVID-19 was defined  based on their being a mention of  COVID-19 anywhere  on the death certificate regardless of prior testing or not, OR death within 28 days of testing positive  (or of the admission date for those SMR01 ascertained cases) regardless of whether COVID-19 was mentioned on the death certificate or not.
-Note that these definitions of COVID-19 deaths are the official ones used by the  National Records of Scotland and Public Health Scotland, respectively.
+In cases with fatal outcome, median time from presentation to death was `r with(cc.all[fatalcase==1], median(as.integer(Date.Death -SPECIMENDATE)))` days.
+The proportion of cases with fatal outcome was calculated over sliding seven-day time windows.
+We denote this fraction as the fatal case proportion, rather than the case fatality ratio which is classically defined without linkage between diagnoses and deaths, as the ratio of deaths from the disease to diagnosed cases over a specified time window.
+
+As those who were notified as having died were removed from the list of those eligible for shielding before letters were sent out, and no records of this filtering were available, we examined the fatal case proportion by presentation date.
+The rationale for this was that as filtering to remove those who had died would misallocate fatal cases from those eligible for shielding to those ineligible, we could estimate the date up to which this misallocation occurred by examining the fatal case proportion by presentation date.
 
 ## Ascertainment of controls
-For every case i.e.
-person with "any evidence of COVID-19" "as defined above, we selected 10 controls matched for sex, one year age band and attending the same GP practice  and alive on the same date of the index case , test, two weeks prior to index case admission or two weeks prior to death for cases ascertained through testing, SMR01 admission data, or NRS death data respectively.
-The sampling frame for controls was the Community Health Care Index.
+As linkage to demographic and morbidity data was not available for the full shielding cohort.
+these associations were examined in the REACT-SCOT case-control study.The design of this study has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+For each case of COVID-19, up to 10 controls matched for sex, one-year age band and primary care practice were selected from the CHI database, and assigned the same presentation date as the case.
+With this incidence density sampling design, it is possible and correct for an individual to appear more than once as a control and subsequently as a case.
 
-## Demographic data
-For all those on the shielding list   and for all COVID-19 cases and controls i)socioeconomic status was evaluated using the postcode-based indicator the Scottish Index of Multiple Deprivation (SIMD) using the residential postcode in the CHI database [@fraser_scottish_2020].
-Postcodes were allocated to which quintile of the socioeconomic distribution of Scottish residential datazones they belonged.
-(ii) Ethnic group
-was captured by linkage to prior SMR01 and SMR00 and SMR04  records   in which the
-patients self assigns ethnicity according to the groups that are used in the
-national census. iii) Residential care home status for those with diabetes was available via the
-care home flag for postal address in the CHI database.
+## Linkage of cases and controls to demographic and morbidity data
+Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].
+Quintile 5 is the least deprived.
+Care home residence was assigned using the field in the CHI database.
+As this does not capture all care homes, those aged over 70 in households with 10 or more adults were also coded as care home residents.
+Cases and controls were linked to hospital discharge codes over the last five years in SMR01, to BNF codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.
+As described previously  [@mckeigue_rapid_2020-1], we used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020].
+These are referred to as "listed conditions".
+Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
 
-## Other morbidity and drug prescribing
-For all those on the shielding list   and for all COVID-19 cases and controls, so as to ascertain other comorbidities the  general hospital admissions data in the  Scottish Morbidity Record 01 (SMR01),  and  all discharge codes over the past five years and Anatomical Therapeutic Chemical codes [@who_anatomical_2016] from Prescribing Information Systems (PIS) data for the past three years were extracted and used to define co-morbid conditions and prior drug exposures .
-
-We first extracted history of a specific list of conditions and drug classes that have been included as moderate risk conditions for COVID-19 by public health agencies hereafter termed "listed conditions" [@nhs_whos_2020].
-These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record, or a dispensed prescription in which that diagnostic code is the only indication (see Supplementary Materials for details).
+To examine possible associations with exposure to health-care facilities, we defined the variable "recent hospital visit/stay" for cases and controls to include those individuals who had been hospital in-patients or attended hospital out-patient clinics at any time in the 14 days before presentation date, excluding admissions or outpatient attendances less than five days before presentation date.
+<!--- check whether outpatient attendances include online consultations --->  This exclusion was intended to exclude admissions caused by COVID-19 but to include anyone exposed to health care facilities in the time interval during which the infection is likely to have been acquired.
+This is broader than the definition used in Scotland and England for "probable hospital-acquired infection" which is restricted to cases who had been hospital inpatients for at least seven days before they first tested positive.
+In contrast to this restrictive definition of probable hospital-acquired infection which identifies individuals who are unlikely to have been infected outside hospital, this inclusive definition of "recent hospital visit/stay" allows us to calculate the fraction of cases attributable to this exposure.
 
 ## Statistical methods
-\begin{itemize}
+The case-control and time series analyses focus on severe COVID-19 as this case definition does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is also likely to differ between those eligible for shielding and those ineligible.
 
-\item First we tabulated the shielding list by age-group  sex and reason  for shielding.
+For the full shielding cohort, we tabulated each category of eligibility by age group, outcome and date of sending advice letter.
+2-day or 3-day time windows of numbers of severe cases were calculated for the study period.
+To examine the effect of advice to shield, the case-control dataset was split into two intervals at a time point fourteen days after the first batch of shielding advice letters were sent on 3 April 2020:  presentation dates up to 17 April 2020 were coded as pre-advice, and presentation dates after that as post-advice.
+Care home residents were excluded from these analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.
 
-\item We then tabulated the \% ( and CI for the \%) of those on the shielding list with "Any evidence of COVID-19" as described above by age and sex   and reason for shielding as well as overall.
-
-\item We also tabulated the \% of those on the shielding list with "Severe i.e. CCU treated or fatal COVID-19 " as described above, by age and sex   and reason for shielding as well as overall.
-
-\item To compare risks in shielded persons before and after shielding was introduced For each shielded person we  marked up their longitudinal record  from Feb 7th with date of shielding letter, date of stop shielding letter (Aug 1st)  and any date of death with the record being censored at data of death.
-We then computed the number of cases of any evidence of COVID-19  per total  person days in pre shielded time, then shielded time , then post shielded time.
-We repeated this for severe cases.
-
-\item Among the shielding to understand whether there are other determinants of becoming a case, we compared the  \% with each of the  other listed conditions,  the \% with more than 1 listed condition, the number of prescribed ATC drug classes, the \% in  residential care home,  the  SIMD  distribution and ethnicity  distribution in those who became cases and those who did not.
-
-\item To provide a comparator for the background population,  we used  the 2019 age band and sex - mid year estimates for the general population and the shielding status in cases to derive  the \% ( and 95\%CI) with any evidence of COVID-19 and severe COVID-19 in the same age and sex groups in the general population  (DAVID this is available from STUART).
-We then used logistic regression of the one year age band counts of cases and non cases in the shielded and general population to compute and OR associated with shielding.
-
-\item The odds of "exposure "to eligibility for shielding in cases  versus  controls was also  quantified as the conditional odds ratio  from a conditional logistic regression.
-We did this because it was easy to extend this regression was then extended to include SIMD, ethnicity, residential care home status,  the "listed "COVID-19 risk conditions.
-
-\item Since the calculation in 4.
-above can be confounded by overall background trends in the Re to further evaluate if there is any evidence that shielding reduced risk, The odds of "exposure "to eligibility for shielding in cases  versus  controls  BY WEEK of epidemic was also quantified as the conditional odds ratio  from a conditional logistic regression.
-
-\end{itemize}
-
-Analyses were conducted using the R statistical package .
+Rate ratios were estimated in the case-control study by conditional logistic regression. For time series analyses, these rate ratios were calculated over 20-day time windows.
 
 # Results
+## Cumulative risk of COVID-19 up to presentation date of `r format(max(cc.all$SPECIMENDATE, na.rm=TRUE), "%d %b %Y")` 
+Supplementary Figure \ref{fig:plotcasefatality} compares the time series of the fatal case proportion in those eligible and ineligible for shielding, by seven-day window of presentation date.
+The fatal case proportion in those eligible for shielding reached a peak in the first week of March and fell after this, suggesting that from mid-March onwards any misallocation of fatal cases to those ineligible for shielding was minimal.
 
-Table \ref{tab:shieldgroups} shows the frequency of shielding categories. The 6 women pregnant with heart disese were allocated to the "clinician identified" category.
+## Shielding eligibility categories and clinical outcome
+Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
+The largest risk category was those with severe respiratory disease.
+Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories.
 
-Table \ref{tab:letterdates} shows a breakdown by date of sending letter
+<!--- Supplement Table \ref{tab:shieldgroupsbyage} put this into supplement or  better still omit  altogether its not really that useful eyond being brodly sensible and not showing any emajor problem with classfcaiton but it stll doesnt say anythng about sensitvity or specifcty of the shelding assignation  shows a validation check of whether the SMR01 and PIS  histories of those n the shielded category among all cases and controls is coherent with their assigned shielding category - data frequency of designated risk conditions (ascertained from diabetes registry, hospital discharge records and dispensed prescriptions) in each shielding eligibility category.
+This shows that only 840 out of 2200 individuals on immunosuppressants were identified from prescriptions in primary care, as expected because many immunosuppressant drugs are prescribed only in hospital.
+However 7816 out of 8062 individuals with severe respiratory disease were identified from hospital discharge records and dispensed prescriptions.
+--->
 
-Table \ref{tab:afterletter} shows cases among shielded individuals classified by whether estimated date of infection was before or after the shielding letter.
+Table \ref{tab:shieldcasegroup} shows the number in each eligibility category by case status classified as not diagnosed, not severe or severe COVID-19 .
+Overall among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
+No cases were diagnosed among those identified as pregnant with heart disease.
+Women in this group were allocated to the "Additional conditions" category in subsequent analyses.
+Those on immunosuppressants had the lowest risks.
 
-Table \ref{tab:regression.allcases} shows the estimated rate ratios for any diagnosis of COVID-19
+<!--I think it would be useful to show among the shielded the distribution of any evidence and of severe cases (In columns)  by  age and sex simd residential care home status  and do a logistic regresson among them  ...
+then I can write somehtng like this ...
+Among the shielded risk  of COVID-19 varied strongly by age  and ???
+adjusted by age it varied by shielding category .
+Adjusted for age and shielding category  and  was / was not higher in men than women, in those in residential care homes than  not   and did /di not show an association  with socioeconomic status.
 
-Table \ref{tab:regression.severecases} shows the estimated rate ratios for severe COVID-19.
+** We can't show a breakdown of demographic factors on the shielding cohort as no linkage has been done 
+--->
+
+Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk group as a ribbon plot.
+The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak daily number by `r dateformat(deaths.quartermax.date)`.
+After the first wave of the epidemic had subsided at the end of May 2020, those eligible for shielding accounted for most of the severe cases still occurring until the second wave begain in late September 2020.
+
+### Association of severe COVID-19 with shielding eligibility status and risk factors related to exposure
+`r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][2, 2]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][2, 2]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.
+In subsequent analyses, care home residents have been excluded as policies relevant to shielding residents of care homes are different from those relevant to shielding individuals living independently.
+
+ With those without risk conditions as reference category, the univariate rate ratio for severe disease was
+`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
+in those with moderate risk conditions and
+`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
+in those eligible for shielding.
+
+Table \ref{tab:regressionseverecasesnocare} shows the rate ratios for severe COVID-19 associated with shielding eligibility categories and other relevant covariates, with care home residents excluded.
+Solid organ transplant recipients were the group at highest risk, with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` for severe COVID-19.
+
+The strongest risk factor (in terms of deviance explained) for severe COVID-19 in the general population was recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.
+
+The risk of severe COVID-19 rose steeply with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.
+The other demographic factor associated with increased risk was socioeconomic deprivation: the rate ratio in the least deprived quintile compared with the most deprived quintile was `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 3], table.shielded.severecases.nocare["5 - least deprived", 4])`..
+
+Figure \ref{fig:plotrateratio} shows the time course of the univariate rate ratios for severe COVID-19 associated with eligibility for shielding, recent hospital exposure and number of school-age children in the household.
+These rate ratios are computed over 20-day windows of presentation dates.
+The rate ratio associated with eligibility for shielding increased from `r round(coeffs.timewindow[date.midpoint==as.Date("2020-04-01"), rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.timewindow[date.midpoint==as.Date("2020-05-01"), rateratio], 2)` in the time window with mid-point 1 May.
+The rate ratio associated with recent hospital exposure increased from `r round(coeffs.hosp.timewindow[date.midpoint==as.Date("2020-04-01"), rateratio], 1)` in the time window with mid-point 1 April to `r round(coeffs.hosp.timewindow[date.midpoint==as.Date("2020-05-01"), rateratio], 0)` in the time window with mid-point 1 May.
+The rate ratio per school-age child, adjusted for number of adults and SIMD deprivation score, fell from about 0.9 in mid-March to around 0.8 by mid-April, and rose to about 1 during October 2020.
+As reported previously, in comparison with other occupations patient-facing health-care workers were at higher risk of severe disease with a rate ratio of `r format.estcipv(table.shielded.severecases.nocare["Health care PF", 3], table.shielded.severecases.nocare["Health care PF", 4])`, and teachers were at lower risk with a rate ratio of `r format.estcipv(table.shielded.severecases.nocare["Teacher", 3], table.shielded.severecases.nocare["Teacher", 4])`
+
+### Associations among those eligible for shielding
+Table \ref{tab:shieldedonlyregressionseverecasesnocare} shows associations with risk factors among those eligible for shielding, with severe respiratory disease (the largest category) as reference category.
+Associations with household composition were modest in size and not statistically significant.
+The only detectable risk factor for severe COVID-19 among those eligible for shielding was recent hospital exposure, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 3], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 4])`.
+Of those eligible for shielding, `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 1]` had a recent hospital exposure compared with  `r table.shielded.severecases.nocare["Recent hospital visit/stay", 1]` of the general population.
+
+Figure \ref{fig:plotexphosp} shows the time course of recent hospital exposure in controls by risk category, the fraction of severe cases among those not resident in care homes that was attributable to recent hospital exposure, and the fraction of severe cases that were resident in care homes.
+In those with no risk condition, the frequency of recent hospital exposure fell from mid-March to mid-April, and remained less than 0.5% for the rest of the year except during June and July.
+In those with moderate risk conditions, the frequency of recent hospital exposure also fell sharply betweem mid-March and mid-April, and remained between 1% and 2% except during the summer months.
+In those eligible for shielding, the frequency of recent hospital admissions fell to about 3% in April, and remained at about 4% from mid-September onwards.
+The population attributable risk fraction of severe cases associated with recent hospital exposure reached a peak in the first wave of about 25% at the beginning of May, declined rapidly over the next few weeks, and began to increase again at the beginning of October.
+The fraction of severe cases who were resident in care homes also reached a peak in early May, but subsequent changes were less steep.
 
 # Discussion
+## Summary and interpretation of findings
+In comparison with those with no risk condition, the highest rate ratio for severe COVID-19 was the 15-fold rate ratio in solid organ transplant recipients.
+For other conditions deemed eligible for shielding, the rate ratios were between 4 and 7, compared to the rate ratios of about 3 associated with conditions designated as moderate risk, such as heart disease, diabetes, chronic kidney disease and disabling neurological conditions.
+Other studies of outcome of COVID-19 in solid organ transplant recipients and other individuals using immunosuppressants have been based only on patients admitted to hospital [@avery_inpatient_2020; @andersen_association_2021].
+Such studies cannot assess the risks to immunosuppressed individuals in the population.
 
-## Summary of findings
+Recent hospital exposure was strongly associated with disease in all risk groups.
+As recent hospital exposure was more common in those with moderate risk conditions and those eligible for shielding, especially after a lockdown had been imposed, this exposure accounted for a higher proportion of severe cases in these groups.
+
+If shielding advice had been effective, we would have expected the rate ratio for severe disease associated with eligibility for shielding to fall within two weeks of the first letters being sent out.
+Our results show however that while presentations of severe cases fell rapidly in the general population from the beginning of April 2020, the fall in presentations among those eligible for shielding advice was delayed so that the rate ratio for severe disease associated with eligibility for shielding rose during April 2020, although the daily deaths fell during April 2020 both in those eligible and those ineligible for shielding.
+The most likely explanation for the rise in the rate ratio for severe disease  associated with eligibility for shielding during April 2020 and again beginning in October 2020 is that those in this group were unable to avoid exposure to hospitals during these periods when hospitals were transmitting infection.
+
+ The association of severe COVID-19 with number of adults in the household is consistent with classic studies of other viral infections showing that secondary cases in a household, where the infecting dose is likely to be high, are more severe than index cases [@aaby_malnutrition_1988].
+The inverse association of severe COVID-19 with number of school-age children in the household is consistent with evidence that infection with other coronaviruses generates cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020].
+As reported previously, teachers are at lower risk of severe disease than others of the same age and sex  [@mcallister_report_2020; @ludvigsson_open_2021].
+A possible explanation for this may be that past contact with school-age children helps to build and maintain immunity to endemic coronaviruses that show T-cell cross-reactivity with SARS-CoV-2.
+However no direct studies of this have been reported.
+The time course of the rate ratio, rising from about 0.8 per child at the beginning of October to about 1 in mid-November 2020, suggests that transmission from school-age children to adults may have increased over this period sufficiently to balance the effect of this pre-existing resistance.
 
 ## Comparison with previous literature
+Other reports have evaluated the theoretical possibilities for shielding but without empirical data on its effectiveness.
+A previous report [@jani_comparison_2020] was based on the population of Greater Glasgow and Clyde health board.
+As no age-standardized rate ratios were reported it is difficult to compare our results with theirs.
+In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 [@forbes_association_2020] after adjusting for covariates, but there was no association of non-fatal COVID-19 requiring critical care with living with children in this age group, and no dose-response relationship was reported.
 
-## Limitations and advantages 
-Some key advantages of our study are the population coverage, the range of
-outcomes assessed, the extensive prior risk factor and clinical history
-data.
+## Strengths and limitations 
+Strengths of our study are the population coverage, the range of outcomes assessed, and the comprehensive linkage to medical records and demographic risk factors.
+Limitations are that we do not have data on specific diagnoses, we do not have primary care data other than encashed prescriptions, and we do not yet have data on exposure to health care facilities other than admissions to hospital.
+Failure to capture all exposure to health care facilities would lead to underestimation of the fraction of cases attributable to this exposure.
+As most immunosuppressant drugs are prescribed through hospitals where linkage to prescribing records is not yet possible, the risks associated with  different immunosuppressant drug classes could not be investigated.
 
-## Policy Implications
+We were able to assess the possible effect of unrecorded filtering of fatalities in causing  misallocation of early fatal cases from those eligible for shielding to those ineligible by examining the time course of the fatal case proportion which reached a peak among those eligible for shielding reached a peak by presentation date of 10 March, and fell from that date onwards.
+With a median 13 days from presentation to death, and additional delay in death notifications, it is therefore unlikely that filtering of notified deaths could have misallocated many fatal cases to those ineligible for shielding as long as their presentation date was in the second half of March.
+
+## Policy implications 
+The designation of six groups as clinically extremely vulnerable was undertaken before any systematic studies of associations with severe disease had been undertaken and before risk scores based on these modelling these associations had been constructed.
+Although the list of "diseases and conditions considered to be very high risk" correctly identified some of the most vulnerable individuals, it did not include other conditions associated with high rate ratios for severe disease, including type 1 diabetes, disabling neurological conditions, and chronic kidney disease.
+
+Now that vaccines are available though supply is limited, the focus of risk stratification has shifted from identifying individuals who should be advised to shield to identifying those who should have priority status for being offered vaccination.
+We have reported previously the construction of a risk score based on electronic health records, and we are now updating this to include the additional information obtained when the shielding list was constructed.
+For such a risk score to be used either to assign priority for vaccination or to designate individuals as eligible for shielding, it is necessary to define the objective function that is to be minimized.
+The [Joint Committee on Vaccination and Immunization](https://app.box.com/s/iddfb4ppwkmtjusir2tc/file/729773393403) meeting in September 2020 did not state an objective function explicitly, but assertions like "vaccinating the very oldest group first was always the best strategy" imply that the objective was to minimize the number of deaths.
+An objective of minimizing disability-adjusted years of life lost, for instance, would give more weight to protecting younger individuals with risk conditions than to protecting older individuals in care homes.
+For older individuals with a life expectancy of only a few years, the risks of fatal disease might be outweighed by the value attached to maintaining contact with their families.
+
+The availability of enough vaccine doses to cover all those designated as vulnerable is likely to take several months.
+For some groups such as solid organ transplant recipients, vaccines are less likely to be effective and vaccines that use a live vector may be contraindicated [@chong_comprehensive_2017].
+These groups will require help with shielding until the risk of transmission has been reduced to a very low level by herd immunity.
+It is thus relevant to examine why advice to shield, combined with offers of support for delivery of food and medicines, failed to protect some individuals who were identified as clinically extremely vulnerable.
+We can identify several points from this study:
+
+* As a lockdown on the general population had been imposed on 23 March 2020, and individuals who considered themselves to be at high risk would have been likely to reduce their contact level before then, there were limited possibilities for risk in this group to be reduced further by additional advice to shield and offers of help in letters sent from 3 April onwards.
+
+* Severe disease was strongly associated with the number of other adults in the household.
+It is unlikely that shielding can be effective unless help is provided for other adults living with an individual designated as clinically vulnerable either to co-isolate with the individual who requires shielding, or to move out temporarily.
+To protect immunosuppressed individuals in whom the efficacy of vaccination is expected to be low, one possible strategy is to vaccinate household contacts.
+The Joint Committee on Vaccination and Immunization did not recommend this in a report published in December 2020, stating that "at this time there are no data on the size of the effect of COVID-19 vaccines on transmission." [@joint_commiteee_on_vaccination_and_immunisation_joint_2020]  Despite this  Committee recommended that frontline health and social care workers who are "at risk of transmitting infection to multiple vulnerable persons or other staff in a health care environment" should receive priority for vaccination.
+Even if the effects on the end-points used in clinical trials were entirely accounted for by a reduction in the proportion of infections that are symptomatic, a reduction in transmission would be expected as infected individuals without symptoms are much less likely than those with symptoms to transmit the infection to other household members [@bi_household_2020].
+As the group at highest risk -- solid organ transplant recipients -- is relatively small, vaccination of household contacts of these individuals would not divert a high proportion of scarce vaccine doses.
+
+* Recent hospital exposure was the strongest risk factor for severe disease in those not resident in care homes.
+In those eligible for shielding, the frequency of recent hospital exposure remained in the range 2% to 4% even after lockdown, compared to less than 0.5% in those with no risk conditions.
+Because vulnerable individuals cannot stay away from hospitals, control of nosocomial transmission is critical to protecting these groups from infection and to minimizing the load on critical care services.
+Possible measures to achieve this include assessing and managing COVID-19 cases in separate units or even in separate hospitals, using immunity tests (not only antibody tests) to assign health-care workers who are already immune to management of individuals at high risk of severe disease, and wider use of established infection control measures including personal protective equipment and barrier nursing.
 
@@
 ## Information governance
+This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.
+Datasets were de-identified before analysis.
+
+## Data availability
+The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public
+
+## Funding 
+No specific funding was received for this study.
 
@@
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG
-database, the Scottish Morbidity Record Data Team, the staff of the National
-Register of Scotland, the Public Health Scotland Terminology Services, the HPS
-COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology
-Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
-
-## Funding 
-There was no specific funder for this study.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
 
 ## Public Health Scotland COVID-19 Health Protection Study Group
-Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew 
-Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin
-Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo
-Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme
-Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade
-Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$,
-Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$,
-Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross
-Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$,
-Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna
-Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
+Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
 
-1 Health Protection Scotland (Public Health Scotland), Meridian Court, 5
-Cadogan Street, Glasgow G2 6QE.
+1 Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
 
@@
 
+\newpage
+
+# Figures
+```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1, fig.cap="Ribbon plot of severe cases by vulnerability category and date of presentation, excluding care home residents"}
+p.byelig
+```
+
+```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.4, fig.cap="Rate ratios for severe COVID-19 in those not resident in care homes, by 20-day window of presentation dates.
+Line thickness is proportional to precision of estimate.
+Arrows indicate dates that shielding advice letters were sent, with thickness proportional to number of letters sent on that date.
+Rate ratio per school-age child is adjusted for number of adults and SIMD deprivation score"}
+gridExtra::grid.arrange(p.rateratio + theme(axis.title.x = element_blank()),
+                        p.hosp.rateratio + theme(axis.title.x = element_blank()),
+                        p.children.rateratio, nrow=3)
+```
+
+```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.35, fig.cap="(a)  Prevalence of recent hospital exposure in controls not resident in care homes, by risk category;(b) fraction of severe cases not resident in care homes attributable to recent hospital exposure; (c) fraction of severe cases who were care home residents."}
+gridExtra::grid.arrange(p.hosp + theme(axis.title.x = element_blank()),
+                        p.fracattr + theme(axis.title.x = element_blank()),
+                        p.carefrac, nrow=3)
+```
+
+\newpage
+
 # Tables
-
-```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shieldgroups, 
+```{r shieldgroupsbyage, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::kable(table.shieldgroups.byage, 
              escape=FALSE, 
              booktabs=TRUE,
-			 label="shieldgroups",
+			 label="shieldgroupsbyage",
              row.names=TRUE,
-             caption="Shielded categories") %>%
-    kable_styling(latex_options="hold_position")
+             align=rep("r", 5),
+             caption="Shielding eligibility cohort by eligibility category and age") %>%
+    kableExtra::group_rows("Shielding eligibility category", 1, 7) %>%
+        add_header_above(c(" "=1, "Age group"=4)) %>%
+        add_footnote(label="Percentages are row percentages",
+                     notation="none") %>%
+       kable_styling(latex_options="HOLD_position")
 
 ```
+
+```{r shieldcasegroup, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.shield.casegr, 
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="shieldcasegroup",
+             row.names=TRUE,
+             align=rep("r", 5),
+             caption="Shielding eligibility cohort by case status") %>%
+    kableExtra::group_rows("Shielding eligibility category", 2, 8) %>%
+    column_spec(column=1, width="4.5cm") %>%
+    column_spec(column=2, width="2.5cm") %>%
+    column_spec(column=3, width="2.2cm") %>%
+    column_spec(column=4, width="2.2cm") %>%
+    column_spec(column=5, width="2.2cm") %>%
+    column_spec(column=6, width="2cm") %>%
+        add_footnote(label="Percentages are row percentages",
+                     notation="none") %>%
+     kable_styling(latex_options=c("scale_down", "hold_position"))
+
+```
+
+\onehalfspacing
+
+```{r regressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.shielded.severecases.nocare,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="regressionseverecasesnocare",
+             caption="Rate ratios for severe COVID-19 in those not resident in care homes. ", 
+             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="2cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="2cm") %>%
+    kableExtra::group_rows("Shielding eligibility category", 3, 8) %>%
+    kableExtra::group_rows("School-age children in household", 10, 12) %>%
+    kableExtra::group_rows("Adults in household", 13, 16) %>%
+    kableExtra::group_rows("SIMD quintile", 18, 22) %>%
+    kableExtra::group_rows("Occupation", 23, 26) %>%
+    kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
+    add_footnote(label=c("Percentages are column percentages for each variable",
+                         "PF, patient-facing"), notation="none")
+
+```
+
+```{r shieldedonlyregressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.shieldedonly.severecases.nocare,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="shieldedonlyregressionseverecasesnocare",
+             caption="Rate ratios for severe COVID-19 associated with risk conditions (with severe respiratory disease as reference category) in those eligible for shielding and  resident in care homes",
+             col.names=c(clean.header(colnames(table.shieldedonly.severecases.nocare)[1:2]),
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="1.8cm") %>%
+    column_spec(column=6, width="2.7cm") %>%
+    column_spec(column=7, width="1.8cm") %>%
+    kableExtra::group_rows("Shielding eligibility category", 1, 6) %>%
+    kableExtra::group_rows("School-age children in household", 8, 10) %>%
+    kableExtra::group_rows("Adults in household", 11, 13) %>%
+    add_footnote(label="Percentages are column percentages for each variable",
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "hold_position"))
+
+```
+
+\newpage
+
+\beginsupplement
+
+# Supplementary Figures
+
+```{r plotcasefatality, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8, fig.cap="Proportion of cases that were fatal in those ineligible and eligible for shielding, by 7-day rolling average of presentation date.
+Line thickness proportional to number of  cases"}
+
+p.casefatality
+
+```
+
+\newpage
+
+# Supplementary Tables
 
@@
              row.names=TRUE,
-             caption="Dates of sending letters") %>%
-    kable_styling(latex_options="hold_position")
+             align=rep("r", 6), 
+             caption="Dates of sending letters to those eligible for shielding (row percentages)") %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2.1cm") %>%
+    column_spec(column=3, width="1.9cm") %>%
+    column_spec(column=4, width="1.9cm") %>%
+    column_spec(column=5, width="1.9cm") %>%
+    column_spec(column=6, width="1.9cm") %>%
+    kable_styling(latex_options=c("scale_down", "hold_position"))
 
 ```
-
-```{r afterletter, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.after.letter,
-escape=FALSE,
-             booktabs=TRUE,
-			 label="afterletter",
-             caption="Proportions of cases advised to shield who were diagnosed > 10 days after date of sending letter")
-
-```
-
-```{r regression.allcases, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shielded.allcases,
-escape=FALSE,
-             booktabs=TRUE,
-			 label="regression.allcases",
-   col.names=c(clean.header(colnames(table.shielded.allcases)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-            caption="Rate ratios for diagnosis associated with shielding categories, with and without adjustment for diagnosis > 10 days after sending letter") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
-    column_spec(column=1, width="3cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-   column_spec(column=4, width="2.5cm") %>%
-    column_spec(column=5, width="2cm") %>%
-   column_spec(column=6, width="2.5cm") %>%
-    column_spec(column=7, width="2cm") %>%
-    kable_styling(latex_options="hold_position")
-
-```
-
-```{r regression.severecases, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shielded.severecases,
-escape=FALSE,
-             booktabs=TRUE,
-			 label="regression.severecases",
-   col.names=c(clean.header(colnames(table.shielded.severecases)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-
-             caption="Rate ratios for severe COVID-19 associated with shielding categories, with and without adjustment for diagnosis > 10 days after sending letter") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
-    column_spec(column=1, width="3cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-   column_spec(column=4, width="2.5cm") %>%
-    column_spec(column=5, width="2cm") %>%
-   column_spec(column=6, width="2.5cm") %>%
-    column_spec(column=7, width="2cm") %>%
-    kable_styling(latex_options="hold_position")
-
-```

### 2021-01-09 shielding_v1.Rmd
updated shielding paper, fixed trainmodel.R

---
title: "Risk of COVID-19 in clinically extremely vulnerable individuals advised to shield during the epidemic in Scotland" 
author:
  - name: Paul M McKeigue
    affiliation: "HPS,Usher"
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow  
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Jim McMenamin  
    affiliation: HPS
  - name: Alison Smith-Palmer 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun
    email: helen.colhoun@igmm.ed.ac.uk (07791 652665)
    affiliation: "HPS,IGMM,NHSFife"

address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology
- code: HPS
  address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: ISD
  address: Information Services Division, NHS Scotland, UK

header-includes:
  \usepackage{newunicodechar}
  \DeclareUnicodeCharacter{039B}{\ensuremath{\mit\Lambda}}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
  \usepackage[nolists, nomarkers]{endfloat}
always_allow_html: true
urlcolor: blue
output:
  bookdown::pdf_book:  
    base_format: rticles::elsevier_article
    keep_tex: true
    fig_caption: yes
    number_sections: false
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./biomed-central.csl
linenumbers: false
linkcolor: cyan

---

On behalf of Public Health Scotland COVID-19 Health Protection Study Group

Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group

```{r functions, echo=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
source("helperfunctions.R")

# rmarkdown::render("shielding.Rmd", output_file="shielding.pdf")

dateformat <- function(x) { # format date and strip leading zero
     gsub("^0", "", format(x, "%d %b %Y"))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

```
# Abstract

• Objectives - to investigate the risk of COVID-19 in individuals designated as clinically extremely vulnerable, and the effectiveness of the advice to shield and support for shielding that was provided in the first wave of the epidemic in Scotland.

• Design - (1) cohort study of all individuals identified as eligible for shielding; (2) matched case-control study (REACT-SCOT) including all cases diagnosed with COVID-19 in Scotland

• Setting - population of Scotland

• Participants - `r nrow(shielded.full)` individuals identified as eligible for shielding, `r nrow(cc.all[CASE==1])` diagnosed cases and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice.

• Interventions - Advice to shield and support for shielding was offered from 3 April 2020 to  1 August 2020.

• Main outcome measures - (1) any diagnosis of COVID-19 by nucleic acid test, hospital discharge record or death certificate; (2) diagnosed cases that required critical care or were fatal.

• Results - There were `r nrow(shielded.full[!is.na(casegr2)])` diagnosed cases in shielded individuals of which `r nrow(shielded.full[!is.na(casegr2) & casegr2=="Severe"])` were severe.   `r nrow(cc.all[shield.any==TRUE & CASE==1 & after.letter==1])` cases were diagnosed more than 14 days after the shielding letter had been sent.  In the case-control study, the rate ratios for severe disease in the six shielding eligibility categories ranged from `r format.estcipv(table.shielded.severecases.nocare["Rare diseases", 3], table.shielded.severecases.nocare["Rare diseases", 4])` for the rare disease category to `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` in solid organ transplant recipients.  In those not resident in care homes the strongest risk factor for severe COVID-19 was recent hospital stay: rate ratio `r format.estcipv(table.shielded.severecases.nocare["recent hospital stay", 3], table.shielded.severecases.nocare["recent hospital stay", 4])`.  The rate ratio associated with eligibility for shielding was higher by a factor of  `r format.estcipv(table.shielded.severecases.nocare["interval2.shielding", 5], table.shielded.severecases.nocare["interval2.shielding", 6])` during the shielding period than before the shielding period.

• Conclusions - Advice to shield and offers of support were not effective in preventing severe COVID-19 in clinically extremely vulnerable individuals in a population where a general lockdown had already been imposed.  From the strong associations of severe COVID-19 with recent hospital stay in it appears that prevention transmission in health-care settings may be one of the most effective ways to shield these individuals.

• Trial registration - ENCEPP registration of case-control study.

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a lockdown on 23 March 2020.

Early in the epidemic it was realised that certain people would likely be extremely vulnerable to severe COVID-19 disease including those receiving certain immunosuppressant drugs or with other conditions.  Therefore Public Health Scotland (PHS) instituted a process of identifying such persons through their primary care physician and through searching of administrative databases  (see appendix for full protocol). Those considered eligible then received a letter detailing the need for self isolation and the support they would receive to achieve this. These letters were sent out beginning on `r dateformat(batches$Date.Sent)[1]` until advice to shield and support for shielding were withdrawn on 1 August 2020.  In October 2020 additional advice was issued to "those at higher risk from coronavirus" to protect themselves when the protection level in their region is 3, 4 or 5, but no additional support was offered (https://www.gov.scot/publications/covid-shielding/)

An initial survey of ~12000 shielding eligible persons indicated that compliance was high but often partial and varied across socio-demographic groups. <!--- cite source from Joke Delvaux --->

Here we take the cohort of  `r nrow(shielded.full)` persons who were deemed clinically extremely vulnerable and thus eligible for shielding at some time during this period. We examine the risk of diagnosed cases of COVID-19 and severe COVID-19.  In the REACT-SCOT study, an ongoing case-control study of all diagnosed cases of COVID-19 with age and sex matched controls, we examine the rate ratios for disease associated with eligibility for shielding and compare these rate ratios before and after advice to shield was issued to these groups.

# Methods
## Ascertainment of eligibility for shielding 
To identify  all those nationally deemed eligible for shielding we took the list generated by PHS at the outset of the epidemic [@health_protection_scotland_search_2020]:

The following groups were identified as eligible for shielding:

* Solid organ transplant recipients

* People with specific cancers

* People with severe respiratory conditions: cystic fibrosis, severe asthma, severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)

* People with rare diseases: inborn errors of metabolism that significantly increase the risk of infections (such as severe combined immunodeficiency, homozygous sickle cell disease (not trait), interstitial Lung Disease and sarcoidosis.

* People on immunosuppression therapies: lists were collected from dermatology, rheumatology, renal, gastroenterology and neurology specialists. Criteria were:

  - Corticosteroid equivalent to prednisolone 20 mg/day for 4 weeks or more.

  - On a single agent that has high risk of causing immunosuppression e.g.  Cyclophosphamide, rituximab, Infliximab, Cladribine, Alemtuzumab. These medications are prescribed only in  secondary care.

  - Corticosteroid equivalent to prednisolone \ensuremath{\ge} 5 mg/day for at least 4 weeks and other immunosuppressive therapy.

  - On two immunosuppressant medications and with a relevant co-morbid condition.

* People who were pregnant with significant heart disease, congenital or acquired.

* Additional conditions.  Patients identified by primary and secondary care clinicians as requiring shielding advice were included in this group.  On 11 May 2020 the list was updated to include people on renal dialysis and those who have had a splenectomy.

As described elsewhere [@health_protection_scotland_search_2020],  patients were identified via searches of hospital discharge and prescribing records, as well as other sources including regional cancer chemotherapy databases, NHS blood and transplant registries and lists prepared by clinicians in relevant specialties.

For this analysis only the shielding eligibility category and the date on which the shielding letter was sent were available.  Additional data on specific diagnoses and immunosuppressant drugs were supplied to Public Health Scotland but were not available for this analysis.

Letters were sent on Fridays, beginning on `r format(batches$Date.Sent[1], "%d %b %Y")`

## COVID-19 case ascertainment and definition
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].  For the total Scottish population cases with any evidence of COVID-19 were ascertained as those with a positive  nucleic acid test  for SARS-CoV-2 in  the national virology laboratory database, a hospital discharge code for COVID-19, or a death with mention of COVID-19 anywhere on the death certificate.  The presentation date was defined as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those diagnosed only through hospital discharge records, and as fourteen days before the date of death for those diagnosed only through death certificates

The databases used were the Electronic Communication of Surveillance in Scotland (ECOSS) database  that captures virology testing in all NHS laboratories nationally, the National Records of Scotland (NRS) death registrations data, the RAPID database which is a daily update of hospitalisation nationwide and the Scottish Morbidity Records-01 which capture hospital discharges  and discharge diagnostic codes nationally.  All these databases use the Community Health Index (CHI) number as identifier.  The master CHI database fields include age, sex, residential postcode, and residential care home status. Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.

Among cases of COVID-19 we defined severe disease as that requiring critical care or  fatal.   Admissions to critical care units -- intensive care, high dependency or combined units -- were obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal COVID-19 was defined as mention of COVID-19 anywhere on the death certificate or death within 28 days of testing positive or death within 28 days of admission for cases ascertained through discharge records.  These definitions of COVID-19 deaths match those used by the National Records of Scotland and Public Health Scotland respectively.

## Ascertainment of controls
For every case defined above, we selected 10 controls matched for sex, one-year age band and attending the same GP practice  and alive on the presentation date of the index case.   The sampling frame for controls was the Community Health Care Index.

## Demographic data
For all those on the shielding list and for all COVID-19 cases and controls socioeconomic status was evaluated using the postcode-based indicator the Scottish Index of Multiple Deprivation (SIMD) using the residential postcode in the CHI database [@fraser_scottish_2020]. Postcodes were allocated to which quintile of the socioeconomic distribution of Scottish residential datazones they belonged. Ethnic group was captured by linkage to prior SMR01 and SMR00 and SMR04 records in which the patients self assigns ethnicity according to the groups that are used in the national census.

## Other morbidity and drug prescribing
For all COVID-19 cases and controls, the records were linked to  general hospital admissions data in the  Scottish Morbidity Record 01 (SMR01),  and  all discharge codes over the past five years and Anatomical Therapeutic Chemical codes [@who_anatomical_2016] from Prescribing Information Systems (PIS) data for the past three years were extracted and used to define co-morbid conditions and prior drug exposures.

We first extracted history of a specific list of conditions and drug classes that have been included as "moderate risk conditions" for COVID-19 by public health agencies [@nhs_whos_2020]. These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record, or a dispensed prescription in which that diagnostic code is the only indication (see Supplementary Materials for details).

We defined a variable termed "recent hospital stay" so as to include cases who were in hospital during the possible time interval for infection given the presentation date, but to exclude admissions caused by COVID-19.  We assume a minimum time of 4 days from exposure to testing positive, and a maximum time of 14 days from exposure to hospital admission in symptomatic cases. This variable was defined as:

1. any hospital inpatient stay during last 14 days before presentation date, with admission date at least 5 days before presentation date
2. recent hospital discharge: a discharge date  <= 14 days before presentation date;

The Scottish NHS "probable or definite" healthcare associated COVID-19 inpatient infection as a positive test in someone who is an inpatient and was admitted at least 7 days earlier.  This is more stringent in excluding cases who test positive 5 to 7 days after admissions, but ignores cases who test positive after discharge.

Cases were grouped by presentation date into three advice intervals: (1) up to 17 April 2020; (14 days after the first letters were sent, (2) from 18 April 2020 to 15 August 2020 (14 days after advice and support for shielding were withdrawn; (3) from 16 August 2020 onwards.

## Statistical methods

### Shielding eligibility cohort

We tabulated the shielding eligibility cohort as follows:

* shielding eligibility category x age group.

* shielding eligibility category x outcome: no diagnosis, not severe, severe COVID-19.

* shielding eligibility category x date of sending letter

* for severe cases, shielding eligibility category x advice interval

### Case-control study

* shielding eligibility category x moderate risk conditions

* unadjusted and adjusted rate ratios for severe disease associated with each shielding eligibility category, with those not eligible as reference category. Interaction terms for shielding eligibility x advice interval, with advice interval 1 as reference category.

* unadjusted and adjusted rate ratios for severe disease associated with each shielding eligibility category in shielded individuals only, with severe respiratory disease as reference category. 
\end{itemize}

Rate ratios were estimated in the case-control study by conditional logistic regression.

# Results
Table \ref{tab:shieldgroupsbyage} shows the frequency of shielding eligibility categories by age.  Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories.  The largest shielding eligibility category was those with severe respiratory disease.  The small number of women pregnant with heart disease were allocated to the "Additional conditions" category in subsequent analyses.

Table \ref{tab:shieldcasegroup} shows a breakdown of cases in each shielding eligibility category by  outcome. Severe respiratory disease and cancers accounted for most of the severe cases.

Table \ref{tab:shieldgroupsbyage} shows the frequency of designated risk conditions (ascertained from diabetes registry, hospital discharge records and dispensed prescriptions) in each shielding eligibility category. This shows that only 840 out of 2200 individuals on immunosuppressants were identified from prescriptions in primary care, as expected because many immunosuppressant drugs are prescribed only in hospital.  However 7816 out of 8062 individuals with severe respiratory disease were identified from hospital discharge records and dispensed prescriptions.

Table \ref{tab:letterdates} shows a breakdown by date of sending letter.  36% of severe cases had been diagnosed by the time the first letters were sent out on 3 April 2020.  Most of the solid organ transplant recipients, cancer patients and those with respiratory disease were identified in this first batch of letters.  Most of those with rare diseases or on immunosuppressants were identified in the second batch of letters sent out one week later. Sending of letters to those identified as in the "additional conditions" category continued for the next four weeks.

Table \ref{tab:afterletter} shows cases among shielded individuals classified by whether the date of diagnosis was more than 14 days after the date that the letter was sent.  The proportion of these post-advice cases was highest in the shielding eligibility categories whose letters were sent out inthe first batch: solid organ transplant recipients, cancer patiens and those with respiratory disease.  There are no other obvious associations with post-advice diagnosis.

Table \ref{tab:regressionseverecasesnocare} shows the estimated rate ratios for severe COVID-19 with care home residents excluded. The rate ratios associated with shielding eligibility categories are larger for severe disease than for any diagnosis, with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` for solid organ transplant recipients.  There was also a strong association with socioeconomic status with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 3], table.shielded.severecases.nocare["5 - least deprived", 4])` in the least deprived quintile of SIMD compared with the most deprived.  The adjusted rate ratios associated with number of adults in the household (1.12 per adult) and recent hospital stay (ratio 19.1) were also large.  The rate ratio for severe disease associated with the "letter sent" variable however was `r format.estcipv(table.shielded.severecases.nocare["Letter sent >= 14 days earlier", 3], table.shielded.severecases.nocare["Letter sent >= 14 days earlier", 4])`.

Figure 1 compares the time series of severe cases in the general Scottish population with that in those eligible for shielding.  Presentations of severe cases in the general population peaked at the end of March 2020 and fell rapidly after that.  In those eligible for shielding the daily number of presentations of severe cases began later and the fall was less rapid.

Figure 2 shows the same result in terms of rate ratios for severe COVID-19 associated with eligibility for shielding, computed over 20-day windows of presentation dates.  The rate ratio asociated with eligibility for shielding increased from about 1.8 in late March to about 2.2 in mid-April.

# Discussion
## Summary of findings
Shielding letters were sent out after the epidemic had peaked and after a general lockdown had been imposed.

In comparison of the risk in groups eligible for shielding to that in the general population, the highest rate ratio for severe disease was that for solid organ transplant recipients.  The absolute risks of severe disease during this first wave of the epidemic were low: even in solid organ transplant recipients the risk was less than 0.5%.

In both ineligible and eligible individuals severe COVID-19 was strongly associated with  recent hospital stay, suggesting that nosocomial transmission was an important risk factor.  In those eligible for shielding, there was a weak association of severe disease with the number of adults in the household.

A possible explanation for the later peaking and slower fall of severe cases among those eligible for shielding compared with the general population, is that those eligible for shielding were exposed while attending health care facilities.

## Comparison with previous literature
A previous report [@jani_comparison_2020] was based on the population of Greater Glasgow and Clyde health board. As no age-standardized rate ratios were reported it is difficult to compare our results with theirs.

## Strengths and limitations 
Strengths of our study are the population coverage, the range of outcomes assessed, the extensive prior risk factor and clinical history data.  Limitations are that we do not have data on specific diagnoses and immunosuppressant drugs, numbers of children in the household were not available for this analysis, and we do not  have data on visits to health care facilities other than admissions to hospital.

## Policy Implications

Lessons that can be learned:

* For shielding advice to the vulnerable to be effective, it should be provided early in an epidemic, before any population-wide measures are imposed. This necessitates maintaining up to date registries of the clinically extremely vulnerable that are accessible to public health agencies have access to them.  Capture of records of hospital prescribing, which include immunosuppressant and cytotoxic drugs, would facilitate this.

* The cases of severe disease in clinically extremely vulnerable individuals during the shielding period could be investigated further with a review of medical records to identify avoidable factors such as exposure to infection while attending health-care facilities.

Control of nosocomial infection should include specific measures to shield clinically extremely vulnerable individuals when they attend hospital.

To protect clinically extremely vulnerable individuals living with others, support could be provided for other household members either to isolate with the vulnerable individual, or to move into alternative acccommodation for the duration of the epidemic. However this study suggests that even those living alone were not adequately shielded.

# Declarations
## Information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.  Datasets were de-identified before analysis.

## Data availability
The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public

## Funding 
No specific funding was received for this study.

## Conflicts of interest
The authors declare no conflicts of interest.

## Acknowledgements
We thank all staff in critical care units who submitted data to the SICSAG
database, the Scottish Morbidity Record Data Team, the staff of the National
Register of Scotland, the Public Health Scotland Terminology Services, the HPS
COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology
Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.

## Public Health Scotland COVID-19 Health Protection Study Group
Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew 
Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin
Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo
Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme
Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade
Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$,
Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$,
Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross
Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$,
Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna
Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$

1 Health Protection Scotland (Public Health Scotland), Meridian Court, 5
Cadogan Street, Glasgow G2 6QE.

2 NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2
6QE.

coeffs.timewindow <- coeffs.timewindow[date.midpoint < as.Date("2020-07-01")]

p.rateratio <-
    ggplot(data=coeffs.timewindow, aes(x=date.midpoint, y=coeff)) +
    geom_line(size=0.01 * coeffs.timewindow[, se.coeff]^-2) +
    scale_y_continuous(breaks=log(c(1, 1.5, 2, 2.5, 3)),
                       labels=c(1, 1.5, 2, 2.5, 3)) + 
    xlab("Presentation date: mid-point of 20-day window") + ylab("Rate ratio (log scale)") +
    ggtitle("Rate ratio for severe COVID-19 associated with eligibility for shielding")

# References
<div id="refs"></div>

```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Severe cases by presentation date"}

gridExtra::grid.arrange(p.all, p.elig, nrow=2)

```

```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Rate ratio for severe COVID-19 associated with elegibility for shielding, by 20-day window of presentation dates.  Line thickness is proportional to precision of estimate."}

p.rateratio

```

\newpage

```{r shieldgroupsbyage, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldgroups.byage, 
escape=FALSE, 
             booktabs=TRUE,
			 label="shieldgroupsbyage",
             row.names=TRUE,
caption="Shielding eligibility categories by age (row percentages)") %>%
        kableExtra::group_rows("shielding eligibility category", 1, 7) %>%
    kable_styling(latex_options="hold_position")

```

```{r shieldcasegroup, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shield.casegr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldcasegroup",
             row.names=TRUE,
             caption="Cases in shielding eligibility cohort by outcome") %>%
    kableExtra::group_rows("shielding eligibility category", 2, 7) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2.2cm") %>%
    column_spec(column=3, width="2.2cm") %>%
    column_spec(column=4, width="2.2cm") %>%
    column_spec(column=5, width="2.2cm") %>%
    column_spec(column=6, width="2.2cm") %>%
    column_spec(column=7, width="2.2cm") %>%
    kable_styling(latex_options="hold_position")

```

```{r shieldgroupconditions, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldgroups.conditions, 
escape=FALSE, 
             booktabs=TRUE,
			 label="shieldgroupsconditions",
             row.names=TRUE,
caption="Shielding eligibility categories in cases and controls by presence of listed conditions") %>%
    kableExtra::group_rows("Shielding eligibility category", 1, 7) %>%
    kable_styling(latex_options="hold_position")

```

```{r letterdates, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.date.letter,
             escape=FALSE, 
             booktabs=TRUE,
			 label="letterdates",
             row.names=TRUE,
             caption="Dates of sending letters, excluding care home residents") %>%
    kableExtra::group_rows("Cumulative numbers of cases who had tested positive on or before date", 1, 2, bold=TRUE) %>%
     kableExtra::group_rows("Numbers of cases and controls sent shielding letters on date", 3, 8, bold=TRUE) %>%

    kable_styling(latex_options="hold_position")

```

```{r afterletter, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldedcases.interval,
             escape=FALSE,
             booktabs=TRUE,
             col.names=c("up to 17 April", "18 April to 15 August", "after 15 August"),
			 label="afterletter",
             caption="Severe cases by shielding eligibility category and advice interval") %>%
    kableExtra::group_rows("shielding eligibility category", 2, 7)

```

```{r regressionallcasesnocare, echo=FALSE, warning=FALSE, message=FALSE}
if(FALSE) {
    knitr::kable(table.shielded.allcases.nocare,
                 escape=FALSE,
                 booktabs=TRUE,
                 label="regressionallcasesnocare",
                 caption="Rate ratios for diagnosis associated with shielding eligibility categories excluding care home residents, with and without adjustment for infection after letter",
                 col.names=c(clean.header(colnames(table.shielded.allcases.nocare)[1:2]),
                             "Rate ratio (95\\% CI)", "h$p$-value",
                             "Rate ratio (95\\% CI)", "$p$-value")) %>%
        add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
        column_spec(column=1, width="3cm") %>%
        column_spec(column=2, width="2cm") %>%
        column_spec(column=3, width="2cm") %>%
        column_spec(column=4, width="2.5cm") %>%
        column_spec(column=5, width="2cm") %>%
        column_spec(column=6, width="2.5cm") %>%
        column_spec(column=7, width="2cm") %>%
        kableExtra::group_rows("Shielding eligibility category", 1, 7) %>%
            kableExtra::group_rows("SIMD quintile", 8, 12) %>%
            kable_styling(latex_options="hold_position")
}

```

```{r regressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shielded.severecases.nocare,
             escape=FALSE,
             booktabs=TRUE,
             label="regressionseverecasesnocare",
             caption="Rate ratios for severe COVID-19 associated with shielding eligibility categories (with no shielding advice as reference category) in those not resident in care homes.  Effects of advice intervals 2 and 3 on the rate ratio associated with any shielding eligibility category are represented by indicator variables", 
             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="2cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="2cm") %>%
    kableExtra::group_rows("Shielding eligibility category", 2, 7) %>%
    kableExtra::group_rows("Household composition", 9, 13) %>%
    kable_styling(latex_options="hold_position")

```

```{r shieldedonlyregressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldedonly.severecases.nocare,
             escape=FALSE,
             booktabs=TRUE,
             label="shieldedonlyregressionseverecasesnocare",
             caption="Rate ratios for severe COVID-19 associated with shielding eligibility categories (with severe respiratory disease as reference category) in clinically extremely vulnerable individuals not resident in care homes", 
             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="2cm") %>%
    column_spec(column=6, width="2.5cm") %>%
    column_spec(column=7, width="2cm") %>%
    kableExtra::group_rows("Shielding eligibility category", 1, 6) %>%
    kableExtra::group_rows("Household composition", 7, 11) %>%
    kable_styling(latex_options="hold_position")

```

### 2021-01-09 shieldingslides.Rmd
updated shielding paper, fixed trainmodel.R

---
title: "Relation of severe COVID-19 to exposure related factors in REACT-SCOT case-control study" 
author: Helen Colhoun and Paul McKeigue 
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
#header-includes:
#  \usepackage{booktabs}
output: 
  beamer_presentation:
#    reference_doc: ~/markdown.pptx
    md_extensions: +raw_tex+tex_math_dollars-raw_tex
    slide_level: 1
    fig_caption: false
    includes:
      in_header: ~/beamer_header.tex
    #   after_body: ~/beamer_afterbody.tex
    keep_tex: yes
urlcolor: blue 
always_allow_html: yes
---

# Severe cases by risk category and date of presentation: ribbon plot
```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.5,fig.asp=0.75, fig.cap="Care home residents excluded"}

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

theme_set(theme_gray(base_size = 12))

p.byelig
```

# Rate ratios for severe COVID-19 by presentation date 
```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Care home residents excluded. Line thickness is proportional to precision of estimate. Arrows indicate dates that shielding advice letters were sent, with thickness proportional to number of letters sent on that date. Rate ratio per school-age child is adjusted for number of adults and SIMD deprivation score"}
gridExtra::grid.arrange(p.rateratio + theme(axis.title.x = element_blank()),
                        p.hosp.rateratio + theme(axis.title.x = element_blank()),
                        p.children.rateratio, nrow=3)
```

# Prevalence of recent hospital exposure by presentation date

```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.5, fig.asp=0.8, fig.cap=" Prevalence of recent hospital exposure in controls not resident in care homes, by risk category"}
p.hosp

```

# Population attributable risk fraction for severe COVID-19 by presentation date

```{r plotfrac, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4, fig.asp=0.7, fig.cap="Fraction of severe cases not resident in care homes attributable to recent hospital exposure"}
p.fracattr

```

# Rate ratios for severe COVID-19 associated with risk groups and exposure factors

```{r regressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(
#           as.data.frame(lapply(
               table.shielded.severecases.nocare[1:8, c(1:2, 5:6)], 
                                        #               function(x) gsub("\\\\", "\\\\\\\\", x))),
           format="latex", 
             escape=FALSE,
             booktabs=TRUE,
             label="regressionseverecasesnocare",
             caption="Rate ratios for severe COVID-19 in those not resident in care homes. ", 
             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
 #                        "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
#    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
#    column_spec(column=4, width="2.5cm") %>%
#    column_spec(column=5, width="2cm") %>%
#    column_spec(column=6, width="2.5cm") %>%
#    column_spec(column=7, width="2cm") %>%
    kableExtra::group_rows("Shielding eligibility category", 3, 8) %>%
#    kableExtra::group_rows("School-age children in household", 10, 12) %>%
#    kableExtra::group_rows("Adults in household", 13, 16) %>%
#    kableExtra::group_rows("SIMD quintile", 18, 22) %>%
#    kableExtra::group_rows("Occupation", 23, 26) %>%
    kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
    add_footnote(label=c("Percentages are column percentages for each variable",
                         "PF, patient-facing"), notation="none")

```

### 2021-01-09 casesonly.R

@@ -1,0 +1,28 @@
+## save cases only table for diabetes paper
+
+casesonly <- as.data.table(readRDS("./data/2020-09-09/all_cases.rds"), key="ANON_ID")
+                                        # latest specimen date 29 Sep 20
+                                        # latest death date 31 July 20
+sicsag.new <- as.data.table(readRDS("./data/2020-09-09/CC_SICSAG_ANON.rds"), key="ANON_ID") # latest hospital date 6 Sep 2020
+casesonly$DATE_OF_DEATH <- as.Date(casesonly$DATE_OF_DEATH)
+casesonly <- casesonly[SPECIMENDATE <= as.Date("2020-08-25") &
+                       (is.na(DATE_OF_DEATH) |
+                        DATE_OF_DEATH <= as.Date("2020-07-31")), ]
+sicsag.new <- sicsag.new[AdmitUnit <= as.Date("2020-08-23"), ]
+casesonly <- sicsag.new[casesonly]
+setnames(casesonly, "AgeYear", "AGE")
+setnames(casesonly, "SEX", "sex")
+
+severe.cases <- casesonly[(!is.na(covidICUorHDU) & (covidICUorHDU == 1 | covidICUorHDU ==3)) |
+                          dead28 == 1 |
+                          covid_cod == 1, ]
+fatal.cases <- casesonly[dead28 == 1 | covid_cod == 1, ]
+
+case.freqs <- with(severe.cases[, .(AGE, sex, SPECIMENDATE)],
+                   table(AGE, sex, exclude=NULL))
+death.freqs <- with(fatal.cases[, .(AGE, sex, SPECIMENDATE)], table(AGE, sex, exclude=NULL))
+severe.cases <- severe.cases[,  .(AGE, sex, SPECIMENDATE)]
+fatal.cases <- fatal.cases[,  .(AGE, sex, SPECIMENDATE)]
+
+save(severe.cases, fatal.cases, case.freqs, death.freqs,
+     file="casefreqs31July.RData") ## national cumulative cases and deaths by sex and one year age group

### 2021-01-09 wc.manuscript.sh

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+pdftotext pharmaco.pdf
+ 
+function wcpdfms() {
+    local firstline=$( grep -n ^.\\?Background$ pharmaco.txt | sed 's/:.*//' )
+    local lastline=$( grep -n ^.\\?Declarations$ pharmaco.txt | sed 's/:.*//' )
+    local numlines=$(( lastline - firstline ))
+    local wordcount=$( tail -n+$firstline pharmaco.txt | head -n $numlines | grep -v ^[0-9]\\+/[0-9]\\+$ | grep -v ^.20$ | wc -w )
+    echo $wordcount
+}
+
+function wcabstr() {
+    local firstline=$( grep -n "^Background .\\+" pharmaco.txt | sed 's/:.*//' )
+    local lastline=$( grep -n "^.\\?Keywords" pharmaco.txt | sed 's/:.*//' )
+    local numlines=$(( lastline - firstline ))
+    local wordcount=$( tail -n+$firstline pharmaco.txt | head -n $numlines | grep -v ^[0-9]\\+/[0-9]\\+$ | grep -v ^.20$ | wc -w )
+    echo $wordcount
+}
+
+
+echo "$(wcpdfms) words, abstract $(wcabstr) words"

### 2021-02-10 beamerkable.Rmd
updated shielding markdown script before circulation

@@
 ---
-title: "beamer kable" 
-author: PM 
+title: "Morbidity and transmission factors associated with variant B1.1.7" 
+author: "Paul McKeigue and Helen Colhoun" 
 output: 
@@
     includes:
-      in_header: /mnt/phs/beamer_header.tex
+      in_header: ../beamer_header.tex
     keep_tex: yes
@@
 
-# Title slide
-
-# Table with kable
+# 28-day fatality by S gene dropout status
+\scriptsize
+\setbeamertemplate{caption}{\raggedright\insertcaption\par}
 
@@
 
-knitr::kable(mtcars[1:4, 1:4], 
+knitr::kable(table.death.coxph[3:5, c(1:2, 5:6)],
              escape=FALSE, 
              booktabs=TRUE,
-	label="regressionseverecasesnocare",
-	caption="Rate ratios for severe COVID-19 in those not resident in care homes" 
-	) %>% 
-    add_header_above(c(" "=2, "Univariate"=2))
+			 label="coxfatal",
+             row.names=TRUE,
+             align=c("r", "r", "r", "l"), 
+             col.names=c(clean.header(colnames(table.death.coxph)[1:2]),
+                        "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
+             caption="Death within 28 days of testing positive: Cox regression on S gene dropout and covariates") %>%
+    column_spec(column=1, width="2cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="1.8cm") %>%
+    column_spec(column=4, width="2.2cm") %>%
+    column_spec(column=5, width="1.5cm") %>%
+    add_footnote(label=c("Adjusted for age and sex",
+                         "Stratified by days since testing positive and calendar week"),
+                 notation="none")
 
 ```
-# more stuff
+# Test-positive cases: severity and cycle threshold
 
-           format="latex", 
-             escape=FALSE,
-             booktabs=TRUE,
-             label="regressionseverecasesnocare",
-             caption="Rate ratios for severe COVID-19 in those not resident in care homes. ", 
-             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
- #                        "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
-#    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
-    column_spec(column=1, width="3cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-#    column_spec(column=4, width="2.5cm") %>%
-#    column_spec(column=5, width="2cm") %>%
-#    column_spec(column=6, width="2.5cm") %>%
-#    column_spec(column=7, width="2cm") %>%
-    kableExtra::group_rows("Shielding eligibility category", 3, 8) %>%
-#    kableExtra::group_rows("School-age children in household", 10, 12) %>%
-#    kableExtra::group_rows("Adults in household", 13, 16) %>%
-#    kableExtra::group_rows("SIMD quintile", 18, 22) %>%
-#    kableExtra::group_rows("Occupation", 23, 26) %>%
-    kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
-    add_footnote(label=c("Percentages are column percentages for each variable",
-                         "PF, patient-facing"), notation="none")
+```{r plotvariantbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=0.8}
+
+theme_set(theme_gray(base_size = 18))
+
+gA <- ggplotGrob(p.variantbydate)
+gB <- ggplotGrob(p.mediansbydate)
+grid::grid.newpage()
+grid::grid.draw(rbind(gA, gB))
+
+```
+
+# Severe cases by source
+
+```{r plotvariantbysource, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=0.75}
+
+theme_set(theme_gray(base_size = 18))
+p.bysource
+
+```

### 2021-02-10 shielding.Rmd
updated shielding markdown script before circulation

--- 2020-11-03 shielding.Rmd
@@
 ---
-title: "Risk of COVID-19 in clinically extremely vulnerable individuals advised to shield during the epidemic: the REACT-SCOT case-control study" 
+title: "Relation of severe COVID-19 in Scotland to transmission-related factors in individuals with and without risk conditions eligible for shielding support: REACT-SCOT case-control study"
 author:
-  - name: Paul M McKeigue
-    affiliation: "HPS,Usher"
+  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
+    affiliation: HPS
+  - name: David McAllister\textsuperscript{\getAff{Glasgow}}  
+    affiliation: HPS 
   - name: David Caldwell 
@@
     affiliation: HPS
-  - name: Sharon Hutchinson 
-    affiliation: GCU
-  - name: David McAllister 
-    affiliation: Glasgow  
-  - name: Chris Robertson 
-    affiliation: Strathclyde
+  - name: Sam Colville
+    affiliation: HPS
+  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
+    affiliation: HPS
+  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
+    affiliation: HPS
   - name: Nazir Lone
     affiliation: Usher
-  - name: Janet Murray
-    affiliation: HPS 
   - name: Jim McMenamin 
-    affiliation: HPS
-  - name: Alison Smith-Palmer 
     affiliation: HPS
@@
     affiliation: HPS
-  - name: Helen M Colhoun
-    affiliation: "HPS,IGMM,NHSFife"
-  - name: Public Health Scotland COVID  -19 Health Protection Study Group
+  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
+    email: helen.colhoun@igmm.ed.ac.uk (+44 7791 652665)
     affiliation: HPS
-
 address: 
-- code: IGMM
-  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland.
-HC - AXA Chair in Medical Informatics and Epidemiology
 - code: Usher
@@
 - code: HPS
-  address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
 - code: Glasgow
   address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
+- code: GCU
+  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
 - code: Strathclyde
   address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
-- code: GCU
-  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
-- code: ISD
-  address: Information Services Division, NHS Scotland, UK
+- code: IGMM
+  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland.
+HC - AXA Chair in Medical Informatics and Epidemiology
 
@@
 
-Prepared for the Shielding Policy Evaluation team and the Clinical Leads for Shielding Group (CLAGS)
-
-Correspondence: Professor Helen Colhoun 07791652665/ helen.colhoun@igmm.ed.ac.uk
-
 ```{r functions, echo=FALSE}
@@
 library(kableExtra)
-library(bookdown)
 
-options(knitr.kable.NA = '')
+options(knitr.kable.NA = '.')
 options(knitr.table.format = "latex") # adds tab: prefix to labels
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(fig.width=6, fig.pos = "H", out.extra = "")
 source("helperfunctions.R")
@@
 
+dateformat <- function(x) { # format date and strip leading zero
+     gsub("^0", "", format(x, "%d %B %Y"))
+}
+
+format.estcipv <- function(x.ci, x.pval) {
+    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
+    x <- gsub(", ", " to ", x)
+    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
+    x <- gsub("times", "\\\\times", x)
+    return(x)
+}
+
+peakdate <- casedates.allgr[which.max(avdaily), date]
+maxdeaths <- casedates.allgr[which.max(avdaily), avdaily]
+deaths.quartermax.date <- casedates.allgr[date > peakdate & avdaily < 0.25 * maxdeaths,
+                                 date][1]
+
+peakdate.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), date]
+maxdeaths.elig <- casedates.byelig[listedgr3=="Eligible for shielding"][which.max(avdaily), avdaily]
+deaths.elig.quartermax.date <- casedates.byelig[listedgr3=="Eligible for shielding" &
+                                         date > peakdate.elig &
+                                         avdaily < 0.25 * maxdeaths.elig,
+                                         date][1]
+
+r <- as.numeric(gsub(" .*", "", table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5]));
+
+pc <- gsub(".*\\(", "", table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2])
+pc <- 0.01 * as.numeric(gsub("\\\\.*", "", pc))
+
 ```
 
+ \newpage
+
+# Abstract
+
+* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent hospital exposure -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.
+
+* _Design_ - Matched case-control study (REACT-SCOT).
+
+* _Setting_ - population of Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)`.
+
+* _Participants_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice, linked with all `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland.
+
+* _Main outcome measure_ - severe COVID-19, defined as cases that entered critical care or were  fatal.
+
+* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was
+`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
+in those with moderate risk conditions and
+`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
+in those eligible for shielding.
+The highest risk for severe disease was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` .
+In all risk groups severe COVID-19 was strongly associated with recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.
+The fraction of severe cases attributable to recent hospital exposure peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  at the beginning of December 2020.
+Risk of severe COVID-19 increased with the number of adults in the household but decreased with the number of school-age children in the household.
+The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.
+
+* _Conclusions_ -  Mitigating the impact of the epidemic requires control of transmission in hospital settings.
+Shielding of vulnerable individuals until they have been vaccinated requires protection from transmission from other adults in the same household.
+For solid organ transplant recipients, who are at very high risk and in whom the efficacy of vaccines is uncertain, vaccination should be offered to household contacts.
+
 # Background  
-The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test being recorded on March 1st.
-Extreme social distancing measures were soon imposed, as in many other countries, with subsequent relaxation now that the effective reproduction number of cases is below 1.
-The overall policy is one of suppression and if monitoring indicates a second wave of COVID-19, more stringent measures are likely to be be re-imposed.
+The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020.
+Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.
+Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].
+Those with "very high risk" conditions were designated as clinically extremely vulnerable and eligible for shielding.
 
-Early in the epidemic it was realised that certain people would likely be extremely vulnerable to severe COVID-19 disease including those receiving certain immunosuppressant drugs or with other conditions.
-Therefore Public Health Scotland  instituted a process of identifying such persons through their primary care physician and through searching of administrative databases  (see appendix for full protocol).
-Those considered eligible then received a letter detailing the need for self isolation and the support they would receive to achieve this.
-These letters were sent out over several weeks between xx-and xx.
+The shielding programme issued individual letters to eligible persons advising on the need for social distancing, recommending strict isolation and giving information on the provided support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
+Whether the programme had an impact on the infection rates in those shielded is unclear.
 
-An initial survey of ~12000 shielding eligible persons indicates that in general some degree of compliance was high but often partial and it varied across different socio-demographic groups.
-
-Here we take the original listing of ~ 180,228 persons deemed eligible for shielding at the outset of the epidemic.
-We examine the cumulative incidence of any evidence of COVID-19 and any serious COVID-19.
-using a case control design of all serious COVID-19 cases and age and sex matched controls in Scotland we examine the odds of disease associated with being shielding eligible.
-We explore whether there is any evidence that shielding reduced risks in these persons.
-
-Using data from the period in which shielding was then stopped ( 1 August 2020) and the next instruction issued to take extra  precautions we examine the number of cases that have occurred.
+The initial objective of this study was to investigate  whether the shielding letter and programme had any demonstrable impact on the risk of COVID-19 in those eligible for shielding  versus those with moderate risk or the rest of the population.
+For this we hypothesized that  a fall in  the risk ratio  for those eligible for  shielding  after the receipt of shielding letters  would be evidence of impact.
+Our second objective was to understand the role of  transmission-related factors: household composition, recent hospital exposure including out-patient consultations, and exposure related to occupations including health care [@shah_risk_2020] and teaching [@mcallister_report_2020]  on COVID-19 among those eligible for shielding.
+As the relevance of these transmission-related factors among the shielded  became clear, the objectives were broadened to investigation of the relation of severe COVID-19 to these transmission-related factors in  the remainder of the population also.
 
 # Methods
-The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+In brief; we linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) conducted by PHS that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020-1]   This case control dataset is refreshed every few weeks and  through  linkage to health and other records already contained extensive covariate data on a wide range of risk factors and settings for COVID-19 and derived variables for   "moderate risk conditions".
+Thus all cases and controls were  classifiable into three categories: no risk condition, moderate risk condition only, eligible for shielding.
+The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio  from the case-control study as described below.
 
 ## Ascertainment of eligibility for shielding 
+We used the dataset of those eligible for shielding that was created by PHS for the purposes of issuing letters.
+This dataset comprised a list of  patient identifiers and the shielding condition and the date the shielding letter was issued.
+The list had been generated  by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries,  for evidence of a list of agreed diseases and conditions  and supplemented by data submitted  to PHS following direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].
 
-To identify  all those nationally deemed eligible for shielding we took the list generated by PHS at the outset of the epidemic (see https://hpspubsrepo.blob.core.windows.net/hps-website/nss/3008/documents/1_covid-19-search-criteria-highest-risk-patients.pdf  for details) .
+The categories designated as eligible for shielding were as listed below [@health_protection_scotland_search_2020; @kanani_caring_2020]:
 
-In brief the following groups were identified
+1. Solid organ transplant recipients
+2. People with cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
+3. People with severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
+4. People with rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
+5. People on immunosuppression therapies sufficient to increase risk of infection
+6. Women who were pregnant with heart disease
+7. Other patients identified by clinicians as requiring shielding advice. 
+8. Those on renal dialysis and those who had had a splenectomy (included from 11 May 2020)
 
-* Solid organ transplant recipients
+The dataset of  those eligible for shielding process  has been regularly refreshed.
+Here we used the dataset comprising  `r nrow(shielded.full)` individuals identified up to `r dateformat(lastdate.specimen)`.
+The first batch of  shielding letters was sent on `r dateformat(batches$Date.Sent)[1]`.
+Further batches were issued on a weekly basis and the programme was paused on  1 August 2020.
+On 25 November 2020 a further letter was issued with "extra protection level advice for people at highest risk" based on the current protection level for the population level in that area.
 
-* People with specific cancers
+## Ascertainment of cases and sampling of controls
+Case ascertainment for the REACT-SCOT study has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS) that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general hospital discharges including day cases and is ICD-10 coded.
+All these databases use the Community Health Index (CHI) number as identifier.
+The CHI database includes age, sex, postcode and care home status, and can be queried to extract numbers of adults and children in the household.
 
-* People with severe respiratory conditions: cystic fibrosis, severe asthma, severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
+Cases of COVID-19 were defined as those with a positive  nucleic acid test for SARS-CoV-2 in ECOSS, a hospital discharge diagnosis of COVID-19 in SMR01, or a death registration with mention of COVID-19 anywhere on the death certificate.
+The presentation date was assigned as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those without a positive test result ascertained through hospital discharge records, and as fourteen days before the date of death for those without a positive test result ascertained through death certificates.
+Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r dateformat(maxdate.specimen)` for test results and `r dateformat(maxdate.death)` for deaths.
 
-* People with rare diseases
+Severe COVID-19 was defined by entry to critical care within 21 days of presentation date, or fatal outcome.
+Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
+Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.
+This definition of severe cases was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
 
-* People on immunosuppression therapies
+For each case of COVID-19, up to 10  community controls matched for sex, one-year age band and primary care practice were selected from the CHI database, and assigned the same presentation date as the case.
+With this incidence density sampling design, it is possible and correct for an individual to appear more than once as a control and subsequently as a case.
 
-* People who are pregnant with significant heart disease, congenital or acquired.
+## Linkage of cases and controls to demographic and morbidity data
+Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020-1].
+Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.
+We used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic airway disease, chronic kidney disease, disabling neurological conditions, and immune deficiency or suppression.
+The codes used are as described previously  [@mckeigue_rapid_2020-1].
+Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
+Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].
+Quintile 5 is the least deprived.
 
-* People on renal dialysis and who have had a splenectomy were also added ** to which group?
+# Transmission-related risk factors 
+Through linkage of the CHI database to the Unique Property Reference Number it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.
+Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.
+Linkage to occupational status for health-care workers and teachers was undertaken as described elsewhere [@shah_risk_2020; @mcallister_report_2020].
 
-* A 7th group for primary and secondary care clinicians to identify patients who should be advised to shield.
-
- As noted in the search criteria document cited above patients were identified via searches of hospital discharge and prescribing records, as well as other sources including regional cancer chemotherapy databases, NHS Blood and Transplant and individual specialist clinician lists.
-
-The list contained the CHI number which enabled linkage to other datasets. it also contained# the reason for eligibility and the date on which the shielding letter was sent to the person concerned.
-
-## COVID-19 case ascertainment and definition
-For the total Scottish population "Evidence of  any detected COVID-19" was obtained as having had a positive  nucleic acid test  for SARS-CoV-2 in  the national virology laboratory database,  or having had a hospital discharge code for COVID-19 in those with no positive test, or a death with COVID-19 mention anywhere on the death certificate even where testing had not been  positive.
-The databases used were  the Electronic Communication of Surveillance in Scotland (ECOSS) database  that captures all virology testing in all NHS laboratories nationally,  the National Records of Scotland (NRS) death registrations data, the RAPID
-database which is a daily update of hospitalisation nationwide and Scottish Morbidity Records-01 which captures hospital discharges  and discharge diagnostic codes nationally.
-
-Databases were queried from March 1st 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to the `r format(max(cc.all$SPECIMENDATE), "%d %b %Y")` for test results and `r format(max(cc.all$DATE_OF_REGISTRATION, na.rm=TRUE), "%d %b %Y")` for death registrations.
-
-All these databases use the  Community Health Index (CHI) identifier.
-The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.
-The CHI database is the master database for the CHI number and  contains several fields including age, sex, residential postcode, and residential care home status.
-
-Among all those with any evidence of detected COVID-19 we then defined severe disease as that requiring critical care or  fatal.
-All admissions to intensive (ICU), high dependency (HDU) or combined ICU/HDU care (hereafter critical care unit-CCU) were obtained by linkage using the CHI number  to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
-Fatal COVID-19 was defined  based on their being a mention of  COVID-19 anywhere  on the death certificate regardless of prior testing or not, OR death within 28 days of testing positive  (or of the admission date for those SMR01 ascertained cases) regardless of whether COVID-19 was mentioned on the death certificate or not.
-Note that these definitions of COVID-19 deaths are the official ones used by the  National Records of Scotland and Public Health Scotland, respectively.
-
-## Ascertainment of controls
-For every case i.e.
-person with "any evidence of COVID-19" "as defined above, we selected 10 controls matched for sex, one year age band and attending the same GP practice  and alive on the same date of the index case , test, two weeks prior to index case admission or two weeks prior to death for cases ascertained through testing, SMR01 admission data, or NRS death data respectively.
-The sampling frame for controls was the Community Health Care Index.
-
-## Demographic data
-For all those on the shielding list   and for all COVID-19 cases and controls i)socioeconomic status was evaluated using the postcode-based indicator the Scottish Index of Multiple Deprivation (SIMD) using the residential postcode in the CHI database [@fraser_scottish_2020].
-Postcodes were allocated to which quintile of the socioeconomic distribution of Scottish residential datazones they belonged.
-(ii) Ethnic group
-was captured by linkage to prior SMR01 and SMR00 and SMR04  records   in which the
-patients self assigns ethnicity according to the groups that are used in the
-national census. iii) Residential care home status for those with diabetes was available via the
-care home flag for postal address in the CHI database.
-
-## Other morbidity and drug prescribing
-For all those on the shielding list   and for all COVID-19 cases and controls, so as to ascertain other comorbidities the  general hospital admissions data in the  Scottish Morbidity Record 01 (SMR01),  and  all discharge codes over the past five years and Anatomical Therapeutic Chemical codes [@who_anatomical_2016] from Prescribing Information Systems (PIS) data for the past three years were extracted and used to define co-morbid conditions and prior drug exposures .
-
-We first extracted history of a specific list of conditions and drug classes that have been included as moderate risk conditions for COVID-19 by public health agencies hereafter termed "listed conditions" [@nhs_whos_2020].
-These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record, or a dispensed prescription in which that diagnostic code is the only indication (see Supplementary Materials for details).
+We used the Scottish Morbidity Records SMR01 and SMR00 (outpatient attendance) to examine possible associations with exposure to health-care facilities.
+We defined the variable "recent hospital exposure" as any hospital in-patient stay or face-to-face out-patient consultation from 5 to 14 days before presentation date.
+Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.
+Our intention was to capture all cases that *could have been* nosocomial rather than the more restrictive definition  used in Scotland and England for "probable healthcare -associated infection" as illness onset (or first positive specimen date at least 8 days after admission  which aims to  identify cases that are unlikely to have been infected outside hospital [@public_health_england_covid-19_2020-1].
 
 ## Statistical methods
-\begin{itemize}
+The analyses presented here focus  mainly on severe COVID-19 as this case definition does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is also likely to differ between those eligible for shielding and those ineligible.
+Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.
 
-\item First we tabulated the shielding list by age-group  sex and reason  for shielding.
+Cases and controls were grouped into three categories: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
+Rate ratios for severe COVID-19 associated with eligibility for shielding are calculated with "no risk condition" as reference category.
+All rate ratios were estimated by conditional logistic regression.
+As the  community controls were drawn by incidence density sampling and matched for age, sex, and general practice, this controls for these variables and for calendar time.
 
-\item We then tabulated the \% ( and CI for the \%) of those on the shielding list with "Any evidence of COVID-19" as described above by age and sex   and reason for shielding as well as overall.
+To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.
+The fraction of severe cases attributable to hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.
+where \ensuremath{p_c} is the frequency of exposure in cases in that time window and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].
+Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.
 
-\item We also tabulated the \% of those on the shielding list with "Severe i.e. CCU treated or fatal COVID-19 " as described above, by age and sex   and reason for shielding as well as overall.
-
-\item To compare risks in shielded persons before and after shielding was introduced For each shielded person we  marked up their longitudinal record  from Feb 7th with date of shielding letter, date of stop shielding letter (Aug 1st)  and any date of death with the record being censored at data of death.
-We then computed the number of cases of any evidence of COVID-19  per total  person days in pre shielded time, then shielded time , then post shielded time.
-We repeated this for severe cases.
-
-\item Among the shielding to understand whether there are other determinants of becoming a case, we compared the  \% with each of the  other listed conditions,  the \% with more than 1 listed condition, the number of prescribed ATC drug classes, the \% in  residential care home,  the  SIMD  distribution and ethnicity  distribution in those who became cases and those who did not.
-
-\item To provide a comparator for the background population,  we used  the 2019 age band and sex - mid year estimates for the general population and the shielding status in cases to derive  the \% ( and 95\%CI) with any evidence of COVID-19 and severe COVID-19 in the same age and sex groups in the general population  (DAVID this is available from STUART).
-We then used logistic regression of the one year age band counts of cases and non cases in the shielded and general population to compute and OR associated with shielding.
-
-\item The odds of "exposure "to eligibility for shielding in cases  versus  controls was also  quantified as the conditional odds ratio  from a conditional logistic regression.
-We did this because it was easy to extend this regression was then extended to include SIMD, ethnicity, residential care home status,  the "listed "COVID-19 risk conditions.
-
-\item Since the calculation in 4.
-above can be confounded by overall background trends in the Re to further evaluate if there is any evidence that shielding reduced risk, The odds of "exposure "to eligibility for shielding in cases  versus  controls  BY WEEK of epidemic was also quantified as the conditional odds ratio  from a conditional logistic regression.
-
-\end{itemize}
-
-Analyses were conducted using the R statistical package .
+Sliding windows of size  `r winsize.casedates`, `r winsize.hosp`, and `r winsize.fatality` days were used to plot the number of severe cases, the frequency of recent hospital exposure, and the rate ratios associated with hospital exposure and household composition.
+In the plots of exposure frequency and rate ratio the data points from 1 June 2020 to 30 September 2020 are omitted as the numbers of cases and controls are too low for estimates of frequencies and rate ratios to be accurate.
 
 # Results
+A methodological concern was that unrecorded filtering of fatalities while the list of those eligible for shielding was being prepared could have led us to misclassify early fatal cases among those eligible for shielding as ineligible, leading to underestimation of the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding.
+As described in the Supplementary Material, by examining the time course of the fatal case proportion we were able to establish that such misclassification was unlikely in cases presenting after the second week of March 2020.
 
-Table \ref{tab:shieldgroups} shows the frequency of shielding categories. The 6 women pregnant with heart disese were allocated to the "clinician identified" category.
+## Shielding eligibility and risk of COVID-19
+Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
+To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women who were identified as in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.
 
-Table \ref{tab:letterdates} shows a breakdown by date of sending letter
+The largest risk category was those with severe respiratory disease.
+Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories.
+Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters on April 3rd 2020, but most of those in other risk categories were included in later batches.
 
-Table \ref{tab:afterletter} shows cases among shielded individuals classified by whether estimated date of infection was before or after the shielding letter.
+Table \ref{tab:shieldcasegroup} shows  among the shielding group the number in each eligibility category by case status classified as not diagnosed, not severe or severe COVID-19.
 
-Table \ref{tab:regression.allcases} shows the estimated rate ratios for any diagnosis of COVID-19
+Overall among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
+Those on immunosuppressants had the lowest risks.
 
-Table \ref{tab:regression.severecases} shows the estimated rate ratios for severe COVID-19.
+Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as a ribbon plot.
+The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak by `r dateformat(deaths.quartermax.date)`.
+After each peak, the number of daily cases fell more slowly in those eligible for shielding than in those without risk conditions.
+
+## Rate ratio for severe COVID-19  by risk condition group   
+Of those eligible for shielding who were included in the case-control dataset, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.
+In all subsequent case-control analyses care home residents and their matched controls have been excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.
+
+Table \ref{tab:regressionseverecasesnocare} shows the rate ratios for severe COVID-19 associated with shielding eligibility and with presence of a moderate risk condition, with those without risk conditions as reference category.
+The univariate rate ratio for severe disease was
+`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
+in those with moderate risk conditions and
+`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
+in those eligible for shielding.
+Among those eligible for shielding, solid organ transplant recipients were the group at highest risk, with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` for severe COVID-19.
+On multivariate adjustment rate ratios for all these groups remained high.
+
+### Time course of the rate ratio associated with eligibility for shielding 
+Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 n the  shielding and moderate risk categories, with those without risk conditions as reference category.
+These rate ratios are computed over `r winsize`-day windows of presentation dates.
+The rate ratio associated with eligibility for shielding increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
+Therefore there was no evidence that  the  shielding letters reduced the rate ratio associated with shielding eligibility.
+A similar trend of rate ratio over time was found for those with moderate risk conditions.
+
+## Associations of severe COVID-19 with transmission setting  and occupation -related factors among those eligible for shielding and in the overall population
+Supplementary Table \ref{tab:shieldedonlyregressionseverecasesnocare} shows associations with risk factors among those eligible for shielding only, with severe respiratory disease (the largest category) as reference category.
+The rate ratio increased with the number of adults in the household  but not with the number of children.
+The strongest risk factor for severe COVID-19 among those eligible for shielding was recent hospital exposure, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 6])` in the multivariable model.
+Among those eligible for shielding who were cases of severe COVID-19, the proportion with recent hospital exposure was `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2]` had a recent hospital exposure.
+The fraction of severe cases attributable to recent hospital exposure can thus be calculated as `r round(100 * pc * (r -1) / r)`%.
+
+Table \ref{tab:regressionseverecasesnocare}  shows the association with risk factors in the case control study overall.
+The strongest risk factor (in terms of deviance explained) for severe COVID-19 in the general population was also recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.
+As shown in Figure \ref{fig:plotexphosp} (middle panel)  the rate ratio associated with recent hospital exposure increased from `r round(coeffs.hosp.timewindow[date==as.Date("2020-04-01"), rateratio], 1)` in the time window with mid-point 1 April to `r round(coeffs.hosp.timewindow[date==as.Date("2020-05-01"), rateratio], 0)` in the time window with mid-point 1 May.
+Of the `r cc.severe[CASE==1 & hosp.recent==TRUE, .N]` cases who were classified as having recent hospital exposure, only `r cc.severe[CASE==1 & hosp.recent==TRUE & hosp.recent==TRUE & prob.hcai==TRUE, .N]` met the criteria for probable health-care associated infection.
+
+Table \ref{tab:regressionseverecasesnocare} also show that the risk of severe COVID-19 increased with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.
+The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.
+The other demographic factor associated with increased risk was socioeconomic deprivation: the rate ratio in the least deprived quintile compared with the most deprived quintile was `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 5], table.shielded.severecases.nocare["5 - least deprived", 6])`.
+
+As shown in Figure \ref{fig:plotrateratio} (lower panel), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease associated with adults remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
+
+As reported previously, in comparison with other occupations patient-facing health-care workers were at higher risk of severe disease with a rate ratio of `r format.estcipv(table.shielded.severecases.nocare["Health care PF", 3], table.shielded.severecases.nocare["Health care PF", 4])`, and teachers were at lower risk with a rate ratio of `r format.estcipv(table.shielded.severecases.nocare["Teacher", 3], table.shielded.severecases.nocare["Teacher", 4])`.
+
+As shown in Figure \ref{fig:plotexphosp}  (top panel) the frequency of recent hospital exposure in controls fell when restrictions were imposed in March-April 2020 and November-December 2020, but remained higher in those eligible for shielding than in those with moderate risk conditions or no risk conditions.
+It is the frequency of exposure in controls that is relevant to estimating background exposure levels, as the frequency of exposure in cases depends on the rate ratio.
+The population attributable risk fraction of severe cases associated with recent hospital exposure reached a peak in the first wave of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% at the beginning of May, declined rapidly over the next few weeks, and increased again to `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  at the beginning of December Figure \ref{fig:plotexphosp} (bottom panel).
+For comparison, over the period of study the population attributable risk fraction of severe cases associated with the narrow definition of "probable health-care associated infection" was only `r round(100 * hcai.popattrfrac, 1)`%.
 
 # Discussion
+## Summary and interpretation of findings
+Key findings of this study were: (i) that the rate ratios for  severe COVID-19 associated with shielding conditions remained high through the epidemic; (ii) that recent hospital exposure and  the number of adults in the household were important  risk factors for severe COVID-19 among those eligible for shielding and also in the general population; (iii) that among those not resident in care homes the about 40% of severe cases among those not resident in care homes were attributable to hospital attendance even in the second wave.
 
-## Summary of findings
+In  comparison with those with no risk condition, the highest rate ratio for severe COVID-19 was the `r round(as.numeric(gsub(" .*", "", table.shielded.severecases.nocare["Solid organ transplant", 3])))`-fold rate ratio in solid organ transplant recipients.
+For other conditions deemed eligible for shielding, the rate ratios were between 4 and 7, compared to the rate ratios of about 3 associated with conditions designated as moderate risk, which include heart disease, diabetes, chronic kidney disease and disabling neurological conditions.
+Other studies of outcome of COVID-19 in solid organ transplant recipients and other individuals using immunosuppressants have been based only on patients admitted to hospital [@avery_inpatient_2020; @andersen_association_2021].
+Such studies cannot assess the risks to immunosuppressed individuals in the population.
+
+If the advice provided and support offered in the shielding letters had been effective (beyond the messages already widely disseminated by this time and  general social distancing measures that were introduced  by then), we would have expected the rate ratio for severe disease associated with eligibility for shielding to fall within two weeks of the first letters being sent out.
+Our results show however that while presentations of severe cases fell rapidly in the general population from the beginning of April 2020, the fall in presentations among those eligible for shielding advice was delayed so that the rate ratio for severe disease associated with eligibility for shielding rose during April 2020, although the daily deaths fell during April 2020 both in those eligible and those ineligible for shielding.
+
+Recent hospital exposure was strongly associated with disease in all risk groups.
+As recent hospital exposure was more common in those with moderate risk conditions and those eligible for shielding, especially after a lockdown had been imposed, this exposure accounted for a higher proportion of severe cases in these groups.
+Thus a possible explanation for this rise in the rate ratio for severe disease  associated with eligibility for shielding during April 2020 and the subsequent rise beginning in October 2020 is that those in this group were not able to avoid exposure to hospitals during periods when hospitals were a source of infection.
+
+The association of severe COVID-19 with number of adults in the household is consistent with classic studies of other viral infections showing that secondary cases in a household, where the infecting dose is likely to be high, are more severe than index cases [@aaby_malnutrition_1988].
+The inverse association of severe COVID-19 with number of school-age children in the household and the lower risk in teachers compared with others of the same age and sex are consistent with a protective effect of past exposure to children  Lower risk in teachers has been reported previously [@mcallister_report_2020; @ludvigsson_open_2021].
+These associations are consistent with evidence that past infection with other coronaviruses generates cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020].
+However no direct studies of the protective effect of these T-cell responses have been reported.
+Although the rate ratio per child has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows.
 
 ## Comparison with previous literature
+Other reports have evaluated the theoretical possibilities for shielding but without empirical data on its effectiveness.
+A previous report [@jani_comparison_2020] was based on the population of Greater Glasgow and Clyde health board.
+As no age-standardized rate ratios were reported it is difficult to compare our results with theirs.
+In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 [@forbes_association_2020] after adjusting for covariates, but there was no association of non-fatal COVID-19 requiring critical care with living with children in this age group, and no dose-response relationship was reported.
 
-## Limitations and advantages 
-Some key advantages of our study are the population coverage, the range of
-outcomes assessed, the extensive prior risk factor and clinical history
-data.
+## Strengths and limitations 
+Strengths of our study are the population coverage, the range of outcomes assessed, and the comprehensive linkage to medical records and demographic risk factors.
+A limitation is that we do not have primary care data other than encashed prescriptions.
+Furthermore, as most immunosuppressant drugs are prescribed through hospitals where linkage to prescribing records is not yet possible, the risks associated with  specific immunosuppressant drug classes could not be investigated.
+We have no data on help with daily activities from non-resident carers as another possible source of exposure of clinically vulnerable individuals attempting to shield themselves.
 
-## Policy Implications
+The calculation of the fraction of cases attributable to an exposure provides an upper bound on the predicted effect of removing that exposure.
+The definition of "recent hospital exposure" for this study is intended to capture all those whose infection could have been acquired in hospital (sensitivity of 1), unlike the official definition of "probable health-care associated infection" which is intended to identify those whose infection was unlikely to have been acquired outside hospital (high specificity).
+This inclusive definition is appropriate as the calculation of the fraction of cases in the population that are attributable to the exposure is valid if the classification of exposure has sensitivity of 1, even if the specificity is less than 1 [@hsieh_effect_1988].
+The association of severe COVID-19 with recent hospital admission is likely to be confounded by pre-existing risk conditions.
+However adjusting for risk conditions in a multivariable analysis reduced the rate ratio associated with recent hospital exposure only slightly (from 11.5 to 9.5) which suggests that this confounding effect is modest in comparison to the causal effect of exposure to hospital.
+The large increase in the rate ratio associated with recent hospital exposure during periods when a high proportion of inpatients were hospital cases also supports causality.
 
-# Declarations
+## Policy implications 
+The data presented here have important policy implications for shielding programmes,  for ongoing attempts to reduce COVID-19 transmission, and for vaccination policy.
+
+Regarding shielding programmes, we could not find any evidence that the programme *per se* reduced COVID-19 rates  above and beyond the factors leading to the general reduction in rates in the background society.
+It is  relevant to examine why advice to shield, combined with offers of support for delivery of food and medicines, failed to protect some individuals who were identified as clinically extremely vulnerable.
+As a lockdown on the general population had been imposed on 23 March 2020, and individuals who considered themselves to be at high risk would have been likely to reduce their contact level before then, there were limited possibilities for risk in this group to be reduced further by advice to shield and offers of help in letters sent from 3 April onwards.
+We have identified two sources of exposure that are associated with severe disease and cannot easily be avoided by those advised to shield: hospital attendance as in-patient or out-patient, and sharing a household with other adults.
+We recommend that consideration should be given to special measures for to allow these  vulnerable individuals to reduce their exposure.
+This could include alternative ways of delivering care without hospital attendance, and support for other household members to co-isolate with the vulnerable individual.
+
+<!---Another important question is whether the correct risk groups were included in the shielded list.
+--->The designation of groups as clinically extremely vulnerable was necessarily  undertaken before any systematic studies of associations with severe disease could have been conducted.
+Our study confirms that all these groups have markedly elevated risks of severe COVID-19, except for pregnant women with heart disease where the numbers at risk are too small for this risk to be reliably estimated, and therefore that assigning priority to these groups is appropriate for a vaccination programme that is aimed at reducing mortality and limiting the requirement for critical care.
+<!---The inability to estimate the risk in pregnant women with heart disease in this study should not alter the   vaccination  priority currently given to pregnant women with severe acquired or congenital heart disease.
+--->
+However we note that others in the population who do not have one of these conditions may nonetheless have similarly elevated risk from multiple factors, for example people with poorly-controlled diabetes and multiple complications.
+To identify vulnerable individuals eligible for shielding or vaccination, the list of risk conditions could be supplanted by  formal risk scores.
+
+Regarding  the importance of this study for policies to reduce overall transmissions we highlight here the different picture that emerges when associations of COVID-19 with any recent hospital exposure are calculated, rather than associations with probable hospital-acquired infection as formally defined.
+Our analysis shows that focusing only on definite nosocomial inpatient-acquired infection may underestimate the impact that reducing infection risks across hospital sites could have.
+
+Our results emphasise the importance of household contacts  in transmission in those with risk conditions and raise the question of the priority that should be given to vaccinating  household contacts of vulnerable individuals.
+The Joint Committee on Vaccination and Immunization did not recommend this in a report published in December 2020, stating that "at this time there are no data on the size of the effect of COVID-19 vaccines on transmission." [@joint_commiteee_on_vaccination_and_immunisation_joint_2020].
+The Committee however recommended vaccination of  health and social care workers who are "at risk of transmitting infection to multiple vulnerable persons or other staff in a health care environment",  implying some expectation that vaccination would reduce transmission.
+
+For most of those designated as clinically vulnerable, vaccination is expected to provide protection against severe COVID-19.
+For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke  an immune response sufficient to be effective and also uncertain whether  vaccines that use a live vector can be used safely [@chong_comprehensive_2017].
+On the basis of our results, we recommend that household members of solid organ transplant recipients should be offered vaccination, without waiting for  more definitive evidence of the effect of vaccines on transmission.
+Even with vaccination of  household contacts, this group will require help with shielding until until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.
+
 ## Information governance
+This study was approved under COVID-19 Rapid Data Protection Impact Assessment​ (DPIA) 20210023 that allows Public Health Scotland staff to link the datasets.
+Datasets were de-identified before analysis.
+
+## Data availability
+The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public
+
+## Funding 
+No specific funding was received for this study.
+
+## Transparency declaration
+PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
+This manuscript has been generated directly from the source data by a reproducible research pipeline.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on GitHub.
+
+# Registration
+The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
 
@@
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG
-database, the Scottish Morbidity Record Data Team, the staff of the National
-Register of Scotland, the Public Health Scotland Terminology Services, the HPS
-COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology
-Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
-
-## Funding 
-There was no specific funder for this study.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
 
 ## Public Health Scotland COVID-19 Health Protection Study Group
-Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew 
-Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin
-Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo
-Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme
-Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade
-Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$,
-Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$,
-Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross
-Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$,
-Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna
-Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
-
-1 Health Protection Scotland (Public Health Scotland), Meridian Court, 5
-Cadogan Street, Glasgow G2 6QE.
-
-2 NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2
-6QE.
+Allan McLeod​, Alice Whettlock April Went, Beth Findlay Chris Sullivan, Ciaran Harvey, David Henderson, Edward McArdle, Eisin McDonald, Emily Griffiths, Genna Drennan, Johanna Young, Kirstin Leslie, Leonardo I Green, Louise Nicol, Melissa Llano, Nick Christofidis, Paul Bett, Ross Cameron, Theresa Ryan, Victoria Ponce Hardy.
+Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
 
@@
 
+\newpage
+
+# Figures
+```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=0.8, fig.cap="Ribbon plot of severe cases by vulnerability category and date of presentation, excluding care home residents"}
+
+theme_set(theme_gray(base_size = 12))
+
+p.byelig
+```
+
+```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Rate ratios for severe COVID-19 in those not resident in care homes, by risk group and sliding window of presentation dates.
+Line thickness is proportional to precision of estimate.
+Arrows indicate dates that shielding advice letters were sent, with thickness proportional to number of letters sent on that date.
+(b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates.
+Data points for dates from 1 June to 30 September 2020 are omitted because the numbers of cases and controls are small."}
+gridExtra::grid.arrange(p.rateratio + theme(axis.title.x = element_blank()),
+                        p.household.rateratio, nrow=2)
+```
+
+```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and sliding windown of presentation dates (b) Rate ratio asociated with recent hospital exposure.
+(c) fraction of severe cases not resident in care homes attributable to recent hospital exposure.
+Data points for dates from 1 June to 30 September 2020 are omitted because the numbers of cases and controls are small."}
+gridExtra::grid.arrange(p.hosp + theme(axis.title.x = element_blank()),
+                        p.hosp.rateratio + theme(axis.title.x = element_blank()),
+                        p.fracattr, nrow=3)
+
+```
+
+\newpage
+
 # Tables
-
-```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shieldgroups, 
+```{r shieldgroupsbyage, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::kable(table.shieldgroups.byage, 
              escape=FALSE, 
              booktabs=TRUE,
-			 label="shieldgroups",
+			 label="shieldgroupsbyage",
              row.names=TRUE,
-             caption="Shielded categories") %>%
-    kable_styling(latex_options="hold_position")
+             align=rep("r", 5),
+             caption="Shielding eligibility cohort by eligibility category and age") %>%
+    kableExtra::group_rows("Shielding eligibility category", 1, 6) %>%
+        add_header_above(c(" "=1, "Age group"=4)) %>%
+        add_footnote(label="Percentages are row percentages",
+                     notation="none") %>%
+       kable_styling(latex_options="HOLD_position")
 
 ```
+
+```{r shieldcasegroup, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.shield.casegr, 
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="shieldcasegroup",
+             row.names=TRUE,
+             align=rep("r", 5),
+             caption="Shielding eligibility cohort by case status. Severe cases are those that were fatal or entered critical care") %>%
+    kableExtra::group_rows("Shielding eligibility category", 2, 7) %>%
+    column_spec(column=1, width="4.5cm") %>%
+    column_spec(column=2, width="2.5cm") %>%
+    column_spec(column=3, width="2.2cm") %>%
+    column_spec(column=4, width="2.2cm") %>%
+    column_spec(column=5, width="2.2cm") %>%
+    column_spec(column=6, width="2cm") %>%
+        add_footnote(label="Percentages are row percentages",
+                     notation="none") %>%
+     kable_styling(latex_options=c("scale_down", "hold_position"))
+
+```
+
+\onehalfspacing
+
+```{r regressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.shielded.severecases.nocare,
+             escape=FALSE,
+             booktabs=TRUE,
+             label="regressionseverecasesnocare",
+             caption="Rate ratios for severe COVID-19 in those not resident in care homes. ", 
+             col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="2.5cm") %>%
+    column_spec(column=5, width="2cm") %>%
+    column_spec(column=6, width="2.5cm") %>%
+    column_spec(column=7, width="2cm") %>%
+    kableExtra::group_rows("Shielding eligibility category", 3, 8, bold=FALSE) %>%
+    kableExtra::group_rows("Number of school age children in household", 10, 12, bold=FALSE) %>%
+    kableExtra::group_rows("Number of adults in household", 13, 17, bold=FALSE) %>%
+    kableExtra::group_rows("SIMD quintile", 18, 22, bold=FALSE) %>%
+    kableExtra::group_rows("Occupation", 23, 26, bold=FALSE) %>%
+    kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
+    add_footnote(label=c("Percentages are column percentages for each variable",
+                         "PF, patient-facing",
+                         "Rate ratios are from conditional logistic regression models with cases and controls matched for age, sex and general care practice",
+                         "Univariate rate ratios are for models with a single covariate",
+                         "Multivariable rate ratios are for a model including all covariates"),
+                 notation="none")
+
+```
+
+\newpage
+
+\beginsupplement
+# Supplementary Text
+## Sensitivity analysis for misclassification of early fatal cases as ineligible for shielding
+As those who were notified as having died were removed from the list of those eligible for shielding before letters were sent out, and no records of this filtering were available, we examined the fatal case proportion by presentation date.
+The rationale for this was that as filtering the list of those eligible for shielding to remove those who had died would misallocate fatal cases to the ineligible for shielding category, we could estimate the date up to which this misallocation occurred by examining the fatal case proportion by presentation date.
+
+Supplementary Figure \ref{fig:plotcasefatality} compares the time series of the fatal case proportion in those eligible and ineligible for shielding, by seven-day window of presentation date.
+The fatal case proportion in those eligible for shielding reached a peak in the first week of March and fell after this, suggesting that from mid-March onwards any misallocation of fatal cases to those ineligible for shielding was minimal.
+With a median 13 days from presentation to death, and further delay in death notifications, any filtering of notified deaths would have been unlikely to misallocate fatal cases whose presentation date was after the second week of March.
+
+# Supplementary Figures
+
+```{r plotcasefatality, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8, fig.cap="Proportion of diagnosed cases that were fatal in those ineligible and eligible for shielding, by sliding window of presentation date.
+For each category, line thickness is proportional to number of cases"}
+
+p.casefatality
+
+```
+
+\newpage
+
+# Supplementary Tables
 
@@
              row.names=TRUE,
-             caption="Dates of sending letters") %>%
-    kable_styling(latex_options="hold_position")
+             align=rep("r", 6), 
+             caption="Dates of sending advice letters to those eligible for shielding") %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2.1cm") %>%
+    column_spec(column=3, width="1.9cm") %>%
+    column_spec(column=4, width="1.9cm") %>%
+    column_spec(column=5, width="1.9cm") %>%
+    column_spec(column=6, width="1.9cm") %>%
+    add_footnote(label="Percentages are row percentages",
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "hold_position"))
 
@@
 
-```{r afterletter, echo=FALSE, warning=FALSE, message=FALSE}
+```{r shieldedonlyregressionseverecasesnocare, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(table.after.letter,
+knitr::kable(table.shieldedonly.severecases.nocare,
              escape=FALSE,
              booktabs=TRUE,
-			 label="afterletter",
-             caption="Proportions of cases advised to shield who were diagnosed > 10 days after date of sending letter")
-
-```
-
-```{r regression.allcases, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shielded.allcases,
-escape=FALSE,
-             booktabs=TRUE,
-			 label="regression.allcases",
-   col.names=c(clean.header(colnames(table.shielded.allcases)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-            caption="Rate ratios for diagnosis associated with shielding categories, with and without adjustment for diagnosis > 10 days after sending letter") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
-    column_spec(column=1, width="3cm") %>%
+             label="shieldedonlyregressionseverecasesnocare",
+             caption="Rate ratios for severe COVID-19 associated with risk conditions (with severe respiratory disease as reference category) in those eligible for shielding and not resident in care homes",
+             col.names=c(clean.header(colnames(table.shieldedonly.severecases.nocare)[1:2]),
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
+    column_spec(column=1, width="3.2cm") %>%
     column_spec(column=2, width="2cm") %>%
@@
     column_spec(column=4, width="2.5cm") %>%
-    column_spec(column=5, width="2cm") %>%
-   column_spec(column=6, width="2.5cm") %>%
-    column_spec(column=7, width="2cm") %>%
-    kable_styling(latex_options="hold_position")
+    column_spec(column=5, width="1.7cm") %>%
+    column_spec(column=6, width="2.7cm") %>%
+    column_spec(column=7, width="1.7cm") %>%
+    kableExtra::group_rows("Shielding eligibility category", 1, 6, bold=FALSE) %>%
+    kableExtra::group_rows("School-age children in household", 8, 10, bold=FALSE) %>%
+    kableExtra::group_rows("Adults in household", 11, 13, bold=FALSE) %>%
+    add_footnote(label=c("Percentages are column percentages for each variable",
+                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
+                         "Univariate rate ratios are for models with a single covariate",
+                         "Multivariable rate ratios are for a model including all covariates"),
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "hold_position"))
 
 ```
-
-```{r regression.severecases, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shielded.severecases,
-escape=FALSE,
-             booktabs=TRUE,
-			 label="regression.severecases",
-   col.names=c(clean.header(colnames(table.shielded.severecases)[1:2]),
-                         "Rate ratio (95\\% CI)", "$p$-value",
-                         "Rate ratio (95\\% CI)", "$p$-value"),
-
-             caption="Rate ratios for severe COVID-19 associated with shielding categories, with and without adjustment for diagnosis > 10 days after sending letter") %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Adjusted"=2)) %>%
-    column_spec(column=1, width="3cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-   column_spec(column=4, width="2.5cm") %>%
-    column_spec(column=5, width="2cm") %>%
-   column_spec(column=6, width="2.5cm") %>%
-    column_spec(column=7, width="2cm") %>%
-    kable_styling(latex_options="hold_position")
-
-```

### 2021-02-18 ct.Rmd
updated for shielding paper

---
title: "Relation of SARS-CoV-2 variant B.1.1.7 to transmission-related factors and to severity of COVID-19 in Scotland: REACT-SCOT case-only linkage study [follow-up to `r dateformat(min(maxdate.death, lastdate.rapid))`]"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: Glasgow 
  - name: Jen Bishop
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}}  
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}  
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    email: helen.colhoun@igmm.ed.ac.uk (07791 652665)
    affiliation: HPS

address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
#    number_sections: false
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./biomed-central.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan

---

On behalf of Public Health Scotland COVID-19 Health Protection Study Group

```{r functions, echo=FALSE}

library(knitr)
library(kableExtra)
library(survminer)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.pos = "H", out.extra = "")
source("helperfunctions.R")

# rmarkdown::render("ct.Rmd", output_file="ct.pdf")

cases.dropout <- with(cases.all, table(severe.case, Sgene.dropout))
cases.dropout <- paste.colpercent(cases.dropout, digits=2)

tests.byct <- with(cc.all[SPECIMENDATE >= as.Date("2020-09-01")],
                   table(flag_lighthouse_labs_testing, ecoss.result))

```

 \newpage

# Abstract
* _Objectives_ - To investigate the transmission of the SARS-CoV-2 variant B.1.1.7 and its relation to the severity of COVID-19 in Scotland.  
* _Design_ - case-only study (REACT-SCOT). 
* _Setting_ - population of Scotland from  1 September 2020 to `r dateformat(min(maxdate.death, lastdate.rapid))`.  
* _Participants_ -  All `r nrow(cc.all[CASE==1])` individuals in Scotland who tested positive for SARS-CoV-2 with the ThermoFisher TaqPath assay.  
* _Main outcome measures_ - death, entry to critical care or hospitalisation within 28 days of first testing positive for SARS-CoV-2.  
* _Results_ - The B.1.1.7 variant, imputed from S gene dropout, largely replaced the original wild-type strain in Scotland over a four-week period beginning in mid-December 2020.     Among those testing positive, the B.1.1.7 variant was less common in those aged over than under 60 years, showed substantial regional variation being more common in the west than east, was  inversely associated with residence in a care home, working in health care, and recent hospital exposure than in those not exposed to these settings.   The proportion with the variant did not vary with the number of children in the household.   With adjustment for age, sex and transmission-related factors the rate ratio for each outcome associated with S gene dropout was 
`r table.death.coxph["S gene dropout", 5]` for death, 
`r table.severe.coxph["S gene dropout", 5]` for death or entry to critical care, and 
`r table.hosp.coxph["S gene dropout", 5]` for death or hospitalisation. These risk ratios showed variation over time having risen and then fallen again the the last two weeks of  the study period. <!--can u add a p value for the interaction of the rates with calendar time?-->
* _Conclusions_ - The relationship of the VOC 2020-12/01 to  hospitalisation rates and  severity is varying over time and the most recent data do not show any elevation in hospitalisation or severity with it. The data  do not support suggestions that this variant is preferentially transmitted through children.

# Background  
<!--- 4 PHE technical briefings on this page
Meera Chand, Susan Hopkins, Gavin Dabrera, Christina Achison
Wendy Barclay, Neil Ferguson, Erik Volz
Nick Loman
Andrew Rambaut
Jeff Barrett
https://www.gov.uk/government/publications/investigation-of-novel-sars-cov-2-variant-variant-of-concern-20201201
Briefing 1 -- the Kent cluster
Briefing 2 -- preliminary fatality and hospitalisation rates
Briefing 3 -- 98% specificity of SGTF by early Dec, analysis of contact tracing data
Briefing 4 -- age and sex distribution
--->
A SARS-Cov-2 variant ([lineage B1.1.7](https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563) on `gisaid.org` was designated as a variant of concern (VOC 202012/01)  in the UK in December 2020.  The variant has 23 mutations, including N501Y that is likely to affect receptor binding affinity of the spike protein.   Further characteristics of the variant were documented in a series of technical briefing notes by Public England from December 2020 onwards [@chand_investigation_2021]  The variant has spread rapidly, becoming  the commonest variant  in  cases detected in the Pillar 2 laboratories in the UK and accounting for over 90% of cases in London by early January. Analysis of contact tracing date showed higher secondary attack rates implying increased transmissibility  of this variant. One of the S gene mutations in the variant, which deletes amino acids 69 and 70 (Δ69-70), causes a reproducible failure of the S gene signal in the ThermoFisher TaqPath assay used in UK Lighthouse laboratories. Although Δ69-70 is not unique to B.1.1.7, by early  December 2020 this variant accounted for  98% of all Δ69-70 sequences. Thus S gene dropout can be used as a  proxy for B.1.1.7, allowing the epidemiology of the variant to be studied without sequencing [@chand_investigation_2021].

A modelling study suggested that "the increase in transmissibility is likely to lead to a large increase in incidence, with COVID-19 hospitalisations and deaths projected to reach higher levels in 2021 than were observed in 2020" and that control measures would not reduce the case reproduction number to less than 1 "unless primary schools, secondary schools, and universities are also closed." [@davies_estimated_2020]

Preliminary analyses by Public Health England and other groups have suggested [@horby_nervtag_2021] that the VOC causes more severe disease. The objectives of this study were to investigate the relation of variant status to exposure to children and other transmission-related factors, and to investigate whether the variant status is associated with more severe disease in Scotland.

# Methods
## Ascertainment of COVID-19 cases evaluable for B.1.1.7 status
From early in the epidemic Public Health Scotland has maintained a regularly-updated case-control study (REACT-SCOT) of all COVID-19 cases in Scotland, with each case matched to up to ten controls and extensive linkage to other health care records and routine datasets   as previously reported [@mckeigue_rapid_2020-1].  The case definition in REACT-SCOT includes all those with a positive nucleic acid test, a hospital admission with a discharge code of COVID-19  or a death certificate with mention of COVID-19.  For cases who had tested positive in a Lighthouse laboratory, cycle threshold (Ct) values for each channel were available.  The analyses presented here focus on the subset of cases evaluable for S gene dropout as described below. As these analyses did not use data on controls, the description of methods is restricted to case ascertainment and linkages in REACT-SCOT.

All cases in Scotland with a positive RT-PCR test regardless of laboratory used were identified from the Electronic Communication of Surveillance in Scotland (ECOSS)  that captures virology testing in all NHS laboratories, hub laboratories and the  Scottish Lighthouse laboratory. From 1 September 2020  to `r dateformat(max(ct$SpecimenDate, na.rm=TRUE))` the Lighthouse laboratory accounted for `r tests.byct[2]` of the `r sum(tests.byct)` positive tests in Scotland.  For this study case ascertainment was complete up to  `r lastdate.specimen` as the date of the first positive test.  Test results from the Lighthouse laboratory were available up to `r dateformat(maxdate.ct)`

## Imputation of B.1.1.7 status from S gene signal
The ThermoFisher TaqPath assay scores results as positive if there are detectable signals from at least two of the three genes: S, N, and ORF1ab.  For this study test-positive cases with a detectable S gene signal were scored as "No dropout". Test-positive cases with no S gene signal were scored as "Definite dropout" if the cycle thresholds for the ORF1ab and N genes were less than 30. All other positive test results with no signal from the S gene were scored as "Undetermined". 
 <!--Test-positive cases with no S gene signal were scored as "Definite dropout" if the Ct for the ORF1ab  gene (\ensuremath{C_{ORF}}) was less than 31, the Ct for the N gene (\ensuremath{C_{N}}) was less than 30, and \ensuremath{C_{ORF} - C_{N} > - 1}. All other positive test results with no signal from the S gene were scored as "Undetermined".   This was based on using the results of repeat tests on the same individual to reclassify borderline results as described in the Supplementary Text.

Chris R: a positive for STGF defined as  negative on the S gene and Ct < 25 for at least one of the OR and N genes.  A weak positive is negative for S gene and Ct <= 30 for both OR and N gene. --->

## Ascertainment of  hospitalisation and severity status
The classification of cases as COVID-symptomatic or asymptomatic is based on the question used in Pillar 2 testing. Using the Community Health Index (CHI) number as identifier all cases were linked to RAPID which is a daily update of hospitalisations, out-patient attendances (SMR00), the Scottish Morbidity Record 01 (SMR01) which records general /acute hospital discharges, the Scottish Intensive Care Society and Audit Group (SICSAG) database which records entry to critical care, and to deaths registered with the National Records of Scotland (NRS).  Fatal COVID-19 was defined as death within 28 days of testing positive, regardless of the recorded cause of death or as any death with COVID-19 coded as underlying cause.

The proportion of cases with fatal outcome was calculated over sliding seven-day time windows.  We denote this fraction as the fatal case proportion.  We use this term  to distinguish it from the case fatality ratio which is classically defined without linkage between diagnoses and deaths, as the ratio of deaths from the disease to diagnosed cases over a specified time window. Data were available up to  `r dateformat(lastdate.rapid)` for hospital admissions, `r dateformat(maxdate.sicsag)`  for entry to critical care and `r dateformat(maxdate.death)` for death registrations.

## Ascertainment of exposure  to  other  COVID-19 risk settings or exposures 
All cases were linked to the CHI database that contains addresses linked to the Unique Property Reference Number, care home status, and the ages of all household members.  Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].  Quintile 5 is the least deprived. Care home residence was assigned using the field in the CHI database,   augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases.

To examine possible associations with exposure to health-care facilities, we defined the variable "recent hospital visit/stay" as any hospital in-patient stay or out-patient attendance (face-to-face consultations only) from 5 to 14 days before presentation date.  Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired. This is broader than the definition used in Scotland and England for "probable healthcare -associated infection" as illness onset or first positive specimen date between 8 and 14 days after admission.  The broad definition we have used here is intended to capture all infections that might have been acquired through hospital exposure, rather than to identify individuals who are unlikely to have been infected outside hospital.

Teachers were identified as individuals of working-age registered with the General Teaching Council of Scotland and currently working, or believed to be currently working, as teachers at a Scottish school in 2020. Healthcare workers were identified using the General Practitioner Contractor Database (GPCD) and Scottish Workforce Information Standard System (SWISS) databases as described previously. [@shah_risk_2020; @mcallister_report_2020]. Prior comorbidities were derived by linkage to prescribing and SMR01 as described previously  [@mckeigue_rapid_2020-1].

## Statistical methods
The three outcomes under study were death, death or entry to critical care (severe case), and death or hospitalisation.  Follow-up for these outcomes was limited to 28 days since first testing  positive.  For each outcome a Cox regression model was fitted with days since first testing positive as the time scale, with follow-up censored at the earliest of three dates: date of first event, last date up to which records of the outcome were available through linkage, or 28 days since first testing positive.  To control for calendar time, the dataset was additionally stratified by one-week time intervals.  This is equivalent to a conditional logistic regression model where the observations are person-time intervals and there is one stratum for each combination of days since testing positive and one-week calendar time interval.  Age, sex and other covariates were included in the model as fixed effects.

Note that because the assignment of definite S gene dropout is based on a minimum level of signal (maximum Ct value) from the ORF1ab and N genes, average Ct values of these two genes in those with S gene dropout will necessarily be lower than average Ct values in those without definite S gene dropout.  As Ct values vary inversely with the quantity of viral RNA in the sample, a crude comparison of those with and without definite S gene dropout will be confounded by viral load.  For this reason associations with disease severity are presented with and without adjustment for the average Ct value of the other two genes. <!--can u insert somewhere a sensitiviy analysis of what heppens if you instead restrict to those below a given ct value threshold as other  have done? -->

# Governance/ ethics
This work was carried out under Public Health Scotland Data Protection Impact Assessment DP20210023.

# Results
Supplementary Table \ref{tab:coeffssymptoms} shows that in comparison with cases without Ct results whose first positive tests were done in NHS hospital laboratories,  cases with Ct results whose first positive tests were done in Lighthouse laboratories had a younger age profile, a higher proportion of females, a higher proportion with children at home, and a lower fatality rate.

## Frequency of S gene dropout and association with transmission-related factors
Figure \ref{fig:variantbydate} shows that from the second half of December 2020 onwards the S gene dropout rapidly replaced the original wild-type strain as the predominant strain among test-positive cases detected in the Lighthouse testing programme. <!--insert here the exact % t accotns for in the last available week of data ** readers can get these numbers off the figure. ---> The rise in median cycle thresholds of the ORF and N genes in the first half of December 2020 antedated the replacement of the wild-type strain by the variant.  As the variant replaced the original strain, the difference of about one cycle between the detection thresholds of the ORF gene and the N gene narrowed to the point where the Ct values overlapped.

Table \ref{tab:coeffsexposure} shows among all test-positive cases with presentation date from 1 December 2020 to `r dateformat(max(cases.all$SPECIMENDATE))` and non-missing values for S gene dropout status the association of S gene dropout with transmission-related factors including age, region, household composition, occupation and recent hospital exposure.   Among those testing positive, the B.1.1.7 variant was less common in those aged over than under 60 years. The odds of the variant were more than twice as high in the other three Scottish regions than in the eastern region.  The B.1.1.7 variant was inversely associated with residence in a care home [rate ratio `r table.coeffs.exposure["Care home", 3]`], with working in health care [rate ratio `r table.coeffs.exposure["Health care, PF", 3]`], , and with recent hospital exposure [rate ratio `r table.coeffs.exposure["Recent hospital visit/stay", 3]`].  The proportion with the variant did not vary with the number of children in the household. Supplementary Figure \ref{fig:plotrateratioSgene} shows that the associations with region and healthcare status did not vary much over time.

## Relation of symptom status to S gene dropout
Table \ref{tab:coeffssymptoms} shows that after adjustment for cycle threshold and other covariates, the proportion of cases with symptoms did not differ between those with the variant and those with the wild-type strain.

## Relation of outcome of infection to S gene dropout
<!--can ytou give a little ore basic data here please ie fill this in. Of all test positive cases examined here between insert date and insert date  the median (interquartile range) of follow up time up to  COVID-19  death or being censored for other death or censored at 28 days post test was xxx in those with and xxx in those without the B.1.1.7 variant  --> the `r cases.all[, .N]` test-positive cases with S gene dropout status, `r cases.all[hosp28==1, .N]` were hospitalised or dead within 28 days of testing positive,
`r cases.all[severe28==1, .N]` were admitted to critical care or dead within 28 days, and `r cases.all[death28==1, .N]` died within 28 days. <!--can you also insert these basic data The incidence rate of death was xxx / person days at risk versus xxx /person days at risk in those with adn without the variant. For the composte outcome of hospitalsiation or death these rates were xxx and xxx respectively and for admssion to CCU or death they were xxx and xxx. -->

Figure \ref{fig:survivalcurves} shows the crude survival curves for these three outcomes and suggests slightly lower survival for all three outcomes for those with the B.1.1.7 variant .

 With adjustment for age, sex and transmission-related factors the rate ratio associated with S gene dropout was 
`r table.death.coxph["S gene dropout", 5]` for death within 28 days of testing positive (Table \ref{tab:coxfatal}),
`r table.severe.coxph["S gene dropout", 5]` for death or entry to critical care within 28 days of testing positive (Table \ref{tab:coxsevere}), and 
`r table.hosp.coxph["S gene dropout", 5]` for death or hospitalisation within 28 days of testing positive (Table \ref{tab:coxhosp}).  The average of Ct values of the ORF1ab and N genes was strongly and inversely associated with both severe disease and hospitalisation.

<!--need to run react scot risk score and further adjust for it in suppementary tables and coment son that here  - need also to show how the react score and other characterstics DIFFERENCES  ie age residential care hcw etc etc varies over the study period   INCLUDE in this differences in the CT value between the strains over time  -->

Supplementary Figure \ref{fig:plotrateratioSgene} shows the rate ratio for death or hospitalisation associated with S gene dropout by sliding 7-day time window of presentation date, adjusted for the  covariates in Table \ref{tab:coxhosp}.  The rate ratio rose from the last week of December until the second week of January, then began to fall. We note that the rate ratios for cases presenting in January are based on follow-up only to `r dateformat(min(maxdate.death, lastdate.rapid))` though the Cox regression allows for censoring.

# Discussion
## Summary of findings and comparison with other reports
### Transmission in different settings
These data show that transmission of the B.1.1.7 variant  in Scotland rose from early December 2020 and that by the middle of January 2021 it had largely replaced the original strain.  From the calendar time-stratified associations of variant status with demographic variables, there is no evidence that higher transmission of the variant is associated with any demographic stratum of the population.  It is however evident that during the period of study the transmission of the variant was inversely associated with exposure to hospitals or residence in care homes, perhaps because the variant had not yet spread to these sites.

Our results showing no association of variant status with the number of children in the household are consistent with a recent report from the Office of National Statistics COVID-19 Infection Survey (CIS), a  randomly selected swab surveillance survey in the UK that found no evidence of differential growth of the variant between children and adults [@walker_increased_2021].  Our results showing no difference in the proportion of symptomatic cases by S gene dropout status are consistent with the findings of the CIS survey.

### Severity of disease
There is no clear evidence from this study that the B.1.1.7 variant is associated with more severe disease: the 28-day rates of fatality, admission to critical care and hospital admission do not differ significantly between test-positive cases with and without the variant.  Our results on the association of S gene dropout with increased hospitalisation and severity  in the first 28 days differ from those from for England presented at the NERVTAG meeting in  January 2021, and from analysis of an earlier version of this dataset presented at the NERVTAG meeting on 5 February 2021.   With ascertainment of more recent cases and wtih longer follow-up, it is now clear that the rate ratio for hospitalisation associated with the variant varied markedly over the time period that this study covers.  
The reasons for this are not clear and we are undertaking further analyses using individual risk scores to assess confounding by susceptibility to severe disease.

### Cycle threshold
Our results show that lower average cycle threshold of the N and ORF1ab genes in the initial positive test is associated with higher rates of fatality, entry to critical care and hospitalisation.  This contrasts with PHE [guidance](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/926410/Understanding_Cycle_Threshold__Ct__in_SARS-CoV-2_RT-PCR_.pdf) that Ct values in a swab taken at a single point in time do not provide information about the trajectory or subsequent course of illness.  Monitoring of the cycle threshold may be useful in predicting future utilisation of hospital care.

## Strengths and limitations 
* Although most positive tests in Scotland are in the Lighthouse labs that use the Thermofisher TaqPath assay and these data are nationally complete,  most of the cases that are hospitalised or fatal are tested in NHS labs that do not use this assay and therefore cannot impute the B.1.1.7 variant.  As this study is based on comparisons only within those who test positive in Lighthouse labs,  comparisons of disease severity between those with and without S gene dropout should not be biased even though the cohort is unrepresentative of all test-positive cases.

* The strongest confounders of the association between S gene dropout and disease severity are likely to be calendar time and transmission-related risk factors such as recent hospital exposure. We controlled for calendar time by stratification in Cox regression models.  The comprehensive linkage of the REACT-SCOT study allowed us to adjust for recent hospital exposure, care home residence, household composition, and geographic region.

* The numbers of hospitalised cases with S gene dropout is large enough for the rate ratio for this outcome to be estimated reliably.  The numbers of fatal cases during the period that both the original strain and the variant were circulating is relatively small, so that the rate ratio cannot be estimated accurately.

## Implications for policy 
These results do not indicate that the variant has been preferentially transmitted to those who are exposed to school-age children.  They are also reassuring with respect to concerns that the variant may be associated with more severe disease.  Even if the higher transmissibility of the variant increases the incidence of infections, the effect on severe disease is likely to be small in comparison to the effect of other factors that are amenable to intervention, including overprescribing of psychotropic drugs [@mckeigue_relation_2021], transmission in hospitals[@mckeigue_relation_2021-1], and vaccination of vulnerable individuals.

# Declarations
## Data availability
The component datasets used here are available via the Public Benefits Privacy Panel for Health and Social Care at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.

## Funding 
No specific funding was received for this study.

## Transparency declaration
PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained. This manuscript has been generated directly from the source data by a reproducible research pipeline. All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on GitHub.

# Registration
The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).

## Conflicts of interest
The authors declare no conflicts of interest.

## Acknowledgements
We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.

## Public Health Scotland COVID-19 Health Protection Study Group
<!--this group needs to be updated-->

Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$

1 Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.

2 NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2
6QE.

# References
<div id="refs"></div>

\newpage

# Figures

```{r plotvariantbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1.25, fig.cap="\\label{fig:variantbydate}(a) Test-positives by date: area plot showing proportion of S gene dropouts in positive tests with reporting of cycle threshold. (b) Median cycle thresholds (reverse scale) in positive tests by date "}

theme_set(theme_gray(base_size = 10))

gA <- ggplotGrob(p.variantbydate)
gB <- ggplotGrob(p.mediansbydate)
#gC <- ggplotGrob(p.bysource)
grid::grid.newpage()
grid::grid.draw(rbind(gA, gB))

```

```{r plotsurvivalcurves, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1.25, fig.cap="\\label{fig:survivalcurves}28-day survival curves comparing those with and without S gene dropout, censored at last date of observation of relevant outcomes: (a) Survival till death; (b) Survival till death or admission to critical care; (c) Survival till hospital admission."}

theme_set(theme_gray(base_size = 10))

arrange_ggsurvplots(list(psurv.fatal, psurv.severe,
                         psurv.hosp),
                    print=TRUE,
                    ncol=1,
                    nrow=3)

```

\newpage

# Tables

```{r coeffsexposure, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.coeffs.exposure, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coeffsexposure",
             row.names=TRUE,
             align=c("r", "r", "r", "l", "r", "l"), 
             col.names=c(clean.header(colnames(table.coeffs.exposure)[1:2]),
                         "Odds ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Odds ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Logistic regression of S gene dropout on transmission-related factors") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    add_footnote(label=c("Excluding those with undetermined status for S gene dropout",
                         "Unconditional logistic regression with S gene dropout as dependent variable",
                         "Univariate odds ratios are for models with single variable + week of specimen as categoric variable", 
                         "Multivariable odds ratios are for model with all covariates + week of specimen as categoric variable",
                         "PF, patient-facing"),
                 notation="none") %>%
    kableExtra::group_rows("Age group", 1, 5, bold=FALSE) %>%
    kableExtra::group_rows("Region", 7, 10, bold=FALSE) %>%
    kableExtra::group_rows("Number of adults in household", 13, 17, bold=FALSE) %>%
    kableExtra::group_rows("Number of school-age children in household", 19, 21, bold=FALSE) %>%
    kableExtra::group_rows("Occupation", 22, 25, bold=FALSE) %>%

    kable_styling(latex_options=c("scale_down", "hold_position"))
```
\newpage

```{r coxfatal, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.death.coxph, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxfatal",
             row.names=TRUE,
             align=c("r", "r", "r", "l", "r", "l"), 
             col.names=c(clean.header(colnames(table.death.coxph)[1:2]),
                        "rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Death within 28 days of testing positive: Cox regression on S gene dropout and covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    add_footnote(label=c("Excluding those with undetermined status for S gene dropout",
                         "Care home residents are excluded", 
                         "Stratified by days since testing positive and calendar week",
                         "Multivariable rate ratios are for model with all covariates"),
                 notation="none") %>%
    kable_styling(latex_options=c("scale_down", "hold_position"))

```

```{r coxsevere, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.severe.coxph, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxsevere",
             row.names=TRUE,
             align=c("r", "r", "r", "l", "r", "l"), 
             col.names=c(clean.header(colnames(table.severe.coxph)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Death or entry to critical care within 28 days of testing positive: Cox regression of time to failure on S gene dropout and covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    add_footnote(label=c("Excluding those with undetermined status for S gene dropout",
                         "Stratified by days since testing positive and calendar week",
                          "Univariate rate ratios are for models with single variable + week of specimen as categoric variable", 
                        "Multivariable rate ratios are for model with all covariates"),
                 notation="none") %>%
    kable_styling(latex_options=c("scale_down", "hold_position"))

```

```{r coxhosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.hosp.coxph, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxhosp",
             row.names=TRUE,
             align=c("r", "r", "r", "l", "r", "l"), 
             col.names=c(clean.header(colnames(table.hosp.coxph)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Death or hospitalisation within 28 days of testing positive: Cox regression on S gene dropout and covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    add_footnote(label=c("Excluding those with undetermined status for S gene dropout",
                         "Stratified by days since testing positive and calendar week",
                         "Univariate rate ratios are for models with single variable + week of specimen as categoric variable", 
                         "Multivariable rate ratios are for model with all covariates + week of specimen as categoric variable"),
                 notation="none") %>%
    kable_styling(latex_options=c("scale_down", "hold_position"))

```

\newpage

\beginsupplement

# Imputation of S gene dropout

Supplementary Figure \ref{fig:plotbyCt} shows that absence of the S gene signal was associated with lower cycle threshold (Ct) for the N gene than for the ORF1ab gene.  We can identify no obvious artefactual explanation for this finding based on published documentation of the assay.

We then plotted the subset of "Undetermined" results in those individuals who had had another test that had been definitively scored (Supplementary Figure \ref{fig:plotcheck}.  Inspection of this plot showed that without loss of specificity the Ct cutoff  for the ORF gene could be increased to 31 and that the criteria for S gene dropout should include an additional restriction that \ensuremath{C_{ORF} - C_{N} > - 2}.  For comparability, the results presented here are based on the standard definition.

\newpage

# Supplementary Figures

```{r plotSgenebyCt, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1, fig.cap="\\label{fig:plotbyCt}Relation of S gene signal to ORF and N gene Ct values"}

p.dropout

```

\newpage

```{r checkcutoff, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1, fig.cap="\\label{fig:plotcheck}Test results with S gene signal absent that were originally classified as Undetermined (\\ensuremath{C_{ORF} >= 31 or \\ensuremath{C_{N} >= 31 or  \\ensuremath{C_{N} - C_{ORF} < 1.5}), colour coded by result at repeat test. New separation lines at \\ensuremath{C_{N} - C_{ORF} = 3} and ORF = 33, used in this analysis, are shown in green"}

p.check
```

```{r plotrrSgene, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1.4, fig.cap="\\label{fig:plotrateratioSgene} (a) Rate ratio for 28-day hospitalisation or death associated with S gene dropout. Line thickness is proportional to precision of estimate. (b) Rate ratio for S gene dropout associated with region (Eastern region as reference category). (c) Rate ratio for S gene dropout associated with health care worker status (other occupations as reference category)"}

gA <- ggplotGrob(p.rateratio)
gB <- ggplotGrob(p.region)
gC <- ggplotGrob(p.hcw)
grid::grid.newpage()
grid::grid.draw(rbind(gA, gB, gC))

```

\newpage

# Supplementary Tables

```{r casesbymonth, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.cases.bymonth, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="casebymonth",
             row.names=TRUE,
             align=rep("r", 5),
             caption="Cases by month") %>%
    add_footnote(label=c("Restricted to cases with specimen date no earlier than 1 Sep 2020",
                         "Outcome is not yet evaluable for cases with specimen date after 31 December 2020"),
                 notation="none") %>%
    kableExtra::group_rows("All cases", 1, 8) %>%
        kableExtra::group_rows("Test-positive cases", 9, 11, bold=FALSE) %>%
        kableExtra::group_rows("Cases with Ct results", 12, 16, bold=FALSE) %>%
        kableExtra::group_rows("Fatal cases and cohort fatality rates by S gene dropout status", 17, 18, bold=FALSE) %>%
        kable_styling(latex_options=c("scale_down", "hold_position"))

```

\newpage

```{r withCt, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.withCt, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="withCt",
             row.names=TRUE,
             align=rep("r", 5),
             caption="Comparison of test-positive cases with and without cycle threshold results") %>%
    add_footnote(label=c("Restricted to cases with specimen date no earlier than 1 Sep 2020"),
                 notation="none") %>%
    kableExtra::group_rows("Age group", 1, 5, bold=FALSE) %>%
        kableExtra::group_rows("Number of adults in household", 9, 13, bold=FALSE) %>%
        kableExtra::group_rows("Number of school-age children in household", 15, 17, bold=FALSE) %>%
        kableExtra::group_rows("Occupation", 18, 21, bold=FALSE) %>%
        kableExtra::group_rows("Outcome", 22, 27, bold=FALSE) %>%
        kable_styling(latex_options=c("scale_down", "hold_position"))
```

\newpage

```{r surv28, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(surv28.table, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="surv28",
             row.names=FALSE,
             align=c("r", "r", "r", "r", "r", "r", "r"),
             caption="28-day survival table for three outcomes: deaths, severe cases, hospitalisations") %>%
    column_spec(column=1, width="1.5cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2cm") %>%
    column_spec(column=5, width="2cm") %>%
    column_spec(column=6, width="2cm") %>%
    column_spec(column=7, width="2cm") %>%
    kableExtra::group_rows("Totals", 30, 30) %>%
        add_footnote(label=c("Columns 2, 4, 6 contain the numbers censored by last date of follow-up on that date",
                             "All remaining test-positive cases are censored at 28 days",
                             "For each outcome, the sum of totals for censored and failures is the number evaluable at baseline",
                             "Restricted to cases with specimen date no earlier than 1 September 2020"),
                     notation="none") %>%
    kable_styling(latex_options=c("scale_down", "hold_position"))

```

\newpage

```{r coeffssymptoms, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.symptoms, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coeffssymptoms",
             row.names=TRUE,
             align=c("r", "r", "r", "l", "r", "l"), 
             col.names=c(clean.header(colnames(table.symptoms)[1:2]),
                         "Odds ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Odds ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Logistic regression of symptomatic versus asymptomatic COVID-19 on S gene dropout and covariates") %>%
        add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    add_footnote(label=c("Symptom status is from the field in the Lighthouse lab tests database",
                         "Excluding those with undetermined status for S gene dropout or presentation date after 31 December 2020",
                         "Unconditional logistic regression with symptom status as dependent variable",
                         "Univariate odds ratios are for models with single variable + week of specimen as categoric variable", 
                         "Multivariable odds ratios are for model with all covariates + week of specimen as categoric variable"),
                 notation="none") %>%
    kable_styling(latex_options=c("scale_down", "hold_position"))

```

### 2021-02-18 shielding.Rmd
updated for shielding paper

@@
 ---
-title: "Relation of severe COVID-19 in Scotland to transmission-related factors in individuals with and without risk conditions eligible for shielding support: REACT-SCOT case-control study"
+title: "Relation of severe COVID-19 in Scotland to transmission-related factors and risk conditions eligible for shielding support: REACT-SCOT case-control study"
 author:
@@
     affiliation: HPS
-  - name: David McAllister\textsuperscript{\getAff{Glasgow}}  
+  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
     affiliation: HPS 
@@
   \usepackage{pdfpages}
+  \linenumbers 
 output:
@@
 
-* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent hospital exposure -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.
+* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent exposure to hospital -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.
 
@@
 The highest risk for severe disease was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` .
-In all risk groups severe COVID-19 was strongly associated with recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.
-The fraction of severe cases attributable to recent hospital exposure peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  at the beginning of December 2020.
 Risk of severe COVID-19 increased with the number of adults in the household but decreased with the number of school-age children in the household.
 The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.
+In all risk groups severe COVID-19 was strongly associated with recent exposure to hospital (5 to 14 days before presentation date): rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.
+In a case-crossover analysis with less recent exposure (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
+The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% of severe cases at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December 2020.
 
-* _Conclusions_ -  Mitigating the impact of the epidemic requires control of transmission in hospital settings.
-Shielding of vulnerable individuals until they have been vaccinated requires protection from transmission from other adults in the same household.
-For solid organ transplant recipients, who are at very high risk and in whom the efficacy of vaccines is uncertain, vaccination should be offered to household contacts.
+* _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.
+For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts.
+<!---These results also provide compelling evidence for a substantial contribution of nosocomial infection to the burden of severe COVID-19---> Mitigating the impact of the epidemic requires control of transmission in hospital settings.
+
+\newpage
 
@@
 Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].
-Those with "very high risk" conditions were designated as clinically extremely vulnerable and eligible for shielding.
+Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters recommending strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
 
-The shielding programme issued individual letters to eligible persons advising on the need for social distancing, recommending strict isolation and giving information on the provided support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
-Whether the programme had an impact on the infection rates in those shielded is unclear.
-
-The initial objective of this study was to investigate  whether the shielding letter and programme had any demonstrable impact on the risk of COVID-19 in those eligible for shielding  versus those with moderate risk or the rest of the population.
-For this we hypothesized that  a fall in  the risk ratio  for those eligible for  shielding  after the receipt of shielding letters  would be evidence of impact.
-Our second objective was to understand the role of  transmission-related factors: household composition, recent hospital exposure including out-patient consultations, and exposure related to occupations including health care [@shah_risk_2020] and teaching [@mcallister_report_2020]  on COVID-19 among those eligible for shielding.
-As the relevance of these transmission-related factors among the shielded  became clear, the objectives were broadened to investigation of the relation of severe COVID-19 to these transmission-related factors in  the remainder of the population also.
+The initial objective of this study was to investigate  whether shielding advice and support had reduced the risk of COVID-19.
+On the hypothesis that shielding was effective, we predicted that the rate ratio associated with eligibility for  shielding  compared to those without risk conditions would fall after the receipt of shielding letters.
+Our second objective was to understand the role of  transmission-related factors: household composition, occupations including health care [@shah_risk_2020], and recent exposure to hospital on COVID-19 among those eligible for shielding.
+As the relevance of these transmission-related factors among those eligible for shielding  became clear, the objectives were broadened to investigate the relation of severe COVID-19 to these transmission-related factors in  the general population.
 
 # Methods
-In brief; we linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) conducted by PHS that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020-1]   This case control dataset is refreshed every few weeks and  through  linkage to health and other records already contained extensive covariate data on a wide range of risk factors and settings for COVID-19 and derived variables for   "moderate risk conditions".
-Thus all cases and controls were  classifiable into three categories: no risk condition, moderate risk condition only, eligible for shielding.
+We linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case control dataset is refreshed every few weeks and linked to health records.
+These linked records are used to derive a variables for a list of designated "moderate risk conditions".
+Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.
 The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio  from the case-control study as described below.
@@
 
-1. Solid organ transplant recipients
-2. People with cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
-3. People with severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
-4. People with rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
+1. Solid organ transplant recipient
+2. Cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
+3. Severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
+4. Rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
 5. People on immunosuppression therapies sufficient to increase risk of infection
-6. Women who were pregnant with heart disease
-7. Other patients identified by clinicians as requiring shielding advice. 
-8. Those on renal dialysis and those who had had a splenectomy (included from 11 May 2020)
+6. Pregnant with heart disease
+7. Additional conditions, including people on renal dialysis, those who had had a splenectomy  and others identified by clinicians as requiring shielding advice.
 
-The dataset of  those eligible for shielding process  has been regularly refreshed.
-Here we used the dataset comprising  `r nrow(shielded.full)` individuals identified up to `r dateformat(lastdate.specimen)`.
 The first batch of  shielding letters was sent on `r dateformat(batches$Date.Sent)[1]`.
@@
 On 25 November 2020 a further letter was issued with "extra protection level advice for people at highest risk" based on the current protection level for the population level in that area.
+The list of those eligible for shielding has been regularly updated: this study is based on the list of `r nrow(shielded.full)` individuals identified up to 28 January 2021.
 
 ## Ascertainment of cases and sampling of controls
-Case ascertainment for the REACT-SCOT study has been described in detail elsewhere [@mckeigue_rapid_2020-1].
+Case ascertainment for the REACT-SCOT study has been described in detail elsewhere [@mckeigue_rapid_2020].
 Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS) that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general hospital discharges including day cases and is ICD-10 coded.
@@
 
-Severe COVID-19 was defined by entry to critical care within 21 days of presentation date, or fatal outcome.
 Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
 Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.
-This definition of severe cases was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
+Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome.
+This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
 
@@
 ## Linkage of cases and controls to demographic and morbidity data
-Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020-1].
+Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020].
 Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.
 We used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic airway disease, chronic kidney disease, disabling neurological conditions, and immune deficiency or suppression.
-The codes used are as described previously  [@mckeigue_rapid_2020-1].
+The codes used are as described previously  [@mckeigue_rapid_2020].
 Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
@@
 # Transmission-related risk factors 
-Through linkage of the CHI database to the Unique Property Reference Number it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.
+Since addresses in the CHI database have been mapped to  Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.
 Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.
@@
 
-We used the Scottish Morbidity Records SMR01 and SMR00 (outpatient attendance) to examine possible associations with exposure to health-care facilities.
-We defined the variable "recent hospital exposure" as any hospital in-patient stay or face-to-face out-patient consultation from 5 to 14 days before presentation date.
+We used the Scottish Morbidity Records SMR01 (inpatients and day cases) and SMR00 (outpatient attendance) together with the RAPID database to derive variables encoding recent exposure to hospitals.
+We defined the variable "recent hospital exposure" as any hospital in-patient stay, day case attendance or face-to-face out-patient consultation from 5 to 14 days before presentation date.
 Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.
-Our intention was to capture all cases that *could have been* nosocomial rather than the more restrictive definition  used in Scotland and England for "probable healthcare -associated infection" as illness onset (or first positive specimen date at least 8 days after admission  which aims to  identify cases that are unlikely to have been infected outside hospital [@public_health_england_covid-19_2020-1].
+Our intention was to capture all cases that *could have been* nosocomial rather than the more stringent definitions for "health-care associated COVID-19" specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
+The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.
+Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated.
+The ECDPC states that "cases with symptom onset within 14 days of discharge from a healthcare facility (e.g.
+re-admission) may be considered as community-associated, probable or definite health-care-associated-COVID-19, or to have an indeterminate association" but does not specify how cases are to be assigned to these categories.
 
 ## Statistical methods
-The analyses presented here focus  mainly on severe COVID-19 as this case definition does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is also likely to differ between those eligible for shielding and those ineligible.
+The analyses presented here focus on severe COVID-19, defined as the main outcome measure of  of the REACT-SCOT study at the design stage.
+Ascertainment of this outcome does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is likely to differ between those eligible for shielding and those ineligible.
 Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.
@@
 
+To test for causality of the association of recent exposure to hospital with severe COVID-19, we estimated rate ratios associated with recent hospital exposure (5 to 14 days before first testing positive) with less recent exposure (15 to 24 days before first testing positive) as reference category.
+This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases.
+It is a variant of the classic case-crossover design, which usually  compares in cases only the frequencies of exposure in recent and less recent time windows.
+Such an analysis can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000].
+The conditional odds ratio can thus be estimated as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only.
+The analysis that we report is a refinement of this that takes advantage of the availability of a matched control group.
+We contrast not the numbers of cases with recent exposure only and nonrecent exposure only, but the conditional odds of case status between those with recent exposure only and those with less recent exposure only.
+This controls for any difference in the population frequencies of exposure or disease incidence between recent and less recent time windows.
+
 # Results
-A methodological concern was that unrecorded filtering of fatalities while the list of those eligible for shielding was being prepared could have led us to misclassify early fatal cases among those eligible for shielding as ineligible, leading to underestimation of the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding.
-As described in the Supplementary Material, by examining the time course of the fatal case proportion we were able to establish that such misclassification was unlikely in cases presenting after the second week of March 2020.
-
 ## Shielding eligibility and risk of COVID-19
 Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
-To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women who were identified as in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.
-
+To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.
 The largest risk category was those with severe respiratory disease.
 Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories.
-Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters on April 3rd 2020, but most of those in other risk categories were included in later batches.
+Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters sent on 3 April 2020.
+Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
 
-Table \ref{tab:shieldcasegroup} shows  among the shielding group the number in each eligibility category by case status classified as not diagnosed, not severe or severe COVID-19.
-
-Overall among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
-Those on immunosuppressants had the lowest risks.
-
-Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as a ribbon plot.
+Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as an area plot.
 The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak by `r dateformat(deaths.quartermax.date)`.
-After each peak, the number of daily cases fell more slowly in those eligible for shielding than in those without risk conditions.
+After the peak, the daily number of severe cases fell more slowly in those eligible for shielding than in those without risk conditions.
 
@@
 Of those eligible for shielding who were included in the case-control dataset, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.
-In all subsequent case-control analyses care home residents and their matched controls have been excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.
+In  subsequent case-control analyses care home residents and their matched controls were excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.
 
@@
 ### Time course of the rate ratio associated with eligibility for shielding 
-Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 n the  shielding and moderate risk categories, with those without risk conditions as reference category.
+Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 in the  shielding and moderate risk categories, with those without risk conditions as reference category.
 These rate ratios are computed over `r winsize`-day windows of presentation dates.
 The rate ratio associated with eligibility for shielding increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
-Therefore there was no evidence that  the  shielding letters reduced the rate ratio associated with shielding eligibility.
-A similar trend of rate ratio over time was found for those with moderate risk conditions.
+The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding.
 
-## Associations of severe COVID-19 with transmission setting  and occupation -related factors among those eligible for shielding and in the overall population
+A methodological concern was that unrecorded removal of fatalities from the original list of those identified as eligible for shielding could have led us to misclassify early fatal cases among those eligible for shielding as ineligible and thus to underestimate the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding.
+As described in Supplementary Methods and Figure \ref{fig:plotcasefatality}, by examining the time course of the fatal case proportion we were able to establish that such misclassification of eligibility was unlikely in cases presenting after the second week of March 2020.
+
+## Associations of severe COVID-19 with transmission-related factors
+### Associations in those eligible for shielding 
 Supplementary Table \ref{tab:shieldedonlyregressionseverecasesnocare} shows associations with risk factors among those eligible for shielding only, with severe respiratory disease (the largest category) as reference category.
 The rate ratio increased with the number of adults in the household but not with the number of children.
-The strongest risk factor for severe COVID-19 among those eligible for shielding was recent hospital exposure, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 6])` in the multivariable model.
-Among those eligible for shielding who were cases of severe COVID-19, the proportion with recent hospital exposure was `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2]` had a recent hospital exposure.
-The fraction of severe cases attributable to recent hospital exposure can thus be calculated as `r round(100 * pc * (r -1) / r)`%.
+The strongest risk factor for severe COVID-19 among those eligible for shielding was recent exposure to hospital, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 6])` in the multivariable model.
+Of severe cases among those eligible for shielding, `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2]` had recent  exposure to hospital.
+Using Miettinen's formula as given above, the Population Attributable Risk Fraction  for recent  exposure to hospital can thus be calculated as `r round(100 * pc * (r -1) / r)`% of severe cases in those eligible for shielding.
+<!--Table \ref{tab:casegr} shows the relationship of hospitalisation, entry to critical care and fatality to recent hospital exposure among all diagnosed cases.-->
 
-Table \ref{tab:regressionseverecasesnocare}  shows the association with risk factors in the case control study overall.
-The strongest risk factor (in terms of deviance explained) for severe COVID-19 in the general population was also recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.
-As shown in Figure \ref{fig:plotexphosp} (middle panel)  the rate ratio associated with recent hospital exposure increased from `r round(coeffs.hosp.timewindow[date==as.Date("2020-04-01"), rateratio], 1)` in the time window with mid-point 1 April to `r round(coeffs.hosp.timewindow[date==as.Date("2020-05-01"), rateratio], 0)` in the time window with mid-point 1 May.
-Of the `r cc.severe[CASE==1 & hosp.recent==TRUE, .N]` cases who were classified as having recent hospital exposure, only `r cc.severe[CASE==1 & hosp.recent==TRUE & hosp.recent==TRUE & prob.hcai==TRUE, .N]` met the criteria for probable health-care associated infection.
-
-Table \ref{tab:regressionseverecasesnocare} also show that the risk of severe COVID-19 increased with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.
+### Associations in the general population 
+Table \ref{tab:regressionseverecasesnocare}  shows the association with risk factors in all those not resident in care homes.
+The risk of severe COVID-19 increased with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.
 The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.
 The other demographic factor associated with increased risk was socioeconomic deprivation: the rate ratio in the least deprived quintile compared with the most deprived quintile was `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 5], table.shielded.severecases.nocare["5 - least deprived", 6])`.
-
-As shown in Figure \ref{fig:plotrateratio} (lower panel), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease associated with adults remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
+As shown in Figure \ref{fig:plotrateratio} (lower panel), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease per adult in household remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
 
@@
 
-As shown in Figure \ref{fig:plotexphosp}  (top panel) the frequency of recent hospital exposure in controls fell when restrictions were imposed in March-April 2020 and November-December 2020, but remained higher in those eligible for shielding than in those with moderate risk conditions or no risk conditions.
-It is the frequency of exposure in controls that is relevant to estimating background exposure levels, as the frequency of exposure in cases depends on the rate ratio.
-The population attributable risk fraction of severe cases associated with recent hospital exposure reached a peak in the first wave of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% at the beginning of May, declined rapidly over the next few weeks, and increased again to `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  at the beginning of December Figure \ref{fig:plotexphosp} (bottom panel).
-For comparison, over the period of study the population attributable risk fraction of severe cases associated with the narrow definition of "probable health-care associated infection" was only `r round(100 * hcai.popattrfrac, 1)`%.
+#### Analysis of the association of exposure to hospital with severe COVID-19
+As Table \ref{tab:regressionseverecasesnocare} shows, the strongest risk factor (in terms of deviance explained) for severe COVID-19 in the general population was recent exposure to hospital: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.
+Exclusion of the `r round(100 * with(cc.all[testpositive.case & deathwithin28], table(covid_ucod)[1] / sum(table(covid_ucod))))`% of those dying within 28 days of a positive test who did not have  COVID-19 as underlying cause of death on their death certificates changed this rate ratio only slightly to `r testpos.coeffs["Recent hospital visit/stay", 3]`.
+As shown in Figure \ref{fig:plotexphosp} (middle panel)  the rate ratio associated with recent exposure to hospital  increased from `r round(coeffs.hosp.timewindow[date==as.Date("2020-04-01"), rateratio], 1)` in the time window with mid-point 1 April to a rate ratio of `r round(coeffs.hosp.timewindow[date==as.Date("2020-05-01"), rateratio], 0)` in the time window with mid-point 1 May.
+
+Figure \ref{fig:plotexphosp}  (top panel) shows the time course of the frequency of recent exposure to hospital in controls; for monitoring background exposure levels it is appropriate to examine the frequency in controls rather than in cases.
+The frequency of recent exposure  to hospital fell when restrictions were imposed in March and November 2020, but remained higher in those eligible for shielding than in those with moderate risk conditions or no risk conditions.
+The Population Attributable Risk Fraction for recent  exposure to hospital reached a peak in the first wave of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`%  of severe cases at the beginning of May, declined rapidly over the next few weeks, and increased again to `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December (Figure \ref{fig:plotexphosp}, bottom panel).
+In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r table.rapid.coeffs["rapid.recent", 3]`]; the  rate ratios associated with day case exposure  [rate ratio `r table.daycase.coeffs["daycase.recent", 3]`] or outpatient exposure [`r table.opd.coeffs["opd.recent", 3]`] were  much lower.
+
+To test if the association of severe COVID-19 with recent inpatient exposure was likely to simply reflect  confounding by time invariant comorbidity, the association was examined by time window as described in the Methods section.
+Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
+This rate ratio based on comparison of time windows is not directly comparable with the rate ratio based on the case-control analysis because restriction to those with discordant exposure between time windows selects those with  fewer days in hospital (i.e.
+a lower dosage of the exposure).
+<!--- tabulate underlying cause of death on death certs of fatal cases
+paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(hosp.recent, covid_ucod)))
+with(cc.severe[CASE==1], table(hosp.recent, inhosp))
+paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(occup, covid_ucod)))
+--->
+The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.
+From this and the  rate ratio of `r round(r.rapid, 1)` given above the Population Attributable Risk Fraction of severe COVID-19 associated with recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
+As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would  usually be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital acquired infection using the ECPDC classification.
+The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.
+Given the `r round(100 * p.hcai)`% frequency of exposure in cases the PARF can be calculated as `r round(100 * parf.hcai)`%.
 
 # Discussion
-## Summary and interpretation of findings
-Key findings of this study were: (i) that the rate ratios for  severe COVID-19 associated with shielding conditions remained high through the epidemic; (ii) that recent hospital exposure and  the number of adults in the household were important  risk factors for severe COVID-19 among those eligible for shielding and also in the general population; (iii) that among those not resident in care homes the about 40% of severe cases among those not resident in care homes were attributable to hospital attendance even in the second wave.
+## Summary of findings and comparison with other studies
+Key findings of this study were: (1) that the rate ratios for severe COVID-19 associated with shielding conditions remained high through the epidemic; (2) that recent hospital exposure and  the number of adults in the household were associated with severe COVID-19 among those eligible for shielding and also in the general population; (3) that among those not resident in care homes the proportion of severe cases that was attributable to hospital attendance reached a peak of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`% during the second wave.
 
+### Risk to those eligible for shielding
 In  comparison with those with no risk condition, the highest rate ratio for severe COVID-19 was the `r round(as.numeric(gsub(" .*", "", table.shielded.severecases.nocare["Solid organ transplant", 3])))`-fold rate ratio in solid organ transplant recipients.
-For other conditions deemed eligible for shielding, the rate ratios were between 4 and 7, compared to the rate ratios of about 3 associated with conditions designated as moderate risk, which include heart disease, diabetes, chronic kidney disease and disabling neurological conditions.
+For other conditions deemed eligible for shielding, the rate ratios were between `r round(min(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[4:8, 3]))))` and `r round(max(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[4:8, 3]))))`, compared to the rate ratio of about `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[2, 3])))` associated with conditions designated as moderate risk, which include heart disease, diabetes, chronic kidney disease and disabling neurological conditions.
+The numbers of pregnant women with heart disease were too small for the risk in this group to be estimated.
 Other studies of outcome of COVID-19 in solid organ transplant recipients and other individuals using immunosuppressants have been based only on patients admitted to hospital [@avery_inpatient_2020; @andersen_association_2021].
 Such studies cannot assess the risks to immunosuppressed individuals in the population.
+Other reports have evaluated the theoretical possibilities for shielding but without empirical data on its effectiveness.
+A previous report [@jani_comparison_2020] based on the population of Glasgow did not report age-adjusted rate ratios.
 
-If the advice provided and support offered in the shielding letters had been effective (beyond the messages already widely disseminated by this time and  general social distancing measures that were introduced  by then), we would have expected the rate ratio for severe disease associated with eligibility for shielding to fall within two weeks of the first letters being sent out.
-Our results show however that while presentations of severe cases fell rapidly in the general population from the beginning of April 2020, the fall in presentations among those eligible for shielding advice was delayed so that the rate ratio for severe disease associated with eligibility for shielding rose during April 2020, although the daily deaths fell during April 2020 both in those eligible and those ineligible for shielding.
+If the advice and support offered in the shielding letters had been effective (beyond the messages already widely disseminated by this time and  general social distancing measures that were introduced  by then), we would have expected the rate ratio for severe disease associated with eligibility for shielding to fall within two weeks of the first letters being sent out.
+Our results show however that while presentations of severe cases fell rapidly in the general population from the beginning of April 2020, the fall in presentations among those eligible for shielding advice was delayed so that the rate ratio for severe COVID-19 associated with eligibility for shielding rose during April 2020, although the daily deaths fell during April 2020 both in those eligible and those ineligible for shielding.
+The rate ratios associated with eligibility for shielding and with moderate risk conditions rose again from October 2020 to early December.
 
-Recent hospital exposure was strongly associated with disease in all risk groups.
-As recent hospital exposure was more common in those with moderate risk conditions and those eligible for shielding, especially after a lockdown had been imposed, this exposure accounted for a higher proportion of severe cases in these groups.
-Thus a possible explanation for this rise in the rate ratio for severe disease  associated with eligibility for shielding during April 2020 and the subsequent rise beginning in October 2020 is that those in this group were not able to avoid exposure to hospitals during periods when hospitals were a source of infection.
+The high frequency of recent hospital exposure in cases of severe COVID-19 who were eligible for shielding, together with the increase in the rate ratio associated with recent hospital  exposure during periods when population-wide social distancing measures were being imposed, suggests that exposure to transmission in hospital settings is at least part of the explanation for the failure to reduce the rate ratio associated with eligibility for shielding.
+Our finding that even in those advised to shield the risk of severe disease was increased in those who were sharing a household with other adults suggests that this was another constraint on the effectiveness of advice to shield, as no support was provided for other household members to co-isolate with the vulnerable individual.
 
+### Associations of transmission settings and household composition with severe  COVID-19
+The increased risk of severe COVID-19 in patient facing  health care workers is consistent with our previous report of a threefold risk for hospitalised COVID-19 earlier in the epidemic [@shah_risk_2020].
+We and others have previously reported  lower risk in teachers compared with others of the same age and sex [@mcallister_report_2020; @ludvigsson_open_2021].
+The inverse association of severe disease with the number of school-age children in the household extends and confirms the findings of an earlier study of health care workers and their families [@wood_living_2021].
+In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 after adjusting for covariates, but no dose-response relationship was reported [@forbes_association_2020].
+The inverse association of severe COVID-19 with past exposure to children is consistent with evidence that  other coronaviruses generate cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020].
+From a public health perspective the most relevant implication is that although the rate ratio per child has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows.
 The association of severe COVID-19 with number of adults in the household is consistent with classic studies of other viral infections showing that secondary cases in a household, where the infecting dose is likely to be high, are more severe than index cases [@aaby_malnutrition_1988].
-The inverse association of severe COVID-19 with number of school-age children in the household and the lower risk in teachers compared with others of the same age and sex are consistent with a protective effect of past exposure to children  Lower risk in teachers has been reported previously [@mcallister_report_2020; @ludvigsson_open_2021].
-These associations are consistent with evidence that past infection with other coronaviruses generates cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020].
-However no direct studies of the protective effect of these T-cell responses have been reported.
-Although the rate ratio per child has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows.
 
-## Comparison with previous literature
-Other reports have evaluated the theoretical possibilities for shielding but without empirical data on its effectiveness.
-A previous report [@jani_comparison_2020] was based on the population of Greater Glasgow and Clyde health board.
-As no age-standardized rate ratios were reported it is difficult to compare our results with theirs.
-In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 [@forbes_association_2020] after adjusting for covariates, but there was no association of non-fatal COVID-19 requiring critical care with living with children in this age group, and no dose-response relationship was reported.
+### Association with recent exposure to hospital
+A striking finding  from our analysis is the  large risk ratio and PARF for severe disease associated with  exposure to hospital.
+Across the epidemic we estimate that the PARF for hospital exposure  using our definition was  30-50% of severe cases.
+Although we had pre-specified our exposure variable to include day case and outpatients the association was driven by inpatient hospital exposure.
+The PARF associated with inpatient exposure was   `r round(100 * parf.rapid)`%,  Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF .
+As we have shown most of this difference is due to the fact that the international definition does not attempt to quantify cases diagnosed shortly after discharge that could have been acquired in hospital.
+The other difference  is that the  international definition classifies  those with onset 5-6 days after admission as indeterminate.
+As the average incubation period for COVID-19 is 5-6 days we consider it is reasonable  to include exposure in this time window as relevant.
+
+<!--I think this para snt really needed they ajhvent looked at rsk ratios or PARF and they have looked at all cases not jsut severe so wahts the pont f ncluding this ??
+see my attempt at an opening sentence that ties it in otherwise delte -->
+
+Comparisons with previous reports are limited by the fact that, apart from using the  ECDPC definition,  such reports are for all  test positives not severe cases as we have analysed  and report exposure frequency not rate ratios or PARF.
+So for example, for the week ending 6 December, when we estimate the population attributable risk fraction  for  recent exposure to hospital  to have peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), "popattr.frac"])`% of severe cases,   the ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland reported that 3% of all test-positive cases were indeterminate, probable or definite hospital onset [@arhai_scotland_hospital_2021].
+
+A paper presented to SAGE on 28 January 2021 from Public Health England (PHE) examined the frequency of recent hospital exposure among cases hospitalised with COVID-19, using Secondary Usage Service hospital episode data linked with COVID-19 tests and CO-CIN
+study data [@phe_contribution_2021].
+The authors broadened the ECDPC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure".
+The authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that  onward transmission of COVID-19 from hospital-acquired cases could account for an even higher proportion of hospitalised cases.
+As their study did not include a control group they were not able to estimate the rate ratio or the population attributable risk fraction.
 
 ## Strengths and limitations 
-Strengths of our study are the population coverage, the range of outcomes assessed, and the comprehensive linkage to medical records and demographic risk factors.
+Strengths of our study are the national coverage, the range of outcomes assessed, and the comprehensive linkage to medical records and demographic risk factors.
 A limitation is that we do not have primary care data other than encashed prescriptions.
@@
 
-The calculation of the fraction of cases attributable to an exposure provides an upper bound on the predicted effect of removing that exposure.
-The definition of "recent hospital exposure" for this study is intended to capture all those whose infection could have been acquired in hospital (sensitivity of 1), unlike the official definition of "probable health-care associated infection" which is intended to identify those whose infection was unlikely to have been acquired outside hospital (high specificity).
-This inclusive definition is appropriate as the calculation of the fraction of cases in the population that are attributable to the exposure is valid if the classification of exposure has sensitivity of 1, even if the specificity is less than 1 [@hsieh_effect_1988].
+The calculation of the PARF for an  exposure provides an upper bound on the predicted effect of removing that exposure.
+The definition of "recent hospital exposure" used in this study was intended to capture all those whose infection could have been acquired in hospital (sensitivity of 1), unlike the international definition of "probable health-care associated infection" which is intended to identify those whose infection was unlikely to have been acquired outside hospital (high specificity).
+This inclusive definition is appropriate, as the calculation of the PARF is valid if the classification of exposure has sensitivity of 1, even if the specificity is less than 1 [@hsieh_effect_1988].
 The association of severe COVID-19 with recent hospital admission is likely to be confounded by pre-existing risk conditions.
-However adjusting for risk conditions in a multivariable analysis reduced the rate ratio associated with recent hospital exposure only slightly (from 11.5 to 9.5) which suggests that this confounding effect is modest in comparison to the causal effect of exposure to hospital.
+However adjusting for risk conditions in a multivariable analysis reduced the rate ratio associated with recent hospital exposure only slightly (from  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 3])), 1)` to  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 5])), 1)`).
 The large increase in the rate ratio associated with recent hospital exposure during periods when a high proportion of inpatients were hospital cases also supports causality.
+The most compelling evidence of causality is the time window analysis.
+This eliminates confounding by time-invariant risk conditions.
+The rate ratio estimated from the time window analysis is likely to be conservative because of non-differential misclassification of exposure.
+Exposure after discharge from hospital (for instance if those recently discharged have more visitors at home) would not be expected to generate an association with inpatient stay in the most recent time window.
+
+As all hospital inpatients are tested for SARS-CoV-2 every few days and the official PHS definition of  deaths involving COVID-19 includes all deaths within 28 days of a positive test  and since  some of the conditions for which people are hospitalised may not be not much less than that for SARS-CoV-2 infection, this could bias the association with exposure to hospital.
+However the rate ratio associated with recent hospital exposure is barely changed when fatal cases without COVID-19 as underlying cause of death on death certificate are omitted.
+<!--- Chris R to run a comparison of 28-day fatality rates in those who test positive and those who test negative
+Nonetheless a very important issue more  generally for official death statistics for COVID is that we lack post mortem derived estimates of the sensitivity and specificity of the working definitions for COVID-19 death for true COIVD-19 death.
+Accordingly the potential for current working definitions to substantially overestimate the true death burden from COIVD-19 is considerable.
+--->
 
 ## Policy implications 
-The data presented here have important policy implications for shielding programmes,  for ongoing attempts to reduce COVID-19 transmission, and for vaccination policy.
+Our results have implications for policies on shielding the vulnerable, vaccination, and control of SARS-CoV-2 transmission in the population.
 
-Regarding shielding programmes, we could not find any evidence that the programme *per se* reduced COVID-19 rates  above and beyond the factors leading to the general reduction in rates in the background society.
+We found no evidence that the shielding programme *per se* reduced COVID-19 rates although it is possible that without shielding advice and support the outcome in this group would have been worse.
 It is relevant to examine why advice to shield, combined with offers of support for delivery of food and medicines, failed to protect some individuals who were identified as clinically extremely vulnerable.
 As a lockdown on the general population had been imposed on 23 March 2020, and individuals who considered themselves to be at high risk would have been likely to reduce their contact level before then, there were limited possibilities for risk in this group to be reduced further by advice to shield and offers of help in letters sent from 3 April onwards.
-We have identified two sources of exposure that are associated with severe disease and cannot easily be avoided by those advised to shield: hospital attendance as in-patient or out-patient, and sharing a household with other adults.
+We have identified two sources of exposure that are associated with severe disease and cannot easily be avoided by those advised to shield: in-patient hospital care, and sharing a household with other adults.
 We recommend that consideration should be given to special measures for to allow these  vulnerable individuals to reduce their exposure.
-This could include alternative ways of delivering care without hospital attendance, and support for other household members to co-isolate with the vulnerable individual.
+This could include special measures to protect these individuals from nosocomial infection, and support for other household members to co-isolate with the vulnerable individual.
 
-<!---Another important question is whether the correct risk groups were included in the shielded list.
---->The designation of groups as clinically extremely vulnerable was necessarily  undertaken before any systematic studies of associations with severe disease could have been conducted.
-Our study confirms that all these groups have markedly elevated risks of severe COVID-19, except for pregnant women with heart disease where the numbers at risk are too small for this risk to be reliably estimated, and therefore that assigning priority to these groups is appropriate for a vaccination programme that is aimed at reducing mortality and limiting the requirement for critical care.
-<!---The inability to estimate the risk in pregnant women with heart disease in this study should not alter the   vaccination  priority currently given to pregnant women with severe acquired or congenital heart disease.
---->
-However we note that others in the population who do not have one of these conditions may nonetheless have similarly elevated risk from multiple factors, for example people with poorly-controlled diabetes and multiple complications.
-To identify vulnerable individuals eligible for shielding or vaccination, the list of risk conditions could be supplanted by  formal risk scores.
-
-Regarding  the importance of this study for policies to reduce overall transmissions we highlight here the different picture that emerges when associations of COVID-19 with any recent hospital exposure are calculated, rather than associations with probable hospital-acquired infection as formally defined.
-Our analysis shows that focusing only on definite nosocomial inpatient-acquired infection may underestimate the impact that reducing infection risks across hospital sites could have.
-
-Our results emphasise the importance of household contacts  in transmission in those with risk conditions and raise the question of the priority that should be given to vaccinating  household contacts of vulnerable individuals.
-The Joint Committee on Vaccination and Immunization did not recommend this in a report published in December 2020, stating that "at this time there are no data on the size of the effect of COVID-19 vaccines on transmission." [@joint_commiteee_on_vaccination_and_immunisation_joint_2020].
-The Committee however recommended vaccination of  health and social care workers who are "at risk of transmitting infection to multiple vulnerable persons or other staff in a health care environment",  implying some expectation that vaccination would reduce transmission.
+Our findings support the policy of assigning highest priority for vaccination to those with risk conditions eligible for shielding, as these groups have markedly elevated risks of severe COVID-19.
+However we note that others in the population who do not have one of these conditions may have similarly elevated risk from multiple factors, for example people with poorly-controlled diabetes and multiple complications.
+To identify vulnerable individuals eligible for shielding or vaccination, lists of risk conditions could be supplemented by  formal risk scores.
 
 For most of those designated as clinically vulnerable, vaccination is expected to provide protection against severe COVID-19.
-For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke  an immune response sufficient to be effective and also uncertain whether  vaccines that use a live vector can be used safely [@chong_comprehensive_2017].
-On the basis of our results, we recommend that household members of solid organ transplant recipients should be offered vaccination, without waiting for  more definitive evidence of the effect of vaccines on transmission.
-Even with vaccination of  household contacts, this group will require help with shielding until until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.
+For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke  an immune response sufficient to be effective  [@chong_comprehensive_2017].
+The relation of severe disease in those eligible for shielding to number of adults in the household raises the question of whether priority should be given to vaccinating  household contacts of solid organ transplant recipients and other vulnerable individuals.
+The Joint Committee on Vaccination and Immunisation did not recommend this in a report published in December 2020, stating that "at this time there are no data on the size of the effect of COVID-19 vaccines on transmission." [@joint_committee_on_vaccination_and_immunisation_joint_2020].
+The Committee however recommended vaccination of  health and social care workers who are "at risk of transmitting infection to multiple vulnerable persons or other staff in a health care environment",  implying some expectation that vaccination would reduce transmission.
+On this basis, there is an equally strong case for offering vaccination to household contacts of solid organ transplant recipients without waiting for more definitive evidence of the effect of vaccines on transmission.
+Even with vaccination of  household contacts, this group will require shielding until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.
 
+Policies for control of transmission in the population and for reducing burden on health services and
+total deaths have been heavily focused on population-wide reduction of social contact. The report by
+PHE shows that infection prevention measures in hospital settings have likely had a major impact on
+reducing COVID-19 [@phe_contribution_2021].
+However our analysis provides compelling evidence for a substantial contribution of nosocomial transmission to the burden of severe COVID-19 even during the second wave.
+In turn this means implies there is considerable scope for reducing severe COVID-19
+cases further if nosocomial transmission can be further reduced. It is hoped that vaccination of health
+care workers will reduce transmissions in this setting but consideration should be given to widespread
+vaccination of all inpatients. We have not attempted to quantify here the onward transmission of
+COVID-19 from those who have acquired COVID-19 in hospital and are then discharged which could be
+considerable. At the least the data suggests more stringent testing before discharge and quarantine post
+discharge. More detailed understanding of how well recommended IP policies are actually being
+operationalised and the on-the grounds constraints of adherence to guidelines are also needed.
+
+# Declarations
 ## Information governance
@@
 We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
+We thank Shona Cairns and Laura Imrie of the Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland for comments on an earlier draft of this manuscript.
 
@@
 
-```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and sliding windown of presentation dates (b) Rate ratio asociated with recent hospital exposure.
-(c) fraction of severe cases not resident in care homes attributable to recent hospital exposure.
+```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and sliding window of presentation dates (b) Rate ratio asociated with recent hospital exposure in those not resident in care homes.
+(c) fraction of severe cases attributable to recent hospital exposure in those not resident in care homes.
 Data points for dates from 1 June to 30 September 2020 are omitted because the numbers of cases and controls are small."}
@@
 
+\doublespacing
+
+```{r timewindow, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.tw[1:3, ],
+             escape=FALSE,
+             booktabs=TRUE,
+             label="timewindow",
+             row.names=FALSE,
+             caption="Rate ratios for severe COVID-19 by time window of hospital in-patient exposure", 
+             col.names=c("Time window of exposure", "Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    column_spec(column=1, width="5cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="3cm") %>%
+    column_spec(column=5, width="2cm") %>%
+#    kableExtra::group_rows("Admissions (RAPID)", 1, 3, bold=TRUE) %>%
+#        kableExtra::group_rows("Day case attendance (SMR01)", 4, 6, bold=TRUE) %>%
+#        kableExtra::group_rows("Admission (SMR01 to be added when we have extract of admission dates)", 7, 9, bold=TRUE) %>%
+ #       kableExtra::group_rows("Out-patient attendance (SMR00)", 7, 9, bold=TRUE) %>%
+ #       kableExtra::group_rows("Psychiatric admisssions (SMR04)", 13, 15, bold=TRUE) %>%
+ #       kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
+        add_footnote(label=c("Rate ratios are from conditional logistic regression models",
+                             "The dataset is restricted to those exposed at some time between 5 and 24 days",
+                             "  before first testing positive.",
+                             "Recent interval is days 5 to 14 before first testing positive.",
+                             "Less recent interval is days 15 to 24 before first testing positive.",
+                             "Reference category is exposure in less recent interval only.",
+                             "The conventional case-crossover estimate of the odds ratio is the ratio of cases with",  "  recent exposure only to cases with less recent exposure only."),
+                     notation="none")
+
+```
+
 \newpage
@@
 \beginsupplement
-# Supplementary Text
+
+\onehalfspacing
+
+# Supplementary Material
 ## Sensitivity analysis for misclassification of early fatal cases as ineligible for shielding
@@
 # Supplementary Tables
+
+```{r hcai_hosp, echo=FALSE, warning=FALSE, message=FALSE}
+if(FALSE) {
+knitr::kable(table.hcai.hosp.recent,
+             escape=FALSE, 
+             booktabs=TRUE,
+             col.names=c("No recent hospital exposure", "Recent hospital exposure"), 
+			 label="hcai_hosp",
+             row.names=TRUE,
+             align=rep("r", 2), 
+             caption="Hospital onset status as defined by ECDPC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
+     add_footnote(label="Percentages are column percentages",
+                 notation="none") %>%
+    kable_styling(latex_options=c("hold_position"))
+}
+```
 
@@
 ```
+
+```{r hcai_rapid, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.hcai.rapid.recent,
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="hcai_rapid",
+             row.names=TRUE,
+             align=rep("r", 2), 
+             caption="Severe test-positive cases classified by hospital onset status (as defined by ECDPC) and recent inpatient exposure") %>%
+	 column_spec(column=1, width="4cm") %>%
+	 column_spec(column=2, width="2.5cm") %>%
+	 column_spec(column=3, width="2.5cm") %>%
+#	 column_spec(column=4, width="2.5cm") %>%
+#	 column_spec(column=5, width="2.5cm") %>%
+     add_footnote(label="Percentages are column percentages",
+                 notation="none") %>%
+    kable_styling(latex_options=c("hold_position"))
+
+```

### 2021-03-06 transmission.Rmd
updates for transmission preprint

--- 2021-02-18 shielding.Rmd
@@
     affiliation: HPS
+  - name: Stuart McGurnaghan
+    affiliation: IGMM
   - name: Matthew Armstrong
@@
   - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
-    email: helen.colhoun@igmm.ed.ac.uk (+44 7791 652665)
+    corresponding: yes
+    email: helen.colhoun@igmm.ed.ac.uk
     affiliation: HPS
@@
 bibliography: ./covid.bib
-csl: ./biomed-central.csl
+csl: ./plos.csl
 always_allow_html: true
@@
 
-r <- as.numeric(gsub(" .*", "", table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5]));
+r <- as.numeric(gsub(" .*", "", table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 3]));
 
@@
 
+r.all <- as.numeric(gsub(" .*", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 3]));
+
+pc.all <- gsub(".*\\(", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 2])
+pc.all <- 0.01 * as.numeric(gsub("\\\\.*", "", pc.all))
+
 ```
+\nolinenumbers
+
+Word counts: abstract 396 words, main text 5603 words.  
+2 figures, 4 tables, 25 references
+
+Keywords: COVID-19 / prevention & control; Transplant Recipient; Nosocomial Infection; Matched Case Control Studies.
 
@@
 
-* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent exposure to hospital -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.
-
+* _Objectives_ -  To investigate:(1) the risk of severe  COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.
 * _Design_ - Matched case-control study (REACT-SCOT). 
-
-* _Setting_ - population of Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)`.
-
+* _Setting_ - Population of Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)`. 
 * _Participants_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice, linked with all `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland.
-
-* _Main outcome measure_ - severe COVID-19, defined as cases that entered critical care or were  fatal.
-
-* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was
-`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
+* _Main outcome measure_ - Severe COVID-19, defined as cases that entered critical care or were  fatal.  
+* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])`
 in those with moderate risk conditions and 
@@
 in those eligible for shielding.
-The highest risk for severe disease was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` .
-Risk of severe COVID-19 increased with the number of adults in the household but decreased with the number of school-age children in the household.
-The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.
-In all risk groups severe COVID-19 was strongly associated with recent exposure to hospital (5 to 14 days before presentation date): rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.
-In a case-crossover analysis with less recent exposure (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
-The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% of severe cases at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December 2020.
-
+The highest rate was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])`.
+In both the shielded and  the general population, the risk of severe COVID-19 increased with the number of adults  but decreased with the number of school-age children in the household.
+Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.
+In a case-crossover analysis with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
+Among those eligible for shielding, the population attributable risk fraction (PARF) of severe cases for recent exposure to hospital was  `r round(100 * pc * (r -1) / r)`%.
+In the general population the PARF for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.
 * _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.
 For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts.
-<!---These results also provide compelling evidence for a substantial contribution of nosocomial infection to the burden of severe COVID-19---> Mitigating the impact of the epidemic requires control of transmission in hospital settings.
+Mitigating the impact of the epidemic requires control of nosocomial transmission.
 
 \newpage
+\linenumbers
+
+# Summary
+
+## What is already known on this topic?
+
+* Individuals designated as "clinically extremely vulnerable" -- about 3% of the population -- have been advised to shield themselves during the COVID-19 epidemic
+
+* The effectiveness of this shielding programme has not been evaluated.
+
+## What this study adds
+
+* Hospital-acquired infection has made a substantial contribution to the burden of COVID-19 in the population, accounting for more than half of severe cases in early December 2020.
+
+* For vulnerable individuals to be shielded, the risk of hospital-acquired infection must be reduced and support must be provided for other adults in the household to co-isolate with the vulnerable individual.
+
+* Solid organ transplant recipients are at the highest risk; vaccination of  household contacts should be considered for this group.
+
+<!--- 
+* don't blame HCW, there are problems 
+* intel failure
+--->
 
@@
 Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].
-Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters recommending strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
+Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters advising strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
 
-The initial objective of this study was to investigate  whether shielding advice and support had reduced the risk of COVID-19.
+The initial objective of this study was to quantify the incidence of severe COVID-19 in the shielding population and to  investigate  whether shielding advice and support had reduced the risk of COVID-19.
 On the hypothesis that shielding was effective, we predicted that the rate ratio associated with eligibility for  shielding  compared to those without risk conditions would fall after the receipt of shielding letters.
-Our second objective was to understand the role of  transmission-related factors: household composition, occupations including health care [@shah_risk_2020], and recent exposure to hospital on COVID-19 among those eligible for shielding.
+Our second objective was to understand the role of  transmission-related factors including household composition and recent exposure to hospital on COVID-19 among those eligible for shielding.
 As the relevance of these transmission-related factors among those eligible for shielding  became clear, the objectives were broadened to investigate the relation of severe COVID-19 to these transmission-related factors in  the general population.
@@
 # Methods
-We linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case control dataset is refreshed every few weeks and linked to health records.
-These linked records are used to derive a variables for a list of designated "moderate risk conditions".
+We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions".
 Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.
-The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio  from the case-control study as described below.
+The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio as described below.
 
@@
 The list had been generated  by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries,  for evidence of a list of agreed diseases and conditions  and supplemented by data submitted  to PHS following direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].
-
 The categories designated as eligible for shielding were as listed below [@health_protection_scotland_search_2020; @kanani_caring_2020]:
@@
 The presentation date was assigned as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those without a positive test result ascertained through hospital discharge records, and as fourteen days before the date of death for those without a positive test result ascertained through death certificates.
-Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r dateformat(maxdate.specimen)` for test results and `r dateformat(maxdate.death)` for deaths.
+Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r dateformat(lastdate.specimen)` for test results and `r dateformat(maxdate.death)` for deaths.
 
@@
 The codes used are as described previously  [@mckeigue_rapid_2020].
-Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
-Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].
-Quintile 5 is the least deprived.
+Three broad risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
+Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020]; quintile 5 is the least deprived.
 
-# Transmission-related risk factors 
+## Transmission-related risk factors 
 Since addresses in the CHI database have been mapped to  Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.
@@
 Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.
-Our intention was to capture all cases that *could have been* nosocomial rather than the more stringent definitions for "health-care associated COVID-19" specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
+Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
 The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.
 Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated.
-The ECDPC states that "cases with symptom onset within 14 days of discharge from a healthcare facility (e.g.
-re-admission) may be considered as community-associated, probable or definite health-care-associated-COVID-19, or to have an indeterminate association" but does not specify how cases are to be assigned to these categories.
+<!---The ECDPC states that "cases with symptom onset within 14 days of discharge from a healthcare facility (e.g.
+re-admission) may be considered as community-associated, probable or definite health-care-associated-COVID-19, or to have an indeterminate association" but does not specify how cases are to be assigned to these categories.--->
 
@@
 To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.
-The fraction of severe cases attributable to hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.
-where \ensuremath{p_c} is the frequency of exposure in cases in that time window and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].
+The population attributable risk fraction (PARF) of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.
+where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].
 Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.
@@
 This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases.
-It is a variant of the classic case-crossover design, which usually  compares in cases only the frequencies of exposure in recent and less recent time windows.
-Such an analysis can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000].
+It is a variant of the classic case-crossover design, which compares in cases only the frequencies of exposure in recent and less recent time windows.
+This can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000].
 The conditional odds ratio can thus be estimated as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only.
-The analysis that we report is a refinement of this that takes advantage of the availability of a matched control group.
+The analysis that we report is a refinement that takes advantage of the availability of a matched control group.
 We contrast not the numbers of cases with recent exposure only and nonrecent exposure only, but the conditional odds of case status between those with recent exposure only and those with less recent exposure only.
@@
 ## Shielding eligibility and risk of COVID-19
-Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
+Supplementary Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
 To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.
-The largest risk category was those with severe respiratory disease.
-Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories.
-Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters sent on 3 April 2020.
 Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
 
-Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as an area plot.
-The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak by `r dateformat(deaths.quartermax.date)`.
-After the peak, the daily number of severe cases fell more slowly in those eligible for shielding than in those without risk conditions.
+As shown in Figure \ref{fig:plotcasesbydate} the time course of daily severe cases in those eligible for shielding paralleled the time course in those without risk conditions.
+However in the first wave the daily number of severe cases fell more slowly in those eligible for shielding than in those without risk conditions.
+Of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases among those eligible for shielding, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` were resident in care homes.
+In  the case-control analyses, care home residents were excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.
 
-## Rate ratio for severe COVID-19  by risk condition group   
-Of those eligible for shielding who were included in the case-control dataset, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.
-In  subsequent case-control analyses care home residents and their matched controls were excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.
-
-Table \ref{tab:regressionseverecasesnocare} shows the rate ratios for severe COVID-19 associated with shielding eligibility and with presence of a moderate risk condition, with those without risk conditions as reference category.
+## Rate ratios for severe COVID-19  by risk group   
+Table \ref{tab:regressionseverecasesnocare} shows the rate ratios for severe COVID-19 associated with each category of shielding eligibility, with those without risk conditions as reference category, excluding care home residents.
 The univariate rate ratio for severe disease was
@@
 
-### Time course of the rate ratio associated with eligibility for shielding 
-Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 in the  shielding and moderate risk categories, with those without risk conditions as reference category.
-These rate ratios are computed over `r winsize`-day windows of presentation dates.
-The rate ratio associated with eligibility for shielding increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
+We examined the time course of the rate ratio associated with eligibility for shielding.
+As shown in Supplementary Table \ref{tab:letterdates} most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of  shielding letters sent on 3 April 2020.
+Supplementary Figure \ref{fig:plotrateratio}(a) shows  the risk ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
 The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding.
-
-A methodological concern was that unrecorded removal of fatalities from the original list of those identified as eligible for shielding could have led us to misclassify early fatal cases among those eligible for shielding as ineligible and thus to underestimate the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding.
-As described in Supplementary Methods and Figure \ref{fig:plotcasefatality}, by examining the time course of the fatal case proportion we were able to establish that such misclassification of eligibility was unlikely in cases presenting after the second week of March 2020.
+Thus there was no evidence that shielding advice reduced the rate ratio.
 
@@
 ### Associations in those eligible for shielding 
-Supplementary Table \ref{tab:shieldedonlyregressionseverecasesnocare} shows associations with risk factors among those eligible for shielding only, with severe respiratory disease (the largest category) as reference category.
+Table \ref{tab:shieldedonlyregressionseverecasesnocare} shows associations with risk factors among those eligible for shielding only, with severe respiratory disease (the largest category) as reference category.
 The rate ratio increased with the number of adults in the household but not with the number of children.
@@
 Of severe cases among those eligible for shielding, `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2]` had recent  exposure to hospital.
-Using Miettinen's formula as given above, the Population Attributable Risk Fraction  for recent  exposure to hospital can thus be calculated as `r round(100 * pc * (r -1) / r)`% of severe cases in those eligible for shielding.
-<!--Table \ref{tab:casegr} shows the relationship of hospitalisation, entry to critical care and fatality to recent hospital exposure among all diagnosed cases.-->
+Using Miettinen's formula as given above, the PARF for recent  exposure to hospital in those eligible for shielding can thus be calculated as `r round(100 * pc * (r -1) / r)`% of severe cases.
 
-### Associations in the general population 
-Table \ref{tab:regressionseverecasesnocare}  shows the association with risk factors in all those not resident in care homes.
+### Associations in the overall population 
+Table \ref{tab:regressionseverecasesnocare}  shows the association of severe COVID-19 with risk factors in the general  population,  including those eligible for shielding but excluding residents in care homes.
 The risk of severe COVID-19 increased with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.
@@
 The other demographic factor associated with increased risk was socioeconomic deprivation: the rate ratio in the least deprived quintile compared with the most deprived quintile was `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 5], table.shielded.severecases.nocare["5 - least deprived", 6])`.
-As shown in Figure \ref{fig:plotrateratio} (lower panel), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease per adult in household remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
+As shown in Supplementary Figure \ref{fig:plotrateratio}(b), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease per adult in household remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
 
@@
 
-#### Analysis of the association of exposure to hospital with severe COVID-19
-As Table \ref{tab:regressionseverecasesnocare} shows, the strongest risk factor (in terms of deviance explained) for severe COVID-19 in the general population was recent exposure to hospital: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.
+#### Association with recent exposure to hospital
+As Table \ref{tab:regressionseverecasesnocare} shows, the strongest risk factor (in terms of deviance explained) for severe COVID-19 in the overall population was recent exposure to hospital: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`.
 Exclusion of the `r round(100 * with(cc.all[testpositive.case & deathwithin28], table(covid_ucod)[1] / sum(table(covid_ucod))))`% of those dying within 28 days of a positive test who did not have  COVID-19 as underlying cause of death on their death certificates changed this rate ratio only slightly to `r testpos.coeffs["Recent hospital visit/stay", 3]`.
-As shown in Figure \ref{fig:plotexphosp} (middle panel)  the rate ratio associated with recent exposure to hospital  increased from `r round(coeffs.hosp.timewindow[date==as.Date("2020-04-01"), rateratio], 1)` in the time window with mid-point 1 April to a rate ratio of `r round(coeffs.hosp.timewindow[date==as.Date("2020-05-01"), rateratio], 0)` in the time window with mid-point 1 May.
 
-Figure \ref{fig:plotexphosp}  (top panel) shows the time course of the frequency of recent exposure to hospital in controls; for monitoring background exposure levels it is appropriate to examine the frequency in controls rather than in cases.
-The frequency of recent exposure  to hospital fell when restrictions were imposed in March and November 2020, but remained higher in those eligible for shielding than in those with moderate risk conditions or no risk conditions.
-The Population Attributable Risk Fraction for recent  exposure to hospital reached a peak in the first wave of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`%  of severe cases at the beginning of May, declined rapidly over the next few weeks, and increased again to `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December (Figure \ref{fig:plotexphosp}, bottom panel).
-In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r table.rapid.coeffs["rapid.recent", 3]`]; the  rate ratios associated with day case exposure  [rate ratio `r table.daycase.coeffs["daycase.recent", 3]`] or outpatient exposure [`r table.opd.coeffs["opd.recent", 3]`] were  much lower.
+Figure \ref{fig:plotexphosp}(a) shows the time course of the frequency of recent exposure to hospital using data from controls;    for monitoring background exposure levels it is appropriate to examine the frequency in controls rather than in cases.
+Recent exposure  to hospital fell  precipitously when restrictions on non-COVID-19 admission were imposed in March, but remained higher in those eligible for shielding than in those with moderate risk conditions or no risk conditions throughout.
 
+As shown in Figure \ref{fig:plotexphosp}(b)  the rate ratio associated with recent exposure to hospital  increased from `r round(coeffs.hosp.timewindow[date==as.Date("2020-04-01"), rateratio], 1)` in the time window with mid-point 1 April to a rate ratio of `r round(coeffs.hosp.timewindow[date==as.Date("2020-05-01"), rateratio], 0)` in the time window with mid-point 1 May.
+
+The  PARF of severe COVID-19 for recent  exposure to hospital in the general population can be calculated as `r round(100 * pc.all * (r.all - 1) / r.all)`% over the study period.
+This fraction reached a peak in the first wave of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`%  of severe cases at the beginning of May, declined rapidly over the next few weeks, and increased again to `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December (Figure \ref{fig:plotexphosp}(c)).
+
+In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r format.estcipv(table.rapid.coeffs["rapid.recent", 3], table.rapid.coeffs["rapid.recent", 4])`]; the  rate ratios associated with day case exposure [`r format.estcipv(table.daycase.coeffs["daycase.recent", 3], table.daycase.coeffs["daycase.recent", 4])`] or outpatient exposure  [`r format.estcipv(table.opd.coeffs["opd.recent", 3], table.opd.coeffs["opd.recent", 4])`]  were  much lower.
 To test if the association of severe COVID-19 with recent inpatient exposure was likely to simply reflect  confounding by time invariant comorbidity, the association was examined by time window as described in the Methods section.
 Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
-This rate ratio based on comparison of time windows is not directly comparable with the rate ratio based on the case-control analysis because restriction to those with discordant exposure between time windows selects those with  fewer days in hospital (i.e.
-a lower dosage of the exposure).
 <!--- tabulate underlying cause of death on death certs of fatal cases
@@
 The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.
-From this and the  rate ratio of `r round(r.rapid, 1)` given above the Population Attributable Risk Fraction of severe COVID-19 associated with recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
-As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would  usually be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital acquired infection using the ECPDC classification.
+From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
+As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital acquired infection using the ECPDC classification.
 The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.
-Given the `r round(100 * p.hcai)`% frequency of exposure in cases the PARF can be calculated as `r round(100 * parf.hcai)`%.
 
@@
 ## Summary of findings and comparison with other studies
-Key findings of this study were: (1) that the rate ratios for severe COVID-19 associated with shielding conditions remained high through the epidemic; (2) that recent hospital exposure and  the number of adults in the household were associated with severe COVID-19 among those eligible for shielding and also in the general population; (3) that among those not resident in care homes the proportion of severe cases that was attributable to hospital attendance reached a peak of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`% during the second wave.
+Key findings of this study were: (1) that the rate ratios for severe COVID-19 associated with shielding conditions remained high through the epidemic; (2) that recent hospital exposure and  the number of adults in the household were associated with severe COVID-19 among those eligible for shielding and also in the general population; (3) that among those not resident in care homes the proportion of severe cases attributable to hospital attendance reached a peak of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`% during the second wave.
 
@@
 
-### Associations of transmission settings and household composition with severe  COVID-19
+### Association with household composition
 The increased risk of severe COVID-19 in patient facing  health care workers is consistent with our previous report of a threefold risk for hospitalised COVID-19 earlier in the epidemic [@shah_risk_2020].
@@
 ### Association with recent exposure to hospital
-A striking finding  from our analysis is the  large risk ratio and PARF for severe disease associated with  exposure to hospital.
-Across the epidemic we estimate that the PARF for hospital exposure  using our definition was  30-50% of severe cases.
-Although we had pre-specified our exposure variable to include day case and outpatients the association was driven by inpatient hospital exposure.
-The PARF associated with inpatient exposure was   `r round(100 * parf.rapid)`%,  Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF .
-As we have shown most of this difference is due to the fact that the international definition does not attempt to quantify cases diagnosed shortly after discharge that could have been acquired in hospital.
+A striking finding from our analysis is the  PARF  of  `r round(100 * pc.all * (r.all - 1) / r.all)` % for severe disease associated with recent exposure to hospital.
+Although we had pre-specified this category to include day case and outpatient exposure, the association was driven by inpatient exposure.
+The PARF associated with inpatient exposure was  `r round(100 * parf.rapid)`%.
+Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF.
+As we have shown most of this difference is due to the fact that the international definition does not capture cases diagnosed shortly after discharge that could have been acquired in hospital.
 The other difference  is that the  international definition classifies  those with onset 5-6 days after admission as indeterminate.
-As the average incubation period for COVID-19 is 5-6 days we consider it is reasonable  to include exposure in this time window as relevant.
+As the average incubation period for COVID-19 is 5-6 days, exposure in this time window is relevant.
 
-<!--I think this para snt really needed they ajhvent looked at rsk ratios or PARF and they have looked at all cases not jsut severe so wahts the pont f ncluding this ??
-see my attempt at an opening sentence that ties it in otherwise delte -->
+Most previous reports have focused on the ECDPC definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].
+The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital acquired infection [@arhai_scotland_hospital_2021].
 
-Comparisons with previous reports are limited by the fact that, apart from using the  ECDPC definition,  such reports are for all  test positives not severe cases as we have analysed  and report exposure frequency not rate ratios or PARF.
-So for example, for the week ending 6 December, when we estimate the population attributable risk fraction  for  recent exposure to hospital  to have peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), "popattr.frac"])`% of severe cases,   the ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland reported that 3% of all test-positive cases were indeterminate, probable or definite hospital onset [@arhai_scotland_hospital_2021].
-
-A paper presented to SAGE on 28 January 2021 from Public Health England (PHE) examined the frequency of recent hospital exposure among cases hospitalised with COVID-19, using Secondary Usage Service hospital episode data linked with COVID-19 tests and CO-CIN
-study data [@phe_contribution_2021].
-The authors broadened the ECDPC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure".
-The authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that  onward transmission of COVID-19 from hospital-acquired cases could account for an even higher proportion of hospitalised cases.
-As their study did not include a control group they were not able to estimate the rate ratio or the population attributable risk fraction.
+A paper presented to SAGE on 28 January 2021 broadened the ECDPC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure" [@knight_contribution_2021].
+From linking hospital episode data to COVID-19 tests and CO-CIN, the authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that onward transmission of COVID-19 from hospital-acquired cases could account for another 5% of hospitalised cases.
 
 ## Strengths and limitations 
-Strengths of our study are the national coverage, the range of outcomes assessed, and the comprehensive linkage to medical records and demographic risk factors.
+Strengths of our study are the national coverage and the comprehensive linkage to medical records and demographic risk factors.
 A limitation is that we do not have primary care data other than encashed prescriptions.
@@
 The definition of "recent hospital exposure" used in this study was intended to capture all those whose infection could have been acquired in hospital (sensitivity of 1), unlike the international definition of "probable health-care associated infection" which is intended to identify those whose infection was unlikely to have been acquired outside hospital (high specificity).
-This inclusive definition is appropriate, as the calculation of the PARF is valid if the classification of exposure has sensitivity of 1, even if the specificity is less than 1 [@hsieh_effect_1988].
+This inclusive definition is appropriate, as the calculation of the PARF is valid if the classification of exposure has sensitivity of 1 even if the specificity is less than 1 [@hsieh_effect_1988].
 The association of severe COVID-19 with recent hospital admission is likely to be confounded by pre-existing risk conditions.
 However adjusting for risk conditions in a multivariable analysis reduced the rate ratio associated with recent hospital exposure only slightly (from  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 3])), 1)` to  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 5])), 1)`).
-The large increase in the rate ratio associated with recent hospital exposure during periods when a high proportion of inpatients were hospital cases also supports causality.
-The most compelling evidence of causality is the time window analysis.
-This eliminates confounding by time-invariant risk conditions.
-The rate ratio estimated from the time window analysis is likely to be conservative because of non-differential misclassification of exposure.
-Exposure after discharge from hospital (for instance if those recently discharged have more visitors at home) would not be expected to generate an association with inpatient stay in the most recent time window.
+The most compelling evidence of causality is the time window analysis, which shows that the association of disease is with elapsed time since exposure corresponding to the known incubation period of SARS-CoV-2.
+The rate ratio estimated from the time window analysis is not directly comparable with the rate ratio based on the case-control analysis because restriction to those with discordant exposure between time windows selects those with  fewer days in hospital (lower dosage of the exposure).
 
-As all hospital inpatients are tested for SARS-CoV-2 every few days and the official PHS definition of  deaths involving COVID-19 includes all deaths within 28 days of a positive test  and since  some of the conditions for which people are hospitalised may not be not much less than that for SARS-CoV-2 infection, this could bias the association with exposure to hospital.
-However the rate ratio associated with recent hospital exposure is barely changed when fatal cases without COVID-19 as underlying cause of death on death certificate are omitted.
-<!--- Chris R to run a comparison of 28-day fatality rates in those who test positive and those who test negative
-Nonetheless a very important issue more  generally for official death statistics for COVID is that we lack post mortem derived estimates of the sensitivity and specificity of the working definitions for COVID-19 death for true COIVD-19 death.
-Accordingly the potential for current working definitions to substantially overestimate the true death burden from COIVD-19 is considerable.
+As all hospital inpatients are tested for SARS-CoV-2 every few days and all deaths within 28 days of a positive test are officially classified as deaths involving COVID-19,  some misclassification of deaths from other causes as COVID-19 deaths is likely in those with recent hospital exposure.
+However the rate ratio associated with recent hospital exposure was barely changed when fatal cases without COVID-19 as underlying cause of death on death certificate were omitted.
+<!--- Chris R to run a comparison of 28-day fatality rates in those who test positive and those who test negative.
 --->
@@
 Our findings support the policy of assigning highest priority for vaccination to those with risk conditions eligible for shielding, as these groups have markedly elevated risks of severe COVID-19.
-However we note that others in the population who do not have one of these conditions may have similarly elevated risk from multiple factors, for example people with poorly-controlled diabetes and multiple complications.
-To identify vulnerable individuals eligible for shielding or vaccination, lists of risk conditions could be supplemented by  formal risk scores.
-
-For most of those designated as clinically vulnerable, vaccination is expected to provide protection against severe COVID-19.
 For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke an immune response sufficient to be effective  [@chong_comprehensive_2017].
@@
 
-Policies for control of transmission in the population and for reducing burden on health services and
-total deaths have been heavily focused on population-wide reduction of social contact. The report by
-PHE shows that infection prevention measures in hospital settings have likely had a major impact on
-reducing COVID-19 [@phe_contribution_2021].
+Policies for control of transmission in the population and for reducing burden on health services and total deaths have focused on population-wide reduction of social contact.
 However our analysis provides compelling evidence for a substantial contribution of nosocomial transmission to the burden of severe COVID-19 even during the second wave.
-In turn this means implies there is considerable scope for reducing severe COVID-19
-cases further if nosocomial transmission can be further reduced. It is hoped that vaccination of health
-care workers will reduce transmissions in this setting but consideration should be given to widespread
-vaccination of all inpatients. We have not attempted to quantify here the onward transmission of
-COVID-19 from those who have acquired COVID-19 in hospital and are then discharged which could be
-considerable. At the least the data suggests more stringent testing before discharge and quarantine post
-discharge. More detailed understanding of how well recommended IP policies are actually being
-operationalised and the on-the grounds constraints of adherence to guidelines are also needed.
+A report for NHS England by the Healthcare Safety Investigation Branch noted the challenges of controlling nosocomial transmission and recommended that a national strategy for infection prevention and control should be developed [@healthcare_safety_investigation_branch_investigation_2020].
+Vaccination of health care workers is likely to reduce nosocomial transmission, and vaccination of all inpatients should also be considered.
+We have not attempted to quantify the onward transmission of COVID-19 from discharged patients, but the PHE modelling study estimated that this could have accounted for 5% of cases requiring hospitalisation in England during the first wave [@knight_contribution_2021].
+This suggests  more stringent testing before discharge and quarantine post discharge should be considered.
+More detailed understanding of how recommended IP policies are being operationalised is also needed [@oliver_could_2021].
 
 # Declarations
-## Information governance
+## Public and Patient Involvement statement:
+This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives.
+
+## Ethics approval and information governance
 This study was approved under COVID-19 Rapid Data Protection Impact Assessment​ (DPIA) 20210023 that allows Public Health Scotland staff to link the datasets.
 Datasets were de-identified before analysis.
+<!---## Exclusive License
+The Corresponding Author has the right to grant on behalf of all authors and does grant on behalf of all authors, a worldwide licence to the Publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution, iii) create any other derivative work(s) based on the Contribution, iv) to exploit all subsidiary rights in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.--->
 
-## Data availability
-The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
-All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public
+## Transparency declaration
+PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
+This manuscript has been generated directly from the source data by a reproducible research pipeline.
 
@@
 
-## Transparency declaration
-PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
-This manuscript has been generated directly from the source data by a reproducible research pipeline.
-All source code used for derivation of variables, statistical analysis and generation of this manuscript is freely available on GitHub.
+## Data sharing statement
+The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.
 
-# Registration
-The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
-
-## Conflicts of interest
-The authors declare no conflicts of interest.
+<!---## Contributors
+PM, DM and HC provided substantial contributions to the conception or design and drafting of the manuscript.
+NL, DC, CG, JB, MA, JD and SC provided substantial contributions to the data acquisition.
+SH, CR, JM and DG provided substantial contributions to the data analysis or interpretation.
+All authors contributed to revising the manuscript critically for important intellectual content and approved the final manuscript.--->
 
@@
 We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
-We thank Shona Cairns and Laura Imrie of the Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland for comments on an earlier draft of this manuscript.
+We thank Shona Cairns and Laura Imrie of the Antimicrobial Resistance and Healthcare Associated Infection unit of NHS National Services Scotland for comments on an earlier draft of this manuscript.
+
+## Competing interest
+All authors have completed and submitted the ICMJE Form for Disclosure of Potential Conflicts of Interest.
+
+## Registration
+The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
 
@@
 Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
+
+\nolinenumbers
 
@@
 # Figures
-```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=0.8, fig.cap="Ribbon plot of severe cases by vulnerability category and date of presentation, excluding care home residents"}
+```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=0.8, fig.cap="Area plot of severe cases by vulnerability category and date of presentation, excluding care home residents"}
 
@@
 
-```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Rate ratios for severe COVID-19 in those not resident in care homes, by risk group and sliding window of presentation dates.
-Line thickness is proportional to precision of estimate.
-Arrows indicate dates that shielding advice letters were sent, with thickness proportional to number of letters sent on that date.
-(b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates.
-Data points for dates from 1 June to 30 September 2020 are omitted because the numbers of cases and controls are small."}
-gridExtra::grid.arrange(p.rateratio + theme(axis.title.x = element_blank()),
-                        p.household.rateratio, nrow=2)
-```
+```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and sliding window of presentation dates (b) Rate ratio asociated with recent hospital exposure in those not resident in care homes.
+(c) fraction of severe cases attributable to recent hospital exposure in those not resident in care homes."}
 
-```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and sliding window of presentation dates (b) Rate ratio asociated with recent hospital exposure in those not resident in care homes.
-(c) fraction of severe cases attributable to recent hospital exposure in those not resident in care homes.
-Data points for dates from 1 June to 30 September 2020 are omitted because the numbers of cases and controls are small."}
+theme_set(theme_gray(base_size = 10))
+
 gridExtra::grid.arrange(p.hosp + theme(axis.title.x = element_blank()),
@@
 
+\doublespacing
 \newpage
@@
 # Tables
-```{r shieldgroupsbyage, echo=FALSE, warning=FALSE, message=FALSE}
-knitr::kable(table.shieldgroups.byage, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="shieldgroupsbyage",
-             row.names=TRUE,
-             align=rep("r", 5),
-             caption="Shielding eligibility cohort by eligibility category and age") %>%
-    kableExtra::group_rows("Shielding eligibility category", 1, 6) %>%
-        add_header_above(c(" "=1, "Age group"=4)) %>%
-        add_footnote(label="Percentages are row percentages",
-                     notation="none") %>%
-       kable_styling(latex_options="HOLD_position")
-
-```
 
@@
 \doublespacing
-
-```{r timewindow, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.tw[1:3, ],
-             escape=FALSE,
-             booktabs=TRUE,
-             label="timewindow",
-             row.names=FALSE,
-             caption="Rate ratios for severe COVID-19 by time window of hospital in-patient exposure", 
-             col.names=c("Time window of exposure", "Controls", "Cases",
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
-    column_spec(column=1, width="5cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-    column_spec(column=4, width="3cm") %>%
-    column_spec(column=5, width="2cm") %>%
-#    kableExtra::group_rows("Admissions (RAPID)", 1, 3, bold=TRUE) %>%
-#        kableExtra::group_rows("Day case attendance (SMR01)", 4, 6, bold=TRUE) %>%
-#        kableExtra::group_rows("Admission (SMR01 to be added when we have extract of admission dates)", 7, 9, bold=TRUE) %>%
- #       kableExtra::group_rows("Out-patient attendance (SMR00)", 7, 9, bold=TRUE) %>%
- #       kableExtra::group_rows("Psychiatric admisssions (SMR04)", 13, 15, bold=TRUE) %>%
- #       kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
-        add_footnote(label=c("Rate ratios are from conditional logistic regression models",
-                             "The dataset is restricted to those exposed at some time between 5 and 24 days",
-                             "  before first testing positive.",
-                             "Recent interval is days 5 to 14 before first testing positive.",
-                             "Less recent interval is days 15 to 24 before first testing positive.",
-                             "Reference category is exposure in less recent interval only.",
-                             "The conventional case-crossover estimate of the odds ratio is the ratio of cases with",  "  recent exposure only to cases with less recent exposure only."),
-                     notation="none")
-
-```
-
 \newpage
-
-\beginsupplement
-
-\onehalfspacing
-
-# Supplementary Material
-## Sensitivity analysis for misclassification of early fatal cases as ineligible for shielding
-As those who were notified as having died were removed from the list of those eligible for shielding before letters were sent out, and no records of this filtering were available, we examined the fatal case proportion by presentation date.
-The rationale for this was that as filtering the list of those eligible for shielding to remove those who had died would misallocate fatal cases to the ineligible for shielding category, we could estimate the date up to which this misallocation occurred by examining the fatal case proportion by presentation date.
-
-Supplementary Figure \ref{fig:plotcasefatality} compares the time series of the fatal case proportion in those eligible and ineligible for shielding, by seven-day window of presentation date.
-The fatal case proportion in those eligible for shielding reached a peak in the first week of March and fell after this, suggesting that from mid-March onwards any misallocation of fatal cases to those ineligible for shielding was minimal.
-With a median 13 days from presentation to death, and further delay in death notifications, any filtering of notified deaths would have been unlikely to misallocate fatal cases whose presentation date was after the second week of March.
-
-# Supplementary Figures
-
-```{r plotcasefatality, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8, fig.cap="Proportion of diagnosed cases that were fatal in those ineligible and eligible for shielding, by sliding window of presentation date.
-For each category, line thickness is proportional to number of cases"}
-
-p.casefatality
-
-```
-
-\newpage
-
-# Supplementary Tables
-
-```{r hcai_hosp, echo=FALSE, warning=FALSE, message=FALSE}
-if(FALSE) {
-knitr::kable(table.hcai.hosp.recent,
-             escape=FALSE, 
-             booktabs=TRUE,
-             col.names=c("No recent hospital exposure", "Recent hospital exposure"), 
-			 label="hcai_hosp",
-             row.names=TRUE,
-             align=rep("r", 2), 
-             caption="Hospital onset status as defined by ECDPC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
-     add_footnote(label="Percentages are column percentages",
-                 notation="none") %>%
-    kable_styling(latex_options=c("hold_position"))
-}
-```
-
-```{r letterdates, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.date.letter,
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="letterdates",
-             row.names=TRUE,
-             align=rep("r", 6), 
-             caption="Dates of sending advice letters to those eligible for shielding") %>%
-    column_spec(column=1, width="3cm") %>%
-    column_spec(column=2, width="2.1cm") %>%
-    column_spec(column=3, width="1.9cm") %>%
-    column_spec(column=4, width="1.9cm") %>%
-    column_spec(column=5, width="1.9cm") %>%
-    column_spec(column=6, width="1.9cm") %>%
-    add_footnote(label="Percentages are row percentages",
-                 notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "hold_position"))
-
-```
 
@@
                  notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "hold_position"))
+    kable_styling(latex_options=c("scale_down", "HOLD_position"))
 
+```
+
+```{r timewindow, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.tw[1:3, ],
+             escape=FALSE,
+             booktabs=TRUE,
+             label="timewindow",
+             row.names=FALSE,
+             caption="Rate ratios for severe COVID-19 by time window of hospital in-patient exposure", 
+             col.names=c("Time window of exposure", "Controls", "Cases",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    column_spec(column=1, width="5cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="3cm") %>%
+    column_spec(column=5, width="2cm") %>%
+    add_footnote(label=c("Rate ratios are from conditional logistic regression models",
+                         "Dataset restricted to those exposed at some time between 5 and 24 days before ",
+                         "first testing positive.",
+                         "Recent interval = days 5 to 14, less recent interval days 15 to 24 before first testing positive.",
+                         "Reference category is exposure in less recent interval only."), 
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "HOLD_position"))
+
+```
+
+\beginsupplement
+\newpage
+\onehalfspacing
+
+# Supplementary Material
+## Supplementary Figures
+
+```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Rate ratios for severe COVID-19 in those not resident in care homes, by risk group and sliding window of presentation dates.
+(b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates.
+Line thickness is proportional to precision of estimate."}
+
+gridExtra::grid.arrange(p.rateratio + theme(axis.title.x = element_blank()),
+                        p.household.rateratio, nrow=2)
+
+```
+
+\newpage
+
+## Supplementary Tables
+
+```{r shieldgroupsbyage, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::kable(table.shieldgroups.byage, 
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="shieldgroupsbyage",
+             row.names=TRUE,
+             align=rep("r", 5),
+             caption="Shielding eligibility cohort by eligibility category and age") %>%
+    kableExtra::group_rows("Shielding eligibility category", 1, 6) %>%
+        add_header_above(c(" "=1, "Age group"=4)) %>%
+        add_footnote(label="Percentages are row percentages",
+                     notation="none") %>%
+       kable_styling(latex_options="HOLD_position")
+
+```
+
+```{r hcai_hosp, echo=FALSE, warning=FALSE, message=FALSE}
+if(FALSE) {
+knitr::kable(table.hcai.hosp.recent,
+             escape=FALSE, 
+             booktabs=TRUE,
+             col.names=c("No recent hospital exposure", "Recent hospital exposure"), 
+			 label="hcai_hosp",
+             row.names=TRUE,
+             align=rep("r", 2), 
+             caption="Hospital onset status as defined by ECDPC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
+     add_footnote(label="Percentages are column percentages",
+                 notation="none") %>%
+    kable_styling(latex_options=c("hold_position"))
+}
 ```
@@
                  notation="none") %>%
-    kable_styling(latex_options=c("hold_position"))
+    kable_styling(latex_options=c("HOLD_position"))
 
 ```
+
+```{r letterdates, echo=FALSE, warning=FALSE, message=FALSE}
+
+knitr::kable(table.date.letter,
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="letterdates",
+             row.names=TRUE,
+             align=rep("r", 6), 
+             caption="Dates of sending advice letters to those eligible for shielding") %>%
+    column_spec(column=1, width="3cm") %>%
+    column_spec(column=2, width="2.1cm") %>%
+    column_spec(column=3, width="1.9cm") %>%
+    column_spec(column=4, width="1.9cm") %>%
+    column_spec(column=5, width="1.9cm") %>%
+    column_spec(column=6, width="1.9cm") %>%
+    add_footnote(label="Percentages are row percentages",
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "hold_position"))
+
+```

### 2021-03-06 wghtalk.Rmd
updates for transmission preprint

---
title: "Controlling and mitigating the COVID-19 epidemic -- insights from record linkage studies in Scotland" 
author: Paul McKeigue 
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output: 
  beamer_presentation:
    slide_level: 1
    includes:
      in_header: ../beamer_header.tex
    latex_engine: lualatex
    keep_tex: yes
---

# REACT-SCOT study

* Case-control study of all diagnosed cases of COVID-19 in Scotland since March 2020

  - Cases include those without a positive test, ascertained through hospital discharge diagnosis or death certificates.

  - Ten controls per case, not previously diagnosed and matched for age, sex and general practice

  - linkage to household composition, medical records including hospital admissions, prescribing in primary care

# Severe COVID-19: main outcome variable

* Fatal outcome defined as death within 28 days of a positive test, or death at any time with COVID-19 coded as underlying cause

* Severe COVID-19 defined as entry to critical care within 21 days of presentation date, or fatal outcome

  - definition ensures that case ascertainment is not affected by triage of those assessed as unlikely to benefit from critical care

# Studies based on REACT-SCOT

* Co-morbidity and construction of a risk score ("COVID age")

* Pharmacoepidemiology: medications compromising COVID

* Shielding and hospital transmission

* Other studies: ethnicity, health care workers, teachers, living with children, variant B.1.1.7, vaccine efficacy

# Risk score as "COVID age": some vulnerable individuals not yet vaccinated

![](COVIDage_vaxstatus.png)

# Relation of severe COVID-19 to polypharmacy

![](plot.numdrugsgr-1.png)

# Laporte & Healy (April 2020): "Medications compromising COVID"

* proton pump inhibitors

* hypnotics, sedatives

* antipsychotics, drugs used in nausea and vertigo

* opioid analgesics

* gabapentinoids

* non-steroidal anti-inflammatory drugs.

* anticholinergic effects likely to increase risk of pneumonia:

  - H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used for parkinsonism, and antiepileptic drugs

# Univariate associations of severe COVID-19 with antipsychotic drugs

\begin{tabular}{lrrrl}
 &  Rate ratio (95\% CI) & $p$-value \\
Aripiprazole & 3.19 (1.61, 6.33) &  $9 \times 10^{-4}$ \\
Olanzapine & 2.68 (1.57, 4.56)  & $3 \times 10^{-4}$ \\
Levomepromazine  & 9.9  (4.7, 20.6) &  $1 \times 10^{-9}$ \\
Risperidone & 4.90  (3.26, 7.38) & $3 \times 10^{-14}$ \\
Haloperidol  &  9.9 (3.6, 26.9) & $7 \times 10^{-6}$ \\
Chlorpromazine hydrochloride  & 3.00  (1.28, 7.03) & 0.01 \\ 
Amisulpride  & 6.0 (1.7, 21.4) &  0.005 \\
\end{tabular}

# Conditions designated as eligible for shielding

1. Solid organ transplant recipient

2. Cancer of the blood or bone marrow, or people with cancer receiving treatments that affect  immune system

3. Severe respiratory conditions: cystic fibrosis, asthma, chronic obstructive airway disease, bronchiectasis, pulmonary hypertension

4. Rare diseases that increase risk of infections: severe combined immunodeficiency, sickle cell disease.

5. People on immunosuppression therapies sufficient to increase risk of infection

6. Pregnant with heart disease

7. Additional conditions: dialysis, splenectomy, others identified by clinicians

# The shielding programme

* List of eligible individuals generated by querying medical records including hospital discharge diagnoses, dispensed prescriptions and transplant registries

  - supplemented by data requested from clinicians

* First batch of shielding letters sent on 4 April 2020

  - advised strict isolation, offered support with home delivery of food and medication.

  - no support for other household members to co-isolate.

* Shielding programme was paused on 1 April 2020,

* Further letter issued on 25 November 2020 advising "extra protection" based on current protection level in that area.

# Shielding eligibility cohort by case status
\scriptsize

```{r shieldcasegroup, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shield.casegr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldcasegroup",
             row.names=TRUE,
             align=rep("r", 5)) %>%
    kableExtra::group_rows("Shielding eligibility category", 2, 7) %>%
    column_spec(column=1, width="2.5cm") %>%
    column_spec(column=2, width="1.9cm") %>%
    column_spec(column=3, width="1.7cm") %>%
    column_spec(column=4, width="1.5cm") %>%
    column_spec(column=5, width="1.4cm") %>%
    column_spec(column=6, width="1.5cm") %>%
        add_footnote(label="Percentages are row percentages",
                     notation="none")

```

# Severe cases by risk category and date of presentation: ribbon plot
```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4,fig.asp=0.75, fig.cap="Care home residents excluded"}

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

theme_set(theme_gray(base_size = 12))

p.byelig
```

# Rate ratios for severe COVID-19 associated with risk conditions
\scriptsize
\setbeamertemplate{caption}{\raggedright\insertcaption\par}

```{r  rateratiosshielded, echo=FALSE, warning=FALSE, message=FALSE}

options(knitr.table.format = "latex") # adds tab: prefix to labels

knitr::kable(table.shielded.severecases.nocare[1:8, c(1:2, 5:6)], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxfatal",
             row.names=TRUE,
             align=c("r", "r", "l", "l"), 
   col.names=c(clean.header(colnames(table.shielded.severecases.nocare)[1:2]),
                "Rate ratio (95\\% CI)", "p-value")) %>%
    column_spec(column=1, width="2cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.5cm") %>%
    column_spec(column=4, width="2.2cm") %>%
    column_spec(column=5, width="1.2cm")

```

# Risk factors for severe COVID-19 in those eligible for shielding 
\scriptsize
\setbeamertemplate{caption}{\raggedright\insertcaption\par}

```{r riskfactorsshieldedonly, echo=FALSE, warning=FALSE, message=FALSE}

library(knitr)
library(kableExtra)

options(knitr.table.format = "latex") # adds tab: prefix to labels

knitr::kable(table.shieldedonly.severecases.nocare[7:15, c(1:2, 5:6)], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxfatal",
             row.names=TRUE,
             align=c("r", "r", "l", "l"), 
   col.names=c(clean.header(colnames(table.shieldedonly.severecases.nocare)[1:2]),
                "Rate ratio (95\\% CI)", "p-value")) %>%
    kableExtra::group_rows("School-age children in household", 2, 4, bold=FALSE) %>%
    kableExtra::group_rows("Adults in household", 5, 7, bold=FALSE) %>%
    column_spec(column=1, width="2.3cm") %>%
    column_spec(column=2, width="1.7cm") %>%
    column_spec(column=3, width="1.5cm") %>%
    column_spec(column=4, width="2.2cm") %>%
    column_spec(column=5, width="1.2cm")

```

# Rate ratios for severe COVID-19 associated with shielding and moderate risk condition 
```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25,fig.asp=0.7, fig.cap="Care home residents excluded. Line thickness is proportional to precision of estimate. Arrows indicate dates that shielding advice letters were sent."}

p.rateratio + theme(plot.title = element_blank())  + 
labs(caption = element_blank(), tag = element_blank()) +   
theme(legend.position = c(0.55, 0.85)) + theme(legend.text=element_text(size=8)) + 
theme(legend.title = element_blank())

```

# Rate ratios for severe COVID-19 associated with recent hospital exposure 
```{r plothosprateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.5,fig.asp=0.65, fig.cap="Care home residents excluded"}

p.hosp.rateratio + theme(plot.title = element_blank()) + 
labs(caption = element_blank(), tag = element_blank())

```

# Prevalence of recent hospital exposure by risk category

```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25, fig.asp=0.7, fig.cap="Prevalence in controls not resident in care homes"}
p.hosp + theme(plot.title = element_blank()) + 
labs(caption = element_blank(), tag = element_blank()) + 
theme(legend.position = c(0.4, 0.8))

```

# Fraction of severe COVID-19 attributable to recent hospital exposure

```{r plotfrac, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25, fig.asp=0.7, fig.cap="Care home residents excluded"}
p.fracattr + theme(plot.title = element_blank())+ 
labs(caption = element_blank(), tag = element_blank())

```

# Conclusions (1)

* For vulnerable individuals to be shielded, the risk of hospital-acquired infection must be reduced and support must be provided for other adults in the household to co-isolate with the vulnerable individual.

* Solid organ transplant recipients are at the highest risk; vaccination of  household contacts should be considered for this group.

* Hospital-acquired infection has made a substantial contribution to the burden of COVID-19 in the population, accounting for more than half of severe cases in early December 2020.

# Conclusions (2)

* Expert advisory panels and policy-makers should give as much attention to nosocomial transmission as to population-wide social distancing and new variants of concern.

* Vaccination of health care workers is expected to help.

* It is difficult to control nosocomial transmission in buildings not designed for preventing aerosol transmission in a pandemic.

  - noted in a report for NHS England by the Healthcare Safety Investigation Branch dated 29 October 2020.

  - Designing or building separate fever hospitals should be considered.

* Analysis of real-world data beats theoretical modelling every time

  - possibilities for collaboration between public health agencies and researchers should be fully exploited.

### 2021-03-06 ct.Rmd
updates to ct draft

@@
 ## Ascertainment of COVID-19 cases evaluable for B.1.1.7 status
-From early in the epidemic Public Health Scotland has maintained a regularly-updated case-control study (REACT-SCOT) of all COVID-19 cases in Scotland, with each case matched to up to ten controls and extensive linkage to other health care records and routine datasets   as previously reported [@mckeigue_rapid_2020-1].
+From early in the epidemic Public Health Scotland has maintained a regularly-updated case-control study (REACT-SCOT) of all COVID-19 cases in Scotland, with each case matched to up to ten controls and extensive linkage to other health care records and routine datasets   as previously reported [@mckeigue_rapid_2020].
 The case definition in REACT-SCOT includes all those with a positive nucleic acid test, a hospital admission with a discharge code of COVID-19  or a death certificate with mention of COVID-19.
@@
 This was based on using the results of repeat tests on the same individual to reclassify borderline results as described in the Supplementary Text.
-
 Chris R: a positive for STGF defined as  negative on the S gene and Ct < 25 for at least one of the OR and N genes.  A weak positive is negative for S gene and Ct <= 30 for both OR and N gene. --->
@@
 The proportion of cases with fatal outcome was calculated over sliding seven-day time windows.
-We denote this fraction as the fatal case proportion.
-We use this term  to distinguish it from the case fatality ratio which is classically defined without linkage between diagnoses and deaths, as the ratio of deaths from the disease to diagnosed cases over a specified time window.
 Data were available up to  `r dateformat(lastdate.rapid)` for hospital admissions, `r dateformat(maxdate.sicsag)`  for entry to critical care and `r dateformat(maxdate.death)` for death registrations.
@@
 Care home residence was assigned using the field in the CHI database,   augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases.
+The definition of Scottish regions was based on the Nomenclature of Units for Territorial Statistics [Level 2](https://www.gov.scot/publications/review-nomenclature-units-territorial-statistics-nuts-boundaries/), with the North-Eastern and Highlands & Islands regions grouped as Northern.
 
@@
 
-Teachers were identified as individuals of working-age registered with the General Teaching Council of Scotland and currently working, or believed to be currently working, as teachers at a Scottish school in 2020.
+Teachers were identified as individuals of working-age registered with the General Teaching Council of Scotland and currently working as teachers at a Scottish school in 2020.
 Healthcare workers were identified using the General Practitioner Contractor Database (GPCD) and Scottish Workforce Information Standard System (SWISS) databases as described previously.
 [@shah_risk_2020; @mcallister_report_2020].
-Prior comorbidities were derived by linkage to prescribing and SMR01 as described previously  [@mckeigue_rapid_2020-1].
+Prior comorbidities were derived by linkage to prescribing and SMR01 as described previously  [@mckeigue_rapid_2020].
 
@@
 They are also reassuring with respect to concerns that the variant may be associated with more severe disease.
-Even if the higher transmissibility of the variant increases the incidence of infections, the effect on severe disease is likely to be small in comparison to the effect of other factors that are amenable to intervention, including overprescribing of psychotropic drugs [@mckeigue_relation_2021], transmission in hospitals[@mckeigue_relation_2021-1], and vaccination of vulnerable individuals.
+Even if the higher transmissibility of the variant increases the incidence of infections, the effect on severe disease is likely to be small in comparison to the effect of other factors that are amenable to intervention, including overprescribing of psychotropic drugs [@mckeigue_relation_2021], transmission in hospitals (McKeigue 2021, submitted) and vaccination of vulnerable individuals.
 
@@
 
-# Registration
+## Registration
 The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
@@
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories.
 
 ## Public Health Scotland COVID-19 Health Protection Study Group
-<!--this group needs to be updated-->
-
-Alice Whettlock$^1$, Allan McLeod$^1$, Andrew Gasiorowski$^1$, Andrew Merrick$^1$, Andy McAuley$^1$, April Went$^1$, Calum Purdie$^1$, Colin Fischbacher$^1$, Colin Ramsay$^1$, David Bailey$^1$, David Henderson$^1$, Diogo Marques$^1$, Eisin McDonald$^1$, Genna Drennan$^1$, Graeme Gowans$^1$, Graeme Reid$^1$, Heather Murdoch$^1$, Jade Carruthers$^1$, Janet Fleming$^1$, Jade Carruthers$^1$, Joseph Jasperse$^1$, Josie Murray$^1$, Karen Heatlie$^1$, Lindsay Mathie$^1$, Lorraine Donaldson$^1$, Martin Paton$^1$, Martin Reid$^1$, Melissa Llano$^1$, Michelle Murphy-Hall$^1$, Paul Smith$^1$, Ros Hall$^1$, Ross Cameron$^1$, Susan Brownlie$^1$, Adam Gaffney$^2$, Aynsley Milne$^2$, Christopher Sullivan$^2$, Edward McArdle$^2$, Elaine Glass$^2$, Johanna Young$^2$, William Malcolm$^2$, Jodie McCoubrey $^2$
-
-1 Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE.
-
-2 NHS National Services Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2
-6QE.
+Allan McLeod​, Alice Whettlock April Went, Beth Findlay Chris Sullivan, Ciaran Harvey, David Henderson, Edward McArdle, Eisin McDonald, Emily Griffiths, Genna Drennan, Johanna Young, Kirstin Leslie, Leonardo I Green, Louise Nicol, Melissa Llano, Nick Christofidis, Paul Bett, Ross Cameron, Theresa Ryan, Victoria Ponce Hardy.
+Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
 
@@
 
-```{r plotvariantbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1.25, fig.cap="\\label{fig:variantbydate}(a) Test-positives by date: area plot showing proportion of S gene dropouts in positive tests with reporting of cycle threshold.
-(b) Median cycle thresholds (reverse scale) in positive tests by date "}
+```{r plotvariantbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1.25, fig.cap="\\label{fig:variantbydate}(a) Test-positive cases with reporting of cycle threshold, by date: area plot showing proportion of S gene dropouts.
+(b) Median cycle thresholds (reverse scale) in positive tests, by date."}
 
@@
     kableExtra::group_rows("Region", 7, 10, bold=FALSE) %>%
-    kableExtra::group_rows("Number of adults in household", 13, 17, bold=FALSE) %>%
-    kableExtra::group_rows("Number of school-age children in household", 19, 21, bold=FALSE) %>%
-    kableExtra::group_rows("Occupation", 22, 25, bold=FALSE) %>%
+    kableExtra::group_rows("Risk category", 13, 15, bold=FALSE) %>%
+    kableExtra::group_rows("Number of adults in household", 16, 20, bold=FALSE) %>%
+    kableExtra::group_rows("Number of school-age children in household", 21, 23, bold=FALSE) %>%
+    kableExtra::group_rows("Occupation", 24, 28, bold=FALSE) %>%
 
@@
                  notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "hold_position"))
+    kable_styling(latex_options=c("scale_down", "HOLD_position"))
 
 ```
+
+\newpage
 
@@
                  notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "hold_position"))
+    kable_styling(latex_options=c("scale_down", "HOLD_position"))
 
@@
 
-# Imputation of S gene dropout
+# Supplementary material
+## Imputation of S gene dropout
 
@@
 
-# Supplementary Figures
+## Supplementary Figures
 
@@
 
-```{r checkcutoff, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1, fig.cap="\\label{fig:plotcheck}Test results with S gene signal absent that were originally classified as Undetermined (\\ensuremath{C_{ORF} >= 31 or \\ensuremath{C_{N} >= 31 or  \\ensuremath{C_{N} - C_{ORF} < 1.5}), colour coded by result at repeat test.
-New separation lines at \\ensuremath{C_{N} - C_{ORF} = 3} and ORF = 33, used in this analysis, are shown in green"}
+```{r checkcutoff, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.asp=1, fig.cap="\\label{fig:plotcheck}Repeat tests on same individual for test-positive results originally classified as Undetermined (\\ensuremath{C_{ORF} \\ge 31} or \\ensuremath{C_N \\ge 31} or  \\ensuremath{C_N - C_{ORF} < 1.5}), colour coded by result at repeat test.
+New separation lines at \\ensuremath{C_N - C_{ORF} = 3} and \\ensuremath{C_{ORF} = 33}, used in this analysis, are shown in green"}
 
@@
 gB <- ggplotGrob(p.region)
-gC <- ggplotGrob(p.hcw)
+gC <- ggplotGrob(p.hosp)
 grid::grid.newpage()
@@
 
-# Supplementary Tables
+## Supplementary Tables
 
@@
              caption="Cases by month") %>%
-    add_footnote(label=c("Restricted to cases with specimen date no earlier than 1 Sep 2020",
-                         "Outcome is not yet evaluable for cases with specimen date after 31 December 2020"),
+    add_footnote(label=c("Restricted to cases with specimen date no earlier than 1 Sep 2020"),
                  notation="none") %>%
     kableExtra::group_rows("All cases", 1, 8) %>%
-        kableExtra::group_rows("Test-positive cases", 9, 11, bold=FALSE) %>%
-        kableExtra::group_rows("Cases with Ct results", 12, 16, bold=FALSE) %>%
+        kableExtra::group_rows("Test-positive cases", 9, 12, bold=FALSE) %>%
+        kableExtra::group_rows("Cases with Ct results", 13, 16, bold=FALSE) %>%
         kableExtra::group_rows("Fatal cases and cohort fatality rates by S gene dropout status", 17, 18, bold=FALSE) %>%

### 2021-04-16 shielding.Rmd
updates to vax efficacy in shielded

@@
   \usepackage{pdfpages}
-  \linenumbers 
 output:
@@
 ```
+
+\nolinenumbers
 
@@
 Although we had pre-specified our exposure variable to include day case and outpatients the association was driven by inpatient hospital exposure.
-The PARF associated with inpatient exposure was   `r round(100 * parf.rapid)`%,  Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF .
+The PARF associated with inpatient exposure was   `r round(100 * parf.rapid)`%.
+Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF.
 As we have shown most of this difference is due to the fact that the international definition does not attempt to quantify cases diagnosed shortly after discharge that could have been acquired in hospital.

### 2021-04-16 vax.Rmd
updates to vax efficacy in shielded

---
title: "Short report: efficacy of vaccination in those eligible for shielding in Scotland"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Jen Bishop
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    corresponding: yes
    email: helen.colhoun@igmm.ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}

rollsum.datewin <- function(dt, k, datevar) {
    ## returns a table of rolling sums of width k centred on date
    ## dt is a dataset of totals (N) by date, which may have no rows for some dates in the
    ## range of interest
    ## add rows for missing dates by a left join of all.dates with dt
    all.dates <- data.table(date=seq(from=min(dt[[datevar]]),
                                     to=max(dt[[datevar]]),
                                     by=1))
    setkey(all.dates, date)
    # setkeyv(dt, datevar) can't set physical key here
    all.dates <- dt[all.dates]

    return(data.table(date=all.dates[[datevar]],
                      rsum=zoo::rollsum(all.dates[, N], k=k, fill=0, align="center")))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

format.estci <- function(x.ci) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    #x <- gsub("times", "\\\\times", x)
    return(x)
}

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

## vaccination analysis

datadir <- "data/2021-03-16/"
load(file=paste0(datadir, "cc.all.RData"))

shield.bydate <- cc.all[CASE==0 & SPECIMENDATE > as.Date("2020-12-01") & shield.any==TRUE,
                           .N,
                           by=c("SPECIMENDATE")]
setkey(shield.bydate, SPECIMENDATE)
shield.bydate <- rollsum.datewin(shield.bydate, 7, "SPECIMENDATE")
setnames(shield.bydate, "rsum", "all")

if(FALSE) {
    shield.bydate <- cc.all[CASE==0 & SPECIMENDATE > as.Date("2020-12-01") & shield.any==TRUE,
                            .N,
                            by=c("SPECIMENDATE", "vax14.dose")]
    setkey(shield.bydate, SPECIMENDATE)
    vax0.bydate <- rollsum.datewin(shield.bydate[vax14.dose==0], 7, "SPECIMENDATE")
    vax1.bydate <- rollsum.datewin(shield.bydate[vax14.dose==1], 7, "SPECIMENDATE")
    vax2.bydate <- rollsum.datewin(shield.bydate[vax14.dose==2], 7, "SPECIMENDATE")
    setkey(vax0.bydate, date)
    setkey(vax1.bydate, date)
    setkey(vax2.bydate, date)
    setnames(vax0.bydate, "rsum", "rsum0")
    setnames(vax1.bydate, "rsum", "rsum1")
    setnames(vax2.bydate, "rsum", "rsum2")
    shield.bydate <- vax2.bydate[vax1.bydate]
    shield.bydate <- shield.bydate[vax0.bydate]
}

shieldvax.bydate <- cc.all[CASE==0 & SPECIMENDATE > as.Date("2020-12-01") &
                           shield.any==TRUE & vax14==1,
                           .N,
                           by=c("SPECIMENDATE")]
setkey(shieldvax.bydate, SPECIMENDATE)
shieldvax.bydate <- rollsum.datewin(shieldvax.bydate, 7, "SPECIMENDATE")
setnames(shieldvax.bydate, "rsum", "vax")
setkey(shield.bydate, date)
setkey(shieldvax.bydate, date)       
shield.bydate <- shieldvax.bydate[shield.bydate]
shield.bydate[, prop.vax := vax / all]

p.shieldvax <-
    ggplot(data=shield.bydate, aes(x=date, y=prop.vax)) +
    geom_line() +
    scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), limits=c(0, 1),
                       expand=c(0, 0)) + 
    scale_x_date(breaks = seq.Date(from = as.Date("2020-12-15"),
                                   to =  as.Date("2021-03-02"), by = "week"),
                 labels=gsub("^0", "", 
                             format.Date(seq.Date(from = as.Date("2020-12-15"),
                                                  to =  as.Date("2021-03-02"),
                                                  by = "week"),
                                         "%d %b")
                             ),
                 limits=c(as.Date("2020-12-15"), as.Date("2021-03-02"))) +
        labs(x=paste0("Presentation date: mid-point of 7-day window"),
         y="Proportion vaccinated among controls eligible for shielding")

keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group",
               listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
cc.vax <- na.omit(cc.severe[SPECIMENDATE > as.Date("2020-12-01"), ..keep.vars])

keep.hosp.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "care.home", "shieldelig.group", 
                    "vax14", "stratum")
cc.hosp.vax <- na.omit(cc.all[(casegroup=="A" | casegroup=="B") &
                              SPECIMENDATE > as.Date("2020-12-01"), ..keep.hosp.vars])
rm(cc.all)

mh.strata <- cc.vax[care.home=="Independent" & shield.any==TRUE, 
                    .(SPECIMENDATE, stratum, CASE, vax14)]
mh.strata <- mh.strata[, c("a", "b", "c", "d") :=
                           list(CASE * as.integer(vax14),
                                (1 - CASE) * as.integer(vax14),
                                CASE * (1 - as.integer(vax14)),
                                (1 - CASE) * (1 - as.integer(vax14)))]
mh.strata <- mh.strata[, lapply(.SD, sum),
                       by=c("SPECIMENDATE", "stratum"), 
                       .SDcols=c("a", "b", "c", "d")]
mh.strata[, N := a + b + c + d]
mh.strata[, a.exp := (a + b) * (a + c) / N]
mh.strata[, c.exp := (a + c) * (c + d) / N]
mh.strata[, var.exp := (a + c) * (b + d) * (a + b) * (c + d) / (N^2 * (N - 1))]
mh.strata[is.nan(var.exp), var.exp := 0]
mh.strata[, week := paste(lubridate::year(SPECIMENDATE),
                          sprintf("%02d", lubridate::week(SPECIMENDATE)), sep=":")]

mh.dates <- mh.strata[, lapply(.SD, sum), by=week,
                      .SDcols=c("a", "a.exp", "c", "c.exp", "var.exp")]
setkey(mh.dates, week) 
mh.all <- data.table(week="All", mh.dates[, lapply(.SD, sum),
                                          .SDcols=c("a", "a.exp", "c", "c.exp", "var.exp")])
mh.dates <- rbind(mh.dates, mh.all)
colnames(mh.dates) <- c("Year:week", "Observed", "Expected", "Observed", "Expected", "Var(expected)")

table.dose <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14.dose + strata(stratum)))$coefficients

table.riskgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.riskgroups)
table.riskgroups <- as.data.table(table.riskgroups)
table.riskgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgroups <- table.riskgroups[, .(rateratio, pvalue)]
table.riskgroups <- data.table(effect=effect, # freqs.riskgroups,
                               table.riskgroups)
table.riskgroups[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.riskgroups[, effect := gsub("^listedgr3", "", effect)]
table.riskgroups[, effect := gsub("TRUE$", "", effect)]
table.riskgroups[, effect := gsub(":vax14$", "", effect)]
table.riskgroups[, effect := replace.names(effect)]

table.adjhosp <- summary(clogit(data=cc.vax, 
                               formula = CASE ~ care.home + listedgr3 / vax14 +
                                   prob.hcai + discharge15to56 + strata(stratum)))$coefficients

table.adjhosp <- as.data.table(table.adjhosp)
table.adjhosp[, rateratio := or.ci(coef, `se(coef)`)]
table.adjhosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.adjhosp <- table.adjhosp[, .(rateratio, pvalue)]

table.shieldgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + shield.group / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.shieldgroups)
table.shieldgroups <- as.data.table(table.shieldgroups)
table.shieldgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.shieldgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.shieldgroups <- table.shieldgroups[, .(rateratio, pvalue)]
table.shieldgroups <- data.table(effect=effect, # freqs.shieldgroups,
                               table.shieldgroups)
table.shieldgroups[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.shieldgroups[, effect := gsub("^shield.group", "", effect)]
table.shieldgroups[, effect := gsub("TRUE$", "", effect)]
table.shieldgroups[, effect := gsub(":vax14$", "", effect)]
table.shieldgroups[, effect := replace.names(effect)]
table.shieldgroups <- as.data.frame(table.shieldgroups)

table.hosp <- summary(clogit(data=cc.hosp.vax, formula = CASE ~ care.home + shieldelig.group / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.hosp)
table.hosp <- as.data.table(table.hosp)
table.hosp[, rateratio := or.ci(coef, `se(coef)`)]
table.hosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.hosp <- table.hosp[, .(rateratio, pvalue)]
table.hosp <- data.table(effect=effect, # freqs.hosp,
                               table.hosp)
table.hosp[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.hosp[, effect := gsub("^shieldelig.group", "", effect)]
table.hosp[, effect := gsub("TRUE$", "", effect)]
table.hosp[, effect := gsub(":vax14$", "", effect)]
table.hosp[, effect := replace.names(effect)]
table.hosp <- as.data.frame(table.hosp)

freqs.riskgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "listedgr3"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "listedgr3"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]

oddsratio.table <- function(freqs.table) {
    freqs.table <- as.matrix(freqs.table)
    or <- numeric(nrow(freqs.table))
    for(i in 1:nrow(freqs.table)) {
        a <- as.numeric(gsub(" .+", "", freqs.table[i, ]))
        or[i] <- a[1] * a[4] / (a[2] * a[3])
    }
    return(round(or, 2))
}

freqs.shieldgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]
rownames(freqs.shieldgroups) <- table.shieldgroups$effect[9:16]
freqs.shieldgroups$"Crude odds ratio" <- oddsratio.table(freqs.shieldgroups)

freqs.hosp <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]
rownames(freqs.hosp) <- table.hosp$effect[8:14]
freqs.hosp$"Crude odds ratio" <- oddsratio.table(freqs.hosp)

if(FALSE) {
cc.vax[`dm.typeOther/unknown type` == 1, `dm.typeType 2 diabetes` := 1]
cc.vax[, `dm.typeOther/unknown type` := NULL]

controls.vax <- cc.kept[CASE==0 & care.home=="Independent", .(AGE, COVID.age, vax_dose_1)]
controls.vax[, vaxnow := !is.na(vax_dose_1)]
controls.vax[, vaxnow := car::recode(vaxnow,
                                   recodes="FALSE='Not vaccinated';
                                            TRUE='At least one dose'",
                                   as.factor=TRUE,
                                   levels=c("Not vaccinated", "At least one dose"))]

theme_set(theme_gray(base_size = 16))
p.covidage <- ggplot(data=controls.vax, aes(x=AGE, y=COVID.age, color=vaxnow)) +
    geom_point(size=0.5, position="jitter") +
    scale_x_continuous(breaks=seq(20, 80, by=10), limits=c(15, 90)) + 
    scale_y_continuous(breaks=seq(20, 80, by=10), limits=c(15, 90)) +
    theme(legend.title = element_blank()) +
    theme(legend.position=c(0.15, 0.85)) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(title=paste("COVID age versus calendar age by vaccination status on",
                        format.Date(max(vacc$vax_dose_1), '%d %b %Y')),
         caption="Controls matched to severe cases, not resident in care homes. Limits of COVID age are 16 and 90 years",
         x="Calendar age (years)",
         y="COVID age (years)")
p.covidage

png("COVIDage_vaxstatus.png")
p.covidage
dev.off()
rm(p.covidage)

vax.firstdoses <- vacc[, .N, by=c("weekdose1", "vax_type_1")]
ggplot(data=vax.firstdoses, aes(x=weekdose1, y=N, color=vax_type_1)) +
    geom_line()

}

```

# Background
In Scotland to date COVD-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.  Both have been shown to be efficacious in clinical trials.  Real-world observational analysis in a primary care-based study in Scotland up to 15 Feb 15 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].  However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as restrictions on the general population are lifted.  This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.

# Methods 
This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to explore risk factors for severe COVID-19 in the population of Scotland.  The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission of death with COVID-19 codes. The dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.  This analysis is based on all cases presenting between 1 December 1st 2020 to 16 March 2021.

We used this matched case-control design to take advantage of extensive data linkages already established. As the design uses incidence density sampling, it controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise confound the association between vaccination and disease.

As previously, to minimise ascertainment bias we focused the analysis on  severe COVID-19,  defined as entry to critical care within 28 days of presentation or  fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).   As described previously, cases and controls were grouped into one of three categories: no risk condition, risk condition designated previously by Public Health Scotland as moderate risk, eligible for shielding.  For further analyses, the shielding category was subdivided as described previously into six categories: solid organ transplant, specific cancers, severe respiratory conditions, rare conditions, on immunosuppressants, additional conditions.

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. Ths assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few too allow further subdivision of time since vaccination.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories.  Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence. For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.  The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$. The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.

As recent hospital admissions or being in hospital before presentation date might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.  Because As another sensitivity analysis and to allow inference on vaccine efficacy in less frequent risk groups, the analysis was repeated with the case definition extended to include all cases that were hospitalized or fatal.  For this analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".

# Results
Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date.  This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population.  By 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.

```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:plotvax}Proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date"}

theme_set(theme_gray(base_size = 8))
p.shieldvax

```

Supplementary Table \ref{tab:freqsshieldgroups} shows the distribution of risk groups among  all severe cases and controls and then vaccinated cases and controls. Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.   These   tables show that the numbers of cases in some shielding subgroups, especially solid organ transplant recipients, are too small for rate ratios to be estimated reliability. Supplementary Table \ref{tab:mhdates} shows the observed and expected numbers of severe cases among vaccinated cases and unvaccinated individuals who eligible for shielding and not resident in care homes, by week of presentation.  This shows that after week 6 of 2021 the number of severe cases was greatere in vaccinated than in unvaccinated individuals.  This is to be expected as by this time more than 70% of those eligible for shielding had been vaccinated; it does not indicate that vaccination was ineffective.

The rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.riskgroups$rateratio[4])` in those with no risk condition, `r format.estci(table.riskgroups$rateratio[5])` in those with moderate risk conditions and `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding.  With adjustment for having been in hospital for more than 7 days before presentation and for any hospital discharge from 15 to 56 days before presentation, the corresponding rate ratios were `r format.estci(table.adjhosp$rateratio[6])`, `r format.estci(table.adjhosp$rateratio[7])` and `r format.estci(table.adjhosp$rateratio[8])`.  
Table \ref{tab:shieldgroups} gives a breakdown by shielding group.  For the categories labelled as specific cancers, severe respiratory disease and additional conditions the numbers are large enough for the effect of vaccination on severe disease to be estimated reliably.  The point estimates for the rate ratio for all risk groups were  lower than for those without risk conditions i.e. vaccine efficacy was at least as high,  with the exception of  solid organ transplant recipients,  those  with rare diseases and  those on immunosuppressants.  For these groups the numbers of cases and controls were small (Supplementary Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.

```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldgroups[9:16, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldgroups",
             row.names=FALSE,
             align=c("l", "r", "l"),
             caption="Rate ratios for severe COVID-19 associated with vaccination, within each risk group",
             col.names=c("Effect", 
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
                         "Vaccine effects nested within each level of risk group",
                         "Care home residence included as covariate", 
                         "Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(column=1, width="6cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

Table \ref{tab:shieldelig} shows the analysis extended to include all hospitalized cases, for which the numbers of cases are larger.   This shows significant efficacy in all groups other than solid organ transplant recipients but in this group the numbers of cases and controls were small even with the broader case definition (Supplementary Table \ref{tab:freqshosp}).

```{r hosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.hosp[8:14, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldelig",
             row.names=FALSE,
             align=c("l", "r", "l"),
             caption="Rate ratios for hospitalized or fatal COVID-19 associated with vaccination within each risk group",
             col.names=c("Effect", "Rate ratio (95\\% CI)", 
                         "\\ensuremath{p}-value")) %>%
    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
                         "Care home residence included as covariate", 
                         "Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    kable_styling(latex_options=c("HOLD_position"))

```

# Conclusions
Efficacy of a single vaccination dose in protecting against severe COVID-19 is as high or higher in those eligible for shielding as in those without risk conditions.  For solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases is still too small for the efficacy of vaccination to be estimated.

The estimates of the rate ratio associated with vaccination or the vaccine efficacy here cannot be directly compared with the previous report for the total population of Scotland as they are summary estimates across all time periods from 14 days post-vaccination, rather than over 7-day intervals.  Nonetheless the estimate is close to that reported at 14 days in the  previous report.

Since we have previously shown that  the shielding group overall have approximately 6 fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1] they would need a vaccine efficacy of  82% (i.e. a rate ratio of 0.18)  with a single dose to achieve a lowering of risk to the level of unvaccinated persons  without risk conditions.   Thus whilst those in shielding  can be reassured that the vaccine is showing efficacy they should also be advised that they remain at increased risk  after a single dose (about twofold compared to those of the same age and sex without risk conditions given the vaccine-associated rate ratio of about 0.3). Thus they should remain vigilant  until they have received their second dose and background transmission rates are low. Also a  larger dataset than the  Scottish population is needed to estimate efficacy in solid organ transplant recipients where we have not been able to demonstrate efficacy.

The vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].  Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]

# References
<div id="refs"></div>

\beginsupplement
\onehalfspacing

# Supplementary Material

```{r freqsshieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(freqs.shieldgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsshieldgroups",
             row.names=TRUE,
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

```

```{r freqshosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

```

Vaccine efficacy cannot be estimated from the crude odds ratios in Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp} because these totals do not take account of the matched case-control design.   The matched design makes it possible to control for age, sex, general practice and crucially for calendar date where the proportion vaccinated and the rates of severe disease were changing day by day during the study period.  The vaccine efficacy is estimated over matched sets, conditional on the marginal totals (cases vs controls, vaccinated versus unvaccinated) in each set. This section is intended to provide a more intuitive explanation of this.  Table \ref{tab:mhdates} shows how information for a statistical test (the Mantel-Haenszel test) of  vaccine efficacy accumulated over matched sets (strata) of cases and controls during the study period.  To keep the working simple, this table is restricted to those not resident in care homes so there is no covariate adjustment.

Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:

 | | Case | Control | All |
 | --- | --- | --- | --- |
 Vaccinated | $a$ | $b$ | $V_1$ |
 Unvaccinated | $c$ | $d$ | $V_0$ |
 All | $C_1$ | $C_0$ | $N$ |

The marginal totals of cases and controls are $C_1, C_0$ and of vaccinated and unvaccinated are $V_1, V_2$.  Conditional on these marginal totals the expected number of vaccinated cases under the null hypothesis of no association is $C_1 V_1 / N$ and the variance of the expected is $C_1 C_0 V_1 V_0 / \left[ N^2 \left(N - 1 \right) \right]$.  The observed numbers, the expected and the variance are summed over matched sets to calculate a test statistic as (observed - expected) divided by the square root of the variance.

```{r mh.dates, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(mh.dates[1:17, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="mhdates",
             row.names=FALSE,
             align=c("l", "r", "r", "r", "r"),
             digits=1,
             caption="Vaccinated and unvaccinated severe cases among those eligible for shielding and not resident in care homes: observed and expected numbers by week") %>%
    add_header_above(c(" "=1, "Vaccinated cases"=2, "Unvaccinated cases"=2)) %>% 
    add_footnote(label=c("Expected numbers (under the null) are calculated in each stratum",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    kable_styling(latex_options=c("hold_position"))

```

### 2021-08-08 vaxshieldupdate.Rmd
separate category for blood cancers

---
title: "Efficacy of COVID-19 vaccination in individuals designated as clinically extremely vulnerable in Scotland: update examining effect of second dose"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
#    corresponding: yes
#    email: helen.colhoun@igmm.ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: IGMM
  address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE

header-includes:
  \usepackage{newunicodechar}
  \usepackage[T1]{fontenc}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./f1000.csl # ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)
library(kableExtra)

source("helperfunctions.R")

rollmean.dtN <- function(dt, k) {
    ## FIXME: add a by= argument
    N.dates <- dt[, .N, by=SPECIMENDATE]
    setkey(N.dates, SPECIMENDATE)
    all.dates <- data.table(date=seq(from=min(N.dates$SPECIMENDATE),
                                     to=max(N.dates$SPECIMENDATE),
                                     by=1))
    setkey(all.dates, date)

    ## add rows for missing dates by a left join
    all.dates <- N.dates[all.dates]
    all.dates[is.na(N), N := 0]
    return(data.table(date=dt$SPECIMENDATE,
                      avdaily=zoo::rollmean(dt$N, k=k, fill=NA)))
}

rollsum.datewin <- function(dt, k, datevar) {
    ## returns a table of rolling sums of width k centred on date
    ## dt is a dataset of totals (N) by date, which may have no rows for some dates in the
    ## range of interest
    ## add rows for missing dates by a left join of all.dates with dt
    all.dates <- data.table(date=seq(from=min(dt[[datevar]]),
                                     to=max(dt[[datevar]]),
                                     by=1))
    setkey(all.dates, date)
    # setkeyv(dt, datevar) can't set physical key here
    all.dates <- dt[all.dates]

    return(data.table(date=all.dates[[datevar]],
                      rsum=zoo::rollsum(all.dates[, N], k=k, fill=0, align="center")))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

format.estci <- function(x.ci) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    #x <- gsub("times", "\\\\times", x)
    return(x)
}

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
lastdate.cases <- as.Date("2021-07-12")

## vaccination analysis
## frequency tables by risk group and dose
## by shield group and dose
## by risk group and dose-product

#cc.all[, agegr20plus5 := car::recode(age_years,
#                                    "0:19='0 to 19'; 20:29='20 to 29'; 30:39='30 to 39'; #40:49='40 to 49'; 50:hi='50 years and over'",
                                        #                                    as.factor=TRUE)]

cc.all[, shield.group := as.character(shield.group)]
cc.all[bloodcancer==1,
       shield.group := car::recode(shield.group, "'Specific cancers'='Blood cancers'")] 
cc.all[bloodcancer==0,
       shield.group := car::recode(shield.group, "'Specific cancers'='Other specific cancers'")]
cc.all[, shield.group := factor(shield.group,
                                levels=c("No risk condition",
                                         "Moderate risk condition",
                                         "Solid organ transplant",
                                         "Blood cancers",
                                         "Other specific cancers",
                                         "Severe respiratory",
                                         "Rare diseases",
                                         "On immunosuppressants",
                                         "Additional conditions"))]
severe.byagevax <- with(cc.all[CASE==1 & casegroup=="A" &
                             specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                      table(agegr20plus5, vax14.factor))
severe.byagevax.rowsums <- rowSums(severe.byagevax)
severe.byagevax <- cbind(severe.byagevax, severe.byagevax.rowsums)
severe.byagevax <- paste.colpercent(severe.byagevax)
colnames(severe.byagevax)[4] <- "All"

severe.byagerisk <- with(cc.all[CASE==1 & casegroup=="A" &
                             specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                      table(agegr20plus5, listedgr3))
severe.byagerisk.colsums <- colSums(severe.byagerisk)
severe.byagerisk <- rbind(severe.byagerisk, severe.byagerisk.colsums)
severe.byagerisk <- paste.rowpercent(severe.byagerisk)
rownames(severe.byagerisk)[6] <- "All"

casesnorisk <- cc.all[casegroup=="A" & listedgr3=="No risk condition" & 
                              specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases &
                              age_years < 50] # 91 cases

setkey(casesnorisk, anon_id, specimen_date)
setkey(cc.diagnoses, anon_id, specimen_date)
casesnorisk.diagnoses <- cc.diagnoses[casesnorisk]

casesnorisk.chnums <- unique(casesnorisk.diagnoses,
                                by=c("anon_id", "specimen_date", "chnum"))

chnum.table <- with(casesnorisk.chnums, table(chnum, CASE))
colnames(chnum.table) <- c("Controls", "Cases")
chnum.table <- data.table(chnum=as.integer(rownames(chnum.table)),
                             as.data.frame.matrix(chnum.table))
chnum.table[, descr := icdchapters$name[as.integer(chnum)]]
chnum.table[, total := Controls + Cases]
chnum.table[, Controls := paste0(Controls, " (",
                                 round(100 * Controls / casesnorisk[CASE==0, .N]),
                                 "\\%)")]
chnum.table[, Cases := paste0(Cases, " (",
                                 round(100 * Cases / casesnorisk[CASE==1, .N]),
                                 "\\%)")]
chnum.table[, chnum := as.roman(chnum)]
chnum.table <- chnum.table[total >= 20, .(chnum, descr, Controls, Cases)]

## 7 with alcohol diagnoses F00

casesnorisk.scrips <- cc.scripsbnf[casesnorisk]
casesnorisk.scrips[, bnfchnum := substr(bnf_item_code, 1, 2)]
casesnorisk.bnfchnums <- unique(casesnorisk.scrips, 
                                by=c("anon_id", "specimen_date", "bnfchnum"))
bnfchnum.table <- with(casesnorisk.bnfchnums, table(bnfchnum, CASE))
colnames(bnfchnum.table) <- c("Controls", "Cases")
bnfchnum.table <- data.table(bnfchnum=as.integer(rownames(bnfchnum.table)),
                             as.data.frame.matrix(bnfchnum.table))
bnfchnum.table[, total := Controls + Cases]
bnfchnum.table[, Controls := paste0(Controls, " (",
                                 round(100 * Controls / casesnorisk[CASE==0, .N]),
                                 "\\%)")]
bnfchnum.table[, Cases := paste0(Cases, " (",
                                 round(100 * Cases / casesnorisk[CASE==1, .N]),
                                 "\\%)")]
bnfchnum.table[, descr := bnfchapters$chaptername[bnfchnum]]
bnfchnum.table <- bnfchnum.table[total >= 20, .(bnfchnum, descr, Controls, Cases)]

table.numdrugs <- cbind(summary(casesnorisk[CASE==0]$numdrugs),
                        summary(casesnorisk[CASE==1]$numdrugs))
colnames(table.numdrugs) <- c("Controls", "Cases")

########################################################
freqs1.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
                                                     "numdrugs", "inpat.recent",
                                                     "listedgr3", 
                                                     "vax14.factor"),
                                        data=cc.all[casegroup=="A" & age_years < 50 & 
                                                #    care.home=="Independent" & 
                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                                        drop.reflevel=FALSE)
freqs2.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
                                                     "numdrugs", "inpat.recent",
                                                     "listedgr3", 
                                                     "vax14.factor"),
                                        data=cc.all[casegroup=="A" & age_years >= 50 & 
                                                #    care.home=="Independent" & 
                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                                        drop.reflevel=FALSE)
freqs.riskgroups <- cbind(freqs1.riskgroups, freqs2.riskgroups)
rownames(freqs.riskgroups) <- replace.names(rownames(freqs.riskgroups))

freqs.vaxgr <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
                                                   "numdrugs", "inpat.recent", "shield.group", 
                                                   "vaxgr"),
                                        data=cc.all[casegroup=="A" &
                                               #     care.home=="Independent" & 
                                                    specimen_date >= as.Date("2020-12-01")],
                                        drop.reflevel=FALSE)
rownames(freqs.vaxgr) <- replace.names(rownames(freqs.vaxgr))

###############################
# rate ratios by risk group x dose
table.riskgr <- summary(clogit(data=cc.all[casegroup=="A" & #care.home=="Independent" & 
                           specimen_date >= as.Date("2020-12-01")],
                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
                               listedgr3/vax14.factor + strata(stratum)))
table.riskgr <- table.riskgr$coefficients
effect <- rownames(table.riskgr)
table.riskgr <- as.data.table(table.riskgr)
table.riskgr[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgr <- table.riskgr[, .(rateratio, pvalue)]
table.riskgr <- data.table(effect=effect, # freqs.riskgr,
                               table.riskgr)
table.riskgr[, effect := gsub("^care.home", "", effect)]
table.riskgr[, effect := gsub("^listedgr3", "", effect)]
table.riskgr[, effect := gsub("TRUE$", "", effect)]
table.riskgr[, effect := gsub(":vax14\\.factor", ": vax dose ", effect)]
table.riskgr[, effect := replace.names(effect)]
table.riskgr <- table.riskgr[c(1:6, 7, 10, 8, 11, 9, 12), ]

# rate ratios by shield group x dose
table.shieldgr <- summary(clogit(data=cc.all[casegroup=="A" & #care.home=="Independent" & 
                           #specimen_date <= as.Date("2021-03-16") &
                           specimen_date >= as.Date("2020-12-01")],
                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
                               shield.group/vax14.dose + strata(stratum)))
table.shieldgr <- table.shieldgr$coefficients
effect <- rownames(table.shieldgr)
table.shieldgr <- as.data.table(table.shieldgr)
table.shieldgr[, rateratio := or.ci(coef, `se(coef)`)]
table.shieldgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.shieldgr <- table.shieldgr[, .(rateratio, pvalue)]
table.shieldgr <- data.table(effect=effect, # freqs.riskgr,
                               table.shieldgr)
table.shieldgr[, effect := gsub("^care.home", "", effect)]
table.shieldgr[, effect := gsub("^shield\\.group", "", effect)]
table.shieldgr[, effect := gsub("TRUE$", "", effect)]
table.shieldgr[, effect := gsub(":vax14\\.dose", ": vax dose", effect)]
table.shieldgr[, effect := replace.names(effect)]

##  rate ratios by risk group x dose-product
table.vaxgr <- summary(clogit(data=cc.all[casegroup=="A" & #care.home=="Independent" & 
                           specimen_date >= as.Date("2020-12-01")],
                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
                               listedgr3/vaxgr + strata(stratum)))
table.vaxgr <- table.vaxgr$coefficients
effect <- rownames(table.vaxgr)
table.vaxgr <- as.data.table(table.vaxgr)
table.vaxgr[, rateratio := or.ci(coef, `se(coef)`)]
table.vaxgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.vaxgr <- table.vaxgr[, .(rateratio, pvalue)]
table.vaxgr <- data.table(effect=effect, # freqs.vaxgr,
                               table.vaxgr)
table.vaxgr[, effect := gsub("^care.home", "", effect)]
table.vaxgr[, effect := gsub("^listedgr3", "", effect)]
table.vaxgr[, effect := gsub("TRUE$", "", effect)]
table.vaxgr[, effect := gsub(":vaxgr", ": ", effect)]
table.vaxgr[, effect := replace.names(effect)]
table.vaxgr <- table.vaxgr[c(1:6,
                             7, 13, 10, 16,
                             8, 14, 11, 17,
                             9, 15, 12, 18), ]

```

This is an update of a report based on data up to 16 March 2021 that was  [posted](https://f1000research.com/articles/10-663) as a preprint.  This update is based on  data extracted on 28 July 2021.  With this updated dataset we are able to look at the effects of two vaccine doses versus one.

# Methods 
This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland [@mckeigue_rapid_2020].  We used this study to take advantage of data linkages already established. The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched for one-year age, sex and primary care practice and alive on the day of presentation of the case that they were matched to were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission or death with COVID-19 ICD-10 codes. The REACT-SCOT case-control dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.  This analysis is based on cases presenting from 1 December 2020 to 14 July 2021, so that we have at least 14 days of follow-up for each case.  This allows case severity status (hospitalisation, entry to critical care or death) to be assigned.

As previously [@mckeigue_rapid_2020], to minimise ascertainment bias we pre-specified the primary outcome measure as severe COVID-19,  defined as diagnosed cases with entry to critical care within 28 days of presentation or fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).  Cases and controls were classified into three broad risk categories: no risk condition; at least one of the moderate risk conditions designated by public health agencies [@mckeigue_rapid_2020]; or eligible for shielding [@mckeigue_relation_2021].   For further analyses, the shielding category was subdivided as described previously  into six categories: solid organ transplant, specific cancers, severe respiratory conditions, other rare conditions, on immunosuppressants, and additional conditions [@mckeigue_relation_2021].

Vaccination status was coded as the number of doses administered at least 14 days before presentation date. Vaccine doses administered less than 14 days before presentation date are ignored.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories. With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.  The strata in the conditional logistic regression model are matched sets: typically one case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.   The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].   The design controls not only for the matching factors of age, sex and general practice but also for calendar time.  Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date.

# Results

Fig \ref{fig:plotcases} shows daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.

Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category, vaccination dose (coded as 0, 1, 2 doses), and recent (5 to 14 days before presentation date) hospital exposure as inpatient.

Table \ref{tab:freqsvaxgr} tabulates risk/shielding group, vaccination status by product, and other risk factors in cases and controls with presentation dates from 1 December 2020 to 14 July 2021.

Table \ref{tab:freqsriskgroups} tabulates risk factors and vaccination status in cases and controls restricted to presentation dates after March 2021

Table \ref{tab:severeagevax} tabulates severe cases presenting after March 2021 by age group and vaccination status.

Table \ref{tab:casesnoriskchnums} tabulates severe cases and controls aged under 50 without risk conditions presenting after March 2021 by frequency of ICD-10 chapters represented in hospital discharge diagnostic codes from 5 years to 25 days before presentation date.

Table \ref{tab:casesnoriskbnfchnums} tabulates severe cases and controls aged under 50 without risk conditions presenting after March 2021 by frequency of BNF chapters represented in prescriptions dispensed from 240 to 15 days before presentation date.

Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories

Table \ref{tab:rateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group

Table \ref{tab:rateratiosvaxgroups} shows rate ratios associated with 1 and 2 doses of each class of vaccine (with 0 doses as reference category) in each of the three broad risk categories.

Supplementary Tables repeat these tables with the case definition broadened to all hospitalized or fatal cases.

Table \ref{tab:hospfreqsvaxgr} tabulates risk/shielding group, vaccination status by product, and other risk factors in cases and controls with presentation dates from 1 December 2020 to 14 July 2021.

Table \ref{tab:hospfreqsriskgroups} tabulates risk factors and vaccination status in cases and controls restricted to presentation dates after March 2021

Table \ref{tab:hosprateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories

Table \ref{tab:hosprateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group

Table \ref{tab:hosprateratiosvaxgroups} shows rate ratios associated with 1 and 2 doses of each class of vaccine (with 0 doses as reference category) in each of the three broad risk categories.

# Conclusions

To reduce the numbers of severe cases, interventions should focus on

* delivering two doses of vaccine to everyone aged over 50, and at least one dose to everyone aged 30-49

* protecting clinically extremely vulnerable individuals, especially by preventing nosocomial transmission and support for other adults in the household to avoid exposure.

To put this in proportion, over 4 months in which there were 122 severe (but mostly non-fatal) cases of COVID-19 in people aged under 50 in Scotland, at current rates there will have been about 400 drug-related deaths in this age group.

# References
<div id="refs"></div>

\newpage

```{r plotcases, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotcases} Daily severe cases by risk category"}

theme_set(theme_gray(base_size = 10))

p.byelig

```

```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window. For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence."}

theme_set(theme_gray(base_size = 10))

p.rateratio

```

\newpage

```{r freqsvaxgroups, echo=FALSE, message=FALSE}

knitr::kable(freqs.vaxgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsvaxgr",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccine product") %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Number of adults in household coded as 1 for care home residents.", 
                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                        "Number of drugs is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Risk/shielding eligibility group", 6, 14, bold=FALSE) %>% 
    kableExtra::group_rows("Vaccination status", 15, 19, bold=FALSE) %>% 
    column_spec(1, width="5cm") %>%
    column_spec(2, width="4cm") %>%
    column_spec(3, width="3cm") %>%
   kable_styling(latex_options=c("HOLD_position"))

```

```{r freqsriskgroups, echo=FALSE, message=FALSE}
knitr::kable(freqs.riskgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsriskgroups",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of severe COVID-19 after March 2021 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Age 0-49 years"=2, "Age 50 years and over"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
                         "Number of adults in household coded as 1 for care home residents.", 
                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Vaccination status", 9, 11, bold=FALSE) %>% 
    column_spec(1, width="6cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="2.5cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2.5cm") %>%
   kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

```{r severebyagevax, echo=FALSE, message=FALSE}

knitr::kable(severe.byagevax, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="severeagevax",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Severe cases after March 2021 by age group and vaccination status") %>%
        add_header_above(c("Age group"=1, "Vaccine doses"=3)) %>%

    kable_styling(latex_options=c("HOLD_position"))

```

```{r casesnoriskchnums, echo=FALSE, message=FALSE}

knitr::kable(chnum.table, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="casesnoriskchnums",
             row.names=FALSE, 
			 align=c('r', 'l', 'r', 'r'), 
             caption="Severe cases and controls aged under 50 without risk conditions presenting after March 2021: ICD-10 chapters represented at least once in discharge diagnoses  from 5 years to 25 days before presentation date") %>%
    column_spec(2, width="7cm") %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Table filtered to retain only ICD-10 chapters represented by at least 20 individuals.",
                         "The most frequent ICD subchapter in cases was F00 (alcohol-related disorders): 7 cases."),
                             notation="none") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r casesnoriskbnfchnums, echo=FALSE, message=FALSE}

knitr::kable(bnfchnum.table, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="casesnoriskbnfchnums",
             row.names=FALSE, 
			 align=c('r', 'l', 'r', 'r'), 
             caption="Severe cases and controls aged under 50 without risk conditions presenting after March 2021: BNF chapters represented at least once in prescriptions dispensed during 240 to 15 days before presentation date") %>%
    column_spec(2, width="7cm") %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Table filtered to retain only BNF chapters represented by at least 20 individuals."),
                             notation="none") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r rateratiosriskgroups, echo=FALSE, message=FALSE}

knitr::kable(table.riskgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosriskgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 nested within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 8, bold=FALSE) %>% 
    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 9, 10, bold=FALSE)  %>% 
    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 11, 12, bold=FALSE)  %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Number of adults in household coded as 1 for care home residents.", 
                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                         "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    column_spec(1, width="6cm") %>%
    column_spec(2, width="4cm") %>%
    column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r rateratiosshieldgroups, echo=FALSE, message=FALSE}

knitr::kable(table.shieldgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 nested within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
    kableExtra::group_rows("Risk / shielding eligibility group, no risk condition as reference category", 5, 12, bold=FALSE) %>% 
    kableExtra::group_rows("Vaccine effect as rate ratio per dose", 13, 20, bold=FALSE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Number of adults in household coded as 1 for care home residents.", 
                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r rateratiosvaxgroups, echo=FALSE, message=FALSE}

knitr::kable(table.vaxgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosvaxgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 nested within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 10, bold=FALSE) %>% 
    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 11, 14, bold=FALSE)  %>% 
    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 15, 18, bold=FALSE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Number of adults in household coded as 1 for care home residents.", 
                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

\newpage

# Supplementary Tables
\beginsupplement

```{r hosp, echo=FALSE, messsage=FALSE}

########################################################
freqs1.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
                                                     "numdrugs", "inpat.recent",
                                                     "listedgr3", 
                                                     "vax14.factor"),
                                        data=cc.all[(casegroup=="A" | casegroup=="B") & age_years < 50 & 
                                                #    care.home=="Independent" & 
                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                                        drop.reflevel=FALSE)
freqs2.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
                                                     "numdrugs", "inpat.recent",
                                                     "listedgr3", 
                                                     "vax14.factor"),
                                        data=cc.all[(casegroup=="A" | casegroup=="B") & age_years >= 50 & 
                                                #    care.home=="Independent" & 
                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                                        drop.reflevel=FALSE)
freqs.riskgroups <- cbind(freqs1.riskgroups, freqs2.riskgroups)
rownames(freqs.riskgroups) <- replace.names(rownames(freqs.riskgroups))

freqs.vaxgr <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
                                                   "numdrugs", "inpat.recent", "shield.group", 
                                                   "vaxgr"),
                                        data=cc.all[(casegroup=="A" | casegroup=="B") &
                                               #     care.home=="Independent" & 
                                                    specimen_date >= as.Date("2020-12-01")],
                                        drop.reflevel=FALSE)
rownames(freqs.vaxgr) <- replace.names(rownames(freqs.vaxgr))

###############################
# rate ratios by risk group x dose
table.riskgr <- summary(clogit(data=cc.all[(casegroup=="A" | casegroup=="B") & #care.home=="Independent" & 
                           specimen_date >= as.Date("2020-12-01")],
                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
                               listedgr3/vax14.factor + strata(stratum)))
table.riskgr <- table.riskgr$coefficients
effect <- rownames(table.riskgr)
table.riskgr <- as.data.table(table.riskgr)
table.riskgr[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgr <- table.riskgr[, .(rateratio, pvalue)]
table.riskgr <- data.table(effect=effect, # freqs.riskgr,
                               table.riskgr)
table.riskgr[, effect := gsub("^care.home", "", effect)]
table.riskgr[, effect := gsub("^listedgr3", "", effect)]
table.riskgr[, effect := gsub("TRUE$", "", effect)]
table.riskgr[, effect := gsub(":vax14\\.factor", ": vax dose ", effect)]
table.riskgr[, effect := replace.names(effect)]
table.riskgr <- table.riskgr[c(1:6, 7, 10, 8, 11, 9, 12), ]

# rate ratios by shield group x dose
table.shieldgr <- summary(clogit(data=cc.all[(casegroup=="A" | casegroup=="B") & #care.home=="Independent" & 
                           #specimen_date <= as.Date("2021-03-16") &
                           specimen_date >= as.Date("2020-12-01")],
                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
                               shield.group/vax14.dose + strata(stratum)))
table.shieldgr <- table.shieldgr$coefficients
effect <- rownames(table.shieldgr)
table.shieldgr <- as.data.table(table.shieldgr)
table.shieldgr[, rateratio := or.ci(coef, `se(coef)`)]
table.shieldgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.shieldgr <- table.shieldgr[, .(rateratio, pvalue)]
table.shieldgr <- data.table(effect=effect, # freqs.riskgr,
                               table.shieldgr)
table.shieldgr[, effect := gsub("^care.home", "", effect)]
table.shieldgr[, effect := gsub("^shield\\.group", "", effect)]
table.shieldgr[, effect := gsub("TRUE$", "", effect)]
table.shieldgr[, effect := gsub(":vax14\\.dose", ": vax dose", effect)]
table.shieldgr[, effect := replace.names(effect)]

##  rate ratios by risk group x dose-product
table.vaxgr <- summary(clogit(data=cc.all[(casegroup=="A" | casegroup=="B") & #care.home=="Independent" & 
                           specimen_date >= as.Date("2020-12-01")],
                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
                               listedgr3/vaxgr + strata(stratum)))
table.vaxgr <- table.vaxgr$coefficients
effect <- rownames(table.vaxgr)
table.vaxgr <- as.data.table(table.vaxgr)
table.vaxgr[, rateratio := or.ci(coef, `se(coef)`)]
table.vaxgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.vaxgr <- table.vaxgr[, .(rateratio, pvalue)]
table.vaxgr <- data.table(effect=effect, # freqs.vaxgr,
                               table.vaxgr)
table.vaxgr[, effect := gsub("^care.home", "", effect)]
table.vaxgr[, effect := gsub("^listedgr3", "", effect)]
table.vaxgr[, effect := gsub("TRUE$", "", effect)]
table.vaxgr[, effect := gsub(":vaxgr", ": ", effect)]
table.vaxgr[, effect := replace.names(effect)]
table.vaxgr <- table.vaxgr[c(1:6,
                             7, 13, 10, 16,
                             8, 14, 11, 17,
                             9, 15, 12, 18), ]

```

```{r hospfreqsvaxgroups, echo=FALSE, message=FALSE}

knitr::kable(freqs.vaxgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hospfreqsvaxgr",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccine product") %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Number of adults in household coded as 1 for care home residents.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                        "Number of drugs is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Risk/shielding eligibility group", 6, 13, bold=FALSE) %>% 
    kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
    column_spec(1, width="5cm") %>%
    column_spec(2, width="4cm") %>%
    column_spec(3, width="3cm") %>%
   kable_styling(latex_options=c("HOLD_position"))

```

```{r hospfreqsriskgroups, echo=FALSE, message=FALSE}
knitr::kable(freqs.riskgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hospfreqsriskgroups",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 after March 2021 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Age 0-49 years"=2, "Age 50 years and over"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
                         "Number of adults in household coded as 1 for care home residents.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Vaccination status", 9, 11, bold=FALSE) %>% 
    column_spec(1, width="6cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="2.5cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2.5cm") %>%
   kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

```{r hosprateratiosriskgroups, echo=FALSE, message=FALSE}

knitr::kable(table.riskgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosriskgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 nested within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 8, bold=FALSE) %>% 
    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 9, 10, bold=FALSE)  %>% 
    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 11, 12, bold=FALSE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Number of adults in household coded as 1 for care home residents.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="6cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r hosprateratiosshieldgroups, echo=FALSE, message=FALSE}

knitr::kable(table.shieldgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 nested within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
    kableExtra::group_rows("Risk / shielding eligibility group, no risk condition as reference category", 5, 12, bold=FALSE) %>% 
    kableExtra::group_rows("Vaccine effect as rate ratio per dose", 13, 20, bold=FALSE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Number of adults in household coded as 1 for care home residents.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r hosprateratiosvaxgroups, echo=FALSE, message=FALSE}

knitr::kable(table.vaxgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosvaxgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 nested within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 10, bold=FALSE) %>% 
    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 11, 14, bold=FALSE)  %>% 
    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 15, 18, bold=FALSE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Number of adults in household coded as 1 for care home residents.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

### 2021-08-08 casecontrol.R

@@ -28,14 +28,8 @@ fatal.predict <- FALSE
 
 ## install all required R packages
 list.of.packages = c("car", 
-                     "survival", 
                      "MASS", 
-                     "wevid", 
                      "rmarkdown",
-                     "bookdown",
-                     "rticles",
-                     "kableExtra",
-                     "pander", 
                      "ggplot2", 
                      "doParallel", 
                      "readxl", 
@@ -53,7 +47,6 @@ library(car)
 library(survival)
 library(MASS)
 library(ggplot2)
-#library(doParallel)
 library(readxl)
 library(DescTools)
 library(icd)
@@ -334,47 +327,17 @@ cc.all[, vaxgr := car::recode(vaxgr,
 #######################################################################################
     
 # source("ct_ecoss.R") # run first time only
-load(paste0(datadir, "ecoss.RData"))
+load(paste0(datadir, "ecoss.firstpos.RData"))
      
 #######################################################################################
 
-if(FALSE) { # import Ct
-    ## left join cc.all with first positive test from ct
-ct <- ct[!is.na(Sgene.dropout)]  ## drop records where the S gene signal is NA
-setkey(ct, SpecimenDate)
-ct.first <- ct[!duplicated(anon_id)]
-setnames(ct.first, "SpecimenDate", "Ct.SpecimenDate") # don't want these fields imported twice
-setnames(ct.first, "flag_lighthouse_labs_testing", "Ct.Lighthouse.flag") 
-                                        #ct.first[, SourceLab := NULL]
-setkey(ct.first, anon_id)
-
-cc.all <- ct.first[, ][cc.all]
-## some controls later tested positive in the Lighthouse lab but were not ascertained as cases by the cutoff date for case ascertainment 
-rm(ct.first)
-rm(ct)
-
-## use max and min sgtf to recode undetermined values where possible
-cc.all[Sgene.dropout=="Undetermined" & max.Sgene.dropout==3, Sgene.dropout := "Definite dropout"]
-cc.all[Sgene.dropout=="Undetermined" & min.Sgene.dropout==1, Sgene.dropout := "No dropout"]
-
-if("flag_covid_symptomatic" %in% names(cc.all)) {
[…]

### 2021-08-08 ct_ecoss.R

@@ -24,6 +24,7 @@ if(linkdate == "jul12") {
     datadir <- "./data/2021-07-28/"
     ct.filename <-  paste0(datadir, "CC_CT_Test_Results_2021-07-28.rds")
     ecoss.tests.filename <- paste0(datadir, "CC_ecoss_tests_2021-07-28.rds")
+    lastdate <- as.Date("2021-07-28")
 }
 
 ################## read Ct table ####################################
@@ -44,23 +45,15 @@ setnames(ct, "test_result", "ct.result")
 ## jan 6 linkage has no variable date_appointment
 setnames(ct, "date_ecoss_specimen", "date_appointment", skip_absent=TRUE)
 
-ct[, date_appointment := as.Date(date_appointment)] # mostly missing
+ct[, date_appointment := as.Date(date_appointment)] # 948 missing
 ct[, date_onset_of_symptoms := as.Date(date_onset_of_symptoms)] 
-ct[, date_reporting := as.Date(date_reporting)] # only 3 missing values and these have all fields missing except ID
+ct[, date_reporting := as.Date(date_reporting)] # no missing values or wrong values
+## mode is 2 days
 
-#print(ct[date_appointment > Sys.Date(), .(date_appointment, date_onset_of_symptoms, date_reporting)])
-#print(ct[date_appointment < as.Date("2020-03-01"), .(date_appointment, date_onset_of_symptoms, date_reporting)])
+ct[date_appointment < as.Date("2020-08-01") | date_appointment > lastdate,
+   date_appointment := date_reporting - 2]
 
-ct[date_appointment > Sys.Date() & lubridate::month(date_appointment) > 9,
-   date_appointment := year.to2020(date_appointment)]
-
-ct[date_appointment < as.Date("2020-03-01") & lubridate::month(date_appointment) < 3,
-   date_appointment := year.to2021(date_appointment)]
-
-ct[date_appointment > Sys.Date() | date_appointment < as.Date("2020-03-01"),
-   date_appointment := NA]
-
-ct[date_onset_of_symptoms < as.Date("2020-02-01"), date_onset_of_symptoms := NA]
+ct[date_onset_of_symptoms < as.Date("2020-03-01"), date_onset_of_symptoms := NA]
 
[…]

### 2021-08-12 vaxshieldupdate.Rmd
vaxshield update

@@
 ---
-title: "Efficacy of COVID-19 vaccination in individuals designated as clinically extremely vulnerable in Scotland: update examining effect of second dose"
+title: "Efficacy of COVID-19 vaccination against severe COVID-19 and risk to  double-vaccinated individuals in Scotland: relationship to co-morbidities conferring  clinically extremely vulnerable status or moderate risk"
 author:
@@
 ```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
-library(data.table)
-library(ggplot2)
 library(kableExtra)
 
-source("helperfunctions.R")
-
-rollmean.dtN <- function(dt, k) {
-    ## FIXME: add a by= argument
-    N.dates <- dt[, .N, by=SPECIMENDATE]
-    setkey(N.dates, SPECIMENDATE)
-    all.dates <- data.table(date=seq(from=min(N.dates$SPECIMENDATE),
-                                     to=max(N.dates$SPECIMENDATE),
-                                     by=1))
-    setkey(all.dates, date)
-
-    ## add rows for missing dates by a left join
-    all.dates <- N.dates[all.dates]
-    all.dates[is.na(N), N := 0]
-    return(data.table(date=dt$SPECIMENDATE,
-                      avdaily=zoo::rollmean(dt$N, k=k, fill=NA)))
-}
-
-rollsum.datewin <- function(dt, k, datevar) {
-    ## returns a table of rolling sums of width k centred on date
-    ## dt is a dataset of totals (N) by date, which may have no rows for some dates in the
-    ## range of interest
-    ## add rows for missing dates by a left join of all.dates with dt
-    all.dates <- data.table(date=seq(from=min(dt[[datevar]]),
-                                     to=max(dt[[datevar]]),
-                                     by=1))
-    setkey(all.dates, date)
-    # setkeyv(dt, datevar) can't set physical key here
-    all.dates <- dt[all.dates]
-
-    return(data.table(date=all.dates[[datevar]],
-                      rsum=zoo::rollsum(all.dates[, N], k=k, fill=0, align="center")))
-}
-
-format.estcipv <- function(x.ci, x.pval) {
-    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
-    x <- gsub(", ", " to ", x)
-    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
-    x <- gsub("times", "\\\\times", x)
-    return(x)
-}
-
-format.estci <- function(x.ci) {
-    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
-    x <- gsub(", ", " to ", x)
-    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
-    #x <- gsub("times", "\\\\times", x)
-    return(x)
-}
+source("vaxshieldupdate.R")
 
@@
 knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
-lastdate.cases <- as.Date("2021-07-12")
-
-## vaccination analysis
-## frequency tables by risk group and dose
-## by shield group and dose
-## by risk group and dose-product
-
-#cc.all[, agegr20plus5 := car::recode(age_years,
-#                                    "0:19='0 to 19'; 20:29='20 to 29'; 30:39='30 to 39'; #40:49='40 to 49'; 50:hi='50 years and over'",
-                                        #                                    as.factor=TRUE)]
-
-cc.all[, shield.group := as.character(shield.group)]
-cc.all[bloodcancer==1,
-       shield.group := car::recode(shield.group, "'Specific cancers'='Blood cancers'")] 
-cc.all[bloodcancer==0,
-       shield.group := car::recode(shield.group, "'Specific cancers'='Other specific cancers'")]
-cc.all[, shield.group := factor(shield.group,
-                                levels=c("No risk condition",
-                                         "Moderate risk condition",
-                                         "Solid organ transplant",
-                                         "Blood cancers",
-                                         "Other specific cancers",
-                                         "Severe respiratory",
-                                         "Rare diseases",
-                                         "On immunosuppressants",
-                                         "Additional conditions"))]
-severe.byagevax <- with(cc.all[CASE==1 & casegroup=="A" &
-                             specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
-                      table(agegr20plus5, vax14.factor))
-severe.byagevax.rowsums <- rowSums(severe.byagevax)
-severe.byagevax <- cbind(severe.byagevax, severe.byagevax.rowsums)
-severe.byagevax <- paste.colpercent(severe.byagevax)
-colnames(severe.byagevax)[4] <- "All"
-
-severe.byagerisk <- with(cc.all[CASE==1 & casegroup=="A" &
-                             specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
-                      table(agegr20plus5, listedgr3))
-severe.byagerisk.colsums <- colSums(severe.byagerisk)
-severe.byagerisk <- rbind(severe.byagerisk, severe.byagerisk.colsums)
-severe.byagerisk <- paste.rowpercent(severe.byagerisk)
-rownames(severe.byagerisk)[6] <- "All"
-
-casesnorisk <- cc.all[casegroup=="A" & listedgr3=="No risk condition" & 
-                              specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases &
-                              age_years < 50] # 91 cases
-
-setkey(casesnorisk, anon_id, specimen_date)
-setkey(cc.diagnoses, anon_id, specimen_date)
-casesnorisk.diagnoses <- cc.diagnoses[casesnorisk]
-
-casesnorisk.chnums <- unique(casesnorisk.diagnoses,
-                                by=c("anon_id", "specimen_date", "chnum"))
-
-chnum.table <- with(casesnorisk.chnums, table(chnum, CASE))
-colnames(chnum.table) <- c("Controls", "Cases")
-chnum.table <- data.table(chnum=as.integer(rownames(chnum.table)),
-                             as.data.frame.matrix(chnum.table))
-chnum.table[, descr := icdchapters$name[as.integer(chnum)]]
-chnum.table[, total := Controls + Cases]
-chnum.table[, Controls := paste0(Controls, " (",
-                                 round(100 * Controls / casesnorisk[CASE==0, .N]),
-                                 "\\%)")]
-chnum.table[, Cases := paste0(Cases, " (",
-                                 round(100 * Cases / casesnorisk[CASE==1, .N]),
-                                 "\\%)")]
-chnum.table[, chnum := as.roman(chnum)]
-chnum.table <- chnum.table[total >= 20, .(chnum, descr, Controls, Cases)]
-
-## 7 with alcohol diagnoses F00
-
-casesnorisk.scrips <- cc.scripsbnf[casesnorisk]
-casesnorisk.scrips[, bnfchnum := substr(bnf_item_code, 1, 2)]
-casesnorisk.bnfchnums <- unique(casesnorisk.scrips, 
-                                by=c("anon_id", "specimen_date", "bnfchnum"))
-bnfchnum.table <- with(casesnorisk.bnfchnums, table(bnfchnum, CASE))
-colnames(bnfchnum.table) <- c("Controls", "Cases")
-bnfchnum.table <- data.table(bnfchnum=as.integer(rownames(bnfchnum.table)),
-                             as.data.frame.matrix(bnfchnum.table))
-bnfchnum.table[, total := Controls + Cases]
-bnfchnum.table[, Controls := paste0(Controls, " (",
-                                 round(100 * Controls / casesnorisk[CASE==0, .N]),
-                                 "\\%)")]
-bnfchnum.table[, Cases := paste0(Cases, " (",
-                                 round(100 * Cases / casesnorisk[CASE==1, .N]),
-                                 "\\%)")]
-bnfchnum.table[, descr := bnfchapters$chaptername[bnfchnum]]
-bnfchnum.table <- bnfchnum.table[total >= 20, .(bnfchnum, descr, Controls, Cases)]
-
-table.numdrugs <- cbind(summary(casesnorisk[CASE==0]$numdrugs),
-                        summary(casesnorisk[CASE==1]$numdrugs))
-colnames(table.numdrugs) <- c("Controls", "Cases")
-
-########################################################
-freqs1.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
-                                                     "numdrugs", "inpat.recent",
-                                                     "listedgr3", 
-                                                     "vax14.factor"),
-                                        data=cc.all[casegroup=="A" & age_years < 50 & 
-                                                #    care.home=="Independent" & 
-                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
-                                        drop.reflevel=FALSE)
-freqs2.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
-                                                     "numdrugs", "inpat.recent",
-                                                     "listedgr3", 
-                                                     "vax14.factor"),
-                                        data=cc.all[casegroup=="A" & age_years >= 50 & 
-                                                #    care.home=="Independent" & 
-                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
-                                        drop.reflevel=FALSE)
-freqs.riskgroups <- cbind(freqs1.riskgroups, freqs2.riskgroups)
-rownames(freqs.riskgroups) <- replace.names(rownames(freqs.riskgroups))
-
-freqs.vaxgr <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
-                                                   "numdrugs", "inpat.recent", "shield.group", 
-                                                   "vaxgr"),
-                                        data=cc.all[casegroup=="A" &
-                                               #     care.home=="Independent" & 
-                                                    specimen_date >= as.Date("2020-12-01")],
-                                        drop.reflevel=FALSE)
-rownames(freqs.vaxgr) <- replace.names(rownames(freqs.vaxgr))
-
-###############################
-# rate ratios by risk group x dose
-table.riskgr <- summary(clogit(data=cc.all[casegroup=="A" & #care.home=="Independent" & 
-                           specimen_date >= as.Date("2020-12-01")],
-                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
-                               listedgr3/vax14.factor + strata(stratum)))
-table.riskgr <- table.riskgr$coefficients
-effect <- rownames(table.riskgr)
-table.riskgr <- as.data.table(table.riskgr)
-table.riskgr[, rateratio := or.ci(coef, `se(coef)`)]
-table.riskgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
-table.riskgr <- table.riskgr[, .(rateratio, pvalue)]
-table.riskgr <- data.table(effect=effect, # freqs.riskgr,
-                               table.riskgr)
-table.riskgr[, effect := gsub("^care.home", "", effect)]
-table.riskgr[, effect := gsub("^listedgr3", "", effect)]
-table.riskgr[, effect := gsub("TRUE$", "", effect)]
-table.riskgr[, effect := gsub(":vax14\\.factor", ": vax dose ", effect)]
-table.riskgr[, effect := replace.names(effect)]
-table.riskgr <- table.riskgr[c(1:6, 7, 10, 8, 11, 9, 12), ]
-
-# rate ratios by shield group x dose
-table.shieldgr <- summary(clogit(data=cc.all[casegroup=="A" & #care.home=="Independent" & 
-                           #specimen_date <= as.Date("2021-03-16") &
-                           specimen_date >= as.Date("2020-12-01")],
-                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
-                               shield.group/vax14.dose + strata(stratum)))
-table.shieldgr <- table.shieldgr$coefficients
-effect <- rownames(table.shieldgr)
-table.shieldgr <- as.data.table(table.shieldgr)
-table.shieldgr[, rateratio := or.ci(coef, `se(coef)`)]
-table.shieldgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
-table.shieldgr <- table.shieldgr[, .(rateratio, pvalue)]
-table.shieldgr <- data.table(effect=effect, # freqs.riskgr,
-                               table.shieldgr)
-table.shieldgr[, effect := gsub("^care.home", "", effect)]
-table.shieldgr[, effect := gsub("^shield\\.group", "", effect)]
-table.shieldgr[, effect := gsub("TRUE$", "", effect)]
-table.shieldgr[, effect := gsub(":vax14\\.dose", ": vax dose", effect)]
-table.shieldgr[, effect := replace.names(effect)]
-
-##  rate ratios by risk group x dose-product
-table.vaxgr <- summary(clogit(data=cc.all[casegroup=="A" & #care.home=="Independent" & 
-                           specimen_date >= as.Date("2020-12-01")],
-                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
-                               listedgr3/vaxgr + strata(stratum)))
-table.vaxgr <- table.vaxgr$coefficients
-effect <- rownames(table.vaxgr)
-table.vaxgr <- as.data.table(table.vaxgr)
-table.vaxgr[, rateratio := or.ci(coef, `se(coef)`)]
-table.vaxgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
-table.vaxgr <- table.vaxgr[, .(rateratio, pvalue)]
-table.vaxgr <- data.table(effect=effect, # freqs.vaxgr,
-                               table.vaxgr)
-table.vaxgr[, effect := gsub("^care.home", "", effect)]
-table.vaxgr[, effect := gsub("^listedgr3", "", effect)]
-table.vaxgr[, effect := gsub("TRUE$", "", effect)]
-table.vaxgr[, effect := gsub(":vaxgr", ": ", effect)]
-table.vaxgr[, effect := replace.names(effect)]
-table.vaxgr <- table.vaxgr[c(1:6,
-                             7, 13, 10, 16,
-                             8, 14, 11, 17,
-                             9, 15, 12, 18), ]
 
 ```
+\newpage
 
-This is an update of a report based on data up to 16 March 2021 that was  [posted](https://f1000research.com/articles/10-663) as a preprint.
-This update is based on  data extracted on 28 July 2021.
-With this updated dataset we are able to look at the effects of two vaccine doses versus one.
+# Abstract
+
+**Objectives** - To determine whether vaccine efficacy varies with risk category and shielding group and to investigate risk factors for severe COVID-19 in those who have received at least two doses of vaccine.
+
+**Design** - Matched case-control study.
+
+**Setting** - Population of Scotland from 1 December 2020 to 28 July 2021.
+
+**Main outcome measure** - Severe COVID-19, defined as cases with entry to critical care or fatal outcome.
+
+**Results** -The rate ratio for severe COVID-19 associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
+Among those eligible for shielding, there was a dose-response relationship and only weak evidence (p=0.04 in a test for interaction) that efficacy against hospitalized/severe disease was lower for the AZ vaccine than for the mRNA vaccines.
+Among the double-vaccinated, 87% of severe cases and 77% of hospitalised cases had designated risk conditions or were eligible for shielding.
+With those with no risk conditions as reference category, the rate ratio associated with receipt of a solid organ transplant among the double-vaccinated was `r format.estci(table.severe.vax2["Solid organ transplant", 3])` for severe disease and `r format.estci(table.hosp.vax2["Solid organ transplant", 3])` for hospitalisation.
+A risk score calculated from health records had moderate predictive performance (C-statistic 0.80) hospitalized cases and age-sex-matched controls among the double-vaccinated.
+
+**Conclusions** - 
+Although efficacy of vaccination against severe disease in those without risk conditions remains high, it is now clear that efficacy in clinically extremely vulnerable individuals is considerably lower.
+Solid organ transplant recipients remain at high risk of severe disease even after two doses of vaccine.
+
+\newpage
+
+# Introduction
+The REACT-SCOT matched case-control study was established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland [@mckeigue_rapid_2020].
+Using this framework, we have reported in detail on the relation of severe COVID-19 to risk factors including designated moderate risk conditions, drug prescribing and eligiblility for shielding [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021].
+We reported a study of vaccine efficacy against severe disease in different risk groups based on data up to 16 March 2021 [@mckeigue_efficacy_2021].
+As most people on the shielding list did not receive their first dose of vaccine until February 2021, there were too few people in this group who had received a second dose of vaccine at least 14 days earlier for the rate ratios associated with two vaccine doses to be estimated at this time.
+
+This report updates and extends the earlier analyses, with the following objectives:
+
+1. To examine the time course, since vaccination began, of the relation of severe COVID-19 to vaccination status and risk conditions.
+
+2. To determine whether vaccine efficacy varies with risk category and shielding group now that more data, including exposure to two doses of vaccine, are available.
+
+3. To investigate risk factors for infections causing severe illness or hospitalisation in those who have received at least two doses of vaccine.
 
 # Methods 
-This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland [@mckeigue_rapid_2020].
-We used this study to take advantage of data linkages already established.
+We used the REACT-SCOT study to take advantage of data linkages already established.
 The design has been described in detail previously [@mckeigue_rapid_2020].
@@
 This allows case severity status (hospitalisation, entry to critical care or death) to be assigned.
+For time-window analyses of length \ensuremath{2n + 1} days the time-window centred on 1 December 2020 includes the \ensuremath{n} days before that date.
 
+## Classification of risk categories
 As previously [@mckeigue_rapid_2020], to minimise ascertainment bias we pre-specified the primary outcome measure as severe COVID-19,  defined as diagnosed cases with entry to critical care within 28 days of presentation or fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).
@@
 For further analyses, the shielding category was subdivided as described previously  into six categories: solid organ transplant, specific cancers, severe respiratory conditions, other rare conditions, on immunosuppressants, and additional conditions [@mckeigue_relation_2021].
+This corresponds to the list used by Public Health Scotland  [@health_protection_scotland_search_2020], with the small numbers in the group "pregnant with heart disease combined with the "other conditions" category.
+For additional analyses the category "specific cancers" was split to allocate cancers of blood-forming organs (ICD-10 codes C81-C88, C90-C96) to a separate category.
 
-Vaccination status was coded as the number of doses administered at least 14 days before presentation date. Vaccine doses administered less than 14 days before presentation date are ignored.
+## Statistical analysis
+Vaccination status was coded as the number of doses administered at least 14 days before presentation date.
+Vaccine doses administered less than 14 days before presentation date are ignored.
+For estimates of effect by risk category, where the numbers of events are large, vaccination status was modelled as a factor comparing 1 dose or 2 doses with 0 doses as reference category.
+For estimates of effect within shielding eligibility groups, and for comparisons between the mRNA and AZ vaccines where the numbers of events in each category are smaller, vaccination status was modelled as a numeric variable taking values \ensuremath{0, \frac{1}{2}, 1}  so that the effect parameter represents the rate ratio associated with 2 doses assuming the effect of 0, 1, 2 doses to be linear on the scale of the log rate ratio.
+
+Covariates included in the models were those that have been previously identified as strong predictors of severe disease in this population: care home residence, number of adults in household, number of drug classes dispensed and recent hospital stay [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021].
+For care home residents the number of of adults in the household was coded as 1.
+The number of drug classes was calculated as the number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.
+As we have previously shown that severe COVID-19 in Scotland is associated with the number of non-cardiovascular but not the number of cardiovascular drug classes [@mckeigue_relation_2021a], the covariate included in the model was the number of non-cardiovascular drug classes.
+The number of hospital diagnoses was calculated as the number of distinct ICD-10 chapters represented at least once in hospital discharge records between 25 days and 5 years before presentation date.
+Recent hospital stay was defined as any in-patient stay from 5 to 14 days before presentation date.
 
@@
 
+# Cohort analysis of the shielding list
+The case-control study can estimate only rate ratios.
+To investigate the absolute rates of severe disease in those listed as clinically extremely vulnerable, we undertook a separate cohort study of all individuals who had ever been on the shielding list.
+At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.
+A Poisson regression model was fitted to a dataset formatted with one observation per 28-day person-time interval, and individuals censored at first event.
+The event was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, age, and shielding eligibility group.
+The baseline hazard rate was specified as a natural spline function with 4 degrees of freedom.
+Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.
+
 # Results
-
-Fig \ref{fig:plotcases} shows daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.
+## Time course of the epidemic and associations with risk factors 
+Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.
+From mid-January to early March 2021 there was a rapid decline in the incidence of severe cases, followed by a slight increase from early June to early July.
 
 Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category, vaccination dose (coded as 0, 1, 2 doses), and recent (5 to 14 days before presentation date) hospital exposure as inpatient.
+These rate ratios are computed from a model that includes all variables shown, together with care home residence.
 
-Table \ref{tab:freqsvaxgr} tabulates risk/shielding group, vaccination status by product, and other risk factors in cases and controls with presentation dates from 1 December 2020 to 14 July 2021.
+The rate ratio for severe disease associated with shielding eligibility did not decline as the winter peak declined, and increased from May 2020 onwards.
+The efficacy of vaccination against severe disease (modelled as the rate ratio per 2 doses) remained high, despite the rapid replacement of the Alpha by the Delta variant during May 2020.
 
-Table \ref{tab:freqsriskgroups} tabulates risk factors and vaccination status in cases and controls restricted to presentation dates after March 2021
+## Relation of vaccine efficacy to risk conditions
+Table \ref{tab:freqsvaxgr} tabulates risk/shielding group, vaccination status by product, and other risk factors in severe cases and controls with presentation dates from 1 December 2020 to 14 July 2021.
+Table \ref{tab:hospfreqsvaxgr} shows the same tabulation with case definition broadened to include all hospitalised cases and their matched controls.
 
-Table \ref{tab:severeagevax} tabulates severe cases presenting after March 2021 by age group and vaccination status.
+Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories, in a model that includes risk categories (no risk conditions as reference category),  care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates.
+The rate ratio associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
+On the scale of efficacy, these estimates are equivalent to 94% in those without risk conditions, 91% in those with risk conditions, and only 68% in those eligible for shielding.
+To test formally for an effect of risk category on vaccine efficacy, a model was fitted with interaction terms for  risk category \ensuremath{\times} times vaccine dose.
+In those with 2 doses, the effect (on the scale of rate ratio) of eligibility for shielding (with no risk condition as reference category) on the rate ratio associated with 2 doses of vaccine (0 doses as reference category) was `r format.estci(vaxriskgr.interaction)`.
 
-Table \ref{tab:casesnoriskchnums} tabulates severe cases and controls aged under 50 without risk conditions presenting after March 2021 by frequency of ICD-10 chapters represented in hospital discharge diagnostic codes from 5 years to 25 days before presentation date.
+To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower efficacy of vaccines in the clinically extremely vulnerable, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.
+With this restricted case definition, the rate ratio associated with two doses of vaccine was `r format.estci(table.nonfatalorucod[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.nonfatalorucod[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.nonfatalorucod[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
+Table \ref{tab:hosprateratiosriskgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls: the efficacy of 2 doses was lower in those eligible for shielding than in the other two risk categories.
 
-Table \ref{tab:casesnoriskbnfchnums} tabulates severe cases and controls aged under 50 without risk conditions presenting after March 2021 by frequency of BNF chapters represented in prescriptions dispensed from 240 to 15 days before presentation date.
-
-Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories
-
-Table \ref{tab:rateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group
+Table \ref{tab:rateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group.
+The efficacy per dose was lower in all shielding eligibility groups than in those without risk conditions  Table \ref{tab:hosprateratiosshieldgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.
+Although there is evidence of efficacy in all shielding eligibility groups, the efficacy estimate was lowest in solid organ transplant recipients.
 
 Table \ref{tab:rateratiosvaxgroups} shows rate ratios associated with 1 and 2 doses of each class of vaccine (with 0 doses as reference category) in each of the three broad risk categories.
+Table \ref{tab:hosprateratiosvaxgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.
+In those eligible for shielding, a separate analysis was undertaken to test for difference in efficacy between the AZ vaccine and mRNA vaccines.
+In conditional logistic regression models with vaccine dose and vaccine product centred at zero mean and other covariates as in Table  \ref{tab:hosprateratiosvaxgroups}, there was no evidence of lower efficacy against severe COVID-19 for the AZ vaccine and only weak evidence for lower efficacy against the broader category of hospitalized or fatal COVID-19: the rate ratio associated with vaccine dose was 1.51 (95% CI 0.66 to 1.03, _p_=0.04) times higher for the AZ vaccine than for the Pfizer vaccine.
 
-Supplementary Tables repeat these tables with the case definition broadened to all hospitalized or fatal cases.
+## Risk factors for severe COVID-19 in the double-vaccinated
+Table \ref{tab:severevax2} shows risk factors for severe COVID-19 in cases and controls who had received 2 doses of vaccine at least 14 days before presentation date.
+Table \ref{tab:hospvax2} shows the same analysis with case definition broadened to include hospitalized cases and matched controls who had received 2 doses of vaccine.
+Of the variables that we have previously reported as risk factors for severe COVID-19 in the general population, care home residence and socioeconomic deprivation were not associated with severe disease in the double-vaccinated.
+In this group the risk factors for severe disease were   designated risk conditions, other indicators of co-morbidity including numbers of hospital diagnoses and drug classes prescribed and transmission-related factors including number of adults in household and recent hospital stay.
+Only 13% of double-vaccinated severe cases had no designated risk condition: 44% had at least one moderate risk condition and 43% were eligible for shielding.
+With those with no risk conditions as reference category, the rate ratio associated with receipt of a solid organ transplant among the double-vaccinated was `r format.estci(table.severe.vax2["Solid organ transplant", 3])` for severe disease and `r format.estci(table.hosp.vax2["Solid organ transplant", 3])` for hospitalisation.
 
-Table \ref{tab:hospfreqsvaxgr} tabulates risk/shielding group, vaccination status by product, and other risk factors in cases and controls with presentation dates from 1 December 2020 to 14 July 2021.
+A risk score for hospitalised or fatal COVID-19 in the double-vaccinated calculated from the conditional logistic regression model had a C-statistic of 0.80.
 
-Table \ref{tab:hospfreqsriskgroups} tabulates risk factors and vaccination status in cases and controls restricted to presentation dates after March 2021
+<!---
+## Associations with severe diseases after the first quarter of 2021
 
-Table \ref{tab:hosprateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories
+Table \ref{tab:freqsriskgroups} tabulates risk factors and vaccination status in severe cases and controls restricted to presentation dates from April to July 2021, by age dichotomized at 50 years.
+Among severe cases aged over 50 years, 75% had been vaccinated with at least one dose, and 74% had a moderate risk condition or were eligible for shielding.
+Among severe cases aged less than 50, 81% had not been vaccinated, and 74% had no risk condition.
 
-Table \ref{tab:hosprateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group
+Table \ref{tab:severeagevax} tabulates severe cases presenting from April to July 2021 by 10-year age group and vaccination status, showing that unvaccinated individuals aged 30-49 made up 129 of of the 171 severe cases among those aged less than 50 years.
 
-Table \ref{tab:hosprateratiosvaxgroups} shows rate ratios associated with 1 and 2 doses of each class of vaccine (with 0 doses as reference category) in each of the three broad risk categories.
+Table \ref{tab:casesnoriskchnums} tabulates severe cases and matched controls aged under 50 years without designated risk conditions presenting from April to July 2021, with frequencies in each chapter of ICD-10 of at least one hospital discharge diagnosis  from 5 years to 25 days before presentation date.
+Table  \ref{tab:casesnoriskbnfchnums} tabulates for the same subset of cases and controls the frequencies in each chapter of the BNF of at least one prescription dispensed  from 240 to 15 days before presentation date.
+In comparison with their matched controls, the cases had higher frequencies of admissions across several chapter headings including diseases of the digestive system, and other more non-specific chapter headings, and higher frequencies of at least one dispensed prescription in most chapters of the BNF.
+--->
 
-# Conclusions
+# Discussion
+## Statement of principal findings
 
-To reduce the numbers of severe cases, interventions should focus on
+1.
+Although the efficacy of vaccination against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is considerably lower (about 70%).
+There is however a dose-response effect in this group: two doses are more effective than one.
 
-* delivering two doses of vaccine to everyone aged over 50, and at least one dose to everyone aged 30-49
+2.
+Solid organ transplant recipients remain at very high risk of severe disease even after two doses of vaccine: approximately 70-fold for severe disease and 30-fold for hospitalisation in comparison with double-vaccinated individuals of the same age and sex without risk conditions.
 
-* protecting clinically extremely vulnerable individuals, especially by preventing nosocomial transmission and support for other adults in the household to avoid exposure.
+3. There is only weak evidence that efficacy against severe disease disease is lower for the AZ vaccine than for the mRNA vaccines.
 
-To put this in proportion, over 4 months in which there were 122 severe (but mostly non-fatal) cases of COVID-19 in people aged under 50 in Scotland, at current rates there will have been about 400 drug-related deaths in this age group.
+4.
+Among the double-vaccinated, 87% of severe cases and 77% of hospitalised cases have designated risk conditions or are eligible for shielding.
+A risk score calculated from health records has moderately high performance as a predictor of severe COVID-19 or hospitalisation among the double-vaccinated, independent of age and sex.
+
+## Strengths and limitations
+Strengths of this study are the focus on severe cases rather than test-positives, the matched case-control design and the comprehensive linkage to e-health records
+
+The main limitation of this study is that no data on hospital prescribing are available.
+Most biologic immunosuppressants and drugs for cancer are prescribed only in hospital, and the records of prescribing are not held in electronic form.
+The national shielding list does not hold any information about drug therapy, and no information about diagnoses other than the seven broad categories listed as eligible for shielding.
+
+## Relation to other studies
+A recent report based on primary care data covering 7.2 million people curated by the Royal College of General Practitioners Research and Surveillance Centre examined vaccine efficacy against symptomatic COVID-19 for which a medical consultation was recorded  [@whitaker_pfizerbiontech_2021].
+For those who had been advised to shield, the efficacy after two doses of any vaccine was estimated as 92% in those aged less than 65 years and 84% in those aged 65 years and over.
+For the "immunocompromised" group the estimated vaccine efficacy was somewhat lower at 74%, with a wide confidence interval.
+No separate analysis of vaccine efficacy in solid organ transplant recipients was reported.
+
+## Policy implications. 
+The dose-response effect suggests that an additional dose might give more protection to double-vaccinated individuals who are clinically extremely vulnerable.
+
+For identifying those among the double-vaccinated who are at high risk of severe disease, a   risk score based on electronic health records has better discrimination than using only shielding eligibility.
+However if vaccine supply is not constrained, a policy of offering booster vaccination to all those above an age threshold and all those with designated risk conditions would achievetogether with acombined would cover more than 80% of those at risk of severe disease.
 
@@
 
-```{r plotcases, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotcases} Daily severe cases by risk category"}
-
-theme_set(theme_gray(base_size = 10))
-
-p.byelig
-
-```
-
-```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window.
-For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence."}
-
-theme_set(theme_gray(base_size = 10))
-
-p.rateratio
-
-```
-
-\newpage
+# Tables
 
@@
     add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Number of adults in household coded as 1 for care home residents.", 
                          "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                          "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                        "Number of drugs is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                          "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                  notation="none") %>%
-    kableExtra::group_rows("Risk/shielding eligibility group", 6, 14, bold=FALSE) %>% 
-    kableExtra::group_rows("Vaccination status", 15, 19, bold=FALSE) %>% 
+    kableExtra::group_rows("Risk/shielding eligibility group", 6, 13, bold=FALSE) %>% 
+    kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
     column_spec(1, width="5cm") %>%
@@
 
-```{r freqsriskgroups, echo=FALSE, message=FALSE}
-knitr::kable(freqs.riskgroups, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="freqsriskgroups",
-             row.names=TRUE, 
-			 align=c('r', 'r'), 
-             caption="Numbers of controls and cases of severe COVID-19 after March 2021 by risk group and vaccination status") %>%
-    add_header_above(c(" "=1, "Age 0-49 years"=2, "Age 50 years and over"=2)) %>%
-    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
-                         "Number of adults in household coded as 1 for care home residents.", 
-                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
-                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                         "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
-                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-                 notation="none") %>%
-    kableExtra::group_rows("Vaccination status", 9, 11, bold=FALSE) %>% 
-    column_spec(1, width="6cm") %>%
-    column_spec(2, width="3cm") %>%
-    column_spec(3, width="2.5cm") %>%
-    column_spec(4, width="3cm") %>%
-    column_spec(5, width="2.5cm") %>%
-   kable_styling(latex_options=c("HOLD_position", "scale_down"))
-
-```
-
-```{r severebyagevax, echo=FALSE, message=FALSE}
-
-knitr::kable(severe.byagevax, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="severeagevax",
-             row.names=TRUE, 
-			 align=c('r', 'r'), 
-             caption="Severe cases after March 2021 by age group and vaccination status") %>%
-        add_header_above(c("Age group"=1, "Vaccine doses"=3)) %>%
-
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r casesnoriskchnums, echo=FALSE, message=FALSE}
-
-knitr::kable(chnum.table, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="casesnoriskchnums",
-             row.names=FALSE, 
-			 align=c('r', 'l', 'r', 'r'), 
-             caption="Severe cases and controls aged under 50 without risk conditions presenting after March 2021: ICD-10 chapters represented at least once in discharge diagnoses  from 5 years to 25 days before presentation date") %>%
-    column_spec(2, width="7cm") %>%
-    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Table filtered to retain only ICD-10 chapters represented by at least 20 individuals.",
-                         "The most frequent ICD subchapter in cases was F00 (alcohol-related disorders): 7 cases."),
-                             notation="none") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r casesnoriskbnfchnums, echo=FALSE, message=FALSE}
-
-knitr::kable(bnfchnum.table, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="casesnoriskbnfchnums",
-             row.names=FALSE, 
-			 align=c('r', 'l', 'r', 'r'), 
-             caption="Severe cases and controls aged under 50 without risk conditions presenting after March 2021: BNF chapters represented at least once in prescriptions dispensed during 240 to 15 days before presentation date") %>%
-    column_spec(2, width="7cm") %>%
-    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Table filtered to retain only BNF chapters represented by at least 20 individuals."),
-                             notation="none") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
 ```{r rateratiosriskgroups, echo=FALSE, message=FALSE}
 
-knitr::kable(table.riskgr, 
+knitr::kable(table.riskgr[7:12, ], 
              escape=FALSE, 
@@
 			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for severe COVID-19 nested within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 8, bold=FALSE) %>% 
-    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 9, 10, bold=FALSE)  %>% 
-    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 11, 12, bold=FALSE)  %>% 
+             caption="Rate ratios for severe COVID-19 within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
+    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 1, 2, bold=TRUE) %>% 
+    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 3, 4, bold=TRUE)  %>% 
+    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 5, 6, bold=TRUE)  %>% 
     add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Number of adults in household coded as 1 for care home residents.", 
                          "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                          "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
+                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                      "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                         "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
-                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
+                         "Vaccination status is doses administered at least 14 days fbefore presentation date"), 
                  notation="none") %>%
@@
 
-knitr::kable(table.shieldgr, 
+knitr::kable(table.shieldgr[14:21, ], 
              escape=FALSE, 
@@
 			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for severe COVID-19 nested within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
-    kableExtra::group_rows("Risk / shielding eligibility group, no risk condition as reference category", 5, 12, bold=FALSE) %>% 
-    kableExtra::group_rows("Vaccine effect as rate ratio per dose", 13, 20, bold=FALSE)  %>% 
+             caption="Rate ratios for severe COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1/2, 1") %>%
 add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Number of adults in household coded as 1 for care home residents.", 
                      "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                      "Controls matched for age, sex, and primary care practice on date of presentation of case.",
+                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                      "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                      "Vaccination status is number of doses administered at least 14 days before presentation date"), 
@@
 
-knitr::kable(table.vaxgr, 
+startrow <- 1
+knitr::kable(table.vaxgr[7:18,], 
              escape=FALSE, 
@@
 			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for severe COVID-19 nested within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 10, bold=FALSE) %>% 
-    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 11, 14, bold=FALSE)  %>% 
-    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 15, 18, bold=FALSE)  %>% 
+             caption="Rate ratios for severe COVID-19 within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
+    kableExtra::group_rows("No risk condition", startrow, startrow+3, bold=TRUE) %>% 
+    kableExtra::group_rows("Moderate risk condition", startrow+4, startrow+7, bold=TRUE)  %>% 
+    kableExtra::group_rows("Eligible for shielding", startrow+8, startrow+11, bold=TRUE)  %>% 
 add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Number of adults in household coded as 1 for care home residents.", 
-                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
+                     "Adjusted for care home residence, number of adults in household, number of drug classes dispensed, recent hospital stay and risk category.", 
                      "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                      "Vaccination status is number of doses administered at least 14 days before presentation date"), 
@@
 
-\newpage
+```{r severevax2, echo=FALSE, message=FALSE}
 
-# Supplementary Tables
-\beginsupplement
-
-```{r hosp, echo=FALSE, messsage=FALSE}
-
-########################################################
-freqs1.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
-                                                     "numdrugs", "inpat.recent",
-                                                     "listedgr3", 
-                                                     "vax14.factor"),
-                                        data=cc.all[(casegroup=="A" | casegroup=="B") & age_years < 50 & 
-                                                #    care.home=="Independent" & 
-                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
-                                        drop.reflevel=FALSE)
-freqs2.riskgroups <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
-                                                     "numdrugs", "inpat.recent",
-                                                     "listedgr3", 
-                                                     "vax14.factor"),
-                                        data=cc.all[(casegroup=="A" | casegroup=="B") & age_years >= 50 & 
-                                                #    care.home=="Independent" & 
-                                                    specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
-                                        drop.reflevel=FALSE)
-freqs.riskgroups <- cbind(freqs1.riskgroups, freqs2.riskgroups)
-rownames(freqs.riskgroups) <- replace.names(rownames(freqs.riskgroups))
-
-freqs.vaxgr <- univariate.tabulate(varnames=c("care.home", "qSIMD.integer", "hh.over18", 
-                                                   "numdrugs", "inpat.recent", "shield.group", 
-                                                   "vaxgr"),
-                                        data=cc.all[(casegroup=="A" | casegroup=="B") &
-                                               #     care.home=="Independent" & 
-                                                    specimen_date >= as.Date("2020-12-01")],
-                                        drop.reflevel=FALSE)
-rownames(freqs.vaxgr) <- replace.names(rownames(freqs.vaxgr))
-
-###############################
-# rate ratios by risk group x dose
-table.riskgr <- summary(clogit(data=cc.all[(casegroup=="A" | casegroup=="B") & #care.home=="Independent" & 
-                           specimen_date >= as.Date("2020-12-01")],
-                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
-                               listedgr3/vax14.factor + strata(stratum)))
-table.riskgr <- table.riskgr$coefficients
-effect <- rownames(table.riskgr)
-table.riskgr <- as.data.table(table.riskgr)
-table.riskgr[, rateratio := or.ci(coef, `se(coef)`)]
-table.riskgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
-table.riskgr <- table.riskgr[, .(rateratio, pvalue)]
-table.riskgr <- data.table(effect=effect, # freqs.riskgr,
-                               table.riskgr)
-table.riskgr[, effect := gsub("^care.home", "", effect)]
-table.riskgr[, effect := gsub("^listedgr3", "", effect)]
-table.riskgr[, effect := gsub("TRUE$", "", effect)]
-table.riskgr[, effect := gsub(":vax14\\.factor", ": vax dose ", effect)]
-table.riskgr[, effect := replace.names(effect)]
-table.riskgr <- table.riskgr[c(1:6, 7, 10, 8, 11, 9, 12), ]
-
-# rate ratios by shield group x dose
-table.shieldgr <- summary(clogit(data=cc.all[(casegroup=="A" | casegroup=="B") & #care.home=="Independent" & 
-                           #specimen_date <= as.Date("2021-03-16") &
-                           specimen_date >= as.Date("2020-12-01")],
-                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
-                               shield.group/vax14.dose + strata(stratum)))
-table.shieldgr <- table.shieldgr$coefficients
-effect <- rownames(table.shieldgr)
-table.shieldgr <- as.data.table(table.shieldgr)
-table.shieldgr[, rateratio := or.ci(coef, `se(coef)`)]
-table.shieldgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
-table.shieldgr <- table.shieldgr[, .(rateratio, pvalue)]
-table.shieldgr <- data.table(effect=effect, # freqs.riskgr,
-                               table.shieldgr)
-table.shieldgr[, effect := gsub("^care.home", "", effect)]
-table.shieldgr[, effect := gsub("^shield\\.group", "", effect)]
-table.shieldgr[, effect := gsub("TRUE$", "", effect)]
-table.shieldgr[, effect := gsub(":vax14\\.dose", ": vax dose", effect)]
-table.shieldgr[, effect := replace.names(effect)]
-
-##  rate ratios by risk group x dose-product
-table.vaxgr <- summary(clogit(data=cc.all[(casegroup=="A" | casegroup=="B") & #care.home=="Independent" & 
-                           specimen_date >= as.Date("2020-12-01")],
-                           formula=CASE ~ care.home + hh.over18 + numdrugs + inpat.recent +
-                               listedgr3/vaxgr + strata(stratum)))
-table.vaxgr <- table.vaxgr$coefficients
-effect <- rownames(table.vaxgr)
-table.vaxgr <- as.data.table(table.vaxgr)
-table.vaxgr[, rateratio := or.ci(coef, `se(coef)`)]
-table.vaxgr[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
-table.vaxgr <- table.vaxgr[, .(rateratio, pvalue)]
-table.vaxgr <- data.table(effect=effect, # freqs.vaxgr,
-                               table.vaxgr)
-table.vaxgr[, effect := gsub("^care.home", "", effect)]
-table.vaxgr[, effect := gsub("^listedgr3", "", effect)]
-table.vaxgr[, effect := gsub("TRUE$", "", effect)]
-table.vaxgr[, effect := gsub(":vaxgr", ": ", effect)]
-table.vaxgr[, effect := replace.names(effect)]
-table.vaxgr <- table.vaxgr[c(1:6,
-                             7, 13, 10, 16,
-                             8, 14, 11, 17,
-                             9, 15, 12, 18), ]
+knitr::kable(table.severe.vax2, 
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="severevax2",
+             row.names=TRUE, 
+            caption="Risk factors for severe COVID-19 in those who had received 2 doses of vaccine at least 14 days before", 
+             col.names=c(clean.header(colnames(table.severe.vax2)[1:2]),
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
+    column_spec(column=1, width="4cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="3cm") %>%
+    column_spec(column=5, width="1.7cm") %>%
+    column_spec(column=6, width="2.7cm") %>%
+    column_spec(column=7, width="1.7cm") %>%
+    kableExtra::group_rows("Shielding eligibility category", 9, 14, bold=FALSE) %>%
+   add_footnote(label=c("Severe COVID-19 is defined by entry to critical care or fatal outcome", 
+                         "Percentages are column percentages for each variable",
+                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
+                         "Univariate rate ratios are for models with a single covariate",
+                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "HOLD_position"))
 
@@
 
+# Supplementary Material
+\beginsupplement
+
+## Supplementary Figures
+
+```{r plotcases, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotcases} Daily severe cases by risk category"}
+
+theme_set(theme_gray(base_size = 10))
+
+p.byelig
+
+```
+
+```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window.
+For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence.
+Line thickness is inversely proportional to standard error"}
+
+theme_set(theme_gray(base_size = 10))
+
+p.rateratio
+
+```
+
+## Supplementary Tables
+
 ```{r hospfreqsvaxgroups, echo=FALSE, message=FALSE}
 
-knitr::kable(freqs.vaxgr, 
+knitr::kable(freqs.hosp.vaxgr, 
              escape=FALSE, 
@@
     add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Number of adults in household coded as 1 for care home residents.", 
                          "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                        "Number of drugs is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                          "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                  notation="none") %>%
-    kableExtra::group_rows("Risk/shielding eligibility group", 6, 13, bold=FALSE) %>% 
+    kableExtra::group_rows("Risk/shielding eligibility group", 6, 12, bold=FALSE) %>% 
     kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
@@
 
-```{r hospfreqsriskgroups, echo=FALSE, message=FALSE}
-knitr::kable(freqs.riskgroups, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hospfreqsriskgroups",
-             row.names=TRUE, 
-			 align=c('r', 'r'), 
-             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 after March 2021 by risk group and vaccination status") %>%
-    add_header_above(c(" "=1, "Age 0-49 years"=2, "Age 50 years and over"=2)) %>%
-    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
-                         "Number of adults in household coded as 1 for care home residents.", 
-                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                         "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                         "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
-                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-                 notation="none") %>%
-    kableExtra::group_rows("Vaccination status", 9, 11, bold=FALSE) %>% 
-    column_spec(1, width="6cm") %>%
-    column_spec(2, width="3cm") %>%
-    column_spec(3, width="2.5cm") %>%
-    column_spec(4, width="3cm") %>%
-    column_spec(5, width="2.5cm") %>%
-   kable_styling(latex_options=c("HOLD_position", "scale_down"))
-
-```
-
 ```{r hosprateratiosriskgroups, echo=FALSE, message=FALSE}
 
-knitr::kable(table.riskgr, 
+knitr::kable(table.hosp.riskgr[7:12, ], 
              escape=FALSE, 
@@
 			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for hospitalized or fatal COVID-19 nested within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 8, bold=FALSE) %>% 
-    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 9, 10, bold=FALSE)  %>% 
-    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 11, 12, bold=FALSE)  %>% 
+             caption="Rate ratios for hospitalized or fatal COVID-19 within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
+    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 1, 2, bold=FALSE) %>% 
+    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 3, 4, bold=FALSE)  %>% 
+    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 5,  6, bold=FALSE)  %>% 
 add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Number of adults in household coded as 1 for care home residents.", 
                      "Controls matched for age, sex, and primary care practice on date of presentation of case.",
+                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                      "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                      "Vaccination status is number of doses administered at least 14 days before presentation date"), 
@@
 
-knitr::kable(table.shieldgr, 
+knitr::kable(table.hosp.shieldgr[13:20, ], 
              escape=FALSE, 
@@
 			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for hospitalized or fatal COVID-19 nested within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
-    kableExtra::group_rows("Risk / shielding eligibility group, no risk condition as reference category", 5, 12, bold=FALSE) %>% 
-    kableExtra::group_rows("Vaccine effect as rate ratio per dose", 13, 20, bold=FALSE)  %>% 
+             caption="Rate ratios for hospitalized or fatal COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
 add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Number of adults in household coded as 1 for care home residents.", 
                      "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                      "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
@@
 
-knitr::kable(table.vaxgr, 
+startrow <- 1
+knitr::kable(table.hosp.vaxgr[7:18,], 
              escape=FALSE, 
@@
 			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for hospitalized or fatal COVID-19 nested within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 7, 10, bold=FALSE) %>% 
-    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 11, 14, bold=FALSE)  %>% 
-    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 15, 18, bold=FALSE)  %>% 
+             caption="Rate ratios for hospitalized or fatal COVID-19 within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
+    kableExtra::group_rows("No risk condition", startrow, startrow+3, bold=TRUE) %>% 
+    kableExtra::group_rows("Moderate risk condition", startrow+4, startrow+7, bold=TRUE)  %>% 
+    kableExtra::group_rows("Eligible for shielding", startrow+8, startrow+11, bold=TRUE)  %>% 
 add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Number of adults in household coded as 1 for care home residents.", 
+                     "Adjusted for care home residence, number of adults in household, number of drug classes dispensed, recent hospital stay and risk category.", 
                      "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Number of drug classes is number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.", 
-                     "Recent hospital stay is any in-patient stay from 5 to 14 days before presentation date.", 
                      "Vaccination status is number of doses administered at least 14 days before presentation date"), 
@@
 ```
+
+```{r hospvax2, echo=FALSE, message=FALSE}
+
+knitr::kable(table.hosp.vax2, 
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="hospvax2",
+             row.names=TRUE, 
+            caption="Risk factors for hospitalised or fatalCOVID-19 in those who had received 2 doses of vaccine at least 14 days before", 
+             col.names=c(clean.header(colnames(table.severe.vax2)[1:2]),
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
+    column_spec(column=1, width="4cm") %>%
+    column_spec(column=2, width="2cm") %>%
+    column_spec(column=3, width="2cm") %>%
+    column_spec(column=4, width="3cm") %>%
+    column_spec(column=5, width="1.7cm") %>%
+    column_spec(column=6, width="2.7cm") %>%
+    column_spec(column=7, width="1.7cm") %>%
+    kableExtra::group_rows("Shielding eligibility category", 9, 14, bold=FALSE) %>%
+   add_footnote(label=c("Percentages are column percentages for each variable",
+                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
+                         "Univariate rate ratios are for models with a single covariate",
+                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
+                 notation="none") %>%
+    kable_styling(latex_options=c("scale_down", "HOLD_position"))
+
+```

### 2021-08-12 casecontrol.R

@@ -192,27 +192,21 @@ if(!controls) { # cases only
     cc.all[, CASE := as.integer(is_case)]
     cc.all[, stratum := as.integer(stratum)]
 }
-cc.all[CASE==1, testpositive.case := diag_case==0 & cod_case==0]
 
 ## include cases ascertained through discharge diagnosis
-    ## case ascertainment uses three sources in order:
-    ##   test positives, discharge diagnoses, death certs
-    ## is.case is defined to assign case status
-    ## diag.case is a 0-1 variable encoding cases ascertained through discharge diagnosis
-    ## cod.case encodes ascertainment through death cert
-with(cc.all, table(is_case, cod_case)) # 137 have is_case==false and cod_case==1
-with(cc.all, table(is_case, diag_case)) # 74 have is.fase==false and diag_case==1 and is_case==0
-#with(cc.all[diag_case==1 & cod_case==0],
-#     table(icu + hdu > 0, !is.na(date_of_death))) # 4 cases ascertained through discharge diagnosis were not in critical care and have a death certificate with no mention of COVID.
-## Sharon & Jen have assigned these a dummy specimen date 7 days before date of admission
-## we reconstruct the date of admission as 7 days after the specimen date,
-## then code as severe if the date of death was within 28 days of admission.
+## case ascertainment uses three sources in order:
+## test positives, discharge diagnoses, death certs
+## is.case is defined to assign case status
+## diag_case is a 0-1 variable encoding cases ascertained through discharge diagnosis
+## these cases are assigned a dummy specimen date 7 days before date of admission
+## cod_case encodes ascertainment through death cert
+## testpositive.case is assigned by exclusion
+cc.all[CASE==1, testpositive.case := diag_case==0 & cod_case==0]
+with(cc.all[CASE==1], table(testpositive.case, cod_case))
+with(cc.all[CASE==1], table(testpositive.case, diag_case))
 
 cc.all[age_years < 0, age_years := 0]
 
-#cc.all[is.na(Prisoner_Flag), Prisoner_Flag := FALSE]
-## all cases not ascertained through diag or cod must be testpositive cases
-## 3052 cases not testpositive cases
[…]

### 2021-08-12 helperfunctions.R

@@ -391,6 +391,8 @@ lookup.names <- data.frame(varname=c("AGE", "sex",
                                      "Sgene.dropout",
                                      "av2channels",
                                      "inpat.recent",
+                                     "numdrugs.notcv",
+                                     "numicdchapters",
                                      "numdrugs"
                                 ),
                            longname=c("Age", "Males",
@@ -438,6 +440,8 @@ lookup.names <- data.frame(varname=c("AGE", "sex",
                                       "S gene dropout",
                                       "Mean Ct of ORF1ab and N genes",
                                       "Recent hospital stay",
+                                      "Number of non-cardiovascular drug classes",
+                                      "Number of hospital diagnoses", 
                                       "Number of drug classes"
                                       ))
 

### 2021-08-12 rrtimewin.R

@@ -1,4 +1,11 @@
+
+datadir <- "data/2021-07-28/"
+
+#load(paste0(datadir, "cc.all.RData"))
+
+
 firstdate <- as.Date("2020-03-01")
+
 ## for coeffs we have to specify explicitly the midpoint of the time window 
 winsize <- 42 
 startdates <- with(cc.all, firstdate:max(specimen_date) - winsize)
@@ -33,12 +40,13 @@ coeffs.timewindow[, linesize := `se(coef)`^-1.5]
 coeffs.timewindow[, linesize := 2 * linesize / max(linesize, na.rm=TRUE), by="effect"]
 setorder(coeffs.timewindow, effect)
 
-breakpoints <- c(0.2, 0.3, 0.5, 1, 2, 3, 5, 10, 20, 30, 50)
+## Figure: rate ratios by date of presentation
+breakpoints <- c(0.05, 0.1, 0.2, 0.5, 1, 2, 3, 5, 10, 20, 30)
 p.rateratio <-
     ggplot(data=coeffs.timewindow, aes(x=date.midpoint, y=coef, color=effect)) +
         geom_line(size=coeffs.timewindow$linesize) +
         labs(x=paste0("Presentation date: mid-point of ", winsize, "-day window"),
-             y="Rate ratio (log scale)") + 
+             y="Rate ratio for severe COVID-19 (log scale)") + 
         scale_y_continuous(breaks=log(breakpoints),
                            labels=breakpoints,
                            limits=log(c(min(breakpoints), max(breakpoints))),
@@ -74,7 +82,7 @@ rollmean.dtN <- function(dt, k) {
                       avdaily=zoo::rollmean(dt$N, k=k, fill=NA)))
 }
 
-## Figure 1: severe cases by date of presentation
+## Figure: severe cases by date of presentation
 winsize.casedates <- 3
 
 casedates.allgr <-  cc.all[CASE==1 & casegroup=="A", .N, by=specimen_date]
@@ -111,6 +119,3 @@ p.byelig <- ggplot(data=casedates.byelig,
          ylab("Daily severe cases")
 p.byelig
 
-png("dailysevere.png")
-p.byelig
-dev.off()

### 2021-08-12 shieldcohort.R

@@ -3,40 +3,83 @@ library(ggplot2)
 library(survival)
 
 datadir <- "data/2021-07-28/"
-
+lastdate <- as.Date("2021-07-14")
+interval.length <- 28
+    
 load(paste0(datadir, "shielded.full.RData"))
 
+shielded.full[, anycase := as.integer(!is.na(specimen_date))]
 min.time <- min(as.integer(shielded.full$entrydate), na.rm=TRUE)
 max.time <- max(as.integer(shielded.full$exitdate), na.rm=TRUE)
 
-interval.length <- 14
-
 shielded.tsplit <-
-			as.data.table(survSplit(formula=Surv(time=as.integer(entrydate),
-					  time2=as.integer(exitdate),	
-                			       	   event=severecase) ~ age_years + sex + shield.group,
-			     	    data=shielded.full,
-						   		cut=seq(min.time, max.time, by=interval.length)))	
-
+    as.data.table(survSplit(formula=Surv(time=as.integer(entrydate),
+                                         time2=as.integer(exitdate),	
+                                         event=anycase) ~ age_years + sex + shield.group + casegr + casegr2,
+                            data=shielded.full,
+                            cut=seq(min.time, max.time, by=interval.length)))	
+nrow(shielded.tsplit)
 
 shielded.tsplit[, tobs := tstop - tstart]
-		  
-severe.poissonmodel <- glm(formula = severecase ~
-                                offset(log(tobs)) + splines::ns(tstart, 4) +
+shielded.tsplit[, severe := as.integer(anycase==1 & casegr2=="Severe")]
+
+## any case ##################################
+anycase.poissonmodel <- glm(formula = anycase ~
+                                offset(log(tobs)) + splines::ns(tstart, 5) +
                                 age_years + sex + shield.group,
                            family = poisson,
                            data = shielded.tsplit)
-
+anycase.summary <- summary(anycase.poissonmodel)
 ## predict on scale of intensity of Poisson arrivals per interval
+shielded.tsplit[, anycase.lambda := as.numeric(predict(anycase.poissonmodel,
+                                                       newdata=shielded.tsplit,
+                                                      type="response"))]
+
+## severe case #################################################
+severe.poissonmodel <- glm(formula = severe ~
+                                offset(log(tobs)) + splines::ns(tstart, 5) +
+                                age_years + sex + shield.group,
[…]

### 2021-08-12 vaxshieldupdate.R

@@ -1,0 +1,417 @@
+library(data.table)
+library(ggplot2)
+
+source("helperfunctions.R")
+
+rollmean.dtN <- function(dt, k) {
+    ## FIXME: add a by= argument
+    N.dates <- dt[, .N, by=SPECIMENDATE]
+    setkey(N.dates, SPECIMENDATE)
+    all.dates <- data.table(date=seq(from=min(N.dates$SPECIMENDATE),
+                                     to=max(N.dates$SPECIMENDATE),
+                                     by=1))
+    setkey(all.dates, date)
+
+    ## add rows for missing dates by a left join
+    all.dates <- N.dates[all.dates]
+    all.dates[is.na(N), N := 0]
+    return(data.table(date=dt$SPECIMENDATE,
+                      avdaily=zoo::rollmean(dt$N, k=k, fill=NA)))
+}
+
+rollsum.datewin <- function(dt, k, datevar) {
+    ## returns a table of rolling sums of width k centred on date
+    ## dt is a dataset of totals (N) by date, which may have no rows for some dates in the
+    ## range of interest
+    ## add rows for missing dates by a left join of all.dates with dt
+    all.dates <- data.table(date=seq(from=min(dt[[datevar]]),
+                                     to=max(dt[[datevar]]),
+                                     by=1))
+    setkey(all.dates, date)
+    # setkeyv(dt, datevar) can't set physical key here
+    all.dates <- dt[all.dates]
+
+    return(data.table(date=all.dates[[datevar]],
+                      rsum=zoo::rollsum(all.dates[, N], k=k, fill=0, align="center")))
+}
+
+format.estcipv <- function(x.ci, x.pval) {
+    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
+    x <- gsub(", ", " to ", x)
+    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
+    x <- gsub("times", "\\\\times", x)
+    return(x)
+}
+
+format.estci <- function(x.ci) {
+    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
+    x <- gsub(", ", " to ", x)
+    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
+    #x <- gsub("times", "\\\\times", x)
+    return(x)
+}
+
+lastdate.cases <- as.Date("2021-07-12")
+
[…]

### 2021-08-13 vaxshieldupdate.Rmd
now loads cc.all

@@
 library(kableExtra)
-
-source("vaxshieldupdate.R")
 
@@
 Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date.
-
+<!---
 # Cohort analysis of the shielding list
@@
 Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.
+--->
 
 # Results
-## Time course of the epidemic and associations with risk factors 
+## Time course of vaccine efficacy against severe disease 
 Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.
-From mid-January to early March 2021 there was a rapid decline in the incidence of severe cases, followed by a slight increase from early June to early July.
+In the first half of July the average daily number of severe cases among the 200,000 people still on the shielded list was about 4, equating to an incidence rate of about 0.6 per 1000 per month.
 
-Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category, vaccination dose (coded as 0, 1, 2 doses), and recent (5 to 14 days before presentation date) hospital exposure as inpatient.
-These rate ratios are computed from a model that includes all variables shown, together with care home residence.
-
-The rate ratio for severe disease associated with shielding eligibility did not decline as the winter peak declined, and increased from May 2020 onwards.
-The efficacy of vaccination against severe disease (modelled as the rate ratio per 2 doses) remained high, despite the rapid replacement of the Alpha by the Delta variant during May 2020.
+Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category and vaccination status, in a model that includes both these variables together with care home residence and recent hospital stay as covariates.
+This shows that the rate ratio for severe disease associated with shielding eligibility, adjusted for vaccination status, has increased since 1 May, even though the absolute rate of severe disease in this group remains much lower than in December-January.
+The rate ratios for severe disease associated with 1 dose and 2 doses of vaccine at least 14 days earlier (with unvaccinated as reference category) declined  steadily from January to June, implying that the efficacy increased despite the rapid replacement of the Alpha variant by the Delta variant during May 2020 [@sheikh_sarscov2_2021].
 
@@
 ## Statement of principal findings
-
 1.
@@
 
-knitr::kable(table.shieldgr[14:21, ], 
+table.shieldgr[, effect := gsub("vax", "vaccine", effect)] 
+knitr::kable(table.shieldgr[14:19, ], 
              escape=FALSE, 
@@
 ```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window.
-For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence.
+For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence and recent hospital stay.
 Line thickness is inversely proportional to standard error"}
@@
 
-knitr::kable(table.hosp.shieldgr[13:20, ], 
+table.hosp.shieldgr[, effect := gsub("vax", "vaccine", effect)] 
+knitr::kable(table.hosp.shieldgr[14:19, ], 
              escape=FALSE, 

### 2021-08-13 rrtimewin.R

@@ -3,14 +3,13 @@ datadir <- "data/2021-07-28/"
 
 #load(paste0(datadir, "cc.all.RData"))
 
-
 firstdate <- as.Date("2020-03-01")
+lastdate <- as.Date("2021-07-14")
 
 ## for coeffs we have to specify explicitly the midpoint of the time window 
 winsize <- 42 
 startdates <- with(cc.all, firstdate:max(specimen_date) - winsize)
 enddates <- startdates + winsize
-lastdate <- as.Date("2021-07-14")
 
 coeffs.timewindow <- NULL
 for(timewin in 1:length(startdates)) {
@@ -20,7 +19,7 @@ for(timewin in 1:length(startdates)) {
                        .(CASE, care.home, listedgr3, vax14.dose, vax14.factor, occup, inpat.recent, stratum)]
     if(length(with(tdata, table(CASE, occup))) > 2) {
         date.midpoint <- startdates[timewin] + floor(winsize / 2)
-        coeffs <- tryCatch(summary(clogit(formula=CASE ~ care.home + listedgr3 + inpat.recent + vax14.dose + strata(stratum), data=tdata))$coefficients,
+        coeffs <- tryCatch(summary(clogit(formula=CASE ~ care.home + listedgr3 + inpat.recent + vax14.factor + strata(stratum), data=tdata))$coefficients,
                            error=function(cond) return(NULL)
                            )
         if(!is.null(coeffs)) {
@@ -32,36 +31,39 @@ for(timewin in 1:length(startdates)) {
 }
 coeffs.timewindow[, date.midpoint := as.Date(date.midpoint, origin="1970-01-01")]
 coeffs.timewindow <- coeffs.timewindow[grep("care.home", effect, invert=TRUE)]
+coeffs.timewindow <- coeffs.timewindow[grep("inpat.recent", effect, invert=TRUE)]
 coeffs.timewindow[, effect := gsub("^care.home", "", effect)]
 coeffs.timewindow[, effect := gsub("^listedgr3", "", effect)]
 coeffs.timewindow[, effect := gsub("TRUE$", "", effect)]
+coeffs.timewindow[, effect := gsub("vax14\\.factor", "Vaccine dose ", effect)]
 coeffs.timewindow[, effect := replace.names(effect)]
 coeffs.timewindow[, linesize := `se(coef)`^-1.5]
[…]

### 2021-08-13 shieldcohort.R

@@ -3,7 +3,7 @@ library(ggplot2)
 library(survival)
 
 datadir <- "data/2021-07-28/"
-lastdate <- as.Date("2021-07-14")
+lastdate <- as.Date("2021-07-25")
 interval.length <- 28
     
 load(paste0(datadir, "shielded.full.RData"))
@@ -55,31 +55,12 @@ shielded.tsplit[, date := as.Date(tstart, origin=as.Date("1970-01-01"))]
 shield.anycase  <- shielded.tsplit[, .(riskyear=mean(anycase.prob, na.rm=TRUE)), by=date]
 shield.severe <- shielded.tsplit[, .(riskyear=mean(severe.prob, na.rm=TRUE)), by=date]
 
-save(anycase.summary, severe.summary, shield.anycase, shield.severe,
-     file=paste0(datadir, "shieldcohort.models.RData"))
 rm(shielded.tsplit)
 gc()
 
-########### plots ##################################
-p.anycase <- ggplot(data=shield.anycase, aes(x=date, y=riskyear)) +
-    geom_line() +
-        labs(x=paste0("Presentation date: start of ", interval.length, "-day interval"),
-             y="Rate per year") + 
-    scale_x_date(breaks = seq.Date(from = as.Date("2020-03-01"),
-                                   to = lastdate, by = "month"),
-                 expand=c(0, 10), 
-                 labels=gsub("^0", "", 
-                             format.Date(seq.Date(from = as.Date("2020-03-01"),
-                                                  to = lastdate, by = "month"),
-                                         "%d %b")
-                             ),
-                 limits=c(as.Date("2020-12-01"), lastdate)) 
-p.anycase
+shield.all <- rbind(shield.anycase[, casegr := "All cases"],
+                    shield.severe[, casegr := "Severe cases"])
 
-p.severe <- ggplot(data=shield.severe, aes(x=date, y=riskyear)) +
-    geom_line()
-p.severe
+save(anycase.summary, severe.summary, shield.all,
+     file=paste0(datadir, "shieldcohort.models.RData"))
 
-png("shieldedrisk.png")
-gridExtra::grid.arrange(p.anycase, p.severe, nrow=2)
-dev.off()

### 2021-08-13 shieldlist.R

@@ -1,4 +1,12 @@
 
+batches <- as.data.table(read.table("data/BatchReference.csv", sep="\t", header=TRUE))
+setnames(batches, "Batch", "shield.batch")
+batches$shield.batch <- as.integer(gsub("Batch", "", batches$shield.batch))
+setkey(batches, shield.batch)
+batches$Date.Sent <- gsub("Friday ", "", batches$Date.Sent)
+batches$Date.Sent <- gsub("^([[:digit:]]+)([stndrdth]{2} )", "\\1 ", batches$Date.Sent)
+batches$Date.Sent <- as.Date(batches$Date.Sent, "%d %B %Y")
+
 datadir <- "./data/2021-07-28/"
 shielding.full.filename <- paste0(datadir,
                                   "CC_shielding_patients_anon_2021-07-28.csv")
@@ -37,6 +45,7 @@ shielded.full <- shielded.full[!(shield.group == "Pregnant with heart disease" &
 
 save(shielded.full, file=paste0(datadir, "shielded.full.RData")) # this will be used for a left join of cc.all with shielding cohort
 
+load(paste0(datadir, "cc.all.RData"))
 ## import case status into shielded.full
 cases <- cc.all[CASE==1, .(anon_id, specimen_date, casegr, casegr2, casegr3)]
 setkey(cases, anon_id)
@@ -69,6 +78,7 @@ shielded.full[is.na(exitdate), exitdate := max(shielded.full$specimen_date, na.r
 shielded.full[, entrydate := as.Date("2020-03-01")]
 shielded.full[entrydate >= exitdate, entrydate := NA]
 
+save(shielded.full, file=paste0(datadir, "shielded.full.RData"))
 rm(shielded.full)
 
 ####### rheumatology ######################################

### 2021-08-13 vaxshieldupdate.R

@@ -1,4 +1,5 @@
 library(data.table)
+library(survival)
 library(ggplot2)
 
 source("helperfunctions.R")
@@ -51,6 +52,14 @@ format.estci <- function(x.ci) {
     return(x)
 }
 
+datadir <- "data/2021-07-28/"
+if(!exists("cc.all")) {
+    load(paste0(datadir, "cc.all.RData"))
+}
+
+source("rrtimewin.R")
+
+interval.length <- 28  # for poisson regression of shielding cohort
 lastdate.cases <- as.Date("2021-07-12")
 
 ## tabulate case groups by vax status in shielding cohort
@@ -62,9 +71,10 @@ paste.rowpercent(with(cc.all[CASE==1 & specimen_date >= as.Date("2020-12-01") &
                              vax14.factor=="2"], 
                       table(shield.group, casegroup)))
 
-
+if(FALSE) {
 ## tabulate severe cases after 1 April
-severe.byagevax <- with(cc.all[CASE==1 & casegroup=="A" &
+ 
+    severe.byagevax <- with(cc.all[CASE==1 & casegroup=="A" &
                              specimen_date >= as.Date("2021-04-01") & specimen_date <= lastdate.cases],
                       table(agegr20plus5, vax14.factor))
 severe.byagevax.rowsums <- rowSums(severe.byagevax)
@@ -145,7 +155,7 @@ bnfchnum.table <- bnfchnum.table[total >= 20, .(bnfchnum, descr, Controls, Cases
 table.numdrugs <- cbind(summary(casesnorisk[CASE==0]$numdrugs),
                         summary(casesnorisk[CASE==1]$numdrugs))
 colnames(table.numdrugs) <- c("Controls", "Cases")
-
+}
 ########################################################
 ## tabulate freqs of risk factors including shielding conditions in cases and controls
 
@@ -415,3 +425,33 @@ model.hosp.shieldonly.vax2 <-
                                            vax14.dose==1 & 
                                specimen_date >= as.Date("2020-12-01")],
                    formula=CASE ~ shield.any + strata(stratum)))
+##################################################
+
+
+########### plots ##################################
+
+load(paste0(datadir, "shieldcohort.models.RData"))
+
+shield.all[, rateper1000 := 1000 * riskyear]
+
+p.anycase <- ggplot(data=shield.all, aes(x=date, y=rateper1000, color=casegr)) +
+    geom_line() +
+    labs(x=paste0("Presentation date: start of ", interval.length, "-day interval"),
[…]

### 2021-08-29 imovecovid.Rmd
reinfections

---
title: "Risk factors for severe COVID-19 in the Scottish population: the REACT-SCOT Study"
author: Paul McKeigue 
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output: 
  beamer_presentation:
    slide_level: 1
    includes:
      in_header: ../beamer_header.tex
    latex_engine: lualatex
    keep_tex: yes
---

# REACT-SCOT Study

* Case-control study of all diagnosed cases of COVID-19 in Scotland since March 2020

  - Cases include those without a positive test, ascertained through hospital discharge diagnosis or death certificates.

  - Ten controls per case, not previously diagnosed and matched for age, sex and general practice

  - linkage to household composition, medical records including hospital admissions, prescribing in primary care

* Main outcome variable: **severe COVID-19** defined as entry to critical care within 21 days of presentation date, or fatal outcome

  - definition ensures that case ascertainment is not affected by triage of those assessed as unlikely to benefit from critical care

# Studies based on REACT-SCOT

* Co-morbidity and construction of a risk score ("COVID age")

* Pharmacoepidemiology: medications compromising COVID

* Effectiveness of shielding clinically extremely vulnerable individuals

* Hospital transmission

* Other studies published, under review or in progress: 
  - associations with occupation: patient-facing health care workers, teachers (inverse association), 
  - (inverse) association with living with children
  - vaccine efficacy in clinically vulnerable individuals

# Laporte & Healy (April 2020): "Medications compromising COVID"

Drug classes associated with community-acquired pneumonia

* proton pump inhibitors

* hypnotics, sedatives

* antipsychotics, drugs used in nausea and vertigo

* opioid analgesics

* gabapentinoids

* non-steroidal anti-inflammatory drugs.

* anticholinergic effects likely to increase risk of pneumonia:

  - H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used for parkinsonism, and antiepileptic drugs

Predictions of associations with severe COVID-19 were confirmed in REACT-SCOT.

# Conditions designated as eligible for shielding (~4% of population)

1. Solid organ transplant recipient

2. Cancer of blood or bone marrow, or cancer treatments that affect immune system

3. Severe respiratory conditions: cystic fibrosis, asthma, chronic obstructive airway disease, bronchiectasis, pulmonary hypertension

4. Rare diseases that increase risk of infections: severe combined immunodeficiency, sickle-cell disease.

5. Immunosuppression therapies sufficient to increase risk of infection

7. Other conditions including dialysis, splenectomy

First batch of shielding letters sent on 4 April 2020 advised strict isolation, offered support with home delivery of food and medication.

# Severe cases by risk category and date of presentation: ribbon plot
```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.5,fig.asp=0.6, fig.cap="Care home residents excluded"}

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

theme_set(theme_gray(base_size = 10))

p.byelig + theme(legend.position = c(0.4, 0.7)) + theme(legend.text=element_text(size=8))

```

# Rate ratios for severe COVID-19 associated with shielding and moderate risk condition 
```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.8,fig.asp=0.58, fig.cap="Care home residents excluded. Line thickness is proportional to precision of estimate. Arrows indicate dates that shielding advice letters were sent."}

theme_set(theme_gray(base_size = 10))

p.rateratio + theme(plot.title = element_blank())  + 
labs(caption = element_blank(), tag = element_blank()) +   
theme(legend.position = c(0.48, 0.87)) + theme(legend.text=element_text(size=8)) + 
theme(legend.title = element_blank())

```

# Risk factors for severe COVID-19 in those eligible for shielding 
\scriptsize
\setbeamertemplate{caption}{\raggedright\insertcaption\par}

```{r riskfactorsshieldedonly, echo=FALSE, warning=FALSE, message=FALSE}

library(knitr)
library(kableExtra)

options(knitr.table.format = "latex") # adds tab: prefix to labels

knitr::kable(table.shieldedonly.severecases.nocare[7:15, c(1:2, 5:6)], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxfatal",
             row.names=TRUE,
             align=c("r", "r", "l", "l"), 
   col.names=c(clean.header(colnames(table.shieldedonly.severecases.nocare)[1:2]),
                "Rate ratio (95\\% CI)", "p-value")) %>%
    kableExtra::group_rows("School-age children in household", 2, 4, bold=FALSE) %>%
    kableExtra::group_rows("Adults in household", 5, 7, bold=FALSE) %>%
    column_spec(column=1, width="2.3cm") %>%
    column_spec(column=2, width="1.7cm") %>%
    column_spec(column=3, width="1.5cm") %>%
    column_spec(column=4, width="2.2cm") %>%
    column_spec(column=5, width="1.3cm")

```

# Prevalence of recent hospital exposure by risk category

```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25, fig.asp=0.7, fig.cap="Prevalence in controls not resident in care homes"}
p.hosp + theme(plot.title = element_blank()) + 
labs(caption = element_blank(), tag = element_blank()) + 
theme(legend.position = c(0.4, 0.82))

```

# Fraction of severe COVID-19 attributable to recent hospital exposure

```{r plotfrac, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25, fig.asp=0.7, fig.cap="Care home residents excluded"}
p.fracattr + theme(plot.title = element_blank())+ 
labs(caption = element_blank(), tag = element_blank())

```

# Conclusions

* Protection of vulnerable individuals was limited by their inability to avoid transmission in hospital and from other adults in the household.

  - Efficacy of single dose vaccine ~ 70% in shielded but a larger study is required to estimate efficacy in solid organ transplant recipients.

* Hospital-acquired infection accounted for more than half of severe cases in early December 2020.

* Role of nosocomial transmission was recognized early in the epidemic:

  - Wang (2020 Feb) JAMA: 41% of hospitalized cases in Wuhan in January 2020
  - Boccia (2020 Apr) JAMA Int Med: "largely a nosocomial infection" in Lombardy

* Lessons to be learned: physically separate hospitals, surveillance, epidemic modelling.

### 2021-08-29 reinfections.Rmd
reinfections

---
title: "Risk factors for reinfection with SARS-CoV-2 in Scotland"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
    corresponding: yes
    email: paul.mckeigue@ed.ac.uk
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: for the PHS COVID-19 Epidemiology and Research Cell
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland.
- code: IGMM
  address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. 
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ.
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
header-includes:
  \usepackage{newunicodechar}
  \usepackage[T1]{fontenc}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
#csl: ./f1000.csl 
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

```

# Abstract

**Objectives** - To estimate the risk of reinfection and risk factors for reinfection in people who had a positive nucleic acid test for SARS-CoV-2.

**Design** - Cohort study

**Setting** - Population of Scotland from June 2020 to 28 July 2021.

**Main outcome measure** - Reinfection with COVID-19, defined as a positive nucleic acid test at least 90 days after first positive test.

**Results** - There were `r nrow(reinfections)` reinfections over an average of `r round(atleast90day.firstpos[, mean(tobs)] * 12 / 365.25, 1)` months of follow-up.

**Conclusions**  
Reinfection rates remain low and the proportion of reinfections that lead to hospitalisation or death is low.  The low efficacy of vaccination against reinfection weakens the rationale for mandatory vaccination in occupational settings of people who have evidence of previous infection with SARS-CoV-2.

# Introduction

# Methods
A cohort of all individuals in Scotland at risk of reinfection was formed from the `r nrow(atleast90day.firstpos)` individuals in Scotland who survived at least 90 days from the date they first tested positive, based on CDC criteria for reinfection [@centers_for_disease_control_and_prevention_common_2020].  These individuals had already been linked to electronic health records for the REACT-SCOT case-control study as described elsewhere [@mckeigue_rapid_2020;@mckeigue_relation_2021b].

Entry date was 90 days after first testing positive, exit date was the earliest of: (1) date of testing positive (reinfection date), (2) date of death, (3) end of the follow-up period (28 July 2021). 
The follow-up time was split by calendar time into 7-day person-time intervals.  With this time-split dataset, a Cox regression model was used to estimate rate ratios. Baseline covariates included were those that we have previously reported to be associated with severe COVID-19 in this population [@mckeigue_rapid_2020;@mckeigue_relation_2021b]: age, sex, care home residence, occuptation, number of adults in household, number of children in household, and risk category as a factor with three levels: no risk condition, moderate risk condition, eligible for shielding   Vaccination status was assigned as the number of doses administered at least 14 days before presentation date. Vaccination status at the start of each person-time interval and months since first infection were encoded as time-updated covariates.

Cox regression estimates only rate ratios. To model the baseline hazard rate as a function of calendar time, a Poisson regression model was fitted with the covariates that were associated with reinfection in the Cox model, plus calendar time as a natural spline function with 6 degrees of freedom.

# Results
Table \ref{tab:fatal} shows the number of reinfections by age group and by fatality within 28 days of testing positive. Of the 24 reinfections that were fatal within 28 days of a positive test, only 9 were coded with COVID-19 as underlying cause and 8 of these were in people aged 75 years and over.

Table \ref{tab:reinf} shows the frequencies of risk factors over all person-time intervals in the cohort, together with the coefficients from a Cox-regression fitted to this dataset. Most of the variables that we have previously reported to be associated with severe COVID-19 in this population -- age, male sex, care home residence, number of adults in household, risk conditions -- were not associated with reinfection.  Reinfection rates were higher in women than in men, and higher in teacher and patient-facing health care workers than in other occupations. With unvaccinated as reference category, the  rate ratios associated with vaccination were 1.05 and 0.81 for one dose and two doses respectively. WThe effect of time since first infection, estimated within strata of calendar time, was aWithin each stratum of person-time interval each The rate

Figure \ref{fig:bycalendartime} shows the fitted incidence rate per month from a Poisson regression model fitted to the dataset, with the same covariates as the Cox regression model together with spline terms for the effect of calendar time.  In unvaccinated individuals, the rate of reinfection follows the course of the epidemic, with no obvious effect of the replacement of the original strain by the Alpha strain in December 2020.  The model predicts  reinfection rates to be higher in the vaccinated than in the unvaccinated in early 2021 because at this time the vaccinated group contained a high proportion of health care workers.

# Discussion
## Statement of principal findings

* The incidence of reinfection so far is about 3 per 1000 per month. The very few cases with fatal outcome are almost all in those aged over 75 years and COVID-19 is not the underlying cause in most of these deaths.

* Most of the risk factors for severe first infection with COVID-19 are not risk factors for reinfection: the strongest associations with reinfection identified in this study are female sex and occupation as a teacher or patient-facing health-care worker:

* The protection against reinfection conferred by natural infection appears to wane at about 3% per month

* The efficacy of two doses of vaccine against reinfection is only about 23%.

## Strengths and limitations
Strengths are the establishment of a cohort based on all test-positive cases arising in the population since the beginning of the epidemic followed till July 2021, the comprehensive linkage to electronic health records and demographic factors including relevant occupations, and a formal cohort analysis based on splitting into short person-time intervals.  This  eliminates confounding by calendar time, but still allows the effect of vaccination and time since first infection to be modelled as time-updated covariates.

One limitation is that ascertainment of both first infections and reinfections may be biased by with respect to factors such as occupation where frequent testing is mandated.

## Comparison with other studies
The results of this study, showing an odds ratio of 0.77 for reinfection in those with 2 doses of vaccine versus none, differ from the results of a case-control study in Kentucky [@cavanaugh_reduced_2021].  In that study the odds ratio for reinfection in fully vaccinated versus unvaccinated individuals was 0.43, equivalent to 57% efficacy. The explanation for this  difference is not clear. The Kentucky dataset was restricted to first infections in 2020, and follow-up was restricted to May-June 2021.  In contrast our analysis included the entire follow-up period, stratified by calendar time, with vaccination status as a time-updated covariate.

An Israeli study [@gazit_comparing_2021] reported an odds ratio of 0.53 in previously infected and vaccinated versus previously infected and unvaccinated individuals.  However the temporal sequence of infection and vaccination was not given.

## Policy implications

, -- abo The low efficacy of vaccination against reinfection weakens the rationale for mandatory vaccination in occupational settings of people who have evidence of previous infection with SARS-CoV-2.

# Declarations
## Ethics approval
This study was performed within Public Health Scotland as part of its statutory duty to monitor and investigate public health problems.  Under the [UK Policy Framework for Health and Social Care Research](https://www.hra.nhs.uk/planning-and-improving-research/policies-standards-legislation/uk-policy-framework-health-social-care-research/) set out by the NHS Health Research Authority, this does not fall within the definition of research and ethical review is not required.  Individual consent is not required for Public Health Scotland staff to process personal data to perform specific tasks in the public interest that fall within its statutory role.  The statutory basis for this is set out in Public Health Scotland's [privacy notice](https://www.publichealthscotland.scot/our-privacy-notice/personal-data-processing/).

# References 
<div id="refs"></div>

# Figures
```{r bycalendartime, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:bycalendartime} Fitted values for incidence of reinfection, based on Poisson regression model with spline terms for calendar time and other covariates as in Table 2.  Vaccination status is defined by number of doses received at least 14 days before."}

theme_set(theme_gray(base_size = 10))

p.reinf

```

```{r fatal, echo=FALSE, message=FALSE}

knitr::kable(table.fatal, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="fatal",
             row.names=TRUE, 
			 #align=c('r', 'r'), 
             caption="Reinfections by death within 28 days of positive test and underlying cause of death") %>%
    column_spec(1, width="3.5cm") %>%
    column_spec(2, width="3.5cm") %>%
    column_spec(3, width="3.5cm") %>%
    column_spec(4, width="3.5cm") %>%
   kable_styling(latex_options=c("HOLD_position"))

```

```{r reinf, echo=FALSE, message=FALSE}

knitr::kable(table.reinf.coeffs, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="reinf",
             row.names=FALSE, 
             col.names=c(colnames(table.reinf.coeffs)[1:3],
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
			 #align=c('r', 'r'), 
             caption="Rate ratios for reinfection in cohort of all individuals who survived at least 90 days after first positive test") %>%
    add_footnote(label=c("Vaccination status defined as number of doses received at least 14 days before presentation date.",
                         "Multivariable model includes all covariates in the table"), 
                 notation="none") %>%
    group_rows("Occupation", 4, 7, bold=FALSE) %>%
    group_rows("Vaccination status", 14, 16, bold=FALSE) %>%
    column_spec(1, width="5cm") %>%
    column_spec(2, width="2.5cm") %>%
    column_spec(3, width="2.5cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2cm") %>%
   kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

### 2021-08-29 rheumatol.Rmd
reinfections

---
title: "Concise Report: COVID-19 in patients on biological and targeted synthetic disease-modifying antirheumatic drugs"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
    corresponding: yes
    email: paul.mckeigue@ed.ac.uk
  - name: Duncan Porter
    affiliation: Gart
  - name: Rosemary Hollick
    affiliation: Aberdeen
  - name: Stuart Ralston 
    affiliation: IGMM
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: PHS COVID-19 Epidemiology and Research Cell
    affiliation: HPS
address: 
  - code: Usher
    address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG  
  - code: Gart
    address: Gartnavel General Hospital, 1053 Gt Western Rd, Glasgow G12 0YN
  - code: Aberdeen
    address: University of Aberdeen, Aberdeen AB25 2ZN
  - code: Glasgow
    address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ 
  - code: IGMM
    address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland.
  - code: HPS
    address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
header-includes:
  \usepackage{newunicodechar}
  \usepackage[T1]{fontenc}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./f1000.csl # ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

```

Main text 1497 words excluding Declarations

\newpage

# Abstract

**Objectives** - To investigate the association of hospitalized COVID-19 to treatment with disease-modifying antirheumatic drugs (DMARDs).

**Methods**  - A list of rheumatology clinic patients on biologic DMARDs or janus kinase (JAK) inhibitors in March 2020 was linked to the REACT-SCOT case-control study which includes all cases of COVID-19 diagnosed in Scotland since the start of the epidemic.

**Main outcome measure** - Hospitalised or fatal COVID-19

**Results** - By July 2021 201 of the 4633 patients on the original list had been diagnosed with COVID-19, of whom 42 had been hospitalised and the outcome was fatal in 5. The adjusted rate ratios for hospitalised COVID-19 associated with conventional synthetic DMARDs ranged from `r astext.or.ci(table.rheumatol.drugs["methotrexate", 5])` for methotrexate to `r astext.or.ci(table.rheumatol.drugs["prednisolone", 5])` for prednisolone.  B cell depletion (rate ratio `r astext.or.ci(table.rheumatol.drugs["B cell depletion", 5])`) and janus kinase inhibitors (rate ratio `r astext.or.ci(table.rheumatol.drugs["JAK inhibitors", 5])`) were associated with increased risk of hospitalisation but tumour necrosis factor(TNF) inhibitors were not (rate ratio `r astext.or.ci(table.rheumatol.drugs["TNF inhibitors", 5])`.

**Conclusion** - On the basis of these results, patients on TNF inhibitors can be reassured that they are not at increased risk of severe COVID-19 in comparison with the general population.

\newpage

# Introduction
Early in the COVID-19 epidemic in the UK, people on "immunosuppression therapies sufficient to increase risk of infection" were designated by public health agencies as clinically extremely vulnerable and thus eligible for shielding.  Because biologic drugs and targeted synthetic immunosuppressants are prescribed only in hospital and these prescriptions are not held in electronic form, Public Health Scotland (PHS) requested  clinicians in relevant specialties to provide lists of patients on these drugs  [@health_protection_scotland_search_2020; @kanani_caring_2020].

The objective of this study was to investigate the association of hospitalized or fatal COVID-19 with disease-modifying antirheumatic drugs (DMARDs)

# Methods
Rheumatologists in three centres covering about one-third of the Scottish population, including the health boards of Lothian, Grampian, Highland, and part of Greater Glasgow and Clyde, sent lists of patients on biologic agents and targeted synthetic DMARDS to PHS in March 2020.  For this study the merged list, hereafter the "biologics list" was linked to the REACT-SCOT case-control study to take advantage of data linkages already established in this study, which now includes about 44% of the Scottish population and all incident cases of COVID-19 in Scotland diagnosed since the start of the epidemic.

The design of the REACT-SCOT study has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched for one-year age, sex and primary care practice and alive on the day of presentation of the case that they were matched to were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, a hospital discharge diagnosis or death with COVID-19 as underlying cause. The REACT-SCOT case-control dataset is linked to national data on vaccinations, hospital discharges and outpatient consultations in the last 5 years before presentation date, dispensed prescriptions written in primary care in the last 240 days, and the list of individuals eligible for shielding.  The dataset used for this study was based on cases presenting up to 28 July 2021 and their matched controls.  The main outcome measure cases of COVID-19 with hospitalisation within 28 days of a positive test or fatal outcome.

ICD-10 codes (as main condition or other condition) for autoimmune rheumatic diseases in hospital discharge records were grouped into three broad categories: rheumatoid arthritis (M05 to M09, M12.3, M13); axial spondyloarthropathies (M07, M45, M46); and connective tissue disorders (M30 to M35).  Rheumatology patients were defined in the case-control study as individuals with any of these discharge diagnoses or an outpatient consultation (face to face or remote) with specialty code AR in the last five years. From drug prescribing records, patients who had received conventional synthetic DMARDs -- methotrexate, hydroxychloroquine, sulfasalazine, leflunomide or prednisolone -- in the last 240 days were identified.

Rate ratios for hospitalised / fatal COVID-19 were estimated by conditional logistic regression. To investigate associations of COVID-19 with use of antirheumatic drugs, the analysis was undertaken: (1) with the entire case-control dataset, comparing rates of hospitalisation for COVID-19 between patients on the drug with individuals sampled from the population who were not on that drug; (2) with restriction of the dataset to rheumatology patients, thus comparing disease rates in rheumatology patients on the drug to disease rates in rheumatology patients not on that drug.

# Results
The biologics list comprised 4633 individuals: the diagnostic category was rheumatoid arthritis in 2702, axial spondyloarthropathies or other diagnoses in 1765, connective tissue disorder in 141, and other conditions in 25.  Of the 4633 individuals on the list, 201 had been diagnosed with COVID-19 by 28 July 2021.  Of these 201 cases, 42 were hospitalized, 5 entered critical care and 5 were fatal. Of the 4633, 2527 had been added by PHS to the shielding list baed on based on criteria laid down in guidance from the British Society of Rheumatology [@british_society_for_rheumatology_covid19_2020].  Of those added to the shielding list 30 (1.2%) were hospitalised with COVID-19, compared with 12 (0.5%) of those not added to the list.  The algorithm used by PHS to identify those eligible for shielding thus discriminated between low-risk and high-risk patients.

Of the 4633 on the biologics list, 1962 were sampled in the REACT-SCOT case-control study.  Of these 1962, all but 3 had a rheumatology outpatient consultation within the last 5 years, but only 550 had a hospital discharge diagnosis with a rheumatology code.

Table \ref{tab:diagnosis}, based on the full case-control dataset, shows that with those with no rheumatologic diagnosis as reference category, the rate ratio for hospitalized or fatal COVID-19 was elevated 2- to 3-fold in all three diagnostic categories.

```{r diagnosis, echo=FALSE, message=FALSE}

knitr::kable(table.rheumatol.diagnosis[4:7, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="diagnosis",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             col.names=c(clean.header(colnames(table.rheumatol.diagnosis)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Rate ratios for hospitalized or fatal COVID-19 by rheumatology diagnosis") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    add_footnote(label=c("Presentation dates up to 28 July 2021",
                         "This analysis includes all hospitalized or fatal COVID-19 cases and their matched controls, irrespective of whether they were on the biologics list", 
                         "Reference category is those without a rheumatology diagnosis: not on the biologics list and without a hospital discharge diagnosis of a rheumatic disease in last 5 years.", 
                         "Multivariable model includes all variables shown together with vaccination status, coded as number of doses received at least 14 days before presentation date."), 
                 notation="none") %>%
    column_spec(1, width="4.5cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="3cm") %>%
    column_spec(7, width="2cm") %>%
    kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

Table \ref{tab:drugs} shows the rate ratios for hospitalized or fatal COVID-19 associated with each drug class based on the full case-control dataset. In this analysis the reference category for each drug is all those in the population who are not taking the drug.  All conventional synthetic DMARDS prescribed in primary care were associated with modest rate ratios (around twofold) for hospitalisation for COVID-19.  The highest rate ratio was associated with prednisolone (rate ratio `r astext.or.ci(table.rheumatol.drugs["prednisolone", 5])`. Tumour necrosis factor (TNF) inhibitors were not associated with increased risk (rate ratio `r astext.or.ci(table.rheumatol.drugs["TNF inhibitors", 5])`, but B cell depletion and janus kinase (JAK) inhibitors were associated with elevated rate ratios (`r astext.or.ci(table.rheumatol.drugs["B cell depletion", 5])` and `r astext.or.ci(table.rheumatol.drugs["JAK inhibitors", 5])`) respectively.    For IL-17 and IL-6 inhibitors there was no evidence of elevated risk but the  confidence intervals were wide.

```{r drugs, echo=FALSE, message=FALSE}

knitr::kable(table.rheumatol.drugs[-(1:3), ],
             escape=FALSE, 
             booktabs=TRUE,
			 label="drugs",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             col.names=c(clean.header(colnames(table.rheumatol.drugs)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Rate ratios for hospitalized or fatal COVID-19 associated with use of disease-modifying anti-rheumatic drugs in the population") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    add_footnote(label=c("Presentation dates up to 28 July 2021.",
                        "This analysis includes all hospitalized or fatal COVID-19 cases and their matched controls", 
                         "Rate ratios are estimated by conditional logistic regression", 
                         "Multivariable model includes all variables shown together with vaccination status, coded as number of doses received at least 14 days before presentation date."), 
                 notation="none") %>%
    group_rows("Drugs prescribed in primary care", 1, 5) %>%
    group_rows("Drugs reported on the biologics list", 6, 10) %>%
     column_spec(1, width="3.5cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="3cm") %>%
    column_spec(7, width="2cm") %>%
    kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

Supplementary Table \ref{tab:diagdrugs} shows the same analysis based on restricting the dataset to those with a rheumatology outpatient consultation or discharge diagnosis. The confidence intervals are wider because cases without a matched control contribute no information. With the exception of prednisolone, conventional synthetic DMARDs were not associated with elevated rates of hospitalisation for COVID-19 in this analysis.  The increased risk associated with B cell depleting agents was still evident.

# Discussion
## Statement of principal findings
We have shown that:

* The risk of hospitalised or fatal COVID-19 is about two-fold higher in people with autoimmune rheumatic diseases. However the absolute risk of mortality from COVID-19 in patients on biologics has remained low: about 1 in 1000 over 16 months.

* In comparison with those treated with conventional synthetic DMARDs, patients treated with TNF inhibitors are at lower risk but patients treated with prednisolone, B cell depleting agents or JAK inhibitors are at higher risk.

## Strengths and limitations
Strengths of this study are the complete ascertainment of cases, the comprehensive linkage to electronic health records, and the incidence density sampling design that controls for general practice and calendar time.  One limitation is that the clinics that provided the biologics list cover only about one third of the Scottish population.  Although individuals prescribed biologics in other centres are misclassified as not on biologics, the effect on the rate ratio will be small because the proportion of the general population who are on biologics is low and because matching on general practice ensures that patients on biologics are compared with others from the catchment population of the clinics that provided the biologics list. In the unrestricted analysis, associations with glucocorticoids may be confounded by other conditions such as asthma that increase the risk of severe COVID-19. However we have shown that the key findings -- increased risk associated with glucocorticoids and B cell depleting agents but not with TNF inhibitors -- do not change when the dataset is restricted to those who have had rheumatology outpatient consultations.

## Comparison with other studies
Others have reported that hospitalisation for COVID-19 is associated with use of glucocorticoids and B cell depleting agents but not with TNF inhibitors. In a registry of 600 rheumatology patients diagnosed with COVID-19, the odds ratio for hospitalisation (with methotrexate use as reference category) was 2.1 for glucocorticoid use equivalent to >10mg/day predisone, and 0.4 for TNF inhibitors [@gianfrancesco_characteristics_2020].  In a later analysis based on 2869 patients with rheumatoid arthritis the odds ratios (with TNF inhibitors as reference category) were 4.5 for rituximab and 2.1 for JAK inhibitors [@sparks_associations_2021].  In a French series of 694 patients the odds ratio for fatal disease was 2.8 for any glucocorticoid, 0.2 for TNF inhibitors and 3.1 for  rituximab.  [@fai2r_/sfr/snfmi/sofremip/cri/imidiate_consortium_and_contributors_severity_2020].  In an Israeli cohort of 6112 psoriasis patients  the rate ratio for hospitalisation associated with TNF inhibitors, with methotrexate as reference category, was 0.08 [@kridin_tumor_2021].

As patients treated with TNF inhibitors generally have more severe rheumatic disease than those treated only with conventional synthetic DMARDs, the low risk of severe COVID-19 in TNF inhibitors users compared with conventional synthetic DMARD users suggests a possible protective effect.  As B cell depleters and JAK inhibitors are generally used as third-line therapy in patients who have not responded to TNF inhibitors, associations of these drug classes with severe COVID-19 may be attributable to pre-existing morbidity.

## Clinical implications
On the basis of these results, patients on TNF inhibitors can be reassured that they are not at increased risk of severe COVID-19 in comparison with the general population.  A larger study is required to investigate the association of severe COVID-19 with B cell depleting agents and JAK inhibitors.

# Declarations
## Public and Patient Involvement statement:
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives.

## Ethics approval and information governance
This study was performed within Public Health Scotland as part of its statutory duty to monitor and investigate public health problems.  Under the [UK Policy Framework for Health and Social Care Research](https://www.hra.nhs.uk/planning-and-improving-research/policies-standards-legislation/uk-policy-framework-health-social-care-research/) set out by the NHS Health Research Authority, this does not fall within the definition of research and ethical review is not required.  Individual consent is not required for Public Health Scotland staff to process personal data to perform specific tasks in the public interest that fall within its statutory role.  The statutory basis for this is set out in Public Health Scotland's [privacy notice](https://www.publichealthscotland.scot/our-privacy-notice/personal-data-processing/).  
A Data Protection Impact Assessment​ (DPIA) allows Public Health Scotland staff to link existing datasets.

## Transparency declaration
PM as the manuscript’s guarantor affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained. This manuscript has been generated directly from the source data by a reproducible research pipeline.

## Funding 
No specific funding was received for this study.

## Data Availability
The component datasets used here are available via the Public Benefits and Privacy Panel for Health and Social Care at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.

## Competing interest
All authors have completed and submitted the ICMJE Form for Disclosure of Potential Conflicts of Interest.

## Contributors
All authors provided substantial contributions to the conception of the study and the drafting of the manuscript. PM undertook the statistical analysis. All authors contributed to revising the manuscript critically for important intellectual content and approved the final manuscript.

## Acknowledgements
??

# References
<div id="refs"></div>

\beginsupplement

# Supplementary Table

```{r diagdrugs, echo=FALSE, message=FALSE}

knitr::kable(table.rheumatoldiag.drugs[-(1:3), ],
             escape=FALSE, 
             booktabs=TRUE,
			 label="diagdrugs",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             col.names=c(clean.header(colnames(table.rheumatoldiag.drugs)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Rate ratios for hospitalized or fatal COVID-19 associated with use of DMARDs, restricted to those with a rheumatology outpatient consultation or discharge diagnosis") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    add_footnote(label=c("Presentation dates up to 28 July 2021.",
                         "Restricted to individuals with a rheumatology outpatient consultation, hospital discharge diagnosis of an autoimmune rheumatic disease, or on the biologics list.",
                         "Rate ratios are estimated by conditional logistic regression", 
                         "Multivariable model includes all variables shown together with vaccination status, coded as number of doses received at least 14 days before presentation date."), 
                 notation="none") %>%
    group_rows("Drugs prescribed in primary care", 1, 5) %>%
    group_rows("Drugs reported on the biologics list", 6, 10) %>%
     column_spec(1, width="3.5cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="3cm") %>%
    column_spec(7, width="2cm") %>%
    kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

### 2021-08-29 transmission17Mar.Rmd
reinfections

--- 2021-03-06 transmission.Rmd
@@
 
-# rmarkdown::render("shielding.Rmd", output_file="shielding.pdf")
-
 dateformat <- function(x) { # format date and strip leading zero
@@
     x <- gsub("times", "\\\\times", x)
+    return(x)
+}
+
+format.estci <- function(x.ci) {
+    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
+    x <- gsub(", ", " to ", x)
+    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
+    #x <- gsub("times", "\\\\times", x)
     return(x)
@@
 
-Word counts: abstract 396 words, main text 5603 words.  
-2 figures, 4 tables, 25 references
+Word counts: abstract 350 words, main text 5771 words.  
+2 figures, 4 tables, 26 references
 
@@
 # Abstract  
-
-* _Objectives_ -  To investigate:(1) the risk of severe  COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.
-* _Design_ - Matched case-control study (REACT-SCOT). 
-* _Setting_ - Population of Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)`. 
-* _Participants_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice, linked with all `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland.
-* _Main outcome measure_ - Severe COVID-19, defined as cases that entered critical care or were  fatal.  
-* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])`
-in those with moderate risk conditions and 
-`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
+* _Background_ -  Clinically vulnerable individuals have been advised to shield themselves during the COVID-19 epidemic.
+The objectives of this study were to investigate: (1) the risk of severe COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.
+* _Methods_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.
+This dataset (REACT-SCOT) was  linked to the list of `r nrow(shielded.full)` individuals identified as eligible for shielding.
+Severe COVID-19 was defined as cases that entered critical care or were fatal.
+* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estci(table.listedgr3.severecases.nocare["Moderate risk condition", 3])` in those with moderate risk conditions and
+`r format.estci(table.listedgr3.severecases.nocare["Eligible for shielding", 3])` 
 in those eligible for shielding.
-The highest rate was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])`.
-In both the shielded and  the general population, the risk of severe COVID-19 increased with the number of adults  but decreased with the number of school-age children in the household.
-Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.
-In a case-crossover analysis with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
-Among those eligible for shielding, the population attributable risk fraction (PARF) of severe cases for recent exposure to hospital was  `r round(100 * pc * (r -1) / r)`%.
-In the general population the PARF for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.
+The highest rate was in solid organ transplant recipients: rate ratio `r format.estci(table.shielded.severecases.nocare["Solid organ transplant", 3])`.
+Risk of severe COVID-19 increased with the number of adults but decreased with the number of school-age children in the household.
+Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estci(table.shielded.severecases.nocare["Recent hospital visit/stay", 3])`  overall.
+To test for causality, a case-crossover analysis was undertaken; with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estci(table.tw["Recent.TW.only", 4])`.
+The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.
 * _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.
@@
 \linenumbers
-
-# Summary
-
-## What is already known on this topic?
-
+<!---
+# Author Summary
+## Why was this study done?  
 * Individuals designated as "clinically extremely vulnerable" -- about 3% of the population -- have been advised to shield themselves during the COVID-19 epidemic
@@
 
-## What this study adds
+## What did the researchers do and find?  
+* Despite the provision of advice and support, the risk of severe COVID-19 remained far higher in those eligible for shielding than in those without risk conditions.
 
-* Hospital-acquired infection has made a substantial contribution to the burden of COVID-19 in the population, accounting for more than half of severe cases in early December 2020.
+* In both those eligible for shielding and in the rest of the population, the risk of severe COVID-19 was higher in those who were sharing a household with other adults, and was far higher in the 5 to 14 days after exposure to hospital than in those without recent exposure to hospital.
 
+* In the second wave of the epidemic in December 2020, more than half of the cases of severe COVID-19 in the population were attributable to recent hospital exposure.
+
+## What do these findings mean?  
 * For vulnerable individuals to be shielded, the risk of hospital-acquired infection must be reduced and support must be provided for other adults in the household to co-isolate with the vulnerable individual.
@@
 
-<!--- 
-* don't blame HCW, there are problems 
-* intel failure
+* Limiting the burden of COVID-19 requires control of hospital-acquired infection. 
 --->
@@
 # Methods
-We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions".
+We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case-control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions".
 Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.
@@
 ## Ascertainment of eligibility for shielding 
-We used the dataset of those eligible for shielding that was created by PHS for the purposes of issuing letters.
-This dataset comprised a list of  patient identifiers and the shielding condition and the date the shielding letter was issued.
-The list had been generated  by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries,  for evidence of a list of agreed diseases and conditions  and supplemented by data submitted  to PHS following direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].
+The list of those eligible for shielding was generated by Public Health Scotland from March 2020 onwards by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries, for a designated list of diseases and conditions, supplemented direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].
 The categories designated as eligible for shielding were as listed below [@health_protection_scotland_search_2020; @kanani_caring_2020]:
@@
 3. Severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
-4. Rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
+4. Rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle-cell disease. 
 5. People on immunosuppression therapies sufficient to increase risk of infection
@@
 ## Transmission-related risk factors 
-Since addresses in the CHI database have been mapped to  Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.
+As addresses in the CHI database have been mapped to Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to augment the coding of care home residence in the CHI database.
 Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.
@@
 The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.
-Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated.
-<!---The ECDPC states that "cases with symptom onset within 14 days of discharge from a healthcare facility (e.g.
-re-admission) may be considered as community-associated, probable or definite health-care-associated-COVID-19, or to have an indeterminate association" but does not specify how cases are to be assigned to these categories.--->
+<!---Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated.
+--->
 
@@
 Ascertainment of this outcome does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is likely to differ between those eligible for shielding and those ineligible.
-Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.
+For the case-control analyses, cases and controls were censored at date of first vaccination and care home residents were excluded, as shielding advice intended for individuals in private households would not have been relevant to care home residents.
 
@@
 where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].
-Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.
+<!---Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.--->
 
@@
 This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases.
-It is a variant of the classic case-crossover design, which compares in cases only the frequencies of exposure in recent and less recent time windows.
+The classic case-crossover design compares in cases only the frequencies of exposure in recent and less recent time windows, and estimates the conditional odds ratio  as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only.
 This can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000].
-The conditional odds ratio can thus be estimated as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only.
-The analysis that we report is a refinement that takes advantage of the availability of a matched control group.
-We contrast not the numbers of cases with recent exposure only and nonrecent exposure only, but the conditional odds of case status between those with recent exposure only and those with less recent exposure only.
+The analysis that we report is a refinement that takes advantage of the availability of a matched control group to contrast the conditional odds of being a case between those with recent exposure only and those with less recent exposure only.
 This controls for any difference in the population frequencies of exposure or disease incidence between recent and less recent time windows.
@@
 From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
-As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital acquired infection using the ECPDC classification.
+As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.
 The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.
@@
 Such studies cannot assess the risks to immunosuppressed individuals in the population.
-Other reports have evaluated the theoretical possibilities for shielding but without empirical data on its effectiveness.
-A previous report [@jani_comparison_2020] based on the population of Glasgow did not report age-adjusted rate ratios.
 
@@
 
-Most previous reports have focused on the ECDPC definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].
-The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital acquired infection [@arhai_scotland_hospital_2021].
+The role of nosocomial transmission was recognized early in the epidemic.
+Hospital-acquired infection was suspected in 41% of hospitalised cases seen in January 2020 in one centre in Wuhan [@wang_clinical_2020-2].
+A commentary in April 2020 noted that "in Lombardy, SARS-CoV-2 became largely a nosocomial infection" [@boccia_what_2020].
+Most previous reports have focused on the narrow definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].
+The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].
 
@@
 For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke an immune response sufficient to be effective  [@chong_comprehensive_2017].
-The relation of severe disease in those eligible for shielding to number of adults in the household raises the question of whether priority should be given to vaccinating  household contacts of solid organ transplant recipients and other vulnerable individuals.
-The Joint Committee on Vaccination and Immunisation did not recommend this in a report published in December 2020, stating that "at this time there are no data on the size of the effect of COVID-19 vaccines on transmission." [@joint_committee_on_vaccination_and_immunisation_joint_2020].
-The Committee however recommended vaccination of  health and social care workers who are "at risk of transmitting infection to multiple vulnerable persons or other staff in a health care environment",  implying some expectation that vaccination would reduce transmission.
-On this basis, there is an equally strong case for offering vaccination to household contacts of solid organ transplant recipients without waiting for more definitive evidence of the effect of vaccines on transmission.
-Even with vaccination of  household contacts, this group will require shielding until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.
+As there is now evidence that vaccination reduces transmission to household contacts [@public_health_scotland_covid-19_2021], one way to reduce risk in this group would be to vaccinate their household contacts.
+Even with vaccination of  household contacts, solid organ transplant recipients will require shielding until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.
 
@@
 A report for NHS England by the Healthcare Safety Investigation Branch noted the challenges of controlling nosocomial transmission and recommended that a national strategy for infection prevention and control should be developed [@healthcare_safety_investigation_branch_investigation_2020].
-Vaccination of health care workers is likely to reduce nosocomial transmission, and vaccination of all inpatients should also be considered.
+Vaccination of health care workers is likely to reduce nosocomial transmission, but vaccination of those booked for elective procedures should also be considered.
 We have not attempted to quantify the onward transmission of COVID-19 from discharged patients, but the PHE modelling study estimated that this could have accounted for 5% of cases requiring hospitalisation in England during the first wave [@knight_contribution_2021].
 This suggests  more stringent testing before discharge and quarantine post discharge should be considered.
-More detailed understanding of how recommended IP policies are being operationalised is also needed [@oliver_could_2021].
+More detailed understanding of how recommended infection control policies are being operationalised is also needed [@oliver_could_2021].
 
 # Declarations
-## Public and Patient Involvement statement:
+<!---## Public and Patient Involvement statement:
 This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives.
+--->
 
@@
 
-## Transparency declaration
+<!---## Transparency declaration
 PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
 This manuscript has been generated directly from the source data by a reproducible research pipeline.
+--->
 
@@
 
-## Data sharing statement
+## Data Availability
 The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
 All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.
+
+## Competing interest
+All authors have completed and submitted the ICMJE Form for Disclosure of Potential Conflicts of Interest.
 
@@
 
+## Registration
+The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
+
 ## Acknowledgements
@@
 We thank Shona Cairns and Laura Imrie of the Antimicrobial Resistance and Healthcare Associated Infection unit of NHS National Services Scotland for comments on an earlier draft of this manuscript.
-
-## Competing interest
-All authors have completed and submitted the ICMJE Form for Disclosure of Potential Conflicts of Interest.
-
-## Registration
-The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
 

### 2021-08-29 transmissionBMED-D-21-00640.Rmd
reinfections

--- 2021-08-29 transmission17Mar.Rmd
@@
 
+table.testpos <- paste.rowpercent(with(cc.severe[CASE==1], 
+table(SPECIMENDATE > as.Date("2020-08-31"), testpositive.case)))
+
 ```
@@
 The objectives of this study were to investigate: (1) the risk of severe COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.
-* _Methods_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.
+* _Methods_ -  In a matched case-control design, all `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.
 This dataset (REACT-SCOT) was  linked to the list of `r nrow(shielded.full)` individuals identified as eligible for shielding.
 Severe COVID-19 was defined as cases that entered critical care or were fatal.
+Rate ratios were estimated by conditional logistic regression.
 * _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estci(table.listedgr3.severecases.nocare["Moderate risk condition", 3])` in those with moderate risk conditions and
@@
 Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estci(table.shielded.severecases.nocare["Recent hospital visit/stay", 3])`  overall.
-To test for causality, a case-crossover analysis was undertaken; with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estci(table.tw["Recent.TW.only", 4])`.
 The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.
 * _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.
-For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts.
 Mitigating the impact of the epidemic requires control of nosocomial transmission.
@@
 Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters advising strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
+Compliance was voluntary.
 
@@
 
+Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome.
+This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
+In the first wave (up end of August 2020), `r table.testpos[1, 2]` of the severe cases were test-positive; in the second wave beginning on 1 September 2020, `r table.testpos[2, 2]` were test-positive.
+
 Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
 Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.
-Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome.
-This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
 
@@
 Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.
-Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
-The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.
+In Scotland people with COVID-19 symptoms were instructed not to visit hospitals, to attend designated test centres for testing within 3 days, and to call an emergency number for admission if their symptoms worsened.
+It is therefore unlikely that hospital visits by people with COVID-19 symptoms could account for the association of first positive tests 5-6 days before.
+
+Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
+The ECDC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.
 <!---Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated.
@@
 where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].
-<!---Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.--->
 
@@
 Supplementary Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
-To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.
+To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been allocated to the "Additional conditions" category; there were no severe cases in this group.
 Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
@@
 As shown in Supplementary Table \ref{tab:letterdates} most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of  shielding letters sent on 3 April 2020.
-Supplementary Figure \ref{fig:plotrateratio}(a) shows  the risk ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
+Supplementary Figure \ref{fig:plotrateratio}(a) shows  the rate ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
 The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding.
@@
 Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.
-<!--- tabulate underlying cause of death on death certs of fatal cases
-paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(hosp.recent, covid_ucod)))
-with(cc.severe[CASE==1], table(hosp.recent, inhosp))
-paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(occup, covid_ucod)))
---->
+
 The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.
 From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
-As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.
-The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.
+As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.
+The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDC classification was `r round(r.hcai, 1)`.
 
@@
 Our finding that even in those advised to shield the risk of severe disease was increased in those who were sharing a household with other adults suggests that this was another constraint on the effectiveness of advice to shield, as no support was provided for other household members to co-isolate with the vulnerable individual.
+A survey conducted online in the first two weeks of June 2020 found that 65% of those living alone or sharing a a household with one other person reported that they followed the guidance completely, but only 55% of those living with with two or more others [@delvaux_covid-19_2020].
 
@@
 We and others have previously reported  lower risk in teachers compared with others of the same age and sex [@mcallister_report_2020; @ludvigsson_open_2021].
-The inverse association of severe disease with the number of school-age children in the household extends and confirms the findings of an earlier study of health care workers and their families [@wood_living_2021].
+The inverse association of severe disease with the number of school-age children in the household extends and confirms the findings of an earlier study of health care workers and their families [@wood_sharing_2021].
 In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 after adjusting for covariates, but no dose-response relationship was reported [@forbes_association_2020].
@@
 ### Association with recent exposure to hospital
-A striking finding from our analysis is the  PARF  of  `r round(100 * pc.all * (r.all - 1) / r.all)` % for severe disease associated with recent exposure to hospital.
+A striking finding from our analysis is the  PARF  of  `r round(100 * pc.all * (r.all - 1) / r.all)`% for severe disease associated with recent exposure to hospital.
 Although we had pre-specified this category to include day case and outpatient exposure, the association was driven by inpatient exposure.
@@
 Most previous reports have focused on the narrow definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].
-The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].
+The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents) during 2020, 30% (1015/3389) met the  ECDC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].
 
-A paper presented to SAGE on 28 January 2021 broadened the ECDPC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure" [@knight_contribution_2021].
+A paper presented to SAGE on 28 January 2021 broadened the ECDC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure" [@knight_contribution_2021].
 From linking hospital episode data to COVID-19 tests and CO-CIN, the authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that onward transmission of COVID-19 from hospital-acquired cases could account for another 5% of hospitalised cases.
@@
 However the rate ratio associated with recent hospital exposure was barely changed when fatal cases without COVID-19 as underlying cause of death on death certificate were omitted.
-<!--- Chris R to run a comparison of 28-day fatality rates in those who test positive and those who test negative.
---->
 
@@
 (b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates.
-Line thickness is proportional to precision of estimate."}
+Line thickness is proportional to precision of estimate.
+Labels on the y axes are untransformed rate ratios, though the axis scales are log-transformed"}
 
@@
              align=rep("r", 2), 
-             caption="Hospital onset status as defined by ECDPC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
+             caption="Hospital onset status as defined by ECDC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
      add_footnote(label="Percentages are column percentages",
@@
              align=rep("r", 2), 
-             caption="Severe test-positive cases classified by hospital onset status (as defined by ECDPC) and recent inpatient exposure") %>%
+             caption="Severe test-positive cases classified by hospital onset status (as defined by ECDC) and recent inpatient exposure") %>%
 	 column_spec(column=1, width="4cm") %>%
@@
 	 column_spec(column=3, width="2.5cm") %>%
-#	 column_spec(column=4, width="2.5cm") %>%
-#	 column_spec(column=5, width="2.5cm") %>%
-     add_footnote(label="Percentages are column percentages",
+    add_footnote(label=c("Percentages are column percentages",
+                         "ECDC, European Centre for Disease Prevention and Control"),
                  notation="none") %>%

### 2021-08-29 transmissionBMED-D-21-00640_withoutsuppl.Rmd
reinfections

--- 2021-08-29 transmissionBMED-D-21-00640.Rmd
@@
   \usepackage{pdfpages}
+  \usepackage{soul}
   \linenumbers 
@@
     x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
+    x <- gsub("=<", "<", x)
     x <- gsub("times", "\\\\times", x)
@@
 * _Background_ -  Clinically vulnerable individuals have been advised to shield themselves during the COVID-19 epidemic.
-The objectives of this study were to investigate: (1) the risk of severe COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.
+The objectives of this study were to investigate:  1) the rate ratio of severe COVID-19  associated with eligibility for  the shielding programme in Scotland across the first and second waves of the epidemic, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general population”.
 * _Methods_ -  In a matched case-control design, all `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.
@@
 \linenumbers
-<!---
-# Author Summary
-## Why was this study done?  
-* Individuals designated as "clinically extremely vulnerable" -- about 3% of the population -- have been advised to shield themselves during the COVID-19 epidemic
-
-* The effectiveness of this shielding programme has not been evaluated.
-
-## What did the researchers do and find?  
-* Despite the provision of advice and support, the risk of severe COVID-19 remained far higher in those eligible for shielding than in those without risk conditions.
-
-* In both those eligible for shielding and in the rest of the population, the risk of severe COVID-19 was higher in those who were sharing a household with other adults, and was far higher in the 5 to 14 days after exposure to hospital than in those without recent exposure to hospital.
-
-* In the second wave of the epidemic in December 2020, more than half of the cases of severe COVID-19 in the population were attributable to recent hospital exposure.
-
-## What do these findings mean?  
-* For vulnerable individuals to be shielded, the risk of hospital-acquired infection must be reduced and support must be provided for other adults in the household to co-isolate with the vulnerable individual.
-
-* Solid organ transplant recipients are at the highest risk; vaccination of  household contacts should be considered for this group.
-
-* Limiting the burden of COVID-19 requires control of hospital-acquired infection. 
---->
 
@@
 Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].
-Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters advising strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
+Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters advising strict isolation, even from other members of the same household,  and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.
+Those living in multiple-occupancy households were advised to isolate from other household members.
 Compliance was voluntary.
 
-The initial objective of this study was to quantify the incidence of severe COVID-19 in the shielding population and to  investigate  whether shielding advice and support had reduced the risk of COVID-19.
-On the hypothesis that shielding was effective, we predicted that the rate ratio associated with eligibility for  shielding  compared to those without risk conditions would fall after the receipt of shielding letters.
-Our second objective was to understand the role of  transmission-related factors including household composition and recent exposure to hospital on COVID-19 among those eligible for shielding.
-As the relevance of these transmission-related factors among those eligible for shielding  became clear, the objectives were broadened to investigate the relation of severe COVID-19 to these transmission-related factors in  the general population.
+For this study the research question was to investigate the effectiveness of the shielding programme individuals and to identify avoidable factors that might have limited the effectiveness of this focused protection of vulnerable individuals.
+The original aim was to  investigate whether shielding advice and support had reduced the risk of COVID-19.
+The first objective was to quantify the incidence of severe COVID-19 in those eligible for shielding and to determine whether the rate ratio associated with eligibility for  shielding  compared to those without risk conditions had fallen after the receipt of shielding letters.
+The second objective was to understand the relation of severe COVID-19 in those eligible for shielding to transmission-related factors including household composition and recent exposure to hospital.
+As the relevance of these transmission-related factors among those eligible for shielding  became clear, the aims were broadened to investigate the relation of severe COVID-19 to these transmission-related factors in  the general population.
 
 # Methods
-We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case-control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions".
+## Overview
+## Registration
+The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
+We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all persons diagnosed with COVID-19 in Scotland.
+Each case was matched to up to 10 controls of the same sex and in the same one-year age band, registered with the same primary care practice and not previously diagnosed with COVID-19, who were alive on the date of presentation of the case controls.[@mckeigue_rapid_2020]  This case-control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions".
 Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.
-The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio as described below.
+As a first step we report the cumulative incidence of diagnosed COVID-19 among those in shielding.
+For all tests of association and comparisons of rates between groups we limit the analysis to severe  cases and their respective controls as described below, and we exclude care home residents.
+
+The study period was from 1 March 2020 (the date of the first case in Scotland ) to 18 February 2021 (date of the latest data extract available for this study).
+Thus it predates the start of the shielding programme.
 
@@
 
-Cases of COVID-19 were defined as those with a positive  nucleic acid test for SARS-CoV-2 in ECOSS, a hospital discharge diagnosis of COVID-19 in SMR01, or a death registration with mention of COVID-19 anywhere on the death certificate.
+Cases diagnosed with COVID-19 were defined as those with a positive  nucleic acid test for SARS-CoV-2 in ECOSS, a hospital discharge diagnosis of COVID-19 in SMR01, or a death registration with mention of COVID-19 anywhere on the death certificate.
+Thus not all those  cases defined as having evidence of COVID-19  had a positive test.
+To restrict cases to those with a positive test would have missed a large proportion of deaths in the first wave as the availability of tests was restricted.
+Some cases with a definite clinical diagnosis of COVID-19 based on typical presentation and radiological signs nonetheless tested negative when they presented late in their illness.
+In  the first wave (up end of August 2020), only `r table.testpos[1, 2]` of the severe cases were test-positive; in the second wave beginning on 1 September 2020, `r table.testpos[2, 2]` were test-positive.
+
 The presentation date was assigned as the date of the first positive test for those ascertained through testing, as seven days before the admission date for those without a positive test result ascertained through hospital discharge records, and as fourteen days before the date of death for those without a positive test result ascertained through death certificates.
 Databases were queried from 1 March 2020  (date of the first diagnosed case of COVID-19 in Scotland) up to `r dateformat(lastdate.specimen)` for test results and `r dateformat(maxdate.death)` for deaths.
+For each case  diagnosed with COVID-19, up to 10  community controls matched for sex, one-year age band and primary care practice were selected from the CHI database, and assigned the same presentation date as the case.
+With this incidence density sampling design, it is possible and correct for an individual to appear more than once as a control and subsequently as a case.
+As primary care practice catchment areas are localized, matching on primary care practice matches for geographic factors also.
+s
+### Inclusion and exclusion criteria for analyses 
+We reported the cumulative incidence of diagnosed COVID-19 in the shielded and further categorised these cases into those that required critical care or were fatal or not.
+We defined severe COVID-19 as those diagnosed with COVID-19 who also required  entry to critical care within 21 days of presentation date, or had a fatal outcome.
+Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
+Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained only through discharge records.
 
-Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome.
-This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.
-In the first wave (up end of August 2020), `r table.testpos[1, 2]` of the severe cases were test-positive; in the second wave beginning on 1 September 2020, `r table.testpos[2, 2]` were test-positive.
-
-Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database.
-Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.
-
-For each case of COVID-19, up to 10  community controls matched for sex, one-year age band and primary care practice were selected from the CHI database, and assigned the same presentation date as the case.
-With this incidence density sampling design, it is possible and correct for an individual to appear more than once as a control and subsequently as a case.
+For all subsequent tests of association and comparisons of rates between groups presented here, we restricted the dataset to cases with severe COVID-19 and their matched controls from the population.
+The narrow definition of severe COVID-19 was defined as the main outcome measure of the REACT-SCOT study at the design stage, to ensure that ascertainment would not be biased by variation in testing policies or selection for hospitalisation.
+In the first wave in Scotland people with symptoms were advised to stay at home and not to seek testing or attend hospital unless their condition deteriorated.
+A broader case definition based on test-positive or hospitalized cases would have given rise to selection bias if, for equivalent severity of symptoms, those with pre-existing risk conditions were more likely to be tested or hospitalised than those without pre-existing risk conditions.
+For all tests of association and comparisons of rates between groups other than the initial tabulation of cumulative incidence of any diagnosis of COVID-19 in the shielded,  we excluded care home residents because policies for shielding  care home residents are different from those relevant to shielding individuals living independently.
 
@@
 Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020].
-Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.
+Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System and to the national register of diabetes.
 We used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic airway disease, chronic kidney disease, disabling neurological conditions, and immune deficiency or suppression.
@@
 It is therefore unlikely that hospital visits by people with COVID-19 symptoms could account for the association of first positive tests 5-6 days before.
+As the average incubation period for COVID-19 is 5-6 days [@dhouib_incubation_2021], exposure in the time window is relevant.
 
-Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
-The ECDC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.
-<!---Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated.
---->
+Our intention was to capture all relevant exposure in cases and controls, rather than to assign individual cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].
+The ECDC definition of probable/definite health-care associated COVID-19 is restricted to those who have been in hospital for at least 8 days before first developing symptoms: all other cases are classified as community onset.
+The definition of "recent hospital exposure" used in this study was intended to capture all those whose infection could have been acquired in hospital (sensitivity of 1), unlike the ECDC definition of "probable/definite health-care associated infection" which is intended to identify those whose infection was unlikely to have been acquired outside hospital (high specificity).
+Because this study includes a control group, the calculation of the population attributable risk fraction is valid if the classification of exposure has sensitivity of 1 even if the specificity is less than 1 [@hsieh_effect_1988].
+For estimating the population attributable risk fraction, it is therefore appropriate to use an inclusive definition of exposure, even if this includes some individuals whose infection was not acquired in hospital.
 
 ## Statistical methods
-The analyses presented here focus on severe COVID-19, defined as the main outcome measure of  of the REACT-SCOT study at the design stage.
-Ascertainment of this outcome does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is likely to differ between those eligible for shielding and those ineligible.
-For the case-control analyses, cases and controls were censored at date of first vaccination and care home residents were excluded, as shielding advice intended for individuals in private households would not have been relevant to care home residents.
+Other than the cumulative incidence analyses the analyses presented here focus on severe COVID-19,  and their respective controls.
+For the matched case-control analyses, severe cases and their respective controls were censored at date of first vaccination.
 
-Cases and controls were grouped into three categories: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.
-Rate ratios for severe COVID-19 associated with eligibility for shielding are calculated with "no risk condition" as reference category.
-All rate ratios were estimated by conditional logistic regression.
-As the  community controls were drawn by incidence density sampling and matched for age, sex, and general practice, this controls for these variables and for calendar time.
+Among all severe cases and controls, conditional odds ratios for severe COVID-19 associated with eligibility for shielding and for other moderate risk conditions were calculated with "no risk condition" as reference category using conditional logistic regression.
+With this  incidence density design the conditional odds ratios are equivalent to rate ratios so are referred to as such throughout the manuscript.
+As the  community controls were drawn by incidence density sampling and matched for age, sex, and general practice, the conditional logistic regression inherently controls for these variables and for calendar time.
+Conditional logistic regression was also used to test for association of  transmission-related factors with severe COVID-19.
+These transmission-related factors were chosen *a priori* as factors that would plausibly constrain the extent to which individuals were able to limit their exposure to infection: the need for hospital care, deprivation score since it may relate to overcrowding or the economic imperative to work, number of persons in the household and usual occupation.
+A multivariable model  was used to adjust for shielding condition, moderate risk condition and transmission-related factors  simultaneously.
+These analyses were repeated restricted to those eligible for shielding.
+The multivariable analyses in Tables 2 and 3 are simultaneously adjusted for all variables shown in the table.
 
-To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.
-The population attributable risk fraction (PARF) of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.
+To plot the time course of the rate ratios for severe COVID-19  associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.
+The population attributable risk fraction of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.
 where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].
 
-Sliding windows of size  `r winsize.casedates`, `r winsize.hosp`, and `r winsize.fatality` days were used to plot the number of severe cases, the frequency of recent hospital exposure, and the rate ratios associated with hospital exposure and household composition.
-In the plots of exposure frequency and rate ratio the data points from 1 June 2020 to 30 September 2020 are omitted as the numbers of cases and controls are too low for estimates of frequencies and rate ratios to be accurate.
+Sliding windows of size  `r winsize.casedates`, `r winsize.hosp`, and `r winsize` days were used to plot the number of severe cases, the frequency of recent hospital exposure, and the rate ratios associated with hospital exposure and household composition.
+In the plots of exposure frequency and rate ratio the data points from 1 June 2020 to 30 September 2020 were omitted as the numbers of severe cases and controls were too low for estimates of frequencies and rate ratios to be accurate.
 
-To test for causality of the association of recent exposure to hospital with severe COVID-19, we estimated rate ratios associated with recent hospital exposure (5 to 14 days before first testing positive) with less recent exposure (15 to 24 days before first testing positive) as reference category.
-This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases.
+To test for causality of the association of recent exposure to hospital with severe COVID-19, we estimated the conditional odds ratios associated with recent hospital exposure (5 to 14 days before first testing positive) with less recent exposure (15 to 24 days before first testing positive) as reference category.
+This analysis was restricted to test-positive severe cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases.
 The classic case-crossover design compares in cases only the frequencies of exposure in recent and less recent time windows, and estimates the conditional odds ratio as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only.
 This can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000].
-The analysis that we report is a refinement that takes advantage of the availability of a matched control group to contrast the conditional odds of being a case between those with recent exposure only and those with less recent exposure only.
+The analysis that we report is a refinement that takes advantage of the availability of a matched control group to estimate the rate ratio for severe COVID-19 associated with exposure only in a recent time window, with those exposed only in a less recent time window as the reference category.
 This controls for any difference in the population frequencies of exposure or disease incidence between recent and less recent time windows.
@@
 ## Shielding eligibility and risk of COVID-19
-Supplementary Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.
+Additional File: Table S1 shows the frequency of risk categories in those eligible for shielding.
 To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been allocated to the "Additional conditions" category; there were no severe cases in this group.
 Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.
-
-As shown in Figure \ref{fig:plotcasesbydate} the time course of daily severe cases in those eligible for shielding paralleled the time course in those without risk conditions.
-However in the first wave the daily number of severe cases fell more slowly in those eligible for shielding than in those without risk conditions.
+As shown in Figure \ref{fig:plotcasesbydate} the time course of  frequency of daily severe cases in those eligible for shielding  broadly paralleled the time course in those without risk conditions.
 Of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases among those eligible for shielding, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` were resident in care homes.
-In  the case-control analyses, care home residents were excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.
 
@@
 `r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
-in those eligible for shielding.
+in those eligible for shielding. The rate ratio associated with eligibility for shielding was 
+`r format.estci(table.listedgr3.wave1["Eligible for shielding", 3])` up to the end of August 2020 and `r format.estci(table.listedgr3.wave2["Eligible for shielding", 3])` after August 2020.
+
 Among those eligible for shielding, solid organ transplant recipients were the group at highest risk, with a univariate rate ratio of  `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` for severe COVID-19.
-On multivariate adjustment rate ratios for all these groups remained high.
+On multivariable adjustment for the covariates shown in Table \ref{tab:regressionseverecasesnocare} the rate ratios for all these shielding groups remained high.
 
 We examined the time course of the rate ratio associated with eligibility for shielding.
-As shown in Supplementary Table \ref{tab:letterdates} most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of  shielding letters sent on 3 April 2020.
-Supplementary Figure \ref{fig:plotrateratio}(a) shows  the rate ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
+As shown in Additional File: Table S2 most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of  shielding letters sent on 3 April 2020.
+Additional File: Figure S1(a) shows that the rate ratio associated with eligibility for shielding   increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.
 The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding.
@@
 The rate ratio increased with the number of adults in the household but not with the number of children.
-The strongest risk factor for severe COVID-19 among those eligible for shielding was recent exposure to hospital, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 6])` in the multivariable model.
+The strongest risk factor for severe COVID-19 among those eligible for shielding was recent exposure to hospital, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 6])` in the multivariable model which all the covariates shown n the table were entered simultaneously.
 Of severe cases among those eligible for shielding, `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2]` had recent  exposure to hospital.
-Using Miettinen's formula as given above, the PARF for recent  exposure to hospital in those eligible for shielding can thus be calculated as `r round(100 * pc * (r -1) / r)`% of severe cases.
+Using Miettinen's formula as given above, the population attributable risk fraction for recent  exposure to hospital in those eligible for shielding can thus be calculated as `r round(100 * pc * (r -1) / r)`% of severe cases.
 
@@
 Table \ref{tab:regressionseverecasesnocare}  shows the association of severe COVID-19 with risk factors in the general  population,  including those eligible for shielding but excluding residents in care homes.
-The risk of severe COVID-19 increased with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model.
+The risk of severe COVID-19 increased with the number of adults in the household, but was inversely associated with the number of school-age children in the household in a  multivariable model in which all the covariates in Table \ref{tab:regressionseverecasesnocare} were entered simultaneously.
 The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.
 The other demographic factor associated with increased risk was socioeconomic deprivation: the rate ratio in the least deprived quintile compared with the most deprived quintile was `r format.estcipv(table.shielded.severecases.nocare["5 - least deprived", 5], table.shielded.severecases.nocare["5 - least deprived", 6])`.
-As shown in Supplementary Figure \ref{fig:plotrateratio}(b), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease per adult in household remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
+As shown in Additional File: Figure S1(b), in a joint model with number of adults in household, number of children in household, and SIMD deprivation score the rate ratio for severe disease per adult in household remained in the range 1.5 to 2 throughout the epidemic, and the rate ratio per school-age child in household remained mostly in the range 0.7 to 1.
 
@@
 
-The  PARF of severe COVID-19 for recent  exposure to hospital in the general population can be calculated as `r round(100 * pc.all * (r.all - 1) / r.all)`% over the study period.
+The population attributable risk fraction of severe COVID-19 for recent  exposure to hospital in the general population can be calculated as `r round(100 * pc.all * (r.all - 1) / r.all)`% over the study period.
 This fraction reached a peak in the first wave of `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`%  of severe cases at the beginning of May, declined rapidly over the next few weeks, and increased again to `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December (Figure \ref{fig:plotexphosp}(c)).
@@
 The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.
-From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
-As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.
-The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDC classification was `r round(r.hcai, 1)`.
+From this and the  rate ratio of `r round(r.rapid, 1)` the population attributable risk fraction for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.
+As Additional File: Table S3 shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECDC classification.
+The rate ratio for severe disease  associated with being in hospital for at least 8 days before presentation date was `r round(r.hcai, 1)`.
 
@@
 
-If the advice and support offered in the shielding letters had been effective (beyond the messages already widely disseminated by this time and  general social distancing measures that were introduced  by then), we would have expected the rate ratio for severe disease associated with eligibility for shielding to fall within two weeks of the first letters being sent out.
-Our results show however that while presentations of severe cases fell rapidly in the general population from the beginning of April 2020, the fall in presentations among those eligible for shielding advice was delayed so that the rate ratio for severe COVID-19 associated with eligibility for shielding rose during April 2020, although the daily deaths fell during April 2020 both in those eligible and those ineligible for shielding.
-The rate ratios associated with eligibility for shielding and with moderate risk conditions rose again from October 2020 to early December.
+The shielding programme was introduced at a time when general social distancing measures were being introduced and after the first lockdown had started.
+We had hypothesised *a priori* that a fall in the rate ratio for severe disease associated with eligibility for shielding  within two weeks of the first letters being sent out would be consistent with the  shielding  programme having some impact beyond the population-wide lockdown that had been imposed by this time.
+However we did not find this; while presentations of severe cases fell rapidly in the general population from the beginning of April 2020, the fall in presentations among those eligible for shielding advice was delayed so that the rate ratio for severe COVID-19 associated with eligibility for shielding rose during April 2020.
+Daily deaths fell during April 2020 both in those eligible and those ineligible for shielding.
+As a lockdown on the general population had been imposed on 23 March 2020, and individuals who considered themselves to be at high risk would have been likely to reduce their contact level before then, there were limited possibilities for risk in this group to be reduced further by advice to shield and offers of help in letters sent from 3 April onwards.
 
-The high frequency of recent hospital exposure in cases of severe COVID-19 who were eligible for shielding, together with the increase in the rate ratio associated with recent hospital  exposure during periods when population-wide social distancing measures were being imposed, suggests that exposure to transmission in hospital settings is at least part of the explanation for the failure to reduce the rate ratio associated with eligibility for shielding.
-Our finding that even in those advised to shield the risk of severe disease was increased in those who were sharing a household with other adults suggests that this was another constraint on the effectiveness of advice to shield, as no support was provided for other household members to co-isolate with the vulnerable individual.
-A survey conducted online in the first two weeks of June 2020 found that 65% of those living alone or sharing a a household with one other person reported that they followed the guidance completely, but only 55% of those living with with two or more others [@delvaux_covid-19_2020].
+Although we did not find any evidence that the shielding programme *per se* reduced COVID-19 rates, it is possible that without shielding advice and support the outcome in this group would have been worse.
+It is still relevant to examine why advice to shield, combined with offers of support for delivery of food and medicines, failed to protect some individuals who were identified as clinically extremely vulnerable.
+The association of severe COVID-19 with recent hospital exposure in those who were eligible for shielding, together with the increase in the rate ratio associated with this exposure during periods when population-wide social distancing measures were being imposed, suggests that exposure to transmission in hospital settings is at least part of the explanation.
+Our finding that the risk of severe COVID-19 was increased in those who were sharing a household with other adults suggests that this was another constraint on the effectiveness of advice to shield, as no support was provided for other household members to co-isolate with the vulnerable individual.
+A survey conducted online in the first two weeks of June 2020 found that 65% of those living alone or sharing a a household with one other person reported that they followed the guidance completely, but only 55% of those living with with two or more others did [@delvaux_covid-19_2020].
+Development of policies to overcome these constraints may lay a basis for focused protection of the most vulnerable individuals to be more effective in future epidemics.
 
-### Association with household composition
+### Association with  occupation and with household composition
 The increased risk of severe COVID-19 in patient facing  health care workers is consistent with our previous report of a threefold risk for hospitalised COVID-19 earlier in the epidemic [@shah_risk_2020].
@@
 In the OPENSAFELY cohort, the rate ratio for fatal COVID-19 associated with living with children aged 0-11 years was 0.75 after adjusting for covariates, but no dose-response relationship was reported [@forbes_association_2020].
-The inverse association of severe COVID-19 with past exposure to children is consistent with evidence that  other coronaviruses generate cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020].
-From a public health perspective the most relevant implication is that although the rate ratio per child has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows.
-The association of severe COVID-19 with number of adults in the household is consistent with classic studies of other viral infections showing that secondary cases in a household, where the infecting dose is likely to be high, are more severe than index cases [@aaby_malnutrition_1988].
+
+The strong association of severe disease with number of adults in the household is consistent with a recent estimate, based on an intensive study of households in the Netherlands, that the secondary attack rate for SARS-CoV-2 infection is as high as 51% in adults [@reukers_high_2021].
+We would expect the risk of infection to increase with the number of other people in the household, especially where population-wide social distancing between households has been imposed.
+We might also expect that this association would be strongest for severe disease, as the infecting dose is likely to be higher in intra-household transmissions than in transmissions between households.
+This is consistent with classic studies of other viral infections showing that secondary cases in a household are more severe than index cases [@aaby_malnutrition_1988].
+The inverse association of severe COVID-19 with number of children in the household, and the lower risk in teachers however suggests that current or past exposure to children may have some protective effect on adults.
+A possible explanation may be that other coronaviruses generate cross-reactive T-cell responses that may confer some resistance to SARS-CoV-2 [@braun_sars-cov-2-reactive_2020].
+From a public health perspective the most relevant implication is that although the rate ratio per child in the household has been higher in the second wave of the epidemic than in the first wave, it has remained below 1 in almost all time windows.
 
 ### Association with recent exposure to hospital
-A striking finding from our analysis is the  PARF  of  `r round(100 * pc.all * (r.all - 1) / r.all)`% for severe disease associated with recent exposure to hospital.
+A striking finding from our analysis is the  population attributable risk fraction of `r round(100 * pc.all * (r.all - 1) / r.all)`% for severe disease associated with recent exposure to hospital.
 Although we had pre-specified this category to include day case and outpatient exposure, the association was driven by inpatient exposure.
-The PARF associated with inpatient exposure was  `r round(100 * parf.rapid)`%.
-Using the accepted international definitions of probable or definite nosocomial acquisition gives a much lower estimate of the PARF.
-As we have shown most of this difference is due to the fact that the international definition does not capture cases diagnosed shortly after discharge that could have been acquired in hospital.
-The other difference  is that the  international definition classifies  those with onset 5-6 days after admission as indeterminate.
-As the average incubation period for COVID-19 is 5-6 days, exposure in this time window is relevant.
+The population attributable risk fraction associated with inpatient exposure was  `r round(100 * parf.rapid)`%.
+The calculation of the population attributable risk fraction for an exposure provides an upper bound on the predicted effect of removing that exposure.
+As explained in the Methods section, for estimating the population attributable risk fraction it is appropriate to use an inclusive definition of exposure that includes all those whose infection could have been acquired in hospital.
+Using the ECDC definition of probable/definite nosocomial acquisition would give a lower estimate of the  proportion of severe cases exposed to hospitals and the population attributable risk fraction, because the ECDC definition excludes cases who were not continuously in hospital for at least 8 days before presenting.
 
-The role of nosocomial transmission was recognized early in the epidemic.
+The association of severe COVID-19 with recent hospital admission is likely to be confounded by pre-existing risk conditions.
+However adjusting for risk conditions in a multivariable analysis reduced the rate ratio associated with recent hospital exposure only slightly (from  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 3])), 1)` to  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 5])), 1)`).
+The most compelling evidence of causality is the time window analysis, which shows that the association of disease is with elapsed time since exposure corresponding to the known incubation period of SARS-CoV-2.
+The rate ratio estimated from the time window analysis is not directly comparable with the rate ratio based on the case-control analysis because restriction to those with discordant exposure between time windows selects those with  fewer days in hospital (lower dosage of the exposure).
+
+The role of nosocomial transmission was recognized early in the COVID-19 epidemic.
 Hospital-acquired infection was suspected in 41% of hospitalised cases seen in January 2020 in one centre in Wuhan [@wang_clinical_2020-2].
 A commentary in April 2020 noted that "in Lombardy, SARS-CoV-2 became largely a nosocomial infection" [@boccia_what_2020].
-Most previous reports have focused on the narrow definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].
-The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents) during 2020, 30% (1015/3389) met the  ECDC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].
-
-A paper presented to SAGE on 28 January 2021 broadened the ECDC definition of health-care associated infection to include cases discharged from hospital up to 14 days before first testing positive as "community-onset suspected healthcare-associated", similar to  the definition that we have used for "recent hospital exposure" [@knight_contribution_2021].
-From linking hospital episode data to COVID-19 tests and CO-CIN, the authors estimated that 40% of cases hospitalised with COVID-19 were exposed by this broad definition, and that onward transmission of COVID-19 from hospital-acquired cases could account for another 5% of hospitalised cases.
+A study of 11 hospitals (10 in the UK) reported that 12.5% of hospitalized cases of COVID-19 up to the end of April 2020 met the ECDC criterion for definite nosocomial infection [@carter_nosocomial_2020].
+Most other reports have used the ECDC criterion for probable/definite nosocomial infection   A study presented to the UK Scientific Advisory Group on Emergencies in January 2021 based on national data for hospital episodes and COVID-19 tests in England linked to symptom onset data from the COVID-19 Clinical Information Network estimated that 27.6% of hospitalized cases up to the end of July 2020 met this criterion, and that onward transmission of COVID-19 from hospital-acquired cases could account for another 5% of hospitalised cases [@knight_contribution_2021].
+No systematic study of nosocomial cases has been reported for England in the second wave: from routinely reported data the proportion of hospitalized cases that met the criterion for probable/definite nosocomial was estimated to be 18% in October 2020.
+For Scotland the ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents) during 2020, 30% met the  ECDC definition of probable or definite nosocomial infection [@arhai_scotland_hospital_2021].
 
@@
 We have no data on help with daily activities from non-resident carers as another possible source of exposure of clinically vulnerable individuals attempting to shield themselves.
-
-The calculation of the PARF for an  exposure provides an upper bound on the predicted effect of removing that exposure.
-The definition of "recent hospital exposure" used in this study was intended to capture all those whose infection could have been acquired in hospital (sensitivity of 1), unlike the international definition of "probable health-care associated infection" which is intended to identify those whose infection was unlikely to have been acquired outside hospital (high specificity).
-This inclusive definition is appropriate, as the calculation of the PARF is valid if the classification of exposure has sensitivity of 1 even if the specificity is less than 1 [@hsieh_effect_1988].
-The association of severe COVID-19 with recent hospital admission is likely to be confounded by pre-existing risk conditions.
-However adjusting for risk conditions in a multivariable analysis reduced the rate ratio associated with recent hospital exposure only slightly (from  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 3])), 1)` to  `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare["Recent hospital visit/stay", 5])), 1)`).
-The most compelling evidence of causality is the time window analysis, which shows that the association of disease is with elapsed time since exposure corresponding to the known incubation period of SARS-CoV-2.
-The rate ratio estimated from the time window analysis is not directly comparable with the rate ratio based on the case-control analysis because restriction to those with discordant exposure between time windows selects those with  fewer days in hospital (lower dosage of the exposure).
+As we do not have individual-level data on compliance with shielding advice, we cannot directly evaluate the extent to which non-compliance or non-receipt of letters might have reduced the effectiveness of shielding advice: we can only evaluate the programme as implemented.
 
@@
 Our results have implications for policies on shielding the vulnerable, vaccination, and control of SARS-CoV-2 transmission in the population.
-
-We found no evidence that the shielding programme *per se* reduced COVID-19 rates although it is possible that without shielding advice and support the outcome in this group would have been worse.
-It is relevant to examine why advice to shield, combined with offers of support for delivery of food and medicines, failed to protect some individuals who were identified as clinically extremely vulnerable.
-As a lockdown on the general population had been imposed on 23 March 2020, and individuals who considered themselves to be at high risk would have been likely to reduce their contact level before then, there were limited possibilities for risk in this group to be reduced further by advice to shield and offers of help in letters sent from 3 April onwards.
 We have identified two sources of exposure that are associated with severe disease and cannot easily be avoided by those advised to shield: in-patient hospital care, and sharing a household with other adults.
-We recommend that consideration should be given to special measures for to allow these  vulnerable individuals to reduce their exposure.
-This could include special measures to protect these individuals from nosocomial infection, and support for other household members to co-isolate with the vulnerable individual.
+Measures to allow vulnerable individuals to reduce their exposure could include extra protection from  nosocomial infection, and support for other household members to co-isolate with the vulnerable individual.
 
@@
 As there is now evidence that vaccination reduces transmission to household contacts [@public_health_scotland_covid-19_2021], one way to reduce risk in this group would be to vaccinate their household contacts.
-Even with vaccination of  household contacts, solid organ transplant recipients will require shielding until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.
 
@@
 
+# Conclusions
+The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.
+Mitigating the impact of the epidemic requires control of nosocomial transmission.
+
+# List of abbreviations  
+CHI, Community Health Index  
+ECDC, European Centre for Disease Prevention and control  
+ECOSS, Electronic Communication of Surveillance in Scotland  
+NHS, National Health Service  
+NRS, National Records of Scotland  
+PHS, Public Health Scotland  
+SICSAG, Scottish Intensive Care Society and Audit Group  
+SIMD, Scottish Index of Multiple Deprivation  
+SMR, Scottish Morbidity Record
+
 # Declarations
-<!---## Public and Patient Involvement statement:
-This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives.
---->
+## Ethics approval and information governance
+Record linkage studies using NHS data in Scotland are governed by the Public Benefit and Privacy Panel for Health and Social Care.
+This study was approved under COVID-19 Rapid Data Protection Impact Assessment​ 20210023 that allows Public Health Scotland staff to link datasets without requiring individual consent and to make de-identified datasets available for analysis within Public Health Scotland.
 
-## Ethics approval and information governance
-This study was approved under COVID-19 Rapid Data Protection Impact Assessment​ (DPIA) 20210023 that allows Public Health Scotland staff to link the datasets.
-Datasets were de-identified before analysis.
-<!---## Exclusive License
-The Corresponding Author has the right to grant on behalf of all authors and does grant on behalf of all authors, a worldwide licence to the Publishers and its licensees in perpetuity, in all forms, formats and media (whether known now or created in the future), to i) publish, reproduce, distribute, display and store the Contribution, ii) translate the Contribution into other languages, create adaptations, reprints, include within collections and create summaries, extracts and/or, abstracts of the Contribution, iii) create any other derivative work(s) based on the Contribution, iv) to exploit all subsidiary rights in the Contribution, v) the inclusion of electronic links from the Contribution to third party material where-ever it may be located; and, vi) licence any third party to do any or all of the above.--->
-
-<!---## Transparency declaration
-PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained.
-This manuscript has been generated directly from the source data by a reproducible research pipeline.
---->
+## Consent for publication
+Not applicable
 
 ## Funding 
-No specific funding was received for this study.
+No specific funding was received for this study. HC is supported by an AXA Foundation fellowship.
 
 ## Data Availability
-The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
+The component datasets used here are available via the Public Benefits and Privacy Panel for Health and Social Care https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
 All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.
 
-## Competing interest
-All authors have completed and submitted the ICMJE Form for Disclosure of Potential Conflicts of Interest.
+## Competing interests
+HC receives research support and honoraria and is a member of advisory panels or speaker bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly.
+HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.
+SH received honoraria from Gilead.
 
-<!---## Contributors
+## Author contributions
 PM, DM and HC provided substantial contributions to the conception or design and drafting of the manuscript.
@@
 SH, CR, JM and DG provided substantial contributions to the data analysis or interpretation.
-All authors contributed to revising the manuscript critically for important intellectual content and approved the final manuscript.--->
-
-## Registration
-The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
+All authors contributed to revising the manuscript critically for important intellectual content and approved the final manuscript.
 
 ## Acknowledgements
-We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories, and Nicola Rowan (HPS) for coordinating this collaboration.
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories.
 We thank Shona Cairns and Laura Imrie of the Antimicrobial Resistance and Healthcare Associated Infection unit of NHS National Services Scotland for comments on an earlier draft of this manuscript.
@@
 
+# Figure legends  
+Fig 1: Area plot of severe cases by vulnerability category and date of presentation, excluding care home residents
+
+Fig 2: (a) Recent hospital exposure by risk group and 7-day sliding window of presentation dates; (b) Rate ratio for severe COVID-19 associated with recent hospital exposure in those not resident in care homes; (c) fraction of severe cases attributable to recent hospital exposure in those not resident in care homes
+
+# Additional files  
+File name: BMED-D-21-00640_SupplementaryMaterial.pdf
+Title: Additional File
+Description: Supplementary Figure and Tables
+
+\newpage
+
 # Figures
@@
 
-```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and sliding window of presentation dates (b) Rate ratio asociated with recent hospital exposure in those not resident in care homes.
-(c) fraction of severe cases attributable to recent hospital exposure in those not resident in care homes."}
+```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Recent hospital exposure by risk group and 7-day sliding window of presentation dates; (b) Rate ratio for severe COVID-19 associated with recent hospital exposure in those not resident in care homes; (c) fraction of severe cases attributable to recent hospital exposure in those not resident in care homes."}
 
@@
 
+colnames(table.shield.casegr)[1] <- "Not diagnosed with COVID-19"
 knitr::kable(table.shield.casegr, 
@@
              align=rep("r", 5),
-             caption="Shielding eligibility cohort by case status. Severe cases are those that were fatal or entered critical care") %>%
+             caption="Frequencies of any diagnosis of COVID-19, entry to critical care and fatal outcome among those eligible for shielding, by eligibility category, up to 18 February 2021") %>%
+    add_header_above(c(" "=2, "Diagnosed with COVID-19"=3)) %>%
     kableExtra::group_rows("Shielding eligibility category", 2, 7) %>%
@@
     column_spec(column=6, width="2cm") %>%
-        add_footnote(label="Percentages are row percentages",
+        add_footnote(label=c("Percentages are row percentages", 
+                             "Severe cases are those that entered critical care or were fatal"),
                      notation="none") %>%
@@
     kable_styling(latex_options=c("scale_down", "HOLD_position")) %>%
-    add_footnote(label=c("Percentages are column percentages for each variable",
+    add_footnote(label=c("Severe COVID-19 is defined by entry to critical care or fatal outcome", 
+                         "Percentages are column percentages for each variable",
                          "PF, patient-facing",
@@
                          "Univariate rate ratios are for models with a single covariate",
-                         "Multivariable rate ratios are for a model including all covariates"),
+                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
                  notation="none")
@@
     kableExtra::group_rows("Adults in household", 11, 13, bold=FALSE) %>%
-    add_footnote(label=c("Percentages are column percentages for each variable",
+    add_footnote(label=c("Severe COVID-19 is defined by entry to critical care or fatal outcome", 
+                         "Percentages are column percentages for each variable",
                          "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
                          "Univariate rate ratios are for models with a single covariate",
-                         "Multivariable rate ratios are for a model including all covariates"),
+                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
                  notation="none") %>%
@@
              row.names=FALSE,
-             caption="Rate ratios for severe COVID-19 by time window of hospital in-patient exposure", 
+             caption="Rate ratios for severe COVID-19 by time window of hospital in-patient exposure, excluding care home residents", 
              col.names=c("Time window of exposure", "Controls", "Cases",
@@
                          "first testing positive.",
-                         "Recent interval = days 5 to 14, less recent interval days 15 to 24 before first testing positive.",
+                         "Recent interval = days 5 to 14 before first testing positive, less recent interval = days 15 to 24 before first testing positive.",
                          "Reference category is exposure in less recent interval only."), 
@@
 ```
-
-\beginsupplement
-\newpage
-\onehalfspacing
-
-# Supplementary Material
-## Supplementary Figures
-
-```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=1.2, fig.cap="(a) Rate ratios for severe COVID-19 in those not resident in care homes, by risk group and sliding window of presentation dates.
-(b) Rate ratio per adult and per child in household, in models with number of adults, number of children and SIMD deprivation score as covariates.
-Line thickness is proportional to precision of estimate.
-Labels on the y axes are untransformed rate ratios, though the axis scales are log-transformed"}
-
-gridExtra::grid.arrange(p.rateratio + theme(axis.title.x = element_blank()),
-                        p.household.rateratio, nrow=2)
-
-```
-
-\newpage
-
-## Supplementary Tables
-
-```{r shieldgroupsbyage, echo=FALSE, warning=FALSE, message=FALSE}
-knitr::kable(table.shieldgroups.byage, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="shieldgroupsbyage",
-             row.names=TRUE,
-             align=rep("r", 5),
-             caption="Shielding eligibility cohort by eligibility category and age") %>%
-    kableExtra::group_rows("Shielding eligibility category", 1, 6) %>%
-        add_header_above(c(" "=1, "Age group"=4)) %>%
-        add_footnote(label="Percentages are row percentages",
-                     notation="none") %>%
-       kable_styling(latex_options="HOLD_position")
-
-```
-
-```{r hcai_hosp, echo=FALSE, warning=FALSE, message=FALSE}
-if(FALSE) {
-knitr::kable(table.hcai.hosp.recent,
-             escape=FALSE, 
-             booktabs=TRUE,
-             col.names=c("No recent hospital exposure", "Recent hospital exposure"), 
-			 label="hcai_hosp",
-             row.names=TRUE,
-             align=rep("r", 2), 
-             caption="Hospital onset status as defined by ECDC, by any recent hospital exposure including day cases, outpatient attendance and psychiatric admissions") %>%
-     add_footnote(label="Percentages are column percentages",
-                 notation="none") %>%
-    kable_styling(latex_options=c("hold_position"))
-}
-```
-
-```{r hcai_rapid, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.hcai.rapid.recent,
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hcai_rapid",
-             row.names=TRUE,
-             align=rep("r", 2), 
-             caption="Severe test-positive cases classified by hospital onset status (as defined by ECDC) and recent inpatient exposure") %>%
-	 column_spec(column=1, width="4cm") %>%
-	 column_spec(column=2, width="2.5cm") %>%
-	 column_spec(column=3, width="2.5cm") %>%
-    add_footnote(label=c("Percentages are column percentages",
-                         "ECDC, European Centre for Disease Prevention and Control"),
-                 notation="none") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r letterdates, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.date.letter,
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="letterdates",
-             row.names=TRUE,
-             align=rep("r", 6), 
-             caption="Dates of sending advice letters to those eligible for shielding") %>%
-    column_spec(column=1, width="3cm") %>%
-    column_spec(column=2, width="2.1cm") %>%
-    column_spec(column=3, width="1.9cm") %>%
-    column_spec(column=4, width="1.9cm") %>%
-    column_spec(column=5, width="1.9cm") %>%
-    column_spec(column=6, width="1.9cm") %>%
-    add_footnote(label="Percentages are row percentages",
-                 notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "hold_position"))
-
-```

### 2021-08-29 under50norisk.Rmd
reinfections

```{r freqsriskgroups, echo=FALSE, message=FALSE}
if(FALSE) {
knitr::kable(freqs.riskgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsriskgroups",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of severe COVID-19 from April to July 2021 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Age 0-49 years"=2, "Age 50 years and over"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Vaccination status", 9, 11, bold=TRUE) %>% 
    column_spec(1, width="6cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="2.5cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2.5cm") %>%
   kable_styling(latex_options=c("HOLD_position", "scale_down"))

##```

##```{r severebyagevax, echo=FALSE, message=FALSE}

knitr::kable(severe.byagevax, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="severeagevax",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Severe cases from April to July 2021 by age group and vaccination status") %>%
        add_header_above(c("Age group"=1, "Vaccine doses"=3)) %>%

    kable_styling(latex_options=c("HOLD_position"))

##```

##``{r casesnoriskchnums, echo=FALSE, message=FALSE}

knitr::kable(chnum.table, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="casesnoriskchnums",
             row.names=FALSE, 
			 align=c('r', 'l', 'r', 'r'), 
             caption="Severe cases and controls aged under 50 without risk conditions presenting from April to July 2021: ICD-10 chapters represented at least once in discharge diagnoses  from 5 years to 25 days before presentation date") %>%
    column_spec(2, width="7cm") %>%
    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
                         "Table filtered to retain only ICD-10 chapters represented by at least 20 individuals.",
                         "The most frequent ICD subchapter in cases was F00 (alcohol-related disorders): 7 cases."),
                             notation="none") %>%
    kable_styling(latex_options=c("HOLD_position"))

##```

##```{r casesnoriskbnfchnums, echo=FALSE, message=FALSE}

knitr::kable(bnfchnum.table, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="casesnoriskbnfchnums",
             row.names=FALSE, 
			 align=c('r', 'l', 'r', 'r'), 
             caption="Severe cases and controls aged under 50 without risk conditions presenting after March 2021: BNF chapters represented at least once in prescriptions dispensed during 240 to 15 days before presentation date") %>%
    column_spec(2, width="7cm") %>%
    add_footnote(label=c("Presentation dates from 1 April 2021 to 14 July 2021.",
                         "Table filtered to retain only BNF chapters represented by at least 20 individuals."),
                             notation="none") %>%
    kable_styling(latex_options=c("HOLD_position")) 
}
```

### 2021-08-29 vax2.Rmd
reinfections

--- 2021-04-16 vax.Rmd
@@
 freqs.shieldgroups <- as.data.frame.matrix(cbind(
-    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax,
-                        drop.reflevel=FALSE),
-    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax[vax14==1],
-                        drop.reflevel=FALSE)))[-1, ]
-rownames(freqs.shieldgroups) <- table.shieldgroups$effect[9:16]
-freqs.shieldgroups$"Crude odds ratio" <- oddsratio.table(freqs.shieldgroups)
+   with(cc.vax[CASE==0], paste.rowpercent(table(shield.group, vax14))),
+   with(cc.vax[CASE==1], paste.rowpercent(table(shield.group, vax14)))
+   ))
 
 freqs.hosp <- as.data.frame.matrix(cbind(
-    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax,
-                        drop.reflevel=FALSE),
-    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax[vax14==1],
-                        drop.reflevel=FALSE)))[-1, ]
+   with(cc.hosp.vax[CASE==0], paste.rowpercent(table(shieldelig.group, vax14))),
+   with(cc.hosp.vax[CASE==1], paste.rowpercent(table(shieldelig.group, vax14)))
+   ))
+
 rownames(freqs.hosp) <- table.hosp$effect[8:14]
-freqs.hosp$"Crude odds ratio" <- oddsratio.table(freqs.hosp)
 
@@
 
+# Key point summary
+
+- This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.
+- This analysis is based on data from 1 December 1st 2020 to 16 March 2021 and therefore covers the initial dose of vaccine with very few second doses in the shielded having been administered as yet  in this period.
+- Efficacy of a single vaccination dose in protecting against severe COVID-19 was found to be as high or higher in those eligible for shielding, considered as a whole,  as in those without risk conditions.
+- Examination of  vaccine efficacy among those eligible for shielding broken down by the reason for shielding showed that for  solid organ transplant recipients the number of cases is still too small for the efficacy of vaccination to be estimated.
+For all other groups there was a statistically significant reduction in hospitalisation for COVID-19 associated with  vaccination and that the magnitude of this reduction was similar to that in those who were not eligible for shielding.
+- Further data in larger populations are needed to evaluate vaccine efficacy in solid organ transplant recipients.
+- Until such data are available it should not be assumed that vaccination is efficacious in this group.
+
 # Background
+
 In Scotland to date COVD-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.
@@
 Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date.
-Ths assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.
+This assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.
 The numbers of cases and controls were too few too allow further subdivision of time since vaccination.
@@
 Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence.
-For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.
+<!---For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.
 The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$.
-The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.
+---> The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.
 
 As recent hospital admissions or being in hospital before presentation date might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.
-Because As another sensitivity analysis and to allow inference on vaccine efficacy in less frequent risk groups, the analysis was repeated with the case definition extended to include all cases that were hospitalized or fatal.
-For this analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".
+To allow inference on vaccine efficacy in less frequent risk groups, the analysis was also repeated with the case definition extended to include all cases that were hospitalized or fatal.
+For this latter analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".
 
@@
 This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population.
-By 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.
+From this we can infer that by 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.
 
@@
 
-Supplementary Table \ref{tab:freqsshieldgroups} shows the distribution of risk groups among  all severe cases and controls and then vaccinated cases and controls.
+Supplementary Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.
+.
 Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.
 These   tables show that the numbers of cases in some shielding subgroups, especially solid organ transplant recipients, are too small for rate ratios to be estimated reliability.
-Supplementary Table \ref{tab:mhdates} shows the observed and expected numbers of severe cases among vaccinated cases and unvaccinated individuals who eligible for shielding and not resident in care homes, by week of presentation.
-This shows that after week 6 of 2021 the number of severe cases was greatere in vaccinated than in unvaccinated individuals.
-This is to be expected as by this time more than 70% of those eligible for shielding had been vaccinated; it does not indicate that vaccination was ineffective.
+<!--delete the misleading crude odds ratios from these tables  alter the tables as discussed AND CHANGE THE TITLES ACCORDNGLY-->
 
@@
 
+For more detail on the computation of the vaccine  efficacy in a matched case control design see  the statistical note in the supplement and the associated  Supplementary Table \ref{tab:mhdates} <!--this comes out as s table  4 but there is no S table 3 please rename--> .
+
 # Conclusions
+
 Efficacy of a single vaccination dose in protecting against severe COVID-19 is as high or higher in those eligible for shielding as in those without risk conditions.
@@
 Thus they should remain vigilant  until they have received their second dose and background transmission rates are low.
+
 Also a  larger dataset than the  Scottish population is needed to estimate efficacy in solid organ transplant recipients where we have not been able to demonstrate efficacy.
+It is important that  such an analysis be conducted so that such persons can be advised appropriately as to their ongoing risk level.
+We note that such data  would also be important to understand the need for development of potential alternatives  to vaccines such as synthetic antibodies that are now being evaluated in clinical trials [@eli_lilly_and_company_phase_2021]
 
-The vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].
+The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].
 Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]
@@
 \onehalfspacing
+\newpage
 
@@
              row.names=TRUE,
+				 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
 				 align=c("l", "r", "r", "r", "r"),
              caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccination status") %>%
-    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
+    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
     add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
@@
              row.names=TRUE,
+  		 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
              align=c("l", "r", "r", "r", "r"),
              caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
-    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
+    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
     add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
@@
 ```
-
-Vaccine efficacy cannot be estimated from the crude odds ratios in Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp} because these totals do not take account of the matched case-control design.
-The matched design makes it possible to control for age, sex, general practice and crucially for calendar date where the proportion vaccinated and the rates of severe disease were changing day by day during the study period.
-The vaccine efficacy is estimated over matched sets, conditional on the marginal totals (cases vs controls, vaccinated versus unvaccinated) in each set.
-This section is intended to provide a more intuitive explanation of this.
-Table \ref{tab:mhdates} shows how information for a statistical test (the Mantel-Haenszel test) of  vaccine efficacy accumulated over matched sets (strata) of cases and controls during the study period.
-To keep the working simple, this table is restricted to those not resident in care homes so there is no covariate adjustment.
-
-Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:
-
- | | Case | Control | All |
- | --- | --- | --- | --- |
- Vaccinated | $a$ | $b$ | $V_1$ |
- Unvaccinated | $c$ | $d$ | $V_0$ |
- All | $C_1$ | $C_0$ | $N$ |
-
-The marginal totals of cases and controls are $C_1, C_0$ and of vaccinated and unvaccinated are $V_1, V_2$.
-Conditional on these marginal totals the expected number of vaccinated cases under the null hypothesis of no association is $C_1 V_1 / N$ and the variance of the expected is $C_1 C_0 V_1 V_0 / \left[ N^2 \left(N - 1 \right) \right]$.
-The observed numbers, the expected and the variance are summed over matched sets to calculate a test statistic as (observed - expected) divided by the square root of the variance.
 
@@
 ```
+
+\newpage
+
+## Statistical note
+Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp}  show the the vaccination status of  distribution of cases and controls by shielding stratum  to illustrate that for solid organ transplant recipients the numbers of cases are too few to estimate vaccine efficacy reliably.
+We have used a matched case-control design, in which each incident case was matched to up to 10 controls of the same age (one-year band), sex and GP practice who were alive on the day of presentation of the case, forming a matched set.
+Analyses of these matched sets are thus  inherently stratified by day.
+Most matched sets have 1 case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set would have 2 cases and up to 20 controls.
+
+<!---
+Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:
+
+ | | Case | Control | All |
+ | --- | --- | --- | --- |
+ Vaccinated | $a$ | $b$ | $V_1$ |
+ Unvaccinated | $c$ | $d$ | $V_0$ |
+ All | $C_1$ | $C_0$ | $N$ |
+--->
+Put simply, the vaccine efficacy is estimated as the most likely value of the odds ratio associated with vaccination given the observed values in each matched set, conditional on the marginal totals (total cases, total controls, total vaccinated, total unvaccinated.
+The log-likelihood of the odds ratio given the data is summed across all sets: matched sets in which all or none are vaccinated  do not contribute to the log-likelihood.
+So for example whilst it is obvious that sets that pertain to days before the vaccination programmes started cannot be informative it may be less obvious that sets pertaining to days when vaccination would be 100% complete are also uninformative.
+The calendar time period that is most informative is that when around 50% of the population   under study is vaccinated.
+
+When estimating the efficacy within a risk group, only cases and controls in that risk group are retained in the matched sets, and only those only matched sets that contain both unvaccinated and vaccinated individuals contribute to the likelihood.
+
+It follows from the above that crude tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the efficacy.
+This is especially true given that vaccination rates are rising sharply over the study period.
+
+Table \ref{tab:mhdates} may give the reader a more intuitive  sense of how the information for computing vaccine efficacy in the shielded evolved over the study period.
+It shows the weekly observed split of vaccinated versus unvaccinated cases among those in shielding versus the expected split conditional upon the marginal totals.
+To keep the working simple, this table is restricted to those 778 cases eligible for shielding who  were not resident in care homes so there is no covariate adjustment.
+The variance of the expected can be interpreted as the information about vaccine efficacy.

### 2021-08-29 vax3.Rmd
reinfections

--- 2021-08-29 vax2.Rmd
@@
 ---
-title: "Short report: efficacy of vaccination in those eligible for shielding in Scotland"
+title: "Brief Report: efficacy of vaccination in those eligible for shielding in Scotland"
 author:
@@
 ```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
+library(data.table)
+library(ggplot2)
+source("helperfunctions.R")
 
@@
 load(file=paste0(datadir, "cc.all.RData"))
+# with(cc.all[CASE==0 & SPECIMENDATE > as.Date("2021-02-28")], paste.rowpercent(table(dm.type, vax14.dose>0)))
 
@@
 
-keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group",
+keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group", "dm.type3",
                listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
 cc.vax <- na.omit(cc.severe[SPECIMENDATE > as.Date("2020-12-01"), ..keep.vars])
+cc.vax[, listedgr5 := listedgr3]
+cc.vax[listedgr3=="Moderate risk condition" & dm.type != "Not diabetic",
+       listedgr5 := dm.type3]
+cc.vax[, listedgr5 := factor(as.character(listedgr5),
+                     levels=c("No risk condition",
+                              "Type 1 diabetes", "Type 2 diabetes",
+                              "Moderate risk condition",
+                              "Eligible for shielding"))]
+
+table.dmriskgroups <- summary(clogit(data=cc.vax[care.home=="Independent"],
+         formula = CASE ~  listedgr5 / vax14 + strata(stratum)))$coefficients # matrix
+effect <- rownames(table.dmriskgroups)
+table.dmriskgroups <- data.table(effect=effect, table.dmriskgroups)
+
+table.dmriskgroups[, rateratio := or.ci(coef, `se(coef)`)]
+table.dmriskgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
+table.dmriskgroups <- table.dmriskgroups[5:9, .(effect, rateratio, pvalue)]
+table.dmriskgroups[, effect := gsub("listedgr5|vax14TRUE", "", effect)]
+
+freqs.riskgroups <- as.data.frame.matrix(cbind(
+    univariate.tabulate(varnames=c("listedgr5"), data=cc.vax,
+                        drop.reflevel=FALSE),
+    univariate.tabulate(varnames=c("listedgr5"), data=cc.vax[vax14==1],
+                        drop.reflevel=FALSE)))[-1, ]
 
@@
 table.riskgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14 + strata(stratum)))$coefficients
+
 effect <- rownames(table.riskgroups)
@@
 
-# Key point summary
+1836 words excluding Appendix
 
+\newpage
+
+# Abstract
 - This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.
@@
 - Efficacy of a single vaccination dose in protecting against severe COVID-19 was found to be as high or higher in those eligible for shielding, considered as a whole,  as in those without risk conditions.
-- Examination of  vaccine efficacy among those eligible for shielding broken down by the reason for shielding showed that for  solid organ transplant recipients the number of cases is still too small for the efficacy of vaccination to be estimated.
-For all other groups there was a statistically significant reduction in hospitalisation for COVID-19 associated with  vaccination and that the magnitude of this reduction was similar to that in those who were not eligible for shielding.
+- Examination of  vaccine efficacy among those eligible for shielding broken down by the reason for shielding showed that for  solid organ transplant recipients the number of cases was too small for the efficacy of vaccination to be estimated.
+For all other groups there was a statistically significant reduction in hospitalisation for COVID-19 associated with  vaccination and the magnitude of this reduction was similar to that in those who were not eligible for shielding.
 - Further data in larger populations are needed to evaluate vaccine efficacy in solid organ transplant recipients.
-- Until such data are available it should not be assumed that vaccination is efficacious in this group.
+Until such data are available it should not be assumed that vaccination is efficacious in this group.
 
-# Background
-
-In Scotland to date COVD-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.
-Both have been shown to be efficacious in clinical trials.
-Real-world observational analysis in a primary care-based study in Scotland up to 15 Feb 15 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].
-However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as restrictions on the general population are lifted.
-This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.
+# Introduction
+In Scotland up to March 2021 COVID-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.
+A primary-care based observational study up to 15 Feb 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].
+However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as population-wide restrictions are lifted.
+This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable in Scotland.
 
@@
 
-We used this matched case-control design to take advantage of extensive data linkages already established.
-As the design uses incidence density sampling, it controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise confound the association between vaccination and disease.
-
+We used this matched case-control design to take advantage of data linkages already established. 
 As previously, to minimise ascertainment bias we focused the analysis on  severe COVID-19,  defined as entry to critical care within 28 days of presentation or  fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).
@@
 
-The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories.
+The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk ategories.
+The  incidence density sampling, design controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise confound the association between vaccination and disease.
 Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence.
@@
 ---> The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.
+More explanation of the estimation of vaccine  efficacy in a matched case-control design are given in the Statistical Note in the Appendix.
 
-As recent hospital admissions or being in hospital before presentation date might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.
-To allow inference on vaccine efficacy in less frequent risk groups, the analysis was also repeated with the case definition extended to include all cases that were hospitalized or fatal.
+As hospital admission might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.
+To allow inference on vaccine efficacy in smaller groups, the analysis was repeated with the case definition extended to include all cases that were hospitalized or fatal.
 For this latter analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".
@@
 Supplementary Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.
-.
 Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.
-These   tables show that the numbers of cases in some shielding subgroups, especially solid organ transplant recipients, are too small for rate ratios to be estimated reliability.
-<!--delete the misleading crude odds ratios from these tables  alter the tables as discussed AND CHANGE THE TITLES ACCORDNGLY-->
+These tables show that the numbers of cases among solid organ transplant recipients, are too small for rate ratios to be estimated reliably.
 
@@
 Table \ref{tab:shieldgroups} gives a breakdown by shielding group.
-For the categories labelled as specific cancers, severe respiratory disease and additional conditions the numbers are large enough for the effect of vaccination on severe disease to be estimated reliably.
-The point estimates for the rate ratio for all risk groups were  lower than for those without risk conditions i.e.
-vaccine efficacy was at least as high,  with the exception of  solid organ transplant recipients,  those  with rare diseases and  those on immunosuppressants.
-For these groups the numbers of cases and controls were small (Supplementary Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.
+Point estimates for the rate ratio associated with vaccination were lower in all risk groups than in those without risk conditions,  with the exception of solid organ transplant recipients,  those  with rare diseases and those on immunosuppressants in whom the numbers of cases and were small (Supplementary Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.
+Among those in the "moderate risk conditions" category who had diabetes, the rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.dmriskgroups$rateratio[2])` and `r format.estci(table.dmriskgroups$rateratio[3])` in those with Type 1 and Type 2 diabetes respectively.
 
@@
 Table \ref{tab:shieldelig} shows the analysis extended to include all hospitalized cases, for which the numbers of cases are larger.
-This shows significant efficacy in all groups other than solid organ transplant recipients but in this group the numbers of cases and controls were small even with the broader case definition (Supplementary Table \ref{tab:freqshosp}).
+This shows significant efficacy in all groups other than solid organ transplant recipients but even with the broader case definition the numbers of cases and controls were too small  to evaluate efficacy.
+(Supplementary Table \ref{tab:freqshosp}).
 
@@
 
-For more detail on the computation of the vaccine  efficacy in a matched case control design see  the statistical note in the supplement and the associated  Supplementary Table \ref{tab:mhdates} <!--this comes out as s table  4 but there is no S table 3 please rename--> .
+# Conclusions
+Efficacy of a single vaccination dose in protecting against severe COVID-19 was as high or higher in those eligible for shielding as in those without risk conditions.
+This is reassuring in the light of other reports that vaccine-induced immune reactivity to SARS-CoV-2 is impaired in immunocompromised individuals [@monin-aldama_interim_2021; @deepak_glucocorticoids_2021].
+However for solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases is too small for the efficacy of vaccination to be estimated.
 
-# Conclusions
+These estimates of vaccine efficacy over all time periods from 14 days post-vaccination are similar to those estimated at 14-20 days post-vaccination for the total population of Scotland [@vasileiou_effectiveness_2021].
+Since we have previously shown that the shielding group overall have approximately six-fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1] efficacy of at least 82% (i.e.
+a rate ratio no more than 0.18) with a single dose would be required to lower risk to the level of unvaccinated persons of the same age and sex without risk conditions.
+Thus whilst those in shielding  can be reassured that the vaccine is showing efficacy they should also be advised that they remain at increased risk after a single dose (about twofold compared to those of the same age and sex without risk conditions given the vaccine-associated rate ratio of about 0.3), and should remain vigilant  until they have received their second dose and background transmission rates are low.
 
-Efficacy of a single vaccination dose in protecting against severe COVID-19 is as high or higher in those eligible for shielding as in those without risk conditions.
-For solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases is still too small for the efficacy of vaccination to be estimated.
+A study in a larger population is needed to estimate efficacy in solid organ transplant recipients.
+Such an analysis would lay the basis for advising people in this group about their ongoing risk level, and for assessing the need for development of alternatives to vaccines such as synthetic antibodies that are now being evaluated in clinical trials [@eli_lilly_and_company_phase_2021]
 
-The estimates of the rate ratio associated with vaccination or the vaccine efficacy here cannot be directly compared with the previous report for the total population of Scotland as they are summary estimates across all time periods from 14 days post-vaccination, rather than over 7-day intervals.
-Nonetheless the estimate is close to that reported at 14 days in the  previous report.
+The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between 4 and 12 weeks after the first dose [@public_health_england_covid-19_2021].
+Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]
 
-Since we have previously shown that  the shielding group overall have approximately 6 fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1] they would need a vaccine efficacy of  82% (i.e.
-a rate ratio of 0.18)  with a single dose to achieve a lowering of risk to the level of unvaccinated persons  without risk conditions.
-Thus whilst those in shielding  can be reassured that the vaccine is showing efficacy they should also be advised that they remain at increased risk  after a single dose (about twofold compared to those of the same age and sex without risk conditions given the vaccine-associated rate ratio of about 0.3).
-Thus they should remain vigilant  until they have received their second dose and background transmission rates are low.
+# Declarations
+## Ethics approval and information governance
+This study was approved under COVID-19 Rapid Data Protection Impact Assessment​ (DPIA) 20210023 that allows Public Health Scotland staff to link the datasets.
+Datasets were de-identified before analysis.
 
-Also a  larger dataset than the  Scottish population is needed to estimate efficacy in solid organ transplant recipients where we have not been able to demonstrate efficacy.
-It is important that  such an analysis be conducted so that such persons can be advised appropriately as to their ongoing risk level.
-We note that such data  would also be important to understand the need for development of potential alternatives  to vaccines such as synthetic antibodies that are now being evaluated in clinical trials [@eli_lilly_and_company_phase_2021]
+## Funding 
+No specific funding was received for this study.
 
-The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].
-Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]
+## Data Availability
+The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data.
+All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.
+
+## Competing interest
+The authors declare no competing interests
+
+## Registration
+The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).
+
+## Acknowledgements
+We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories.
 
@@
 
-# Supplementary Material
+# Appendix
 
@@
 --->
-Put simply, the vaccine efficacy is estimated as the most likely value of the odds ratio associated with vaccination given the observed values in each matched set, conditional on the marginal totals (total cases, total controls, total vaccinated, total unvaccinated.
-The log-likelihood of the odds ratio given the data is summed across all sets: matched sets in which all or none are vaccinated  do not contribute to the log-likelihood.
-So for example whilst it is obvious that sets that pertain to days before the vaccination programmes started cannot be informative it may be less obvious that sets pertaining to days when vaccination would be 100% complete are also uninformative.
-The calendar time period that is most informative is that when around 50% of the population   under study is vaccinated.
-
-When estimating the efficacy within a risk group, only cases and controls in that risk group are retained in the matched sets, and only those only matched sets that contain both unvaccinated and vaccinated individuals contribute to the likelihood.
+In brief vaccine efficacy is estimated as the most likely value of the odds ratio associated with vaccination given the observed values in each matched set, conditional on the marginal totals (total cases, total controls, total vaccinated, total unvaccinated).
+The log-likelihood of the odds ratio given the data is summed across all sets: matched sets in which all are vaccinated or none are vaccinated do not contribute information.
+When estimating the efficacy within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets
 
 It follows from the above that crude tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the efficacy.
-This is especially true given that vaccination rates are rising sharply over the study period.
-
-Table \ref{tab:mhdates} may give the reader a more intuitive  sense of how the information for computing vaccine efficacy in the shielded evolved over the study period.
+This is especially true given that vaccination rates were rising sharply over the study period and rates of severe disease were falling.
+Table \ref{tab:mhdates} may give the reader a more intuitive sense of how the information for computing vaccine efficacy in the shielded accumulated over the study period.
 It shows the weekly observed split of vaccinated versus unvaccinated cases among those in shielding versus the expected split conditional upon the marginal totals.
-To keep the working simple, this table is restricted to those 778 cases eligible for shielding who  were not resident in care homes so there is no covariate adjustment.
+To keep the illustration simple, this table is restricted to the 778 cases eligible for shielding who  were not resident in care homes so there is no covariate adjustment.
 The variance of the expected can be interpreted as the information about vaccine efficacy.
+As shown fewer of the cases were vaccinated and more were unvaccinated than expected if vaccination had no impact on disease rates.

### 2021-08-29 vax4.Rmd
reinfections

--- 2021-08-29 vax3.Rmd
@@
 ---
-title: "Brief Report: efficacy of vaccination in those eligible for shielding in Scotland"
+title: "Brief Report: efficacy of COVID-19 vaccination in individuals designated as clinically extremely vulnerable in Scotland"
 author:
@@
   \usepackage{longtable}
-  \usepackage{lscape}
-  \usepackage{pdfpages}
 output: 
@@
 load(file=paste0(datadir, "cc.all.RData"))
-# with(cc.all[CASE==0 & SPECIMENDATE > as.Date("2021-02-28")], paste.rowpercent(table(dm.type, vax14.dose>0)))
 
@@
                  limits=c(as.Date("2020-12-15"), as.Date("2021-03-02"))) +
-        labs(x=paste0("Presentation date: mid-point of 7-day window"),
+        labs(x=paste0("Sampling date: mid-point of 7-day window"),
          y="Proportion vaccinated among controls eligible for shielding")
 
-keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group", "dm.type3",
+keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group", "dm.type3", "fatalcase",
                listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
@@
                               SPECIMENDATE > as.Date("2020-12-01"), ..keep.hosp.vars])
-rm(cc.all)
+#rm(cc.all)
 
@@
 
-1836 words excluding Appendix
+2201 words, abstract 270 words
 
@@
 # Abstract
-- This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.
-- This analysis is based on data from 1 December 1st 2020 to 16 March 2021 and therefore covers the initial dose of vaccine with very few second doses in the shielded having been administered as yet  in this period.
-- Efficacy of a single vaccination dose in protecting against severe COVID-19 was found to be as high or higher in those eligible for shielding, considered as a whole,  as in those without risk conditions.
-- Examination of  vaccine efficacy among those eligible for shielding broken down by the reason for shielding showed that for  solid organ transplant recipients the number of cases was too small for the efficacy of vaccination to be estimated.
-For all other groups there was a statistically significant reduction in hospitalisation for COVID-19 associated with  vaccination and the magnitude of this reduction was similar to that in those who were not eligible for shielding.
-- Further data in larger populations are needed to evaluate vaccine efficacy in solid organ transplant recipients.
-Until such data are available it should not be assumed that vaccination is efficacious in this group.
+* _Objectives_ - The objective of this study was to estimate the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable in Scotland.
+* _Methods_ -  In a matched case-control design (REACT-SCOT), all `r nrow(cc.all[CASE==1 & SPECIMENDATE >= as.Date("2020-12-01")])` cases of COVID-19 in Scotland diagnosed from 1 December 2020 to 16 March 2021 were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0 & SPECIMENDATE >= as.Date("2020-12-01")])` controls from the general population.
+This dataset was linked to vaccination records and to the list of those identified as eligible for shielding.
+Severe COVID-19 was  defined by entry to critical care or fatal outcome.
+Rate ratios were estimated by conditional logistic regression.
+* _Results_ -  The rate ratio for severe COVID-19 associated with vaccination at least 14 days before was `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding, compared with `r format.estci(table.shieldgroups[9, 2])` in those without risk conditions.
+In a breakdown by shielding condition, the rate ratio for the broader category of hospitalization or fatal outcome ranged from `r format.estci(table.hosp[10, 2])` in those with specific cancers to `r format.estci(table.hosp[9, 2])` in solid organ transplants recipients, and  `r format.estci(table.hosp[13, 2])` in others on immunosuppressants (excluding solid organ transplant recipients)  compared with `r format.estci(table.hosp[8, 2])` in those ineligible for shielding.
+* _Conclusions_ -  These results, obtained when very few second doses of vaccine had been administered, are reassuring with respect to efficacy in clinically vulnerable individuals including immunocompromised individuals, but studies in larger populations are needed to estimate vaccine efficacy in solid organ transplant recipients.
 
 # Introduction
-In Scotland up to March 2021 COVID-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.
+In Scotland up to March 2021 only the ChAdOx1 nCoV-19 (Astra-Zeneca) and BNT162b2 mRNA (Pfizer-BioNtech) vaccines were used.
 A primary-care based observational study up to 15 Feb 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].
@@
 # Methods 
-This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to explore risk factors for severe COVID-19 in the population of Scotland.
+This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland.
+We used this study to take advantage of data linkages already established.
 The design has been described in detail previously [@mckeigue_rapid_2020].
-In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice were selected using the Community Health Index database.
+In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice and alive on the day of presentation of the case were selected using the Community Health Index database.
 COVID-19 cases are those with a positive nucleic acid test, or a hospital admission of death with COVID-19 codes.
 The dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.
-This analysis is based on all cases presenting between 1 December 1st 2020 to 16 March 2021.
+This analysis is based on all cases presenting from 1 December 2020 to 16 March 2021.
 
-We used this matched case-control design to take advantage of data linkages already established. 
 As previously, to minimise ascertainment bias we focused the analysis on  severe COVID-19,  defined as entry to critical care within 28 days of presentation or  fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).
-As described previously, cases and controls were grouped into one of three categories: no risk condition, risk condition designated previously by Public Health Scotland as moderate risk, eligible for shielding.
+As previously, cases and controls were grouped into one of three categories: no risk condition, risk condition designated previously by Public Health Scotland as moderate risk, eligible for shielding.
 For further analyses, the shielding category was subdivided as described previously into six categories: solid organ transplant, specific cancers, severe respiratory conditions, rare conditions, on immunosuppressants, additional conditions.
@@
 This assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.
-The numbers of cases and controls were too few too allow further subdivision of time since vaccination.
+The numbers of cases and controls were too few to allow further subdivision of time since vaccination.
 
-The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk ategories.
-The  incidence density sampling, design controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise confound the association between vaccination and disease.
+The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories.
+The strata in this conditional logistic regression model are matched sets: typically one case and up to ten controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.
+When estimating the efficacy within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.
+Matched sets in which all are vaccinated or none are vaccinated do not contribute to the conditional odds ratio.
+
+With this incidence density sampling design, the conditional odds ratio is the rate ratio and the efficacy of vaccination is 1 minus the rate ratio.
+The unconditional odds ratio calculated from summary tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the rate ratio.
+
+The design controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise strongly confound the association between vaccination and disease.
 Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence.
-<!---For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.
-The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$.
----> The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.
-More explanation of the estimation of vaccine  efficacy in a matched case-control design are given in the Statistical Note in the Appendix.
 
@@
 # Results
-Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date.
-This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population.
-From this we can infer that by 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.
+Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by seven-day sliding window of the date on which the controls were sampled.
+From this we can estimate that by 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.
 
-```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:plotvax}Proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date"}
+Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.
+`r cc.vax[CASE==1 & fatalcase, .N, by=listedgr3][2, 2]` of the `r cc.vax[CASE==1, .N, by=listedgr3][2, 2]` severe cases among those eligible for shielding were fatal.
+For comparison, `r cc.vax[CASE==1 & fatalcase, .N, by=listedgr3][3, 2]` of the `r cc.vax[CASE==1, .N, by=listedgr3][3, 2]` severe cases among with no risk condition were fatal.
 
-theme_set(theme_gray(base_size = 8))
-p.shieldvax
+Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls. 
+The rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.shieldgroups[9, 2])` in those with no risk condition, `r format.estci(table.shieldgroups[10, 2])` in those with moderate risk conditions and `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding.
+With adjustment for having been in hospital for more than 7 days before presentation and for any hospital discharge from 15 to 56 days before presentation, the corresponding rate ratios were `r format.estci(table.adjhosp$rateratio[6])`, `r format.estci(table.adjhosp$rateratio[7])` and `r format.estci(table.adjhosp$rateratio[8])`.
 
-```
-
-Supplementary Table \ref{tab:freqsshieldgroups} shows  the vaccination status of  severe cases and  their controls  by risk stratum.
-Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.
-These tables show that the numbers of cases among solid organ transplant recipients, are too small for rate ratios to be estimated reliably.
-
-The rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.riskgroups$rateratio[4])` in those with no risk condition, `r format.estci(table.riskgroups$rateratio[5])` in those with moderate risk conditions and `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding.
-With adjustment for having been in hospital for more than 7 days before presentation and for any hospital discharge from 15 to 56 days before presentation, the corresponding rate ratios were `r format.estci(table.adjhosp$rateratio[6])`, `r format.estci(table.adjhosp$rateratio[7])` and `r format.estci(table.adjhosp$rateratio[8])`.
 Table \ref{tab:shieldgroups} gives a breakdown by shielding group.
-Point estimates for the rate ratio associated with vaccination were lower in all risk groups than in those without risk conditions,  with the exception of solid organ transplant recipients,  those  with rare diseases and those on immunosuppressants in whom the numbers of cases and were small (Supplementary Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.
+Point estimates for the rate ratio associated with vaccination were lower in all risk groups than in those without risk conditions,  with the exception of solid organ transplant recipients,  those  with rare diseases and those on immunosuppressants in whom the numbers of cases were small (Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.
 Among those in the "moderate risk conditions" category who had diabetes, the rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.dmriskgroups$rateratio[2])` and `r format.estci(table.dmriskgroups$rateratio[3])` in those with Type 1 and Type 2 diabetes respectively.
 
-```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.shieldgroups[9:16, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="shieldgroups",
-             row.names=FALSE,
-             align=c("l", "r", "l"),
-             caption="Rate ratios for severe COVID-19 associated with vaccination, within each risk group",
-             col.names=c("Effect", 
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
-    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
-                         "Vaccine effects nested within each level of risk group",
-                         "Care home residence included as covariate", 
-                         "Presentation dates from 1 December 2020 to 16 March 2021",
-                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
-                 notation="none") %>%
-    column_spec(column=1, width="6cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
 Table \ref{tab:shieldelig} shows the analysis extended to include all hospitalized cases, for which the numbers of cases are larger.
-This shows significant efficacy in all groups other than solid organ transplant recipients but even with the broader case definition the numbers of cases and controls were too small  to evaluate efficacy.
-(Supplementary Table \ref{tab:freqshosp}).
-
-```{r hosp, echo=FALSE, warning=FALSE, message=FALSE}
-
-knitr::kable(table.hosp[8:14, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="shieldelig",
-             row.names=FALSE,
-             align=c("l", "r", "l"),
-             caption="Rate ratios for hospitalized or fatal COVID-19 associated with vaccination within each risk group",
-             col.names=c("Effect", "Rate ratio (95\\% CI)", 
-                         "\\ensuremath{p}-value")) %>%
-    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
-                         "Care home residence included as covariate", 
-                         "Presentation dates from 1 December 2020 to 16 March 2021",
-                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
-                 notation="none")  %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
+This shows evidence for efficacy in all groups except solid organ transplant recipients in whom even with this broader case definition the numbers of cases and controls were too small for efficacy to be estimated reliably.
 
@@
 
+As rates of severe disease in Scotland have now fallen to very low levels, the efficacy of two doses compared with one cannot be estimated accurately.
+The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between 4 and 12 weeks after the first dose [@public_health_england_covid-19_2021].
+Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021].
+
 A study in a larger population is needed to estimate efficacy in solid organ transplant recipients.
-Such an analysis would lay the basis for advising people in this group about their ongoing risk level, and for assessing the need for development of alternatives to vaccines such as synthetic antibodies that are now being evaluated in clinical trials [@eli_lilly_and_company_phase_2021]
-
-The Vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between 4 and 12 weeks after the first dose [@public_health_england_covid-19_2021].
-Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]
+Such an analysis would lay the basis for advising people in this group about their ongoing risk level, and for assessing the need for development of alternatives to vaccines such as synthetic antibodies that are now being evaluated in clinical trials  [@eli_lilly_and_company_phase_2021].
 
@@
 
-## Competing interest
+## Competing interests
 The authors declare no competing interests
@@
 
-\beginsupplement
-\onehalfspacing
 \newpage
 
-# Appendix
+```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:plotvax}Proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by 7-day window of date on which they were sampled (presentation date of the cases they were matched to)"}
 
-```{r freqsshieldgroups, echo=FALSE, warning=FALSE, message=FALSE}
+theme_set(theme_gray(base_size = 8))
+p.shieldvax
+
+```
+
+\newpage
+
+```{r freqsshieldgroups, echo=FALSE, message=FALSE}
 
@@
     column_spec(6, width="1cm") %>%
-    kable_styling(latex_options=c("hold_position"))
+    kable_styling(latex_options=c("HOLD_position"))
 
@@
 
-```{r mh.dates, echo=FALSE, warning=FALSE, message=FALSE}
+```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}
 
-knitr::kable(mh.dates[1:17, ], 
+knitr::kable(table.shieldgroups[9:16, ], 
              escape=FALSE, 
              booktabs=TRUE,
-			 label="mhdates",
+			 label="shieldgroups",
              row.names=FALSE,
-             align=c("l", "r", "r", "r", "r"),
-             digits=1,
-             caption="Vaccinated and unvaccinated severe cases among those eligible for shielding and not resident in care homes: observed and expected numbers by week") %>%
-    add_header_above(c(" "=1, "Vaccinated cases"=2, "Unvaccinated cases"=2)) %>% 
-    add_footnote(label=c("Expected numbers (under the null) are calculated in each stratum",
+             align=c("l", "r", "l"),
+             caption="Rate ratios for severe COVID-19 associated with vaccination, within each risk group",
+             col.names=c("Effect", 
+                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
+    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
+                         "Vaccine effects nested within each level of risk group",
+                         "Care home residence included as covariate", 
+                         "Presentation dates from 1 December 2020 to 16 March 2021",
                          "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                  notation="none") %>%
-    kable_styling(latex_options=c("hold_position"))
+    column_spec(column=1, width="6cm") %>%
+    kable_styling(latex_options=c("HOLD_position"))
 
@@
 
-\newpage
+```{r hosp, echo=FALSE, warning=FALSE, message=FALSE}
 
-## Statistical note
-Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp}  show the the vaccination status of  distribution of cases and controls by shielding stratum  to illustrate that for solid organ transplant recipients the numbers of cases are too few to estimate vaccine efficacy reliably.
-We have used a matched case-control design, in which each incident case was matched to up to 10 controls of the same age (one-year band), sex and GP practice who were alive on the day of presentation of the case, forming a matched set.
-Analyses of these matched sets are thus  inherently stratified by day.
-Most matched sets have 1 case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set would have 2 cases and up to 20 controls.
+knitr::kable(table.hosp[8:14, ], 
+             escape=FALSE, 
+             booktabs=TRUE,
+			 label="shieldelig",
+             row.names=FALSE,
+             align=c("l", "r", "l"),
+             caption="Rate ratios for hospitalized or fatal COVID-19 associated with vaccination within each risk group",
+             col.names=c("Effect", "Rate ratio (95\\% CI)", 
+                         "\\ensuremath{p}-value")) %>%
+    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
+                         "Care home residence included as covariate", 
+                         "Presentation dates from 1 December 2020 to 16 March 2021",
+                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
+                 notation="none")  %>%
+    kable_styling(latex_options=c("HOLD_position"))
 
-<!---
-Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$:
-
- | | Case | Control | All |
- | --- | --- | --- | --- |
- Vaccinated | $a$ | $b$ | $V_1$ |
- Unvaccinated | $c$ | $d$ | $V_0$ |
- All | $C_1$ | $C_0$ | $N$ |
---->
-In brief vaccine efficacy is estimated as the most likely value of the odds ratio associated with vaccination given the observed values in each matched set, conditional on the marginal totals (total cases, total controls, total vaccinated, total unvaccinated).
-The log-likelihood of the odds ratio given the data is summed across all sets: matched sets in which all are vaccinated or none are vaccinated do not contribute information.
-When estimating the efficacy within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets
-
-It follows from the above that crude tables of the vaccination status of cases and controls in each risk group cannot be used to estimate the efficacy.
-This is especially true given that vaccination rates were rising sharply over the study period and rates of severe disease were falling.
-Table \ref{tab:mhdates} may give the reader a more intuitive sense of how the information for computing vaccine efficacy in the shielded accumulated over the study period.
-It shows the weekly observed split of vaccinated versus unvaccinated cases among those in shielding versus the expected split conditional upon the marginal totals.
-To keep the illustration simple, this table is restricted to the 778 cases eligible for shielding who  were not resident in care homes so there is no covariate adjustment.
-The variance of the expected can be interpreted as the information about vaccine efficacy.
-As shown fewer of the cases were vaccinated and more were unvaccinated than expected if vaccination had no impact on disease rates.
+```

### 2021-08-29 vax7.Rmd
reinfections

---
title: "Brief Report: Efficacy of COVID-19 vaccination in individuals designated as clinically extremely vulnerable in Scotland"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Jen Bishop
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    corresponding: yes
    email: helen.colhoun@igmm.ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \usepackage[T1]{fontenc}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)
source("helperfunctions.R")

rollmean.dtN <- function(dt, k) {
    ## FIXME: add a by= argument
    N.dates <- dt[, .N, by=SPECIMENDATE]
    setkey(N.dates, SPECIMENDATE)
    all.dates <- data.table(date=seq(from=min(N.dates$SPECIMENDATE),
                                     to=max(N.dates$SPECIMENDATE),
                                     by=1))
    setkey(all.dates, date)

    ## add rows for missing dates by a left join
    all.dates <- N.dates[all.dates]
    all.dates[is.na(N), N := 0]
    return(data.table(date=dt$SPECIMENDATE,
                      avdaily=zoo::rollmean(dt$N, k=k, fill=NA)))
}

rollsum.datewin <- function(dt, k, datevar) {
    ## returns a table of rolling sums of width k centred on date
    ## dt is a dataset of totals (N) by date, which may have no rows for some dates in the
    ## range of interest
    ## add rows for missing dates by a left join of all.dates with dt
    all.dates <- data.table(date=seq(from=min(dt[[datevar]]),
                                     to=max(dt[[datevar]]),
                                     by=1))
    setkey(all.dates, date)
    # setkeyv(dt, datevar) can't set physical key here
    all.dates <- dt[all.dates]

    return(data.table(date=all.dates[[datevar]],
                      rsum=zoo::rollsum(all.dates[, N], k=k, fill=0, align="center")))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

format.estci <- function(x.ci) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    #x <- gsub("times", "\\\\times", x)
    return(x)
}

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

## vaccination analysis

datadir <- "data/2021-03-16/"
load(file=paste0(datadir, "cc.all.RData"))

winsize.casedates <- 3
## use by= to get rolling means by category

firstdate <- as.Date("2020-12-15")
lastdate <- as.Date("2021-03-01")

casedates.byriskgroup <- cc.severe[CASE==1, .N, by=.(SPECIMENDATE, listedgr3)]
setkey(casedates.byriskgroup, SPECIMENDATE)
casedates.byriskgroup <- casedates.byriskgroup[, rollmean.dtN(dt=.SD, k=winsize.casedates),
                                         by=listedgr3,
                                         .SDcols=c("SPECIMENDATE", "N")]

p.byriskgroup <- ggplot(data=casedates.byriskgroup[date <= lastdate],
                     aes(x=date,
                         y=avdaily, fill=listedgr3)) +
    geom_area() +
    scale_x_date(expand=c(0, 3), breaks = seq.Date(from = as.Date("2021-01-01"),
                                   to = as.Date("2021-03-01"),
                                   by = "month"),
                 labels=gsub("^0", "", 
                             format.Date(seq.Date(from = as.Date("2021-01-01"),
                                                  to = as.Date("2021-03-01"), by = "month"),
                                         "%d %b")
                 ),
                 limits=c(firstdate, lastdate)) +
    theme(legend.position = c(0.8, 0.8)) +
    theme(legend.title = element_blank()) +
    scale_fill_manual(values=c("grey", "blue", "red")) +
    ggtitle("(a) Daily severe cases in Scotland") + 
    xlab(paste0("Specimen date: mid-point of ", winsize.casedates, "-day window")) +
         ylab("Daily severe cases")

table.vaxshield <- with(cc.all[CASE==0], table(vax14.dose, shield.any))
table.vaxshield <- rbind(paste.colpercent(table.vaxshield, digits=2),
                         colSums(table.vaxshield))

shield.bydate <- cc.all[CASE==0 & SPECIMENDATE >= as.Date("2020-12-01") & shield.any==TRUE,
                           .N, by=c("SPECIMENDATE")]
setkey(shield.bydate, SPECIMENDATE)
shield.bydate <- rollsum.datewin(shield.bydate, 7, "SPECIMENDATE")
setnames(shield.bydate, "rsum", "all")
setkey(shield.bydate, date)

shieldvax.bydate <- cc.all[CASE==0 & SPECIMENDATE >= as.Date("2020-12-01") &
                           shield.any==TRUE & vax14==1,
                           .N,
                           by=c("SPECIMENDATE")] # 6193 vaccinated
setkey(shieldvax.bydate, SPECIMENDATE)
shieldvax.bydate <- rollsum.datewin(shieldvax.bydate, 7, "SPECIMENDATE")
setnames(shieldvax.bydate, "rsum", "vax")
setkey(shieldvax.bydate, date)       
shield1.bydate <- shieldvax.bydate[shield.bydate]
shield1.bydate[, prop.vax := vax / all]

p.shieldvax <-
    ggplot(data=shield1.bydate, aes(x=date, y=prop.vax)) +
    geom_line() +
    scale_y_continuous(breaks=c(0, 0.5, 1), limits=c(0, 1),
                       expand=c(0, 0)) + 
    scale_x_date(expand=c(0, 3), breaks = seq.Date(from = as.Date("2021-01-01"),
                                   to = as.Date("2021-03-01"),
                                   by = "month"),
                 labels=gsub("^0", "", 
                             format.Date(seq.Date(from = as.Date("2021-01-01"),
                                                  to =  as.Date("2021-03-01"),
                                                  by = "month"),
                                         "%d %b")
                             ),
                 limits=c(firstdate, lastdate)) +
    ggtitle("(b) Controls eligible for shielding: proportion vaccinated") + 
        labs(x=paste0("Sampling date: mid-point of 7-day window"),
         y=stringr::str_wrap("Proportion vaccinated", width=35))

keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group", "shieldelig.group", "dm.type3", "fatalcase",
               listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
cc.vax <- na.omit(cc.severe[SPECIMENDATE > as.Date("2020-12-01"), ..keep.vars])
cc.vax[, listedgr5 := listedgr3]
cc.vax[listedgr3=="Moderate risk condition" & dm.type != "Not diabetic",
       listedgr5 := dm.type3]
cc.vax[, listedgr5 := factor(as.character(listedgr5),
                     levels=c("No risk condition",
                              "Type 1 diabetes", "Type 2 diabetes",
                              "Moderate risk condition",
                              "Eligible for shielding"))]

table.dmriskgroups <- summary(clogit(data=cc.vax[care.home=="Independent"],
         formula = CASE ~  listedgr5 / vax14 + strata(stratum)))$coefficients # matrix
effect <- rownames(table.dmriskgroups)
table.dmriskgroups <- data.table(effect=effect, table.dmriskgroups)

table.dmriskgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.dmriskgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.dmriskgroups <- table.dmriskgroups[5:9, .(effect, rateratio, pvalue)]
table.dmriskgroups[, effect := gsub("listedgr5|vax14TRUE", "", effect)]

freqs.riskgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("listedgr5"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("listedgr5"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]

keep.hosp.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "care.home", "shield.any", "shieldelig.group", 
                    "vax14", "stratum")
cc.hosp.vax <- na.omit(cc.all[(casegroup=="A" | casegroup=="B") &
                              SPECIMENDATE > as.Date("2020-12-01"), ..keep.hosp.vars])
#rm(cc.all)

mh.strata <- cc.vax[care.home=="Independent" & shield.any==TRUE, 
                    .(SPECIMENDATE, stratum, CASE, vax14)]
mh.strata <- mh.strata[, c("a", "b", "c", "d") :=
                           list(CASE * as.integer(vax14),
                                (1 - CASE) * as.integer(vax14),
                                CASE * (1 - as.integer(vax14)),
                                (1 - CASE) * (1 - as.integer(vax14)))]
mh.strata <- mh.strata[, lapply(.SD, sum),
                       by=c("SPECIMENDATE", "stratum"), 
                       .SDcols=c("a", "b", "c", "d")]
mh.strata[, N := a + b + c + d]
mh.strata[, a.exp := (a + b) * (a + c) / N]
mh.strata[, c.exp := (a + c) * (c + d) / N]
mh.strata[, var.exp := (a + c) * (b + d) * (a + b) * (c + d) / (N^2 * (N - 1))]
mh.strata[is.nan(var.exp), var.exp := 0]
mh.strata[, week := paste(lubridate::year(SPECIMENDATE),
                          sprintf("%02d", lubridate::week(SPECIMENDATE)), sep=":")]

mh.dates <- mh.strata[, lapply(.SD, sum), by=week,
                      .SDcols=c("a", "a.exp", "c", "c.exp", "var.exp")]
setkey(mh.dates, week) 
mh.all <- data.table(week="All", mh.dates[, lapply(.SD, sum),
                                          .SDcols=c("a", "a.exp", "c", "c.exp", "var.exp")])
mh.dates <- rbind(mh.dates, mh.all)
colnames(mh.dates) <- c("Year:week", "Observed", "Expected", "Observed", "Expected", "Var(expected)")

table.dose <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14.dose + strata(stratum)))$coefficients

table.riskgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14 + strata(stratum)))$coefficients

effect <- rownames(table.riskgroups)
table.riskgroups <- as.data.table(table.riskgroups)
table.riskgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgroups <- table.riskgroups[, .(rateratio, pvalue)]
table.riskgroups <- data.table(effect=effect, # freqs.riskgroups,
                               table.riskgroups)
table.riskgroups[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.riskgroups[, effect := gsub("^listedgr3", "", effect)]
table.riskgroups[, effect := gsub("TRUE$", "", effect)]
table.riskgroups[, effect := gsub(":vax14$", "", effect)]
table.riskgroups[, effect := replace.names(effect)]

table.riskgroups2 <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + shield.any / vax14 + strata(stratum)))$coefficients

effect <- rownames(table.riskgroups2)
table.riskgroups2 <- as.data.table(table.riskgroups2)
table.riskgroups2[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgroups2[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgroups2 <- table.riskgroups2[, .(rateratio, pvalue)]
table.riskgroups2 <- data.table(effect=effect, # freqs.riskgroups2,
                               table.riskgroups2)
table.riskgroups2[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.riskgroups2[, effect := gsub("^listedgr3", "", effect)]
table.riskgroups2[, effect := gsub("TRUE$", "", effect)]
table.riskgroups2[, effect := gsub(":vax14$", "", effect)]
table.riskgroups2[, effect := replace.names(effect)]

table.adjhosp <- summary(clogit(data=cc.vax, 
                               formula = CASE ~ care.home + shield.any / vax14 +
                                   prob.hcai + discharge15to56 + strata(stratum)))$coefficients

table.adjhosp <- as.data.table(table.adjhosp)
table.adjhosp[, rateratio := or.ci(coef, `se(coef)`)]
table.adjhosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.adjhosp <- table.adjhosp[, .(rateratio, pvalue)]

table.shieldgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + shieldelig.group / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.shieldgroups)
table.shieldgroups <- as.data.table(table.shieldgroups)
table.shieldgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.shieldgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.shieldgroups <- table.shieldgroups[, .(rateratio, pvalue)]
table.shieldgroups <- data.table(effect=effect, # freqs.shieldgroups,
                               table.shieldgroups)
table.shieldgroups[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.shieldgroups[, effect := gsub("^shieldelig.group", "", effect)]
table.shieldgroups[, effect := gsub("TRUE$", "", effect)]
table.shieldgroups[, effect := gsub(":vax14$", "", effect)]
table.shieldgroups[, effect := replace.names(effect)]
table.shieldgroups <- as.data.frame(table.shieldgroups)

table.riskgroups2.hosp <- summary(clogit(data=cc.hosp.vax, formula = CASE ~ care.home + shield.any / vax14 + strata(stratum)))$coefficients

effect <- rownames(table.riskgroups2.hosp)
table.riskgroups2.hosp <- as.data.table(table.riskgroups2.hosp)
table.riskgroups2.hosp[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgroups2.hosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgroups2.hosp <- table.riskgroups2.hosp[, .(rateratio, pvalue)]
table.riskgroups2.hosp <- data.table(effect=effect, # freqs.riskgroups2.hosp,
                               table.riskgroups2.hosp)
table.riskgroups2.hosp[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.riskgroups2.hosp[, effect := gsub("^listedgr3", "", effect)]
table.riskgroups2.hosp[, effect := gsub("TRUE$", "", effect)]
table.riskgroups2.hosp[, effect := gsub(":vax14$", "", effect)]
table.riskgroups2.hosp[, effect := replace.names(effect)]

table.hosp <- summary(clogit(data=cc.hosp.vax, formula = CASE ~ care.home + shieldelig.group / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.hosp)
table.hosp <- as.data.table(table.hosp)
table.hosp[, rateratio := or.ci(coef, `se(coef)`)]
table.hosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.hosp <- table.hosp[, .(rateratio, pvalue)]
table.hosp <- data.table(effect=effect, # freqs.hosp,
                               table.hosp)
table.hosp[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.hosp[, effect := gsub("^shieldelig.group", "", effect)]
table.hosp[, effect := gsub("TRUE$", "", effect)]
table.hosp[, effect := gsub(":vax14$", "", effect)]
table.hosp[, effect := replace.names(effect)]
table.hosp <- as.data.frame(table.hosp)

freqs.riskgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "listedgr3"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "listedgr3"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]

oddsratio.table <- function(freqs.table) {
    freqs.table <- as.matrix(freqs.table)
    or <- numeric(nrow(freqs.table))
    for(i in 1:nrow(freqs.table)) {
        a <- as.numeric(gsub(" .+", "", freqs.table[i, ]))
        or[i] <- a[1] * a[4] / (a[2] * a[3])
    }
    return(round(or, 2))
}

freqs.shieldgroups <- as.data.frame.matrix(cbind(
   with(cc.vax[CASE==0], paste.rowpercent(table(shieldelig.group, vax14))),
   with(cc.vax[CASE==1], paste.rowpercent(table(shieldelig.group, vax14)))
   ))

freqs.hosp <- as.data.frame.matrix(cbind(
   with(cc.hosp.vax[CASE==0], paste.rowpercent(table(shieldelig.group, vax14))),
   with(cc.hosp.vax[CASE==1], paste.rowpercent(table(shieldelig.group, vax14)))
   ))

rownames(freqs.hosp) <- table.hosp$effect[8:14]

if(FALSE) {
cc.vax[`dm.typeOther/unknown type` == 1, `dm.typeType 2 diabetes` := 1]
cc.vax[, `dm.typeOther/unknown type` := NULL]

controls.vax <- cc.kept[CASE==0 & care.home=="Independent", .(AGE, COVID.age, vax_dose_1)]
controls.vax[, vaxnow := !is.na(vax_dose_1)]
controls.vax[, vaxnow := car::recode(vaxnow,
                                   recodes="FALSE='Not vaccinated';
                                            TRUE='At least one dose'",
                                   as.factor=TRUE,
                                   levels=c("Not vaccinated", "At least one dose"))]

theme_set(theme_gray(base_size = 16))
p.covidage <- ggplot(data=controls.vax, aes(x=AGE, y=COVID.age, color=vaxnow)) +
    geom_point(size=0.5, position="jitter") +
    scale_x_continuous(breaks=seq(20, 80, by=10), limits=c(15, 90)) + 
    scale_y_continuous(breaks=seq(20, 80, by=10), limits=c(15, 90)) +
    theme(legend.title = element_blank()) +
    theme(legend.position=c(0.15, 0.85)) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(title=paste("COVID age versus calendar age by vaccination status on",
                        format.Date(max(vacc$vax_dose_1), '%d %b %Y')),
         caption="Controls matched to severe cases, not resident in care homes. Limits of COVID age are 16 and 90 years",
         x="Calendar age (years)",
         y="COVID age (years)")
p.covidage

png("COVIDage_vaxstatus.png")
p.covidage
dev.off()
rm(p.covidage)

vax.firstdoses <- vacc[, .N, by=c("weekdose1", "vax_type_1")]
ggplot(data=vax.firstdoses, aes(x=weekdose1, y=N, color=vax_type_1)) +
    geom_line()

}

```
2069 words, abstract 280 words

\newpage

# Abstract
_Objectives_ - The objective of this study was to estimate the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable  in Scotland.

_Methods_ -  In a matched case-control design (REACT-SCOT), all `r nrow(cc.all[CASE==1 & SPECIMENDATE >= as.Date("2020-12-01")])` incident cases of COVID-19 in Scotland diagnosed from 1 December 2020 to 16 March 2021 were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0 & SPECIMENDATE >= as.Date("2020-12-01")])` controls from the general population. This was linked to national data on vaccinations and those designated as clinically extremely vulnerable and thus eligible for shielding support.  Severe COVID-19 was defined as cases with entry to critical care or fatal outcome.  Rate ratios associated with vaccination within risk groups were estimated by conditional logistic regression.

_Results_ -  The rate ratio for severe COVID-19 associated with vaccination at least 14 days before was `r format.estci(table.riskgroups2$rateratio[4])` in those eligible for shielding, compared with `r format.estci(table.riskgroups2$rateratio[3])` in those ineligible for shielding.  The rate ratio for hospitalised or fatal COVID-19 was `r format.estci(table.riskgroups2.hosp$rateratio[4])` in those  eligible and  `r format.estci(table.riskgroups2.hosp$rateratio[3])` in those not eligible for shielding.  Examined by specific shielding conditions, the rate ratio for hospitalized or fatal COVID-19  ranged from `r format.estci(table.hosp[10, 2])` in those with specific cancers to `r format.estci(table.hosp[9, 2])` in solid organ transplants recipients, and  `r format.estci(table.hosp[13, 2])` in others on immunosuppressants (excluding solid organ transplant recipients).

_Conclusions_ -  These results are reassuring with respect to efficacy in clinically vulnerable individuals including immunocompromised individuals, but studies in larger populations are needed to estimate vaccine efficacy in solid organ transplant recipients.

# Introduction
 A primary-care based observational study up to 15 Feb 2021  in Scotland found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_interim_2021]. Whether such high levels of vaccine efficacy pertain among those designated as clinically extremely vulnerable to COVID-19  is unclear.   In Scotland early in the epidemic a register of all those with certain conditions or on certain drugs that were likely to lead to  extreme vulnerability to COVID-19 was established based on electronic health care records and primary care physician reporting.  This register was used to determine eligibility for the national shielding support programme. We previously reported on the risk of severe COVID-19 in this shielding eligibility cohort compared to the rest of the population [@mckeigue_relation_2021-1]. If vaccine efficacy were estimated to be lower in clinically extremely vulnerable individuals than in the rest of the population, additional measures to shield these individuals would be needed until overall  transmission rates in the population are low.  Lower vaccine efficacy is possible since many such persons are immunocompromised or on immunosuppressants.  Therefore the aim of this study was to examine the efficacy of  COVID-19 vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable in Scotland.

# Methods 
This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland.  We used this study to take advantage of data linkages already established. The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice and alive on the day of presentation of the case were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission or death with COVID-19 ICD-10 codes. The dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.

As previously, to minimise ascertainment bias we pre-specified the primary outcome measure as severe COVID-19,  defined as cases with entry to critical care within 28 days of presentation or fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause) [@mckeigue_rapid_2020].  As previously, cases and controls were grouped into those eligible or not for the  shielding  programme.   For further analyses, the shielding category was subdivided as described previously  into six categories: solid organ transplant, specific cancers, severe respiratory conditions, other rare conditions, on immunosuppressants, additional conditions [@mckeigue_relation_2021-1].  Those ineligible for shielding were classified according to the presence of conditions designated by public health agencies as "moderate risk conditions" as described previously [@mckeigue_rapid_2020].  This analysis is based on cases presenting from 1 December 2020 to 16 March 2021.

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. This assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few to allow further subdivision of time since vaccination.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories. With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.  The strata in the conditional logistic regression model are matched sets: typically one case and up to ten controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.   We emphasize that the unconditional odds ratios calculated from summary tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].   The design controls not only for the matching factors of age, sex and general practice but also for calendar time.  Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes care home residence as a covariate.

Severe COVID-19 is strongly associated with recent hospital admission.  This might confound the association of severe COVID-19 with vaccination if hospital admission causes vaccination appointments to be missed or alternatively if hospital admission increases the probability of being vaccinated by a given date.  Therefore the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date, and for being an inpatient for at least 8 days before presentation date (the criterion for probable/definite hospital-acquired infection [@european_centre_for_disease_prevention_and_control_surveillance_2020]). We also reported the vaccine efficacy for severe COVID-19 separating out those with conditions designated previously by Public Health Scotland as moderate risk from the remainder not eligible for shielding.

To allow inference on vaccine efficacy in smaller groups, the analysis was repeated with the case definition extended to include all cases that were hospitalized or fatal.

# Results
Over the study period there were  `r cc.vax[CASE==1, .N, by=shield.any][2, 2]` severe cases among those eligible for shielding `r cc.vax[CASE==1 & fatalcase, .N, by=shield.any][2, 2]` of which were fatal and  `r cc.vax[CASE==1, .N, by=shield.any][1, 2]`  severe cases  among those not eligible for shielding of which  `r cc.vax[CASE==1 & fatalcase, .N, by=shield.any][1, 2]` were fatal. Figure \ref{fig:plotvax} shows the time course of daily severe cases in Scotland and vaccination among those eligible for shielding.  The proportion vaccinated at least 14 days before is estimated from those sampled as controls who were eligible for shielding.  From January to February 2020 the daily number of severe cases declined rapidly and the proportion vaccinated among those eligible for shielding rose from 2% to 87%.   The association between vaccination and severe disease is thus confounded by calendar time.  This confounding is controlled by the matching of cases and controls on calendar date but we reiterate that because of this the unconditional odds ratios calculated from Tables \ref{tab:freqsshieldgroups} and \ref{tab:freqshosp} cannot be used to estimate rate ratios [@greenland_need_1982].  Information about vaccine efficacy in those eligible for shielding is contributed by matched sets of cases and controls in the narrow time window during which incidence of severe COVID-19 was high and at least some of those eligible for shielding had been vaccinated at least 14 days before.  Up to 16 March 2021 the proportion who had received a second dose at least 14 days before the date of sampling was very low: `r table.vaxshield[3, 1]` of the `r table.vaxshield[4, 1]` controls ineligible for shielding and `r table.vaxshield[3, 2]` of the `r table.vaxshield[4, 2]` controls eligible for shielding.

Table \ref{tab:freqsshieldgroups} shows  the vaccination status of severe cases and  their controls  by risk stratum.  Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.  These tables illustrate that the data are very sparse for certain shielding groups especially solid organ transplant recipients.

 The rate ratio for severe COVID-19 associated with vaccination at least 14 days before was `r format.estci(table.riskgroups2$rateratio[4])` in those eligible for shielding, compared with `r format.estci(table.riskgroups2$rateratio[3])` in those ineligible for shielding.   When inpatient stay for at least 8 days before presentation and any hospital discharge from 15 to 56 days before presentation were included as covariates, the corresponding rate ratios were  `r format.estci(table.adjhosp$rateratio[6])` in those eligible and  `r format.estci(table.adjhosp$rateratio[5])` in those ineligible for shielding. Table \ref{tab:shieldgroups} gives a breakdown of the rate ratios for severe COVID-19  associated with vaccination by shielding condition.  Point estimates for the rate ratio associated with vaccination in risk groups were similar to the estimated rate ratio in those not eligible for shielding,  with the exception of solid organ transplant recipients and others on immunosuppressants in whom the numbers of cases were small (Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide.

When those ineligible for shielding were separated into those with moderate risk conditions  and those with no risk condition the rate ratios  for severe COVID-19 associated with vaccination were `r format.estci(table.riskgroups[5, 2])` and `r format.estci(table.riskgroups[4, 2])` respectively.  Among people with diabetes who were ineligible for shielding (and thus in the "moderate risk conditions" category), the rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.dmriskgroups$rateratio[2])` and `r format.estci(table.dmriskgroups$rateratio[3])` in those with Type 1 and Type 2 diabetes respectively.

For a secondary analysis with larger numbers of cases, the case definition was broadened to include all hospitalized or fatal cases.  The rate ratio for hospitalised or fatal COVID-19 was `r format.estci(table.riskgroups2.hosp$rateratio[4])` in those  eligible and  `r format.estci(table.riskgroups2.hosp$rateratio[3])` in those not eligible for shielding.   Table \ref{tab:shieldelig} shows that there was evidence for efficacy in all groups except solid organ transplant recipients in whom even with this broader case definition the numbers of cases and controls were too small for efficacy to be estimated reliably.

# Conclusions
Efficacy of a single vaccination dose in protecting against severe COVID-19 was as high or higher in those eligible for shielding as in those without risk conditions. This is reassuring in the light of reports that vaccine-induced antibody conversion is impaired in clinically vulnerable groups such as those with haematologic malignancies [@monin-aldama_interim_2021] and those on immunosuppressants for chronic inflammatory disease [@deepak_glucocorticoids_2021].  However for solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases in the Scottish population is too small for the efficacy of vaccination to be estimated.

These estimates of vaccine efficacy over all time periods from 14 days post-vaccination were similar to those estimated at 14-20 days post-vaccination for the total population of Scotland [@vasileiou_interim_2021].  As we have previously shown that the shielding group overall have approximately six-fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1], efficacy of at least 82% (a rate ratio no more than 0.18) with a single dose would be required to lower risk to the level of unvaccinated persons of the same age and sex without risk conditions.  Thus whilst these results give reassurance that the vaccine is showing efficacy overall in people with extreme clinical vulnerability  to COVID-19  such persons should also be advised that they remain at increased risk after a single dose (about twofold compared to those of the same age and sex without risk conditions, given the vaccine-associated rate ratio of about 0.3), and should remain vigilant  until they have received their second dose and background transmission rates  in their locale are low.  Although data published by Public Health Scotland show that by 29 May 2021 90% of those eligible for shielding had received two vaccine doses, there have been too few cases of severe disease after February 2020 for an updated estimate of the efficacy of two doses in those eligible for shielding to be reliable.

Strengths of  this analysis are the linkage to a pre-existing register of individuals designated as clinically extremely vulnerable, the ability to adjust for recent hospital exposure for which there was evidence of some confounding effect and the focus on severe COVID-19 rather than test-positive cases.  Limitations of our analysis are the small numbers with specific risk conditions, especially solid organ transplant recipients, and that the estimates of efficacy are effectively for first dose only as very few of those eligible for shielding had received a second dose of vaccine during the study period.   A study in a larger population is needed to estimate efficacy in solid organ transplant recipients. Such an analysis would lay the basis for advising people in this group about their ongoing risk level, and for assessing the need for development of alternatives to vaccines such as synthetic antibodies.

# Declarations
## Ethics approval and information governance
This study was approved under COVID-19 Rapid Data Protection Impact Assessment​ (DPIA) 20210023 that allows Public Health Scotland staff to link the datasets.  Datasets were de-identified before analysis.

## Funding 
No specific funding was received for this study. Professor Colhoun is supported by an endowed chair from the AXA Research Foundation.

## Data Availability
The component datasets used here are available via the Public Benefits Privacy Panel for Health and Social Care at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.

## Competing interests
The authors declare no competing interests

## Registration
The original protocol for the REACT-SCOT case-control study was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number EUPAS35558).

## Acknowledgements
We thank all staff in critical care units who submitted data to the SICSAG database, the Scottish Morbidity Record Data Team, the staff of the National Register of Scotland, the Public Health Scotland Terminology Services, the HPS COVID-19 Laboratory & Testing cell and the NHS Scotland Diagnostic Virology Laboratories.

# References
<div id="refs"></div>

\newpage

```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=5, fig.asp=1.2, fig.cap="\\label{fig:plotvax}(a) Daily severe cases by risk category; (b) Controls eligible for shielding: proportion vaccinated at least 14 days before, by 7-day window of date on which they were sampled as controls"}

theme_set(theme_gray(base_size = 10))

gridExtra::grid.arrange(p.byriskgroup, p.shieldvax, nrow=2, heights=3:2)

```

\newpage

```{r freqsshieldgroups, echo=FALSE, message=FALSE}

knitr::kable(freqs.shieldgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsshieldgroups",
             row.names=TRUE,
				 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
				 align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

```

```{r freqshosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
  		 col.names=c("Unvaccinated", "Vaccinated", "Unvaccinated", "Vaccinated"),
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "Controls"=2, "Cases"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position"))

```
\newpage

```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldgroups[8:14, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldgroups",
             row.names=FALSE,
             align=c("l", "r", "l"),
             caption="Rate ratios for severe COVID-19 associated with vaccination, within each risk group",
             col.names=c("Effect", 
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
                         "Vaccine effects nested within each level of risk group",
                         "Care home residence included as covariate", 
                         "Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(column=1, width="6cm") %>%
    kable_styling(latex_options=c("hold_position"))

```

```{r hosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.hosp[8:14, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldelig",
             row.names=FALSE,
             align=c("l", "r", "l"),
             caption="Rate ratios for hospitalized or fatal COVID-19 associated with vaccination within each risk group",
             col.names=c("Effect", "Rate ratio (95\\% CI)", 
                         "\\ensuremath{p}-value")) %>%
    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
                         "Care home residence included as covariate", 
                         "Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    kable_styling(latex_options=c("hold_position"))

```

### 2021-08-29 vaxshield230821.Rmd
reinfections

---
title: "Variation in efficacy of COVID-19 vaccination against severe COVID-19 by clinical vulnerability status: the REACT-SCOT case-control study"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
    corresponding: yes
    email: paul.mckeigue@ed.ac.uk
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}} 
    affiliation: HPS 
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}} 
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: for the PHS COVID-19 Epidemiology and Research Cell
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. 
- code: IGMM
  address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. 
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ.

header-includes:
  \usepackage{newunicodechar}
  \usepackage[T1]{fontenc}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./f1000.csl # ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan 
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

```
\newpage

# Abstract

**Objectives** - To determine whether COVID-19 efficacy varies with clinical risk category and to investigate risk factors for severe COVID-19 in those who have received two doses of vaccine.

**Design** - Matched case-control study (REACT-SCOT).

**Setting** - Population of Scotland from 1 December 2020 to 28 July 2021.

**Main outcome measure** - Severe COVID-19, defined as cases with entry to critical care or  fatal outcome.

**Results** - Efficacy against severe COVID-19 of two doses of vaccine was `r pctefficacy.or.ci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without designated risk conditions, `r pctefficacy.or.ci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, but only `r pctefficacy.or.ci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those designated as clinically extremely vulnerable (CEV) and eligible for shielding.  Efficacy of two doses against severe COVID-19 in the CEV group did not differ between the AstraZeneca vaccine (`r pctefficacy.or.ci(table.vaxgr[effect=="Eligible for shielding: 2 doses AZ vaccine", rateratio])`) and the mRNA vaccines (`r pctefficacy.or.ci(table.vaxgr[effect=="Eligible for shielding: 2 doses mRNA vaccine", rateratio])`).  Of the 221 cases of severe COVID-19 in double-vaccinated individuals, 44% had moderate risk conditions and 43% were CEV.  In the doubly-vaccinated CEV group, the rate ratio for severe disease (with no risk condition as reference category) was highest in solid organ transplants at `r astext.or.ci(table.severe.vax2["Solid organ transplant", 3])` but even in this subgroup the absolute risk of severe COVID-19 in the double-vaccinated was low (12 cases in 14386 person-months of follow-up).

**Conclusions** - 
Two doses of vaccine protect against severe COVID-19 in clinically extremely vulnerable persons but the efficacy is somewhat lower than in those who are not CEV.  These results support a policy of offering booster doses to doubly-vaccinated vulnerable individuals, who can be identified from their health records.

\newpage

# Introduction
The REACT-SCOT matched case-control study was established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland [@mckeigue_rapid_2020].  Using this framework, we have reported in detail on the relation of severe COVID-19 to risk factors including designated moderate risk conditions, drug prescribing and those designated as clinically extremely vulnerable (CEV) and therefore eligible for shielding [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021b].

We have previously reported an initial comparison of the efficacy of vaccination against severe COVID-19 between CEV people and those with moderate risk conditions or no risk conditions, based on data up to 16 March 2021 [@mckeigue_efficacy_2021].   At this time few individuals had received a second dose of vaccine, but since then most CEV individuals have received two doses. This report updates and extends the earlier analyses, with the following objectives:

1. To determine whether vaccine efficacy varies with risk category and CEV status now that more data, including exposure to two doses of vaccine, are available  and now that the delta variant is the dominant variant in Scotland.

2. To investigate risk factors for severe or hospitalised COVID-19 in those who have received two doses of vaccine.

In an accompanying report we have investigated how efficacy of vaccination against severe COVID-19 changed with the appearance of the Delta variant and how  efficacy of two doses varies with time since second dose.

# Methods 
We used the REACT-SCOT study to take advantage of data linkages already established. The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched for one-year age, sex and primary care practice and alive on the day of presentation of the case that they were matched to were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission or death with COVID-19 ICD-10 codes. The REACT-SCOT case-control dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.  This analysis is based on cases presenting from 1 December 2020 to 14 July 2021, so that we have at least 14 days of follow-up for each case.  This allows case severity status (hospitalisation, entry to critical care or death) to be assigned.

## Classification of risk categories
As previously [@mckeigue_rapid_2020], to minimise ascertainment bias we pre-specified the primary outcome measure as severe COVID-19,  defined as diagnosed cases with entry to critical care within 28 days of presentation or fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).  Cases and controls were classified into three broad risk categories: no risk condition; at least one of the moderate risk conditions designated by public health agencies [@mckeigue_rapid_2020]; or eligible for shielding [@mckeigue_relation_2021b].  For further analyses, the shielding category was subdivided as described previously  into six categories: solid organ transplant, specific cancers, severe respiratory conditions, other rare conditions, on immunosuppressants, and additional conditions [@mckeigue_relation_2021b].   This corresponds to the list used by Public Health Scotland  [@health_protection_scotland_search_2020], after combining the small numbers in the group "pregnant with heart disease with the "other conditions" category.  For additional analyses the category "specific cancers" was split to allocate cancers of blood-forming organs (ICD-10 codes C81-C88, C90-C96) to a separate category.

## Statistical analysis
Vaccination status was coded as the number of doses administered at least 14 days before presentation date. Vaccine doses administered less than 14 days before presentation date are ignored.  For estimates of effect by risk category, where the numbers of events are large, vaccination status was modelled as a factor comparing 1 dose or 2 doses with 0 doses as reference category.  For estimates of effect within shielding eligibility groups where the numbers of events in each category are smaller, vaccination status was modelled as a numeric variable taking values \ensuremath{0, \frac{1}{2}, 1}  so that the effect parameter represents the rate ratio associated with 2 doses assuming the effect of 0, 1, 2 doses to be linear on the scale of the log rate ratio.

Covariates included in the models were those that have been previously identified as strong predictors of severe disease in this population: care home residence, number of adults in household, number of non-cardiovascular drug classes dispensed and recent hospital stay [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021b].  For care home residents the number of of adults in the household was coded as 1 to ensure that these two variables are not confounded.  The number of non-cardiovascular drug classes was calculated as the number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.  The number of hospital diagnoses was calculated as the number of distinct ICD-10 chapters represented at least once in hospital discharge records between 25 days and 5 years before presentation date. Recent hospital stay was defined as any in-patient stay from 5 to 14 days before presentation date.

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories. With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.   The design controls not only for the matching factors of age, sex and general practice but also for calendar time.  The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].

# Cohort analysis of the shielding list
The case-control study estimates only rate ratios.  To investigate  how the absolute rates of severe disease in those listed as clinically extremely vulnerable  have changed with the vaccination programme rollout, we  also undertook a cohort analysis of all individuals who had ever been on the shielding/ CEV list.  A Poisson regression model was fitted to the cohort formatted with one observation per 28-day person-time interval, and individuals censored at first diagnosis of COVID-19.  Event status was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, baseline age, and shielding eligibility group.  The baseline hazard rate was modelled as a natural spline function of calendar time with 6 degrees of freedom. Similar models were fitted with event status defined as any diagnosed case.

# Results
## Relation of vaccine efficacy to risk conditions
As shown in Table \ref{tab:freqsvaxgr} between 1 December 2020 to 14 July 2021  there were 4992 cases of severe COVID-19 in the total population of Scotland of whom 27% had no designated risk condition, 53%  had a moderate risk condition and 20%  were CEV.  The distributions in cases and controls  of other risk factors included as covariates in the models are shown for completeness.   Overall 86% of severe cases arising in this period were not yet vaccinated.   Table \ref{tab:hospfreqsvaxgr} shows the same tabulation with case definition broadened to include all hospitalised cases and their matched controls.

Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories  from the conditional logistic regression model.  The  model included risk categories (no risk conditions as reference category),  care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates. The rate ratio associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding (i.e. CEV).  On the scale of efficacy, these estimates are equivalent to 94% in those without risk conditions, 91% in those with risk conditions, and 68% in those eligible for shielding.  <!---To test formally for an effect of risk category on vaccine efficacy, a model was fitted with interaction terms for  risk category \ensuremath{\times} times vaccine dose.  In those with 2 doses, the effect (on the scale of rate ratio) of eligibility for shielding (with no risk condition as reference category) on the rate ratio associated with 2 doses of vaccine (0 doses as reference category) was `r format.estci(vaxriskgr.interaction)`.  ---> 
Table \ref{tab:hosprateratiosriskgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls: the efficacy of 2 doses  against hospitalisation  was also lower in those eligible for shielding than in the other two risk categories.

As the estimates of efficacy for a single dose in Table \ref{tab:rateratiosriskgroups} for this entire period  of 1 Dec 2020 up to 14 July 2021 differ from those we reported  previously based only on data up to 16 March 2021 [@mckeigue_efficacy_2021], we examined whether our earlier results were reproduced in this updated dataset.  Restricting the dataset to exclude cases and controls sampled after 16 March 2021, the point estimates of the rate ratios associated with one dose of vaccine in the same model for those with no risk conditions, moderate risk conditions and eligible for shielding were 0.28, 0.25 and 0.31 respectively,  very similar to the estimates previously reported with slightly different covariate adjustment and less complete follow-up of outcome in cases after presentation date [@mckeigue_efficacy_2021]

<!---To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower vaccine efficacy of vaccines in the CEV group, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.  With this restricted case definition, the rate ratio associated with two doses of vaccine was `r format.estci(table.nonfatalorucod[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.nonfatalorucod[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.nonfatalorucod[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.  --->

Table \ref{tab:rateratiosshieldgroups} shows rate ratios associated with vaccination by CEV subgroup.  Because the numbers in some CEV subgroups are small, the rate ratios are estimated as linear effects of dose rather than with separate terms for 1 or 2 doses as in analyses where the CEV group is not subdivided.  Rate ratios were higher (and thus efficacy was lower) in all shielding eligibility subgroups than in those without risk conditions or with moderate risk conditions,  but the confidence intervals were too wide for comparisons of efficacy between CEV subgroups.  Table \ref{tab:hosprateratiosshieldgroups} shows  shows rate ratios  for the first and second dose with the case definition broadened to include all hospitalised cases and their matched controls. In this table the highest rate ratio (lowest efficacy) is in solid organ transplant recipients but again the confidence intervals are too wide for comparisons of efficacy between CEV subgroups.

We examined whether there was any difference in efficacy by class of vaccine and whether any such differences varied by risk category.  Table \ref{tab:rateratiosvaxgroups} shows rate ratios associated with 1 and 2 doses of each class of vaccine (with 0 doses as reference category) in each of the three broad risk categories.  Table \ref{tab:hosprateratiosvaxgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.  Although the confidence intervals for the rate ratios associated with 1 or 2 doses of each vaccine product  in the CEV group are wide, it is clear that the efficacy of 2 doses is lower in the CEV group than in those with moderate risk conditions or no risk conditions.  There was no evidence that efficacy against severe COVID-19 of 2 doses in CEV individuals was lower for the AZ than for the mRNA vaccine, and only weak evidence for lower efficacy of the AZ vaccine against the broader category of hospitalized or fatal COVID-19:  rate ratio 1.51 (95% CI 0.66 to 1.03, _p_=0.04) times higher for the AZ vaccine than for the Pfizer vaccine.

## Risk factors for severe COVID-19 in the double-vaccinated
Table \ref{tab:severevax2} shows risk factors for severe COVID-19 in cases and controls who had received 2 doses of vaccine at least 14 days before presentation date.  Table \ref{tab:hospvax2} shows the same analysis with case definition broadened to include hospitalized cases and matched controls who had received 2 doses of vaccine. Only 13% of double-vaccinated severe cases had no designated risk condition: 44% had at least one moderate risk condition and 43% were eligible for shielding.    Of the variables that we have previously reported as risk factors for severe COVID-19 in the general population, care home residence and socioeconomic deprivation were not associated with severe disease in the double-vaccinated.  In this group the risk factors for severe disease were designated risk conditions, other indicators of co-morbidity including numbers of hospital diagnoses and drug classes prescribed and transmission-related factors including number of adults in household and recent hospital stay.  A risk score for hospitalised or fatal COVID-19 in the double-vaccinated calculated from the multivariable conditional logistic regression model in Table \ref{tab:hospvax2} was able to classify cases and noncases with a C-statistic of 0.80.  Among the CEV groups, the highest rate ratio for severe disease was solid organ transplant recipients: with no risk conditions as reference category, the unadjusted rate ratio associated with receipt of a solid organ transplant among the double-vaccinated was `r format.estci(table.severe.vax2["Solid organ transplant", 3])` for severe disease and `r format.estci(table.hosp.vax2["Solid organ transplant", 3])` for hospitalisation.

## Shielding cohort
By the date of the latest extract of the shielding list in July 2021, 8230 (4%) of the 202510 individuals who had ever been on the shielding list had been diagnosed with COVID-19, and 18893 (9%) had been removed from the list for other reasons.  Fig \ref{fig:caseseverity} shows the fitted incidence per month of any diagnosis and of severe cases in Poisson regression models fitted to the entire cohort.   The incidence of any diagnosed COVID-19 in the shielding cohort  fell from 1 December to a nadir in late Aprll.  Despite the incidence of any diagnosed COVID-19 rising steeply from 1 May onwards, the incidence of severe COVID-19 rose only  slightly .

Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category. Table \ref{tab:cohortseverevax2} shows the incidence of severe COVID-19 in doubly-vaccinated CEV individuals, by subgroup.  The incidence in solid organ transplant recipients was 0.8 per 1000 per month (12 cases in 14386 person-months of follow-up).

# Discussion
## Statement of principal findings
1. Although the efficacy of two doses of vaccine against severe COVID-19 in those without risk conditions remains high (at least 90%), it is now clear that efficacy in clinically extremely vulnerable individuals is somewhat lower.

2. Among the double-vaccinated, 87% of severe cases and 77% of hospitalised cases have designated risk conditions or are eligible for shielding.

2. CEV individuals remain at high risk of severe disease even after two doses of vaccine: for solid organ transplant recipients that rate ratio is approximately 70-fold for severe disease and 30-fold for hospitalisation in comparison with double-vaccinated individuals of the same age and sex without risk conditions. However even in this group the absolute risk of severe disease in the double-vaccinated is less than 1 in 1000 per month.

The explanation for why the differences in efficacy between the CEV group and those without risk conditions have become apparent only after March 2021 is not clear.  We have shown in an accompanying paper that the replacement of the Alpha variant by the Delta variant over a few weeks in May 2021 was accompanied by a decrease in efficacy of vaccination against severe COVID-19, though in the general population this decrease was only temporary.

## Strengths and limitations
Strengths of this study are the elimination of calendar time effects by the matched case-control design, the comprehensive linkage to e-health records and the focus on severe cases as main outcome measure. Hospitalised cases may include some test-positive individuals whose admission or continued stay in hospital was for another underlying diagnosis: this in turn may lead to underestimation of vaccine efficacy against COVID-19, especially in risk groups with comorbidities.  Such misclassification is less likely to occur with the narrow definition of severe COVID-19 used in the REACT-SCOT study.

The main limitation  is that no data on hospital prescribing are available and thus we cannot study the relation of risk to different classes of immunosuppressant drugs.   Most immunosuppressants and drugs for cancer are prescribed only in hospital and records are not held in electronic form.  The national shielding list does not hold any information about drug therapy, and no information about diagnoses other than the seven broad categories listed as eligible for shielding.

## Relation to other studies
A recent report based on primary care data covering 7.2 million people curated by the Royal College of General Practitioners Research and Surveillance Centre examined vaccine efficacy against symptomatic COVID-19 for which a medical consultation was recorded  [@whitaker_pfizerbiontech_2021].  For those who had been advised to shield, the efficacy after two doses of any vaccine was estimated as 92% in those aged less than 65 years and 84% in those aged 65 years and over.  For the "immunocompromised" group the estimated vaccine efficacy was somewhat lower at 74%, with a wide confidence interval. No separate analysis of vaccine efficacy in solid organ transplant recipients was reported.

## Policy implications. 
As in the general population, two doses are more effective than one in CEV individuals. It is thus plausible that an additional dose might give more protection to double-vaccinated individuals in this group. For identifying those among the double-vaccinated who are at high risk of severe disease and should be offered a booster dose, a multivariable risk score based on electronic health records could easily be constructed, but a simple risk classifier based on presence of designated risk conditions or CEV status would cover 87% of those at risk of severe disease.

# Declarations
## Ethics approval
This study was performed within Public Health Scotland as part of its statutory duty to monitor and investigate public health problems.  Under the [UK Policy Framework for Health and Social Care Research](https://www.hra.nhs.uk/planning-and-improving-research/policies-standards-legislation/uk-policy-framework-health-social-care-research/) set out by the NHS Health Research Authority, this does not fall within the definition of research and ethical review is not required.  Individual consent is not required for Public Health Scotland staff to process personal data to perform specific tasks in the public interest that fall within its statutory role.  The statutory basis for this is set out in Public Health Scotland's [privacy notice](https://www.publichealthscotland.scot/our-privacy-notice/personal-data-processing/).

# References
<div id="refs"></div>

\newpage

# Figures

```{r caseseverity, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:caseseverity} Poisson regression model for 60-day incidence of any case and severe case in the shielding cohort"}

theme_set(theme_gray(base_size = 10))

p.anycase

```

# Tables

```{r freqsvaxgroups, echo=FALSE, message=FALSE}

knitr::kable(freqs.vaxgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsvaxgr",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccine product") %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Risk/shielding eligibility group", 6, 13, bold=FALSE) %>% 
    kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
    column_spec(1, width="5cm") %>%
    column_spec(2, width="4cm") %>%
    column_spec(3, width="3cm") %>%
   kable_styling(latex_options=c("HOLD_position"))

```

```{r rateratiosriskgroups, echo=FALSE, message=FALSE}

knitr::kable(table.riskgr[7:12, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosriskgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 1, 2, bold=TRUE) %>% 
    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 3, 4, bold=TRUE)  %>% 
    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 5, 6, bold=TRUE)  %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                         "Vaccination status is doses administered at least 14 days fbefore presentation date"), 
                 notation="none") %>%
    column_spec(1, width="6cm") %>%
    column_spec(2, width="4cm") %>%
    column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r rateratiosshieldgroups, echo=FALSE, message=FALSE}

table.shieldgr[, effect := gsub("vax", "vaccine", effect)] 
knitr::kable(table.shieldgr[12:19, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1/2, 1") %>%
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r rateratiosvaxgroups, echo=FALSE, message=FALSE}

startrow <- 1
knitr::kable(table.vaxgr[7:18,], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="rateratiosvaxgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for severe COVID-19 within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition", startrow, startrow+3, bold=TRUE) %>% 
    kableExtra::group_rows("Moderate risk condition", startrow+4, startrow+7, bold=TRUE)  %>% 
    kableExtra::group_rows("Clinically extremely vulnerable", startrow+8, startrow+11, bold=TRUE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Adjusted for care home residence, number of adults in household, number of drug classes dispensed, recent hospital stay and risk category.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r severevax2, echo=FALSE, message=FALSE}

knitr::kable(table.severe.vax2, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="severevax2",
             row.names=TRUE, 
            caption="Risk factors for severe COVID-19 in those who had received 2 doses of vaccine at least 14 days before", 
             col.names=c(clean.header(colnames(table.severe.vax2)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="4cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="3cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    kableExtra::group_rows("Shielding eligibility category", 9, 14, bold=FALSE) %>%
   add_footnote(label=c("Severe COVID-19 is defined by entry to critical care or fatal outcome", 
                         "Percentages are column percentages for each variable",
                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
                         "Univariate rate ratios are for models with a single covariate",
                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
                 notation="none") %>%
    kable_styling(latex_options=c("scale_down", "HOLD_position"))

```

# Supplementary Material
\beginsupplement

## Supplementary Figures

```{r plotcases, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotcases} Daily severe cases by risk category"}

theme_set(theme_gray(base_size = 10))

p.byelig + theme(legend.title = element_blank())

```

```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window. For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence and recent hospital stay. Line thickness is inversely proportional to standard error"}

theme_set(theme_gray(base_size = 10))
# p.rateratio

```

## Supplementary Tables

```{r hospfreqsvaxgroups, echo=FALSE, message=FALSE}

knitr::kable(freqs.hosp.vaxgr, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hospfreqsvaxgr",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccine product") %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
                 notation="none") %>%
    kableExtra::group_rows("Risk/shielding eligibility group", 6, 12, bold=FALSE) %>% 
    kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
    column_spec(1, width="5cm") %>%
    column_spec(2, width="4cm") %>%
    column_spec(3, width="3cm") %>%
   kable_styling(latex_options=c("HOLD_position"))

```

```{r hosprateratiosriskgroups, echo=FALSE, message=FALSE}

knitr::kable(table.hosp.riskgr[7:12, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosriskgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition, unvaccinated within this group as reference category", 1, 2, bold=FALSE) %>% 
    kableExtra::group_rows("Moderate risk condition, unvaccinated within this group as reference category", 3, 4, bold=FALSE)  %>% 
    kableExtra::group_rows("Clinically extremely vulnerable, unvaccinated within this group as reference category", 5,  6, bold=FALSE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="6cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r hosprateratiosshieldgroups, echo=FALSE, message=FALSE}

table.hosp.shieldgr[, effect := gsub("vax", "vaccine", effect)] 
knitr::kable(table.hosp.shieldgr[12:19, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosshieldgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
                    "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r hosprateratiosvaxgroups, echo=FALSE, message=FALSE}

startrow <- 1
knitr::kable(table.hosp.vaxgr[7:18,], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosprateratiosvaxgroups",
             row.names=FALSE,
             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
			 align=c('l', 'c', 'c'), 
             caption="Rate ratios for hospitalized or fatal COVID-19 within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
    kableExtra::group_rows("No risk condition", startrow, startrow+3, bold=TRUE) %>% 
    kableExtra::group_rows("Moderate risk condition", startrow+4, startrow+7, bold=TRUE)  %>% 
    kableExtra::group_rows("Clinically extremely vulnerable", startrow+8, startrow+11, bold=TRUE)  %>% 
add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
                     "Adjusted for care home residence, number of adults in household, number of drug classes dispensed, recent hospital stay and risk category.", 
                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
             notation="none") %>%
    column_spec(1, width="8cm") %>%
column_spec(2, width="4cm") %>%
column_spec(3, width="3cm") %>%
    kable_styling(latex_options=c("HOLD_position"))

```

```{r hospvax2, echo=FALSE, message=FALSE}

knitr::kable(table.hosp.vax2, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hospvax2",
             row.names=TRUE, 
            caption="Risk factors for hospitalised or fatalCOVID-19 in those who had received 2 doses of vaccine at least 14 days before", 
             col.names=c(clean.header(colnames(table.severe.vax2)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    column_spec(column=1, width="4cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="3cm") %>%
    column_spec(column=5, width="1.7cm") %>%
    column_spec(column=6, width="2.7cm") %>%
    column_spec(column=7, width="1.7cm") %>%
    kableExtra::group_rows("Shielding eligibility category", 9, 14, bold=FALSE) %>%
   add_footnote(label=c("Percentages are column percentages for each variable",
                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
                         "Univariate rate ratios are for models with a single covariate",
                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
                 notation="none") %>%
    kable_styling(latex_options=c("scale_down", "HOLD_position"))

```

```{r cohortseverevax2, echo=FALSE, message=FALSE}

knitr::kable(severe.2vax, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="cohortseverevax2",
             row.names=TRUE, 
             col.names=c("Risk group", "Number of severe cases",
                         "Person-months", "Rate / 1000"),
             caption="Incidence of severe COVID-19 in clinically extremely vulnerable individuals who had received 2 doses of vaccine at least 14 days before, by risk subgroup") %>% 
     kable_styling(latex_options=c("HOLD_position"))

```

### 2021-08-29 vaxshieldupdate.Rmd
reinfections

@@
 
+The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories.
+The design controls not only for the matching factors of age, sex and general practice but also for calendar time.
+
 Covariates included in the models were those that have been previously identified as strong predictors of severe disease in this population: care home residence, number of adults in household, number of drug classes dispensed and recent hospital stay [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021].
@@
 
-The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories.
 With this incidence density sampling design, the conditional odds ratio is the rate ratio.
@@
 The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].
-The design controls not only for the matching factors of age, sex and general practice but also for calendar time.
-Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date.
-<!---
 # Cohort analysis of the shielding list
@@
 At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.
-A Poisson regression model was fitted to a dataset formatted with one observation per 28-day person-time interval, and individuals censored at first event.
-The event was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, age, and shielding eligibility group.
-The baseline hazard rate was specified as a natural spline function with 4 degrees of freedom.
-Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.
---->
+A Poisson regression model was fitted to the cohort formatted with one observation per 28-day person-time interval, and individuals censored at first diagnosis of COVID-19.
+Event status was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, baseline age, and shielding eligibility group.
+The baseline hazard rate was modelled as a natural spline function of calendar time with 6 degrees of freedom, allowing the model to fit an epidemic with multiple waves.
+Similar models were fitted with event status defined as hospitalized or fatal cases, or as any diagnosed case.
+To present the fitted hazard rates in a form that is interpretable as the risk associated with a single wave of the epidemic, the 60-day incidence rate was calculated for each time interval.
 
 # Results
-## Time course of vaccine efficacy against severe disease 
+## Shielding cohort
+By the date of the latest extract of the shielding list in July 2021, 8230 (4%) of the 202510 individuals who had ever been on the shielding list had been diagnosed with COVID-19, and 18893 (9%) had been removed from the list for other reasons.
+Fig \ref{fig:caseseverity} shows the fitted 60-day incidence of any diagnosis and of severe cases in Poisson regression models fitted to the entire cohort.
 Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.
-In the first half of July the average daily number of severe cases among the 200,000 people still on the shielded list was about 4, equating to an incidence rate of about 0.6 per 1000 per month.
+Although the incidence of any diagnosed COVID-19 in the shielding cohort rose steeply from 1 May onwards, the incidence of severe COVID-19 remained relatively low (less than 2 per 1000 per 60 days).
 
-Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category and vaccination status, in a model that includes both these variables together with care home residence and recent hospital stay as covariates.
+<!---Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category and vaccination status, in a model that includes both these variables together with care home residence and recent hospital stay as covariates.
 This shows that the rate ratio for severe disease associated with shielding eligibility, adjusted for vaccination status, has increased since 1 May, even though the absolute rate of severe disease in this group remains much lower than in December-January.
-The rate ratios for severe disease associated with 1 dose and 2 doses of vaccine at least 14 days earlier (with unvaccinated as reference category) declined  steadily from January to June, implying that the efficacy increased despite the rapid replacement of the Alpha variant by the Delta variant during May 2020 [@sheikh_sarscov2_2021].
+The rate ratios for severe disease associated with 1 dose and 2 doses of vaccine at least 14 days earlier (with unvaccinated as reference category) declined  steadily from January to June, implying that the efficacy against severe disease increased despite the rapid replacement of the Alpha variant by the Delta variant during May 2020 [@sheikh_sarscov2_2021].--->
 
@@
 On the scale of efficacy, these estimates are equivalent to 94% in those without risk conditions, 91% in those with risk conditions, and only 68% in those eligible for shielding.
+
+The estimates of efficacy for a single dose in Table \ref{tab:rateratiosriskgroups} differ from those we reported based on data up to 16 March 2021, with slightly different covariate adjustments [@mckeigue_efficacy_2021].
+However when we restricted the current dataset to exclude cases and controls sampled after 16 March 2021, the point estimates of the rate ratios associated with one dose of vaccine in the same model for those with no risk conditions, moderate risk conditions and eligible for shielding were 0.28, 0.25 and 0.31 respectively, similar to the estimates reported based on data without follow-up of outcome in cases after 16 March 2021 .
+[@mckeigue_efficacy_2021]
+
 To test formally for an effect of risk category on vaccine efficacy, a model was fitted with interaction terms for  risk category \ensuremath{\times} times vaccine dose.
@@
 
-To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower efficacy of vaccines in the clinically extremely vulnerable, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.
+The rate ratios for severe disease associated with one dose of vaccine were To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower efficacy of vaccines in the clinically extremely vulnerable, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.
 With this restricted case definition, the rate ratio associated with two doses of vaccine was `r format.estci(table.nonfatalorucod[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.nonfatalorucod[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.nonfatalorucod[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
@@
 The efficacy per dose was lower in all shielding eligibility groups than in those without risk conditions  Table \ref{tab:hosprateratiosshieldgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.
-Although there is evidence of efficacy in all shielding eligibility groups, the efficacy estimate was lowest in solid organ transplant recipients.
 
@@
 1.
-Although the efficacy of vaccination against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is considerably lower (about 70%).
-There is however a dose-response effect in this group: two doses are more effective than one.
+Although the efficacy of 2 doses of vaccine against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is lower (about 70%).
+Comparison between this updated report and our earlier report indicates that efficacy in the clinically extremely vulnerable is now lower than when the Alpha variant was the dominant strain in Scotland.
+There is however a dose-response effect in this group as in those with no risk conditions: two doses are more effective than one.
 
@@
 The main limitation of this study is that no data on hospital prescribing are available.
-Most biologic immunosuppressants and drugs for cancer are prescribed only in hospital, and the records of prescribing are not held in electronic form.
+Most immunosuppressants and drugs for cancer are prescribed only in hospital, and the records of prescribing are not held in electronic form.
 The national shielding list does not hold any information about drug therapy, and no information about diagnoses other than the seven broad categories listed as eligible for shielding.
@@
 \newpage
+
+# Figures
+
+```{r caseseverity, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:caseseverity} Poisson regression model for 90-day incidence of any case and severe case in the shielding cohort"}
+
+theme_set(theme_gray(base_size = 10))
+
+p.anycase
+
+```
 
@@
 
-p.byelig
+p.byelig + theme(legend.title = element_blank())
 
@@
 theme_set(theme_gray(base_size = 10))
-
-p.rateratio
+# p.rateratio
 

### 2021-08-29 vaxtimetrend.Rmd
reinfections

--- 2021-08-12 vaxshieldupdate.Rmd
@@
 ---
-title: "Efficacy of COVID-19 vaccination against severe COVID-19 and risk to  double-vaccinated individuals in Scotland: relationship to co-morbidities conferring  clinically extremely vulnerable status or moderate risk"
+title: "Time course of efficacy of vaccination against severe COVID-19 in Scotland"
 author:
@@
 
-source("vaxshieldupdate.R")
-
 options(knitr.kable.NA = '.')
@@
 
-# Abstract
+# Vaccine efficacy against Delta variant (or while Delta was the predominant variant)
 
-**Objectives** - To determine whether vaccine efficacy varies with risk category and shielding group and to investigate risk factors for severe COVID-19 in those who have received at least two doses of vaccine.
-
-**Design** - Matched case-control study.
-
-**Setting** - Population of Scotland from 1 December 2020 to 28 July 2021.
-
-**Main outcome measure** - Severe COVID-19, defined as cases with entry to critical care or fatal outcome.
-
-**Results** -The rate ratio for severe COVID-19 associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
-Among those eligible for shielding, there was a dose-response relationship and only weak evidence (p=0.04 in a test for interaction) that efficacy against hospitalized/severe disease was lower for the AZ vaccine than for the mRNA vaccines.
-Among the double-vaccinated, 87% of severe cases and 77% of hospitalised cases had designated risk conditions or were eligible for shielding.
-With those with no risk conditions as reference category, the rate ratio associated with receipt of a solid organ transplant among the double-vaccinated was `r format.estci(table.severe.vax2["Solid organ transplant", 3])` for severe disease and `r format.estci(table.hosp.vax2["Solid organ transplant", 3])` for hospitalisation.
-A risk score calculated from health records had moderate predictive performance (C-statistic 0.80) hospitalized cases and age-sex-matched controls among the double-vaccinated.
-
-**Conclusions** - 
-Although efficacy of vaccination against severe disease in those without risk conditions remains high, it is now clear that efficacy in clinically extremely vulnerable individuals is considerably lower.
-Solid organ transplant recipients remain at high risk of severe disease even after two doses of vaccine.
-
-\newpage
-
-# Introduction
-The REACT-SCOT matched case-control study was established at the beginning of the epidemic to investigate risk factors for severe COVID-19 in the population of Scotland [@mckeigue_rapid_2020].
-Using this framework, we have reported in detail on the relation of severe COVID-19 to risk factors including designated moderate risk conditions, drug prescribing and eligiblility for shielding [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021].
-We reported a study of vaccine efficacy against severe disease in different risk groups based on data up to 16 March 2021 [@mckeigue_efficacy_2021].
-As most people on the shielding list did not receive their first dose of vaccine until February 2021, there were too few people in this group who had received a second dose of vaccine at least 14 days earlier for the rate ratios associated with two vaccine doses to be estimated at this time.
-
-This report updates and extends the earlier analyses, with the following objectives:
-
-1. To examine the time course, since vaccination began, of the relation of severe COVID-19 to vaccination status and risk conditions.
-
-2. To determine whether vaccine efficacy varies with risk category and shielding group now that more data, including exposure to two doses of vaccine, are available.
-
-3. To investigate risk factors for infections causing severe illness or hospitalisation in those who have received at least two doses of vaccine.
-
-# Methods 
-We used the REACT-SCOT study to take advantage of data linkages already established.
-The design has been described in detail previously [@mckeigue_rapid_2020].
-In brief, for every incident case of COVID-19 in the population ten controls matched for one-year age, sex and primary care practice and alive on the day of presentation of the case that they were matched to were selected using the Community Health Index database.
-COVID-19 cases are those with a positive nucleic acid test, or a hospital admission or death with COVID-19 ICD-10 codes.
-The REACT-SCOT case-control dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.
-This analysis is based on cases presenting from 1 December 2020 to 14 July 2021, so that we have at least 14 days of follow-up for each case.
-This allows case severity status (hospitalisation, entry to critical care or death) to be assigned.
-For time-window analyses of length \ensuremath{2n + 1} days the time-window centred on 1 December 2020 includes the \ensuremath{n} days before that date.
-
-## Classification of risk categories
-As previously [@mckeigue_rapid_2020], to minimise ascertainment bias we pre-specified the primary outcome measure as severe COVID-19,  defined as diagnosed cases with entry to critical care within 28 days of presentation or fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).
-Cases and controls were classified into three broad risk categories: no risk condition; at least one of the moderate risk conditions designated by public health agencies [@mckeigue_rapid_2020]; or eligible for shielding [@mckeigue_relation_2021].
-For further analyses, the shielding category was subdivided as described previously  into six categories: solid organ transplant, specific cancers, severe respiratory conditions, other rare conditions, on immunosuppressants, and additional conditions [@mckeigue_relation_2021].
-This corresponds to the list used by Public Health Scotland  [@health_protection_scotland_search_2020], with the small numbers in the group "pregnant with heart disease combined with the "other conditions" category.
-For additional analyses the category "specific cancers" was split to allocate cancers of blood-forming organs (ICD-10 codes C81-C88, C90-C96) to a separate category.
-
-## Statistical analysis
-Vaccination status was coded as the number of doses administered at least 14 days before presentation date.
-Vaccine doses administered less than 14 days before presentation date are ignored.
-For estimates of effect by risk category, where the numbers of events are large, vaccination status was modelled as a factor comparing 1 dose or 2 doses with 0 doses as reference category.
-For estimates of effect within shielding eligibility groups, and for comparisons between the mRNA and AZ vaccines where the numbers of events in each category are smaller, vaccination status was modelled as a numeric variable taking values \ensuremath{0, \frac{1}{2}, 1}  so that the effect parameter represents the rate ratio associated with 2 doses assuming the effect of 0, 1, 2 doses to be linear on the scale of the log rate ratio.
-
-Covariates included in the models were those that have been previously identified as strong predictors of severe disease in this population: care home residence, number of adults in household, number of drug classes dispensed and recent hospital stay [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021].
-For care home residents the number of of adults in the household was coded as 1.
-The number of drug classes was calculated as the number of distinct BNF subparagraph codes for which a prescription was dispensed between 15 and 240 days before presentation date.
-As we have previously shown that severe COVID-19 in Scotland is associated with the number of non-cardiovascular but not the number of cardiovascular drug classes [@mckeigue_relation_2021a], the covariate included in the model was the number of non-cardiovascular drug classes.
-The number of hospital diagnoses was calculated as the number of distinct ICD-10 chapters represented at least once in hospital discharge records between 25 days and 5 years before presentation date.
-Recent hospital stay was defined as any in-patient stay from 5 to 14 days before presentation date.
-
-The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories.
-With this incidence density sampling design, the conditional odds ratio is the rate ratio.
-The efficacy of vaccination is 1 minus the rate ratio.
-The strata in the conditional logistic regression model are matched sets: typically one case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.
-When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.
-Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.
-The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].
-The design controls not only for the matching factors of age, sex and general practice but also for calendar time.
-Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date.
-
-# Cohort analysis of the shielding list
-The case-control study can estimate only rate ratios.
-To investigate the absolute rates of severe disease in those listed as clinically extremely vulnerable, we undertook a separate cohort study of all individuals who had ever been on the shielding list.
-At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.
-A Poisson regression model was fitted to a dataset formatted with one observation per 28-day person-time interval, and individuals censored at first event.
-The event was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, age, and shielding eligibility group.
-The baseline hazard rate was specified as a natural spline function with 4 degrees of freedom.
-Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.
-
-# Results
-## Time course of the epidemic and associations with risk factors 
-Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.
-From mid-January to early March 2021 there was a rapid decline in the incidence of severe cases, followed by a slight increase from early June to early July.
-
-Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category, vaccination dose (coded as 0, 1, 2 doses), and recent (5 to 14 days before presentation date) hospital exposure as inpatient.
-These rate ratios are computed from a model that includes all variables shown, together with care home residence.
-
-The rate ratio for severe disease associated with shielding eligibility did not decline as the winter peak declined, and increased from May 2020 onwards.
-The efficacy of vaccination against severe disease (modelled as the rate ratio per 2 doses) remained high, despite the rapid replacement of the Alpha by the Delta variant during May 2020.
-
-## Relation of vaccine efficacy to risk conditions
-Table \ref{tab:freqsvaxgr} tabulates risk/shielding group, vaccination status by product, and other risk factors in severe cases and controls with presentation dates from 1 December 2020 to 14 July 2021.
-Table \ref{tab:hospfreqsvaxgr} shows the same tabulation with case definition broadened to include all hospitalised cases and their matched controls.
-
-Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories, in a model that includes risk categories (no risk conditions as reference category),  care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates.
-The rate ratio associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
-On the scale of efficacy, these estimates are equivalent to 94% in those without risk conditions, 91% in those with risk conditions, and only 68% in those eligible for shielding.
-To test formally for an effect of risk category on vaccine efficacy, a model was fitted with interaction terms for  risk category \ensuremath{\times} times vaccine dose.
-In those with 2 doses, the effect (on the scale of rate ratio) of eligibility for shielding (with no risk condition as reference category) on the rate ratio associated with 2 doses of vaccine (0 doses as reference category) was `r format.estci(vaxriskgr.interaction)`.
-
-To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower efficacy of vaccines in the clinically extremely vulnerable, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.
-With this restricted case definition, the rate ratio associated with two doses of vaccine was `r format.estci(table.nonfatalorucod[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.nonfatalorucod[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.nonfatalorucod[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.
-Table \ref{tab:hosprateratiosriskgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls: the efficacy of 2 doses was lower in those eligible for shielding than in the other two risk categories.
-
-Table \ref{tab:rateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group.
-The efficacy per dose was lower in all shielding eligibility groups than in those without risk conditions  Table \ref{tab:hosprateratiosshieldgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.
-Although there is evidence of efficacy in all shielding eligibility groups, the efficacy estimate was lowest in solid organ transplant recipients.
-
-Table \ref{tab:rateratiosvaxgroups} shows rate ratios associated with 1 and 2 doses of each class of vaccine (with 0 doses as reference category) in each of the three broad risk categories.
-Table \ref{tab:hosprateratiosvaxgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.
-In those eligible for shielding, a separate analysis was undertaken to test for difference in efficacy between the AZ vaccine and mRNA vaccines.
-In conditional logistic regression models with vaccine dose and vaccine product centred at zero mean and other covariates as in Table  \ref{tab:hosprateratiosvaxgroups}, there was no evidence of lower efficacy against severe COVID-19 for the AZ vaccine and only weak evidence for lower efficacy against the broader category of hospitalized or fatal COVID-19: the rate ratio associated with vaccine dose was 1.51 (95% CI 0.66 to 1.03, _p_=0.04) times higher for the AZ vaccine than for the Pfizer vaccine.
-
-## Risk factors for severe COVID-19 in the double-vaccinated
-Table \ref{tab:severevax2} shows risk factors for severe COVID-19 in cases and controls who had received 2 doses of vaccine at least 14 days before presentation date.
-Table \ref{tab:hospvax2} shows the same analysis with case definition broadened to include hospitalized cases and matched controls who had received 2 doses of vaccine.
-Of the variables that we have previously reported as risk factors for severe COVID-19 in the general population, care home residence and socioeconomic deprivation were not associated with severe disease in the double-vaccinated.
-In this group the risk factors for severe disease were   designated risk conditions, other indicators of co-morbidity including numbers of hospital diagnoses and drug classes prescribed and transmission-related factors including number of adults in household and recent hospital stay.
-Only 13% of double-vaccinated severe cases had no designated risk condition: 44% had at least one moderate risk condition and 43% were eligible for shielding.
-With those with no risk conditions as reference category, the rate ratio associated with receipt of a solid organ transplant among the double-vaccinated was `r format.estci(table.severe.vax2["Solid organ transplant", 3])` for severe disease and `r format.estci(table.hosp.vax2["Solid organ transplant", 3])` for hospitalisation.
-
-A risk score for hospitalised or fatal COVID-19 in the double-vaccinated calculated from the conditional logistic regression model had a C-statistic of 0.80.
-
-<!---
-## Associations with severe diseases after the first quarter of 2021
-
-Table \ref{tab:freqsriskgroups} tabulates risk factors and vaccination status in severe cases and controls restricted to presentation dates from April to July 2021, by age dichotomized at 50 years.
-Among severe cases aged over 50 years, 75% had been vaccinated with at least one dose, and 74% had a moderate risk condition or were eligible for shielding.
-Among severe cases aged less than 50, 81% had not been vaccinated, and 74% had no risk condition.
-
-Table \ref{tab:severeagevax} tabulates severe cases presenting from April to July 2021 by 10-year age group and vaccination status, showing that unvaccinated individuals aged 30-49 made up 129 of of the 171 severe cases among those aged less than 50 years.
-
-Table \ref{tab:casesnoriskchnums} tabulates severe cases and matched controls aged under 50 years without designated risk conditions presenting from April to July 2021, with frequencies in each chapter of ICD-10 of at least one hospital discharge diagnosis  from 5 years to 25 days before presentation date.
-Table  \ref{tab:casesnoriskbnfchnums} tabulates for the same subset of cases and controls the frequencies in each chapter of the BNF of at least one prescription dispensed  from 240 to 15 days before presentation date.
-In comparison with their matched controls, the cases had higher frequencies of admissions across several chapter headings including diseases of the digestive system, and other more non-specific chapter headings, and higher frequencies of at least one dispensed prescription in most chapters of the BNF.
---->
-
-# Discussion
-## Statement of principal findings
-
-1.
-Although the efficacy of vaccination against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is considerably lower (about 70%).
-There is however a dose-response effect in this group: two doses are more effective than one.
-
-2.
-Solid organ transplant recipients remain at very high risk of severe disease even after two doses of vaccine: approximately 70-fold for severe disease and 30-fold for hospitalisation in comparison with double-vaccinated individuals of the same age and sex without risk conditions.
-
-3. There is only weak evidence that efficacy against severe disease disease is lower for the AZ vaccine than for the mRNA vaccines.
-
-4.
-Among the double-vaccinated, 87% of severe cases and 77% of hospitalised cases have designated risk conditions or are eligible for shielding.
-A risk score calculated from health records has moderately high performance as a predictor of severe COVID-19 or hospitalisation among the double-vaccinated, independent of age and sex.
-
-## Strengths and limitations
-Strengths of this study are the focus on severe cases rather than test-positives, the matched case-control design and the comprehensive linkage to e-health records
-
-The main limitation of this study is that no data on hospital prescribing are available.
-Most biologic immunosuppressants and drugs for cancer are prescribed only in hospital, and the records of prescribing are not held in electronic form.
-The national shielding list does not hold any information about drug therapy, and no information about diagnoses other than the seven broad categories listed as eligible for shielding.
-
-## Relation to other studies
-A recent report based on primary care data covering 7.2 million people curated by the Royal College of General Practitioners Research and Surveillance Centre examined vaccine efficacy against symptomatic COVID-19 for which a medical consultation was recorded  [@whitaker_pfizerbiontech_2021].
-For those who had been advised to shield, the efficacy after two doses of any vaccine was estimated as 92% in those aged less than 65 years and 84% in those aged 65 years and over.
-For the "immunocompromised" group the estimated vaccine efficacy was somewhat lower at 74%, with a wide confidence interval.
-No separate analysis of vaccine efficacy in solid organ transplant recipients was reported.
-
-## Policy implications. 
-The dose-response effect suggests that an additional dose might give more protection to double-vaccinated individuals who are clinically extremely vulnerable.
-
-For identifying those among the double-vaccinated who are at high risk of severe disease, a   risk score based on electronic health records has better discrimination than using only shielding eligibility.
-However if vaccine supply is not constrained, a policy of offering booster vaccination to all those above an age threshold and all those with designated risk conditions would achievetogether with acombined would cover more than 80% of those at risk of severe disease.
+| Study | Infection | Hospitalisation |  
+| ----  | --------- | --------------- |  
+| @nanduri_effectiveness_2021 | 53% |. | 
+| @tenforde_sustained_2021 | . | 84% |  
+| @rosenberg_new_2021 | 80% during July | 95%  |
+| @lopez_bernal_effectiveness_2021 | Pfizer 88%, AZ 67% | . |  
+| @puranik_comparison_2021 | Pfizer 42%, Moderna 76% | Pfizer 75%, Moderna 81% |  
+| @pouwels_impact_2021 | Pfizer 80%, AZ 67% | . |
 
@@
 
-# Tables
+# Figures
 
-```{r freqsvaxgroups, echo=FALSE, message=FALSE}
-
-knitr::kable(freqs.vaxgr, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="freqsvaxgr",
-             row.names=TRUE, 
-			 align=c('r', 'r'), 
-             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccine product") %>%
-    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
-                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-                 notation="none") %>%
-    kableExtra::group_rows("Risk/shielding eligibility group", 6, 13, bold=FALSE) %>% 
-    kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
-    column_spec(1, width="5cm") %>%
-    column_spec(2, width="4cm") %>%
-    column_spec(3, width="3cm") %>%
-   kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r rateratiosriskgroups, echo=FALSE, message=FALSE}
-
-knitr::kable(table.riskgr[7:12, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="rateratiosriskgroups",
-             row.names=FALSE,
-             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
-			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for severe COVID-19 within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 1, 2, bold=TRUE) %>% 
-    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 3, 4, bold=TRUE)  %>% 
-    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 5, 6, bold=TRUE)  %>% 
-    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
-                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
-                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                         "Vaccination status is doses administered at least 14 days fbefore presentation date"), 
-                 notation="none") %>%
-    column_spec(1, width="6cm") %>%
-    column_spec(2, width="4cm") %>%
-    column_spec(3, width="3cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r rateratiosshieldgroups, echo=FALSE, message=FALSE}
-
-knitr::kable(table.shieldgr[14:21, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="rateratiosshieldgroups",
-             row.names=FALSE,
-             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
-			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for severe COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1/2, 1") %>%
-add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Severe COVID-19 defined as entry to critical care or death within 28 days of presentation.", 
-                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
-                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
-                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-             notation="none") %>%
-    column_spec(1, width="8cm") %>%
-column_spec(2, width="4cm") %>%
-column_spec(3, width="3cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r rateratiosvaxgroups, echo=FALSE, message=FALSE}
-
-startrow <- 1
-knitr::kable(table.vaxgr[7:18,], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="rateratiosvaxgroups",
-             row.names=FALSE,
-             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
-			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for severe COVID-19 within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition", startrow, startrow+3, bold=TRUE) %>% 
-    kableExtra::group_rows("Moderate risk condition", startrow+4, startrow+7, bold=TRUE)  %>% 
-    kableExtra::group_rows("Eligible for shielding", startrow+8, startrow+11, bold=TRUE)  %>% 
-add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Adjusted for care home residence, number of adults in household, number of drug classes dispensed, recent hospital stay and risk category.", 
-                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-             notation="none") %>%
-    column_spec(1, width="8cm") %>%
-column_spec(2, width="4cm") %>%
-column_spec(3, width="3cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r severevax2, echo=FALSE, message=FALSE}
-
-knitr::kable(table.severe.vax2, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="severevax2",
-             row.names=TRUE, 
-            caption="Risk factors for severe COVID-19 in those who had received 2 doses of vaccine at least 14 days before", 
-             col.names=c(clean.header(colnames(table.severe.vax2)[1:2]),
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
-    column_spec(column=1, width="4cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-    column_spec(column=4, width="3cm") %>%
-    column_spec(column=5, width="1.7cm") %>%
-    column_spec(column=6, width="2.7cm") %>%
-    column_spec(column=7, width="1.7cm") %>%
-    kableExtra::group_rows("Shielding eligibility category", 9, 14, bold=FALSE) %>%
-   add_footnote(label=c("Severe COVID-19 is defined by entry to critical care or fatal outcome", 
-                         "Percentages are column percentages for each variable",
-                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
-                         "Univariate rate ratios are for models with a single covariate",
-                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
-                 notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "HOLD_position"))
-
-```
-
-# Supplementary Material
-\beginsupplement
-
-## Supplementary Figures
-
-```{r plotcases, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotcases} Daily severe cases by risk category"}
+```{r bycalendartime, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:caseseverity} Time course of efficacy of 1 and 2 doses of each vaccine against severe disease: rate ratios in multivariable model admPoisson regression model for 90-day incidence of any case and severe case in the shielding cohort"}
 
@@
 
-p.byelig
+gridExtra::grid.arrange(p.rateratio, p.rateratio.hosp)
 
 ```
-
-```{r plotrateratios, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:plotrateratios} Rate ratios by 42-day sliding window.
-For each time window, the conditional logistic regression model includes all variables shown in the plot, together with care home residence.
-Line thickness is inversely proportional to standard error"}
-
-theme_set(theme_gray(base_size = 10))
-
-p.rateratio
-
-```
-
-## Supplementary Tables
-
-```{r hospfreqsvaxgroups, echo=FALSE, message=FALSE}
-
-knitr::kable(freqs.hosp.vaxgr, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hospfreqsvaxgr",
-             row.names=TRUE, 
-			 align=c('r', 'r'), 
-             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccine product") %>%
-    add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                         "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-                 notation="none") %>%
-    kableExtra::group_rows("Risk/shielding eligibility group", 6, 12, bold=FALSE) %>% 
-    kableExtra::group_rows("Vaccination status", 14, 18, bold=FALSE) %>% 
-    column_spec(1, width="5cm") %>%
-    column_spec(2, width="4cm") %>%
-    column_spec(3, width="3cm") %>%
-   kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r hosprateratiosriskgroups, echo=FALSE, message=FALSE}
-
-knitr::kable(table.hosp.riskgr[7:12, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hosprateratiosriskgroups",
-             row.names=FALSE,
-             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
-			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for hospitalized or fatal COVID-19 within risk groups: vaccine dose as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition, unvaccinated as reference category", 1, 2, bold=FALSE) %>% 
-    kableExtra::group_rows("Moderate risk condition, unvaccinated as reference category", 3, 4, bold=FALSE)  %>% 
-    kableExtra::group_rows("Eligible for shielding, unvaccinated as reference category", 5,  6, bold=FALSE)  %>% 
-add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
-                     "Adjusted for care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates", 
-                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-             notation="none") %>%
-    column_spec(1, width="6cm") %>%
-column_spec(2, width="4cm") %>%
-column_spec(3, width="3cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r hosprateratiosshieldgroups, echo=FALSE, message=FALSE}
-
-knitr::kable(table.hosp.shieldgr[13:20, ], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hosprateratiosshieldgroups",
-             row.names=FALSE,
-             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
-			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for hospitalized or fatal COVID-19 within shielding eligibility groups: vaccine dose encoded as numeric variable taking values 0, 1, 2") %>%
-add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Controls matched for age, sex, and primary care practice on date of presentation of case.",
-                     "Rate ratios estimated in multivariable conditional logistic regression with all variables in model",  
-                    "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-             notation="none") %>%
-    column_spec(1, width="8cm") %>%
-column_spec(2, width="4cm") %>%
-column_spec(3, width="3cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r hosprateratiosvaxgroups, echo=FALSE, message=FALSE}
-
-startrow <- 1
-knitr::kable(table.hosp.vaxgr[7:18,], 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hosprateratiosvaxgroups",
-             row.names=FALSE,
-             col.names=c("Effect", "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"),
-			 align=c('l', 'c', 'c'), 
-             caption="Rate ratios for hospitalized or fatal COVID-19 within risk groups: vaccine dose and product encoded as categoric variable with unvaccinated as reference category") %>%
-    kableExtra::group_rows("No risk condition", startrow, startrow+3, bold=TRUE) %>% 
-    kableExtra::group_rows("Moderate risk condition", startrow+4, startrow+7, bold=TRUE)  %>% 
-    kableExtra::group_rows("Eligible for shielding", startrow+8, startrow+11, bold=TRUE)  %>% 
-add_footnote(label=c("Presentation dates from 1 December 2020 to 14 July 2021.",
-                     "Adjusted for care home residence, number of adults in household, number of drug classes dispensed, recent hospital stay and risk category.", 
-                     "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
-                     "Vaccination status is number of doses administered at least 14 days before presentation date"), 
-             notation="none") %>%
-    column_spec(1, width="8cm") %>%
-column_spec(2, width="4cm") %>%
-column_spec(3, width="3cm") %>%
-    kable_styling(latex_options=c("HOLD_position"))
-
-```
-
-```{r hospvax2, echo=FALSE, message=FALSE}
-
-knitr::kable(table.hosp.vax2, 
-             escape=FALSE, 
-             booktabs=TRUE,
-			 label="hospvax2",
-             row.names=TRUE, 
-            caption="Risk factors for hospitalised or fatalCOVID-19 in those who had received 2 doses of vaccine at least 14 days before", 
-             col.names=c(clean.header(colnames(table.severe.vax2)[1:2]),
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
-                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
-    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
-    column_spec(column=1, width="4cm") %>%
-    column_spec(column=2, width="2cm") %>%
-    column_spec(column=3, width="2cm") %>%
-    column_spec(column=4, width="3cm") %>%
-    column_spec(column=5, width="1.7cm") %>%
-    column_spec(column=6, width="2.7cm") %>%
-    column_spec(column=7, width="1.7cm") %>%
-    kableExtra::group_rows("Shielding eligibility category", 9, 14, bold=FALSE) %>%
-   add_footnote(label=c("Percentages are column percentages for each variable",
-                         "Rate ratios are from conditional logistic regression models matched for age, sex and general practice",
-                         "Univariate rate ratios are for models with a single covariate",
-                         "Multivariable rate ratios are for a model including all covariates shown in the table"),
-                 notation="none") %>%
-    kable_styling(latex_options=c("scale_down", "HOLD_position"))
-
-```

### 2021-08-29 vaxtrend.Rmd
reinfections

---
title: "Efficacy of vaccination against severe COVID-19 in relation to Delta variant and to time since second dose: the REACT-SCOT case-control study"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
    corresponding: yes
    email: paul.mckeigue@ed.ac.uk
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    affiliation: HPS
  - name: for the PHS COVID-19 Epidemiology and Research Cell
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland.
- code: IGMM
  address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. 
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ.
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
header-includes:
  \usepackage{newunicodechar}
  \usepackage[T1]{fontenc}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
#csl: ./f1000.csl 
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---

```{r vax, echo=FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

```
\newpage

# Abstract

**Objectives** -  To investigate: (1) whether vaccine efficacy against severe COVID-19 has decreased since Delta became the predominant variant; (2) whether efficacy wanes with time since second dose.

**Design** - Matched case-control study.

**Setting** - Population of Scotland from 1 December 2020 to 28 July 2021.

**Main outcome measure** - Severe COVID-19, defined as cases with entry to critical care or fatal outcome.

**Results** - Efficacy of vaccination against severe COVID-19 increased up to July 2021, despite a temporary decrease coinciding with the replacement of the B.1.1.7 (Alpha) by the B.1.617.2 (Delta) variant in May 2021.  In the most recent time window, the efficacy of two doses against severe COVID-19 was `r az.efficacy.severe` for the AstraZeneca product and `r mrna.efficacy.severe` for mRNA (Pfizer or Moderna) products.  Efficacy against severe COVID-19 decreased over the first two months since the second dose but stabilized thereafter: the model with best fit was one with a change-point at `r bestfit.cp` weeks since second dose.  Against the broader category of hospitalised or fatal COVID-19, the results were similar except that  efficacy in the most recent time window was lower for the AZ product at `r az.efficacy.hosp` than for mRNA vaccines at `r mrna.efficacy.hosp`.

**Conclusions** - 
These results are reassuring with respect to concerns that efficacy against severe COVID-19 might have fallen since the Delta variant became predominant, or that efficacy against severe disease may continue to wane with time since second dose.  This weakens the rationale for policies that offer booster doses to the entire population, rather than to vulnerable individuals for focused protection.

\newpage

# Introduction
Recent reports have suggested that efficacy of vaccines against COVID-19 infection may have fallen since the B.1.617.2 (Delta) variant became predominant [@nanduri_effectiveness_2021;@lopez_bernal_effectiveness_2021;@pouwels_impact_2021;@tenforde_sustained_2021; @rosenberg_new_2021;@puranik_comparison_2021].  These reports are summarized below.

## Vaccine efficacy before and after Delta became the predominant variant

| Study | Setting | Infection || Hospitalisation ||   
| --  | ------- |----- | -----| ----- | ----- |
| | | **Before Delta** | **After Delta** | **Before Delta** | **After Delta** |   
| @nanduri_effectiveness_2021 | Care home residents | Pfizer 74%, Moderna 75% | Pfizer 52%, Moderna 51% | . | 
| @tenforde_sustained_2021  | IVY Network (hospitals) | . | . | 87% | 84%  |  
| @rosenberg_new_2021 | New York State| 92%. | 80% | 92% | 95% |
| @puranik_comparison_2021 | Mayo Clinic Health System | Pfizer 76%, Moderna 86% | Pfizer 42%, Moderna 76% | Pfizer 85%, Moderna 92% | Pfizer 75%, Moderna 81% |  
| @lopez_bernal_effectiveness_2021  | England | Pfizer 94%, AZ 75% | Pfizer 88%, AZ 67% | . |  
| @pouwels_impact_2021 | UK ONS Covid-19 Infection Survey | Pfizer 78%, AZ 79% | Pfizer 80%, AZ 67% | . |

Other studies have raised concerns that efficacy of two doses may wane with time since dose.  A study of individuals registered with an Israeli health-care organization reported lower efficacy of two doses against infection during June-July 2021 among those who had received their second dose in January-February compared with those who had received their second dose in March-April [@mizrahi_correlation_2021].  A study in England reported that efficacy against infection fell with time since second dose by 22% and 7% per 30 days for the AZ and Pfizer vaccines respectively [@puranik_comparison_2021].  These concerns have has led public health agencies in the United States to recommend booster doses [@office_of_the_commissioner_joint_2021a].

The outcomes assessed by these studies -- positive nucleic acid test [@nanduri_effectiveness_2021;@lopez_bernal_effectiveness_2021], positive test with low PCR cycle threshold [@pouwels_impact_2021], or hospitalisation [@tenforde_sustained_2021;@rosenberg_new_2021;@puranik_comparison_2021] -- are subject to ascertainment bias.  Severe COVID-19, defined as cases that are fatal or require critical care, is less susceptible to ascertainment bias and is also the most important outcome for policy purposes, as the original rationale for lockdowns was to prevent critical care services from being overwhelmed.

The objectives of this study were to investigate for the two main classes of vaccine: (1) whether efficacy against severe COVID-19 has decreased since Delta became the predominant variant; (2) whether efficacy of two doses against severe COVID-19 wanes with time since second dose.

# Methods
The design of the REACT-SCOT case-control study has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the Scottish population ten controls matched for one-year age, sex and primary care practice and alive on the day of presentation of the case that they were matched to were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission or death with COVID-19 ICD-10 codes.  As previously [@mckeigue_rapid_2020], to minimise ascertainment bias we pre-specified the primary outcome measure as severe COVID-19,  defined as diagnosed cases with entry to critical care within 28 days of presentation or fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).  We also examined the broader category of hospitalised or fatal COVID-19. Analyses reported here are restricted to cases and controls presenting from 1 December 2020 onwards.

Vaccination status was defined by the number of doses received at least 14 days before presentation date.  Of those who had received two doses of an mRNA vaccine, less than 1% had received the Moderna product. The Pfizer and Moderna products were therefore grouped together as "mRNA vaccine".

The incidence density sampling design controls not only for the matching factors of age, sex and primary care practice but also for calendar time. Rate ratios for severe COVID-19 were estimated from conditional logistic regression models.  The efficacy of vaccination is 1 minus the rate ratio. Covariates included in each model were those that have been previously identified as strong predictors of severe disease in this population: care home residence, risk category (no risk condition, moderate risk condition, clinically extremely vulnerable), number of non-cardiovascular drug classes dispensed and recent hospital stay [@mckeigue_rapid_2020;@mckeigue_relation_2021a;@mckeigue_relation_2021b].

To investigate the effect of the Delta variant we examined how efficacy varied with calendar time, and to investigate possible waning we examined how the efficacy varied with time since last dose.  To allow the data to speak for themselves, the initial analysis presents line plots of rate ratios estimated within sliding 42-day time windows against calendar time and against time since last dose.  For a formal comparison between time periods before and after Delta became the predominant variant, we estimated rate ratios before and after 19 May 2021: the date on which Delta became the main variant in Scotland [@sheikh_sarscov2_2021].  For modelling the relation of efficacy to time since dose, the results of the line plots led us to adopt the modelling approach described in the Results section.

# Results
Tables \ref{tab:severe} and \ref{tab:hosp} show the distributions of risk factors in cases and their matched controls, for severe cases and the broader category of hospitalised or fatal cases respectively. These results are provided for reference only: the reader is cautioned that unconditional odds ratios calculated from these tables cannot be used to estimate rate ratios because of the matched design [@breslow_estimation_1978;@greenland_need_1982].

## Vaccine efficacy before and after the Delta variant became predominant
Figure \ref{fig:bycalendartime} (a) shows that the rate ratio for severe COVID-19 associated with a single dose of vaccine declined (and thus efficacy increased) from March to July 2021. This trend was briefly interrupted during May 2021, corresponding to the period when the Alpha variant was replaced by the Delta variant.  The rate ratios for severe COVID-19 associated with two doses of vaccine show a similar trend, but the estimates of rate ratios for time windows before May 2021 lack precision (represented in the plot by line thicknesses) because at this time few individuals at risk of severe disease had received their second dose.  To compare the rate ratio before and after the date that the Delta variant became predominant, a conditional logistic regression model was fitted with the effect of two doses nested within each level of an indicator variable defined as presentation date on or after 19 May 2021.   The rate ratio for severe disease associated with two doses did not differ between the periods before and after this cutoff date: `r estcipv.severe.before19may` before 19 May and `r estcipv.severe.after19may` from 19 May onwards.

Figure \ref{fig:bycalendartime} (b) shows that the rate ratio for the broader category of hospitalised or fatal disease associated with two doses of vaccine increased (and thus efficacy was lower) after the Delta variant became predominant.  The rate ratio was estimated as `r estcipv.hosp.before19may` before 19 May and `r estcipv.hosp.after19may` from 19 May onwards

Figure \ref{fig:bycalendartime} (a) shows that against severe COVID-19, the efficacy of the AZ and mRNA vaccines did not differ after May 2021, but Figure \ref{fig:bycalendartime} (b) shows that against the broader category of hospitalised or fatal COVID-19 the AZ vaccine had lower efficacy than the mRNA vaccines.

## Vaccine efficacy by time since second dose 
Figure \ref{fig:bytimesincedose} (a) shows that the rate ratio for severe COVID-19 increases (and thus efficacy decreased) up to about 12 weeks after the second dose for both AZ and mRNA vaccines, but that there is no clear trend after that time point. The slope of the increase in rate ratio is less steep for the AZ vaccine than for the mRNA vaccines, so that the difference in efficacy narrows with time since second dose.  Figure \ref{fig:bytimesincedose} (b) shows the same analysis for the broader category of hospitalized or fatal COVID-19.  As for severe COVID-19 the rate ratio associated with vaccine increases up to about 12 weeks and there is no clear trend or even a decreasing rate ratio (increasing efficacy) thereafter.  As in Figure \ref{fig:bycalendartime} (b), efficacy against hospitalisation remains higher for the mRNA than for the AZ vaccine.

These results led us to test formally for a change-point in the slope of the relationship of log rate ratio for severe disease associated with vaccination to time since second dose.  For this we compared the fit of models with change points from 8 to 16 weeks after the second dose.  In each model the slopes before and after the change-point were encoded as interaction terms for vaccination status (2 doses versus none) \ensuremath{\times} weeks since second dose.  Models with a change-point were compared to a model with no change-point. These models included the same covariates as before.

The results of this model comparison are shown in Table \ref{tab:changepoints}. The best-fitting model was one with a change-point at  `r bestfit.cp` weeks since second dose, with deviance `r 2 * max(loglik.changepoints$loglik)` units lower (_p_=`r pvalue.latex(1 - pchisq(2 * max(loglik.changepoints$loglik), df=1))`) than the model with no change point.  The slope of the relationship, expressed as a multiplicative effect on the rate ratio associated with two doses of vaccine,  was `r bestfit.effectbefore` per week since second dose up to the `r bestfit.cp`-week change-point and `r bestfit.effectafter` per week thereafter.

This procedure was repeated for efficacy against the broader outcome of hospitalized or fatal COVID-19 ( Table \ref{tab:changepointshosp}).  Again the best-fitting model was one with a change-point at  `r bestfit.cp.hosp` weeks since second dose, with deviance `r 2 * max(loglik.changepoints.hosp$loglik)` units lower (_p_=`r pvalue.latex(1 - pchisq(2 * max(loglik.changepoints.hosp$loglik), df=1))`) than the model with no change point.

# Conclusions
## Statement of principal findings
* Efficacy of vaccination against severe COVID-19 has increased from March to July 2021 despite the replacement of the Alpha by the Delta variant in May 2021.

* Efficacy of a second dose against severe COVID-19 declines during the first `r bestfit.cp` weeks since vaccination but stabilizes thereafter.  This early decline of efficacy is faster for the mRNA vaccines than for the AZ product.

*  The efficacy of two doses against severe COVID-19 in the most recent time windows is around 95% and does not differ between AZ and mRNA products.  Efficacy against the broader category of hospitalised or fatal COVID-19 remains lower for the AZ than the mRNA products (85% versus 91%).

## Strengths and limitations
Strengths of this study are the focus on severe COVID-19 as main outcome measure, the elimination of confounding by calendar time in the matched case-control design, and the ability to control for multiple comorbidities through linkage to electronic health records including prescribing.  A limitation of the incidence density case-control design is that it excludes reinfections: the sampling of cases and controls excludes those who have previously tested positive for COVID-19.  For investigating the possible effect of the Delta variant on efficacy, a limitation is that we do not have direct measurements of variant type; however in Scotland the Alpha variant was almost completely replaced by Delta during over a few weeks in May 2021 [@sheikh_sarscov2_2021], and we were able to detect the effect of this as a temporary interruption of the upward trend in vaccine efficacy seen in the plot of efficacy against calendar time.

The effects of calendar time and time since second dose are confounded with each other and with other factors not considered in this analysis including seasonality, the build-up of natural immunity, and the changing demographic and morbidity profile of cases.  We would expect the build-up of natural immunity to decrease vaccine efficacy by lowering risk in the unvaccinated. On the basis of recent analyses (to be reported elsewhere) showing lower efficacy against severe COVID-19 in those designated as clinically extremely vulnerable, we would expect the efficacy to decrease where these groups are exposed to infection. Investigating these factors is important but is not necessarily relevant to the immediate policy issue addressed by this paper. For this it is sufficient to show that the efficacy of two doses has not decreased with calendar time and that efficacy does not continue to wane with time since second dose.

## Comparison with other studies
We have previously reported that efficacy of vaccines is higher against severe COVID-19 than against the broader category of hospitalised or fatal COVID-19.  Other studies have reported that efficacy against hospitalisation in turn is higher than efficacy against all test-positive infections [@puranik_comparison_2021;@lopez_bernal_effectiveness_2021].

In three recent studies from the UK [@lopez_bernal_effectiveness_2021; @pouwels_impact_2021;@sheikh_sarscov2_2021] the efficacy against infection with the Delta variant, or against any infection since Delta became the predominant variant, was estimated to be lower for the AZ product than for the Pfizer product.  Only one of these studies, based on Scottish data up to 6 June 2021, was able to examine  efficacy against hospitalisation [@sheikh_sarscov2_2021]; the efficacy against hospitalisation with an S gene positive PCR result (a proxy for Delta) was 62%, 95% CI 42% to 76%).

The ONS study reported that vaccine efficacy against test-positive cases with PCR cycle threshold < 30 or against symptomatic cases declined faster from 14 days after second vaccination for the Pfizer mRNA vaccine than for the AZ vaccine [@pouwels_impact_2021].  The authors suggested that if these results were extrapolated, these two vaccines would have equal efficacy against these outcomes after 16-20 weeks.  We have extended this finding by showing that efficacy against severe COVID-19 and hospitalisation declines faster for mRNA vaccines than for the AZ product, and we show that the two classes of vaccine have about equal efficacy against severe disease after about 12 weeks.  However for efficacy against hospitalisation, the difference in efficacy persists: the gap between the two vaccines (on the scale of log rate ratio) narrows only up to about 6 weeks since second dose.

## Policy implications
A joint CDC/FDA statement on 18 August asserted that [@office_of_the_commissioner_joint_2021a]:

> The available data make very clear that protection against SARS-CoV-2 infection begins to decrease over time following the initial doses of vaccination, and in association with the dominance of the Delta variant, we are starting to see evidence of reduced protection against mild and moderate disease.

On this basis, the statement concluded that "a booster shot will be needed to maximize vaccine-induced protection and prolong its durability", reversing an earlier position [@office_of_the_commissioner_joint_2021].  Our results are more reassuring, in that the decrease in efficacy against severe COVID-19 or hospitalised / fatal COVID-19 is limited to the first 9 weeks since second dose, and the efficacy thereafter appears to stabilize.  We also find no evidence that the replacement of the Alpha by the Delta variant has done more than temporarily interrupt the trend of vaccine efficacy against severe COVID-19 to increase with calendar time.  On this basis the rationale for offering booster doses to the entire population, rather than to vulnerable individuals for focused protection, is open to question.

# Declarations
## Ethics approval
This study was performed within Public Health Scotland as part of its statutory duty to monitor and investigate public health problems.  Under the [UK Policy Framework for Health and Social Care Research](https://www.hra.nhs.uk/planning-and-improving-research/policies-standards-legislation/uk-policy-framework-health-social-care-research/) set out by the NHS Health Research Authority, this does not fall within the definition of research and ethical review is not required.  Individual consent is not required for Public Health Scotland staff to process personal data to perform specific tasks in the public interest that fall within its statutory role.  The statutory basis for this is set out in Public Health Scotland's [privacy notice](https://www.publichealthscotland.scot/our-privacy-notice/personal-data-processing/).

# References 
<div id="refs"></div>

# Figures
```{r bycalendartime, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=1.3, fig.cap="\\label{fig:bycalendartime} Relation of  vaccine efficacy to calendar time: (a) severe COVID-19; (b) hospitalised or fatal COVID-19.  Vaccination status is defined by number of doses received at least 14 days before. Rate ratios in conditional logistic regression model, adjusted for covariates. Thickness of each line is proportional to precision of estimate"}

theme_set(theme_gray(base_size = 10))

gridExtra::grid.arrange(p.rateratio + labs(tag="(a)"),
                        p.rateratio.hosp + labs(tag="(b)") +
                        theme(legend.position = "none"))

```

```{r bytimesincedose, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=1.3, fig.cap="\\label{fig:bytimesincedose} Relation of vaccine efficacy to time since last dose: (a) severe COVID-19; (b) hospitalised or fatal COVID-19.  Rate ratios in conditional logistic regression model, adjusted for covariates. Thickness of each line is proportional to precision of estimate"}

theme_set(theme_gray(base_size = 10))

gridExtra::grid.arrange(p.rateratio.sincedose + labs(tag="(a)"),
                        p.rateratio.sincedose.hosp + labs(tag="(b)") +
                        theme(legend.position = "none"))

```

# Supplementary Tables
\beginsupplement

```{r severe, echo=FALSE, message=FALSE}

knitr::kable(table.severe, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="severe",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             col.names=c(clean.header(colnames(table.severe)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Rate ratios for severe COVID-19") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 28 July 2021.",
                         "Vaccination status defined as number of doses received at least 14 days before presentation date.",
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Multivariable model includes all covariates in the table"), 
                 notation="none") %>%
    group_rows("Vaccination status", 7, 11, bold=FALSE) %>%
    column_spec(1, width="4cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="3cm") %>%
    column_spec(7, width="2cm") %>%
    kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

```{r hosp, echo=FALSE, message=FALSE}

knitr::kable(table.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="hosp",
             row.names=TRUE, 
			 align=c('r', 'r'), 
             col.names=c(clean.header(colnames(table.hosp)[1:2]),
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value",
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value"), 
             caption="Rate ratios for hospitalised or fatal COVID-19") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariable"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 28 July 2021.",
                         "Vaccination status defined as number of doses received at least 14 days before presentation date.",
                         "Controls matched for age, sex, and primary care practice on date of presentation of case.", 
                         "Multivariable model includes all covariates in the table"), 
                 notation="none") %>%
    group_rows("Vaccination status", 7, 11, bold=FALSE) %>%
    column_spec(1, width="4cm") %>%
    column_spec(2, width="3cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="3cm") %>%
    column_spec(7, width="2cm") %>%
    kable_styling(latex_options=c("HOLD_position", "scale_down"))

```

```{r changepoints, echo=FALSE, message=FALSE}

knitr::kable(loglik.changepoints[, .(week.changepoint, deviance, effect.before, effect.after)],
             escape=FALSE, 
             booktabs=TRUE,
			 label="changepoints",
             row.names=FALSE, 
			 # align=c('r', 'r'), 
             col.names=c("Change-point (weeks since second dose)",
                         "Deviance of model",
                         "Before change-point",
                         "After change-point"),
             caption="Relationship of efficacy of two doses of vaccine against severe COVID-19 to time since second dose: comparison of models with change-points at 8 to 16 weeks since second dose") %>%
    add_header_above(c(" "=2, "Effect (95% CI) per week on rate ratio"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 28 July 2021.",
                         "Restricted to those who had received 0 doses or 2 doses at least 14 days before presentation date.",
			 "Variables in model: care home residence + risk category + number of drug classes + vaccination status (0 doses, 2 doses mRNA, 2 doses AZ) + any vaccine x weeks since dose", 
                         "Slopes of relationship before and after changepoint are encoded as  interaction terms for vaccination status x weeks since second dose.",
                         "Effects shown are multiplicative effects on the rate ratio associated with vaccination status, obtained by taking exponents of the coefficients for the interaction terms.",
                         "Deviance is scaled to have value zero for a model with no changepoint.",
                         "Model with best fit (lowest deviance) is shown in bold."),
                 notation="none") %>%
    column_spec(1, width="3cm") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    row_spec(2, bold=TRUE)

```

```{r changepointshosp, echo=FALSE, message=FALSE}

knitr::kable(loglik.changepoints.hosp[, .(week.changepoint, deviance, effect.before, effect.after)],
             escape=TRUE, 
             booktabs=TRUE,
			 label="changepointshosp",
             row.names=FALSE, 
			 # align=c('r', 'r'), 
              col.names=c("Change-point (weeks since second dose)",
                         "Deviance of model",
                         "Before change-point",
                         "After change-point"),
             caption="Relationship of efficacy of two doses of vaccine against hospitalised or fatal COVID-19 to time since second dose: comparison of models with change-points at 8 to 16 weeks since second dose") %>%
    add_header_above(c(" "=2, "Effect (95% CI) per week on rate ratio"=2)) %>%
    add_footnote(label=c("Presentation dates from 1 December 2020 to 28 July 2021.",
                         "Restricted to those who had received 0 doses or 2 doses at least 14 days before presentation date.",
			 "Variables in model: care home residence + risk category + number of drug classes + vaccination status (0 doses, 2 doses mRNA, 2 doses AZ) + any vaccine x weeks since dose", 
                         "Slopes of relationship before and after changepoint are encoded as  interaction terms for vaccination status x weeks since second dose.",
                         "Effects shown are multiplicative effects on the rate ratio associated with vaccination status, obtained by taking exponents of the coefficients for the interaction terms.",
                         "Deviance is scaled to have value zero for a model with no changepoint.",
                         "Model with best fit (lowest deviance) is shown in bold."),
                 notation="none") %>%
    column_spec(1, width="3cm") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="3cm") %>%
    column_spec(4, width="3cm") %>%
    row_spec(2, bold=TRUE)

```

### 2021-11-02 cvst/cvst_medrxiv.Rmd
updated with hardcoded chi numbers sourced from separate fi

---
title: | 
   | Association of cerebral venous thrombosis with recent COVID-19 vaccination: case-crossover study using ascertainment through neuroimaging in Scotland 
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: Raj Burgul
    affiliation: FV 
  - name: Jen Bishop
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Jim McMenamin 
    affiliation: HPS
  - name: Maureen O'Leary
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    corresponding: yes
    email: helen.colhoun@ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. 
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Professor of Clinical Epidemiology and Medical Informatics 
- code: FV
  address: Forth Valley Royal Hospital, Larbert FK5 4WR, Scotland. RB - Consultant Radiologist
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Cancer, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
  \usepackage{colortbl}
  \usepackage{soul}
output: 
  rticles::plos_article:
    includes:
      in_header: ../preamble.tex
    latex_engine: lualatex
    number_sections: false
    keep_tex: true
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ../covid.bib
csl: ../plos.csl
always_allow_html: true
urlcolor: blue

---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

linesep <-function(x,y=character()){
  if(!length(x))
    return(y)
  s <- rep('', max(sep.at))
  s[sep.at] <- "\\addlinespace"
  return(s[-1])
}

options(knitr.kable.NA = '.')
table.vax.bymonth <- with(vaxpop,
                          table(paste0(lubridate::year(vaxdate), " ", 
                                       lubridate::month(vaxdate, #label=TRUE,
                                                        abbr=TRUE)),
                                vacc_product_name, exclude=NULL))
table.vax.bymonth <- rbind(table.vax.bymonth, colSums(table.vax.bymonth))
table.vax.bymonth <- cbind(table.vax.bymonth, rowSums(table.vax.bymonth))
rownames(table.vax.bymonth)[nrow(table.vax.bymonth)] <- "Total doses"
rownames(table.vax.bymonth)[nrow(table.vax.bymonth) - 1] <- "Missing date"
colnames(table.vax.bymonth)[ncol(table.vax.bymonth)] <- "Total doses"

vaxpop[, agegr.vax := cut(age.vax, c(0, 40, 60, 120), right=FALSE,
                          labels=c("0-39", "40-59", "60 or more"))]

table.vax.byage <- with(vaxpop,
                          table(agegr.vax, vacc_product_name, exclude=NULL))
table.vax.byage <- cbind(table.vax.byage, rowSums(table.vax.byage))
colsums.vax.byage <- colSums(table.vax.byage)
table.vax.byage <- paste.colpercent(table.vax.byage)
table.vax.byage <- rbind(table.vax.byage, colsums.vax.byage)
rownames(table.vax.byage)[nrow(table.vax.byage)] <- "Total doses"
colnames(table.vax.byage)[ncol(table.vax.byage)] <- "Total doses"

table.byyear <- with(cvst.disch, table(lubridate::year(ADMISSION_DATE), SEX))
All <- rowSums(table.byyear)
table.byyear <- paste.rowpercent(table.byyear)
table.byyear <- data.frame(as.data.frame.matrix(table.byyear), All)

setorder(cvst.leftjoin, ADMISSION_DATE)
table.disch <- cvst.leftjoin[ADMISSION_DATE >= as.Date("2020-12-01"),
                             .(chi, ADMISSION_DATE, study_date, site_name, SEX, age.seen,
                               MAIN_CONDITION,
                               hfeventcode, scancheck)]
table.disch[, agegr20 := floor(age.seen / 20)]
table.disch[, agegr20 := car::recode(agegr20,
                                 "0='0-19'; 1='20-39'; 2='40-59'; 3='60-79'; 4='80-'")]
table.disch <- table.disch[,
                           .(ADMISSION_DATE, study_date, site_name, SEX, agegr20,
                             MAIN_CONDITION,
                               hfeventcode, scancheck)]
table.disch <- as.data.frame(table.disch)

scans.definite <- scans.possible[hfeventcode=="primary acute"] 
setkey(scans.definite, chi)
setkey(cvst.disch, CHI)

scans.leftjoin <- cvst.disch[scans.definite] ## 49 records of which 20 have cvst code as main diagnosis
scans.leftjoin <- unique(scans.leftjoin, by="CHI")
scans.leftjoin <- admissions.other.leftjoin[scans.leftjoin] # 49 records
scans.leftjoin[is.na(ADMISSION_DATE), ADMISSION_DATE := admissiondate.other]
scans.leftjoin[, CVT.mention := !is.na(MAIN_CONDITION)]
scans.leftjoin[is.na(MAIN_CONDITION), MAIN_CONDITION := diagnosis.other]
scans.leftjoin <- unique(scans.leftjoin, by="chi")
setorder(scans.leftjoin, study_date)

chi.scans <- scans.leftjoin[hfeventcode=="primary acute" & !is.na(MAIN_CONDITION) & is.na(diagnosis.other), chi]
chi.smr <- cvst.leftjoin[hfeventcode=="primary acute", chi]

table.scans <- scans.leftjoin[hfeventcode=="primary acute" &
                                            study_date >= as.Date("2020-12-01"),
                              .(ADMISSION_DATE, study_date, site_name,
                                sex,
                                age.scan, hfeventcode, MAIN_CONDITION,
                                diagnosis.other)]
table.scans[!is.na(MAIN_CONDITION) & is.na(diagnosis.other), smr.status := "CVT diagnosis"]
table.scans[!is.na(MAIN_CONDITION) & !is.na(diagnosis.other), smr.status := "Other diagnosis"]
table.scans[is.na(MAIN_CONDITION), smr.status := "No coded discharge record"]
table.scans[, smr.status := factor(smr.status, 
                                   levels=c("CVT diagnosis", "Other diagnosis", "No coded discharge record"))]

c.chi <- cvst.leftjoin[ADMISSION_DATE >= as.Date("2020-12-01") & hfeventcode=="primary acute", chi]
s.chi <- scans.leftjoin[!is.na(MAIN_CONDITION) & hfeventcode=="primary acute", chi]

##################################################
events.tw <- with(cvst.all[is.na(hfeventcode) | hfeventcode=="primary acute"],
                  table(vacc_product_name, ceiling(daysfromvax / 14)))
productnames <- rownames(events.tw)
events.tw <- cbind(events.tw[, 1:2], rowSums(events.tw[, 3:5]))
colnames(events.tw) <- c("1-14", "15-28", "29-84")
ratepermillion <- data.table(as.data.frame.matrix(round(1E6 *
                                                        events.tw / pop.atrisk[, -1],
                                                        1)))
ci <- credint95.ratepermillion(events.tw[, 1], pop.atrisk[, 2])
ratepermillion[, `1-14` := paste(`1-14`, ci)]
rateratio.mle <- c(
    mle.rateratio(as.numeric(events.tw[1, 1]), as.numeric(sum(events.tw[1, 2:3])),
                  as.numeric(pop.atrisk[1, 2]), as.numeric(sum(pop.atrisk[1, 3:4]))),
    mle.rateratio(as.numeric(events.tw[2, 1]), as.numeric(sum(events.tw[2, 2:3])),
                  as.numeric(pop.atrisk[2, 2]), as.numeric(sum(pop.atrisk[2, 3:4]))),
        NA)
rateratio.ci <- c(
    credint.rateratio(as.numeric(events.tw[1, 1]), as.numeric(sum(events.tw[1, 2:3])),
                  as.numeric(pop.atrisk[1, 2]), as.numeric(sum(pop.atrisk[1, 3:4]))),
    credint.rateratio(as.numeric(events.tw[2, 1]), as.numeric(sum(events.tw[2, 2:3])),
                      as.numeric(pop.atrisk[2, 2]), as.numeric(sum(pop.atrisk[2, 3:4]))),
        NA)
rr <- paste0(rateratio.mle, " (", rateratio.ci, ")")
rr[3] <- "."
ratepermillion <- cbind(ratepermillion, rr)

#############################################
events.possible.tw <- with(cvst.all, table(vacc_product_name, ceiling(daysfromvax / 14)))
events.possible.tw <- cbind(events.possible.tw[, 1:2], rowSums(events.possible.tw[, 3:5]))
colnames(events.possible.tw) <- c("1-14", "15-28", "29-84")
ratepermillion.possible <- data.table(as.data.frame.matrix(round(1E6 *
                                                                 events.possible.tw /
                                                                 pop.atrisk[, -1], 1)))
ci <- credint95.ratepermillion(events.possible.tw[, 1], pop.atrisk[, 2])
ratepermillion.possible[, `1-14` := paste(`1-14`, ci)]
rateratio.mle <- c(
    mle.rateratio(as.numeric(events.possible.tw[1, 1]),
                  as.numeric(sum(events.possible.tw[1, 2:3])),
                  as.numeric(pop.atrisk[1, 2]), as.numeric(sum(pop.atrisk[1, 3:4]))),
    mle.rateratio(as.numeric(events.possible.tw[2, 1]),
                  as.numeric(sum(events.possible.tw[2, 2:3])),
                  as.numeric(pop.atrisk[2, 2]), as.numeric(sum(pop.atrisk[2, 3:4]))),
    NA)
rateratio.ci <- c(
    credint.rateratio(as.numeric(events.possible.tw[1, 1]),
                      as.numeric(sum(events.possible.tw[1, 2:3])),
                      as.numeric(pop.atrisk[1, 2]), as.numeric(sum(pop.atrisk[1, 3:4]))),
    credint.rateratio(as.numeric(events.possible.tw[2, 1]),
                      as.numeric(sum(events.possible.tw[2, 2:3])),
                      as.numeric(pop.atrisk[2, 2]), as.numeric(sum(pop.atrisk[2, 3:4]))),
    NA)
rr <- paste0(rateratio.mle, " (", rateratio.ci, ")")
rr[3] <- "."
ratepermillion.possible <- cbind(ratepermillion.possible, rr)

############################################

events.tw <- as.data.table(as.data.frame.matrix(events.tw))
for (j in 1:3) set(events.tw, j = j, value = as.character(round(events.tw[[j]])))
events.possible.tw <- as.data.table(as.data.frame.matrix(events.possible.tw))
for (j in 1:3) set(events.possible.tw, j = j,
                   value = as.character(round(events.possible.tw[[j]])))

table.tw <- data.table(rbind(round(pop.atrisk[, -1]),
                             events.tw, 
                             ratepermillion,
                             events.possible.tw,
                             ratepermillion.possible,
                             fill=TRUE))
table.tw <- data.table(Product=rep(productnames, 5), table.tw)

credint95.pf <-  credint95.ratepermillion(2, as.numeric(pop.atrisk[1, 2])) 
credint95.az <- credint95.ratepermillion(events=7, popatrisk=as.numeric(pop.atrisk[2, 2]))

printdate <- function(date) {
    gsub("^0", "", format(date, "%d %B %Y"))
}

eventsunder60.tw <- with(cvst.all[age.scan < 60 &
                                  (is.na(hfeventcode) | hfeventcode=="primary acute")],
                  table(vacc_product_name, ceiling(daysfromvax / 14)))
credint95.az.28day.under60 <-
    credint95.ratepermillion(events=8,
                             popatrisk=as.numeric(popunder60.atrisk[2, 2] +
                                                  popunder60.atrisk[2, 3]))

fci <- function(x) {
    gsub("\\(", "\\(95\\% credible interval ", x)
}

```
\newpage

# Abstract
**Objectives** - To investigate the association of primary acute cerebral venous thrombosis (CVT) with COVID-19 vaccination through complete ascertainment of all diagnosed CVT in the population of Scotland.

**Design** - Case-crossover study comparing recent (1-14 days after vaccination) with less recent exposure to vaccination among cases of CVT.

**Setting** - National data for Scotland from 1 December 2020, with diagnosed  CVT case ascertainment through neuroimaging studies up to 17 May 2021 and diagnostic coding of hospital discharges  up to 28 April 2021 and with linkage to vaccination records.

**Main outcome measure** - Primary acute cerebral venous thrombosis

**Results** Of 50 primary acute CVT cases, 29 were ascertained only from neuroimaging  studies, 2 were ascertained only from hospital discharges, and 19 were ascertained from both sources.  Of these 50 cases, 14 had received the Astra-Zeneca ChAdOx1 vaccine and 3 the Pfizer BNT162b2 vaccine.  The incidence of CVT per million doses in the first 14 days after vaccination was `r fci(table.tw[8, 2])` for  ChAdOx1  and `r fci(table.tw[7, 2])` for  BNT162b2.  The rate ratio for CVT associated with exposure to ChAdOx1 in the first 14 days compared with exposure 15-84 days after vaccination was `r fci(table.tw[8, 5])`.

**Conclusions** - These findings support a causal association between CVT and the AstraZeneca vaccine.  The absolute risk of post-vaccination CVT in  this  population-wide study in Scotland  was lower than has been reported for populations in Scandinavia and Germany; the explanation for this is not clear.

\newpage

\begin{myquote}
\subsection{What is already known on this topic}

The risk of cerebral venous  thrombosis (CVT) within 28 days of receiving the AstraZeneca ChAdOx1 vaccine has been estimated as 18 to 25 per million doses in Germany and Scandinavia, but only 5 per million doses in the UK based on the Yellow Card reporting scheme.  Risk estimates based on adverse event reporting systems are subject to under-ascertainment and other biases.

\subsection{What this study adds}

All diagnosed cases of CVT in Scotland were ascertained by searching neuroimaging studies from December 2020 to May 2021 and linked to national vaccination records.  The risk of CVT within 28 days of vaccination with ChAdOx1 was estimated as 3.5 per million doses with an upper bound of 6 per million doses, against a background incidence of about 12 per million adults per year.  This indicates that the Yellow Card system has not seriously underestimated the risk in the UK; the explanation for higher risk in other European countries is not clear.

\end{myquote}

# Introduction
The first reports in international media  of an association of COVID-19 vaccines with thrombotic events  appeared on 7 March 2021, when the Austrian Federal Office for Safety in Health Care announced that it had suspended use of a batch  of AstraZeneca ChAdOx1 vaccine  after cases of thromboembolic events.  On 7 April 2021 the UK Medicines and Healthcare products Regulatory Agency (MHRA) and the Joint Committee on Vaccines and Immunisation (JCVI) concluded that there was "a possible link"  between the vaccine and cerebral venous thrombotic events   and issued new guidance [@medicines_and_healthcare_products_regulatory_agency_mhra_2021].  By 18 May 2021 two countries in the EU/ EEA had discontinued use of the vaccine and 15 had limited its use to older age groups [@ecdc_overview_2021].   In the US cerebral venous thrombosis has  also been reported following administration of the AD26.COV2.S Johnson & Johnson (JJ) vaccine  which like the AstraZeneca ChAdOx1 vaccine utilises recombinant adenoviral vectors encoding the SARS-CoV-2 spike protein [@see_us_2021].  An underlying syndrome denoted "vaccine-induced immune thrombotic thrombocytopenia" (VITT) [@schultz_thrombosis_2021; @greinacher_thrombotic_2021]  or "thrombosis with thrombocytopenia syndrome" (TTS)  [@chen_interim_2021] was described, consisting of thrombosis, thrombocytopenia and platelet-activating antibodies to platelet factor 4–polyanion complexes.  It was recognized that the spectrum of this syndrome may include thrombocytopenia alone or thrombosis without thrombocytopenia.

Risk / benefit assessments of the use of the  AstraZeneca ChAdOx1 vaccine depend on estimating the incidence of CVT and TTS syndrome post-vaccination and evaluating evidence for causality.   However, most such estimates have come from voluntary reporting schemes such as the  MHRA Yellow Card scheme.  These reporting schemes  have limitations: vaccine-associated cases may be under-ascertained if the connection with vaccine exposure is not made by the patient or clinician and,  once there  has been publicity about any possible association with a particular medicine,  reporting can be biased thereafter. Although studies based on linking vaccination records to hospital discharge records ascertained either directly[@pottegard_arterial_2021] or via primary care [[@simpson_firstdose_2021] are less subject to reporting bias, these estimates of incidence depend on the accuracy of coding of hospital diagnoses.  Therefore here we took advantage of a nationwide database  that captures radiologists' reports of all imaging studies conducted in the National Health Service (NHS) in Scotland to obtain an unbiased and comprehensive ascertainment of  diagnosed CVT since the start of the vaccination programme, linked to the national vaccination database.  The aim of this study was to understand the epidemiology of CVT and its association with COVID-19 vaccination. The specific objectives were:

1. to estimate incidence of diagnosed primary acute CVT in Scotland  during the vaccination programme  
2. to ascertain whether diagnostic coding of hospital discharges  or deaths adequately captures diagnosed CVT and allows differentiation of secondary and chronic cases from primary acute cases.  
3. to investigate the association of diagnosed CVT with vaccination.

# Methods
In brief, we retrieved  all potentially relevant scan reports from the nationwide Picture Archiving and Communication System (PACS) to identify diagnoses of CVT from the  study start date of  1 December 2020  up to 17 May 2021.  The scan reports contain a summary of the clinical findings, the radiologist's initial report, and a review by the consultant if the first radiologist was not a consultant.  \hl{To allow the sensitivity and specificity of discharge coding to be assessed against scan reports as ground truth, we also retrieved all hospital discharge  and death records with diagnostic codes for CVT in this time period.}  As all health care records in Scotland use the Community Health Index (CHI) as a unique identifier, we were able to link PACS reports and hospital discharge  and death records to the daily updated national  vaccination programme database held by Public Health Scotland/ National Services Scotland.  We pre-specified a case-crossover design that would compare exposure to vaccine in a recent time window (1 to 14 days since vaccination) with exposure in a less recent time window. The methods are described in detail below.

## Vaccination doses
The COVID-19 vaccination programme in Scotland began on 8 December 2021.   Initial priority groups included care home residents and staff, front line health and social  care workers and  clinically extremely vulnerable individuals.  Initially the Pfizer BNT162b2 product  (hereafter Pfizer vaccine) was used. From 8 January 2021 the  AstraZeneca ChAdOx1 product  (hereafter AstraZeneca vaccine) was introduced and from 7 April 2021 the Moderna mRNA-1273 product (hereafter Moderna vaccine)  was introduced. CHI number, date of administration, age at vaccination, vaccine product name, whether first or second dose were extracted for  all `r nrow(vaxpop)` vaccination records from `r printdate(min(vaxpop$vaxdate, na.rm=TRUE))` to `r printdate(max(vaxpop$vaxdate, na.rm=TRUE))`  from the national vaccination database.

## Ascertainment of CVT from CT scan and MRI reports
The key first-line investigation for patients presenting with any acute neurological syndrome is a non-contrast CT scan of the head.  However, this is positive in only about 30% of patients with CVT [@saposnik_diagnosis_2011]. For this reason current guidance recommends CT or MR venography in all patients where CVT is suspected  [@american_college_of_radiology_acr_2016;@ferro_european_2017] unless the initial CT shows strong evidence of CVT and a venogram is contra-indicated.

Fig \ref{flow} shows a flow diagram of the process of  neuroimaging report extraction and processing. In the pilot stage of the project  13 April 2021 we initially restricted the query of the PACS database to records where the study type (RIS_CODE field) was coded as CT or MR venogram. Comparison with hospital discharge records indicated that  this query was insensitive.  After discussion with radiologists the query was therefore broadened to extract reports with a wider range of potentially relevant study type codes as listed in Supplementary Material and any reports with a study modality of CT or MRI where the study description field  contained any of the strings "head", "brain" or "cerebr".  All such reports of CT and MRI studies  from 1 Dec 2020 to 17 May 2021 were retrieved from the PACS database.

For three of the 14 health boards in Scotland, the PACS database contained the radiology report only as an image that could not be queried.  For these boards  extractions were carried out directly from the local radiology information system that feeds the national PACS instead. For Lanarkshire (12.1% of the Scottish population)  the same query was run as for the national PACS .  From the two other boards that do not encode reports as text in  PACS -- Forth Valley and Dumfries & Galloway, covering 8.3% of the Scottish population -- only the initial RIS code query for venograms up to 13 April 2021 was completed; a subsequent broader query could not be obtained at the time of this study.

The extracted scan reports were filtered to retain any scan likely to be informative for the presence of CVT i.e.  all those with a study type or study description including a venogram or where any of the following strings appeared in the the radiologist's report: "sinus thrombosis", "sinus thrombus", "venous thrombosis", "venous thrombus", "venogr".   For any individual  with at least one scan meeting these criteria all scan reports for that person since 1 Dec 2020 were retained for review.

After arraying scans for each individual chronologically, each clinical event was scored by a doctor as primary acute, possible, follow-up, chronic, no valid result, secondary or negative (see supplementary material).  A new clinical event was defined by a new illness leading to a hospital visit.   The date of onset of the event was assigned as the date of onset of symptoms where this was recorded in the scan report, otherwise as the earliest date of hospital admission or scan. \hl{The reviewing doctor (HC) had no access to vaccination status except where a mention of this was embedded in the scan report. A second doctor (DM with adjudication by RB) re-scored all scans that had been coded non-negative and a random sample of 30 negative scans, with the scoring of the first reviewer masked.}

## Ascertainment of CVT from discharge diagnoses and death certificates
Records of all hospital discharges in Scottish Morbidity Record 01 (SMR01) from 1 December 2020 to `r max(smr01$DISCHARGE_DATE)` and National Register of Scotland death registrations from 1 December 2020 to `r max(deaths15$DATE_OF_DEATH)` were queried for any mention of a CVT code as discharge diagnosis or cause of death.   The ICD-10 codes used were: I63.6 (cerebral infarction due to cerebral venous thrombosis, nonpyogenic); I67.6 (nonpyogenic thrombosis of intracranial venous system); and G08 (intracranial and intraspinal phlebitis and thrombophlebitis).  Records with mention of a local secondary cause such as infection, intracranial abscess or brain tumour were excluded by matching of the regular expression `^A|^B|^C7[012]|^D3[23]|G06[02]` in any of the diagnostic codes.

## Statistical methods
The case-crossover design compares event rates in time windows of recent and less recent exposure; this is equivalent to a matched case-control study in which the case and the control are the same person in different time windows [@maclure_should_2000].  We pre-specified the time window for recent exposure to vaccine as 1 to 14 days before the onset of symptoms as given in the summary of the clinical history in the scan report.

To allow comparison with other studies that have used a 28-day time window, we also tabulated events and rates for the period 15-28 days.  All post-vaccination cases were within 84 days of vaccination, so for the case-crossover analysis the 1-14 day time window is compared with the 15-84 day time window.  For each vaccine product and each time window, the number of person-fortnights at risk was calculated to the end of follow up.   To allow for the variation between health boards in latest date of case ascertainment via scans, the person-days in each exposure category from 14 April to 17 May 2021 were multiplied by 0.917 (the proportion of the population not covered by Forth Valley or Dumfries & Galloway  where only records up to 14 April were extracted) and the person-days from 17 May 2021 to 1 June 2021 were multiplied by 0.121 (the proportion covered by Lanarkshire where records up to 1 June 2021 were extracted).

As the numbers of observed events were too small for confidence intervals to be equivalent to credible intervals, Bayesian credible intervals were calculated directly from the quantiles of the posterior distribution.  With a flat prior on the logarithm of the event rate and \ensuremath{r} observed events in \ensuremath{n} million person-fortnights, the posterior distribution of the event rate is a gamma distribution with shape parameter \ensuremath{r} and inverse scale parameter \ensuremath{n}.   Conditional on the total number of events, the likelihood given \ensuremath{r_1} events in \ensuremath{n_1} person-days in the recently exposed time window and \ensuremath{r_0} events in \ensuremath{n_0} person-days in the less recently exposed time window is \ensuremath{p^{r_1}(1 - p)^{r_0}} where \ensuremath{p = n_1 e^\beta / (n_0 + n_1 e^\beta)} and \ensuremath{\beta} is the log rate ratio.  With a flat prior on the log rate ratio \ensuremath{\beta}, normalizing this likelihood as a function of \ensuremath{\beta} gives the posterior distribution of \ensuremath{\beta}.

# Results
## CVT case ascertainment
As shown in Supplementary Figure \ref{flow}, 142964 neuroimaging reports  pertaining to 80905 individuals were retrieved.  After de-duplication and  retention of all venogram reports,  and any reports containing keywords  as described in the Methods section, 2760 scans pertaining to 1607 individuals were retained for  manual coding.   Table \ref{tab:scansevents} shows the codes assigned to these individuals after review.  The independent coding validation by a second masked coder resulted in only one reassignment where a case with metastatic cancer but no brain metastases was reassigned as primary acute rather than secondary.  48 cases ascertained through scans were coded as having a primary acute CVT after 1 December 2020, and 18 were coded as "possible CVT".  No additional cases in the study period were ascertained through death certificates with CVT codes.

The yield of cases  through  SMR01 discharge reports  up  to the latest discharge in the SMR01 extract is shown in Table \ref{tab:dischbyscan}  In the extract used for this study there were 34 individuals with first mention after 1 December 2020 of a CVT code, of whom 29 had no mention of a secondary code on any discharge record. Of these 29, 7 had neuroimaging reports that showed a  secondary cause, chronic thrombus,  or stenosis rather than thrombus.  Only 3 of the 29 events ascertained from SMR01 discharge records had no neuroimaging report retrieved by the PACS query.  Of these one had an unrelated condition as main diagnosis, one had a CVT code as main diagnosis but had stayed only one night in hospital, and one had a CVT code as main diagnosis and a scan record but no report, from one of the two health boards that did not include the text of scan reports in the PACS database.  On the basis that CVT diagnostic code as main condition had high specificity (14/17) the two cases with CVT coded as main condition were retained as primary acute. As none of these cases without scan reports were post-vaccination, their inclusion or exclusion does not affect the case-crossover results.

Table \ref{tab:scansbydisch} shows that of the 48 primary acute CVT cases ascertained through  neuroimaging, 19 had a discharge record with a CVT code, 8 had a discharge record with no CVT code, and 21 had no discharge record.  Note that as coding  of hospital admissions is done after discharge with varying lags, no exact cutoff  date for the latest date of admission of cases ascertained through SMR01 can be defined.  However  as shown in \ref{tab:scansbydisch} of 35  primary acute CVT case ascertained through PACS up to the end of the first quarter of 2021 ( a period probably fully captured by the SMR01 extraction), only 17 had had a CVT diagnosis code on an SMR01 record.

Combining the 49 cases ascertained through scans and scored as primary acute with the two additional cases with CVT coded as main diagnosis that had been ascertained only through SMR01, 50 cases classified as primary acute were retained for analysis, together with 18 scan-ascertained cases coded as possible.

## Relation of CVT to vaccine exposure
Table \ref{tab:vaxbyage} shows the number of doses by age group and vaccine product  from 1 Dec 2020 to `r max(vaxpop$vaxdate, na.rm=TRUE)`.  The vaccine product was missing for about 0.4% of recorded doses.

Table \ref{tab:tw} tabulates the post-vaccination cases by vaccine product and time since vaccination.  As pre-specified, the time window of 1 to 14 days was used to define recent exposure.  All other events in vaccinated individuals occurred between 15 and 84 days after vaccination.  Case-crossover estimates of rate ratio were based on comparing the 1-14 day time window with the 15-84 day time window. In each time window, the number of person-fortnights exposed in that time window is the denominator from which the rate is calculated.

17 definite  CVT events and 25 definite or possible CVT events occurred post vaccination: the others were in individuals who had not been vaccinated at the time of onset.  Of the 7 cases with onset within 14 days of exposure to AstraZeneca vaccine, 6 were in people aged less than 60 years and 4 were in women.

The observed rate of definite (primary acute) CVT diagnoses in the 14 days after exposure to  AstraZeneca vaccine is estimated as `r fci(table.tw[8, 2])` per million doses.  For the Pfizer vaccine, the corresponding rate was `r fci(table.tw[7, 2])` per million doses.  There were no cases in the small numbers exposed to Moderna vaccine.  For comparison, the observed rate of primary acute CVT in the 15-84 day time window after exposure was 12 per million per year (95% credible interval `r credint95.ratepermillion(8, (1784505 + 4942003+2698048+7390991+17801 + 21013) / 26)`.  The rate ratio associated with exposure in the last 14 days to the AstraZeneca vaccine was estimated as `r fci(table.tw[8, 5])` for primary acute CVT, and `r fci(table.tw[14, 5])` for primary acute or possible cases.

For comparison with other studies that used a 28-day time window, the rate of primary acute CVT within 28 days of vaccination with the AstraZeneca product, based on 9 events in `r p28days` 28-day time intervals, was 3.1 (95% credible interval  `r credint95.ratepermillion(9, p28days)` per million doses. In those aged under 60 years the rate, based on 8 events in `r p28days.under60` 28-day time intervals was 7 (95% credible interval `r credint95.ratepermillion(9, p28days.under60)` per million doses.

## Checking for bias in case ascertainment
Fig \ref{fig:scansbydate} shows the total number of venograms (panel a) and other head scans (panel b) in boards  with complete ascertainment using PACS,  by 7-day sliding window of study date.  The average number of venogram studies was about 50 /day from December 2020 to March 2021, but increased sharply in the first two weeks of April to a peak of nearly 150 /day day in mid-April, around the time that the association of CVT with vaccination first received wide publicity in the UK.  However the number of individuals exposed to the AstraZeneca vaccine in the last 14 days had peaked earlier in late March.  Of the 9 cases that occurred within 28 days of vaccination with the AstraZeneca product, 6 were before April 2021, when the increase in number of venograms per day began.

# Discussion 
## Statement of principal findings
In this population-wide study we have captured all diagnosed CVTs nationally since before the start of the vaccination programme.  We have shown that ascertainment of CVTs from diagnostic codes on hospital discharge records has low sensitivity for CVT; systematic review of scan reports is required for complete ascertainment of diagnosed cases and to distinguish primary acute from secondary cases.  The number of events in those recently exposed to vaccine is small but we can still calculate a credible interval for the incidence rate and thus an upper bound.   For the first 14 days after AZ vaccination, we calculate an upper bound of 4.9 per million doses on the absolute rate,  and 9.1 for the rate ratio.  Although the rate of venogram studies increased markedly in early April 2021 after the association of CVT with vaccination had received wide publicity, most of the vaccine-associated cases in this study were diagnosed before this, and thus it is unlikely that ascertainment could have biased the association with recent vaccine exposure.   When combined with prior evidence of excess risk of CVT associated with the AstraZeneca vaccine in other populations, the case-crossover analysis supports a causal interpretation of this association.  Although the number of cases exposed to the Pfizer vaccine is too small for any comparison between the two products, an upper bound can still be placed on the absolute risk associated with the Pfizer vaccine.

## Strengths and limitations of this study
A strength of this study is the complete ascertainment of cases in the population, not reliant on adverse event reporting or diagnostic coding of hospital discharge records.  Manual review of scan reports, which typically included a summary of the clinical history allowed some CVT cases to be recoded as secondary to a local lesion such as a tumour or as chronic rather than acute CVT, and allowed the date of onset to be determined from the summary of the clinical history.  Accurate assignment of date of onset is required for the case-crossover analysis to be valid.   \hl{Frequently the onset of symptoms was a few days before presentation to hospital.  The time window for onset of vaccine-associated CVT may thus be overestimated in studies that have reported the median time from vaccination to hospitalisation to be more than 10 days} [hwang_comparison_2021;@pavord_clinical_2021].

The case-crossover design  eliminates confounding by time-invariant factors, which may be strong where vaccine allocation is based on pre-existing risk conditions that were used to allocate priority for vaccination, and thus provides evidence for causality where an association is detected.  Note that our analysis assumes an ignorance prior; it ignores any prior evidence of association.  Unlike the self-controlled case series design, which has also been used to study the associations of adverse events with COVID-19 vaccine [@hippisley-cox_risk_2021], the case-crossover design can be used to study events such as CVT that are likely to affect the probability of being vaccinated within the next few weeks.

Limitations of our study are that we do not have access to data on platelet counts, D-dimer or platelet factor 4 antibody levels, allowing enumeration of the number of CVT events that are part of the formally defined VITT / TTS syndrome as meeting associated  haematological criteria.  We note that the Brighton Collaborative however acknowledges that the TTS syndrome definition may be too restrictive by excluding isolated thrombotic events that are causally related to vaccine.  Also we did not ascertain venous thromboses at sites other than brain.

## Relation to other studies
From this study the  background incidence of primary acute CVT in Scotland (excluding the 14-day time window after vaccination) can be estimated as about 12 per million adults per year: this is similar to the estimate of 16 per million per year in Australia in 2016 [@devasagayam_cerebral_2016] based on ascertainment of cases via neuroimaging records.  Most other estimates of CVT incidence have relied on ascertainment via diagnostic coding in health informatics systems, which is likely to underestimate the incidence of CVT.  A recent study from Scotland based on primary care, hospitalisation and death records in the total population reported 19 CVT events between 8 December 2020 and 14 April 2021 [@simpson_firstdose_2021].  For the same period our study ascertained 41 primary acute cases of CVT in Scotland.

Adverse event reporting systems can give early warning of unexpected effects but cannot be relied on to estimate incidence rates. By early April 169 CVTs  had been reported to the EMA by which time about 34 million doses of Astra Zeneca vaccine had been administered, giving a rate of about 5 per million doses.  In the UK as of 9 June 2021 the Yellow Card scheme operated by the MHRA had received 390 reports of cases  of "major thromboembolic events with concurrent thrombocytopenia". Cerebral venous sinus thrombosis was reported in 140 of these.  By 2 June 2021 30.1 million doses had been administered, so allowing for a one-week lag the incidence was 13.0 per million doses for the syndrome and 4.7 per million doses for CVT [@medicines__healthcare_products_regulatory_agency_coronavirus_2021].  \hl{A subset of 102 of these cases were reported in more detail by haematologists from 96 hospital trusts detailed study: the denominator was estimated as 24 million doses, giving a rate of 4.3 per million doses in the first 30 days} [@pavord_clinical_2021].  Our estimate of 3.5 per million doses in the first 28 days is similar to these estimates, indicating that the Yellow Card system, for all its limitations, has not seriously underestimated the incidence of vaccine-associated CVT.

The upper bound of 6.2 per million doses for incidence of CVT within 28 days of vaccination with the AstraZeneca product estimated in this study is rather lower than estimates reported from Scandinavia and Germany.  In a study of all 281264 individuals aged 18-65 years in Denmark and Norway who received a first dose of the AstraZeneca vaccine, there were were 7 observed CVT events within 28 days of vaccination compared with 0.3 expected from rates in the general population [@pottegard_arterial_2021]: a rate of 25 per million doses.  A study of CVT events ascertained through neurologists from nine states in Germany where 2320535 first doses had been administered reported 27 cases within 31 days of AstraZeneca vaccine recipients, giving a rate of `r round(1E6 * 27 / (365.25 * (82511.94 + 33851.19 + 19491.37 + 14858.32) / 31))` per million doses within 31 days of vaccination [@schulz_covid19_2021].  In those aged under 60 years the rate in the German study was 18 per million doses, compared with 7 per million doses for the same age group and 28-day time interval.  The reasons for this difference between the UK and EU countries are not clear.   \hl{Process-related impurities have been reported in batches of the AstraZeneca vaccine} [@krutzke_processrelated_2021].  A report from the European Medicines Agency dated 24 March 2021 stated that they had requested AstraZeneca to provide "a full batch analysis for specific lots and batch data from UK supplied lots to understand if there are any clear differences between that and the EU products" [@pharmacovigilance_risk_assessment_committee_signal_2021].

## Policy implications
This study based on complete ascertainment of CVT cases makes it possible to set a definitive upper bound on the rate of vaccine-associated CVT in Scotland.  The results reinforce the importance of establishing comprehensive surveillance of adverse events occurring after vaccination.  By using e-health record systems we were able to obtain all neuroimaging reports for the population and to report preliminary results to public health agencies within a few weeks of initiating this study.  This entailed labour-intensive manual coding of scan reports.  For longer-term surveillance and scaling to larger populations a natural language processing algorithm could be developed to identify CVTs in imaging reports, but some manual coding would still be required especially to assign the date of onset.

Policy on the continued use of the AstraZeneca vaccine has been driven by estimates of the risk/benefit ratio, with risk of TTS estimated from adverse event reporting schemes  [@european_technical_advisory_group_of_experts_on_immunization_adhoc_2021].  Thus on 7 May 2021 the Joint Committee on Vaccination and Immunisation  advised that those under age 40 should be offered an alternative to the  AstraZeneca vaccine [@department_of_health_and_social_care_use_2021].  Evaluating the risk / benefit ratio of COVID-19 vaccination in healthy young adults and children depends on being able to detect rare adverse events post-vaccination through surveillance, so that the risk of such events can be compared with the low risk of severe complications of COVID-19 in these groups.

# Declarations
## Public and Patient Involvement statement:
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives.

## Ethics approval and information governance
This study was performed within Public Health Scotland as part of its statutory duty to monitor and investigate public health problems.  Under the [UK Policy Framework for Health and Social Care Research](https://www.hra.nhs.uk/planning-and-improving-research/policies-standards-legislation/uk-policy-framework-health-social-care-research/) set out by the NHS Health Research Authority, this does not fall within the definition of research and ethical review is not required.  Individual consent is not required for Public Health Scotland staff to process personal data to perform specific tasks in the public interest that fall within its statutory role.  The statutory basis for this is set out in Public Health Scotland's [privacy notice](https://www.publichealthscotland.scot/our-privacy-notice/personal-data-processing/).  
A Data Protection Impact Assessment​ (DPIA) allows Public Health Scotland staff to link existing datasets.  This study was approved under COVID-19 Rapid DPIA 20210023.

## Transparency declaration
HC as the manuscript’s guarantor affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained. This manuscript has been generated directly from the source data by a reproducible research pipeline.

## Funding 
No specific funding was received for this study.

## Data Availability
The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.

## Competing interest
All authors have completed and submitted the ICMJE Form for Disclosure of Potential Conflicts of Interest.

## Contributors
PM, RB, DM and HC provided substantial contributions to the conception or design and drafting of the manuscript.  JB provided substantial contributions to the data acquisition. CR, JM and MO provided substantial contributions to the interpretation. All authors contributed to revising the manuscript critically for important intellectual content and approved the final manuscript.

## Acknowledgements
We thank Matteo Abideni (Philips Ltd), Alan Fleming (National Services Scotland), and Kenny Scott (National Services Scotland) for help with querying the PACS database

# References
<div id="refs"></div>

\newpage

```{r scansbydate, echo=FALSE, message=FALSE, fig.asp=1.2, fig.width=5, fig.cap="\\label{fig:scansbydate}(a) Venograms in the 11 PACS boards by 7-day sliding window of dates; (b) Other head scans; (c) Population recently exposed: millions vaccinated in last 14 days"}
gridExtra::grid.arrange(p.veno + theme(axis.title.x = element_blank()),
                        p.nonveno + xlab("Mid-point of 7-day sliding window of study date"),
                        p.vaxdates + theme(legend.position = c(0.2, 0.7)) +
                        theme(legend.title = element_blank()), 
                        nrow=3)
```

\newpage

```{r events, echo=FALSE, message=FALSE}

toupper.first <- function(x){
    paste0(toupper(substring(x, 1, 1)),
          substring(x, 2, nchar(x)))
}

numevents <- scans.events[study_date >= as.Date("2020-12-01"), .(count=.N, hfeventcode), by=c("chi", "birth_date", "sex")]
setorder(scans.events, chi, study_date)
#scans.events[chi %in% numevents[count==2, chi],
#             .(chi, sex, birth_date, study_date, hfeventcode, hfeventonset)]

table.events <- with(numevents, table(hfeventcode, count, exclude=NULL))
table.events <- cbind(table.events, rowSums(table.events)) 
table.events <- rbind(table.events, colSums(table.events))
rownames(table.events) <- toupper.first(rownames(table.events))
rownames(table.events)[nrow(table.events)] <- "All"
colnames(table.events)[ncol(table.events)] <- "All"

knitr::kable(table.events[, 1],
    booktabs=TRUE, 
	label="scansevents",
    col.names="Number of individuals",
    caption="Scoring of CVST diagnostic categories on 1605 individuals with scans after 1 December 2020 that were filtered by study description or text strings in report for manual review ") %>%
    kable_styling(latex_options=c("hold_position"))

```

```{r dischbyscan, echo=FALSE, message=FALSE}
# cvst.leftjoin has 32 records
cvst.leftjoin[, maindiag := ifelse(grepl("G08X|I636|I676", MAIN_CONDITION), 
"As main condition", "As other condition")]

disch.byscan <- with(cvst.leftjoin[ADMISSION_DATE >= as.Date("2020-12-01")],
                    table(maindiag, hfeventcode, exclude=NULL))
disch.byscan <- disch.byscan[, colSums(disch.byscan) > 0]
disch.byscan[, 2] <- disch.byscan[, 2] + disch.byscan[, 3] + + disch.byscan[, 4]
disch.byscan <- disch.byscan[, c(1, 2, 5)]
disch.byscan <- rbind(disch.byscan, colSums(disch.byscan))
rownames(disch.byscan)[nrow(disch.byscan)] <- "Total" 
disch.byscan <- cbind(disch.byscan, rowSums(disch.byscan))
colnames(disch.byscan) <- c("Primary acute", "Chronic / secondary / negative", "No scan", "All")

knitr::kable(disch.byscan, 
             booktabs=TRUE,
             escape=FALSE,
             label="dischbyscan",
             caption=paste(disch.byscan["Total", 4], 
                           "SMR01 discharges with CVT diagnoses from 1 December 2020 by scan report")) %>%
    add_footnote("Ascertainment based on mention of CVT codes G08, I63.6, I67.6 excluding records with mention of codes for secondary causes as described in Methods",
                 notation="none") %>%
    kable_styling(latex_options=c("hold_position"))

```
```{r scansbysdisch, echo=FALSE, message=FALSE}
scans.bydisch <- with(table.scans, 
table(paste0(lubridate::year(study_date), "-Q", 
              ceiling(lubridate::month(study_date) / 3)), smr.status, exclude=NULL))
scans.bydisch <- rbind(scans.bydisch, colSums(scans.bydisch))
rownames(scans.bydisch)[nrow(scans.bydisch)] <- "Total"
scans.bydisch <- cbind(scans.bydisch, rowSums(scans.bydisch))
colnames(scans.bydisch)[ncol(scans.bydisch)] <- "All"

knitr::kable(scans.bydisch, 
             booktabs=TRUE,
             escape=FALSE,
             label="scansbydisch", 
             caption=paste(scans.bydisch["Total", 4], 
                           "scans coded as primary acute by 3-month calendar period and mention of a CVT code on discharge record")) %>%
    kable_styling(latex_options=c("hold_position"))

```

```{r vaxbyage, echo=FALSE, message=FALSE}

knitr::kable(table.vax.byage, 
             booktabs=TRUE, 
             escape=FALSE,
             align=rep("r", 5),
             label="vaxbyage", 
             caption=paste("Vaccine doses by age group up to",
                           max(vaxpop$vaxdate, na.rm=TRUE))) %>%
    kable_styling(latex_options=c("hold_position"))

```

```{r timewin, echo=FALSE, message=FALSE}

knitr::kable(table.tw, 
             booktabs=TRUE, 
             label="tw",
             row.names=FALSE,
             align=c("l", rep("r", 3), "c"),
             col.names=c("Product", "1-14", "15-28", "29-84", "1-14 vs 15-84"),
             caption="CVT cases after vaccination, by time since vaccine dose") %>%
    group_rows("Person-fortnights exposed", 1, 3) %>% 
    group_rows("Cases scored as primary acute", 4, 6) %>% 
    group_rows("Rate (95% credible interval) primary acute per million person-fortnights", 7, 9, bold=FALSE) %>%
    group_rows("Cases scored as primary acute or possible", 10, 12) %>% 
    group_rows("Rate (95% credible interval) primary acute or possible per million person-fortnights", 13, 15, bold=FALSE) %>%
    add_header_above(c(" ", "Days since vaccination"=3, "Rate ratio")) %>% 
    kable_styling(latex_options=c("hold_position"))

```

\newpage
\beginsupplement

# Supplementary Material

```{r fig1, out.width="0.7\\linewidth", include=TRUE, fig.align="center", fig.cap="\\label{flow}Flow diagram for case ascertainment from scan reports", echo=FALSE}

knitr::include_graphics("./ctv_flow.png")

```

\newpage

```{r venodesc, echo=FALSE, message=FALSE}
if(FALSE) {
veno.ris <- unique(scans.pacs[grep("veno", STUDY_DESCRIPTION, ignore.case=TRUE),
                       .(RIS_ID, STUDY_DESCRIPTION)])
setorder(veno.ris)
knitr::kable(veno.ris, 
             booktabs=TRUE, 
			 label="risveno", 
			 row.names=FALSE,
			 caption="Venogram codes used to query PACS")
}

```

## PACS query for ascertainment of scans possibly informative for CVT

* (Modality=CT OR Modality=MR) AND (Study description contains "head" or "cerebr*" or "brain")

**OR**

* RIS code = ANY OF

  - CVENO CT Venogram
  - CVEIC CT Venogram Intracranial
  - CVECE CT Venogram cerebral
  - CTHVG CT Head Venogram
  - MVENO MRV Venogram
  - CBNTA CT Brain neck thorax Abdo and pelvis 
  - CHAP CT Head abdomen and pelvis
  - CHAPC CT Head abdomen pelvis with contrast
  - CHNTAP CT Head neck thorax abdomen and pelvis
  - CHTA CT Head thorax and abdomen
  - CHTAP CT Head thorax abdomen and pelvis
  - CHTAPC CT Head thorax Abdo pelvis with contrast
  - CHTH CT Head and thorax
  - CHTHAC CT Head thorax abdomen with contrast
  - CHTHC CT Head and thorax with contast
  - CSKNE CT Head and neck
  - CSKPE CT Brain perfusion study
  - CSKUH CT Head
  - CSKUHC CT Head with contrast
  - CSKUC CT Head with contrast
  - MAICA MRA Head
  - MAICAC MRA Head with contrast
  - MBRCC MRI Brain and cervical cord
  - MSKPE MRI Head brain perfusion study
  - MSKUH MRI Head
  - MSKUHC MRI Head with contrast
  - MSKUS MRI Head spectroscopy
  - MVSKU MRV Cerebral veins

## Protocol for scoring neuroimaging studies
### Scoring at scan level
* **negative** -  no evidence of any venous sinus thrombosis in the report

* **secondary** - evidence of acute or chronic CVT but there is another local pathology contributing such as meningioma, mastoiditis, sinusitis or encephalitis.  Systemic factors such as Factor V Leiden deficiency,  post-partum state, use of oral contraceptives, or pro-thrombotic drugs are not classified as secondary causes.  Trauma caused by a external event such as traffic accident or assault is classified as a secondary cause, but history of a fall that could have resulted from a primary brain event is not classified as a secondary cause.

* **chronic** - this code is assigned when there is a new presentation but the scan shows evidence of unchanged or resolving thrombosis or recanalising thrombosis.  There may or may not be a definitive history of an earlier acute primary event.

* **follow-up** - this code is assigned where the scan shows evidence of thrombosis but has been done as a follow up of a prior event either during the same hospitalisation or later as a routine follow-up. This code was assigned to reports without the primary scan necessarily being available at time of report coding.

* **possible** - this code is used where there is some evidence that may suggest thrombosis but which is not definitive; many such reports will recommend further investigation

* **primary acute** -  scan reported as consistent with venous thrombosis, not assigned as chronic and which cannot definitely be assigned  to  other local  cause.

* **no valid result** - there is a report but the report text is missing or the scan was declared to be a technical failure or unreadable

### Coding at person level
For this process all available scans extracted for an individual were arrayed at person level. An event  encompasses all scans pertaining to a new presentation. A follow up scan done several months or years later is a new event.

* **negative** - none of the scans have found evidence of a  CVT or an initial scan with possible CVT is   followed by a definitive scan such as venogram that rules out  CVT

* **secondary** - an event where there is evidence of acute or chronic CVT but there is another local brain pathology contributing such as meningioma or mastoiditis or sinusitis or encephalitis

* **chronic** - there is a new presentation and one of the scans shows chronic thrombus

* **follow-up** - Most events that include follow up scans during that admission will be coded as primary acute events with the date of the event being that of the  presentation date of the originating thrombus event.  For follow up scans in a separate hospital attendance these are coded as follow up but the  event onset date is given as the original primary event date where it is available. Where the primary event is not referred to or given a date the final event code maybe left as follow-up with the date of onset left blank

* **possible** - one of the scans for the event found evidence of a possible CVT and no subsequent scan ruled this out or resolved whether or not a thrombus was present

* **primary acute** – a new presentation where any scan shows changes consistent with  thrombus that is not assigned as chronic and is not attributed to a secondary local cause

### Examples of how the protocol is applied

1. Patient presents with a seizure – all scans are negative. Two weeks later they present with new onset confusion – CT scan  negative, MR venogram shows bleed and CVT .  The first admission is coded as negative  for each scan and for the event.  The second set of scans is coded as negative and primary acute  respectively and the event is coded as primary acute.

2. Patient presents with confusion and headache  after a fall – CT head  has no abnormality , subsequent CT venogram shows filling  defect in venous sinus – a scan three days later shows similar picture.  The reports are coded as negative, primary acute, follow up.  The event is coded as primary acute.

```{r dischbydate, echo=FALSE, message=FALSE}

if(FALSE) {
knitr::kable(table.byyear, 
             booktabs=TRUE,
             escape=FALSE,
             label="dischbydate", 
             caption="SMR01 discharges with CVT diagnoses by year") %>%
    kable_styling(latex_options=c("hold_position"))
}

```

```{r disch, echo=FALSE, message=FALSE}
if(FALSE) {
colnames(table.disch) <- c("Admission date", "Scan date", "Site", "Sex", "Age", "Main diagnosis", "Scan report", "Scan check")
knitr::kable(table.disch, 
             booktabs=TRUE,
             escape=TRUE,
             label="CVTdisch", 
             caption=paste(nrow(table.disch),
                           "SMR01 discharges admitted from 1 December 2020 to",
						   max(table.disch[['Admission date']]))) %>%
    kable_styling(latex_options=c("hold_position", "scale_down"))
}

```

```{r scans, echo=FALSE, message=FALSE}
if(FALSE) {
table.scans[, agegr20 := floor(age.scan / 20)]
table.scans[, agegr20 := car::recode(agegr20,
                                 "0='0-19'; 1='20-39'; 2='40-59'; 3='60-79'; 4='80-'")]

knitr::kable(table.scans[, .(ADMISSION_DATE, study_date, site_name, sex, agegr20, 
                             MAIN_CONDITION,  smr.status)], 
             booktabs=TRUE,
             escape=TRUE,
             label="CVTscans", 
			 col.names=c("Admission date", "Scan date", "Site", "Sex", "Age", "Main diagnosis",  "Discharge coding"),
             caption=paste(nrow(table.scans),
                           "scans from from 1 December 2020 scored as primary acute, left joined to SMR01 discharges")) %>%
    kable_styling(latex_options=c("hold_position", "scale_down"))
}

```

```{r allcvst, echo=FALSE, message=FALSE}

if(FALSE) {
allCVT <- cvst.all[(is.na(hfeventcode) | (hfeventcode=="primary acute")) &
                    dateseen >= as.Date("2020-12-01"),
                    .(dateseen, MAIN_CONDITION, 
                      study_date, SEX, age.seen, daysfromvax,
                      vacc_product_name)]
setorder(allCVT, dateseen)
allCVT[, monthonset := lubridate::month(dateseen, label=TRUE)]
allCVT[, agegr20 := floor(age.seen / 20)]
allCVT[, agegr20 := car::recode(agegr20,
                                 "0='0-19'; 1='20-39'; 2='40-59'; 3='60-79'; 4='80-'")]

yearmonth <- paste(lubridate::year(allCVT$dateseen),
                   month(allCVT$dateseen), sep="_")
uniquemonth <- sort(unique(yearmonth))
sep.at <- match(uniquemonth, yearmonth)

az.rows1 <- which(allCVT$daysfromvax < 15 & allCVT$vacc_product_name == "AstraZeneca")
az.rows2 <- which(allCVT$daysfromvax >=15 & allCVT$daysfromvax < 29 &
                  allCVT$vacc_product_name == "AstraZeneca")

allCVT <- allCVT[, .(dateseen, SEX, agegr20, daysfromvax, vacc_product_name)]
colnames(allCVT) <- c("Month of onset", "Sex", "Age", "Days from vax", "Product")
knitr::kable(as.data.frame(allCVT), 
             booktabs=TRUE, 
             label="allCVT",
             row.names=FALSE, 
             linesep=linesep(sep.at),
             caption=paste(nrow(allCVT),
                           "primary acute CVT cases ascertained through discharge diagnoses or scan reports since December 2020")) %>%
    row_spec(az.rows1, background = "yellow") %>% 
    row_spec(az.rows2, background = "orange") %>% 
    kable_styling(latex_options=c("hold_position"))
}

```
